### 2015-11-04 paper.Rmd
first hack at first 3 sections

---
author:
  - name: David L Miller
    affiliation: Centre for Research into Ecological and Environmental Modelling, University of St Andrews
    address: >
      The Observatory,
      St Andrews, Fife KY16 9LZ, Scotland
    email: dave@ninepointeightone.net
    url: http://converged.yt
title:
  formatted: "Distance Sampling in \\proglang{R}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Distance Sampling in R"
  # For running headers, if needed
  #short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [distance sampling, abundance estimation, line transects, point transects, "\\proglang{R}"]
  plain:     [distance sampling, abundance estimation, line transects, point transects, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
output: rticles::jss_article
bibliography: jstatsoft.bib
---

# Introduction

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(cache=TRUE)
```

Distance sampling [@Buckland:2001vm; @Buckland:2004ts] encompases a suite of field methods and statistical models used to estimate the abundance of biological populations. Distance sampling field procedure can be thought of as an extension of plot sampling, where we wish to take into account the decreasing probability of detecting objects at increasing distance from the sampler. We do this by building a model for detectability and use quantities calculated from the detection function to adjust the observed counts to obtain an estimate of abundance.

For many years distance sampling analyses have been available via the Windows program Distance (or "DISTANCE"); for clarity henceforth "Distance for Windows" @Thomas:2010cf). From version 5 of Distance for Windows, R packages have been included to perform particular analyses (CITE Distance user manual). This paper shows how to fit detection functions, perform model checking and selection and estimate abundance in \proglang{R} using the package \pkg{Distance2}.

## Distance sampling

It is clear that census-type surveys (quadrat or strip transects) are inefficient (requiring a lot of effort on the part of those in the field) and we should expect that not all animals can be observed. Accounting for imperfect detectability is an important consideration when want to obtain accurate estimates of abundance [@LahozMonfort:2013bg]. Using the extra information gained by recording distance from the sampler to the observation, it is possible to model detectability. Since we expect detectability to decrease with increasing distance from the sampler, we model detectability as a function of distance (plus perhaps other covariates, see below).

Distance sampling comes in two main "flavours": line and point transects. In line transect sampling observers walk (or fly, sail, etc) down lines observing objects and recording the distances to the line; whereas in point transect sampling observers stand stationary at a location and record distances from that point. Field methods are chosen to be suitable to species and habitat constraints [@buckland2015distance].

```{r points-and-lines, fig.height=6, fig.width=7, echo=FALSE, message=FALSE, fig.cap="Left side plots show an example of a survey of an area containing a population of 500 objects, blue indicates sampler placement (top lines, bottom points) and red dots indicate detected individuals. The right side of the figure shows histograms of observed distances (again, lines top and points bottom)."}
opar <- par(mfrow=c(2,2))
# line and point transect examples
# top row points/samplers in space
# bottom row histograms

# generate some animals to sample
set.seed(131) # same results every time
library(mgcv) # for inSide

# generate population locations
N <- 500
x <- runif(N)
y <- runif(N)

## line transects
opar2 <- par(mar=c(1, 1, 1, 1) + 0.1)
plot(x, y, pch=19, asp=1, cex=0.6, main="",col="grey", axes=FALSE, xlab="", ylab="")
polygon(x=c(0,0,1,1,0), y=c(0,1,1,0,0))
# generate some lines
# in this case we don't randomise the offset of the grid
lt <- list(x=c(-0.0125,-0.0125,0.0125,0.0125,-0.0125), y=c(0,1,1,0,0))
# set sigma
sigma <- 0.01
# storage for detected distances
detected_distances <- c()
for(i in 1:5){
  # calculate next strip location
  lt$x <- lt$x+0.15
  # plot the line transect
  lines(x=rep(mean(range(lt$x)),2), y=c(0,1), col="blue",lty=2)
  # calculate the distances to animals from the line
  distances <- abs(lt$x - x)
  # randomly decide which were detected
  detected <- exp(-distances^2/(2*sigma^2)) > runif(length(distances))
  # plot those animals detected
  points(x[detected], y[detected], pch=19, cex=0.6, col="red")
  # collect the distances to detected objects
  detected_distances <- c(detected_distances, distances[detected])
}

#par(opar2)
par(mar=c(4, 4, 1, 2) + 0.1)
hist(detected_distances, main="", xlab="Observed distances")

## point transects
par(mar=c(1, 1, 1, 1) + 0.1)
plot(x, y, pch=19, asp=1, cex=0.6, main="", col="grey", axes=FALSE, xlab="", ylab="")
polygon(x=c(0,0,1,1,0), y=c(0,1,1,0,0))

# set sigma
sigma <- 0.05
# storage for detected distances
detected_distances <- c()
# lay out a grid of points
pt <- as.list(expand.grid(x=seq(0.15, 0.85, len=3), y=seq(0.15, 0.85, len=3)))
for(i in 1:length(pt$x)){
  # generate point location
  # plot the line transect
  points(pt$x[i], pt$y[i], pch=19, col="blue",cex=0.7)
  # calculate the distances to animals from the line
  distances <- sqrt((pt$x[i] - x)^2+(pt$y[i]-y)^2)
  # randomly decide which were detected
  detected <- exp(-distances^2/(2*sigma^2)) > runif(length(distances))
  # plot those animals detected
  points(x[detected], y[detected], pch=19, cex=0.6, col="red")
  # collect the distances to detected objects
  detected_distances <- c(detected_distances, distances[detected])
}

par(mar=c(4, 4, 1, 2) + 0.1)
hist(detected_distances, main="", xlab="Observed distances")

par(opar)

```

For both points and lines, once the geometry of the sampler has been taken into account, the histogram of distances should show a decreasing number of observations with increasing distance from the sampler. For line transects we expect objects to be distributed unformly with respect to distance from the line and what makes our histogram decrease is the detectability. For point transects, we must take into account the fact that as distance from the point increases, the area of the circle encompassed increases with distance squared.

Using this histogram we can crudely estimate the drop-off in detectability by eye by tracing a line that approximates the tops of the histogram bars -- this is the detection function. \pkg{Distance2} estimates the parameters for a fixed-form detection function using maximum likelihood estimation. We address possible models in detail below

## Data

We demonstrate \pkg{Distance2} using two data sets: one line transect and one point transect. These data sets have been chosen to be representative of the kind of data seen in practice.

### Minke whales

The line transect data is simulated data based on a survey of Antarctic minke whales (*Balaenoptera bonaerensis*). The data is simulated from models fitted to data from the International Whaling Commission’s International Decade of Cetacean Research Southern Ocean Whale and Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer surveys [@Branch:2001ua]. They consist of 99 observations and include information on the effort expended and whether observations were in one of two geographical strata (near or far from land).

### Amakihi

The point transect data set consists of 1485 observations of Amakihi (*Hemignathus virens*; a Hawaiian songbird), collected at 41 points between 1992 and 1995. The data include distances and three covariates collected during the survey: observer ID (a three level factor), minutes after sunrise (continuous) and hours after sunrise (a six level factor). Data are analysed comprehensively in @Marques:2007ey.

## The rest of the paper

The rest of the paper is structured as follows: we first look into how data needs to be organised to allow it to be used with \pkg{Distance2}; models for the detection function are then covered including their formulation and examples of fitting in \proglang{R}. We then spend some time looking at model checking and goodness of fit testing, as well as model selection. Having illustrated how to obtain a good detection function, we show how to estimate abundance using that model, including using post-stratification. The final two sections of the article look at extensions (both in terms of methodology and software) and put the package in a broader context amongst other software packages used for animal abundance estimation.

# Data setup

The two example data sets used here are distributed with \pkg{Distance2} so readers can reproduce our results easiy. However in general we expect that data will be collected in the field and need to be formatted correctly for use with \pkg{Distance2}. The package allows for a flexible format for data input ranging from very simple to complex:

* In the simplest case, where one would simply like to estimate a detection function, all that is needed is a vector of distances.
* To include additional covariates into the detection function (see "Detection functions") we need to use a \code{data.frame}, this needs to have a column called \code{distance} and named columns for each covariate collected (for example \code{observer} or \code{seastate}). Some column names are reserved: \code{object} for an observation identifier (see "Extensions"), \code{size} for group or cluster size (see "Detection functions" and "Abundance and variance estimation"), \code{detected} for whether or not an observation was detected (see "Extensions") and the columns described in the next bullet.
* If one would like to estimate abundance for some area, this information must be given at the same time. This consists of information about which transect and occasion the observation was made (the \code{Sample.Label}), a column named \code{Effort} which gives the effort associated with that sample (for lines their length and for points the number of times that point was visited), the stratum the sample was located in (this may have any name and may be from pre- or post-survey stratification, see "Estimating abundance and variance") and that stratum's area (which has the same name as the stratum column, appended with \code{.Area}). (We refer to this data format as "flatfile" is all information we have is contained in one table.)

As we will see in "Extensions", further information is also required when we start using more complex models.

Looking at an example of the data format given in the last bullet, we can examine the minke whale data:

```{r minke-data-head}
library(Distance2)
head(minke)
```
and the amakihi data:
```{r amakihi-data-head}
head(amakihi)
```

* what do we want to say here?

# Detection functions

As mentioned above, the detection function describes the relationship between observed distances and probability of detection. The detection function itself models the probability $\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)$ and is usually denoted $g(y; \boldsymbol{\theta})$ where $y$ is distance and $\boldsymbol{\theta}$ is a vector of parmeters to be estimated. Our goal is to estimate an average probability of detection (average in the sense of an average over the distances), so we must integrate out distance ($y$) from the detection function:
$$
p = \int_0^w \pi(y) g(y; \boldsymbol{\theta}) dy
$$
where $\pi(y)$ describes the distribution of objects with respect to the sampler and is $1/w$ for line transects and $\frac{2r}{w}$ for point transects [@Buckland:2001vm, Chapter 3].

Models for the detection function are expected to have the following properties [@buckland2015distance, Chapter 5]:

* *Shoulder*: we expect observers to be able to see objects near them, not just those directly infront of them. For this reason, we expect the detection function to be "flat" near zero distance.
* *Non-increasing*: we don't think that observers should be more likely to see things further away than those nearer to them. This usually indicates an issue with field procedure (that the distribution of animals with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this.
* *Model robust*: models should be flexible enough to make many different shapes.
* *Pooling robust*: many factors can affect the probability of detection and it is not possible to measure all of these. We would like our models to be robust to us not including these factors.
* *Estimator efficiency*: we would like our models to have low variances, but only given they satisfy the other properties above (which, if they are satisfied, would give low bias).

There is a wide literature on possible formulations for the detection function [@Buckland:1992fa; @Eidous:2005bj; @Becker:2009cj; @Giammarino:2014eg; @Miller:2015hw; @Becker:2015fi]. \code{Distance2} includes the most popular of these models, and includes an extendable class system to add new detection functions while avoiding code duplication (see "Extensions").

Here we'll show two popular detection function approaches: "key function plus adjustments" (K+A) and mixture models.

## Detection functions: key function plus adjustments (K+A)

Key function plus an adjustments (or adjustment series) models are formulated by taking a "key" function and optionally adding "adjustments" to it to improve the fit. Mathematically we formulate this as:
$$
g(y; \boldsymbol{\theta}) = k(y; \boldsymbol{\theta}_k)\left( 1+ \alpha_K(y; \boldsymbol{\theta}_\alpha)\right),
$$
where $k$ is the key function and $\alpha_K$ is sum series of functions and the subscripts on the parameter vector indicate those parameters belonging to each part of the model.

Models for $k$ are as follows
$$
k(y) = \left\{
\begin{array}{l l}
  \exp\left(-\frac{y^2}{2 \sigma^2}\right) & \quad \text{half-normal} \\
  1-\exp\left(\left(-\frac{y}{\sigma}\right)^{-b}\right) & \quad \text{hazard-rate} \\
  1/w & \quad \text{uniform}
\end{array} \right.
$$
Possible modelling options for key and adjustments are given in Table \ref{ka-tab} and illustrated in **Figure ???**.

```{r hn-hr-par-comp, fig.width=10, fig.height=6.5, fig.cap="Half-normal (top row) and hazard-rate (bottom row) detection functions without adjustments, varying scale ($\\sigma$) and (for hazard-rate) shape ($b$) parameters (values are given above the plots). On the top row from left to right, the study species becomes more detectable (higher probability of detection at larger distances). The bottom row shows the hazard-rate model's more pronounced shoulder.", echo=FALSE}
par(mfrow=c(2,4), mar=c(3.5, 3, 2, 1) + 0.1)

## half-normal
g.hn <- function(x,sigma) exp(-x^2/(2*sigma^2))
x <- seq(0,1,len=200)
for(this.sig in c(0.05, 0.25, 1, 10)){
  plot(x, g.hn(x,this.sig), main=bquote(sigma == .(this.sig)),
       xlab="", ylab="",
       type="l", xlim=c(0,1), ylim=c(0,1), asp=1)
  title(xlab="Distance", line=2)
  title(ylab="Detection probability", line=2)
}

## hazard-rate
g.hr <- function(x, sigma, b) 1 - exp(-(x/sigma)^-b)
for(this.sig in c(0.1, 0.5)){
  for(this.b in c(5, 1)){
    plot(x, g.hr(x, this.sig, this.b),
         main=bquote(sigma == .(this.sig) ~ .(", b") == .(this.b)),
         xlab="", ylab="",
         type="l", xlim=c(0,1), ylim=c(0,1), asp=1)
    title(xlab="Distance", line=2)
    title(ylab="Detection probability", line=2)
  }
}

```

\begin{table}
\caption{Modelling options for key plus adjustments models for the detection function.}
\begin{tabular}{llll}
\hline
key function   & form   & adjustment & form\\
\hline
 uniform  & $1/w$   & cosine  & $\sum_{k=1}^K a_k \cos(k \pi y/w)$ \\
 & & Simple polynomial & $\sum_{k=1}^K a_k (y/w)^{2k}$ \\
 half-normal  & $\exp\left(-\frac{y^2}{2 \sigma^2}\right)$ & cosine  & $\sum_{k=2}^K a_k \cos(k \pi y/w)$ \\
 & & Hermite polynomial & $\sum_{k=2}^K a_k H_{2k}(y/\sigma)$ \\
 hazard-rate  & $1-\exp\left[-\left(\frac{y}{\sigma}\right)^{-b}\right]$ & cosine  & $\sum_{k=2}^K a_k \cos(k \pi y/w)$ \\
 & & Simple polynomial & $\sum_{k=2}^K a_k (y/w)^{2k}$ \\
\hline
\end{tabular}
\label{ka-tab}
\end{table}

When adjustment terms are used it may be necessary to standardise the results to ensure that $g(0)=1$, so we can redefine the detection function as:
$$
g(y; \boldsymbol{\theta}) = \frac{k(y; \boldsymbol{\theta})\left( 1+ \alpha_K(y; \boldsymbol{\theta})\right)}{k(0; \boldsymbol{\theta})\left( 1+ \alpha_K(0; \boldsymbol{\theta})\right)}
$$
A distinct disadvantage of K+A models is that we must resort to constrained optimisation (via the \pkg{Rsolnp} package) in order to ensure that the resulting detection function is monotonic non-increasing over the whole range.

We do not always use adjustments (except in the case of the uniform key), in which case we refer to "key only" models; see also "Covariates" and "Model checking and model selection" below.

## Detection functions: mixture models

We can also formulate the detection function as a mixture of (say) half-normal functions [@Miller:2015hw]:
$$
g(y; \boldsymbol{\theta}) = \sum_{k=1}^K \varphi_j \exp\left(-\frac{y^2}{2 \sigma_j^2}\right)
$$
where $\sum_{k=1}^K \varphi_j =1$ and $\boldsymbol{\theta} = (\varphi_1,\ldots,\varphi_{K-1},\sigma_1,\ldots,\sigma_K$. Such models do not require constrained optimisation (as K+A models do) as the mixtures are monotonic decreasing by construction.

As with the K+A model, we describe model selection (in terms of number of mixture components) in "Model checking and selection" below.

## Covariates

As mentioned when we talked about pooling robustness, above, there are many factors that can affect the probability of detecting an object. These include things like the observer, the vessel or platform used, the sea state, weather conditions and time of day (to name but a few). We assume that these variables affect detection only via the scale of the detection function (and that they don't affect the shape).

Covariates can be included in this formulation by considering the scale parameter from the half-normal or hazard-rate detection functions as a linear model (on the exponential scale) of the ($k$) covariates ($\mathbf{z}$):
$$
\sigma(\mathbf{z}) = \exp(\beta_0 + \sum_k \beta_k z_k).
$$
In the next section we'll discuss model selection for these models.

Including covariates has an important implication for our calculation of detectability. Since we don't know what the true distribution of the covariates is we must calculate the probability of detection conditional on the observed values of the covariates:
$$
p(\mathbf{z_i}) = \int_0^w \pi(y) g(y, \mathbf{z_i}; \boldsymbol{\theta}) dy
$$
so we now calculate a value of "average" probability of detection (again the average is in the sense of distance) per observation. There will be as many unique values of $p(\mathbf{z_i})$ as there are unique covariate combinations in our data.

Another important consideration is that K+A models that include covariates and one or more adjustments cannot be guaranteed to be monotonic non-increasing for all covariate combinations, as we don't have any model for the distribution of the covariates. For this reason, we advise against using both adjustments and covariates in a detection function [see @Miller:2015hw for an example of when this can be problematic].

# Model checking and model selection

* qq plots
* hessian/convergence
* monotonicity
* AIC -- what else?

# Estimating abundance and variance

* Horvitz-Thompson
* Refer to Fewster et al 2009

# Extensions

* writing new detection functions
* Interface to dsm etc

# Conclusion

* Other software? \pkg{unmarked}, \pkg{RDistance} etc

### 2015-11-04 make.sh

@@ -1,0 +1,4 @@
+#!/bin/sh
+
+R -e 'rmarkdown::render("paper.Rmd")'
+open paper.pdf

### 2015-11-04 paper.Rmd
plots...

@@
 
-```{r minke-data-head}
+```{r minke-data-head, message=FALSE}
 library(Distance2)
@@
 
+It is important that the detection function accurately models detectability near zero distance.
+We are less worried by its behaviour further away from 0.
+To ensure that the model is not overly influence by distances far from zero we truncate the distances beyond a given distance $w$ (known as the *truncation distance*).
+Further details on truncation can be found in **CITATION**.
+
 Models for the detection function are expected to have the following properties [@buckland2015distance, Chapter 5]:
@@
 
+Given these criteria, we can start to think about models for $g$.
+
+## Formulations
+
 There is a wide literature on possible formulations for the detection function [@Buckland:1992fa; @Eidous:2005bj; @Becker:2009cj; @Giammarino:2014eg; @Miller:2015hw; @Becker:2015fi].
@@
 
-## Detection functions: key function plus adjustments (K+A)
+### Key function plus adjustments (K+A)
 
@@
 g.hn <- function(x,sigma) exp(-x^2/(2*sigma^2))
-x <- seq(0,1,len=200)
 for(this.sig in c(0.05, 0.25, 1, 10)){
-  plot(x, g.hn(x,this.sig), main=bquote(sigma == .(this.sig)),
-       xlab="", ylab="",
-       type="l", xlim=c(0,1), ylim=c(0,1), asp=1)
+  this.g <- function(x) g.hn(x, sigma=this.sig)
+  curve(this.g, from=0, to=1, xlab="", ylab="",
+        main=bquote(sigma == .(this.sig)),
+        xlim=c(0,1), ylim=c(0,1), asp=1)
   title(xlab="Distance", line=2)
@@
   for(this.b in c(5, 1)){
-    plot(x, g.hr(x, this.sig, this.b),
+    this.g <- function(x) g.hr(x, sigma=this.sig, b=this.b)
+    curve(this.g, from=0, to=1, xlab="", ylab="",
           main=bquote(sigma == .(this.sig) ~ .(", b") == .(this.b)),
-         xlab="", ylab="",
-         type="l", xlim=c(0,1), ylim=c(0,1), asp=1)
+          xlim=c(0,1), ylim=c(0,1), asp=1)
     title(xlab="Distance", line=2)
@@
 
-## Detection functions: mixture models
+```{r adjust-mix-comp, fig.width=10, fig.height=5, fig.cap="Top row shows possible shapes for the detection function when adjustments are included for half-normal and hazard-rate models.
+Bottom row shows possible shapes when 2-component mixtures of half-normal functions are used.
+Notation is as in main text.", echo=FALSE}
+par(mfrow=c(2,4), mar=c(3.5, 3, 2, 1) + 0.1)
+
+## half-normal with cosine
+g <- function(x) exp(-x^2/(2*0.01^2))*(1+0.5*cos((2*pi*x)/0.025))
+f <- function(x) g(x)/g(0)
+curve(f, from=0, to=0.025, xlab="", ylab="", main="Half-normal with 1 cosine adjustment",xlim=c(0,0.025), ylim=c(0,1))
+title(xlab="Distance", line=2)
+title(ylab="Detection probability", line=2)
+# 2 cosines
+g <- function(x) exp(-x^2/(2*0.01^2))*(1-0.06*cos((2*pi*x)/0.025)+0.25*cos((3*pi*x)/0.025))
+f <- function(x) g(x)/g(0)
+curve(f, from=0, to=0.025, xlab="", ylab="", main="Half-normal with 2 cosine adjustments",xlim=c(0,0.025), ylim=c(0,1))
+title(xlab="Distance", line=2)
+title(ylab="Detection probability", line=2)
+
+## hazard-rate with cosine
+g <- function(x) (1 - exp(-(x/0.005)^-1.1))*(1-0.05*cos((2*pi*x)/0.025))
+f <- function(x) g(x)/g(0)
+curve(f, from=0, to=0.025, xlab="", ylab="", main="Hazard-rate with 1 cosine adjustment",xlim=c(0,0.025), ylim=c(0,1))
+title(xlab="Distance", line=2)
+title(ylab="Detection probability", line=2)
+# 2 cosines
+g <- function(x) (1 - exp(-(x/0.005)^-1.9))*
+                  (1+0.05*cos((2*pi*x)/0.025) + 0.25*cos((3*pi*x)/0.025))
+f <- function(x) g(x)/g(0)
+curve(f, from=0, to=0.025, xlab="", ylab="", main="Hazard-rate with 2 cosine adjustments",xlim=c(0,0.025), ylim=c(0,1))
+title(xlab="Distance", line=2)
+title(ylab="Detection probability", line=2)
+
+# mixture models
+plotmix <- function(sig1, sig2, mix){
+
+  gmix <- function(x) mix*g.hn(x, sig1) + (1-mix)*g.hn(x,sig2)
+
+  curve(gmix, from=0, to=1, xlab="", ylab="",
+        main=bquote(sigma[1] == .(sig1) ~ sigma[2] == .(sig2) ~  phi[1] == .(mix)),
+        xlim=c(0,1), ylim=c(0,1), asp=1)
+  title(xlab="Distance", line=2)
+  title(ylab="Detection probability", line=2)
+}
+plotmix(0.8, 0.15, 0.3)
+plotmix(0.6, 0.1,0.7)
+plotmix(10, 0.2, 0.15)
+plotmix(0.7, 0.05, 0.6)
+
+```
+
+### Mixture models
 
@@
 
-## Covariates
+### Covariates
 
@@
 
+## Fitting detection functions in R
+
+The workhorse of detection function fitting in \pkg{Distance2} is the \code{ds} function.
+Here we show off the formulations for the detection function that we've seen above for both the minke whale and amakihi data.
+
+We can fit a model to the minke whale data, setting the truncation at 1.5km and using the default options in \code{ds} very simply:
+```{r minke-hn}
+minke_hn <- ds(minke, truncation=1.5)
+```
+**Figure???** shows the result of calling \code{plot} on the resulting model object.
+
+```{r minke-hn-plot, fig.width=4, fig.height=4, fig.cap="Plot of fitted detection function overlayed on the histogram of observed distances for the minke whale data."}
+plot(minke_hn)
+```
+
 # Model checking and model selection
@@
 * Interface to dsm etc
+* MRDS?
+* why is + important?
 

### 2015-11-04 paper.Rmd
starting to show analyses

@@
 
+### Minke whale
+
 We can fit a model to the minke whale data, setting the truncation at 1.5km and using the default options in \code{ds} very simply:
@@
 **Figure???** shows the result of calling \code{plot} on the resulting model object.
+We can also call \code{summary} on the model object to get summary information about the fitted model (though we postpone this until the next section).
 
@@
 
+We can specify the form of the detection function via the \code{model=df(...)} argument to \code{ds}. For example, fitting a hazard rate model with cosine adjustments of order 2 can be specified as:
+```{r minke-hr-cos}
+minke_hrcos <- ds(minke, truncation=1.5, model=df(model=~hr+cos(2)))
+```
+
+### Amakihi
+
+By default \code{ds} assumes that the data given to it is line transect, but we can switch to points using \code{transect="points"}.
+Including covariates in the scale is via the `scale=~...` argument to `df`; a hazard-rate model for the amakihi that includes observer as a covariate (and truncating at 82.5m, given in @Marques:2007ey) can be specified by:
+```{r amakihi-hr}
+amakihi_hr_obs <- ds(amakihi, truncation=82.5, model=df(model=~hr, scale=~obs), transect="point")
+```
+
 # Model checking and model selection
 
+```{r minke-hn-summary}
+summary(minke_hn)
+```
+
+* summary results
 * qq plots

### 2015-11-04 paper.Rmd
more plotting

@@
 
-```{r hn-hr-par-comp, fig.width=10, fig.height=6.5, fig.cap="Half-normal (top row) and hazard-rate (bottom row) detection functions without adjustments, varying scale ($\\sigma$) and (for hazard-rate) shape ($b$) parameters (values are given above the plots).
+```{r hn-hr-par-comp, fig.width=10, fig.height=5, fig.cap="Half-normal (top row) and hazard-rate (bottom row) detection functions without adjustments, varying scale ($\\sigma$) and (for hazard-rate) shape ($b$) parameters (values are given above the plots).
 On the top row from left to right, the study species becomes more detectable (higher probability of detection at larger distances).
@@
 
-```{r minke-hn-plot, fig.width=4, fig.height=4, fig.cap="Plot of fitted detection function overlayed on the histogram of observed distances for the minke whale data."}
-plot(minke_hn)
-```
-
 We can specify the form of the detection function via the \code{model=df(...)} argument to \code{ds}. For example, fitting a hazard rate model with cosine adjustments of order 2 can be specified as:
@@
 Including covariates in the scale is via the `scale=~...` argument to `df`; a hazard-rate model for the amakihi that includes observer as a covariate (and truncating at 82.5m, given in @Marques:2007ey) can be specified by:
-```{r amakihi-hr}
-amakihi_hr_obs <- ds(amakihi, truncation=82.5, model=df(model=~hr, scale=~obs), transect="point")
+```{r amakihi-hr-obs}
+amakihi_hr_obs <- ds(amakihi, truncation=82.5, transect="point",
+                     model=df(model=~hr, scale=~obs))
+```
+As with the minke whale model, we can plot the resulting detection function.
+Since for the amakihi we used covariates in the detection function, the plot will show the detection function averaged over levels/values of the covariate and separately for each level set (for continuous covariates, the 25\%, 50\% and 75\% quantiles are displayed).
+
+When multiple covariates are used in the model, a series of plots are produced. We can see this in the bottom row of **Figure ???**, where we plot the results of fitting following model:
+```{r amakihi-hr-obs-mas}
+amakihi_hr_obs_mas <- ds(amakihi, truncation=82.5, transect="point",
+                         model=df(model=~hr, scale=~obs+mas))
+```
+
+```{r minke-hn-plot, fig.width=9, fig.height=9, fig.cap="Plot of fitted detection functions overlayed on the histograms of observed distances.
+Top row the minke whale data, half-normal model.
+Bottom row: average detection function for the amakihi data and marginal plot showing the levels of observer."}
+par(mfrow=c(3,3))
+plot(minke_hn, nomf=TRUE)
+plot(1:10,1:10, main="",xlab="",ylab="",type="n", axes=FALSE)
+plot(1:10,1:10, main="",xlab="",ylab="",type="n", axes=FALSE)
+plot(amakihi_hr_obs, nomf=TRUE)
+plot(1:10,1:10, main="",xlab="",ylab="",type="n", axes=FALSE)
+plot(amakihi_hr_obs_mas, nomf=TRUE)
 ```

### 2015-11-05 paper.Rmd
add pdf

@@
 plotmix(0.7, 0.05, 0.6)
-
 ```
@@
 When multiple covariates are used in the model, a series of plots are produced. We can see this in the bottom row of **Figure ???**, where we plot the results of fitting following model:
-```{r amakihi-hr-obs-mas}
+```{r amakihi-hr-obs-mas, warning=FALSE}
 amakihi_hr_obs_mas <- ds(amakihi, truncation=82.5, transect="point",
@@
 
-```{r minke-hn-plot, fig.width=9, fig.height=9, fig.cap="Plot of fitted detection functions overlayed on the histograms of observed distances.
+```{r minke-hn-plot, echo=FALSE, warning=FALSE, fig.width=9, fig.height=9, fig.cap="Plot of fitted detection functions overlayed on the histograms of observed distances.
 Top row the minke whale data, half-normal model.

### 2015-11-10 paper.Rmd
gof and model selection first hash

--- 2015-11-04 paper.Rmd
@@
   \usepackage{amssymb}
+  \usepackage{longtable}
 output: rticles::jss_article
@@
 From version 5 of Distance for Windows, R packages have been included to perform particular analyses (CITE Distance user manual).
-This paper shows how to fit detection functions, perform model checking and selection and estimate abundance in \proglang{R} using the package \pkg{Distance2}.
+This paper shows how to fit detection functions, perform model checking and selection and estimate abundance in \proglang{R} using the package \pkg{Distance}.
+\pkg{Distance} is a wrapper package around the more complex (and more powerful) \pkg{mrds} and offers a subset of the analyses possible with that package.
+
+**Need to include more history here (TRANSECT etc)?**
 
@@
 Using this histogram we can crudely estimate the drop-off in detectability by eye by tracing a line that approximates the tops of the histogram bars -- this is the detection function.
-\pkg{Distance2} estimates the parameters for a fixed-form detection function using maximum likelihood estimation.
-We address possible models in detail below
+\pkg{Distance} estimates the parameters for a fixed-form detection function using maximum likelihood estimation.
+We address possible models in detail below.
 
@@
 
-We demonstrate \pkg{Distance2} using two data sets: one line transect and one point transect. These data sets have been chosen to be representative of the kind of data seen in practice.
+We demonstrate \pkg{Distance} using two data sets: one line transect and one point transect. These data sets have been chosen to be representative of the kind of data seen in practice.
 
@@
 
-The rest of the paper is structured as follows: we first look into how data needs to be organised to allow it to be used with \pkg{Distance2}; models for the detection function are then covered including their formulation and examples of fitting in \proglang{R}.
+The rest of the paper is structured as follows: we first look into how data needs to be organised to allow it to be used with \pkg{Distance}; models for the detection function are then covered including their formulation and examples of fitting in \proglang{R}.
 We then spend some time looking at model checking and goodness of fit testing, as well as model selection.
@@
 
-The two example data sets used here are distributed with \pkg{Distance2} so readers can reproduce our results easiy.
-However in general we expect that data will be collected in the field and need to be formatted correctly for use with \pkg{Distance2}.
+The two example data sets used here are distributed with \pkg{Distance} so readers can reproduce our results easiy.
+However in general we expect that data will be collected in the field and need to be formatted correctly for use with \pkg{Distance}.
 The package allows for a flexible format for data input ranging from very simple to complex:
@@
 
+```{r load-data, echo=FALSE}
+# quietly load the data
+load("data/minke.RData")
+load("data/amakihi.RData")
+```
 ```{r minke-data-head, message=FALSE}
-library(Distance2)
+library(Distance)
 head(minke)
@@
 # Detection functions
+
+**Do we need to talk about binned distances too?**
 
@@
 To ensure that the model is not overly influence by distances far from zero we truncate the distances beyond a given distance $w$ (known as the *truncation distance*).
-Further details on truncation can be found in **CITATION**.
+Fitting to distances further away from zero does not improve the precision of abundance estimates considerably [@Buckland:2001vm, 103–108, 151–153].
 
@@
 There is a wide literature on possible formulations for the detection function [@Buckland:1992fa; @Eidous:2005bj; @Becker:2009cj; @Giammarino:2014eg; @Miller:2015hw; @Becker:2015fi].
-\code{Distance2} includes the most popular of these models, and includes an extendable class system to add new detection functions while avoiding code duplication (see "Extensions").
+\code{Distance} includes the most popular of these models, and includes an extendable class system to add new detection functions while avoiding code duplication (see "Extensions").
 
-Here we'll show two popular detection function approaches: "key function plus adjustments" (K+A) and mixture models.
+Here we'll detail the most popular detection function approach: "key function plus adjustments" (K+A).
 
@@
 
-Key function plus an adjustments (or adjustment series) models are formulated by taking a "key" function and optionally adding "adjustments" to it to improve the fit.
+Key function plus adjustment terms (or adjustment series) models are formulated by taking a "key" function and optionally adding "adjustments" to it to improve the fit.
 Mathematically we formulate this as:
@@
 Possible modelling options for key and adjustments are given in Table \ref{ka-tab} and illustrated in **Figure ???**.
+We select the number of adjustment terms ($K$) by AIC (further details in "Model checking and model selection").
 
-```{r hn-hr-par-comp, fig.width=10, fig.height=6.5, fig.cap="Half-normal (top row) and hazard-rate (bottom row) detection functions without adjustments, varying scale ($\\sigma$) and (for hazard-rate) shape ($b$) parameters (values are given above the plots).
+```{r hn-hr-par-comp, fig.width=10, fig.height=5, fig.cap="Half-normal (top row) and hazard-rate (bottom row) detection functions without adjustments, varying scale ($\\sigma$) and (for hazard-rate) shape ($b$) parameters (values are given above the plots).
 On the top row from left to right, the study species becomes more detectable (higher probability of detection at larger distances).
@@
 }
-
 ```
@@
 $$
-A distinct disadvantage of K+A models is that we must resort to constrained optimisation (via the \pkg{Rsolnp} package) in order to ensure that the resulting detection function is monotonic non-increasing over the whole range.
+A disadvantage of K+A models is that we must resort to constrained optimisation (via the \pkg{Rsolnp} package) in order to ensure that the resulting detection function is monotonic non-increasing over the whole range.
 
@@
 
-```{r adjust-mix-comp, fig.width=10, fig.height=5, fig.cap="Top row shows possible shapes for the detection function when adjustments are included for half-normal and hazard-rate models.
-Bottom row shows possible shapes when 2-component mixtures of half-normal functions are used.
-Notation is as in main text.", echo=FALSE}
-par(mfrow=c(2,4), mar=c(3.5, 3, 2, 1) + 0.1)
+```{r adjust-mix-comp, fig.width=9, fig.height=3, fig.cap="Possible shapes for the detection function when adjustments are included for half-normal and hazard-rate models.", echo=FALSE}
+par(mfrow=c(1,4), mar=c(3.5, 3, 2, 1) + 0.1)
 
@@
 title(ylab="Detection probability", line=2)
-
-# mixture models
-plotmix <- function(sig1, sig2, mix){
-
-  gmix <- function(x) mix*g.hn(x, sig1) + (1-mix)*g.hn(x,sig2)
-
-  curve(gmix, from=0, to=1, xlab="", ylab="",
-        main=bquote(sigma[1] == .(sig1) ~ sigma[2] == .(sig2) ~  phi[1] == .(mix)),
-        xlim=c(0,1), ylim=c(0,1), asp=1)
-  title(xlab="Distance", line=2)
-  title(ylab="Detection probability", line=2)
-}
-plotmix(0.8, 0.15, 0.3)
-plotmix(0.6, 0.1,0.7)
-plotmix(10, 0.2, 0.15)
-plotmix(0.7, 0.05, 0.6)
-
 ```
-
-### Mixture models
-
-We can also formulate the detection function as a mixture of (say) half-normal functions [@Miller:2015hw]:
-$$
-g(y; \boldsymbol{\theta}) = \sum_{k=1}^K \varphi_j \exp\left(-\frac{y^2}{2 \sigma_j^2}\right)
-$$
-where $\sum_{k=1}^K \varphi_j =1$ and $\boldsymbol{\theta} = (\varphi_1,\ldots,\varphi_{K-1},\sigma_1,\ldots,\sigma_K$.
-Such models do not require constrained optimisation (as K+A models do) as the mixtures are monotonic decreasing by construction.
-
-As with the K+A model, we describe model selection (in terms of number of mixture components) in "Model checking and selection" below.
 
@@
 
-Covariates can be included in this formulation by considering the scale parameter from the half-normal or hazard-rate detection functions as a linear model (on the exponential scale) of the ($k$) covariates ($\mathbf{z}$):
+Covariates can be included in this formulation by considering the scale parameter from the half-normal or hazard-rate detection functions as a linear model (on the exponential scale) of the ($J$) covariates ($\mathbf{z}$):
 $$
-\sigma(\mathbf{z}) = \exp(\beta_0 + \sum_k \beta_k z_k).
+\sigma(\mathbf{z}) = \exp(\beta_0 + \sum_{j=1}^J \beta_j z_j).
 $$
-In the next section we'll discuss model selection for these models.
+In the next section we'll discuss model selection.
 
@@
 
-Another important consideration is that K+A models that include covariates and one or more adjustments cannot be guaranteed to be monotonic non-increasing for all covariate combinations, as we don't have any model for the distribution of the covariates.
+Another important consideration is that K+A models which include covariates and one or more adjustments cannot be guaranteed to be monotonic non-increasing for all covariate combinations, as we don't have any model for the distribution of the covariates.
 For this reason, we advise against using both adjustments and covariates in a detection function [see @Miller:2015hw for an example of when this can be problematic].
@@
 
-The workhorse of detection function fitting in \pkg{Distance2} is the \code{ds} function.
+The workhorse of detection function fitting in \pkg{Distance} is the \code{ds} function.
 Here we show off the formulations for the detection function that we've seen above for both the minke whale and amakihi data.
@@
 ```
+Note that \code{ds} will automatically select adjustment terms by AIC and shows it's selection steps as it goes.
+
 **Figure???** shows the result of calling \code{plot} on the resulting model object.
@@
 
-```{r minke-hn-plot, fig.width=4, fig.height=4, fig.cap="Plot of fitted detection function overlayed on the histogram of observed distances for the minke whale data."}
-plot(minke_hn)
+We can specify the form of the detection function via the \code{key=} argument to \code{ds}. For example, a hazard rate model can be fitted as:
+```{r minke-hr-cos}
+minke_hrcos <- ds(minke, truncation=1.5, key="hr")
 ```
-
-We can specify the form of the detection function via the \code{model=df(...)} argument to \code{ds}. For example, fitting a hazard rate model with cosine adjustments of order 2 can be specified as:
-```{r minke-hr-cos}
-minke_hrcos <- ds(minke, truncation=1.5, model=df(model=~hr+cos(2)))
-```
+Again, \code{ds} fits a model with cosine adjustments (the default) but finds the AIC improvement to be insufficient to select the adjustment.
 
@@
 
-By default \code{ds} assumes that the data given to it is line transect, but we can switch to points using \code{transect="points"}.
-Including covariates in the scale is via the `scale=~...` argument to `df`; a hazard-rate model for the amakihi that includes observer as a covariate (and truncating at 82.5m, given in @Marques:2007ey) can be specified by:
-```{r amakihi-hr}
-amakihi_hr_obs <- ds(amakihi, truncation=82.5, model=df(model=~hr, scale=~obs), transect="point")
+By default \code{ds} assumes that the data given to it is line transect, but we can switch to points using \code{transect="point"}.
+Including covariates in the scale is via the \code{formula=~...} argument to \code{ds}; a hazard-rate model for the amakihi that includes observer as a covariate (and truncating at 82.5m, given in @Marques:2007ey) can be specified by:
+```{r amakihi-hr-obs}
+amakihi_hr_obs <- ds(amakihi, truncation=82.5, transect="point",
+                     key="hr", formula=~obs)
+```
+As with the minke whale model, we can plot the resulting detection function.
+**Since for the amakihi we used covariates in the detection function, the plot will show the detection function averaged over levels/values of the covariate and separately for each level set (for continuous covariates, the 25\%, 50\% and 75\% quantiles are displayed).**
+
+**It would be nice if I could get "nice" plotting working for `Distance` as they are much easier to interpret (IMO)**
+
+**When multiple covariates are used in the model, a series of plots are produced. We can see this in the bottom row of Figure ???, where we plot the results of fitting following model:**
+```{r amakihi-hr-obs-mas, warning=FALSE}
+amakihi_hr_obs_mas <- ds(amakihi, truncation=82.5, transect="point",
+                         key="hr", formula=~obs+mas)
+```
+
+```{r minke-hn-plot, echo=FALSE, warning=FALSE, fig.width=9, fig.height=3, fig.cap="Plot of fitted detection functions overlayed on the histograms of observed distances.
+Left minke whale data, half-normal model; centre, amakihi data hazard-rate with observer as a covariate; right, amakihi data, hazard-rate model with observer and minutes after sunrise as covariates"}
+par(mfrow=c(1,3))
+plot(minke_hn)
+plot(amakihi_hr_obs)
+plot(amakihi_hr_obs_mas)
 ```
@@
 
-```{r minke-hn-summary}
-summary(minke_hn)
+As with models fitted using \code{lm} or \code{glm} in \proglang{R} we can use \code{summary} to give us some useful information about our fitted model.
+For example for our hazard-rate model for the amakihi, with observer as a covariate:
+```{r amakihi-summary}
+summary(amakihi_hr_obs)
+```
+This summary information includes various details of the data and model specification, as well as the values of the coefficents ($\beta_j$) and their uncertainties, an "average" value for the detectability (see "Estimating abundance and variance" for details on how this is calculated) and it's uncertainty.
+The final line gives an estimate of abundance for the area covered by the survey (again, this will be covered in detail in the next section).
+
+## Goodness of fit
+
+To judge goodness of fit for detection functions when exact distances are used, we want to compare the cumulative distribution function (CDF) and empirical distribution function (EDF) for the detection function via a quantile-quantile plot (Q-Q plot).
+In our case the CDF evaluates the probability of observing an animal at a distance less than or equal to some given value.
+The EDF gives the proportion of observations for which the CDF is less than or equal to that of a given distance.
+In essence we're asking "is the number of observations up to a given distance in line with what the model says they should be?" (where “given values” are the observed distances).
+As usual for Q-Q plots, "good" models will have values close to the line $y=x$, poor models will show more deviations from that line.
+
+We can inspect Q-Q plots visually, though for a large number of models this can be tiresome and prone to subjective judgements.
+Instead we can quantify the Q-Q plot's information using a Kolmogorov-Smirnov or Cramer-von Mises test [@Burnham:2004vd].
+Both test to see if the points from the EDF and CDF are from the same distribution.
+The Kolmogorov-Smirnov uses the test statistic of the largest difference between a point on the Q-Q plot and the line $y=x$, whereas the Cramer-von Mises test uses the sum of all the differences.
+As it takes into account more information and is therefore more powerful, the Cramer-von Mises is generally preferred.
+A significant result from either tests indicates that the EDF and CDF do not come from the same distribution (and therefore the model is does not fit the data well).
+
+We can generate a Q-Q plot and test results using the \code{gof_ddf} function.
+We can see the goodness of fit tests below for two models for the amakihi data, first fitting a half-normal model without covariates or adjustments (not setting \code{adjustment=NULL} will force \code{ds} to fit a model with no adjustments), then calculating goodness of fit for that model and our hazard-rate model with observer and minutes after sunrise included:
+```{r amakihi-gof, fig.keep="none"}
+amakihi_hr <- ds(amakihi, truncation=82.5, transect="point", key="hn", adjustment=NULL)
+gof_ds(amakihi_hr)
+gof_ds(amakihi_hr_obs_mas)
+```
+The corresponding Q-Q plots are shown in **Figure ???**.
+
+```{r, amakihi-qq-comp, fig.width=6, fig.height=3, fig.cap="Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates) and hazard-rate model with observer and minutes after sunrise for the amakihi data.", echo=FALSE, results="hide"}
+par(mfrow=c(1,2))
+gof_ds(amakihi_hr, asp=1)
+gof_ds(amakihi_hr_obs_mas, asp=1)
 ```
 
-* summary results
-* qq plots
-* hessian/convergence
-* monotonicity
-* AIC -- what else?
+**Do we need to add something about binned distances and Q-Q plots here?**
+
+## Model selection
+
+Once we have a set of models which fit well, we can use Akaike's Information Criterion (AIC) to select between models.
+\pkg{Distance} includes a function to create table of summary information for fitted models, making it easy to get an overview of a large number of models at once.
+The \code{summarize_ds_models} function takes models as input and can be especially useful when paired with \pkg{knitr}'s \code{kable} function to create summary tables for publication.
+An example of this output is shown in **Table ???**.
+
+```{r results="asis", echo=FALSE}
+kable(summarize_ds_models(amakihi_hr, amakihi_hr_obs, amakihi_hr_obs_mas),
+      digits = 3, format = "latex", row.names = FALSE, escape=FALSE,
+      caption="Summary for the detection function models fitted to the amakihi data. \"C-vM\" stands for Cramer-von Mises. Models are sorted according to AIC.")
+```
+
+* monotonicity -- do we need this?
 
@@
 * Refer to Fewster et al 2009
+* calculation of $P_a$
 

### 2015-11-10 paper.Rmd
HT estimation

@@
       digits = 3, format = "latex", row.names = FALSE, escape=FALSE,
-      caption="Summary for the detection function models fitted to the amakihi data. \"C-vM\" stands for Cramer-von Mises. Models are sorted according to AIC.")
+      caption="Summary for the detection function models fitted to the amakihi data.
+\"C-vM\" stands for Cramer-von Mises, $P_a$ indicates average detectability (see \"Estimating abundance and variance\"), se indicates standard error.
+Models are sorted according to AIC.")
 ```
@@
 
-* Horvitz-Thompson
+Though fiting the detection function is the primary modelling step in distance sampling, we are really interested in estimating detectability, and from that abundance.
+We also wish to calculate our certainty in each abundance estimate.
+This section addresses these issues.
+
+## Abundance
+
+In order to estimate abundance ($\hat{N}$), we use the estimates of detection probability (the $\{p_i; i=1,\ldots,n\}$, above) in a Horvitz-Thompson like estimator:
+$$
+\hat{N} = \sum_{i=1}^n\frac{s_i}{\hat{p}_i}
+$$
+where $s_i$ are the sizes of the observed groups of objects, which is equal to 1 if objects only occur singly [@Borchers:2004wr].
+@thompson2012sampling is the cannonical reference to this type of estimator; intuitively, we can think of the estimates of detectability ($\hat{p}_i$) as "inflating" the group sizes ($s_i$) to account for incomplete detection -- we then sum to obtain the abundance estimate.
+For models that do not include covariates, $\hat{p}_i$ is equal for all $i$, so this is equivalent to summing the groups and inflating by a single value for detectability.
+
+We can use the Horvitz-Thompson-like estimator to obtain an "average probability of detection" for models which include covariates.
+We can consider what single detectability value would given the $\hat{N}$ and therefore calculate:
+$$
+\hat{P_a} = n/\hat{N}.
+$$
+This can be a useful summary statistic and is included in the \code{summary} output and the table produced by \code{summarize_ds_models}.
+
+**detectability vs. average detectability vs. detection probability -- clarify**
+
+## Variance
+
 * Refer to Fewster et al 2009
-* calculation of $P_a$
 

### 2015-11-11 paper.Rmd
variance and abundance, but running into mrds doc issues

@@
 
-**Do we need to talk about binned distances too?**
-
 As mentioned above, the detection function describes the relationship between observed distances and probability of detection.
 The detection function itself models the probability $\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)$ and is usually denoted $g(y; \boldsymbol{\theta})$ where $y$ is distance and $\boldsymbol{\theta}$ is a vector of parmeters to be estimated.
-Our goal is to estimate an average probability of detection (average in the sense of an average over the distances), so we must integrate out distance ($y$) from the detection function:
+Our goal is to estimate an *average probability of detection* ($p$, average in the sense of an average over the distances), so we must integrate out distance ($y$) from the detection function:
 $$
@@
 $$
-where $\pi(y)$ describes the distribution of objects with respect to the sampler and is $1/w$ for line transects and $\frac{2r}{w}$ for point transects [@Buckland:2001vm, Chapter 3].
+where $\pi(y)$ describes the distribution of objects with respect to the sampler and is $1/w$ for line transects and $\frac{2r}{w}$ for point transects, taking into account the geometry of the sampler [@Buckland:2001vm, Chapter 3].
 
@@
 $$
-g(y; \boldsymbol{\theta}) = k(y; \boldsymbol{\theta}_k)\left( 1+ \alpha_K(y; \boldsymbol{\theta}_\alpha)\right),
+g(y; \boldsymbol{\theta}) = k(y; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(y; \boldsymbol{\theta}_\text{adjust})\right),
 $$
-where $k$ is the key function and $\alpha_K$ is sum series of functions and the subscripts on the parameter vector indicate those parameters belonging to each part of the model.
+where $k$ is the key function and $\alpha_O$ is sum series of functions (given in **Table ???**), described as an *adjustment of order $O$*.
+Subscripts on the parameter vector indicate those parameters belonging to each part of the model (i.e.
+$\boldsymbol{\theta} = (\boldsymbol{\theta}_\text{key}, \boldsymbol{\theta}_\text{adjust})$).
 
-Models for $k$ are as follows
+Some models for the key are as follows:
 $$
@@
 \begin{array}{l l}
-  \exp\left(-\frac{y^2}{2 \sigma^2}\right) & \quad \text{half-normal} \\
-  1-\exp\left(\left(-\frac{y}{\sigma}\right)^{-b}\right) & \quad \text{hazard-rate} \\
-  1/w & \quad \text{uniform}
+  \exp\left(-\frac{y^2}{2 \sigma^2}\right) & \quad \text{half-normal,} \\
+  1-\exp\left(\left(-\frac{y}{\sigma}\right)^{-b}\right) & \quad \text{hazard-rate,} \\
+  1/w & \quad \text{uniform.}
 \end{array} \right.
@@
 \hline
- uniform  & $1/w$   & cosine  & $\sum_{k=1}^K a_k \cos(k \pi y/w)$ \\
- & & Simple polynomial & $\sum_{k=1}^K a_k (y/w)^{2k}$ \\
- half-normal  & $\exp\left(-\frac{y^2}{2 \sigma^2}\right)$ & cosine  & $\sum_{k=2}^K a_k \cos(k \pi y/w)$ \\
- & & Hermite polynomial & $\sum_{k=2}^K a_k H_{2k}(y/\sigma)$ \\
- hazard-rate  & $1-\exp\left[-\left(\frac{y}{\sigma}\right)^{-b}\right]$ & cosine  & $\sum_{k=2}^K a_k \cos(k \pi y/w)$ \\
- & & Simple polynomial & $\sum_{k=2}^K a_k (y/w)^{2k}$ \\
+ uniform  & $1/w$   & cosine  & $\sum_{o=1}^O a_o \cos(o \pi y/w)$ \\
+ & & Simple polynomial & $\sum_{o=1}^O a_o (y/w)^{2o}$ \\
+ half-normal  & $\exp\left(-\frac{y^2}{2 \sigma^2}\right)$ & cosine  & $\sum_{o=2}^O a_o \cos(o \pi y/w)$ \\
+ & & Hermite polynomial & $\sum_{o=2}^O a_o H_{2o}(y/\sigma)$ \\
+ hazard-rate  & $1-\exp\left[-\left(\frac{y}{\sigma}\right)^{-b}\right]$ & cosine  & $\sum_{o=2}^O a_o \cos(o \pi y/w)$ \\
+ & & Simple polynomial & $\sum_{o=2}^O a_o (y/w)^{2o}$ \\
 \hline
@@
 $$
-g(y; \boldsymbol{\theta}) = \frac{k(y; \boldsymbol{\theta})\left( 1+ \alpha_K(y; \boldsymbol{\theta})\right)}{k(0; \boldsymbol{\theta})\left( 1+ \alpha_K(0; \boldsymbol{\theta})\right)}
+g(y; \boldsymbol{\theta}) = \frac{k(y; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(y; \boldsymbol{\theta}_\text{adjust})\right)}{k(0; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(0; \boldsymbol{\theta}_\text{adjust})\right)}.
 $$
+
 A disadvantage of K+A models is that we must resort to constrained optimisation (via the \pkg{Rsolnp} package) in order to ensure that the resulting detection function is monotonic non-increasing over the whole range.
@@
 
-As mentioned when we talked about pooling robustness, above, there are many factors that can affect the probability of detecting an object.
+There are many factors that can affect the probability of detecting an object.
 These include things like the observer, the vessel or platform used, the sea state, weather conditions and time of day (to name but a few).
@@
 
-Covariates can be included in this formulation by considering the scale parameter from the half-normal or hazard-rate detection functions as a linear model (on the exponential scale) of the ($J$) covariates ($\mathbf{z}$):
+Covariates can be included in this formulation by considering the scale parameter from the half-normal or hazard-rate detection functions as a linear model (on the exponential scale) of the ($J$) covariates ($\mathbf{z}$; a vector of length $J$ for each observation):
 $$
@@
 $$
-p(\mathbf{z_i}) = \int_0^w \pi(y) g(y, \mathbf{z_i}; \boldsymbol{\theta}) dy
+p(\mathbf{z_i}) = \int_0^w \pi(y) g(y, \mathbf{z_i}; \boldsymbol{\theta}) dy,
 $$
-so we now calculate a value of "average" probability of detection (again the average is in the sense of distance) per observation.
+where $\mathbf{z_i}$ is the vector of $J$ covariates associated with observation $i$.
+So for covariate models, we are calculating a value of "average" probability of detection (again the average is in the sense of distance) per observation.
 There will be as many unique values of $p(\mathbf{z_i})$ as there are unique covariate combinations in our data.
@@
 
-**Do we need to add something about binned distances and Q-Q plots here?**
-
 ## Model selection
@@
 The \code{summarize_ds_models} function takes models as input and can be especially useful when paired with \pkg{knitr}'s \code{kable} function to create summary tables for publication.
-An example of this output is shown in **Table ???**.
+An example of this output is shown in **Table ???** and was generated by:
+```{r summary-table-null, eval=FALSE}
+summarize_ds_models(amakihi_hr, amakihi_hr_obs, amakihi_hr_obs_mas)
+```
 
-```{r results="asis", echo=FALSE}
+```{r summary-table, results="asis", echo=FALSE}
 kable(summarize_ds_models(amakihi_hr, amakihi_hr_obs, amakihi_hr_obs_mas),
@@
 
-* monotonicity -- do we need this?
-
 # Estimating abundance and variance
@@
 We also wish to calculate our certainty in each abundance estimate.
-This section addresses these issues.
+This section addresses these issues mathematically before showing how to estimate abundance and its variance in R.
 
@@
 
-In order to estimate abundance ($\hat{N}$), we use the estimates of detection probability (the $\{p_i; i=1,\ldots,n\}$, above) in a Horvitz-Thompson like estimator:
-$$
-\hat{N} = \sum_{i=1}^n\frac{s_i}{\hat{p}_i}
-$$
+We wish to obtain the abundance in a study region, which we have sampled some (representative) subset of.
+To do this we first calculate the abundance in the area we have surveyed (the *covered area*) to obtain $\hat{N}_\text{C}$, we can then scale this up to the full study area by multiplying it by the ratio of covered area to study area.
+
+First, in order to estimate abundance in the covered area ($\hat{N}_\text{C}$), we use the estimates of detection probability (the $\{p_i; i=1,\ldots,n\}$, above) in a Horvitz-Thompson like estimator:
+\begin{equation}
+\hat{N}_\text{C} = \sum_{i=1}^n\frac{s_i}{\hat{p}_i},
+\label{ht}
+\end{equation}
 where $s_i$ are the sizes of the observed groups of objects, which is equal to 1 if objects only occur singly [@Borchers:2004wr].
 @thompson2012sampling is the cannonical reference to this type of estimator; intuitively, we can think of the estimates of detectability ($\hat{p}_i$) as "inflating" the group sizes ($s_i$) to account for incomplete detection -- we then sum to obtain the abundance estimate.
-For models that do not include covariates, $\hat{p}_i$ is equal for all $i$, so this is equivalent to summing the groups and inflating by a single value for detectability.
+For models that do not include covariates, $\hat{p}_i$ is equal for all $i$, so this is equivalent to summing the groups and inflating that sum by the corresponding $\hat{p} (=\hat{p}_i \forall i)$.
 
-We can use the Horvitz-Thompson-like estimator to obtain an "average probability of detection" for models which include covariates.
-We can consider what single detectability value would given the $\hat{N}$ and therefore calculate:
+Having obtained the abundance in the covered area, we can then scale-up to the study area:
 $$
-\hat{P_a} = n/\hat{N}.
+\hat{N} = \frac{A}{a} \hat{N}_\text{C},
+$$
+where $A$ is the area of the study region to extrapolate the abundance estimate to and $a$ is the covered area.
+For line transects $a=2wL$ (twice the truncation distance multiplied by the total length of transects surveyed) and for points $a=\pi w^2 K$.
+
+We can use the Horvitz-Thompson-like estimator to calculate the "average" detectability for models which include covariates.
+We can consider what single detectability value would give the estimated $\hat{N}$ and therefore calculate:
+$$
+\hat{P_a} = n/\hat{N}_\text{C}.
 $$
@@
 
-**detectability vs. average detectability vs. detection probability -- clarify**
+**stratification**
 
@@
 
-* Refer to Fewster et al 2009
+Uncertainty in $\hat{N}$ comes from two sources:
+
+1. *Model uncertainty*, from the estimation of the detection function parameters $\boldsymbol{\theta}$
+2. *Sampling uncertainty*, from the distribution of animals along the transect lines or between visiting occasions for points.
+
+We can see this by looking at the Horvitz-Thompson estimation in (\ref{ht}) and consider the terms which are random.
+These are: the detectability $\hat{p}_i$ (and hence the parameters of the detection function it is derived from) and $n$, the number of observations.
+We assume that the observed group size ($s_i$) is recorded without error.
+
+Model uncertainty can be addressed using standard maximum likelihood theory.
+We can invert the Hessian matrix of the likelihood to obtain a variance-covariance matrix.
+We can then pre- and post-multiple this by the derivatives of $\hat{N}_\text{C}$
+$$
+\widehat{\text{Var}}_\text{model}\left( \hat{N}_\text{C}\right) = \left(\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}\right)^\text{T} \left(\hat{\mathbf{H}}(\hat{\boldsymbol{\theta}})^{-1} \right)\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}
+$$
+where the partial derivatives of $\hat{N}_\text{C}$ are evaluated at the MLE ($\hat{\theta}$) and $\mathbf{H}$ is the first partial Hessian (outer product of first derivatives of the log likelihood) for numerical stability [@Buckland:2001vm, p 62].
+Note that although we calculate uncertainty in $\hat{N}_\text{C}$, we can trivially scale-up to variance of $\hat{N}$.
+
+Sampling uncertainty can be characterised (rather than as simply $n$) by the *encounter rate*: $n/L$ for line transects (where $L$ is the total line length) or $n/K$ for point transects.
+Various formulations for the encounter rate variance are discussed in detail in @Fewster:2009ku.
+\pkg{Distance} implements all of the estimators of encounter rate variance given in that article plus some extras.
+
+For line transects, by default, \pkg{Distance} uses a variation of the estimator "R2" from @Fewster:2009ku replacing number of observations per sample with the estimated abundance per sample (per @Innes:2002ka); that is for line transects:
+$$
+\widehat{\text{Var}}_{\text{encounter},R2} = \frac{K}{L^2(K-1)} \sum_{k=1}^{K} l_k^2 \left( \frac{N_k}{l_k} - \frac{N}{L}\right)^2,
+$$
+where $l_k$ are the lengths of the $K$ transects (such that $L = \sum_{k=1}^K l_k$).
+
+For points,
+
+The \code{varn} manual page gives further advice and technical detail on this matter.
+
+## Estimating abundance and variance in R
+
+Going back to the minke whale data, we have the information we require to calculate $A$ and $a$ above, so we can estimate abundance and its variance.
+When we supply data to \code{ds} in the ``flat file'' format given above, \code{ds} will automatically calculate abundance estimates based on the survey setup in the data.
+
+Having already fitted a model to the minke whale data, we can see the results of the abundance estimation by simply looking at a model summary:
+```{r minke-N-summary}
+summary(minke_hn)
+```
 
@@
 
-* writing new detection functions
 * Interface to dsm etc

### 2015-11-12 paper.Rmd
barebones paper with holes for various things
see issues for what to do next

@@
 \pkg{Distance} is a wrapper package around the more complex (and more powerful) \pkg{mrds} and offers a subset of the analyses possible with that package.
-
-**Need to include more history here (TRANSECT etc)?**
 
@@
 ```{r summary-table-null, eval=FALSE}
+library(knitr)
 summarize_ds_models(amakihi_hr, amakihi_hr_obs, amakihi_hr_obs_mas)
@@
 ```{r summary-table, results="asis", echo=FALSE}
+library(knitr)
 kable(summarize_ds_models(amakihi_hr, amakihi_hr_obs, amakihi_hr_obs_mas),
@@
 
-**stratification**
+### Post-stratification
+
+We may wish to calculate abundance estimates for some sub-regions of the study region, we call these areas *strata*.
+A stratum may be defined by, for example, a different habitat type or by the sex of the animal (or some combination) which may be of interest for biological or management reasons.
+In order to calculate estimates for a given stratification each observation must occur in a stratum which must be labelled with a \code{Region.Label} and have a corresponding \code{Area} (if we are using an animal characteristic like sex, we would have the areas be the same but if we were using say forested vs.
+wetland habitat the areas of those strata would be different).
+Finally, we must also know which stratum a given transect lies in.
+
+As an example the minke whale data consists of two strata: \code{North} and \code{South} relating to a stratum further away and nearer the Antarctic ice edge, respectively.
+Figure \ref{minke-strata} shows the two strata, along with observations and transect lines.
+
+\begin{figure}
+\includegraphics{minke-strata}
+\label{minke-strata}
+\caption{Strata used for the minke whale data adapted from @Hedley:2004et.
+Points show the locations of observations along transect lines.
+The stepped line shows the boundary between North and South strata.
+Further details on the survey are available in @Branch:2001ua (simulated data is based on ``1992/93 Area III'' therein).}
+\end{figure}
 
@@
 We can invert the Hessian matrix of the likelihood to obtain a variance-covariance matrix.
-We can then pre- and post-multiple this by the derivatives of $\hat{N}_\text{C}$
+We can then pre- and post-multiply this by the derivatives of $\hat{N}_\text{C}$
 $$
@@
 
-The \code{varn} manual page gives further advice and technical detail on this matter.
+The \code{varn} manual page gives further advice and technical detail on encounter rate variance.
 
@@
 Going back to the minke whale data, we have the information we require to calculate $A$ and $a$ above, so we can estimate abundance and its variance.
-When we supply data to \code{ds} in the ``flat file'' format given above, \code{ds} will automatically calculate abundance estimates based on the survey setup in the data.
+When we supply data to \code{ds} in the ``flatfile'' format given above, \code{ds} will automatically calculate abundance estimates based on the survey setup in the data.
 
@@
 ```
+This prints a rather large amount of information.
+One useful feature is that the returned object is a \code{list} of \code{data.frames}, so we can again use \pkg{knitr}s \code{kable} function to create useful tables for publication.
+For example **Table ???**.
+
+```{r minke-abundance-table}
+minke_table <- summary(minke_hn)$dht$individuals$N
+minke_table$lcl <- minke_table$ucl <- minke_table$df <- NULL
+colnames(minke_table) <- c("Stratum", "$\\hat{N}$", "$\\text{se}(\\hat{N}$)", 
+                           "$\\text{CV}(\\hat{N}$)")
+kable(minke_table, digits = 3, format = "latex",
+      row.names = FALSE, escape=FALSE,caption="Summary of abundance estimation for the half-normal model for the minke whale data.")
+```
 
@@
 
-* Interface to dsm etc
-* MRDS?
-* why is + important?
+The features of \pkg{Distance} are delibarately limited to provide a simplified interface for users.
+However, for more complex distance sampling-based analyses there are further related packages for modelling in \proglang{R}.
+
+We noted at the start of the article that \pkg{Distance} is a simpler wrapper around the package \pkg{mrds}.
+Additional features are available in \pkg{mrds} including the ability to model data where the assumption that detection is certain at zero distance from the line or point by using mark-recapture type methods when two observers are used [see @Burt:2014gu for an introduction].
+
+The abundance estimates calculated here are based on the assumption that within a given stratum abundance is constant.
+We may extend this approach to many strata, making the area of each very small to account for small-scale variation in space.
+A more rigourous approach is to build a spatial model incorporating spatially-referenced environmental data (for example derived from GIS products).
+\pkg{Distance} interfaces with once such package to perform this type of analysis: \pkg{dsm}.
+So-called "density surface modelling" uses the generalized additive model framework [e.g.
+@Wood:2006vg] to build models of abundance (that take into account detectability) as a function of environmental covariates, as part of a two stage model [@Hedley:2004et; @Miller:2013fq].
 
@@
 
-* Other software? \pkg{unmarked}, \pkg{RDistance} etc
+Here we have given an introduction as to how to perform a distance sampling analysis in \proglang{R}.
+We have covered the possible models for detectability, model checking and selection and finally abundance and variance estimation.
+
+In combination with tools such as \pkg{knitr} and \pkg{rmarkdown}, the helper functions in \pkg{Distance} provide a useful set of tools to perform reproducible analyses of wildlife abundance for both managers and ecologists.
+We hope that this paper (itself written in RMarkdown) will provide a useful set of examples for those wishing to persue this.
+
+We note that there are other packages available for performing distance sampling analyses in \proglang{R} but believe that \pkg{Distance} is the most flexible and feature-complete.
+**Appendix ???** gives a feature comparison between \pkg{Distance} and the other packages available.
+
+# Acknowledgements
+
+The authors would like to thank the many users of \pkg{Distance}, \pkg{mrds} and DISTANCE who have contributed bug reports and suggestions for improvements over the years.

### 2015-11-16 paper.Rmd
encounter rate variance update

@@
 
-Uncertainty in $\hat{N}$ comes from two sources:
+Here we take an intuitive approach to uncertainty estimation, for a full derivations consult @Marques:2003vb. Uncertainty in $\hat{N}$ comes from two sources:
 
@@
 
-Sampling uncertainty can be characterised (rather than as simply $n$) by the *encounter rate*: $n/L$ for line transects (where $L$ is the total line length) or $n/K$ for point transects.
-Various formulations for the encounter rate variance are discussed in detail in @Fewster:2009ku.
-\pkg{Distance} implements all of the estimators of encounter rate variance given in that article plus some extras.
+Sampling uncertainty can be characterised (rather than as simply $n$) by the *encounter rate*: the number of animals per unit transect.
+When covariates are not included in the detection function we can define the encounter rate as $n/L$ for line transects (where $L$ is the total line length) or $n/T$ for point transects (where $T$ is the total number of visits to all points).
+When covariates are included in the detection function, it is recommended that we substitute the $n$ in the previous expressions with the estimated abundance $\hat{N}$ as this will take into account the effects of the covariates.
 
-For line transects, by default, \pkg{Distance} uses a variation of the estimator "R2" from @Fewster:2009ku replacing number of observations per sample with the estimated abundance per sample (per @Innes:2002ka); that is for line transects:
+For line transects, by default, \pkg{Distance} uses a variation of the estimator "R2" from @Fewster:2009ku replacing number of observations per sample with the estimated abundance per sample [@Innes:2002ka; @Marques:2003vb]:
 $$
@@
 where $l_k$ are the lengths of the $K$ transects (such that $L = \sum_{k=1}^K l_k$).
+Whereas for points we use estimator "P3" from @Fewster:2009ku but again replacing $n$ by $\hat{N}$ to obtain the following estimator:
+$$
+\widehat{\text{Var}}_{\text{encounter},P3} = \frac{1}{T(K-1)} \sum_{k=1}^{K} t_k \left( \frac{N_k}{t_k} - \frac{N}{T}\right)^2,
+$$
+where $t_k$ is the number of visits to point $k$ and $T = \sum_{k=1}^K t_k$.
 
-For points,
-
+Various formulations for the encounter rate variance are discussed in detail in @Fewster:2009ku.
+\pkg{Distance} implements all of the estimators of encounter rate variance given in that article.
 The \code{varn} manual page gives further advice and technical detail on encounter rate variance.

### 2015-11-17 paper.Rmd
table and figure references

@@
 ```{r points-and-lines, fig.height=6, fig.width=7, echo=FALSE, message=FALSE, fig.cap="Left side plots show an example of a survey of an area containing a population of 500 objects, blue indicates sampler placement (top lines, bottom points) and red dots indicate detected individuals.
-The right side of the figure shows histograms of observed distances (again, lines top and points bottom)."}
+The right side of the figure shows histograms of observed distances (again, lines top and points bottom).\\label{fig:pointslines}"}
 opar <- par(mfrow=c(2,2))
@@
 We address possible models in detail below.
+
+Figure \ref{fig:pointslines} shows examples of 500 individuals sampled using line and point transects (left column) and their corresponding histograms (right column).
 
@@
 $$
-where $k$ is the key function and $\alpha_O$ is sum series of functions (given in **Table ???**), described as an *adjustment of order $O$*.
+where $k$ is the key function and $\alpha_O$ is sum series of functions (given in Table \ref{tab:keyadj}), described as an *adjustment of order $O$*.
 Subscripts on the parameter vector indicate those parameters belonging to each part of the model (i.e.
@@
 $$
-Possible modelling options for key and adjustments are given in Table \ref{ka-tab} and illustrated in **Figure ???**.
+Possible modelling options for key and adjustments are given in Table \ref{tab:keyadj} and illustrated in Figure \ref{fig:hnhr} and \ref{fig:keyadj}.
 We select the number of adjustment terms ($K$) by AIC (further details in "Model checking and model selection").
@@
 On the top row from left to right, the study species becomes more detectable (higher probability of detection at larger distances).
-The bottom row shows the hazard-rate model's more pronounced shoulder.", echo=FALSE}
+The bottom row shows the hazard-rate model's more pronounced shoulder.\\label{fig:hnhr}", echo=FALSE}
 par(mfrow=c(2,4), mar=c(3.5, 3, 2, 1) + 0.1)
@@
 \begin{table}
-\caption{Modelling options for key plus adjustments models for the detection function.}
+\caption{Modelling options for key plus adjustment series models for the detection function.}
 \begin{tabular}{llll}
@@
 \end{tabular}
-\label{ka-tab}
+\label{tab:keyadj}
 \end{table}
@@
 
-```{r adjust-mix-comp, fig.width=9, fig.height=3, fig.cap="Possible shapes for the detection function when adjustments are included for half-normal and hazard-rate models.", echo=FALSE}
+```{r adjust-mix-comp, fig.width=9, fig.height=3, fig.cap="Possible shapes for the detection function when adjustments are included for half-normal and hazard-rate models.\\label{fig:keyadj}", echo=FALSE}
 par(mfrow=c(1,4), mar=c(3.5, 3, 2, 1) + 0.1)
@@
 
-**Figure???** shows the result of calling \code{plot} on the resulting model object.
+Figure \ref{fig:minkeamakihi} (left panel) shows the result of calling \code{plot} on the resulting model object.
 We can also call \code{summary} on the model object to get summary information about the fitted model (though we postpone this until the next section).
@@
 ```
-As with the minke whale model, we can plot the resulting detection function.
-**Since for the amakihi we used covariates in the detection function, the plot will show the detection function averaged over levels/values of the covariate and separately for each level set (for continuous covariates, the 25\%, 50\% and 75\% quantiles are displayed).**
+As with the minke whale model, we can plot the resulting detection function (Figure \ref{fig:minkeamakihi}, centre and right panels).
+*Since for the amakihi we used covariates in the detection function, the plot will show the detection function averaged over levels/values of the covariate and separately for each level set (for continuous covariates, the 25\%, 50\% and 75\% quantiles are displayed).*
 
@@
 
-**When multiple covariates are used in the model, a series of plots are produced. We can see this in the bottom row of Figure ???, where we plot the results of fitting following model:**
+*When multiple covariates are used in the model, a series of plots are produced.
+We can see this in the bottom row of Figure \ref{fig:minkeamakihi}, where we plot the results of fitting following model:*
 ```{r amakihi-hr-obs-mas, warning=FALSE}
@@
 
-```{r minke-hn-plot, echo=FALSE, warning=FALSE, fig.width=9, fig.height=3, fig.cap="Plot of fitted detection functions overlayed on the histograms of observed distances.
-Left minke whale data, half-normal model; centre, amakihi data hazard-rate with observer as a covariate; right, amakihi data, hazard-rate model with observer and minutes after sunrise as covariates"}
+```{r minke-amakihi-hn-plot, echo=FALSE, warning=FALSE, fig.width=9, fig.height=3, fig.cap="Plot of fitted detection functions overlayed on the histograms of observed distances.
+Left minke whale data, half-normal model; centre, amakihi data hazard-rate with observer as a covariate; right, amakihi data, hazard-rate model with observer and minutes after sunrise as covariates.\\label{fig:minkeamakihi}"}
 par(mfrow=c(1,3))
@@
 ```
-The corresponding Q-Q plots are shown in **Figure ???**.
+The corresponding Q-Q plots are shown in Figure \ref{amakihi-qq}.
 
-```{r, amakihi-qq-comp, fig.width=6, fig.height=3, fig.cap="Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates) and hazard-rate model with observer and minutes after sunrise for the amakihi data.", echo=FALSE, results="hide"}
+```{r, amakihi-qq-comp, fig.width=6, fig.height=3, fig.cap="Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates) and hazard-rate model with observer and minutes after sunrise for the amakihi data.\\label{amakihi-qq}", echo=FALSE, results="hide"}
 par(mfrow=c(1,2))
@@
 The \code{summarize_ds_models} function takes models as input and can be especially useful when paired with \pkg{knitr}'s \code{kable} function to create summary tables for publication.
-An example of this output is shown in **Table ???** and was generated by:
+An example of this output is shown in Table \ref{tab:amakihi} and was generated by:
 ```{r summary-table-null, eval=FALSE}
@@
 \"C-vM\" stands for Cramer-von Mises, $P_a$ indicates average detectability (see \"Estimating abundance and variance\"), se indicates standard error.
-Models are sorted according to AIC.")
+Models are sorted according to AIC.\\label{tab:amakihi}")
 ```
@@
 As an example the minke whale data consists of two strata: \code{North} and \code{South} relating to a stratum further away and nearer the Antarctic ice edge, respectively.
-Figure \ref{minke-strata} shows the two strata, along with observations and transect lines.
+Figure \ref{fig:minke-strata} shows the two strata, along with observations and transect lines.
 
@@
 \includegraphics{minke-strata}
-\label{minke-strata}
+\label{fig:minke-strata}
 \caption{Strata used for the minke whale data adapted from @Hedley:2004et.
@@
 One useful feature is that the returned object is a \code{list} of \code{data.frames}, so we can again use \pkg{knitr}s \code{kable} function to create useful tables for publication.
-For example **Table ???**.
+For example Table \ref{minke-abund}.
 
@@
 kable(minke_table, digits = 3, format = "latex",
-      row.names = FALSE, escape=FALSE,caption="Summary of abundance estimation for the half-normal model for the minke whale data.")
+      row.names = FALSE, escape=FALSE,caption="Summary of abundance estimation for the half-normal model for the minke whale data.\\label{minke-abund}")
 ```

### 2015-11-17 paper.Rmd
seems like pandoc knows what to do there anyway...
tables look less gross

@@
   \usepackage{longtable}
+  \usepackage{booktabs}
 output: rticles::jss_article
@@
 \hline
-key function   & form   & adjustment & form\\
+Key function   & Form   & Adjustment series & Form\\
 \hline
- uniform  & $1/w$   & cosine  & $\sum_{o=1}^O a_o \cos(o \pi y/w)$ \\
+ Uniform  & $1/w$   & cosine  & $\sum_{o=1}^O a_o \cos(o \pi y/w)$ \\
  & & Simple polynomial & $\sum_{o=1}^O a_o (y/w)^{2o}$ \\
- half-normal  & $\exp\left(-\frac{y^2}{2 \sigma^2}\right)$ & cosine  & $\sum_{o=2}^O a_o \cos(o \pi y/w)$ \\
+ Half-normal  & $\exp\left(-\frac{y^2}{2 \sigma^2}\right)$ & cosine  & $\sum_{o=2}^O a_o \cos(o \pi y/w)$ \\
  & & Hermite polynomial & $\sum_{o=2}^O a_o H_{2o}(y/\sigma)$ \\
- hazard-rate  & $1-\exp\left[-\left(\frac{y}{\sigma}\right)^{-b}\right]$ & cosine  & $\sum_{o=2}^O a_o \cos(o \pi y/w)$ \\
+ Hazard-rate  & $1-\exp\left[-\left(\frac{y}{\sigma}\right)^{-b}\right]$ & cosine  & $\sum_{o=2}^O a_o \cos(o \pi y/w)$ \\
  & & Simple polynomial & $\sum_{o=2}^O a_o (y/w)^{2o}$ \\
@@
       digits = 3, format = "latex", row.names = FALSE, escape=FALSE,
+      booktabs=TRUE,
       caption="Summary for the detection function models fitted to the amakihi data.
-\"C-vM\" stands for Cramer-von Mises, $P_a$ indicates average detectability (see \"Estimating abundance and variance\"), se indicates standard error.
+``C-vM'' stands for Cramer-von Mises, $P_a$ indicates average detectability (see ``Estimating abundance and variance''), se indicates standard error.
 Models are sorted according to AIC.\\label{tab:amakihi}")
@@
                            "$\\text{CV}(\\hat{N}$)")
-kable(minke_table, digits = 3, format = "latex",
-      row.names = FALSE, escape=FALSE,caption="Summary of abundance estimation for the half-normal model for the minke whale data.\\label{minke-abund}")
+kable(minke_table, digits=3, format="latex", booktabs=TRUE,
+      row.names=FALSE, escape=FALSE,caption="Summary of abundance estimation for the half-normal model for the minke whale data.\\label{minke-abund}")
 ```
@@
 The authors would like to thank the many users of \pkg{Distance}, \pkg{mrds} and DISTANCE who have contributed bug reports and suggestions for improvements over the years.
+
+# Bibliography

### 2015-11-17 paper.Rmd
objects!

@@
 
-It is clear that census-type surveys (quadrat or strip transects) are inefficient (requiring a lot of effort on the part of those in the field) and we should expect that not all animals can be observed.
+It is clear that census-type surveys (quadrat or strip transects) are inefficient (requiring a lot of effort on the part of those in the field) and we should expect that not all objects (animals, plants, dung, etc) can be observed.
 Accounting for imperfect detectability is an important consideration when want to obtain accurate estimates of abundance [@LahozMonfort:2013bg].
@@
   lines(x=rep(mean(range(lt$x)),2), y=c(0,1), col="blue",lty=2)
-  # calculate the distances to animals from the line
+  # calculate the distances to objects from the line
   distances <- abs(lt$x - x)
@@
   detected <- exp(-distances^2/(2*sigma^2)) > runif(length(distances))
-  # plot those animals detected
+  # plot those objects detected
   points(x[detected], y[detected], pch=19, cex=0.6, col="red")
@@
   points(pt$x[i], pt$y[i], pch=19, col="blue",cex=0.7)
-  # calculate the distances to animals from the line
+  # calculate the distances to objects from the line
   distances <- sqrt((pt$x[i] - x)^2+(pt$y[i]-y)^2)
@@
 Having illustrated how to obtain a good detection function, we show how to estimate abundance using that model, including using post-stratification.
-The final two sections of the article look at extensions (both in terms of methodology and software) and put the package in a broader context amongst other software packages used for animal abundance estimation.
+The final two sections of the article look at extensions (both in terms of methodology and software) and put the package in a broader context amongst other software packages used for estimating the abundance of biological populations.
 
@@
 * *Non-increasing*: we don't think that observers should be more likely to see things further away than those nearer to them.
-This usually indicates an issue with field procedure (that the distribution of animals with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this.
+This usually indicates an issue with field procedure (that the distribution of objects with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this.
 * *Model robust*: models should be flexible enough to make many different shapes.
@@
 To judge goodness of fit for detection functions when exact distances are used, we want to compare the cumulative distribution function (CDF) and empirical distribution function (EDF) for the detection function via a quantile-quantile plot (Q-Q plot).
-In our case the CDF evaluates the probability of observing an animal at a distance less than or equal to some given value.
+In our case the CDF evaluates the probability of observing an object at a distance less than or equal to some given value.
 The EDF gives the proportion of observations for which the CDF is less than or equal to that of a given distance.
@@
 1. *Model uncertainty*, from the estimation of the detection function parameters $\boldsymbol{\theta}$
-2. *Sampling uncertainty*, from the distribution of animals along the transect lines or between visiting occasions for points.
+2. *Sampling uncertainty*, from the distribution of objects along the transect lines or between visiting occasions for points.
 
@@
 
-Sampling uncertainty can be characterised (rather than as simply $n$) by the *encounter rate*: the number of animals per unit transect.
+Sampling uncertainty can be characterised (rather than as simply $n$) by the *encounter rate*: the number of objects per unit transect.
 When covariates are not included in the detection function we can define the encounter rate as $n/L$ for line transects (where $L$ is the total line length) or $n/T$ for point transects (where $T$ is the total number of visits to all points).

### 2015-11-20 paper.Rmd
update with new plots
* some corrections
* add Jeff to authors
* PDF plots (needs mrds/Distance from github)

@@
   - name: David L Miller
-    affiliation: Centre for Research into Ecological and Environmental Modelling, University of St Andrews
+    affiliation: University of St Andrews
     address: >
+      Centre for Research into Ecological and Environmental Modelling,
       The Observatory,
@@
     url: http://converged.yt
+  - name: Jeffrey L Laake
+    affiliation: National Marine Mammal Laboratory
+    address: >
+      Alaska Fisheries Science Center
+      7600 Sand Point Way N.E., Seattle, WA 98115, USA
+    email: Jeff.Laake@noaa.gov
 title:
@@
 
-Looking at an example of the data format given in the last bullet, we can examine the minke whale data:
-
+The minke whale data follows the "flatfile" format given in the last bullet point:
 ```{r load-data, echo=FALSE}
@@
 ```
-and the amakihi data:
+Where as the amakihi data lacks effort and stratum data:
 ```{r amakihi-data-head}
@@
 
-* what do we want to say here?
+We'll explore what the effects of including effort data are during analysis below.
 
@@
 As mentioned above, the detection function describes the relationship between observed distances and probability of detection.
-The detection function itself models the probability $\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)$ and is usually denoted $g(y; \boldsymbol{\theta})$ where $y$ is distance and $\boldsymbol{\theta}$ is a vector of parmeters to be estimated.
+The detection function itself models the probability $\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)$ and is usually denoted $g(y; \boldsymbol{\theta})$ where $y$ is distance (either from a line or point) and $\boldsymbol{\theta}$ is a vector of parmeters to be estimated.
 Our goal is to estimate an *average probability of detection* ($p$, average in the sense of an average over the distances), so we must integrate out distance ($y$) from the detection function:
@@
 $$
-where $\pi(y)$ describes the distribution of objects with respect to the sampler and is $1/w$ for line transects and $\frac{2r}{w}$ for point transects, taking into account the geometry of the sampler [@Buckland:2001vm, Chapter 3].
+where $\pi(y)$ describes the distribution of objects with respect to the sampler and $\pi(x)=1/w$ for line transects and $\pi(r)=\frac{2r}{w^2}$ for point transects, taking into account the geometry of the sampler [@Buckland:2001vm, Chapter 3] (letting $x$ denote a perpendicular distance from a line and $r$ denote radial distance from a point).
 
@@
 There is a wide literature on possible formulations for the detection function [@Buckland:1992fa; @Eidous:2005bj; @Becker:2009cj; @Giammarino:2014eg; @Miller:2015hw; @Becker:2015fi].
-\code{Distance} includes the most popular of these models, and includes an extendable class system to add new detection functions while avoiding code duplication (see "Extensions").
-
+\code{Distance} includes the most popular of these models.
 Here we'll detail the most popular detection function approach: "key function plus adjustments" (K+A).
@@
 
-Key function plus adjustment terms (or adjustment series) models are formulated by taking a "key" function and optionally adding "adjustments" to it to improve the fit.
+Key function plus adjustment terms (or adjustment series) models are formulated by taking a "key" function and optionally adding "adjustments" to it to improve the fit [@Buckland:1992fa].
 Mathematically we formulate this as:
@@
 
-Some models for the key are as follows:
+Available models for the key are as follows:
 $$
@@
 
-We do not always use adjustments (except in the case of the uniform key), in which case we refer to "key only" models; see also "Covariates" and "Model checking and model selection" below.
+We do not always include adjustments (except in the case of the uniform key), in which case we refer to "key only" models (see the next section and "Model checking and model selection").
 
@@
 
-Figure \ref{fig:minkeamakihi} (left panel) shows the result of calling \code{plot} on the resulting model object.
+Figure \ref{fig:minkeamakihi} (top left plot) shows the result of calling \code{plot} on the resulting model object.
 We can also call \code{summary} on the model object to get summary information about the fitted model (though we postpone this until the next section).
@@
 By default \code{ds} assumes that the data given to it is line transect, but we can switch to points using \code{transect="point"}.
-Including covariates in the scale is via the \code{formula=~...} argument to \code{ds}; a hazard-rate model for the amakihi that includes observer as a covariate (and truncating at 82.5m, given in @Marques:2007ey) can be specified by:
+Including covariates in the scale is via the \code{formula=~...} argument to \code{ds}; a hazard-rate model for the amakihi that includes observer as a covariate can be specified by [truncating at 82.5m, given in @Marques:2007ey]:
 ```{r amakihi-hr-obs}
@@
 ```
-As with the minke whale model, we can plot the resulting detection function (Figure \ref{fig:minkeamakihi}, centre and right panels).
-*Since for the amakihi we used covariates in the detection function, the plot will show the detection function averaged over levels/values of the covariate and separately for each level set (for continuous covariates, the 25\%, 50\% and 75\% quantiles are displayed).*
-
-**It would be nice if I could get "nice" plotting working for `Distance` as they are much easier to interpret (IMO)**
-
-*When multiple covariates are used in the model, a series of plots are produced.
-We can see this in the bottom row of Figure \ref{fig:minkeamakihi}, where we plot the results of fitting following model:*
+As with the minke whale model, we can plot the resulting detection function (Figure \ref{fig:minkeamakihi}, centre column).
+Since for the amakihi we used covariates in the detection function, the plot will show the detection function averaged over levels/values of the covariate.
+Points on the plot indicate probability of detection  for each observation.
+For the `amakihi_hr_obs` model we see fairly clear levels of the observer covariate in the points.
+Looking at the left panel of Figure \ref{fig:minkeamakihi}, we can see this is less clear when adding minutes after sunrise as a covariate into the model:
 ```{r amakihi-hr-obs-mas, warning=FALSE}
@@
 
-```{r minke-amakihi-hn-plot, echo=FALSE, warning=FALSE, fig.width=9, fig.height=3, fig.cap="Plot of fitted detection functions overlayed on the histograms of observed distances.
-Left minke whale data, half-normal model; centre, amakihi data hazard-rate with observer as a covariate; right, amakihi data, hazard-rate model with observer and minutes after sunrise as covariates.\\label{fig:minkeamakihi}"}
-par(mfrow=c(1,3))
-plot(minke_hn)
+The bottom row of Figure \ref{fig:minkeamakihi} shows the probability density functions for the amakihi models.
+These often are easier to interpret than the detection functions, as for point data we must rescale the histogram when plotting the detection function to take into account the geometry of the point sampler.
+
+```{r minke-amakihi-hn-plot, echo=FALSE, warning=FALSE, fig.width=9, fig.height=6, fig.cap="Top row plots of fitted detection functions overlayed on the histograms of observed distances; left minke whale data, half-normal model; centre, amakihi data hazard-rate with observer as a covariate; right, amakihi data, hazard-rate model with observer and minutes after sunrise as covariates (for amakihi histograms are rescale to take into account sampler geometry).
+Bottom row shows plots of the probability density function for the amakihi models (as top row).
+For covariate models (centre and right columns) points indicate probability of detection for a given observation (given that observations covariate values) and lines indicate the average detection function.\\label{fig:minkeamakihi}"}
+par(mfrow=c(2,3))
+plot(minke_hn, showpoints=FALSE)
 plot(amakihi_hr_obs)
 plot(amakihi_hr_obs_mas)
+plot(1,1, type="n", axes=FALSE, xlab="", ylab="")
+plot(amakihi_hr_obs, pdf=TRUE)
+plot(amakihi_hr_obs_mas, pdf=TRUE)
 ```
@@
 
-As with models fitted using \code{lm} or \code{glm} in \proglang{R} we can use \code{summary} to give us some useful information about our fitted model.
+As with models fitted using \code{lm} or \code{glm} in \proglang{R}, we can use \code{summary} to give us some useful information about our fitted model.
 For example for our hazard-rate model for the amakihi, with observer as a covariate:
@@
 
-### Post-stratification
+### Stratification
 
@@
 
+We combine these two sources of variance by noting that squared coefficients of variation (approximately) add [@Goodman:1960dd].
+
 ## Estimating abundance and variance in R

### 2015-11-22 paper.Rmd
ER edits up to "Detection functions"
some edits simply done to .rmd, other thoughts include as italicised
sentences

@@
 
-It is clear that census-type surveys (quadrat or strip transects) are inefficient (requiring a lot of effort on the part of those in the field) and we should expect that not all objects (animals, plants, dung, etc) can be observed.
-Accounting for imperfect detectability is an important consideration when want to obtain accurate estimates of abundance [@LahozMonfort:2013bg].
+_before getting too deep into detection functions, would be nice to describe the 2-stage nature; adjust counts by detectability to get $\hat{N_c}$ then scale up $\hat{N_c}$ to get $\hat{N}$_
+
+Census-type surveys (quadrat or strip transects assuming perfect detectability within strips or quadrats) are inefficient (requiring considerable field effort) and we should expect that not all objects (animals, plants, dung, etc) can be observed.
+Accounting for imperfect detectability is an important consideration when obtaining accurate estimates of abundance [@LahozMonfort:2013bg].
 Using the extra information gained by recording distance from the sampler to the observation, it is possible to model detectability.
-Since we expect detectability to decrease with increasing distance from the sampler, we model detectability as a function of distance (plus perhaps other covariates, see below).
+Because we expect detectability to decrease with increasing distance from the sampler, we model detectability as a function of distance (plus perhaps other covariates, see below).
 
-Distance sampling comes in two main "flavours": line and point transects.
-In line transect sampling observers walk (or fly, sail, etc) down lines observing objects and recording the distances to the line; whereas in point transect sampling observers stand stationary at a location and record distances from that point.
+Distance sampling surveys are conducted using two transect types: line and point transects.
+In line transect sampling observers walk (or fly, sail, etc)  lines observing objects and recording the distances to the line; whereas in point transect sampling observers remain stationary at a location and record distances from that point.
 Field methods are chosen to be suitable to species and habitat constraints [@buckland2015distance].
@@
 
-For both points and lines, once the geometry of the sampler has been taken into account, the histogram of distances should show a decreasing number of observations with increasing distance from the sampler.
+For both points and lines, once the geometry of the sampler _<likely spell out geom of sampler>_ has been taken into account, the histogram of distances should show a decreasing number of observations with increasing distance from the sampler.
 For line transects we expect objects to be distributed unformly with respect to distance from the line and what makes our histogram decrease is the detectability.
-For point transects, we must take into account the fact that as distance from the point increases, the area of the circle encompassed increases with distance squared.
+For point transects, we recognise that as distance from the point increases, the area of the circle encompassed increases with distance squared; hence the number of objects available to be detected is a linearly increasing function of distance from the point.
 
@@
 
-The rest of the paper is structured as follows: we first look into how data needs to be organised to allow it to be used with \pkg{Distance}; models for the detection function are then covered including their formulation and examples of fitting in \proglang{R}.
-We then spend some time looking at model checking and goodness of fit testing, as well as model selection.
-Having illustrated how to obtain a good detection function, we show how to estimate abundance using that model, including using post-stratification.
+The rest of the paper has this structure: we describe data organisation for use with \pkg{Distance}; models for the detection function are described in terms of formulation and examples of fitting in \proglang{R}.
+We then examine model checking, goodness of fit and model selection.
+Having illustrated how to obtain a good detection function, we show how to estimate abundance using that model, including stratification.
 The final two sections of the article look at extensions (both in terms of methodology and software) and put the package in a broader context amongst other software packages used for estimating the abundance of biological populations.
@@
 
-The two example data sets used here are distributed with \pkg{Distance} so readers can reproduce our results easiy.
-However in general we expect that data will be collected in the field and need to be formatted correctly for use with \pkg{Distance}.
+The two example data sets used here are distributed with \pkg{Distance} so readers can reproduce our results.
+However, data will be collected in the field and formatted correctly for use with \pkg{Distance}.
 The package allows for a flexible format for data input ranging from very simple to complex:
@@
 * In the simplest case, where one would simply like to estimate a detection function, all that is needed is a vector of distances.
-* To include additional covariates into the detection function (see "Detection functions") we need to use a \code{data.frame}, this needs to have a column called \code{distance} and named columns for each covariate collected (for example \code{observer} or \code{seastate}).
-Some column names are reserved: \code{object} for an observation identifier (see "Extensions"), \code{size} for group or cluster size (see "Detection functions" and "Abundance and variance estimation"), \code{detected} for whether or not an observation was detected (see "Extensions") and the columns described in the next bullet.
-* If one would like to estimate abundance for some area, this information must be given at the same time.
+* To include additional covariates into the detection function (see "Detection functions") we need to use a \code{data.frame}.
+The data.frame contains a column called \code{distance} and additional named columns for covariates potentially useful in modelling detectability (for example \code{observer} or \code{seastate}).
+Some column names are reserved: \code{object} for an observation identifier (see "Extensions"), \code{size} for group or cluster size (see "Detection functions" and "Abundance and variance estimation"), \code{detected} for whether  an observation was detected (see "Extensions") and the columns described in the next bullet.
+* If one would like to estimate abundance beyond area covered by expended survey effort, additional information is required.
 This consists of information about which transect and occasion the observation was made (the \code{Sample.Label}), a column named \code{Effort} which gives the effort associated with that sample (for lines their length and for points the number of times that point was visited), the stratum the sample was located in (this may have any name and may be from pre- or post-survey stratification, see "Estimating abundance and variance") and that stratum's area (which has the same name as the stratum column, appended with \code{.Area}).
-(We refer to this data format as "flatfile" is all information we have is contained in one table.)
+(We refer to this data format as "flatfile" where all information is contained in one table.)
 
@@
 ```
-Where as the amakihi data lacks effort and stratum data:
+Whereas the amakihi data lacks effort and stratum data:
 ```{r amakihi-data-head}
@@
 To ensure that the model is not overly influence by distances far from zero we truncate the distances beyond a given distance $w$ (known as the *truncation distance*).
-Fitting to distances further away from zero does not improve the precision of abundance estimates considerably [@Buckland:2001vm, 103–108, 151–153].
+Fitting to distances at great distances from zero does not demonstrably improve the precision of abundance estimates [@Buckland:2001vm, 103–108, 151–153].
 
@@
 
-We wish to obtain the abundance in a study region, which we have sampled some (representative) subset of.
+We wish to obtain the abundance in a study region, of which we have sampled a (representative) subset.
 To do this we first calculate the abundance in the area we have surveyed (the *covered area*) to obtain $\hat{N}_\text{C}$, we can then scale this up to the full study area by multiplying it by the ratio of covered area to study area.

### 2015-11-22 paper.Rmd
Non-ascii characters replaced
Long dashes for 103-108 and 151-153 replaced.

@@
 
-For both points and lines, once the geometry of the sampler _<likely spell out geom of sampler>_ has been taken into account, the histogram of distances should show a decreasing number of observations with increasing distance from the sampler.
+For both points and lines, once the geometry of the sampler _likely spell out geom of sampler, or note it specification is forthcoming_ has been taken into account, the histogram of distances should show a decreasing number of observations with increasing distance from the sampler.
 For line transects we expect objects to be distributed unformly with respect to distance from the line and what makes our histogram decrease is the detectability.
@@
 The line transect data is simulated data based on a survey of Antarctic minke whales (*Balaenoptera bonaerensis*).
-The data is simulated from models fitted to data from the International Whaling Commission’s International Decade of Cetacean Research Southern Ocean Whale and Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer surveys [@Branch:2001ua].
+The data is simulated from models fitted to data from the International Whaling Commission's International Decade of Cetacean Research Southern Ocean Whale and Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer surveys [@Branch:2001ua].
 They consist of 99 observations and include information on the effort expended and whether observations were in one of two geographical strata (near or far from land).
@@
 To ensure that the model is not overly influence by distances far from zero we truncate the distances beyond a given distance $w$ (known as the *truncation distance*).
-Fitting to distances at great distances from zero does not demonstrably improve the precision of abundance estimates [@Buckland:2001vm, 103–108, 151–153].
+Fitting to distances at great distances from zero does not demonstrably improve the precision of abundance estimates [@Buckland:2001vm, 103-108, 151-153].
 
@@
 These include things like the observer, the vessel or platform used, the sea state, weather conditions and time of day (to name but a few).
-We assume that these variables affect detection only via the scale of the detection function (and that they don't affect the shape).
+We assume that these variables affect detection only via the scale of the detection function (and do not affect the shape).
 
@@
 The EDF gives the proportion of observations for which the CDF is less than or equal to that of a given distance.
-In essence we're asking "is the number of observations up to a given distance in line with what the model says they should be?" (where “given values” are the observed distances).
+In essence we're asking "is the number of observations up to a given distance in line with what the model says they should be?" (where 'given values' are the observed distances).
 As usual for Q-Q plots, "good" models will have values close to the line $y=x$, poor models will show more deviations from that line.

### 2015-11-22 paper.tex
Non-ascii characters replaced
Long dashes for 103-108 and 151-153 replaced.

\documentclass[article]{jss}
\usepackage[utf8]{inputenc}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\author{
David L Miller\\University of St Andrews \And Jeffrey L Laake\\National Marine Mammal Laboratory
}
\title{Distance Sampling in \proglang{R}}
\Keywords{distance sampling, abundance estimation, line transects, point transects, \proglang{R}}

\Abstract{
The abstract of the article.
}

\Plainauthor{David L Miller, Jeffrey L Laake}
\Plaintitle{Distance Sampling in R}
\Plainkeywords{distance sampling, abundance estimation, line transects, point transects, R}

%% publication information
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
\Submitdate{}
%% \Acceptdate{2012-06-04}

\Address{
    David L Miller\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling, The
  Observatory, St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \href{mailto:dave@ninepointeightone.net}{\nolinkurl{dave@ninepointeightone.net}}\\
  URL: \url{http://converged.yt}\\~\\
      Jeffrey L Laake\\
  National Marine Mammal Laboratory\\
  Alaska Fisheries Science Center 7600 Sand Point Way N.E., Seattle, WA
  98115, USA\\
  E-mail: \href{mailto:Jeff.Laake@noaa.gov}{\nolinkurl{Jeff.Laake@noaa.gov}}\\

  }

\usepackage{amsmath} \usepackage{amssymb} \usepackage{longtable}
\usepackage{booktabs}

\begin{document}

\section{Introduction}\label{introduction}

Distance sampling (Buckland et al. 2001; Buckland et al. 2004)
encompases a suite of field methods and statistical models used to
estimate the abundance of biological populations. Distance sampling
field procedure can be thought of as an extension of plot sampling,
where we wish to take into account the decreasing probability of
detecting objects at increasing distance from the sampler. We do this by
building a model for detectability and use quantities calculated from
the detection function to adjust the observed counts to obtain an
estimate of abundance.

For many years distance sampling analyses have been available via the
Windows program Distance (or ``DISTANCE''); for clarity henceforth
``Distance for Windows'' Thomas et al. (2010)). From version 5 of
Distance for Windows, R packages have been included to perform
particular analyses (CITE Distance user manual). This paper shows how to
fit detection functions, perform model checking and selection and
estimate abundance in \proglang{R} using the package \pkg{Distance}.
\pkg{Distance} is a wrapper package around the more complex (and more
powerful) \pkg{mrds} and offers a subset of the analyses possible with
that package.

\subsection{Distance sampling}\label{distance-sampling}

\emph{before getting too deep into detection functions, would be nice to
describe the 2-stage nature; adjust counts by detectability to get
\(\hat{N_c}\) then scale up \(\hat{N_c}\) to get \(\hat{N}\)}

Census-type surveys (quadrat or strip transects assuming perfect
detectability within strips or quadrats) are inefficient (requiring
considerable field effort) and we should expect that not all objects
(animals, plants, dung, etc) can be observed. Accounting for imperfect
detectability is an important consideration when obtaining accurate
estimates of abundance (Lahoz-Monfort, Guillera-Arroita, and Wintle
2013). Using the extra information gained by recording distance from the
sampler to the observation, it is possible to model detectability.
Because we expect detectability to decrease with increasing distance
from the sampler, we model detectability as a function of distance (plus
perhaps other covariates, see below).

Distance sampling surveys are conducted using two transect types: line
and point transects. In line transect sampling observers walk (or fly,
sail, etc) lines observing objects and recording the distances to the
line; whereas in point transect sampling observers remain stationary at
a location and record distances from that point. Field methods are
chosen to be suitable to species and habitat constraints (Buckland et
al. 2015).

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics{paper_files/figure-latex/points-and-lines-1}

}

\caption{Left side plots show an example of a survey of an area containing a population of 500 objects, blue indicates sampler placement (top lines, bottom points) and red dots indicate detected individuals. The right side of the figure shows histograms of observed distances (again, lines top and points bottom).\label{fig:pointslines}}\label{fig:points-and-lines}
\end{figure}
\end{CodeChunk}

For both points and lines, once the geometry of the sampler \emph{likely
spell out geom of sampler, or note it specification is forthcoming} has
been taken into account, the histogram of distances should show a
decreasing number of observations with increasing distance from the
sampler. For line transects we expect objects to be distributed unformly
with respect to distance from the line and what makes our histogram
decrease is the detectability. For point transects, we recognise that as
distance from the point increases, the area of the circle encompassed
increases with distance squared; hence the number of objects available
to be detected is a linearly increasing function of distance from the
point.

Using this histogram we can crudely estimate the drop-off in
detectability by eye by tracing a line that approximates the tops of the
histogram bars -- this is the detection function. \pkg{Distance}
estimates the parameters for a fixed-form detection function using
maximum likelihood estimation. We address possible models in detail
below.

Figure \ref{fig:pointslines} shows examples of 500 individuals sampled
using line and point transects (left column) and their corresponding
histograms (right column).

\subsection{Data}\label{data}

We demonstrate \pkg{Distance} using two data sets: one line transect and
one point transect. These data sets have been chosen to be
representative of the kind of data seen in practice.

\subsubsection{Minke whales}\label{minke-whales}

The line transect data is simulated data based on a survey of Antarctic
minke whales (\emph{Balaenoptera bonaerensis}). The data is simulated
from models fitted to data from the International Whaling Commission's
International Decade of Cetacean Research Southern Ocean Whale and
Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer
surveys (Branch and Butterworth 2001). They consist of 99 observations
and include information on the effort expended and whether observations
were in one of two geographical strata (near or far from land).

\subsubsection{Amakihi}\label{amakihi}

The point transect data set consists of 1485 observations of Amakihi
(\emph{Hemignathus virens}; a Hawaiian songbird), collected at 41 points
between 1992 and 1995. The data include distances and three covariates
collected during the survey: observer ID (a three level factor), minutes
after sunrise (continuous) and hours after sunrise (a six level factor).
Data are analysed comprehensively in Marques et al. (2007).

\subsection{The rest of the paper}\label{the-rest-of-the-paper}

The rest of the paper has this structure: we describe data organisation
for use with \pkg{Distance}; models for the detection function are
described in terms of formulation and examples of fitting in
\proglang{R}. We then examine model checking, goodness of fit and model
selection. Having illustrated how to obtain a good detection function,
we show how to estimate abundance using that model, including
stratification. The final two sections of the article look at extensions
(both in terms of methodology and software) and put the package in a
broader context amongst other software packages used for estimating the
abundance of biological populations.

\section{Data setup}\label{data-setup}

The two example data sets used here are distributed with \pkg{Distance}
so readers can reproduce our results. However, data will be collected in
the field and formatted correctly for use with \pkg{Distance}. The
package allows for a flexible format for data input ranging from very
simple to complex:

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  In the simplest case, where one would simply like to estimate a
  detection function, all that is needed is a vector of distances.
\item
  To include additional covariates into the detection function (see
  ``Detection functions'') we need to use a \code{data.frame}. The
  data.frame contains a column called \code{distance} and additional
  named columns for covariates potentially useful in modelling
  detectability (for example \code{observer} or \code{seastate}). Some
  column names are reserved: \code{object} for an observation identifier
  (see ``Extensions''), \code{size} for group or cluster size (see
  ``Detection functions'' and ``Abundance and variance estimation''),
  \code{detected} for whether an observation was detected (see
  ``Extensions'') and the columns described in the next bullet.
\item
  If one would like to estimate abundance beyond area covered by
  expended survey effort, additional information is required. This
  consists of information about which transect and occasion the
  observation was made (the \code{Sample.Label}), a column named
  \code{Effort} which gives the effort associated with that sample (for
  lines their length and for points the number of times that point was
  visited), the stratum the sample was located in (this may have any
  name and may be from pre- or post-survey stratification, see
  ``Estimating abundance and variance'') and that stratum's area (which
  has the same name as the stratum column, appended with \code{.Area}).
  (We refer to this data format as ``flatfile'' where all information is
  contained in one table.)
\end{itemize}

As we will see in ``Extensions'', further information is also required
when we start using more complex models.

The minke whale data follows the ``flatfile'' format given in the last
bullet point:

\begin{CodeChunk}
\begin{CodeInput}
library(Distance)
head(minke)
\end{CodeInput}
\begin{CodeOutput}
  Region.Label  Area Sample.Label Effort distance
1        South 84734            1  86.75     0.10
2        South 84734            1  86.75     0.22
3        South 84734            1  86.75     0.16
4        South 84734            1  86.75     0.78
5        South 84734            1  86.75     0.21
6        South 84734            1  86.75     0.95
\end{CodeOutput}
\end{CodeChunk}

Whereas the amakihi data lacks effort and stratum data:

\begin{CodeChunk}
\begin{CodeInput}
head(amakihi)
\end{CodeInput}
\begin{CodeOutput}
   survey object distance obs mas has detected
1 July 92      1       40 TJS  50   1        1
2 July 92      2       60 TJS  50   1        1
3 July 92      3       45 TJS  50   1        1
4 July 92      4      100 TJS  50   1        1
5 July 92      5      125 TJS  50   1        1
6 July 92      6      120 TJS  50   1        1
\end{CodeOutput}
\end{CodeChunk}

We'll explore what the effects of including effort data are during
analysis below.

\section{Detection functions}\label{detection-functions}

As mentioned above, the detection function describes the relationship
between observed distances and probability of detection. The detection
function itself models the probability
\(\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)\)
and is usually denoted \(g(y; \boldsymbol{\theta})\) where \(y\) is
distance (either from a line or point) and \(\boldsymbol{\theta}\) is a
vector of parmeters to be estimated. Our goal is to estimate an
\emph{average probability of detection} (\(p\), average in the sense of
an average over the distances), so we must integrate out distance
(\(y\)) from the detection function: \[
p = \int_0^w \pi(y) g(y; \boldsymbol{\theta}) dy
\] where \(\pi(y)\) describes the distribution of objects with respect
to the sampler and \(\pi(x)=1/w\) for line transects and
\(\pi(r)=\frac{2r}{w^2}\) for point transects, taking into account the
geometry of the sampler (Buckland et al. 2001, Chapter 3) (letting \(x\)
denote a perpendicular distance from a line and \(r\) denote radial
distance from a point).

It is important that the detection function accurately models
detectability near zero distance. We are less worried by its behaviour
further away from 0. To ensure that the model is not overly influence by
distances far from zero we truncate the distances beyond a given
distance \(w\) (known as the \emph{truncation distance}). Fitting to
distances at great distances from zero does not demonstrably improve the
precision of abundance estimates (Buckland et al. 2001, 103--8,
151--53).

Models for the detection function are expected to have the following
properties (Buckland et al. 2015, Chapter 5):

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  \emph{Shoulder}: we expect observers to be able to see objects near
  them, not just those directly infront of them. For this reason, we
  expect the detection function to be ``flat'' near zero distance.
\item
  \emph{Non-increasing}: we don't think that observers should be more
  likely to see things further away than those nearer to them. This
  usually indicates an issue with field procedure (that the distribution
  of objects with respect to the line, \(\pi(y)\) is not what we
  expect), so we do not want the detection function to model this.
\item
  \emph{Model robust}: models should be flexible enough to make many
  different shapes.
\item
  \emph{Pooling robust}: many factors can affect the probability of
  detection and it is not possible to measure all of these. We would
  like our models to be robust to us not including these factors.
\item
  \emph{Estimator efficiency}: we would like our models to have low
  variances, but only given they satisfy the other properties above
  (which, if they are satisfied, would give low bias).
\end{itemize}

Given these criteria, we can start to think about models for \(g\).

\subsection{Formulations}\label{formulations}

There is a wide literature on possible formulations for the detection
function (Buckland 1992; Eidous 2005; Becker and Quang 2009; Giammarino
and Quatto 2014; Miller and Thomas 2015; Becker and Christ 2015).
\code{Distance} includes the most popular of these models. Here we'll
detail the most popular detection function approach: ``key function plus
adjustments'' (K+A).

\subsubsection{Key function plus adjustments
(K+A)}\label{key-function-plus-adjustments-ka}

Key function plus adjustment terms (or adjustment series) models are
formulated by taking a ``key'' function and optionally adding
``adjustments'' to it to improve the fit (Buckland 1992). Mathematically
we formulate this as: \[
g(y; \boldsymbol{\theta}) = k(y; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(y; \boldsymbol{\theta}_\text{adjust})\right),
\] where \(k\) is the key function and \(\alpha_O\) is sum series of
functions (given in Table \ref{tab:keyadj}), described as an
\emph{adjustment of order \(O\)}. Subscripts on the parameter vector
indicate those parameters belonging to each part of the model (i.e.
\(\boldsymbol{\theta} = (\boldsymbol{\theta}_\text{key}, \boldsymbol{\theta}_\text{adjust})\)).

Available models for the key are as follows: \[
k(y) = \left\{
\begin{array}{l l}
  \exp\left(-\frac{y^2}{2 \sigma^2}\right) & \quad \text{half-normal,} \\
  1-\exp\left(\left(-\frac{y}{\sigma}\right)^{-b}\right) & \quad \text{hazard-rate,} \\
  1/w & \quad \text{uniform.}
\end{array} \right.
\] Possible modelling options for key and adjustments are given in Table
\ref{tab:keyadj} and illustrated in Figure \ref{fig:hnhr} and
\ref{fig:keyadj}. We select the number of adjustment terms (\(K\)) by
AIC (further details in ``Model checking and model selection'').

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics{paper_files/figure-latex/hn-hr-par-comp-1}

}

\caption{Half-normal (top row) and hazard-rate (bottom row) detection functions without adjustments, varying scale ($\sigma$) and (for hazard-rate) shape ($b$) parameters (values are given above the plots). On the top row from left to right, the study species becomes more detectable (higher probability of detection at larger distances). The bottom row shows the hazard-rate model's more pronounced shoulder.\label{fig:hnhr}}\label{fig:hn-hr-par-comp}
\end{figure}
\end{CodeChunk}

\begin{table}
\caption{Modelling options for key plus adjustment series models for the detection function.}
\begin{tabular}{llll}
\hline
Key function   & Form   & Adjustment series & Form\\
\hline
 Uniform  & $1/w$   & cosine  & $\sum_{o=1}^O a_o \cos(o \pi y/w)$ \\
 & & Simple polynomial & $\sum_{o=1}^O a_o (y/w)^{2o}$ \\
 Half-normal  & $\exp\left(-\frac{y^2}{2 \sigma^2}\right)$ & cosine  & $\sum_{o=2}^O a_o \cos(o \pi y/w)$ \\
 & & Hermite polynomial & $\sum_{o=2}^O a_o H_{2o}(y/\sigma)$ \\
 Hazard-rate  & $1-\exp\left[-\left(\frac{y}{\sigma}\right)^{-b}\right]$ & cosine  & $\sum_{o=2}^O a_o \cos(o \pi y/w)$ \\
 & & Simple polynomial & $\sum_{o=2}^O a_o (y/w)^{2o}$ \\
\hline
\end{tabular}
\label{tab:keyadj}
\end{table}

When adjustment terms are used it may be necessary to standardise the
results to ensure that \(g(0)=1\), so we can redefine the detection
function as: \[
g(y; \boldsymbol{\theta}) = \frac{k(y; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(y; \boldsymbol{\theta}_\text{adjust})\right)}{k(0; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(0; \boldsymbol{\theta}_\text{adjust})\right)}.
\]

A disadvantage of K+A models is that we must resort to constrained
optimisation (via the \pkg{Rsolnp} package) in order to ensure that the
resulting detection function is monotonic non-increasing over the whole
range.

We do not always include adjustments (except in the case of the uniform
key), in which case we refer to ``key only'' models (see the next
section and ``Model checking and model selection'').

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics{paper_files/figure-latex/adjust-mix-comp-1}

}

\caption{Possible shapes for the detection function when adjustments are included for half-normal and hazard-rate models.\label{fig:keyadj}}\label{fig:adjust-mix-comp}
\end{figure}
\end{CodeChunk}

\subsubsection{Covariates}\label{covariates}

There are many factors that can affect the probability of detecting an
object. These include things like the observer, the vessel or platform
used, the sea state, weather conditions and time of day (to name but a
few). We assume that these variables affect detection only via the scale
of the detection function (and do not affect the shape).

Covariates can be included in this formulation by considering the scale
parameter from the half-normal or hazard-rate detection functions as a
linear model (on the exponential scale) of the (\(J\)) covariates
(\(\mathbf{z}\); a vector of length \(J\) for each observation): \[
\sigma(\mathbf{z}) = \exp(\beta_0 + \sum_{j=1}^J \beta_j z_j).
\] In the next section we'll discuss model selection.

Including covariates has an important implication for our calculation of
detectability. Since we don't know what the true distribution of the
covariates is we must calculate the probability of detection conditional
on the observed values of the covariates: \[
p(\mathbf{z_i}) = \int_0^w \pi(y) g(y, \mathbf{z_i}; \boldsymbol{\theta}) dy,
\] where \(\mathbf{z_i}\) is the vector of \(J\) covariates associated
with observation \(i\). So for covariate models, we are calculating a
value of ``average'' probability of detection (again the average is in
the sense of distance) per observation. There will be as many unique
values of \(p(\mathbf{z_i})\) as there are unique covariate combinations
in our data.

Another important consideration is that K+A models which include
covariates and one or more adjustments cannot be guaranteed to be
monotonic non-increasing for all covariate combinations, as we don't
have any model for the distribution of the covariates. For this reason,
we advise against using both adjustments and covariates in a detection
function (see Miller and Thomas 2015 for an example of when this can be
problematic).

\subsection{Fitting detection functions in
R}\label{fitting-detection-functions-in-r}

The workhorse of detection function fitting in \pkg{Distance} is the
\code{ds} function. Here we show off the formulations for the detection
function that we've seen above for both the minke whale and amakihi
data.

\subsubsection{Minke whale}\label{minke-whale}

We can fit a model to the minke whale data, setting the truncation at
1.5km and using the default options in \code{ds} very simply:

\begin{CodeChunk}
\begin{CodeInput}
minke_hn <- ds(minke, truncation=1.5)
\end{CodeInput}
\begin{CodeOutput}
Starting AIC adjustment term selection.
Fitting half-normal key function
Key only models do not require monotonicity contraints. Not constraining model for monotonicity.
AIC= 46.872
Fitting half-normal key function with cosine(2) adjustments
AIC= 48.872

half-normal key function selected!
\end{CodeOutput}
\end{CodeChunk}

Note that \code{ds} will automatically select adjustment terms by AIC
and shows it's selection steps as it goes.

Figure \ref{fig:minkeamakihi} (top left plot) shows the result of
calling \code{plot} on the resulting model object. We can also call
\code{summary} on the model object to get summary information about the
fitted model (though we postpone this until the next section).

We can specify the form of the detection function via the \code{key=}
argument to \code{ds}. For example, a hazard rate model can be fitted
as:

\begin{CodeChunk}
\begin{CodeInput}
minke_hrcos <- ds(minke, truncation=1.5, key="hr")
\end{CodeInput}
\begin{CodeOutput}
Starting AIC adjustment term selection.
Fitting hazard-rate key function
Key only models do not require monotonicity contraints. Not constraining model for monotonicity.
AIC= 48.637
Fitting hazard-rate key function with cosine(2) adjustments
AIC= 50.386

hazard-rate key function selected!
\end{CodeOutput}
\end{CodeChunk}

Again, \code{ds} fits a model with cosine adjustments (the default) but
finds the AIC improvement to be insufficient to select the adjustment.

\subsubsection{Amakihi}\label{amakihi-1}

By default \code{ds} assumes that the data given to it is line transect,
but we can switch to points using \code{transect="point"}. Including
covariates in the scale is via the \code{formula=~...} argument to
\code{ds}; a hazard-rate model for the amakihi that includes observer as
a covariate can be specified by (truncating at 82.5m, given in Marques
et al. 2007):

\begin{CodeChunk}
\begin{CodeInput}
amakihi_hr_obs <- ds(amakihi, truncation=82.5, transect="point",
                     key="hr", formula=~obs)
\end{CodeInput}
\begin{CodeOutput}
Cannot perform AIC adjustment term selection when covariates are used.
Fitting hazard-rate key function
AIC= 10778.448
No survey area information supplied, only estimating detection function.
\end{CodeOutput}
\end{CodeChunk}

As with the minke whale model, we can plot the resulting detection
function (Figure \ref{fig:minkeamakihi}, centre column). Since for the
amakihi we used covariates in the detection function, the plot will show
the detection function averaged over levels/values of the covariate.
Points on the plot indicate probability of detection for each
observation. For the \texttt{amakihi\_hr\_obs} model we see fairly clear
levels of the observer covariate in the points. Looking at the left
panel of Figure \ref{fig:minkeamakihi}, we can see this is less clear
when adding minutes after sunrise as a covariate into the model:

\begin{CodeChunk}
\begin{CodeInput}
amakihi_hr_obs_mas <- ds(amakihi, truncation=82.5, transect="point",
                         key="hr", formula=~obs+mas)
\end{CodeInput}
\begin{CodeOutput}
Cannot perform AIC adjustment term selection when covariates are used.
Fitting hazard-rate key function
AIC= 10777.376
No survey area information supplied, only estimating detection function.
\end{CodeOutput}
\end{CodeChunk}

The bottom row of Figure \ref{fig:minkeamakihi} shows the probability
density functions for the amakihi models. These often are easier to
interpret than the detection functions, as for point data we must
rescale the histogram when plotting the detection function to take into
account the geometry of the point sampler.

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics{paper_files/figure-latex/minke-amakihi-hn-plot-1}

}

\caption{Top row plots of fitted detection functions overlayed on the histograms of observed distances; left minke whale data, half-normal model; centre, amakihi data hazard-rate with observer as a covariate; right, amakihi data, hazard-rate model with observer and minutes after sunrise as covariates (for amakihi histograms are rescale to take into account sampler geometry). Bottom row shows plots of the probability density function for the amakihi models (as top row). For covariate models (centre and right columns) points indicate probability of detection for a given observation (given that observations covariate values) and lines indicate the average detection function.\label{fig:minkeamakihi}}\label{fig:minke-amakihi-hn-plot}
\end{figure}
\end{CodeChunk}

\section{Model checking and model
selection}\label{model-checking-and-model-selection}

As with models fitted using \code{lm} or \code{glm} in \proglang{R}, we
can use \code{summary} to give us some useful information about our
fitted model. For example for our hazard-rate model for the amakihi,
with observer as a covariate:

\begin{CodeChunk}
\begin{CodeInput}
summary(amakihi_hr_obs)
\end{CodeInput}
\begin{CodeOutput}

Summary for distance analysis 
Number of observations :  1243 
Distance range         :  0  -  82.5

Model : Hazard-rate key function 
AIC   : 10778.45

Detection function parameters
Scale coefficient(s):  
              estimate         se
(Intercept) 3.06441705 0.10878121
obsTJS      0.53017364 0.09956539
obsTKP      0.08885471 0.18071851

Shape coefficient(s):  
             estimate         se
(Intercept) 0.8690009 0.06261764

                        Estimate          SE         CV
Average p              0.3142723   0.0204413 0.06504326
N in covered region 3955.1685709 274.2284029 0.06933419
\end{CodeOutput}
\end{CodeChunk}

This summary information includes various details of the data and model
specification, as well as the values of the coefficents (\(\beta_j\))
and their uncertainties, an ``average'' value for the detectability (see
``Estimating abundance and variance'' for details on how this is
calculated) and it's uncertainty. The final line gives an estimate of
abundance for the area covered by the survey (again, this will be
covered in detail in the next section).

\subsection{Goodness of fit}\label{goodness-of-fit}

To judge goodness of fit for detection functions when exact distances
are used, we want to compare the cumulative distribution function (CDF)
and empirical distribution function (EDF) for the detection function via
a quantile-quantile plot (Q-Q plot). In our case the CDF evaluates the
probability of observing an object at a distance less than or equal to
some given value. The EDF gives the proportion of observations for which
the CDF is less than or equal to that of a given distance. In essence
we're asking ``is the number of observations up to a given distance in
line with what the model says they should be?'' (where `given values'
are the observed distances). As usual for Q-Q plots, ``good'' models
will have values close to the line \(y=x\), poor models will show more
deviations from that line.

We can inspect Q-Q plots visually, though for a large number of models
this can be tiresome and prone to subjective judgements. Instead we can
quantify the Q-Q plot's information using a Kolmogorov-Smirnov or
Cramer-von Mises test (Burnham et al. 2004). Both test to see if the
points from the EDF and CDF are from the same distribution. The
Kolmogorov-Smirnov uses the test statistic of the largest difference
between a point on the Q-Q plot and the line \(y=x\), whereas the
Cramer-von Mises test uses the sum of all the differences. As it takes
into account more information and is therefore more powerful, the
Cramer-von Mises is generally preferred. A significant result from
either tests indicates that the EDF and CDF do not come from the same
distribution (and therefore the model is does not fit the data well).

We can generate a Q-Q plot and test results using the \code{gof_ddf}
function. We can see the goodness of fit tests below for two models for
the amakihi data, first fitting a half-normal model without covariates
or adjustments (not setting \code{adjustment=NULL} will force \code{ds}
to fit a model with no adjustments), then calculating goodness of fit
for that model and our hazard-rate model with observer and minutes after
sunrise included:

\begin{CodeChunk}
\begin{CodeInput}
amakihi_hr <- ds(amakihi, truncation=82.5, transect="point", key="hn", adjustment=NULL)
\end{CodeInput}
\begin{CodeOutput}
Fitting half-normal key function
Key only models do not require monotonicity contraints. Not constraining model for monotonicity.
AIC= 10833.841
No survey area information supplied, only estimating detection function.
\end{CodeOutput}
\begin{CodeInput}
gof_ds(amakihi_hr)
\end{CodeInput}
\begin{CodeOutput}

Goodness of fit results for ddf object

Distance sampling Kolmogorov-Smirnov test
Test statistic =  0.059345  P =  0.00031527

Distance sampling Cramer-von Mises test (unweighted)
Test statistic =  0.93083  P =  0.003578 
\end{CodeOutput}
\begin{CodeInput}
gof_ds(amakihi_hr_obs_mas)
\end{CodeInput}
\begin{CodeOutput}

Goodness of fit results for ddf object

Distance sampling Kolmogorov-Smirnov test
Test statistic =  0.036251  P =  0.076237

Distance sampling Cramer-von Mises test (unweighted)
Test statistic =  0.15016  P =  0.38908 
\end{CodeOutput}
\end{CodeChunk}

The corresponding Q-Q plots are shown in Figure \ref{amakihi-qq}.

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics{paper_files/figure-latex/amakihi-qq-comp-1}

}

\caption[Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates) and hazard-rate model with observer and minutes after sunrise for the amakihi data]{Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates) and hazard-rate model with observer and minutes after sunrise for the amakihi data.\label{amakihi-qq}}\label{fig:amakihi-qq-comp}
\end{figure}
\end{CodeChunk}

\subsection{Model selection}\label{model-selection}

Once we have a set of models which fit well, we can use Akaike's
Information Criterion (AIC) to select between models. \pkg{Distance}
includes a function to create table of summary information for fitted
models, making it easy to get an overview of a large number of models at
once. The \code{summarize_ds_models} function takes models as input and
can be especially useful when paired with \pkg{knitr}'s \code{kable}
function to create summary tables for publication. An example of this
output is shown in Table \ref{tab:amakihi} and was generated by:

\begin{CodeChunk}
\begin{CodeInput}
library(knitr)
summarize_ds_models(amakihi_hr, amakihi_hr_obs, amakihi_hr_obs_mas)
\end{CodeInput}
\end{CodeChunk}

\begin{table}

\caption{Summary for the detection function models fitted to the amakihi data. ``C-vM'' stands for Cramer-von Mises, $P_a$ indicates average detectability (see ``Estimating abundance and variance''), se indicates standard error. Models are sorted according to AIC.\label{tab:amakihi}}
\begin{tabular}{llrrrrr}
\toprule
Key function & Formula & C-vM $p$-value & $\hat{P_a}$ & se($\hat{P_a}$) & AIC & $\Delta$AIC\\
\midrule
Hazard-rate key function & ~obs + mas & 0.389 & 0.319 & 0.020 & 10777.38 & 0.000\\
Hazard-rate key function & ~obs & 0.271 & 0.314 & 0.020 & 10778.45 & 1.073\\
Half-normal key function & ~1 & 0.004 & 0.351 & 0.011 & 10833.84 & 56.465\\
\bottomrule
\end{tabular}
\end{table}

\section{Estimating abundance and
variance}\label{estimating-abundance-and-variance}

Though fiting the detection function is the primary modelling step in
distance sampling, we are really interested in estimating detectability,
and from that abundance. We also wish to calculate our certainty in each
abundance estimate. This section addresses these issues mathematically
before showing how to estimate abundance and its variance in R.

\subsection{Abundance}\label{abundance}

We wish to obtain the abundance in a study region, of which we have
sampled a (representative) subset. To do this we first calculate the
abundance in the area we have surveyed (the \emph{covered area}) to
obtain \(\hat{N}_\text{C}\), we can then scale this up to the full study
area by multiplying it by the ratio of covered area to study area.

First, in order to estimate abundance in the covered area
(\(\hat{N}_\text{C}\)), we use the estimates of detection probability
(the \(\{p_i; i=1,\ldots,n\}\), above) in a Horvitz-Thompson like
estimator:

\begin{equation}
\hat{N}_\text{C} = \sum_{i=1}^n\frac{s_i}{\hat{p}_i},
\label{ht}
\end{equation}

where \(s_i\) are the sizes of the observed groups of objects, which is
equal to 1 if objects only occur singly (Borchers and Burnham 2004).
Thompson (2002) is the cannonical reference to this type of estimator;
intuitively, we can think of the estimates of detectability
(\(\hat{p}_i\)) as ``inflating'' the group sizes (\(s_i\)) to account
for incomplete detection -- we then sum to obtain the abundance
estimate. For models that do not include covariates, \(\hat{p}_i\) is
equal for all \(i\), so this is equivalent to summing the groups and
inflating that sum by the corresponding
\(\hat{p} (=\hat{p}_i \forall i)\).

Having obtained the abundance in the covered area, we can then scale-up
to the study area: \[
\hat{N} = \frac{A}{a} \hat{N}_\text{C},
\] where \(A\) is the area of the study region to extrapolate the
abundance estimate to and \(a\) is the covered area. For line transects
\(a=2wL\) (twice the truncation distance multiplied by the total length
of transects surveyed) and for points \(a=\pi w^2 K\).

We can use the Horvitz-Thompson-like estimator to calculate the
``average'' detectability for models which include covariates. We can
consider what single detectability value would give the estimated
\(\hat{N}\) and therefore calculate: \[
\hat{P_a} = n/\hat{N}_\text{C}.
\] This can be a useful summary statistic and is included in the
\code{summary} output and the table produced by
\code{summarize_ds_models}.

\subsubsection{Stratification}\label{stratification}

We may wish to calculate abundance estimates for some sub-regions of the
study region, we call these areas \emph{strata}. A stratum may be
defined by, for example, a different habitat type or by the sex of the
animal (or some combination) which may be of interest for biological or
management reasons. In order to calculate estimates for a given
stratification each observation must occur in a stratum which must be
labelled with a \code{Region.Label} and have a corresponding \code{Area}
(if we are using an animal characteristic like sex, we would have the
areas be the same but if we were using say forested vs.~wetland habitat
the areas of those strata would be different). Finally, we must also
know which stratum a given transect lies in.

As an example the minke whale data consists of two strata: \code{North}
and \code{South} relating to a stratum further away and nearer the
Antarctic ice edge, respectively. Figure \ref{fig:minke-strata} shows
the two strata, along with observations and transect lines.

\begin{figure}
\includegraphics{minke-strata}
\label{fig:minke-strata}
\caption{Strata used for the minke whale data adapted from @Hedley:2004et. Points show the locations of observations along transect lines. The stepped line shows the boundary between North and South strata. Further details on the survey are available in @Branch:2001ua (simulated data is based on ``1992/93 Area III'' therein).}
\end{figure}

\subsection{Variance}\label{variance}

Here we take an intuitive approach to uncertainty estimation, for a full
derivations consult F. F. C. Marques and Buckland (2003). Uncertainty in
\(\hat{N}\) comes from two sources:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  \emph{Model uncertainty}, from the estimation of the detection
  function parameters \(\boldsymbol{\theta}\)
\item
  \emph{Sampling uncertainty}, from the distribution of objects along
  the transect lines or between visiting occasions for points.
\end{enumerate}

We can see this by looking at the Horvitz-Thompson estimation in
(\ref{ht}) and consider the terms which are random. These are: the
detectability \(\hat{p}_i\) (and hence the parameters of the detection
function it is derived from) and \(n\), the number of observations. We
assume that the observed group size (\(s_i\)) is recorded without error.

Model uncertainty can be addressed using standard maximum likelihood
theory. We can invert the Hessian matrix of the likelihood to obtain a
variance-covariance matrix. We can then pre- and post-multiply this by
the derivatives of \(\hat{N}_\text{C}\) \[
\widehat{\text{Var}}_\text{model}\left( \hat{N}_\text{C}\right) = \left(\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}\right)^\text{T} \left(\hat{\mathbf{H}}(\hat{\boldsymbol{\theta}})^{-1} \right)\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}
\] where the partial derivatives of \(\hat{N}_\text{C}\) are evaluated
at the MLE (\(\hat{\theta}\)) and \(\mathbf{H}\) is the first partial
Hessian (outer product of first derivatives of the log likelihood) for
numerical stability (Buckland et al. 2001, p 62). Note that although we
calculate uncertainty in \(\hat{N}_\text{C}\), we can trivially scale-up
to variance of \(\hat{N}\).

Sampling uncertainty can be characterised (rather than as simply \(n\))
by the \emph{encounter rate}: the number of objects per unit transect.
When covariates are not included in the detection function we can define
the encounter rate as \(n/L\) for line transects (where \(L\) is the
total line length) or \(n/T\) for point transects (where \(T\) is the
total number of visits to all points). When covariates are included in
the detection function, it is recommended that we substitute the \(n\)
in the previous expressions with the estimated abundance \(\hat{N}\) as
this will take into account the effects of the covariates.

For line transects, by default, \pkg{Distance} uses a variation of the
estimator ``R2'' from Fewster et al. (2009) replacing number of
observations per sample with the estimated abundance per sample (Innes
et al. 2002; F. F. C. Marques and Buckland 2003): \[
\widehat{\text{Var}}_{\text{encounter},R2} = \frac{K}{L^2(K-1)} \sum_{k=1}^{K} l_k^2 \left( \frac{N_k}{l_k} - \frac{N}{L}\right)^2,
\] where \(l_k\) are the lengths of the \(K\) transects (such that
\(L = \sum_{k=1}^K l_k\)). Whereas for points we use estimator ``P3''
from Fewster et al. (2009) but again replacing \(n\) by \(\hat{N}\) to
obtain the following estimator: \[
\widehat{\text{Var}}_{\text{encounter},P3} = \frac{1}{T(K-1)} \sum_{k=1}^{K} t_k \left( \frac{N_k}{t_k} - \frac{N}{T}\right)^2,
\] where \(t_k\) is the number of visits to point \(k\) and
\(T = \sum_{k=1}^K t_k\).

Various formulations for the encounter rate variance are discussed in
detail in Fewster et al. (2009). \pkg{Distance} implements all of the
estimators of encounter rate variance given in that article. The
\code{varn} manual page gives further advice and technical detail on
encounter rate variance.

We combine these two sources of variance by noting that squared
coefficients of variation (approximately) add (Goodman 1960).

\subsection{Estimating abundance and variance in
R}\label{estimating-abundance-and-variance-in-r}

Going back to the minke whale data, we have the information we require
to calculate \(A\) and \(a\) above, so we can estimate abundance and its
variance. When we supply data to \code{ds} in the ``flatfile'' format
given above, \code{ds} will automatically calculate abundance estimates
based on the survey setup in the data.

Having already fitted a model to the minke whale data, we can see the
results of the abundance estimation by simply looking at a model
summary:

\begin{CodeChunk}
\begin{CodeInput}
summary(minke_hn)
\end{CodeInput}
\begin{CodeOutput}

Summary for distance analysis 
Number of observations :  88 
Distance range         :  0  -  1.5

Model : Half-normal key function 
AIC   : 46.87216

Detection function parameters
Scale coefficient(s):  
              estimate        se
(Intercept) -0.3411766 0.1070304

                       Estimate          SE         CV
Average p             0.5733038  0.04980421 0.08687229
N in covered region 153.4962706 17.08959835 0.11133559

Summary statistics:
  Region   Area CoveredArea  Effort  n  k         ER      se.ER     cv.ER
1  North 630582     4075.14 1358.38 49 12 0.03607238 0.01317937 0.3653591
2  South  84734     1453.23  484.41 39 13 0.08051031 0.01809954 0.2248102
3  Total 715316     5528.37 1842.79 88 25 0.04775368 0.01129627 0.2365529

Abundance:
  Label Estimate        se        cv      lcl      ucl       df
1 North 13225.44 4966.7495 0.3755450 6005.590 29124.93 12.27398
2 South  3966.46  955.9616 0.2410113 2395.606  6567.36 15.80275
3 Total 17191.90 5135.5862 0.2987212 9183.475 32184.07 14.00459

Density:
  Label   Estimate          se        cv         lcl        ucl       df
1 North 0.02097339 0.007876453 0.3755450 0.009523884 0.04618738 12.27398
2 South 0.04681073 0.011281913 0.2410113 0.028272077 0.07750560 15.80275
3 Total 0.02403400 0.007179465 0.2987212 0.012838347 0.04499280 14.00459
\end{CodeOutput}
\end{CodeChunk}

This prints a rather large amount of information. One useful feature is
that the returned object is a \code{list} of \code{data.frames}, so we
can again use \pkg{knitr}s \code{kable} function to create useful tables
for publication. For example Table \ref{minke-abund}.

\begin{CodeChunk}
\begin{CodeInput}
minke_table <- summary(minke_hn)$dht$individuals$N
minke_table$lcl <- minke_table$ucl <- minke_table$df <- NULL
colnames(minke_table) <- c("Stratum", "$\\hat{N}$", "$\\text{se}(\\hat{N}$)", 
                           "$\\text{CV}(\\hat{N}$)")
kable(minke_table, digits=3, format="latex", booktabs=TRUE,
      row.names=FALSE, escape=FALSE,caption="Summary of abundance estimation for the half-normal model for the minke whale data.\\label{minke-abund}")
\end{CodeInput}
\begin{table}

\caption{Summary of abundance estimation for the half-normal model for the minke whale data.\label{minke-abund}}
\begin{tabular}{lrrr}
\toprule
Stratum & $\hat{N}$ & $\text{se}(\hat{N}$) & $\text{CV}(\hat{N}$)\\
\midrule
North & 13225.44 & 4966.750 & 0.376\\
South & 3966.46 & 955.962 & 0.241\\
Total & 17191.90 & 5135.586 & 0.299\\
\bottomrule
\end{tabular}
\end{table}

\end{CodeChunk}

\section{Extensions}\label{extensions}

The features of \pkg{Distance} are delibarately limited to provide a
simplified interface for users. However, for more complex distance
sampling-based analyses there are further related packages for modelling
in \proglang{R}.

We noted at the start of the article that \pkg{Distance} is a simpler
wrapper around the package \pkg{mrds}. Additional features are available
in \pkg{mrds} including the ability to model data where the assumption
that detection is certain at zero distance from the line or point by
using mark-recapture type methods when two observers are used (see Burt
et al. 2014 for an introduction).

The abundance estimates calculated here are based on the assumption that
within a given stratum abundance is constant. We may extend this
approach to many strata, making the area of each very small to account
for small-scale variation in space. A more rigourous approach is to
build a spatial model incorporating spatially-referenced environmental
data (for example derived from GIS products). \pkg{Distance} interfaces
with once such package to perform this type of analysis: \pkg{dsm}.
So-called ``density surface modelling'' uses the generalized additive
model framework (e.g. Wood 2006) to build models of abundance (that take
into account detectability) as a function of environmental covariates,
as part of a two stage model (Hedley and Buckland 2004; Miller et al.
2013).

\section{Conclusion}\label{conclusion}

Here we have given an introduction as to how to perform a distance
sampling analysis in \proglang{R}. We have covered the possible models
for detectability, model checking and selection and finally abundance
and variance estimation.

In combination with tools such as \pkg{knitr} and \pkg{rmarkdown}, the
helper functions in \pkg{Distance} provide a useful set of tools to
perform reproducible analyses of wildlife abundance for both managers
and ecologists. We hope that this paper (itself written in RMarkdown)
will provide a useful set of examples for those wishing to persue this.

We note that there are other packages available for performing distance
sampling analyses in \proglang{R} but believe that \pkg{Distance} is the
most flexible and feature-complete. \textbf{Appendix ???} gives a
feature comparison between \pkg{Distance} and the other packages
available.

\section{Acknowledgements}\label{acknowledgements}

The authors would like to thank the many users of \pkg{Distance},
\pkg{mrds} and DISTANCE who have contributed bug reports and suggestions
for improvements over the years.

\section*{Bibliography}\label{bibliography}
\addcontentsline{toc}{section}{Bibliography}

Becker, Earl F, and Aaron M Christ. 2015. ``A Unimodal Model for Double
Observer Distance Sampling Surveys.'' \emph{PLoS ONE} 10 (8):
e0136403--18.

Becker, Earl F, and P X Quang. 2009. ``A gamma-shaped detection function
for line-transect surveys with mark-recapture and covariate data.''
\emph{Journal of Agricultural, Biological, and Environmental Statistics}
14 (2): 207--23.

Borchers, David L, and Kenneth P Burnham. 2004. ``General formulation
for distance sampling.'' In \emph{Advanced Distance Sampling}, 6--30.
Oxford University Press, Oxford, UK.

Branch, T A, and D S Butterworth. 2001. ``Southern Hemisphere minke
whales: standardised abundance estimates from the 1978/79 to 1997/98
IDCR-SOWER surveys.'' \emph{Journal of Cetacean Research and
Management}.

Buckland, Stephen T. 1992. ``Fitting Density Functions with
Polynomials.'' \emph{Applied Statistics} 41 (1): 63.

Buckland, Stephen T, David R Anderson, Kenneth P Burnham, David L
Borchers, and Len Thomas. 2001. \emph{Introduction to Distance
Sampling}. Estimating Abundance of Biological Populations. Oxford
University Press, Oxford, UK.

Buckland, Stephen T, David R Anderson, Kenneth P Burnham, Jeffrey L
Laake, David L Borchers, and Len Thomas. 2004. \emph{Advanced Distance
Sampling}. Estimating Abundance of Biological Populations. Oxford
University Press, Oxford, UK.

Buckland, Stephen T, E A Rexstad, Tiago A Marques, and Cornelia S
Oedekoven. 2015. \emph{Distance Sampling: Methods and Applications}.
Methods in Statistical Ecology. Springer International Publishing.

Burnham, Kenneth P, Stephen T Buckland, Jeffrey L Laake, David L
Borchers, Jon R B Bishop, and Len Thomas. 2004. ``Further topics in
distance sampling.'' In \emph{Advanced Distance Sampling}, 385--89.
Oxford University Press, Oxford, UK.

Burt, Mary Louise, David L Borchers, Kurt J Jenkins, and Tiago A
Marques. 2014. ``Using mark-recapture distance sampling methods on line
transect surveys.'' \emph{Methods in Ecology and Evolution} 5 (11):
1180--91.

Eidous, Omar M. 2005. ``On Improving Kernel Estimators Using Line
Transect Sampling.'' \emph{Communications in Statistics - Theory and
Methods} 34 (4): 931--41.

Fewster, Rachel M, Stephen T Buckland, Kenneth P Burnham, David L
Borchers, Peter E Jupp, Jeffrey L Laake, and Len Thomas. 2009.
``Estimating the Encounter Rate Variance in Distance Sampling.''
\emph{Biometrics} 65 (1): 225--36.

Giammarino, Mauro, and Piero Quatto. 2014. ``On estimating Hooded crow
density from line transect data through exponential mixture models.''
\emph{Environmental and Ecological Statistics} 21 (4): 689--96.

Goodman, Leo A. 1960. ``On the Exact Variance of Products.''
\emph{Journal of the American Statistical Association} 55 (292): 708.

Hedley, Sharon L, and Stephen T Buckland. 2004. ``Spatial models for
line transect sampling.'' \emph{Journal of Agricultural, Biological, and
Environmental Statistics} 9 (2): 181--99.

Innes, Stuart, M P Heide-J{ø}rgensen, Jeffrey L Laake, Kristin L Laidre,
Holly J Cleator, Pierre Richard, and Robert EA Stewart. 2002. ``Surveys
of belugas and narwhals in the Canadian High Arctic in 1996.''
\emph{NAMMCO Scientific Publications} 4 (0): 169--90.

Lahoz-Monfort, Jos{é} J, Gurutzeta Guillera-Arroita, and Brendan A
Wintle. 2013. ``Imperfect detection impacts the performance of species
distribution models.'' \emph{Global Ecology and Biogeography} 23 (4):
504--15.

Marques, Fernanda F C, and Stephen T Buckland. 2003. ``Incorporating
covariates into standard line transect analyses.'' \emph{Biometrics} 59
(4): 924--35.

Marques, Tiago A, Len Thomas, Steven G Fancy, and Stephen T Buckland.
2007. ``Improving estimates of bird density using multiple-covariate
distance sampling.'' \emph{The Auk} 124 (4): 1229.

Miller, David L, and Len Thomas. 2015. ``Mixture models for distance
sampling detection functions.'' \emph{PLoS ONE}.

Miller, David L, Mary Louise Burt, Eric A Rexstad, and Len Thomas. 2013.
``Spatial models for distance sampling data: recent developments and
future directions.'' \emph{Methods in Ecology and Evolution} 4 (11):
1001--10.

Thomas, Len, Stephen T Buckland, Eric A Rexstad, Jeffrey L Laake,
Samantha Strindberg, Sharon L Hedley, Jon R B Bishop, Tiago A Marques,
and Kenneth P Burnham. 2010. ``Distance software: design and analysis of
distance sampling surveys for estimating population size.''
\emph{Journal of Applied Ecology} 47 (1): 5--14.

Thompson, S K. 2002. \emph{Sampling}. 2nd ed. Wiley.

Wood, Simon N. 2006. \emph{Generalized Additive Models}. An Introduction
with R. CRC Press.

\end{document}

### 2015-11-24 paper.Rmd
Eric comments addressed up to "Detection functions"

@@
 
-_before getting too deep into detection functions, would be nice to describe the 2-stage nature; adjust counts by detectability to get $\hat{N_c}$ then scale up $\hat{N_c}$ to get $\hat{N}$_
-
 Census-type surveys (quadrat or strip transects assuming perfect detectability within strips or quadrats) are inefficient (requiring considerable field effort) and we should expect that not all objects (animals, plants, dung, etc) can be observed.
@@
 Because we expect detectability to decrease with increasing distance from the sampler, we model detectability as a function of distance (plus perhaps other covariates, see below).
+We use the probabilities of detection calculated from the detection function to estimate an abundance for the area covered by the survey, which we then scale-up to the study region of interest.
 
@@
 
-For both points and lines, once the geometry of the sampler _likely spell out geom of sampler, or note it specification is forthcoming_ has been taken into account, the histogram of distances should show a decreasing number of observations with increasing distance from the sampler.
+For both points and lines, once the geometry of the sampler (see "Detection functions") has been taken into account, the histogram of distances should show a decreasing number of observations with increasing distance from the sampler.
 For line transects we expect objects to be distributed unformly with respect to distance from the line and what makes our histogram decrease is the detectability.

### 2015-11-24 paper.tex
Eric comments addressed up to "Detection functions"

@@
 \usepackage[utf8]{inputenc}
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
@@
 
-\emph{before getting too deep into detection functions, would be nice to
-describe the 2-stage nature; adjust counts by detectability to get
-\(\hat{N_c}\) then scale up \(\hat{N_c}\) to get \(\hat{N}\)}
-
 Census-type surveys (quadrat or strip transects assuming perfect
@@
 from the sampler, we model detectability as a function of distance (plus
-perhaps other covariates, see below).
+perhaps other covariates, see below). We use the probabilities of
+detection calculated from the detection function to estimate an
+abundance for the area covered by the survey, which we then scale-up to
+the study region of interest.
 
@@
 
-For both points and lines, once the geometry of the sampler \emph{likely
-spell out geom of sampler, or note it specification is forthcoming} has
-been taken into account, the histogram of distances should show a
-decreasing number of observations with increasing distance from the
-sampler. For line transects we expect objects to be distributed unformly
-with respect to distance from the line and what makes our histogram
-decrease is the detectability. For point transects, we recognise that as
-distance from the point increases, the area of the circle encompassed
-increases with distance squared; hence the number of objects available
-to be detected is a linearly increasing function of distance from the
-point.
+For both points and lines, once the geometry of the sampler (see
+``Detection functions'') has been taken into account, the histogram of
+distances should show a decreasing number of observations with
+increasing distance from the sampler. For line transects we expect
+objects to be distributed unformly with respect to distance from the
+line and what makes our histogram decrease is the detectability. For
+point transects, we recognise that as distance from the point increases,
+the area of the circle encompassed increases with distance squared;
+hence the number of objects available to be detected is a linearly
+increasing function of distance from the point.
 

### 2015-11-24 paper.Rmd
add Eric

@@
     url: http://converged.yt
+  - name: Eric A Rexstad
+    affiliation: University of St Andrews
+    address: >
+      Centre for Research into Ecological and Environmental Modelling,
+      The Observatory,
+      St Andrews, Fife KY16 9LZ, Scotland
+    email: Eric.Rexstad@st-andrews.ac.uk
   - name: Jeffrey L Laake

### 2015-11-28 paper.Rmd
ER finishes first run through of edits
some questions included for DLM included in italics in .rmd

--- 2015-11-24 paper.Rmd
@@
     url: http://converged.yt
+  - name: Eric Rexstad
+    affiliation: University of St Andrews
+    address: >
+      Centre for Research into Ecological and Environmental Modelling,
+      The Observatory,
+      St Andrews, Fife KY16 9LZ, Scotland
+    email: Eric.Rexstad@st-andrews.ac.uk
   - name: Jeffrey L Laake
@@
 
-Distance sampling [@Buckland:2001vm; @Buckland:2004ts] encompases a suite of field methods and statistical models used to estimate the abundance of biological populations.
+Distance sampling [@Buckland:2001vm; @Buckland:2004ts] encompasses a suite of field methods and statistical models used to estimate the abundance of biological populations.
 Distance sampling field procedure can be thought of as an extension of plot sampling, where we wish to take into account the decreasing probability of detecting objects at increasing distance from the sampler.
@@
 
-For both points and lines, once the geometry of the sampler (see "Detection functions") has been taken into account, the histogram of distances should show a decreasing number of observations with increasing distance from the sampler.
-For line transects we expect objects to be distributed unformly with respect to distance from the line and what makes our histogram decrease is the detectability.
+For both points and lines, given the geometry of the sampler (see "Detection functions"), the histogram of distances should show a decreasing number of observations with increasing distance from the sampler.
+For line transects we expect objects to be distributed uniformly with respect to distance from the line with detectability causing the number of detections to decrease.
 For point transects, we recognise that as distance from the point increases, the area of the circle encompassed increases with distance squared; hence the number of objects available to be detected is a linearly increasing function of distance from the point.
@@
 \pkg{Distance} estimates the parameters for a fixed-form detection function using maximum likelihood estimation.
-We address possible models in detail below.
+We address possible models below.
 
-Figure \ref{fig:pointslines} shows examples of 500 individuals sampled using line and point transects (left column) and their corresponding histograms (right column).
+Figure \ref{fig:pointslines} shows sampling of a population of 500 individuals using line and point transects (left column) and their corresponding histograms (right column).
 
@@
 
-We demonstrate \pkg{Distance} using two data sets: one line transect and one point transect. These data sets have been chosen to be representative of the kind of data seen in practice.
+We demonstrate \pkg{Distance} using two data sets: one line transect and one point transect. These data sets have been chosen to be representative of data seen in practice.
 
@@
 
-The line transect data is simulated data based on a survey of Antarctic minke whales (*Balaenoptera bonaerensis*).
-The data is simulated from models fitted to data from the International Whaling Commission's International Decade of Cetacean Research Southern Ocean Whale and Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer surveys [@Branch:2001ua].
-They consist of 99 observations and include information on the effort expended and whether observations were in one of two geographical strata (near or far from land).
+The line transect data is based on a survey of Antarctic minke whales (*Balaenoptera bonaerensis*).
+The data are simulated from models fitted to data from the International Whaling Commission's International Decade of Cetacean Research Southern Ocean Whale and Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer surveys [@Branch:2001ua].
+They consist of 99 observations and include geographical strata (near or distant from ice edge) and transect lengths on the effort expended.
 
@@
 The point transect data set consists of 1485 observations of Amakihi (*Hemignathus virens*; a Hawaiian songbird), collected at 41 points between 1992 and 1995.
-The data include distances and three covariates collected during the survey: observer ID (a three level factor), minutes after sunrise (continuous) and hours after sunrise (a six level factor).
+The data include distances and two covariates collected during the survey: observer (a three level factor), time after sunrise (transformed to minutes (continuous) or hours (factor) covariates).
 Data are analysed comprehensively in @Marques:2007ey.
@@
 
-* In the simplest case, where one would simply like to estimate a detection function, all that is needed is a vector of distances.
-* To include additional covariates into the detection function (see "Detection functions") we need to use a \code{data.frame}.
+* In the simplest case, where the objective is to estimate a detection function, all that is needed is a vector of distances.
+* To include additional covariates into the detection function (see "Detection functions") we use a \code{data.frame}.
 The data.frame contains a column called \code{distance} and additional named columns for covariates potentially useful in modelling detectability (for example \code{observer} or \code{seastate}).
 Some column names are reserved: \code{object} for an observation identifier (see "Extensions"), \code{size} for group or cluster size (see "Detection functions" and "Abundance and variance estimation"), \code{detected} for whether an observation was detected (see "Extensions") and the columns described in the next bullet.
-* If one would like to estimate abundance beyond area covered by expended survey effort, additional information is required.
-This consists of information about which transect and occasion the observation was made (the \code{Sample.Label}), a column named \code{Effort} which gives the effort associated with that sample (for lines their length and for points the number of times that point was visited), the stratum the sample was located in (this may have any name and may be from pre- or post-survey stratification, see "Estimating abundance and variance") and that stratum's area (which has the same name as the stratum column, appended with \code{.Area}).
-(We refer to this data format as "flatfile" where all information is contained in one table.)
+* To estimate abundance beyond area covered by expended survey effort, additional information is required.
+This consists of information about which transect and occasion the observation was made (the \code{Sample.Label}), a column named \code{Effort} which gives the effort associated with that sample (for lines their length and for points the number of times that point was visited), the stratum the sample was located in (this may have any name and may be from pre- or post-survey stratification, see "Estimating abundance and variance") and strata areas (which has the same name as the stratum column, appended with \code{.Area}).
+We refer to this data format as "flatfile" where all information is contained in one table.
 
-As we will see in "Extensions", further information is also required when we start using more complex models.
+As we will see in "Extensions", further information is also required when using more complex models.
 
@@
 
-We'll explore what the effects of including effort data are during analysis below.
+We will explore below the effects of including effort data during analysis.
 
@@
 
-As mentioned above, the detection function describes the relationship between observed distances and probability of detection.
-The detection function itself models the probability $\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)$ and is usually denoted $g(y; \boldsymbol{\theta})$ where $y$ is distance (either from a line or point) and $\boldsymbol{\theta}$ is a vector of parmeters to be estimated.
-Our goal is to estimate an *average probability of detection* ($p$, average in the sense of an average over the distances), so we must integrate out distance ($y$) from the detection function:
+The detection function describes the relationship between observed distances and probability of detection.
+The detection function itself models the probability $\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)$ and is usually denoted $g(y; \boldsymbol{\theta})$ where $y$ is distance (either from a line or point) and $\boldsymbol{\theta}$ is a vector of parameters to be estimated.
+Our goal is to estimate an *average probability of detection* ($p$, average in the sense of an average over distances), so we must integrate out distance ($y$) from the detection function:
 $$
@@
 
-It is important that the detection function accurately models detectability near zero distance.
+It is crucial that the detection function accurately models detectability at small distances.
 We are less worried by its behaviour further away from 0.
-To ensure that the model is not overly influence by distances far from zero we truncate the distances beyond a given distance $w$ (known as the *truncation distance*).
+To ensure that the model is not overly influence by distances far from zero, we truncate the distances beyond a given distance $w$ (known as the *truncation distance*).
 Fitting to distances at great distances from zero does not demonstrably improve the precision of abundance estimates [@Buckland:2001vm, 103-108, 151-153].
@@
 
-* *Shoulder*: we expect observers to be able to see objects near them, not just those directly infront of them. For this reason, we expect the detection function to be "flat" near zero distance.
-* *Non-increasing*: we don't think that observers should be more likely to see things further away than those nearer to them.
-This usually indicates an issue with field procedure (that the distribution of objects with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this.
+* *Shoulder*: we expect observers to be able to see objects near them, not just those directly adjacent to them. For this reason, we expect the detection function to be horizontal near zero distance.
+* *Non-increasing*: we do not think that observers should be more likely to see things distant from the transect than those near the transect.
+If this occurs, it usually indicates an issue with field procedure (that the distribution of objects with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this.
 * *Model robust*: models should be flexible enough to make many different shapes.
-* *Pooling robust*: many factors can affect the probability of detection and it is not possible to measure all of these. We would like our models to be robust to us not including these factors.
-* *Estimator efficiency*: we would like our models to have low variances, but only given they satisfy the other properties above (which, if they are satisfied, would give low bias).
+* *Pooling robust*: many factors can affect the probability of detection and it is not possible to measure all of these.
+We would like models to produce unbiased results without inclusion of these factors.
+* *Estimator efficiency*: we would like models to have low variances, but only given they satisfy the other properties above (which, if they are satisfied, would give low bias).
 
-Given these criteria, we can start to think about models for $g$.
+Given these criteria, we can formulate models for $g$.
 
@@
 \code{Distance} includes the most popular of these models.
-Here we'll detail the most popular detection function approach: "key function plus adjustments" (K+A).
+Here we will detail the most popular detection function approach: "key function plus adjustments" (K+A).
 
@@
 
-A disadvantage of K+A models is that we must resort to constrained optimisation (via the \pkg{Rsolnp} package) in order to ensure that the resulting detection function is monotonic non-increasing over the whole range.
+A disadvantage of K+A models is that we must resort to constrained optimisation (via the \pkg{Rsolnp} package) to ensure that the resulting detection function is monotonic non-increasing over the whole range.
 
@@
 f <- function(x) g(x)/g(0)
-curve(f, from=0, to=0.025, xlab="", ylab="", main="Half-normal with 1 cosine adjustment",xlim=c(0,0.025), ylim=c(0,1))
+curve(f, from=0, to=0.025, xlab="", ylab="", main="Half-normal with \n1 cosine adjustment",xlim=c(0,0.025), ylim=c(0,1))
 title(xlab="Distance", line=2)
@@
 f <- function(x) g(x)/g(0)
-curve(f, from=0, to=0.025, xlab="", ylab="", main="Half-normal with 2 cosine adjustments",xlim=c(0,0.025), ylim=c(0,1))
+curve(f, from=0, to=0.025, xlab="", ylab="", main="Half-normal with \n2 cosine adjustments",xlim=c(0,0.025), ylim=c(0,1))
 title(xlab="Distance", line=2)
@@
 f <- function(x) g(x)/g(0)
-curve(f, from=0, to=0.025, xlab="", ylab="", main="Hazard-rate with 1 cosine adjustment",xlim=c(0,0.025), ylim=c(0,1))
+curve(f, from=0, to=0.025, xlab="", ylab="", main="Hazard-rate with \n1 cosine adjustment",xlim=c(0,0.025), ylim=c(0,1))
 title(xlab="Distance", line=2)
@@
 f <- function(x) g(x)/g(0)
-curve(f, from=0, to=0.025, xlab="", ylab="", main="Hazard-rate with 2 cosine adjustments",xlim=c(0,0.025), ylim=c(0,1))
+curve(f, from=0, to=0.025, xlab="", ylab="", main="Hazard-rate with \n2 cosine adjustments",xlim=c(0,0.025), ylim=c(0,1))
 title(xlab="Distance", line=2)
@@
 There are many factors that can affect the probability of detecting an object.
-These include things like the observer, the vessel or platform used, the sea state, weather conditions and time of day (to name but a few).
+These include things like the observer, the vessel or platform used, the sea state, weather conditions and time of day.
 We assume that these variables affect detection only via the scale of the detection function (and do not affect the shape).
@@
 $$
-In the next section we'll discuss model selection.
+In the next section we will discuss model selection.
 
 Including covariates has an important implication for our calculation of detectability.
-Since we don't know what the true distribution of the covariates is we must calculate the probability of detection conditional on the observed values of the covariates:
+Because we do not know what the true distribution of the covariates is we must calculate the probability of detection conditional on the observed values of the covariates:
 $$
@@
 where $\mathbf{z_i}$ is the vector of $J$ covariates associated with observation $i$.
-So for covariate models, we are calculating a value of "average" probability of detection (again the average is in the sense of distance) per observation.
+For covariate models, we are calculating a value of "average" probability of detection (average in the sense of distance) per observation.
 There will be as many unique values of $p(\mathbf{z_i})$ as there are unique covariate combinations in our data.
 
-Another important consideration is that K+A models which include covariates and one or more adjustments cannot be guaranteed to be monotonic non-increasing for all covariate combinations, as we don't have any model for the distribution of the covariates.
+K+A models that include covariates and one or more adjustments cannot be guaranteed to be monotonic non-increasing for all covariate combinations, as we do not have any model for the distribution of the covariates.
 For this reason, we advise against using both adjustments and covariates in a detection function [see @Miller:2015hw for an example of when this can be problematic].
@@
 The workhorse of detection function fitting in \pkg{Distance} is the \code{ds} function.
-Here we show off the formulations for the detection function that we've seen above for both the minke whale and amakihi data.
+Here we show off the formulations for the detection function we have seen above applied to the minke whale and amakihi data.
 
@@
 ```
-Note that \code{ds} will automatically select adjustment terms by AIC and shows it's selection steps as it goes.
+Note that \code{ds} will automatically select adjustment terms by AIC and shows the selection steps.
 
 Figure \ref{fig:minkeamakihi} (top left plot) shows the result of calling \code{plot} on the resulting model object.
-We can also call \code{summary} on the model object to get summary information about the fitted model (though we postpone this until the next section).
+We can also call \code{summary} on the model object to get summary information about the fitted model (we postpone this to the next section).
 
@@
 ```
-Again, \code{ds} fits a model with cosine adjustments (the default) but finds the AIC improvement to be insufficient to select the adjustment.
+\code{ds} fits a model with cosine adjustments (the default) but finds the AIC improvement to be insufficient to select the adjustment.
 
@@
 
-By default \code{ds} assumes that the data given to it is line transect, but we can switch to points using \code{transect="point"}.
-Including covariates in the scale is via the \code{formula=~...} argument to \code{ds}; a hazard-rate model for the amakihi that includes observer as a covariate can be specified by [truncating at 82.5m, given in @Marques:2007ey]:
+By default \code{ds} assumes the data given to it is line transect, but we can switch to points using \code{transect="point"}.
+Including covariates in the scale is via the \code{formula=~...} argument to \code{ds}; a hazard-rate model for the amakihi that includes observer as a covariate can be specified by [truncating sightings at 82.5m, given in @Marques:2007ey]:
 ```{r amakihi-hr-obs}
@@
 As with the minke whale model, we can plot the resulting detection function (Figure \ref{fig:minkeamakihi}, centre column).
-Since for the amakihi we used covariates in the detection function, the plot will show the detection function averaged over levels/values of the covariate.
+Because the amakihi model included covariates, the plot shows the detection function averaged over levels/values of the covariate.
 Points on the plot indicate probability of detection for each observation.
@@
 The bottom row of Figure \ref{fig:minkeamakihi} shows the probability density functions for the amakihi models.
-These often are easier to interpret than the detection functions, as for point data we must rescale the histogram when plotting the detection function to take into account the geometry of the point sampler.
+Probability density function plots give a better sense of fit of the model than the detection functions, as for point data we must rescale the histogram when plotting the detection function to take into account the geometry of the point sampler.
+
+** ER suggests the g(x) plots for the amakihi be deleted **
 
@@
 
-As with models fitted using \code{lm} or \code{glm} in \proglang{R}, we can use \code{summary} to give us some useful information about our fitted model.
+As with models fitted using \code{lm} or \code{glm} in \proglang{R}, we can use \code{summary} to give useful information about our fitted model.
 For example for our hazard-rate model for the amakihi, with observer as a covariate:
@@
 ```
-This summary information includes various details of the data and model specification, as well as the values of the coefficents ($\beta_j$) and their uncertainties, an "average" value for the detectability (see "Estimating abundance and variance" for details on how this is calculated) and it's uncertainty.
-The final line gives an estimate of abundance for the area covered by the survey (again, this will be covered in detail in the next section).
+This summary information includes details of the data and model specification, as well as the values of the coefficients ($\beta_j$) and their uncertainties, an "average" value for the detectability (see "Estimating abundance and variance" for details on how this is calculated) and its uncertainty.
+The final line gives an estimate of abundance for the area covered by the survey (covered in the next section).
 
@@
 The EDF gives the proportion of observations for which the CDF is less than or equal to that of a given distance.
-In essence we're asking "is the number of observations up to a given distance in line with what the model says they should be?" (where 'given values' are the observed distances).
+This can be interpreted as assessing whether the number of observations up to a given distance in line with what the model says they should be (where 'given values' are the observed distances).
 As usual for Q-Q plots, "good" models will have values close to the line $y=x$, poor models will show more deviations from that line.
 
-We can inspect Q-Q plots visually, though for a large number of models this can be tiresome and prone to subjective judgements.
+We can inspect Q-Q plots visually, though this is prone to subjective judgments.
 Instead we can quantify the Q-Q plot's information using a Kolmogorov-Smirnov or Cramer-von Mises test [@Burnham:2004vd].
-Both test to see if the points from the EDF and CDF are from the same distribution.
+Both test whether points from the EDF and CDF are from the same distribution.
 The Kolmogorov-Smirnov uses the test statistic of the largest difference between a point on the Q-Q plot and the line $y=x$, whereas the Cramer-von Mises test uses the sum of all the differences.
@@
 We can generate a Q-Q plot and test results using the \code{gof_ddf} function.
-We can see the goodness of fit tests below for two models for the amakihi data, first fitting a half-normal model without covariates or adjustments (not setting \code{adjustment=NULL} will force \code{ds} to fit a model with no adjustments), then calculating goodness of fit for that model and our hazard-rate model with observer and minutes after sunrise included:
+We can see the goodness of fit tests below for two models for the amakihi data (Figure \ref{amakihi-qq}), first fitting a half-normal model without covariates or adjustments (not setting \code{adjustment=NULL} will force \code{ds} to fit a model with no adjustments), then calculating goodness of fit for that model and our hazard-rate model with observer and minutes after sunrise included:
 ```{r amakihi-gof, fig.keep="none"}
@@
 par(mfrow=c(1,2))
-gof_ds(amakihi_hr, asp=1)
-gof_ds(amakihi_hr_obs_mas, asp=1)
+gof_ds(amakihi_hr, asp=1, cex=0.3)  # shrinking dot size to obliterate 45-deg line less
+gof_ds(amakihi_hr_obs_mas, asp=1, cex=0.3)
 ```
@@
 
-Though fiting the detection function is the primary modelling step in distance sampling, we are really interested in estimating detectability, and from that abundance.
+Though fitting the detection function is the primary modelling step in distance sampling, we are really interested in estimating detectability, and from that abundance.
 We also wish to calculate our certainty in each abundance estimate.
-This section addresses these issues mathematically before showing how to estimate abundance and its variance in R.
+This section addresses these issues mathematically before showing how to estimate abundance and its variance in \proglang{R}.
 
@@
 
-First, in order to estimate abundance in the covered area ($\hat{N}_\text{C}$), we use the estimates of detection probability (the $\{p_i; i=1,\ldots,n\}$, above) in a Horvitz-Thompson like estimator:
+First, to estimate abundance in the covered area ($\hat{N}_\text{C}$), we use the estimates of detection probability (the $\{p_i; i=1,\ldots,n\}$, above) in a Horvitz-Thompson like estimator:
 \begin{equation}
@@
 where $s_i$ are the sizes of the observed groups of objects, which is equal to 1 if objects only occur singly [@Borchers:2004wr].
-@thompson2012sampling is the cannonical reference to this type of estimator; intuitively, we can think of the estimates of detectability ($\hat{p}_i$) as "inflating" the group sizes ($s_i$) to account for incomplete detection -- we then sum to obtain the abundance estimate.
+@thompson2012sampling is the canonical reference to this type of estimator; intuitively, we can think of the estimates of detectability ($\hat{p}_i$) as "inflating" the group sizes ($s_i$) to account for incomplete detection -- we then sum over all $n$ detections to obtain the abundance estimate.
 For models that do not include covariates, $\hat{p}_i$ is equal for all $i$, so this is equivalent to summing the groups and inflating that sum by the corresponding $\hat{p} (=\hat{p}_i \forall i)$.
@@
 
+*have L and K been previously defined?*
+
 We can use the Horvitz-Thompson-like estimator to calculate the "average" detectability for models which include covariates.
@@
 This can be a useful summary statistic and is included in the \code{summary} output and the table produced by \code{summarize_ds_models}.
+*anything else to say about this, seems to mess up folks, including me*
 
@@
 We may wish to calculate abundance estimates for some sub-regions of the study region, we call these areas *strata*.
-A stratum may be defined by, for example, a different habitat type or by the sex of the animal (or some combination) which may be of interest for biological or management reasons.
-In order to calculate estimates for a given stratification each observation must occur in a stratum which must be labelled with a \code{Region.Label} and have a corresponding \code{Area} (if we are using an animal characteristic like sex, we would have the areas be the same but if we were using say forested vs.
+For example, strata may be defined by habitat types or animal gender (or some combination) which may be of interest for biological or management reasons.
+To calculate estimates for a given stratification each observation must occur in a stratum which must be labelled with a \code{Region.Label} and have a corresponding \code{Area} (if we are using an animal characteristic like gender, we would have the areas be the same but if we were using say forested vs.
 wetland habitat the areas of those strata would be different).
-Finally, we must also know which stratum a given transect lies in.
+Finally, we must also know the stratum in which transects lie.
 
@@
 Figure \ref{fig:minke-strata} shows the two strata, along with observations and transect lines.
+*figure number messed up*
 
@@
 We can invert the Hessian matrix of the likelihood to obtain a variance-covariance matrix.
-We can then pre- and post-multiply this by the derivatives of $\hat{N}_\text{C}$
+We can then pre- and post-multiply this by the derivatives of $\hat{N}_\text{C}$ with respect to the parameters of the detection function
 $$
@@
 Sampling uncertainty can be characterised (rather than as simply $n$) by the *encounter rate*: the number of objects per unit transect.
-When covariates are not included in the detection function we can define the encounter rate as $n/L$ for line transects (where $L$ is the total line length) or $n/T$ for point transects (where $T$ is the total number of visits to all points).
+When covariates are not included in the detection function we can define the encounter rate as $n/L$ for line transects (where $L$ is the total line length) or $n/T$ for point transects (where $T$ is the total number of visits to all points) *earlier eqn uses notation $K$ for number of points*.
 When covariates are included in the detection function, it is recommended that we substitute the $n$ in the previous expressions with the estimated abundance $\hat{N}$ as this will take into account the effects of the covariates.
@@
 $$
-where $t_k$ is the number of visits to point $k$ and $T = \sum_{k=1}^K t_k$.
+where $t_k$ is the number of visits to point $k$ and $T = \sum_{k=1}^K t_k$. *this messes with the $K$ notation even further*
 
@@
 
-We combine these two sources of variance by noting that squared coefficients of variation (approximately) add [@Goodman:1960dd].
+We combine detection function variance with encounter rate variance by use of the $\delta$-method approximation [@Goodman:1960dd].
 
@@
 
-Going back to the minke whale data, we have the information we require to calculate $A$ and $a$ above, so we can estimate abundance and its variance.
-When we supply data to \code{ds} in the ``flatfile'' format given above, \code{ds} will automatically calculate abundance estimates based on the survey setup in the data.
+Returning to the minke whale data, we have the necessary information to calculate $A$ and $a$ above, so we can estimate abundance and its variance.
+When we supply data to \code{ds} in the ``flatfile'' format given above, \code{ds} will automatically calculate abundance estimates based on the survey information in the data.
 
-Having already fitted a model to the minke whale data, we can see the results of the abundance estimation by simply looking at a model summary:
+Having already fitted a model to the minke whale data, we can see the results of the abundance estimation by viewing the model summary:
 ```{r minke-N-summary}
@@
 This prints a rather large amount of information.
-One useful feature is that the returned object is a \code{list} of \code{data.frames}, so we can again use \pkg{knitr}s \code{kable} function to create useful tables for publication.
-For example Table \ref{minke-abund}.
+The summary can be more concisely expressed by extracting information from the summary object.
+This object is a \code{list} of \code{data.frames}, so we can again use \pkg{knitr}s \code{kable} function to create summary tables of abundance estimates and measures of precision, such as Table \ref{minke-abund}.
 
@@
 
-The features of \pkg{Distance} are delibarately limited to provide a simplified interface for users.
-However, for more complex distance sampling-based analyses there are further related packages for modelling in \proglang{R}.
+The features of \pkg{Distance} are deliberately limited to provide a simplified interface for users.
+For more complex analyses of distance sampling data, there are further related packages for modelling in \proglang{R}.
 
@@
 
-The abundance estimates calculated here are based on the assumption that within a given stratum abundance is constant.
+The abundance estimates calculated here are based on the assumption that within a given stratum density is uniform.
 We may extend this approach to many strata, making the area of each very small to account for small-scale variation in space.
@@
 So-called "density surface modelling" uses the generalized additive model framework [e.g.
-@Wood:2006vg] to build models of abundance (that take into account detectability) as a function of environmental covariates, as part of a two stage model [@Hedley:2004et; @Miller:2013fq].
+@Wood:2006vg] to build models of abundance (adjusting counts for imperfect detectability) as a function of environmental covariates, as part of a two stage model [@Hedley:2004et; @Miller:2013fq].
 
@@
 
-Here we have given an introduction as to how to perform a distance sampling analysis in \proglang{R}.
+We have given an introduction as to how to perform a distance sampling analysis in \proglang{R}.
 We have covered the possible models for detectability, model checking and selection and finally abundance and variance estimation.
@@
 In combination with tools such as \pkg{knitr} and \pkg{rmarkdown}, the helper functions in \pkg{Distance} provide a useful set of tools to perform reproducible analyses of wildlife abundance for both managers and ecologists.
-We hope that this paper (itself written in RMarkdown) will provide a useful set of examples for those wishing to persue this.
+We hope that this paper provides useful examples for those wishing to pursue this.
 
 We note that there are other packages available for performing distance sampling analyses in \proglang{R} but believe that \pkg{Distance} is the most flexible and feature-complete.
-**Appendix ???** gives a feature comparison between \pkg{Distance} and the other packages available.
+**Appendix ???** gives a feature comparison between \pkg{Distance} and other \proglang{R} packages for analysis of distance sampling data.
 

### 2015-11-28 paper.tex
ER finishes first run through of edits
some questions included for DLM included in italics in .rmd

@@
 
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
 \author{
-David L Miller\\University of St Andrews \And Jeffrey L Laake\\National Marine Mammal Laboratory
+David L Miller\\University of St Andrews \And Eric Rexstad\\University of St Andrews \And Jeffrey L Laake\\National Marine Mammal Laboratory
 }
@@
 
-\Plainauthor{David L Miller, Jeffrey L Laake}
+\Plainauthor{David L Miller, Eric Rexstad, Jeffrey L Laake}
 \Plaintitle{Distance Sampling in R}
@@
   URL: \url{http://converged.yt}\\~\\
+      Eric Rexstad\\
+  University of St Andrews\\
+  Centre for Research into Ecological and Environmental Modelling, The
+  Observatory, St Andrews, Fife KY16 9LZ, Scotland\\
+  E-mail: \href{mailto:Eric.Rexstad@st-andrews.ac.uk}{\nolinkurl{Eric.Rexstad@st-andrews.ac.uk}}\\
+
       Jeffrey L Laake\\
@@
 Distance sampling (Buckland et al. 2001; Buckland et al. 2004)
-encompases a suite of field methods and statistical models used to
+encompasses a suite of field methods and statistical models used to
 estimate the abundance of biological populations. Distance sampling
@@
 
-For both points and lines, once the geometry of the sampler (see
-``Detection functions'') has been taken into account, the histogram of
-distances should show a decreasing number of observations with
-increasing distance from the sampler. For line transects we expect
-objects to be distributed unformly with respect to distance from the
-line and what makes our histogram decrease is the detectability. For
-point transects, we recognise that as distance from the point increases,
-the area of the circle encompassed increases with distance squared;
-hence the number of objects available to be detected is a linearly
-increasing function of distance from the point.
+For both points and lines, given the geometry of the sampler (see
+``Detection functions''), the histogram of distances should show a
+decreasing number of observations with increasing distance from the
+sampler. For line transects we expect objects to be distributed
+uniformly with respect to distance from the line with detectability
+causing the number of detections to decrease. For point transects, we
+recognise that as distance from the point increases, the area of the
+circle encompassed increases with distance squared; hence the number of
+objects available to be detected is a linearly increasing function of
+distance from the point.
 
@@
 estimates the parameters for a fixed-form detection function using
-maximum likelihood estimation. We address possible models in detail
-below.
-
-Figure \ref{fig:pointslines} shows examples of 500 individuals sampled
-using line and point transects (left column) and their corresponding
-histograms (right column).
+maximum likelihood estimation. We address possible models below.
+
+Figure \ref{fig:pointslines} shows sampling of a population of 500
+individuals using line and point transects (left column) and their
+corresponding histograms (right column).
 
@@
 one point transect. These data sets have been chosen to be
-representative of the kind of data seen in practice.
+representative of data seen in practice.
 
@@
 
-The line transect data is simulated data based on a survey of Antarctic
-minke whales (\emph{Balaenoptera bonaerensis}). The data is simulated
-from models fitted to data from the International Whaling Commission's
-International Decade of Cetacean Research Southern Ocean Whale and
-Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer
-surveys (Branch and Butterworth 2001). They consist of 99 observations
-and include information on the effort expended and whether observations
-were in one of two geographical strata (near or far from land).
+The line transect data is based on a survey of Antarctic minke whales
+(\emph{Balaenoptera bonaerensis}). The data are simulated from models
+fitted to data from the International Whaling Commission's International
+Decade of Cetacean Research Southern Ocean Whale and Ecosystem Research
+(IWC IDCR-SOWER) programme 1992-1993 austral summer surveys (Branch and
+Butterworth 2001). They consist of 99 observations and include
+geographical strata (near or distant from ice edge) and transect lengths
+on the effort expended.
 
@@
 (\emph{Hemignathus virens}; a Hawaiian songbird), collected at 41 points
-between 1992 and 1995. The data include distances and three covariates
-collected during the survey: observer ID (a three level factor), minutes
-after sunrise (continuous) and hours after sunrise (a six level factor).
-Data are analysed comprehensively in Marques et al. (2007).
+between 1992 and 1995. The data include distances and two covariates
+collected during the survey: observer (a three level factor), time after
+sunrise (transformed to minutes (continuous) or hours (factor)
+covariates). Data are analysed comprehensively in Marques et al. (2007).
 
@@
 \item
-  In the simplest case, where one would simply like to estimate a
-  detection function, all that is needed is a vector of distances.
+  In the simplest case, where the objective is to estimate a detection
+  function, all that is needed is a vector of distances.
 \item
   To include additional covariates into the detection function (see
-  ``Detection functions'') we need to use a \code{data.frame}. The
-  data.frame contains a column called \code{distance} and additional
-  named columns for covariates potentially useful in modelling
-  detectability (for example \code{observer} or \code{seastate}). Some
-  column names are reserved: \code{object} for an observation identifier
-  (see ``Extensions''), \code{size} for group or cluster size (see
+  ``Detection functions'') we use a \code{data.frame}. The data.frame
+  contains a column called \code{distance} and additional named columns
+  for covariates potentially useful in modelling detectability (for
+  example \code{observer} or \code{seastate}). Some column names are
+  reserved: \code{object} for an observation identifier (see
+  ``Extensions''), \code{size} for group or cluster size (see
   ``Detection functions'' and ``Abundance and variance estimation''),
@@
 \item
-  If one would like to estimate abundance beyond area covered by
-  expended survey effort, additional information is required. This
-  consists of information about which transect and occasion the
-  observation was made (the \code{Sample.Label}), a column named
-  \code{Effort} which gives the effort associated with that sample (for
-  lines their length and for points the number of times that point was
-  visited), the stratum the sample was located in (this may have any
-  name and may be from pre- or post-survey stratification, see
-  ``Estimating abundance and variance'') and that stratum's area (which
-  has the same name as the stratum column, appended with \code{.Area}).
-  (We refer to this data format as ``flatfile'' where all information is
-  contained in one table.)
+  To estimate abundance beyond area covered by expended survey effort,
+  additional information is required. This consists of information about
+  which transect and occasion the observation was made (the
+  \code{Sample.Label}), a column named \code{Effort} which gives the
+  effort associated with that sample (for lines their length and for
+  points the number of times that point was visited), the stratum the
+  sample was located in (this may have any name and may be from pre- or
+  post-survey stratification, see ``Estimating abundance and variance'')
+  and strata areas (which has the same name as the stratum column,
+  appended with \code{.Area}). We refer to this data format as
+  ``flatfile'' where all information is contained in one table.
 \end{itemize}
@@
 As we will see in ``Extensions'', further information is also required
-when we start using more complex models.
+when using more complex models.
 
@@
 
-We'll explore what the effects of including effort data are during
-analysis below.
+We will explore below the effects of including effort data during
+analysis.
 
@@
 
-As mentioned above, the detection function describes the relationship
-between observed distances and probability of detection. The detection
-function itself models the probability
+The detection function describes the relationship between observed
+distances and probability of detection. The detection function itself
+models the probability
 \(\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)\)
@@
 distance (either from a line or point) and \(\boldsymbol{\theta}\) is a
-vector of parmeters to be estimated. Our goal is to estimate an
+vector of parameters to be estimated. Our goal is to estimate an
 \emph{average probability of detection} (\(p\), average in the sense of
-an average over the distances), so we must integrate out distance
-(\(y\)) from the detection function: \[
+an average over distances), so we must integrate out distance (\(y\))
+from the detection function: \[
 p = \int_0^w \pi(y) g(y; \boldsymbol{\theta}) dy
@@
 
-It is important that the detection function accurately models
-detectability near zero distance. We are less worried by its behaviour
+It is crucial that the detection function accurately models
+detectability at small distances. We are less worried by its behaviour
 further away from 0. To ensure that the model is not overly influence by
-distances far from zero we truncate the distances beyond a given
+distances far from zero, we truncate the distances beyond a given
 distance \(w\) (known as the \emph{truncation distance}). Fitting to
@@
   \emph{Shoulder}: we expect observers to be able to see objects near
-  them, not just those directly infront of them. For this reason, we
-  expect the detection function to be ``flat'' near zero distance.
+  them, not just those directly adjacent to them. For this reason, we
+  expect the detection function to be horizontal near zero distance.
 \item
-  \emph{Non-increasing}: we don't think that observers should be more
-  likely to see things further away than those nearer to them. This
-  usually indicates an issue with field procedure (that the distribution
-  of objects with respect to the line, \(\pi(y)\) is not what we
-  expect), so we do not want the detection function to model this.
+  \emph{Non-increasing}: we do not think that observers should be more
+  likely to see things distant from the transect than those near the
+  transect. If this occurs, it usually indicates an issue with field
+  procedure (that the distribution of objects with respect to the line,
+  \(\pi(y)\) is not what we expect), so we do not want the detection
+  function to model this.
 \item
@@
   detection and it is not possible to measure all of these. We would
-  like our models to be robust to us not including these factors.
+  like models to produce unbiased results without inclusion of these
+  factors.
 \item
-  \emph{Estimator efficiency}: we would like our models to have low
+  \emph{Estimator efficiency}: we would like models to have low
   variances, but only given they satisfy the other properties above
@@
 
-Given these criteria, we can start to think about models for \(g\).
+Given these criteria, we can formulate models for \(g\).
 
@@
 and Quatto 2014; Miller and Thomas 2015; Becker and Christ 2015).
-\code{Distance} includes the most popular of these models. Here we'll
+\code{Distance} includes the most popular of these models. Here we will
 detail the most popular detection function approach: ``key function plus
@@
 A disadvantage of K+A models is that we must resort to constrained
-optimisation (via the \pkg{Rsolnp} package) in order to ensure that the
-resulting detection function is monotonic non-increasing over the whole
-range.
+optimisation (via the \pkg{Rsolnp} package) to ensure that the resulting
+detection function is monotonic non-increasing over the whole range.
 
@@
 object. These include things like the observer, the vessel or platform
-used, the sea state, weather conditions and time of day (to name but a
-few). We assume that these variables affect detection only via the scale
-of the detection function (and do not affect the shape).
+used, the sea state, weather conditions and time of day. We assume that
+these variables affect detection only via the scale of the detection
+function (and do not affect the shape).
 
@@
 \sigma(\mathbf{z}) = \exp(\beta_0 + \sum_{j=1}^J \beta_j z_j).
-\] In the next section we'll discuss model selection.
+\] In the next section we will discuss model selection.
 
 Including covariates has an important implication for our calculation of
-detectability. Since we don't know what the true distribution of the
+detectability. Because we do not know what the true distribution of the
 covariates is we must calculate the probability of detection conditional
@@
 \] where \(\mathbf{z_i}\) is the vector of \(J\) covariates associated
-with observation \(i\). So for covariate models, we are calculating a
-value of ``average'' probability of detection (again the average is in
-the sense of distance) per observation. There will be as many unique
-values of \(p(\mathbf{z_i})\) as there are unique covariate combinations
-in our data.
-
-Another important consideration is that K+A models which include
-covariates and one or more adjustments cannot be guaranteed to be
-monotonic non-increasing for all covariate combinations, as we don't
-have any model for the distribution of the covariates. For this reason,
-we advise against using both adjustments and covariates in a detection
-function (see Miller and Thomas 2015 for an example of when this can be
-problematic).
+with observation \(i\). For covariate models, we are calculating a value
+of ``average'' probability of detection (average in the sense of
+distance) per observation. There will be as many unique values of
+\(p(\mathbf{z_i})\) as there are unique covariate combinations in our
+data.
+
+K+A models that include covariates and one or more adjustments cannot be
+guaranteed to be monotonic non-increasing for all covariate
+combinations, as we do not have any model for the distribution of the
+covariates. For this reason, we advise against using both adjustments
+and covariates in a detection function (see Miller and Thomas 2015 for
+an example of when this can be problematic).
 
@@
 \code{ds} function. Here we show off the formulations for the detection
-function that we've seen above for both the minke whale and amakihi
-data.
+function we have seen above applied to the minke whale and amakihi data.
 
@@
 Note that \code{ds} will automatically select adjustment terms by AIC
-and shows it's selection steps as it goes.
+and shows the selection steps.
 
@@
 \code{summary} on the model object to get summary information about the
-fitted model (though we postpone this until the next section).
+fitted model (we postpone this to the next section).
 
@@
 
-Again, \code{ds} fits a model with cosine adjustments (the default) but
-finds the AIC improvement to be insufficient to select the adjustment.
+\code{ds} fits a model with cosine adjustments (the default) but finds
+the AIC improvement to be insufficient to select the adjustment.
 
@@
 
-By default \code{ds} assumes that the data given to it is line transect,
-but we can switch to points using \code{transect="point"}. Including
+By default \code{ds} assumes the data given to it is line transect, but
+we can switch to points using \code{transect="point"}. Including
 covariates in the scale is via the \code{formula=~...} argument to
 \code{ds}; a hazard-rate model for the amakihi that includes observer as
-a covariate can be specified by (truncating at 82.5m, given in Marques
-et al. 2007):
+a covariate can be specified by (truncating sightings at 82.5m, given in
+Marques et al. 2007):
 
@@
 As with the minke whale model, we can plot the resulting detection
-function (Figure \ref{fig:minkeamakihi}, centre column). Since for the
-amakihi we used covariates in the detection function, the plot will show
-the detection function averaged over levels/values of the covariate.
-Points on the plot indicate probability of detection for each
-observation. For the \texttt{amakihi\_hr\_obs} model we see fairly clear
-levels of the observer covariate in the points. Looking at the left
-panel of Figure \ref{fig:minkeamakihi}, we can see this is less clear
-when adding minutes after sunrise as a covariate into the model:
+function (Figure \ref{fig:minkeamakihi}, centre column). Because the
+amakihi model included covariates, the plot shows the detection function
+averaged over levels/values of the covariate. Points on the plot
+indicate probability of detection for each observation. For the
+\texttt{amakihi\_hr\_obs} model we see fairly clear levels of the
+observer covariate in the points. Looking at the left panel of Figure
+\ref{fig:minkeamakihi}, we can see this is less clear when adding
+minutes after sunrise as a covariate into the model:
 
@@
 The bottom row of Figure \ref{fig:minkeamakihi} shows the probability
-density functions for the amakihi models. These often are easier to
-interpret than the detection functions, as for point data we must
-rescale the histogram when plotting the detection function to take into
-account the geometry of the point sampler.
+density functions for the amakihi models. Probability density function
+plots give a better sense of fit of the model than the detection
+functions, as for point data we must rescale the histogram when plotting
+the detection function to take into account the geometry of the point
+sampler.
+
+** ER suggests the g(x) plots for the amakihi be deleted **
 
@@
 As with models fitted using \code{lm} or \code{glm} in \proglang{R}, we
-can use \code{summary} to give us some useful information about our
-fitted model. For example for our hazard-rate model for the amakihi,
-with observer as a covariate:
+can use \code{summary} to give useful information about our fitted
+model. For example for our hazard-rate model for the amakihi, with
+observer as a covariate:
 
@@
 
-This summary information includes various details of the data and model
-specification, as well as the values of the coefficents (\(\beta_j\))
+This summary information includes details of the data and model
+specification, as well as the values of the coefficients (\(\beta_j\))
 and their uncertainties, an ``average'' value for the detectability (see
 ``Estimating abundance and variance'' for details on how this is
-calculated) and it's uncertainty. The final line gives an estimate of
-abundance for the area covered by the survey (again, this will be
-covered in detail in the next section).
+calculated) and its uncertainty. The final line gives an estimate of
+abundance for the area covered by the survey (covered in the next
+section).
 
@@
 some given value. The EDF gives the proportion of observations for which
-the CDF is less than or equal to that of a given distance. In essence
-we're asking ``is the number of observations up to a given distance in
-line with what the model says they should be?'' (where `given values'
-are the observed distances). As usual for Q-Q plots, ``good'' models
-will have values close to the line \(y=x\), poor models will show more
-deviations from that line.
-
-We can inspect Q-Q plots visually, though for a large number of models
-this can be tiresome and prone to subjective judgements. Instead we can
-quantify the Q-Q plot's information using a Kolmogorov-Smirnov or
-Cramer-von Mises test (Burnham et al. 2004). Both test to see if the
-points from the EDF and CDF are from the same distribution. The
-Kolmogorov-Smirnov uses the test statistic of the largest difference
+the CDF is less than or equal to that of a given distance. This can be
+interpreted as assessing whether the number of observations up to a
+given distance in line with what the model says they should be (where
+`given values' are the observed distances). As usual for Q-Q plots,
+``good'' models will have values close to the line \(y=x\), poor models
+will show more deviations from that line.
+
+We can inspect Q-Q plots visually, though this is prone to subjective
+judgments. Instead we can quantify the Q-Q plot's information using a
+Kolmogorov-Smirnov or Cramer-von Mises test (Burnham et al. 2004). Both
+test whether points from the EDF and CDF are from the same distribution.
+The Kolmogorov-Smirnov uses the test statistic of the largest difference
 between a point on the Q-Q plot and the line \(y=x\), whereas the
@@
 function. We can see the goodness of fit tests below for two models for
-the amakihi data, first fitting a half-normal model without covariates
-or adjustments (not setting \code{adjustment=NULL} will force \code{ds}
-to fit a model with no adjustments), then calculating goodness of fit
-for that model and our hazard-rate model with observer and minutes after
-sunrise included:
+the amakihi data (Figure \ref{amakihi-qq}), first fitting a half-normal
+model without covariates or adjustments (not setting
+\code{adjustment=NULL} will force \code{ds} to fit a model with no
+adjustments), then calculating goodness of fit for that model and our
+hazard-rate model with observer and minutes after sunrise included:
 
@@
 
-Though fiting the detection function is the primary modelling step in
+Though fitting the detection function is the primary modelling step in
 distance sampling, we are really interested in estimating detectability,
@@
 abundance estimate. This section addresses these issues mathematically
-before showing how to estimate abundance and its variance in R.
+before showing how to estimate abundance and its variance in
+\proglang{R}.
 
@@
 
-First, in order to estimate abundance in the covered area
-(\(\hat{N}_\text{C}\)), we use the estimates of detection probability
-(the \(\{p_i; i=1,\ldots,n\}\), above) in a Horvitz-Thompson like
-estimator:
+First, to estimate abundance in the covered area (\(\hat{N}_\text{C}\)),
+we use the estimates of detection probability (the
+\(\{p_i; i=1,\ldots,n\}\), above) in a Horvitz-Thompson like estimator:
 
@@
 equal to 1 if objects only occur singly (Borchers and Burnham 2004).
-Thompson (2002) is the cannonical reference to this type of estimator;
+Thompson (2002) is the canonical reference to this type of estimator;
 intuitively, we can think of the estimates of detectability
 (\(\hat{p}_i\)) as ``inflating'' the group sizes (\(s_i\)) to account
-for incomplete detection -- we then sum to obtain the abundance
-estimate. For models that do not include covariates, \(\hat{p}_i\) is
-equal for all \(i\), so this is equivalent to summing the groups and
-inflating that sum by the corresponding
+for incomplete detection -- we then sum over all \(n\) detections to
+obtain the abundance estimate. For models that do not include
+covariates, \(\hat{p}_i\) is equal for all \(i\), so this is equivalent
+to summing the groups and inflating that sum by the corresponding
 \(\hat{p} (=\hat{p}_i \forall i)\).
@@
 
+\emph{have L and K been previously defined?}
+
 We can use the Horvitz-Thompson-like estimator to calculate the
@@
 \code{summary} output and the table produced by
-\code{summarize_ds_models}.
+\code{summarize_ds_models}. \emph{anything else to say about this, seems
+to mess up folks, including me}
 
@@
 We may wish to calculate abundance estimates for some sub-regions of the
-study region, we call these areas \emph{strata}. A stratum may be
-defined by, for example, a different habitat type or by the sex of the
-animal (or some combination) which may be of interest for biological or
-management reasons. In order to calculate estimates for a given
-stratification each observation must occur in a stratum which must be
-labelled with a \code{Region.Label} and have a corresponding \code{Area}
-(if we are using an animal characteristic like sex, we would have the
-areas be the same but if we were using say forested vs.~wetland habitat
-the areas of those strata would be different). Finally, we must also
-know which stratum a given transect lies in.
+study region, we call these areas \emph{strata}. For example, strata may
+be defined by habitat types or animal gender (or some combination) which
+may be of interest for biological or management reasons. To calculate
+estimates for a given stratification each observation must occur in a
+stratum which must be labelled with a \code{Region.Label} and have a
+corresponding \code{Area} (if we are using an animal characteristic like
+gender, we would have the areas be the same but if we were using say
+forested vs.~wetland habitat the areas of those strata would be
+different). Finally, we must also know the stratum in which transects
+lie.
 
@@
 Antarctic ice edge, respectively. Figure \ref{fig:minke-strata} shows
-the two strata, along with observations and transect lines.
+the two strata, along with observations and transect lines. \emph{figure
+number messed up}
 
@@
 variance-covariance matrix. We can then pre- and post-multiply this by
-the derivatives of \(\hat{N}_\text{C}\) \[
+the derivatives of \(\hat{N}_\text{C}\) with respect to the parameters
+of the detection function \[
 \widehat{\text{Var}}_\text{model}\left( \hat{N}_\text{C}\right) = \left(\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}\right)^\text{T} \left(\hat{\mathbf{H}}(\hat{\boldsymbol{\theta}})^{-1} \right)\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}
@@
 total line length) or \(n/T\) for point transects (where \(T\) is the
-total number of visits to all points). When covariates are included in
-the detection function, it is recommended that we substitute the \(n\)
-in the previous expressions with the estimated abundance \(\hat{N}\) as
+total number of visits to all points) \emph{earlier eqn uses notation
+\(K\) for number of points}. When covariates are included in the
+detection function, it is recommended that we substitute the \(n\) in
+the previous expressions with the estimated abundance \(\hat{N}\) as
 this will take into account the effects of the covariates.
@@
 \] where \(t_k\) is the number of visits to point \(k\) and
-\(T = \sum_{k=1}^K t_k\).
+\(T = \sum_{k=1}^K t_k\). \emph{this messes with the \(K\) notation even
+further}
 
@@
 
-We combine these two sources of variance by noting that squared
-coefficients of variation (approximately) add (Goodman 1960).
+We combine detection function variance with encounter rate variance by
+use of the \(\delta\)-method approximation (Goodman 1960).
 
@@
 
-Going back to the minke whale data, we have the information we require
-to calculate \(A\) and \(a\) above, so we can estimate abundance and its
+Returning to the minke whale data, we have the necessary information to
+calculate \(A\) and \(a\) above, so we can estimate abundance and its
 variance. When we supply data to \code{ds} in the ``flatfile'' format
 given above, \code{ds} will automatically calculate abundance estimates
-based on the survey setup in the data.
+based on the survey information in the data.
 
 Having already fitted a model to the minke whale data, we can see the
-results of the abundance estimation by simply looking at a model
-summary:
+results of the abundance estimation by viewing the model summary:
 
@@
 
-This prints a rather large amount of information. One useful feature is
-that the returned object is a \code{list} of \code{data.frames}, so we
-can again use \pkg{knitr}s \code{kable} function to create useful tables
-for publication. For example Table \ref{minke-abund}.
+This prints a rather large amount of information. The summary can be
+more concisely expressed by extracting information from the summary
+object. This object is a \code{list} of \code{data.frames}, so we can
+again use \pkg{knitr}s \code{kable} function to create summary tables of
+abundance estimates and measures of precision, such as Table
+\ref{minke-abund}.
 
@@
 
-The features of \pkg{Distance} are delibarately limited to provide a
-simplified interface for users. However, for more complex distance
-sampling-based analyses there are further related packages for modelling
-in \proglang{R}.
+The features of \pkg{Distance} are deliberately limited to provide a
+simplified interface for users. For more complex analyses of distance
+sampling data, there are further related packages for modelling in
+\proglang{R}.
 
@@
 The abundance estimates calculated here are based on the assumption that
-within a given stratum abundance is constant. We may extend this
-approach to many strata, making the area of each very small to account
-for small-scale variation in space. A more rigourous approach is to
-build a spatial model incorporating spatially-referenced environmental
-data (for example derived from GIS products). \pkg{Distance} interfaces
-with once such package to perform this type of analysis: \pkg{dsm}.
-So-called ``density surface modelling'' uses the generalized additive
-model framework (e.g. Wood 2006) to build models of abundance (that take
-into account detectability) as a function of environmental covariates,
-as part of a two stage model (Hedley and Buckland 2004; Miller et al.
-2013).
+within a given stratum density is uniform. We may extend this approach
+to many strata, making the area of each very small to account for
+small-scale variation in space. A more rigourous approach is to build a
+spatial model incorporating spatially-referenced environmental data (for
+example derived from GIS products). \pkg{Distance} interfaces with once
+such package to perform this type of analysis: \pkg{dsm}. So-called
+``density surface modelling'' uses the generalized additive model
+framework (e.g. Wood 2006) to build models of abundance (adjusting
+counts for imperfect detectability) as a function of environmental
+covariates, as part of a two stage model (Hedley and Buckland 2004;
+Miller et al. 2013).
 
@@
 
-Here we have given an introduction as to how to perform a distance
-sampling analysis in \proglang{R}. We have covered the possible models
-for detectability, model checking and selection and finally abundance
-and variance estimation.
+We have given an introduction as to how to perform a distance sampling
+analysis in \proglang{R}. We have covered the possible models for
+detectability, model checking and selection and finally abundance and
+variance estimation.
 
@@
 perform reproducible analyses of wildlife abundance for both managers
-and ecologists. We hope that this paper (itself written in RMarkdown)
-will provide a useful set of examples for those wishing to persue this.
+and ecologists. We hope that this paper provides useful examples for
+those wishing to pursue this.
 
@@
 most flexible and feature-complete. \textbf{Appendix ???} gives a
-feature comparison between \pkg{Distance} and the other packages
-available.
+feature comparison between \pkg{Distance} and other \proglang{R}
+packages for analysis of distance sampling data.
 

### 2015-12-05 paper.Rmd
first look at Eric's comments

@@
 * To include additional covariates into the detection function (see "Detection functions") we use a \code{data.frame}.
-The data.frame contains a column called \code{distance} and additional named columns for covariates potentially useful in modelling detectability (for example \code{observer} or \code{seastate}).
+The \code{data.frame} contains a column called \code{distance} and additional named columns for covariates potentially useful in modelling detectability (for example \code{observer} or \code{seastate}).
 Some column names are reserved: \code{object} for an observation identifier (see "Extensions"), \code{size} for group or cluster size (see "Detection functions" and "Abundance and variance estimation"), \code{detected} for whether an observation was detected (see "Extensions") and the columns described in the next bullet.
@@
 
-* *Shoulder*: we expect observers to be able to see objects near them, not just those directly adjacent to them. For this reason, we expect the detection function to be horizontal near zero distance.
+* *Shoulder*: we expect observers to be able to see objects near them, not just those directly in front of them. For this reason, we expect the detection function to be horizontal near zero distance.
 * *Non-increasing*: we do not think that observers should be more likely to see things distant from the transect than those near the transect.
@@
 f <- function(x) g(x)/g(0)
-curve(f, from=0, to=0.025, xlab="", ylab="", main="Half-normal with \n1 cosine adjustment",xlim=c(0,0.025), ylim=c(0,1))
+curve(f, from=0, to=0.025, xlab="", ylab="", main="Half-normal with \n1 cosine adjustment", xlim=c(0,0.025), ylim=c(0,1))
 title(xlab="Distance", line=2)
@@
 f <- function(x) g(x)/g(0)
-curve(f, from=0, to=0.025, xlab="", ylab="", main="Hazard-rate with \n1 cosine adjustment",xlim=c(0,0.025), ylim=c(0,1))
+curve(f, from=0, to=0.025, xlab="", ylab="", main="Hazard-rate with \n1 cosine adjustment", xlim=c(0,0.025), ylim=c(0,1))
 title(xlab="Distance", line=2)
@@
 f <- function(x) g(x)/g(0)
-curve(f, from=0, to=0.025, xlab="", ylab="", main="Hazard-rate with \n2 cosine adjustments",xlim=c(0,0.025), ylim=c(0,1))
+curve(f, from=0, to=0.025, xlab="", ylab="", main="Hazard-rate with \n2 cosine adjustments", xlim=c(0,0.025), ylim=c(0,1))
 title(xlab="Distance", line=2)
@@
 where $s_i$ are the sizes of the observed groups of objects, which is equal to 1 if objects only occur singly [@Borchers:2004wr].
-@thompson2012sampling is the canonical reference to this type of estimator; intuitively, we can think of the estimates of detectability ($\hat{p}_i$) as "inflating" the group sizes ($s_i$) to account for incomplete detection -- we then sum over all $n$ detections to obtain the abundance estimate.
+@thompson2012sampling is the canonical reference to this type of estimator; intuitively, we can think of the estimates of detectability ($\hat{p}_i$) as "inflating" the group sizes ($s_i$) to account for incomplete detection -- we then sum over the detections ($i$) to obtain the abundance estimate.
 For models that do not include covariates, $\hat{p}_i$ is equal for all $i$, so this is equivalent to summing the groups and inflating that sum by the corresponding $\hat{p} (=\hat{p}_i \forall i)$.
@@
 Figure \ref{fig:minke-strata} shows the two strata, along with observations and transect lines.
-*figure number messed up*
 
@@
 \includegraphics{minke-strata}
-\label{fig:minke-strata}
-\caption{Strata used for the minke whale data adapted from @Hedley:2004et.
+\caption{Strata used for the minke whale data adapted from Hedley and Buckland (2004).
 Points show the locations of observations along transect lines.
 The stepped line shows the boundary between North and South strata.
-Further details on the survey are available in @Branch:2001ua (simulated data is based on ``1992/93 Area III'' therein).}
+Further details on the survey are available in Branch and Butterworth (2001) (simulated data is based on ``1992/93 Area III'' therein).\label{fig:minke-strata}}
 \end{figure}
@@
 
-We combine detection function variance with encounter rate variance by use of the $\delta$-method approximation [@Goodman:1960dd].
+We combine these two sources of variance by noting that squared coefficients of variation (approximately) add [@Goodman:1960dd] (sometimes referred to as "the delta method").
 
@@
 Returning to the minke whale data, we have the necessary information to calculate $A$ and $a$ above, so we can estimate abundance and its variance.
-When we supply data to \code{ds} in the ``flatfile'' format given above, \code{ds} will automatically calculate abundance estimates based on the survey information in the data.
+When we supply data to \code{ds} in the "flatfile" format given above, \code{ds} will automatically calculate abundance estimates based on the survey information in the data.
 

### 2015-12-05 paper.Rmd
remove g(x) plots for amakihi

@@
 
-** ER suggests the g(x) plots for the amakihi be deleted **
-
-```{r minke-amakihi-hn-plot, echo=FALSE, warning=FALSE, fig.width=9, fig.height=6, fig.cap="Top row plots of fitted detection functions overlayed on the histograms of observed distances; left minke whale data, half-normal model; centre, amakihi data hazard-rate with observer as a covariate; right, amakihi data, hazard-rate model with observer and minutes after sunrise as covariates (for amakihi histograms are rescale to take into account sampler geometry).
-Bottom row shows plots of the probability density function for the amakihi models (as top row).
-For covariate models (centre and right columns) points indicate probability of detection for a given observation (given that observations covariate values) and lines indicate the average detection function.\\label{fig:minkeamakihi}"}
-par(mfrow=c(2,3))
+```{r minke-amakihi-hn-plot, echo=FALSE, warning=FALSE, fig.width=9, fig.height=4, fig.cap="Left: fitted detection functions overlayed on the histogram of observed distances for the minke whale data using half-normal model.
+Centre, right: plots of the probability density function for the amakihi models.
+Centre, hazard-rate with observer as a covariate; right, hazard-rate model with observer and minutes after sunrise as covariates.
+Points indicate probability of detection for a given observation (given that observations covariate values) and lines indicate the average detection function.\\label{fig:minkeamakihi}"}
+par(mfrow=c(1, 3))
 plot(minke_hn, showpoints=FALSE)
-plot(amakihi_hr_obs)
-plot(amakihi_hr_obs_mas)
-plot(1,1, type="n", axes=FALSE, xlab="", ylab="")
-plot(amakihi_hr_obs, pdf=TRUE)
-plot(amakihi_hr_obs_mas, pdf=TRUE)
+plot(amakihi_hr_obs, pdf=TRUE, cex=0.7)
+plot(amakihi_hr_obs_mas, pdf=TRUE, cex=0.7)
 ```

### 2015-12-05 paper.Rmd
$K$ -> $T$

@@
 where $A$ is the area of the study region to extrapolate the abundance estimate to and $a$ is the covered area.
-For line transects $a=2wL$ (twice the truncation distance multiplied by the total length of transects surveyed) and for points $a=\pi w^2 K$.
-
-*have L and K been previously defined?*
+For line transects $a=2wL$ (twice the truncation distance multiplied by the total length of transects surveyed, $L$) and for points $a=\pi w^2 T$ (where $\pi w^2$ is the area of a single surveyed circle and $T$ is the total number of circles sampled).
 

### 2015-12-05 paper.Rmd
more clarification of $K$/$T$

--- 2015-12-05 paper.Rmd
@@
 where $A$ is the area of the study region to extrapolate the abundance estimate to and $a$ is the covered area.
-For line transects $a=2wL$ (twice the truncation distance multiplied by the total length of transects surveyed) and for points $a=\pi w^2 K$.
-
-*have L and K been previously defined?*
+For line transects $a=2wL$ (twice the truncation distance multiplied by the total length of transects surveyed, $L$) and for points $a=\pi w^2 T$ (where $\pi w^2$ is the area of a single surveyed circle and $T$ is the total number of visits to the sampled points).
 
@@
 Sampling uncertainty can be characterised (rather than as simply $n$) by the *encounter rate*: the number of objects per unit transect.
-When covariates are not included in the detection function we can define the encounter rate as $n/L$ for line transects (where $L$ is the total line length) or $n/T$ for point transects (where $T$ is the total number of visits to all points) *earlier eqn uses notation $K$ for number of points*.
+When covariates are not included in the detection function we can define the encounter rate as $n/L$ for line transects (where $L$ is the total line length) or $n/T$ for point transects (where $T$ is the total number of visits to all points).
 When covariates are included in the detection function, it is recommended that we substitute the $n$ in the previous expressions with the estimated abundance $\hat{N}$ as this will take into account the effects of the covariates.
@@
 $$
-where $t_k$ is the number of visits to point $k$ and $T = \sum_{k=1}^K t_k$. *this messes with the $K$ notation even further*
+where $t_k$ is the number of visits to point $k$ and $T = \sum_{k=1}^K t_k$ (the total number of visits to all points is the sum of the visits to each point).
 

### 2015-12-05 paper.Rmd
add readdst

@@
 
+Distance for Windows has many users (**STAT HERE**) and they may be overwhelmed by the prospect of switching existing analyses to \proglang{R}.
+For that reason we have created the \pkg{readdst} package to interface with projects created by Distance for Windows'.
+The package can take analyses created using the CDS, MCDS and MRDS engines in Distance for Windows and extract data and create equivalent model definitions.
+\pkg{readdst} can also run these analyses and test the resulting statistics (for example, $\hat{N}$ or $\hat{P_a}$) calculated in \proglang{R} against those calculated by Distance for Windows.
+We hope that \pkg{readdst} will provide a useful transition to \proglang{R} for interested users.
+\pkg{readdst} is currently available on GitHub at \url{https://github.com/distancedevelopment/readdst}.
+
 # Conclusion

### 2015-12-06 paper.Rmd
add Jon and Lorenzo thanks
add STB, DLB and TAM to acknowledgements
reference to history in the JAE paper

@@
 
-For many years distance sampling analyses have been available via the Windows program Distance (or "DISTANCE"); for clarity henceforth "Distance for Windows" @Thomas:2010cf).
-From version 5 of Distance for Windows, R packages have been included to perform particular analyses (CITE Distance user manual).
+For many years distance sampling analyses have been available via the Windows program Distance (or "DISTANCE"); for clarity henceforth "Distance for Windows").
+From version 5 of Distance for Windows, R packages have been included to perform particular analyses [@Thomas:2010cf, further details of the history of Distance for Windows are included in that paper].
 This paper shows how to fit detection functions, perform model checking and selection and estimate abundance in \proglang{R} using the package \pkg{Distance}.
@@
 The authors would like to thank the many users of \pkg{Distance}, \pkg{mrds} and DISTANCE who have contributed bug reports and suggestions for improvements over the years.
+We would particularly like to thank Steve Buckland, David Borchers, Tiago Marques, Jon Bishop and Lorenzo Milazzo for their contributions.
 

### 2015-12-06 paper.Rmd
included download stat

@@
 
-Distance for Windows has many users (**STAT HERE**) and they may be overwhelmed by the prospect of switching existing analyses to \proglang{R}.
+Distance for Windows has many users (over 45,000 downloads since 2002) and they may be overwhelmed by the prospect of switching existing analyses to \proglang{R}.
 For that reason we have created the \pkg{readdst} package to interface with projects created by Distance for Windows'.

### 2015-12-07 paper.Rmd
fig 4 caption

@@
 
-```{r minke-amakihi-hn-plot, echo=FALSE, warning=FALSE, fig.width=9, fig.height=4, fig.cap="Left: fitted detection functions overlayed on the histogram of observed distances for the minke whale data using half-normal model.
+```{r minke-amakihi-hn-plot, echo=FALSE, warning=FALSE, fig.width=9, fig.height=4, fig.cap="Left: fitted detection function overlayed on the histogram of observed distances for the minke whale data using half-normal model.
 Centre, right: plots of the probability density function for the amakihi models.

### 2015-12-07 paper.Rmd
binned distances

@@
 As we will see in "Extensions", further information is also required when using more complex models.
+
+It is also possible to use distances collected in intervals ("binned", as opposed to "exact" distance data) to model the detection function.
+In this case a set of cutpoints for the bins must be provided rather than the \code{distance} column, these are specified as two columns \code{distbegin} and \code{distend}.
 

### 2015-12-07 paper.Rmd
first hack at abstract

@@
 abstract: >
-  The abstract of the article.
+  The estimation of the abundance of biological population is an essential part of applied ecology and wildlife management.
+To ensure that abundance estimates are accurate, we must account for uncertain detectability of plants and animals; distance sampling is a popular method for doing this.
+Here we show how models for detectabilty can be fitted using the \proglang{R} package \pkg{Distance}.
+We also show how \pkg{Distance} can be used to calculate Horvitz-Thompson-like estimates of abundance, interfaces to other packages to provide sophisticated analyses and a migration from previous software.
 keywords:
   # at least one keyword must be supplied
-  formatted: [distance sampling, abundance estimation, line transects, point transects, "\\proglang{R}"]
-  plain:     [distance sampling, abundance estimation, line transects, point transects, R]
+  formatted: [distance sampling, abundance estimation, line transects, point transects, detection functions, "\\proglang{R}"]
+  plain:     [distance sampling, abundance estimation, line transects, point transects, detection functions, R]
 preamble: >

### 2015-12-07 paper.Rmd
add Laura and Len as authors

@@
     email: Eric.Rexstad@st-andrews.ac.uk
+  - name: Len Thomas
+    affiliation: University of St Andrews
+    address: >
+      Centre for Research into Ecological and Environmental Modelling,
+      The Observatory,
+      St Andrews, Fife KY16 9LZ, Scotland
+    email: len.thomas@st-andrews.ac.uk
+  - name: Laura Marshall
+    affiliation: University of St Andrews
+    address: >
+      Centre for Research into Ecological and Environmental Modelling,
+      The Observatory,
+      St Andrews, Fife KY16 9LZ, Scotland
+    email: lhm@st-andrews.ac.uk
   - name: Jeffrey L Laake
-    affiliation: National Marine Mammal Laboratory
+    affiliation: NOAA
     address: >
+      National Marine Mammal Laboratory
       Alaska Fisheries Science Center

### 2015-12-07 paper.Rmd
migration what?

@@
 Here we show how models for detectabilty can be fitted using the \proglang{R} package \pkg{Distance}.
-We also show how \pkg{Distance} can be used to calculate Horvitz-Thompson-like estimates of abundance, interfaces to other packages to provide sophisticated analyses and a migration from previous software.
+We also show how \pkg{Distance} can be used to calculate Horvitz-Thompson-like estimates of abundance, interfaces to other packages to provide sophisticated analyses and a migration pathway from previous software.
 keywords:

### 2015-12-10 paper.Rmd
DLM edits up to "Detection Functions"

--- 2015-12-07 paper.Rmd
@@
 Here we show how models for detectabilty can be fitted using the \proglang{R} package \pkg{Distance}.
-We also show how \pkg{Distance} can be used to calculate Horvitz-Thompson-like estimates of abundance, interfaces to other packages to provide sophisticated analyses and a migration from previous software.
+We also show how \pkg{Distance} can be used to calculate Horvitz-Thompson-like estimates of abundance, its interface to other packages to provide sophisticated analyses and a migration pathway from previous software.
 keywords:
   # at least one keyword must be supplied
-  formatted: [distance sampling, abundance estimation, line transects, point transects, detection functions, "\\proglang{R}"]
-  plain:     [distance sampling, abundance estimation, line transects, point transects, detection functions, R]
+  formatted: [distance sampling, abundance estimation, line transects, point transects, detection functions, Horvitz-Thomspn, "\\proglang{R}"]
+  plain:     [distance sampling, abundance estimation, line transects, point transects, detection functions, Horvitz-Thomspn, R]
 preamble: >
@@
 
-For many years distance sampling analyses have been available via the Windows program Distance (or "DISTANCE"); for clarity henceforth "Distance for Windows").
-From version 5 of Distance for Windows, R packages have been included to perform particular analyses [@Thomas:2010cf, further details of the history of Distance for Windows are included in that paper].
-This paper shows how to fit detection functions, perform model checking and selection and estimate abundance in \proglang{R} using the package \pkg{Distance}.
+The Windows program Distance (or "DISTANCE"; for clarity henceforth "Distance for Windows") has been available for many years to perform distance sampling analyses (and before that program "TRANSECT").
+From version 5 of Distance for Windows, \proglang{R} packages have been included to perform some parts of an analysis [@Thomas:2010cf, further details of the history of Distance for Windows are included in that article].
+This article shows how to fit detection functions, perform model checking and selection, and estimate abundance in \proglang{R} using the package \pkg{Distance}.
 \pkg{Distance} is a wrapper package around the more complex (and more powerful) \pkg{mrds} and offers a subset of the analyses possible with that package.
@@
 Because we expect detectability to decrease with increasing distance from the sampler, we model detectability as a function of distance (plus perhaps other covariates, see below).
-We use the probabilities of detection calculated from the detection function to estimate an abundance for the area covered by the survey, which we then scale-up to the study region of interest.
+We use the probabilities of detection calculated from the detection function to estimate abundance in the area covered by the survey, which we then scale-up to the study region of interest.
 
 Distance sampling surveys are conducted using two transect types: line and point transects.
-In line transect sampling observers walk (or fly, sail, etc)  lines observing objects and recording the distances to the line; whereas in point transect sampling observers remain stationary at a location and record distances from that point.
+In line transect sampling observers walk (or fly, sail, etc)  lines, observing objects and recording the distances to the line; whereas in point transect sampling observers remain stationary at a location and record distances to objects from that point.
 Field methods are chosen to be suitable to species and habitat constraints [@buckland2015distance].
 
-```{r points-and-lines, fig.height=6, fig.width=7, echo=FALSE, message=FALSE, fig.cap="Left side plots show an example of a survey of an area containing a population of 500 objects, blue indicates sampler placement (top lines, bottom points) and red dots indicate detected individuals.
+```{r points-and-lines, fig.height=6, fig.width=7, echo=FALSE, message=FALSE, fig.cap="Left side plots show an example of a survey of an area containing a population of 500 objects, blue indicates sampler placement (top lines, bottom points), red dots indicate detected individuals and grey dots give the locations of unobserved individuals.
 The right side of the figure shows histograms of observed distances (again, lines top and points bottom).\\label{fig:pointslines}"}
@@
 For both points and lines, given the geometry of the sampler (see "Detection functions"), the histogram of distances should show a decreasing number of observations with increasing distance from the sampler.
-For line transects we expect objects to be distributed uniformly with respect to distance from the line with detectability causing the number of detections to decrease.
-For point transects, we recognise that as distance from the point increases, the area of the circle encompassed increases with distance squared; hence the number of objects available to be detected is a linearly increasing function of distance from the point.
+For line transects we expect objects to be distributed uniformly with respect to distance from the line with detectability causing the number of detections to decrease (as distance from the sampler increases).
+For point transects we recognise that as distance from the point increases, the area of the circle with a corresponding radius increases with distance squared, hence the number of objects available to be detected is a linearly increasing function of distance from the point.
 
-Using this histogram we can crudely estimate the drop-off in detectability by eye by tracing a line that approximates the tops of the histogram bars -- this is the detection function.
+Plotting a histogram of observed distances, we can crudely estimate the drop-off in detectability by eye if we trace a line that approximates the tops of the histogram bars: this is a detection function.
 \pkg{Distance} estimates the parameters for a fixed-form detection function using maximum likelihood estimation.
@@
 Figure \ref{fig:pointslines} shows sampling of a population of 500 individuals using line and point transects (left column) and their corresponding histograms (right column).
+Blue indicates the transects (lines or points), red points indicate that an object was observed and grey points give the locations of unobserved objects.
 
@@
 
-We demonstrate \pkg{Distance} using two data sets: one line transect and one point transect. These data sets have been chosen to be representative of data seen in practice.
+We show examples of analyses in \pkg{Distance} using two data sets: one line transect and one point transect. These data sets have been chosen to be representative of data seen in practice.
 
@@
 The data are simulated from models fitted to data from the International Whaling Commission's International Decade of Cetacean Research Southern Ocean Whale and Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer surveys [@Branch:2001ua].
-They consist of 99 observations and include geographical strata (near or distant from ice edge) and transect lengths on the effort expended.
+They consist of 99 observations and include a stratification based on location (near or distant from ice edge) and effort data (transect lengths).
 
@@
 The rest of the paper has this structure: we describe data organisation for use with \pkg{Distance}; models for the detection function are described in terms of formulation and examples of fitting in \proglang{R}.
-We then examine model checking, goodness of fit and model selection.
-Having illustrated how to obtain a good detection function, we show how to estimate abundance using that model, including stratification.
+We then show how to perform model checking, goodness of fit and model selection.
+We go on to show how to estimate abundance, including stratified estimates of abundance.
 The final two sections of the article look at extensions (both in terms of methodology and software) and put the package in a broader context amongst other software packages used for estimating the abundance of biological populations.
@@
 The two example data sets used here are distributed with \pkg{Distance} so readers can reproduce our results.
-However, data will be collected in the field and formatted correctly for use with \pkg{Distance}.
+However, generally data will be collected in the field and will need to be formatted correctly for use with \pkg{Distance}.
 The package allows for a flexible format for data input ranging from very simple to complex:
@@
 * To include additional covariates into the detection function (see "Detection functions") we use a \code{data.frame}.
-The \code{data.frame} contains a column called \code{distance} and additional named columns for covariates potentially useful in modelling detectability (for example \code{observer} or \code{seastate}).
+The \code{data.frame} contains a column called \code{distance} (containing the observed distances) and additional named columns for covariates potentially useful in modelling detectability (for example \code{observer} or \code{seastate}).
 Some column names are reserved: \code{object} for an observation identifier (see "Extensions"), \code{size} for group or cluster size (see "Detection functions" and "Abundance and variance estimation"), \code{detected} for whether an observation was detected (see "Extensions") and the columns described in the next bullet.
 * To estimate abundance beyond area covered by expended survey effort, additional information is required.
-This consists of information about which transect and occasion the observation was made (the \code{Sample.Label}), a column named \code{Effort} which gives the effort associated with that sample (for lines their length and for points the number of times that point was visited), the stratum the sample was located in (this may have any name and may be from pre- or post-survey stratification, see "Estimating abundance and variance") and strata areas (which has the same name as the stratum column, appended with \code{.Area}).
-We refer to this data format as "flatfile" where all information is contained in one table.
+This consists of information about which transect and occasion the observation was made (the \code{Sample.Label}), a column named \code{Effort} which gives the effort associated with that sample (for lines their length and for points the number of times that point was visited), the stratum the sample was located in (the \code{Region.Label}, which may be from pre- or post-survey stratification, see "Estimating abundance and variance") and the stratum areas (\code{Area}).
+For transects which were surveyed but have no observations we enter \code{NA} in their \code{distance} column.
+We refer to this data format (where all information is contained in one table) as "flatfile" as it can be easily created in a single spreadsheet.
 
@@
 
-It is also possible to use distances collected in intervals ("binned", as opposed to "exact" distance data) to model the detection function.
+It is also possible to use distances collected in intervals ("binned" or "grouped" data, as opposed to "exact" distances) to model the detection function.
 In this case a set of cutpoints for the bins must be provided rather than the \code{distance} column, these are specified as two columns \code{distbegin} and \code{distend}.
+More information on binned data is included in @Buckland:2001vm sections 4.5 and 7.4.1.2.
+
+The columns \code{distance}, \code{Area} and (in the case of line transects) \code{Effort} have associated units.
+We recommended that these are converted to SI units before starting any analysis to ensure that resulting abundance and density estimates have sensible units.
 
@@
 
-We will explore below the effects of including effort data during analysis.
+We will explore the effects of including effort and stratum data during analysis below.
 
@@
 
-```{r adjust-mix-comp, fig.width=9, fig.height=3, fig.cap="Possible shapes for the detection function when adjustments are included for half-normal and hazard-rate models.\\label{fig:keyadj}", echo=FALSE}
-par(mfrow=c(1,4), mar=c(3.5, 3, 2, 1) + 0.1)
+```{r adjust-mix-comp, fig.width=10, fig.height=3, fig.cap="Possible shapes for the detection function when adjustments are included for half-normal and hazard-rate models.\\label{fig:keyadj}", echo=FALSE}
+par(mfrow=c(1,4), mar=c(3.5, 3, 3, 1) + 0.1)
 
@@
 $$
-This can be a useful summary statistic and is included in the \code{summary} output and the table produced by \code{summarize_ds_models}.
-*anything else to say about this, seems to mess up folks, including me*
+This can be a useful summary statistic, giving us an idea of how detectable our $n$ observed animals would have to be in order to estimate the same $\hat{N}$ if there were no observed covariates.
+It can also be compared to similar estimates in mark-recapture studies.
+$P_a$ is included in the \code{summary} output and the table produced by \code{summarize_ds_models}.
 

### 2015-12-11 paper.Rmd
DLM checking done

@@
   # at least one keyword must be supplied
-  formatted: [distance sampling, abundance estimation, line transects, point transects, detection functions, Horvitz-Thomspn, "\\proglang{R}"]
-  plain:     [distance sampling, abundance estimation, line transects, point transects, detection functions, Horvitz-Thomspn, R]
+  formatted: [distance sampling, abundance estimation, line transects, point transects, detection functions, Horvitz-Thompson, "\\proglang{R}"]
+  plain:     [distance sampling, abundance estimation, line transects, point transects, detection functions, Horvitz-Thompson, R]
 preamble: >
@@
 ```{r setup, include=FALSE}
-library(knitr)
+library("knitr")
 opts_chunk$set(cache=TRUE)
@@
 From version 5 of Distance for Windows, \proglang{R} packages have been included to perform some parts of an analysis [@Thomas:2010cf, further details of the history of Distance for Windows are included in that article].
-This article shows how to fit detection functions, perform model checking and selection, and estimate abundance in \proglang{R} using the package \pkg{Distance}.
-\pkg{Distance} is a wrapper package around the more complex (and more powerful) \pkg{mrds} and offers a subset of the analyses possible with that package.
+This article shows how to fit detection functions, perform model checking and selection, and estimate abundance in \proglang{R} [@rcore] using the package \pkg{Distance} [@distance-pkg].
+\pkg{Distance} is a wrapper package around the more complex (and more powerful) \pkg{mrds} [@mrds-pkg] and offers a subset of the analyses possible with that package.
 
@@
 set.seed(131) # same results every time
-library(mgcv) # for inSide
+library("mgcv") # for inSide
 
@@
 ```{r minke-data-head, message=FALSE}
-library(Distance)
+library("Distance")
 head(minke)
@@
 The detection function describes the relationship between observed distances and probability of detection.
-The detection function itself models the probability $\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)$ and is usually denoted $g(y; \boldsymbol{\theta})$ where $y$ is distance (either from a line or point) and $\boldsymbol{\theta}$ is a vector of parameters to be estimated.
+The detection function itself models the probability $\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)$ and is usually denoted $g(y; \boldsymbol{\theta})$ where $y$ is distance (from a line or point) and $\boldsymbol{\theta}$ is a vector of parameters to be estimated.
 Our goal is to estimate an *average probability of detection* ($p$, average in the sense of an average over distances), so we must integrate out distance ($y$) from the detection function:
@@
 $$
-where $\pi(y)$ describes the distribution of objects with respect to the sampler and $\pi(x)=1/w$ for line transects and $\pi(r)=\frac{2r}{w^2}$ for point transects, taking into account the geometry of the sampler [@Buckland:2001vm, Chapter 3] (letting $x$ denote a perpendicular distance from a line and $r$ denote radial distance from a point).
+where $\pi(y)$ describes the distribution of objects with respect to the sampler; $\pi(x)=1/w$ for line transects and $\pi(r)=\frac{2r}{w^2}$ for point transects, taking into account the geometry of the sampler [@Buckland:2001vm, Chapter 3] (letting $x$ denote a perpendicular distance from a line and $r$ denote radial distance from a point).
 
-It is crucial that the detection function accurately models detectability at small distances.
-We are less worried by its behaviour further away from 0.
-To ensure that the model is not overly influence by distances far from zero, we truncate the distances beyond a given distance $w$ (known as the *truncation distance*).
-Fitting to distances at great distances from zero does not demonstrably improve the precision of abundance estimates [@Buckland:2001vm, 103-108, 151-153].
+It is crucial that the detection function accurately models detectability at small distances; we are less worried by its behaviour further away from 0.
+To ensure that the model is not overly influenced by distances far from zero, we truncate the distances beyond a given distance $w$ (known as the *truncation distance*).
+Including these larger distances in out analysis does not demonstrably improve the precision of abundance estimates [@Buckland:2001vm, sections 4.3 and 5.3].
 
@@
 
-* *Shoulder*: we expect observers to be able to see objects near them, not just those directly in front of them. For this reason, we expect the detection function to be horizontal near zero distance.
-* *Non-increasing*: we do not think that observers should be more likely to see things distant from the transect than those near the transect.
+* *Shoulder*: we expect observers to be able to see objects near them, not just those directly in front of them. For this reason, we expect the detection function to be flat near the line or point.
+* *Non-increasing*: we do not think that observers should be more likely to see distant objects than those nearer the transect.
 If this occurs, it usually indicates an issue with field procedure (that the distribution of objects with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this.
@@
 \begin{table}
-\caption{Modelling options for key plus adjustment series models for the detection function.}
+\caption{Modelling options for key plus adjustment series models for the detection function. Adapted from Buckland et al. (2001), section 2.4.}
 \begin{tabular}{llll}
@@
 
-A disadvantage of K+A models is that we must resort to constrained optimisation (via the \pkg{Rsolnp} package) to ensure that the resulting detection function is monotonic non-increasing over the whole range.
+A disadvantage of K+A models is that we must resort to constrained optimisation [via the \pkg{Rsolnp} package; @rsolnp-pkg] to ensure that the resulting detection function is monotonic non-increasing over its range.
 
@@
 The workhorse of detection function fitting in \pkg{Distance} is the \code{ds} function.
-Here we show off the formulations for the detection function we have seen above applied to the minke whale and amakihi data.
+Here we show some of the possible formulations for the detection function we have seen above applied to the minke whale and amakihi data.
 
@@
 
-Figure \ref{fig:minkeamakihi} (top left plot) shows the result of calling \code{plot} on the resulting model object.
+Figure \ref{fig:minkeamakihi} (left panel) shows the result of calling \code{plot} on the resulting model object.
 We can also call \code{summary} on the model object to get summary information about the fitted model (we postpone this to the next section).
@@
 
-By default \code{ds} assumes the data given to it is line transect, but we can switch to points using \code{transect="point"}.
-Including covariates in the scale is via the \code{formula=~...} argument to \code{ds}; a hazard-rate model for the amakihi that includes observer as a covariate can be specified by [truncating sightings at 82.5m, given in @Marques:2007ey]:
+By default \code{ds} assumes the data given to it has been collected as line transects, but we can switch to point transects using \code{transect="point"}.
+We can include covariates in the scale parameter via the \code{formula=~...} argument to \code{ds}.
+A hazard-rate model for the amakihi that includes observer as a covariate can be specified by [truncating sightings at 82.5m, given in @Marques:2007ey]:
 ```{r amakihi-hr-obs}
@@
 ```
-As with the minke whale model, we can plot the resulting detection function (Figure \ref{fig:minkeamakihi}, centre column).
-Because the amakihi model included covariates, the plot shows the detection function averaged over levels/values of the covariate.
+Note that here \code{ds} warns us that we have not supplied enough information to estimate abundance, unlike with the minke whale data.
+
+As with the minke whale model, we can plot the resulting model (Figure \ref{fig:minkeamakihi}, middle panel).
+Rather than detection functions, Figure \ref{fig:minkeamakihi} shows the probability density functions for the amakihi models.
+Probability density function plots give a better sense of fit of the model than the detection functions, as for point data we must rescale the histogram when plotting the detection function to take into account the geometry of the point sampler.
+The amakihi model included covariates, so the plot shows the detection function averaged over levels/values of the covariate.
 Points on the plot indicate probability of detection for each observation.
 For the `amakihi_hr_obs` model we see fairly clear levels of the observer covariate in the points.
-Looking at the left panel of Figure \ref{fig:minkeamakihi}, we can see this is less clear when adding minutes after sunrise as a covariate into the model:
+Looking at the right panel of Figure \ref{fig:minkeamakihi}, we can see this is less clear when adding minutes after sunrise as a covariate into the model:
 ```{r amakihi-hr-obs-mas, warning=FALSE}
@@
 
-The bottom row of Figure \ref{fig:minkeamakihi} shows the probability density functions for the amakihi models.
-Probability density function plots give a better sense of fit of the model than the detection functions, as for point data we must rescale the histogram when plotting the detection function to take into account the geometry of the point sampler.
-
 ```{r minke-amakihi-hn-plot, echo=FALSE, warning=FALSE, fig.width=9, fig.height=4, fig.cap="Left: fitted detection function overlayed on the histogram of observed distances for the minke whale data using half-normal model.
-Centre, right: plots of the probability density function for the amakihi models.
+Centre and right: plots of the probability density function for the amakihi models.
 Centre, hazard-rate with observer as a covariate; right, hazard-rate model with observer and minutes after sunrise as covariates.
@@
 To judge goodness of fit for detection functions when exact distances are used, we want to compare the cumulative distribution function (CDF) and empirical distribution function (EDF) for the detection function via a quantile-quantile plot (Q-Q plot).
-In our case the CDF evaluates the probability of observing an object at a distance less than or equal to some given value.
+The CDF evaluates the probability of observing an object at a distance less than or equal to some given value.
 The EDF gives the proportion of observations for which the CDF is less than or equal to that of a given distance.
-This can be interpreted as assessing whether the number of observations up to a given distance in line with what the model says they should be (where 'given values' are the observed distances).
+This can be interpreted as assessing whether the number of observations up to a given distance in line with what the model says they should be (where the "given values" are the observed distances).
 As usual for Q-Q plots, "good" models will have values close to the line $y=x$, poor models will show more deviations from that line.
@@
 We can generate a Q-Q plot and test results using the \code{gof_ddf} function.
-We can see the goodness of fit tests below for two models for the amakihi data (Figure \ref{amakihi-qq}), first fitting a half-normal model without covariates or adjustments (not setting \code{adjustment=NULL} will force \code{ds} to fit a model with no adjustments), then calculating goodness of fit for that model and our hazard-rate model with observer and minutes after sunrise included:
+Figure \ref{amakihi-qq} shows the goodness of fit tests for two models for the amakihi data.
+We first fitting a half-normal model without covariates or adjustments (not setting \code{adjustment=NULL} will force \code{ds} to fit a model with no adjustments):
 ```{r amakihi-gof, fig.keep="none"}
@@
 ```
-The corresponding Q-Q plots are shown in Figure \ref{amakihi-qq}.
+The corresponding Q-Q plots are shown in Figure \ref{amakihi-qq}, comparing the above model with the hazard-rate model with observer and minutes after sunrise included.
 
@@
 \pkg{Distance} includes a function to create table of summary information for fitted models, making it easy to get an overview of a large number of models at once.
-The \code{summarize_ds_models} function takes models as input and can be especially useful when paired with \pkg{knitr}'s \code{kable} function to create summary tables for publication.
-An example of this output is shown in Table \ref{tab:amakihi} and was generated by:
+The \code{summarize_ds_models} function takes models as input and can be especially useful when paired with \pkg{knitr}'s \code{kable} function to create summary tables for publication [@knitr-pkg].
+An example of this output is shown in Table \ref{tab:amakihi} and was generated by the following call to \code{summarize_ds_models}:
 ```{r summary-table-null, eval=FALSE}
-library(knitr)
 summarize_ds_models(amakihi_hr, amakihi_hr_obs, amakihi_hr_obs_mas)
@@
 ```{r summary-table, results="asis", echo=FALSE}
-library(knitr)
+library("knitr")
 kable(summarize_ds_models(amakihi_hr, amakihi_hr_obs, amakihi_hr_obs_mas),
@@
 
-First, to estimate abundance in the covered area ($\hat{N}_\text{C}$), we use the estimates of detection probability (the $\{p_i; i=1,\ldots,n\}$, above) in a Horvitz-Thompson like estimator:
+First, to estimate abundance in the covered area ($\hat{N}_\text{C}$), we use the estimates of detection probability (the $\{p_i; i=1,\ldots,n\}$, above) in a Horvitz-Thompson-like estimator:
 \begin{equation}
@@
 where $s_i$ are the sizes of the observed groups of objects, which is equal to 1 if objects only occur singly [@Borchers:2004wr].
-@thompson2012sampling is the canonical reference to this type of estimator; intuitively, we can think of the estimates of detectability ($\hat{p}_i$) as "inflating" the group sizes ($s_i$) to account for incomplete detection -- we then sum over the detections ($i$) to obtain the abundance estimate.
-For models that do not include covariates, $\hat{p}_i$ is equal for all $i$, so this is equivalent to summing the groups and inflating that sum by the corresponding $\hat{p} (=\hat{p}_i \forall i)$.
+@thompson2012sampling is the canonical reference to this type of estimator.
+Intuitively, we can think of the estimates of detectability ($\hat{p}_i$) as "inflating" the group sizes ($s_i$), we then sum over the detections ($i$) to obtain the abundance estimate.
+For models that do not include covariates, $\hat{p}_i$ is equal for all $i$, so this is equivalent to summing the groups and inflating that sum by dividing through by the corresponding $\hat{p} (=\hat{p}_i \forall i)$.
 
@@
 wetland habitat the areas of those strata would be different).
-Finally, we must also know the stratum in which transects lie.
+Finally, we must also know which stratum each observation occurs in.
 
@@
 
-1. *Model uncertainty*, from the estimation of the detection function parameters $\boldsymbol{\theta}$
+1. *Model uncertainty*, from the estimation of the detection function parameters $\boldsymbol{\theta}$.
 2. *Sampling uncertainty*, from the distribution of objects along the transect lines or between visiting occasions for points.
 
-We can see this by looking at the Horvitz-Thompson estimation in (\ref{ht}) and consider the terms which are random.
+We can see this by looking at the Horvitz-Thompson-like estimation in (\ref{ht}) and consider the terms which are random.
 These are: the detectability $\hat{p}_i$ (and hence the parameters of the detection function it is derived from) and $n$, the number of observations.
@@
 where the partial derivatives of $\hat{N}_\text{C}$ are evaluated at the MLE ($\hat{\theta}$) and $\mathbf{H}$ is the first partial Hessian (outer product of first derivatives of the log likelihood) for numerical stability [@Buckland:2001vm, p 62].
-Note that although we calculate uncertainty in $\hat{N}_\text{C}$, we can trivially scale-up to variance of $\hat{N}$.
+Note that although we calculate uncertainty in $\hat{N}_\text{C}$, we can trivially scale-up to variance of $\hat{N}$ (by noting that $\hat{N} = \frac{A}{a} \hat{N}_\text{C}$ and therefore $\widehat{\text{Var}}_\text{model}\left(\hat{N}\right) = \left( \frac{A}{a} \right)^2 \widehat{\text{Var}}_\text{model} \left(\hat{N}_\text{C} \right)$).
 
@@
 When covariates are not included in the detection function we can define the encounter rate as $n/L$ for line transects (where $L$ is the total line length) or $n/T$ for point transects (where $T$ is the total number of visits to all points).
-When covariates are included in the detection function, it is recommended that we substitute the $n$ in the previous expressions with the estimated abundance $\hat{N}$ as this will take into account the effects of the covariates.
+When covariates are included in the detection function, it is recommended that we substitute the $n$ in the encounter rate with the estimated abundance $\hat{N}$ as this will take into account the effects of the covariates [@Innes:2002ka].
 
@@
 $$
-\widehat{\text{Var}}_{\text{encounter},R2} = \frac{K}{L^2(K-1)} \sum_{k=1}^{K} l_k^2 \left( \frac{N_k}{l_k} - \frac{N}{L}\right)^2,
+\widehat{\text{Var}}_{\text{encounter},R2}\left( \hat{N}_\text{C}\right) = \frac{K}{L^2(K-1)} \sum_{k=1}^{K} l_k^2 \left( \frac{\hat{N_{\text{C}, k}}}{l_k} - \frac{\hat{N_\text{C}}}{L}\right)^2,
 $$
-where $l_k$ are the lengths of the $K$ transects (such that $L = \sum_{k=1}^K l_k$).
+where $l_k$ are the lengths of the $K$ transects (such that $L = \sum_{k=1}^K l_k$) and $\hat{N_{C,k}}$ is the abundance in the covered area for transect $k$.
 Whereas for points we use estimator "P3" from @Fewster:2009ku but again replacing $n$ by $\hat{N}$ to obtain the following estimator:
 $$
-\widehat{\text{Var}}_{\text{encounter},P3} = \frac{1}{T(K-1)} \sum_{k=1}^{K} t_k \left( \frac{N_k}{t_k} - \frac{N}{T}\right)^2,
+\widehat{\text{Var}}_{\text{encounter},P3}\left( \hat{N}_\text{C}\right) = \frac{1}{T(K-1)} \sum_{k=1}^{K} t_k \left( \frac{\hat{N_{\text{C}, k}}}{t_k} - \frac{\hat{N_\text{C}}}{T}\right)^2,
 $$
 where $t_k$ is the number of visits to point $k$ and $T = \sum_{k=1}^K t_k$ (the total number of visits to all points is the sum of the visits to each point).
+Again, it is straightforward to calculate the encounter rate variance for $\hat{N}$ from the encounter rate variance for $\hat{N}_\text{C}$.
 
-Various formulations for the encounter rate variance are discussed in detail in @Fewster:2009ku.
+Other formulations for the encounter rate variance are discussed in detail in @Fewster:2009ku.
 \pkg{Distance} implements all of the estimators of encounter rate variance given in that article.
@@
 ```
-This prints a rather large amount of information.
+This prints a rather large amount of information: first the detection function summary, then three tables:
+
+1.
+\code{Summary statistics}: giving the areas, covered areas, effort, number of observations, number of transects, encounter rate, its standard error and coefficient of variation for each stratum, then a total for the whole study area.
+2.
+\code{Abundance}: giving estimates, standard errors, coefficients of variation, lower and upper confidence intervals and finally the degrees of freedom for each stratum's abundance estimate, then a total for the whole study area.
+3. \code{Density}: lists the same statistics as \code{Abundance} but for a density estimate.
+
 The summary can be more concisely expressed by extracting information from the summary object.
-This object is a \code{list} of \code{data.frames}, so we can again use \pkg{knitr}s \code{kable} function to create summary tables of abundance estimates and measures of precision, such as Table \ref{minke-abund}.
+This object is a \code{list} of \code{data.frames}, so we can again use the \code{kable} function from \pkg{knitr} to create summary tables of abundance estimates and measures of precision, such as Table \ref{minke-abund}.
+We prepare the \code{data.frame} as follows before using \code{kable}:
 
-```{r minke-abundance-table}
+```{r minke-abundance-table-code, eval=FALSE}
+minke_table <- summary(minke_hn)$dht$individuals$N
+minke_table$lcl <- minke_table$ucl <- minke_table$df <- NULL
+colnames(minke_table) <- c("Stratum", "$\\hat{N}$", "$\\text{se}(\\hat{N}$)", 
+                           "$\\text{CV}(\\hat{N}$)")
+```
+
+```{r minke-abundance-table, echo=FALSE}
 minke_table <- summary(minke_hn)$dht$individuals$N
@@
 We may extend this approach to many strata, making the area of each very small to account for small-scale variation in space.
-A more rigourous approach is to build a spatial model incorporating spatially-referenced environmental data (for example derived from GIS products).
-\pkg{Distance} interfaces with once such package to perform this type of analysis: \pkg{dsm}.
+A more rigorous approach is to build a spatial model incorporating spatially-referenced environmental data (for example derived from GIS products).
+\pkg{Distance} interfaces with once such package to perform this type of analysis: \pkg{dsm} [@dsm-pkg].
 So-called "density surface modelling" uses the generalized additive model framework [e.g.
@@
 Distance for Windows has many users (over 45,000 downloads since 2002) and they may be overwhelmed by the prospect of switching existing analyses to \proglang{R}.
-For that reason we have created the \pkg{readdst} package to interface with projects created by Distance for Windows'.
+For that reason we have created the \pkg{readdst} [@readdst-pkg] package to interface with projects created by Distance for Windows.
 The package can take analyses created using the CDS, MCDS and MRDS engines in Distance for Windows and extract data and create equivalent model definitions.
@@
 
-In combination with tools such as \pkg{knitr} and \pkg{rmarkdown}, the helper functions in \pkg{Distance} provide a useful set of tools to perform reproducible analyses of wildlife abundance for both managers and ecologists.
+In combination with tools such as \pkg{knitr} and \pkg{rmarkdown} [@rmarkdown-pkg], the helper functions in \pkg{Distance} provide a useful set of tools to perform reproducible analyses of wildlife abundance for both managers and ecologists.
 We hope that this paper provides useful examples for those wishing to pursue this.
+More information on distance sampling can be found at \url{http://distancesampling.org} and a mailing list is maintained at \url{https://groups.google.com/forum/#!forum/distance-sampling}.
 
 We note that there are other packages available for performing distance sampling analyses in \proglang{R} but believe that \pkg{Distance} is the most flexible and feature-complete.
-**Appendix ???** gives a feature comparison between \pkg{Distance} and other \proglang{R} packages for analysis of distance sampling data.
+Appendix A gives a feature comparison between \pkg{Distance} and other \proglang{R} packages for analysis of distance sampling data.
 

### 2015-12-12 paper.Rmd
Minor changes to manuscript
Small wording changes:  this line doesn't sound right to me, but didn't
know how to correct:

"where $k$ is the key function and $\alpha_O$ is sum series of functions

@@
 * To estimate abundance beyond area covered by expended survey effort, additional information is required.
-This consists of information about which transect and occasion the observation was made (the \code{Sample.Label}), a column named \code{Effort} which gives the effort associated with that sample (for lines their length and for points the number of times that point was visited), the stratum the sample was located in (the \code{Region.Label}, which may be from pre- or post-survey stratification, see "Estimating abundance and variance") and the stratum areas (\code{Area}).
+This consists of information about transect and occasion the observation was made (the \code{Sample.Label}), a column named \code{Effort} which gives the effort associated with that sample (for lines their length and for points the number of times that point was visited), the stratum the sample was located in (the \code{Region.Label}, which may be from pre- or post-survey stratification, see "Estimating abundance and variance") and the stratum areas (\code{Area}).
 For transects which were surveyed but have no observations we enter \code{NA} in their \code{distance} column.
@@
 
-We will explore the effects of including effort and stratum data during analysis below.
+We will explore the consequences of including effort and stratum data during analysis below.
 
@@
 To ensure that the model is not overly influenced by distances far from zero, we truncate the distances beyond a given distance $w$ (known as the *truncation distance*).
-Including these larger distances in out analysis does not demonstrably improve the precision of abundance estimates [@Buckland:2001vm, sections 4.3 and 5.3].
+Including these larger distances in our analysis does not demonstrably improve the precision of abundance estimates [@Buckland:2001vm, sections 4.3 and 5.3].
 
@@
 We would like models to produce unbiased results without inclusion of these factors.
-* *Estimator efficiency*: we would like models to have low variances, but only given they satisfy the other properties above (which, if they are satisfied, would give low bias).
+* *Estimator efficiency*: we would like models to have low variances, given they satisfy the other properties above (which, if satisfied, would give low bias).
 
@@
 \code{Distance} includes the most popular of these models.
-Here we will detail the most popular detection function approach: "key function plus adjustments" (K+A).
+Here we detail the most popular detection function approach: "key function plus adjustments" (K+A).
 
@@
 Including covariates has an important implication for our calculation of detectability.
-Because we do not know what the true distribution of the covariates is we must calculate the probability of detection conditional on the observed values of the covariates:
+Because we do not know  the true distribution of the covariates,we must calculate the probability of detection conditional on the observed values of the covariates:
 $$
@@
 ```
-Note that \code{ds} will automatically select adjustment terms by AIC and shows the selection steps.
+Note that \code{ds} will automatically select adjustment terms by AIC and show the selection steps.
 
@@
 As with models fitted using \code{lm} or \code{glm} in \proglang{R}, we can use \code{summary} to give useful information about our fitted model.
-For example for our hazard-rate model for the amakihi, with observer as a covariate:
+For our hazard-rate model for the amakihi, with observer as a covariate:
 ```{r amakihi-summary}
@@
 This summary information includes details of the data and model specification, as well as the values of the coefficients ($\beta_j$) and their uncertainties, an "average" value for the detectability (see "Estimating abundance and variance" for details on how this is calculated) and its uncertainty.
-The final line gives an estimate of abundance for the area covered by the survey (covered in the next section).
+The final line gives an estimate of abundance for the area covered by the survey (see the next section).
 
@@
 To judge goodness of fit for detection functions when exact distances are used, we want to compare the cumulative distribution function (CDF) and empirical distribution function (EDF) for the detection function via a quantile-quantile plot (Q-Q plot).
-The CDF evaluates the probability of observing an object at a distance less than or equal to some given value.
+The CDF evaluates the probability of observing an object at a distance less than or equal to some value.
 The EDF gives the proportion of observations for which the CDF is less than or equal to that of a given distance.
 This can be interpreted as assessing whether the number of observations up to a given distance in line with what the model says they should be (where the "given values" are the observed distances).
-As usual for Q-Q plots, "good" models will have values close to the line $y=x$, poor models will show more deviations from that line.
+As usual for Q-Q plots, "good" models will have values close to the line $y=x$, poor models will show greater deviations from that line.
 
@@
 Figure \ref{amakihi-qq} shows the goodness of fit tests for two models for the amakihi data.
-We first fitting a half-normal model without covariates or adjustments (not setting \code{adjustment=NULL} will force \code{ds} to fit a model with no adjustments):
+We first fitting a half-normal model without covariates or adjustments (setting \code{adjustment=NULL} will force \code{ds} to fit a model with no adjustments):
 ```{r amakihi-gof, fig.keep="none"}
@@
 Once we have a set of models which fit well, we can use Akaike's Information Criterion (AIC) to select between models.
-\pkg{Distance} includes a function to create table of summary information for fitted models, making it easy to get an overview of a large number of models at once.
+\pkg{Distance} includes a function to create table of summary information for fitted models, making it easy to get an overview of a large number of models.
 The \code{summarize_ds_models} function takes models as input and can be especially useful when paired with \pkg{knitr}'s \code{kable} function to create summary tables for publication [@knitr-pkg].
@@
 $$
-where $A$ is the area of the study region to extrapolate the abundance estimate to and $a$ is the covered area.
+where $A$ is the area of the study region to which to extrapolate the abundance estimate and $a$ is the covered area.
 For line transects $a=2wL$ (twice the truncation distance multiplied by the total length of transects surveyed, $L$) and for points $a=\pi w^2 T$ (where $\pi w^2$ is the area of a single surveyed circle and $T$ is the total number of visits to the sampled points).
@@
 $$
-This can be a useful summary statistic, giving us an idea of how detectable our $n$ observed animals would have to be in order to estimate the same $\hat{N}$ if there were no observed covariates.
+This can be a useful summary statistic, giving us an idea of how detectable our $n$ observed animals would have to be to estimate the same $\hat{N}$ if there were no observed covariates.
 It can also be compared to similar estimates in mark-recapture studies.
@@
 wetland habitat the areas of those strata would be different).
-Finally, we must also know which stratum each observation occurs in.
+Finally, we must also know the stratum in which each observation occurs.
 

### 2015-12-13 paper.Rmd
start Tiago changes, need to merge Eric first...

--- 2015-12-11 paper.Rmd
@@
 abstract: >
-  The estimation of the abundance of biological population is an essential part of applied ecology and wildlife management.
+  The estimation of the abundance of a biological population is an essential part of applied ecology and wildlife management.
 To ensure that abundance estimates are accurate, we must account for uncertain detectability of plants and animals; distance sampling is a popular method for doing this.
@@
   # at least one keyword must be supplied
-  formatted: [distance sampling, abundance estimation, line transects, point transects, detection functions, Horvitz-Thompson, "\\proglang{R}"]
-  plain:     [distance sampling, abundance estimation, line transects, point transects, detection functions, Horvitz-Thompson, R]
+  formatted: [distance sampling, abundance estimation, line transects, point transects, detection functions, Horvitz-Thompson, "\\proglang{R}", Distance]
+  plain:     [distance sampling, abundance estimation, line transects, point transects, detection functions, Horvitz-Thompson, R, Distance]
 preamble: >
@@
 
-Distance sampling [@Buckland:2001vm; @Buckland:2004ts] encompasses a suite of field methods and statistical models used to estimate the abundance of biological populations.
+Distance sampling [@Buckland:2001vm; @Buckland:2004ts; @buckland2015distance] encompasses a suite of field methods and statistical models used to estimate the abundance of biological populations.
 Distance sampling field procedure can be thought of as an extension of plot sampling, where we wish to take into account the decreasing probability of detecting objects at increasing distance from the sampler.
@@
 The Windows program Distance (or "DISTANCE"; for clarity henceforth "Distance for Windows") has been available for many years to perform distance sampling analyses (and before that program "TRANSECT").
-From version 5 of Distance for Windows, \proglang{R} packages have been included to perform some parts of an analysis [@Thomas:2010cf, further details of the history of Distance for Windows are included in that article].
-This article shows how to fit detection functions, perform model checking and selection, and estimate abundance in \proglang{R} [@rcore] using the package \pkg{Distance} [@distance-pkg].
-\pkg{Distance} is a wrapper package around the more complex (and more powerful) \pkg{mrds} [@mrds-pkg] and offers a subset of the analyses possible with that package.
+From version 5 of Distance for Windows, \proglang{R} packages have been included to perform some parts of an analysis [further details of the history of Distance for Windows are included in @Thomas:2010cf].
+Here we show how to fit detection functions, perform model checking and selection, and estimate abundance in \proglang{R} [@rcore] using the package \pkg{Distance} [@distance-pkg].
+\pkg{Distance} is a wrapper package around the more complex (and more powerful) \pkg{mrds} package [@mrds-pkg] and offers a subset of the analyses possible with that package.
 

### 2015-12-13 paper.Rmd
incorporate Tiago's comments

@@
 Distance sampling surveys are conducted using two transect types: line and point transects.
-In line transect sampling observers walk (or fly, sail, etc)  lines, observing objects and recording the distances to the line; whereas in point transect sampling observers remain stationary at a location and record distances to objects from that point.
+In line transect sampling observers walk (or fly, sail, etc)  lines, observing objects and recording their perpendicular distance to the line; whereas in point transect sampling observers remain stationary at a location and record distances to objects from that point.
 Field methods are chosen to be suitable to species and habitat constraints [@buckland2015distance].
@@
 
-For both points and lines, given the geometry of the sampler (see "Detection functions"), the histogram of distances should show a decreasing number of observations with increasing distance from the sampler.
-For line transects we expect objects to be distributed uniformly with respect to distance from the line with detectability causing the number of detections to decrease (as distance from the sampler increases).
+For both points and lines, given the geometry of the sampler (see "Detection functions"), the histogram of distances should show a decreasing number of observations with increasing distance from the sampler (having taken into account sample geometry, see below).
+For line transects we expect objects to be distributed uniformly with respect to distance from the line (assuming that lines and points are placed independently of animal distribution) with detectability causing the number of detections to decrease (as distance from the sampler increases).
 For point transects we recognise that as distance from the point increases, the area of the circle with a corresponding radius increases with distance squared, hence the number of objects available to be detected is a linearly increasing function of distance from the point.
@@
 
-Figure \ref{fig:pointslines} shows sampling of a population of 500 individuals using line and point transects (left column) and their corresponding histograms (right column).
-Blue indicates the transects (lines or points), red points indicate that an object was observed and grey points give the locations of unobserved objects.
+Figure \ref{fig:pointslines} shows sampling of a population of 500 individuals using line and point transects and their corresponding histograms.
 
-## Data
+The rest of the paper has the following structure: we describe data organisation for use with \pkg{Distance}; models for the detection function are described in terms of formulation and examples of fitting in \proglang{R}.
+We then show how to perform model checking, goodness of fit and model selection.
+We go on to show how to estimate abundance, including stratified estimates of abundance.
+The final two sections of the article look at extensions (both in terms of methodology and software) and put the package in a broader context amongst other software packages used for estimating the abundance of biological populations.
+
+# Data
 
@@
 They consist of 99 observations and include a stratification based on location (near or distant from ice edge) and effort data (transect lengths).
+The survey is shown in Figure \ref{fig:minke-strata}.
+
+\begin{figure}
+\includegraphics{minke-strata}
+\caption{Strata used for the minke whale data adapted from Hedley and Buckland (2004).
+Points show the (simulated) locations of observations along transect lines.
+The stepped line shows the boundary between North and South strata.
+Further details on the survey are available in Branch and Butterworth (2001) (simulated data is based on ``1992/93 Area III'' therein).\label{fig:minke-strata}}
+\end{figure}
 
@@
 
-## The rest of the paper
-
-The rest of the paper has this structure: we describe data organisation for use with \pkg{Distance}; models for the detection function are described in terms of formulation and examples of fitting in \proglang{R}.
-We then show how to perform model checking, goodness of fit and model selection.
-We go on to show how to estimate abundance, including stratified estimates of abundance.
-The final two sections of the article look at extensions (both in terms of methodology and software) and put the package in a broader context amongst other software packages used for estimating the abundance of biological populations.
-
-# Data setup
+## Data setup
 
@@
 Some column names are reserved: \code{object} for an observation identifier (see "Extensions"), \code{size} for group or cluster size (see "Detection functions" and "Abundance and variance estimation"), \code{detected} for whether an observation was detected (see "Extensions") and the columns described in the next bullet.
-* To estimate abundance beyond area covered by expended survey effort, additional information is required.
+* To estimate abundance beyond the area covered by expended survey effort, additional information is required.
 This consists of information about transect and occasion the observation was made (the \code{Sample.Label}), a column named \code{Effort} which gives the effort associated with that sample (for lines their length and for points the number of times that point was visited), the stratum the sample was located in (the \code{Region.Label}, which may be from pre- or post-survey stratification, see "Estimating abundance and variance") and the stratum areas (\code{Area}).
@@
 
-The columns \code{distance}, \code{Area} and (in the case of line transects) \code{Effort} have associated units.
-We recommended that these are converted to SI units before starting any analysis to ensure that resulting abundance and density estimates have sensible units.
+The columns \code{distance}, \code{Area} and (in the case of line transects) \code{Effort} have associated units (though these are not explicitly included in a \pkg{Distance} analysis.
+For this reason, we recommended that these are converted to SI units before starting any analysis to ensure that resulting abundance and density estimates have sensible units.
 
@@
 The detection function itself models the probability $\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)$ and is usually denoted $g(y; \boldsymbol{\theta})$ where $y$ is distance (from a line or point) and $\boldsymbol{\theta}$ is a vector of parameters to be estimated.
-Our goal is to estimate an *average probability of detection* ($p$, average in the sense of an average over distances), so we must integrate out distance ($y$) from the detection function:
+Our goal is to estimate an *average probability of detection* ($p$, average in the sense of an average over distance from $0$ to $w$), so we must integrate out distance ($y$) from the detection function:
 $$
@@
 $$
-where $\pi(y)$ describes the distribution of objects with respect to the sampler; $\pi(x)=1/w$ for line transects and $\pi(r)=\frac{2r}{w^2}$ for point transects, taking into account the geometry of the sampler [@Buckland:2001vm, Chapter 3] (letting $x$ denote a perpendicular distance from a line and $r$ denote radial distance from a point).
+where $\pi(y)$ describes the distribution of objects with respect to the sampler; $\pi(x)=1/w$ for line transects and $\pi(r)=\frac{2r}{w^2}$ for point transects, taking into account the geometry of the sampler [@Buckland:2001vm, Chapter 3].
+When considering a particular transect type we let $x$ denote a perpendicular distance from a line and $r$ denote radial distance from a point (rather than using $y$).
 
@@
 If this occurs, it usually indicates an issue with field procedure (that the distribution of objects with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this.
-* *Model robust*: models should be flexible enough to make many different shapes.
+* *Model robust*: models should be flexible enough to fit many different shapes.
 * *Pooling robust*: many factors can affect the probability of detection and it is not possible to measure all of these.
@@
 
-```{r adjust-mix-comp, fig.width=10, fig.height=3, fig.cap="Possible shapes for the detection function when adjustments are included for half-normal and hazard-rate models.\\label{fig:keyadj}", echo=FALSE}
+```{r adjust-mix-comp, fig.width=10, fig.height=3, fig.cap="Possible (though not necessarily plausible) shapes for the detection function when adjustments are included for half-normal and hazard-rate models.\\label{fig:keyadj}", echo=FALSE}
 par(mfrow=c(1,4), mar=c(3.5, 3, 3, 1) + 0.1)
@@
 There are many factors that can affect the probability of detecting an object.
-These include things like the observer, the vessel or platform used, the sea state, weather conditions and time of day.
+These include things like the observer, size of group (if objects occur in groups), the vessel or platform used, the sea state, weather conditions and time of day.
 We assume that these variables affect detection only via the scale of the detection function (and do not affect the shape).
@@
 $$
-In the next section we will discuss model selection.
 
@@
 where $\mathbf{z_i}$ is the vector of $J$ covariates associated with observation $i$.
-For covariate models, we are calculating a value of "average" probability of detection (average in the sense of distance) per observation.
+For covariate models, we are calculating a value of "average" probability of detection (average in the sense of distance being integrated out) per observation.
 There will be as many unique values of $p(\mathbf{z_i})$ as there are unique covariate combinations in our data.
@@
 K+A models that include covariates and one or more adjustments cannot be guaranteed to be monotonic non-increasing for all covariate combinations, as we do not have any model for the distribution of the covariates.
+Without a distribution for the covariates, it's not possible to know what the behaviour of the detection function will be, so we cannot setup meaningful constraints -- so we cannot constrain the detection function to be monotonic.
 For this reason, we advise against using both adjustments and covariates in a detection function [see @Miller:2015hw for an example of when this can be problematic].
@@
 The EDF gives the proportion of observations for which the CDF is less than or equal to that of a given distance.
-This can be interpreted as assessing whether the number of observations up to a given distance in line with what the model says they should be (where the "given values" are the observed distances).
+This can be interpreted as assessing whether the number of observations up to a given distance is in line with what the model says they should be (where the "given values" are the observed distances).
 As usual for Q-Q plots, "good" models will have values close to the line $y=x$, poor models will show greater deviations from that line.
@@
 As it takes into account more information and is therefore more powerful, the Cramer-von Mises is generally preferred.
-A significant result from either tests indicates that the EDF and CDF do not come from the same distribution (and therefore the model is does not fit the data well).
+A significant result from either test indicates that the EDF and CDF do not come from the same distribution (and therefore the model is does not fit the data well).
 
-We can generate a Q-Q plot and test results using the \code{gof_ddf} function.
+We can generate a Q-Q plot and test results using the \code{gof_ds} function.
 Figure \ref{amakihi-qq} shows the goodness of fit tests for two models for the amakihi data.
-We first fitting a half-normal model without covariates or adjustments (setting \code{adjustment=NULL} will force \code{ds} to fit a model with no adjustments):
+We first fit a half-normal model without covariates or adjustments (setting \code{adjustment=NULL} will force \code{ds} to fit a model with no adjustments):
 ```{r amakihi-gof, fig.keep="none"}
-amakihi_hr <- ds(amakihi, truncation=82.5, transect="point", key="hn", adjustment=NULL)
-gof_ds(amakihi_hr)
+amakihi_hn <- ds(amakihi, truncation=82.5, transect="point", key="hn", adjustment=NULL)
+gof_ds(amakihi_hn)
 gof_ds(amakihi_hr_obs_mas)
 ```
-The corresponding Q-Q plots are shown in Figure \ref{amakihi-qq}, comparing the above model with the hazard-rate model with observer and minutes after sunrise included.
+So we can therefore conclude that the half-normal model does not pass our goodness of fit tests and should be discarded.
+The corresponding Q-Q plots are shown in Figure \ref{amakihi-qq}, comparing the half-normal model with the hazard-rate model with observer and minutes after sunrise included.
 
-```{r, amakihi-qq-comp, fig.width=6, fig.height=3, fig.cap="Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates) and hazard-rate model with observer and minutes after sunrise for the amakihi data.\\label{amakihi-qq}", echo=FALSE, results="hide"}
+```{r, amakihi-qq-comp, fig.width=6, fig.height=3, fig.cap="Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates; left) and hazard-rate model with observer and minutes after sunrise (right) for the amakihi data.\\label{amakihi-qq}", echo=FALSE, results="hide"}
 par(mfrow=c(1,2))
-gof_ds(amakihi_hr, asp=1, cex=0.3)  # shrinking dot size to obliterate 45-deg line less
+gof_ds(amakihi_hn, asp=1, cex=0.3)  # shrinking dot size to obliterate 45-deg line less
 gof_ds(amakihi_hr_obs_mas, asp=1, cex=0.3)
@@
 ```{r summary-table-null, eval=FALSE}
-summarize_ds_models(amakihi_hr, amakihi_hr_obs, amakihi_hr_obs_mas)
+summarize_ds_models(amakihi_hn, amakihi_hr_obs, amakihi_hr_obs_mas)
 ```
@@
 library("knitr")
-kable(summarize_ds_models(amakihi_hr, amakihi_hr_obs, amakihi_hr_obs_mas),
+kable(summarize_ds_models(amakihi_hn, amakihi_hr_obs, amakihi_hr_obs_mas),
       digits = 3, format = "latex", row.names = FALSE, escape=FALSE,
@@
 Though fitting the detection function is the primary modelling step in distance sampling, we are really interested in estimating detectability, and from that abundance.
-We also wish to calculate our certainty in each abundance estimate.
+We also wish to calculate our uncertainty for each abundance estimate.
 This section addresses these issues mathematically before showing how to estimate abundance and its variance in \proglang{R}.
@@
 We wish to obtain the abundance in a study region, of which we have sampled a (representative) subset.
-To do this we first calculate the abundance in the area we have surveyed (the *covered area*) to obtain $\hat{N}_\text{C}$, we can then scale this up to the full study area by multiplying it by the ratio of covered area to study area.
+To do this we first calculate the abundance in the area we have surveyed (the *covered area*) to obtain $\hat{N}_\text{C}$, we can then scale this up (based on the survey design) to the full study area by multiplying it by the ratio of covered area to study area.
+We discuss other methods for spatially explicit abundance estimation in "Extensions".
 
-First, to estimate abundance in the covered area ($\hat{N}_\text{C}$), we use the estimates of detection probability (the $\{p_i; i=1,\ldots,n\}$, above) in a Horvitz-Thompson-like estimator:
+First, to estimate abundance in the covered area ($\hat{N}_\text{C}$), we use the estimates of detection probability (the $\{\hat{p}(\mathbf{z}_i); i=1,\ldots,n\}$, above) in a Horvitz-Thompson-like estimator:
 \begin{equation}
-\hat{N}_\text{C} = \sum_{i=1}^n\frac{s_i}{\hat{p}_i},
+\hat{N}_\text{C} = \sum_{i=1}^n\frac{s_i}{\hat{p}(\mathbf{z}_i)},
 \label{ht}
@@
 @thompson2012sampling is the canonical reference to this type of estimator.
-Intuitively, we can think of the estimates of detectability ($\hat{p}_i$) as "inflating" the group sizes ($s_i$), we then sum over the detections ($i$) to obtain the abundance estimate.
-For models that do not include covariates, $\hat{p}_i$ is equal for all $i$, so this is equivalent to summing the groups and inflating that sum by dividing through by the corresponding $\hat{p} (=\hat{p}_i \forall i)$.
+Intuitively, we can think of the estimates of detectability ($\hat{p}(\mathbf{z}_i)$) as "inflating" the group sizes ($s_i$), we then sum over the detections ($i$) to obtain the abundance estimate.
+For models that do not include covariates, $\hat{p}(\mathbf{z}_i)$ is equal for all $i$, so this is equivalent to summing the groups and inflating that sum by dividing through by the corresponding $\hat{p} (=\hat{p}(\mathbf{z}_i) \forall i)$.
 
@@
 where $A$ is the area of the study region to which to extrapolate the abundance estimate and $a$ is the covered area.
-For line transects $a=2wL$ (twice the truncation distance multiplied by the total length of transects surveyed, $L$) and for points $a=\pi w^2 T$ (where $\pi w^2$ is the area of a single surveyed circle and $T$ is the total number of visits to the sampled points).
+For line transects $a=2wL$ (twice the truncation distance multiplied by the total length of transects surveyed, $L$) and for points $a=\pi w^2 T$ (where $\pi w^2$ is the area of a single surveyed circle and $T$ is the sum of the number of visits to the sampled points).
 
 We can use the Horvitz-Thompson-like estimator to calculate the "average" detectability for models which include covariates.
-We can consider what single detectability value would give the estimated $\hat{N}$ and therefore calculate:
+We can consider what single detectability value would give the estimated $\hat{N}_\text{C}$ and therefore calculate:
 $$
@@
 
-\begin{figure}
-\includegraphics{minke-strata}
-\caption{Strata used for the minke whale data adapted from Hedley and Buckland (2004).
-Points show the locations of observations along transect lines.
-The stepped line shows the boundary between North and South strata.
-Further details on the survey are available in Branch and Butterworth (2001) (simulated data is based on ``1992/93 Area III'' therein).\label{fig:minke-strata}}
-\end{figure}
-
 ## Variance
@@
 
-1. *Model uncertainty*, from the estimation of the detection function parameters $\boldsymbol{\theta}$.
+1. *Model parameter uncertainty*, from the estimation of the detection function parameters $\boldsymbol{\theta}$.
 2. *Sampling uncertainty*, from the distribution of objects along the transect lines or between visiting occasions for points.
@@
 We can see this by looking at the Horvitz-Thompson-like estimation in (\ref{ht}) and consider the terms which are random.
-These are: the detectability $\hat{p}_i$ (and hence the parameters of the detection function it is derived from) and $n$, the number of observations.
+These are: the detectability $\hat{p}(\mathbf{z}_i)$ (and hence the parameters of the detection function it is derived from) and $n$, the number of observations.
 We assume that the observed group size ($s_i$) is recorded without error.
 
-Model uncertainty can be addressed using standard maximum likelihood theory.
+Model parameter uncertainty can be addressed using standard maximum likelihood theory.
 We can invert the Hessian matrix of the likelihood to obtain a variance-covariance matrix.
@@
 
-Sampling uncertainty can be characterised (rather than as simply $n$) by the *encounter rate*: the number of objects per unit transect.
-When covariates are not included in the detection function we can define the encounter rate as $n/L$ for line transects (where $L$ is the total line length) or $n/T$ for point transects (where $T$ is the total number of visits to all points).
-When covariates are included in the detection function, it is recommended that we substitute the $n$ in the encounter rate with the estimated abundance $\hat{N}$ as this will take into account the effects of the covariates [@Innes:2002ka].
+Sampling uncertainty can be characterised by the *encounter rate*: the number of objects per unit transect (rather than just $n$).
+When covariates are not included in the detection function we can define the encounter rate as $n/L$ for line transects (where $L$ is the total line length) or $n/T$ for point transects (where $T$ is the total number of visits summed over all points).
+When covariates are included in the detection function, it is recommended that we substitute the $n$ in the encounter rate with the estimated abundance $\hat{N}_\text{C}$ as this will take into account the effects of the covariates [@Innes:2002ka].
 
@@
 where $l_k$ are the lengths of the $K$ transects (such that $L = \sum_{k=1}^K l_k$) and $\hat{N_{C,k}}$ is the abundance in the covered area for transect $k$.
-Whereas for points we use estimator "P3" from @Fewster:2009ku but again replacing $n$ by $\hat{N}$ to obtain the following estimator:
+Whereas for points we use estimator "P3" from @Fewster:2009ku but again replacing $n$ by $\hat{N}_\text{C}$ in the encounter rate definition, we obtain the following estimator:
 $$
@@
 We noted at the start of the article that \pkg{Distance} is a simpler wrapper around the package \pkg{mrds}.
-Additional features are available in \pkg{mrds} including the ability to model data where the assumption that detection is certain at zero distance from the line or point by using mark-recapture type methods when two observers are used [see @Burt:2014gu for an introduction].
+Additional features are available in \pkg{mrds} including the ability to model data where the assumption that detection is certain at zero distance from the line or point is violated, using mark-recapture type methods for double observer surveys [see @Burt:2014gu for an introduction].
 
@@
 A more rigorous approach is to build a spatial model incorporating spatially-referenced environmental data (for example derived from GIS products).
-\pkg{Distance} interfaces with once such package to perform this type of analysis: \pkg{dsm} [@dsm-pkg].
+\pkg{Distance} interfaces with one such package to perform this type of analysis: \pkg{dsm} [@dsm-pkg].
 So-called "density surface modelling" uses the generalized additive model framework [e.g.
@@
 For that reason we have created the \pkg{readdst} [@readdst-pkg] package to interface with projects created by Distance for Windows.
-The package can take analyses created using the CDS, MCDS and MRDS engines in Distance for Windows and extract data and create equivalent model definitions.
+The package can take analyses created using the CDS, MCDS and MRDS engines in Distance for Windows and extract data and create equivalent models in \proglang{R}.
 \pkg{readdst} can also run these analyses and test the resulting statistics (for example, $\hat{N}$ or $\hat{P_a}$) calculated in \proglang{R} against those calculated by Distance for Windows.

### 2015-12-13 paper.Rmd
First round of edits/comments - just got as far as Data section.

@@
 abstract: >
-  The estimation of the abundance of a biological population is an essential part of applied ecology and wildlife management.
-To ensure that abundance estimates are accurate, we must account for uncertain detectability of plants and animals; distance sampling is a popular method for doing this.
+  (LT: I have gone through putting comments to myself about things I need to come back and edit.
+In the end, I have written them within brackets in the text because I couldn't get the todo package to work, because square brackets are interpreted as markdown.
+In general, we might have been better to have used .Rnw but och well.
+Anyway DLM or others, feel free to comment on my comments!
+(Sorry if comments seem a bit negative -- I'd rather have had time to do the editing on this round so you can see my suggested text -- but it always takes me ages to do this bit and I had limited time right now -- I'll come back to it!...))
+  (LT: Need to define abundance; need to introduce concept of density also.) The estimation of the abundance of a biological population is an essential part of applied ecology and wildlife management.
+(LT: This sentence not strictly accurate, probably needs some editing.) To ensure that abundance estimates are accurate, we must account for uncertain detectability of plants and animals; distance sampling is a popular method for doing this.
 Here we show how models for detectabilty can be fitted using the \proglang{R} package \pkg{Distance}.
-We also show how \pkg{Distance} can be used to calculate Horvitz-Thompson-like estimates of abundance, its interface to other packages to provide sophisticated analyses and a migration pathway from previous software.
+(LT: Is it actually helpful to use the jargon like Horvitz-Thompson-like in the abstract?
+Almost no-one will know what we mean.) We also show how \pkg{Distance} can be used to calculate Horvitz-Thompson-like estimates of abundance, its interface to other packages to provide sophisticated analyses and a migration pathway from previous software.
 keywords:
@@
   \usepackage{booktabs}
+  \usepackage[colorinlistoftodos, color=green!40, textwidth=60, textsize=small]{todonotes}
+
 output: rticles::jss_article
@@
 
-Distance sampling [@Buckland:2001vm; @Buckland:2004ts; @buckland2015distance] encompasses a suite of field methods and statistical models used to estimate the abundance of biological populations.
-Distance sampling field procedure can be thought of as an extension of plot sampling, where we wish to take into account the decreasing probability of detecting objects at increasing distance from the sampler.
-We do this by building a model for detectability and use quantities calculated from the detection function to adjust the observed counts to obtain an estimate of abundance.
+(LT: I wonder who the intended audience is for this paper?
+The way it is currently written, it ``kinda'' introduces distance sampling as well as the relevant commands.
+Is this standard for a paper in this journal -- to introduce the methods as well as the software?
+Can we assume a reasonable level of analytical sophistication on the part of the reader, and so start with a few more formulae?
+Can we assume they know the methods, more or less, or point them at somewher where this is already introduced.
+(Perhaps I should go back and have a look at the EcolApps paper to see what approach we took there...))
 
-The Windows program Distance (or "DISTANCE"; for clarity henceforth "Distance for Windows") has been available for many years to perform distance sampling analyses (and before that program "TRANSECT").
+(LT: Need to define abundance, and add in the concept of density as well as population size.) Distance sampling [@Buckland:2001vm; @Buckland:2004ts; @buckland2015distance] encompasses a suite of field methods and statistical models used to estimate the abundance of biological populations.
+(LT: I would think we might want to put a little more here, defining plot sampling, fo rexample.) Distance sampling field procedure can be thought of as an extension of plot sampling, where we wish to take into account the decreasing probability of detecting objects at increasing distance from the sampler.
+(LT: Detection function is used without defining it.) We do this by building a model for detectability and use quantities calculated from the detection function to adjust the observed counts to obtain an estimate of abundance.
+
+(LT: Possibly edit to add the actual years.) The Windows program Distance (or "DISTANCE"; for clarity henceforth "Distance for Windows") has been available for many years to perform distance sampling analyses (and before that program "TRANSECT").
 From version 5 of Distance for Windows, \proglang{R} packages have been included to perform some parts of an analysis [further details of the history of Distance for Windows are included in @Thomas:2010cf].
-Here we show how to fit detection functions, perform model checking and selection, and estimate abundance in \proglang{R} [@rcore] using the package \pkg{Distance} [@distance-pkg].
+(LT: This won't mean much unless we've defined detection function first -- possibly need to do a little more work introducing the method and being explicit about where to go to find out more -- e.g., encyclopaedia articles, free online stuff, etc.
+The concepts mostly are explained a little further down here, so possibly just some re-ordering needed.) Here we show how to fit detection functions, perform model checking and selection, and estimate abundance in \proglang{R} [@rcore] using the package \pkg{Distance} [@distance-pkg].
 \pkg{Distance} is a wrapper package around the more complex (and more powerful) \pkg{mrds} package [@mrds-pkg] and offers a subset of the analyses possible with that package.
@@
 
-Census-type surveys (quadrat or strip transects assuming perfect detectability within strips or quadrats) are inefficient (requiring considerable field effort) and we should expect that not all objects (animals, plants, dung, etc) can be observed.
+(LT: a census is a complete count; I don't really think that sample surveys fall under this.
+So, a bit of editing needed here.) Census-type surveys (quadrat or strip transects assuming perfect detectability within strips or quadrats) are inefficient (requiring considerable field effort) and we should expect that not all objects (animals, plants, dung, etc) can be observed.
 Accounting for imperfect detectability is an important consideration when obtaining accurate estimates of abundance [@LahozMonfort:2013bg].
@@
 Because we expect detectability to decrease with increasing distance from the sampler, we model detectability as a function of distance (plus perhaps other covariates, see below).
-We use the probabilities of detection calculated from the detection function to estimate abundance in the area covered by the survey, which we then scale-up to the study region of interest.
+(LT: The word detection function is used again without it having been defined explicitly.) We use the probabilities of detection calculated from the detection function to estimate abundance in the area covered by the survey, which we then scale-up to the study region of interest.
 
@@
 
-```{r points-and-lines, fig.height=6, fig.width=7, echo=FALSE, message=FALSE, fig.cap="Left side plots show an example of a survey of an area containing a population of 500 objects, blue indicates sampler placement (top lines, bottom points), red dots indicate detected individuals and grey dots give the locations of unobserved individuals.
+```{r points-and-lines, fig.height=6, fig.width=7, echo=FALSE, message=FALSE, fig.cap="Left side plots show an example of a survey of an area containing a population of 500 objects; blue lines (top plot) and crosses (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals.
 The right side of the figure shows histograms of observed distances (again, lines top and points bottom).\\label{fig:pointslines}"}
@@
   # generate point location
-  # plot the line transect
-  points(pt$x[i], pt$y[i], pch=19, col="blue",cex=0.7)
+  # plot the point transect
+##LT: Changed from pch19 to pch4 so that colour-blind people can see where the samples are
+#  points(pt$x[i], pt$y[i], pch=19, col="blue",cex=0.7)
+  points(pt$x[i], pt$y[i], pch=17, col="blue",cex=1.5)
   # calculate the distances to objects from the line
@@
 
-For both points and lines, given the geometry of the sampler (see "Detection functions"), the histogram of distances should show a decreasing number of observations with increasing distance from the sampler (having taken into account sample geometry, see below).
+(LT: What is said about the decreasing number of observations is incorrect, really, and the "taking account of" just makes it confusing, I think.
+Needs a bit more editing - possibly re-order to explain what we see first.) For both points and lines, given the geometry of the sampler (see "Detection functions"), the histogram of distances should show a decreasing number of observations with increasing distance from the sampler (having taken into account sample geometry, see below).
 For line transects we expect objects to be distributed uniformly with respect to distance from the line (assuming that lines and points are placed independently of animal distribution) with detectability causing the number of detections to decrease (as distance from the sampler increases).
@@
 
-Plotting a histogram of observed distances, we can crudely estimate the drop-off in detectability by eye if we trace a line that approximates the tops of the histogram bars: this is a detection function.
+(LT: A detection function is not a line; it is also scaled differently.
+This needs editing.) Plotting a histogram of observed distances, we can crudely estimate the drop-off in detectability by eye if we trace a line that approximates the tops of the histogram bars: this is a detection function.
 \pkg{Distance} estimates the parameters for a fixed-form detection function using maximum likelihood estimation.
@@
 
-Figure \ref{fig:pointslines} shows sampling of a population of 500 individuals using line and point transects and their corresponding histograms.
+(LT: stranded one-sentence paragraph -- needs integrating.) Figure \ref{fig:pointslines} shows sampling of a population of 500 individuals using line and point transects and their corresponding histograms.
 
@@
 The final two sections of the article look at extensions (both in terms of methodology and software) and put the package in a broader context amongst other software packages used for estimating the abundance of biological populations.
+(LT: Do we mean software packages in general (e.g., Mark) or R packages?
+Do we actually do this -- i.e., talk about how Distance fits in among all this, or is this now an appendix, and does it cover all methods or just distance-based methods.)
+
+(LT: Only got this far: will do more later...)
 
@@
 
-We show examples of analyses in \pkg{Distance} using two data sets: one line transect and one point transect. These data sets have been chosen to be representative of data seen in practice.
+We show examples of analyses in \pkg{Distance} using two data sets: one line transect and one point transect.
+(LT: I think it might be better to delete this next sentence -- after all, two datasets can't possibly represent all the possibilities there are in the real world -- CDS, MCDS, MRDS, with and without covariates, problems such as spiked data, animal movement, measurement error, etc etc.) These data sets have been chosen to be representative of data seen in practice.
 

### 2016-01-05 paper.Rmd
Added paragraphs about mads and DSsim
Added paragraphs about mads and DSsim to Extensions section

@@
 
+Further distance sampling complexities such as uncertainty in measured covariates (e.g.
+cluster size) and model uncertainty (when two models have similar fit but substantially different estimates) can be incorporated using the multi-analysis distance sampling package \pkg{mads} [@mads-pkg].
+In addition, \pkg{mads} can also incorporate sightings for which species identification could not be achieved.
+This is done by estimating the abundance of these 'unidentified' sightings and pro-rating them as described in [@Gerrodette:2005].
+
+In addition to the analyses, an equally if not more important stage of a distance sampling survey is the survey design.
+We have therefore developed a package which allows users to test out different designs in their specific study region and tailor population attributes to reflect the species they are working with.
+In this way \pkg{DSsim} [@DSsim-pkg] allows users to more easily identify challenges unique to their study and select a survey design which is likely to yield the most accurate and precise results.
+
 Distance for Windows has many users (over 45,000 downloads since 2002) and they may be overwhelmed by the prospect of switching existing analyses to \proglang{R}.

### 2016-01-18 paper.Rmd
Updated abstract

@@
 abstract: >
-  (LT: I have gone through putting comments to myself about things I need to come back and edit.
-In the end, I have written them within brackets in the text because I couldn't get the todo package to work, because square brackets are interpreted as markdown.
-In general, we might have been better to have used .Rnw but och well.
-Anyway DLM or others, feel free to comment on my comments!
-(Sorry if comments seem a bit negative -- I'd rather have had time to do the editing on this round so you can see my suggested text -- but it always takes me ages to do this bit and I had limited time right now -- I'll come back to it!...))
-  (LT: Need to define abundance; need to introduce concept of density also.) The estimation of the abundance of a biological population is an essential part of applied ecology and wildlife management.
-(LT: This sentence not strictly accurate, probably needs some editing.) To ensure that abundance estimates are accurate, we must account for uncertain detectability of plants and animals; distance sampling is a popular method for doing this.
-Here we show how models for detectabilty can be fitted using the \proglang{R} package \pkg{Distance}.
-(LT: Is it actually helpful to use the jargon like Horvitz-Thompson-like in the abstract?
-Almost no-one will know what we mean.) We also show how \pkg{Distance} can be used to calculate Horvitz-Thompson-like estimates of abundance, its interface to other packages to provide sophisticated analyses and a migration pathway from previous software.
+
+  The study of plants or animals often involves the estimation of population size and/or their spatial distribution.
+These characteristics are important considerations in wildlife management.
+We introduce the \proglang{R} package \pkg{Distance} that uses distance sampling data to estimate populations size (abundance) and can be used in conjunction with another \proglang{R} package \pkg{dsm} to estimate spatial distribution.
+We describe example analyses using this package as well demonstrating the links with other \proglang{R} packages offering specialist solutions to specific distance sampling challenges.
+We demonstrate how this package provides a migration pathway from previous software, thereby allowing us to deliver cutting edge methods to the users more quickly.
+
 keywords:

### 2016-01-22 paper.Rmd
Edits to Intro
Edited initial section of into

@@
 (LT: I wonder who the intended audience is for this paper?
-The way it is currently written, it ``kinda'' introduces distance sampling as well as the relevant commands.
+(LM: - Distance for Windows users and other people looking do distance sampling in R?) The way it is currently written, it ``kinda'' introduces distance sampling as well as the relevant commands.
 Is this standard for a paper in this journal -- to introduce the methods as well as the software?
@@
 Can we assume they know the methods, more or less, or point them at somewher where this is already introduced.
-(Perhaps I should go back and have a look at the EcolApps paper to see what approach we took there...))
+(Perhaps I should go back and have a look at the EcolApps paper to see what approach we took there...
+(LM: The paper on Distance in JAppEcol?
+There is quite a big intro on distance sampling there.)))
 
-(LT: Need to define abundance, and add in the concept of density as well as population size.) Distance sampling [@Buckland:2001vm; @Buckland:2004ts; @buckland2015distance] encompasses a suite of field methods and statistical models used to estimate the abundance of biological populations.
-(LT: I would think we might want to put a little more here, defining plot sampling, fo rexample.) Distance sampling field procedure can be thought of as an extension of plot sampling, where we wish to take into account the decreasing probability of detecting objects at increasing distance from the sampler.
-(LT: Detection function is used without defining it.) We do this by building a model for detectability and use quantities calculated from the detection function to adjust the observed counts to obtain an estimate of abundance.
+Distance sampling [@Buckland:2001vm; @Buckland:2004ts; @buckland2015distance] encompasses a suite of methods used to estimate the density and/or abundance of biological populations.
+Distance sampling can be thought of as an extension of plot sampling.
+Plot sampling involves selecting a number of plots (small areas) at random within the study area and counting the objects of interest that are contained within each plot.
+By selecting the plots at random we can assume that the density of objects in the plots is representative of the study area as a whole.
+One of the key assumptions of plot sampling is that all objects within each of the plots are counted.
+Distance sampling relaxes this assumption in that observers are no longer required to see and count everything.
+
+Conventional distance sampling assumes the observer is either stood at a point or moving along a line and will only see everything that occurs exactly at that point or on the line.
+It is then expected that the further an object is from the sampler (the point or line) the less likely it is that the observer will see it.
+By recording the distance to each of the detected objects we can fit a model to describe the probability of detection given distance from the sampler, which we refer to as the ‘detection function’.
+The detection function can be used to infer how many objects were missed and thereby produce estimates of density and/or abundance.
+Further information on distance sampling, including its application, various extensions to deal with more complex situations and recent methodological developments can be found at http://distancesampling.org/dbib.html.
+
+The Windows program Distance (or "DISTANCE"; for clarity henceforth "Distance for Windows") was first released in November 1998 as Distance for Windows version 3.5 and was initially based on its predecessor “TRANSECT”.
+Since this time it has evolved to include various design and analysis features.
+Distance for Windows versions 5 onwards have included \proglang{R} packages as the analysis engines providing additional more complex analysis options than those offered by the original FROTRAN engines.
+(LM Add part here to say what this paper describes… pathway to sustainability using R?) Further details of the history of Distance for Windows are included in @Thomas:2010cf.
+
+(LM: still need to incorporate this)Here we show how to fit detection functions, perform model checking and selection, and estimate abundance in \proglang{R} [@rcore] using the package \pkg{Distance} [@distance-pkg].
+\pkg{Distance} is a wrapper package around the more complex (and more powerful) \pkg{mrds} package [@mrds-pkg] and offers a subset of the analyses possible with that package.
 
@@
 From version 5 of Distance for Windows, \proglang{R} packages have been included to perform some parts of an analysis [further details of the history of Distance for Windows are included in @Thomas:2010cf].
+
 (LT: This won't mean much unless we've defined detection function first -- possibly need to do a little more work introducing the method and being explicit about where to go to find out more -- e.g., encyclopaedia articles, free online stuff, etc.
-The concepts mostly are explained a little further down here, so possibly just some re-ordering needed.) Here we show how to fit detection functions, perform model checking and selection, and estimate abundance in \proglang{R} [@rcore] using the package \pkg{Distance} [@distance-pkg].
-\pkg{Distance} is a wrapper package around the more complex (and more powerful) \pkg{mrds} package [@mrds-pkg] and offers a subset of the analyses possible with that package.
+The concepts mostly are explained a little further down here, so possibly just some re-ordering needed.)
 

### 2016-01-22 paper.Rmd
More abstract updates
A few changes

@@
 These characteristics are important considerations in wildlife management.
-We introduce the \proglang{R} package \pkg{Distance} that uses distance sampling data to estimate populations size (abundance) and can be used in conjunction with another \proglang{R} package \pkg{dsm} to estimate spatial distribution.
-We describe example analyses using this package as well demonstrating the links with other \proglang{R} packages offering specialist solutions to specific distance sampling challenges.
+We introduce the \proglang{R} package \pkg{Distance} that uses distance sampling data to estimate populations size (abundance) and is compatible with another \proglang{R} package \pkg{dsm} which can be used to model spatial distribution.
+We describe how users can obtain estimates of abundance and/or density using the package \pkg{Distance} as well documenting the links it provides with other \proglang{R} packages offering specialist solutions to specific distance sampling challenges.
 We demonstrate how this package provides a migration pathway from previous software, thereby allowing us to deliver cutting edge methods to the users more quickly.
@@
 
-Conventional distance sampling assumes the observer is either stood at a point or moving along a line and will only see everything that occurs exactly at that point or on the line.
+In conventional distance sampling the observer is either stood at a point or moving along a line and we only assume they see everything that occurs exactly at that point or on the line.
 It is then expected that the further an object is from the sampler (the point or line) the less likely it is that the observer will see it.
@@
 The detection function can be used to infer how many objects were missed and thereby produce estimates of density and/or abundance.
-Further information on distance sampling, including its application, various extensions to deal with more complex situations and recent methodological developments can be found at http://distancesampling.org/dbib.html.
+(LM: perhaps move this?
+Further information on distance sampling, including its application, various extensions to deal with more complex situations and recent methodological developments can be found at http://distancesampling.org/dbib.html.)
 
@@
 Distance for Windows versions 5 onwards have included \proglang{R} packages as the analysis engines providing additional more complex analysis options than those offered by the original FROTRAN engines.
+(LM to finish: The current version of Distance links to 4 R packages...
+Development in R progressing rapidly so allowing users to access new tools faster?)
+
 (LM Add part here to say what this paper describes… pathway to sustainability using R?) Further details of the history of Distance for Windows are included in @Thomas:2010cf.

### 2016-01-25 paper.Rmd
Edited Intro Section 1

@@
 One of the key assumptions of plot sampling is that all objects within each of the plots are counted.
-Distance sampling relaxes this assumption in that observers are no longer required to see and count everything.
+Distance sampling relaxes this assumption in that observers are no longer required to see and count everything within selected plots.
 
-In conventional distance sampling the observer is either stood at a point or moving along a line and we only assume they see everything that occurs exactly at that point or on the line.
-It is then expected that the further an object is from the sampler (the point or line) the less likely it is that the observer will see it.
+Conventional distance sampling assumes the observer is either stood at a point or moving along a line and will only see everything that occurs exactly at that point or on the line.
+It is then expected that the further away an object is from the point or line (the sampler) the less likely it is that the observer will see it.
 By recording the distance to each of the detected objects we can fit a model to describe the probability of detection given distance from the sampler, which we refer to as the ‘detection function’.
 The detection function can be used to infer how many objects were missed and thereby produce estimates of density and/or abundance.
-(LM: perhaps move this?
-Further information on distance sampling, including its application, various extensions to deal with more complex situations and recent methodological developments can be found at http://distancesampling.org/dbib.html.)
 
-The Windows program Distance (or "DISTANCE"; for clarity henceforth "Distance for Windows") was first released in November 1998 as Distance for Windows version 3.5 and was initially based on its predecessor “TRANSECT”.
+The Windows program Distance (or "DISTANCE"; for clarity henceforth "Distance for Windows") can be used to fit detection functions to such distance sampling data.
+It was first released in November 1998 as Distance for Windows version 3.5 and was initially based on its predecessor “TRANSECT”.
 Since this time it has evolved to include various design and analysis features.
-Distance for Windows versions 5 onwards have included \proglang{R} packages as the analysis engines providing additional more complex analysis options than those offered by the original FROTRAN engines.
-(LM to finish: The current version of Distance links to 4 R packages...
-Development in R progressing rapidly so allowing users to access new tools faster?)
+Distance for Windows versions 5 onwards have included \proglang{R} [@rcore]  packages as the analysis engines providing additional more complex analysis options than those offered by the original FROTRAN engine.
+Further details of the history of Distance for Windows are included in @Thomas:2010cf.
 
-(LM Add part here to say what this paper describes… pathway to sustainability using R?) Further details of the history of Distance for Windows are included in @Thomas:2010cf.
-
-(LM: still need to incorporate this)Here we show how to fit detection functions, perform model checking and selection, and estimate abundance in \proglang{R} [@rcore] using the package \pkg{Distance} [@distance-pkg].
-\pkg{Distance} is a wrapper package around the more complex (and more powerful) \pkg{mrds} package [@mrds-pkg] and offers a subset of the analyses possible with that package.
-
-(LT: Possibly edit to add the actual years.) The Windows program Distance (or "DISTANCE"; for clarity henceforth "Distance for Windows") has been available for many years to perform distance sampling analyses (and before that program "TRANSECT").
-From version 5 of Distance for Windows, \proglang{R} packages have been included to perform some parts of an analysis [further details of the history of Distance for Windows are included in @Thomas:2010cf].
-
-(LT: This won't mean much unless we've defined detection function first -- possibly need to do a little more work introducing the method and being explicit about where to go to find out more -- e.g., encyclopaedia articles, free online stuff, etc.
-The concepts mostly are explained a little further down here, so possibly just some re-ordering needed.)
+As Distance for Windows becomes increasingly reliant on analyses performed in \proglang{R} and new methods are being developed at a rate where it is unfeasible to make all these available in the Windows interface, we are encouraging the use of our \proglang{R} packages directly.
+Until now those wishing to use \proglang{R} for straight forward distance sampling analyses would have had to negotiate the complex package \pkg{mrds} [@mrds-pkg] designed for mark-recapture distance sampling.
+\pkg{Distance} is a wrapper package around \pkg{mrds} making it easier to get started with basic distance sampling analyses in \proglang{R}.
+We demonstrate how to use \pkg{Distance} to fit detection functions, perform model checking and selection, and estimate abundance and/or density.
 

### 2016-01-27 paper.Rmd
Further editing on Intro section 1
Now I have edited the first part of the intro I am not sure if Section
1.1 wants to go in a different direction rather than being a basic
intro to distance sampling. I think that the point transect figures
were a bit confusing without quite a bit more explanation and I’m not
sure we want to go there but could still talk about line transects. But
I did wonder if section 1.1 wanted to focus on why use our distance
package rather than the other options? What do you all think (do you
like the intro after the restructuring and if so what should section
1.1 focus o

@@
 
+<!--
 (LT: I wonder who the intended audience is for this paper?
@@
 (LM: The paper on Distance in JAppEcol?
-There is quite a big intro on distance sampling there.)))
+There is quite a big intro on distance sampling there.))) -->
 
@@
 Distance sampling relaxes this assumption in that observers are no longer required to see and count everything within selected plots.
+While plot sampling techniques are adequate for static populations occuring at high density they are inefficient for more sparsely distributed populations.
+Distance sampling provides a more efficient solution in such circumstances.
 
 Conventional distance sampling assumes the observer is either stood at a point or moving along a line and will only see everything that occurs exactly at that point or on the line.
-It is then expected that the further away an object is from the point or line (the sampler) the less likely it is that the observer will see it.
+It is then expected that the further away an object is from the point or line (also known as the sampler or transect) the less likely it is that the observer will see it.
 By recording the distance to each of the detected objects we can fit a model to describe the probability of detection given distance from the sampler, which we refer to as the ‘detection function’.
 The detection function can be used to infer how many objects were missed and thereby produce estimates of density and/or abundance.
+Further information on distance sampling, including the theory, various extensions to deal with more complex situations and recent methodological developments can be found at http://distancesampling.org/dbib.html
 
 The Windows program Distance (or "DISTANCE"; for clarity henceforth "Distance for Windows") can be used to fit detection functions to such distance sampling data.
-It was first released in November 1998 as Distance for Windows version 3.5 and was initially based on its predecessor “TRANSECT”.
-Since this time it has evolved to include various design and analysis features.
-Distance for Windows versions 5 onwards have included \proglang{R} [@rcore]  packages as the analysis engines providing additional more complex analysis options than those offered by the original FROTRAN engine.
-Further details of the history of Distance for Windows are included in @Thomas:2010cf.
+It was first released in November 1998 as Distance for Windows version 3.5.
+Since this time it has evolved to include various design and analysis features [@Thomas:2010cf].
+Distance for Windows versions 5 onwards have included \proglang{R} [@rcore]  packages as the analysis engines providing additional, more complex analysis options than those offered by the original FORTRAN engine.
 
 As Distance for Windows becomes increasingly reliant on analyses performed in \proglang{R} and new methods are being developed at a rate where it is unfeasible to make all these available in the Windows interface, we are encouraging the use of our \proglang{R} packages directly.
-Until now those wishing to use \proglang{R} for straight forward distance sampling analyses would have had to negotiate the complex package \pkg{mrds} [@mrds-pkg] designed for mark-recapture distance sampling.
+In addition, the core packages in \proglang{R} provide a variety of functions useful for data exploration with minimal data formatting.
+The plotting of a histogram of detection distances only requires these values to be stored in a numerical vector, an exercise that is recommended not only prior to analyses but also early on in the data collection process _(LM need ref here @buckland2015distance ?)_.
+
+Until now those wishing to use our \proglang{R} packages for straight forward distance sampling analyses would have had to negotiate the complex package \pkg{mrds} [@mrds-pkg] designed for mark-recapture distance sampling.
+In addition, \pkg{mrds} requires a complex data structure before analyses can be completed.
 \pkg{Distance} is a wrapper package around \pkg{mrds} making it easier to get started with basic distance sampling analyses in \proglang{R}.
-We demonstrate how to use \pkg{Distance} to fit detection functions, perform model checking and selection, and estimate abundance and/or density.
+Like the histogram function, the most basic detection function estimation only requires a numeric vector of distances.
+Here we demonstrate how to use \pkg{Distance} to fit detection functions, perform model checking and selection, and estimate abundance and/or density.
 
-## Distance sampling
+## Distance sampling in R?
 
-(LT: a census is a complete count; I don't really think that sample surveys fall under this.
-So, a bit of editing needed here.) Census-type surveys (quadrat or strip transects assuming perfect detectability within strips or quadrats) are inefficient (requiring considerable field effort) and we should expect that not all objects (animals, plants, dung, etc) can be observed.
-Accounting for imperfect detectability is an important consideration when obtaining accurate estimates of abundance [@LahozMonfort:2013bg].
-Using the extra information gained by recording distance from the sampler to the observation, it is possible to model detectability.
-Because we expect detectability to decrease with increasing distance from the sampler, we model detectability as a function of distance (plus perhaps other covariates, see below).
-(LT: The word detection function is used again without it having been defined explicitly.) We use the probabilities of detection calculated from the detection function to estimate abundance in the area covered by the survey, which we then scale-up to the study region of interest.
+_(LM I wonder if here we should focus on comparing Distance to other R packages offering distance sampling analyses here?
+Some of the info below could be incorporated into the paragraphs above (I have done this to some extent..
+do we need more on distance sampling theory above?)...
+not done much with this section in case we want to do something different here.)_
 
-Distance sampling surveys are conducted using two transect types: line and point transects.
-In line transect sampling observers walk (or fly, sail, etc)  lines, observing objects and recording their perpendicular distance to the line; whereas in point transect sampling observers remain stationary at a location and record distances to objects from that point.
+<!-- While plot sampling techniques are adequate for static populations occuring at high density they are inefficient for more sparsely distributed populations.
+Distance sampling provides a more efficient solution in such circumstances.
+In addition, for cryptic species it simply may not be possible to detect all individuals within a given plot and we must find a way to account for imperfect detectability when estimating abundance [@LahozMonfort:2013bg].
+This paper does not provide a detailed explanation of the theory behind distance sampling but instead concentrates on it's application using the package \pkg{Distance}.
+Further information on distance sampling, including the theory, various extensions to deal with more complex situations and recent methodological developments can be found at http://distancesampling.org/dbib.html -->
+
+Conventional distance sampling analysis simply relies on estimating the probability of detection given how far an object is from the sampler.
+Note that in the case of populations that occur in groups (clusters) the whole group or cluster is treated as an individual object.
+Figure \ref{fig:lines} shows example data collection on a simulated population of 500 individuals using line transects and the corresponding histogram of distances recorded by the observer.
+As expected we see in the histogram of Fig.
+\ref{fig:lines} that as the perpendicular distance from the transect line increases, the number of detections made in each distance interval decreases.
+\pkg{Distance} provides a selection of functions to model this probability of detection detailed in Section...
+below.
+
+```{r lines, fig.height=3, fig.width=7, echo=FALSE, message=FALSE, fig.cap="Left side plot shows an example of a survey of an area containing a population of 500 objects; blue lines indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals.
+The right side of the figure shows a histogram of observed distances.\\label{fig:lines}"}
+opar <- par(mfrow=c(1,2))
+# line and point transect examples
+# top row points/samplers in space
+# bottom row histograms
+
+# generate some animals to sample
+set.seed(131) # same results every time
+library("mgcv") # for inSide
+
+# generate population locations
+N <- 500
+x <- runif(N)
+y <- runif(N)
+
+## line transects
+opar2 <- par(mar=c(1, 1, 1, 1) + 0.1)
+plot(x, y, pch=19, asp=1, cex=0.6, main="",col="grey", axes=FALSE, xlab="", ylab="")
+polygon(x=c(0,0,1,1,0), y=c(0,1,1,0,0))
+# generate some lines
+# in this case we don't randomise the offset of the grid
+lt <- list(x=c(-0.0125,-0.0125,0.0125,0.0125,-0.0125), y=c(0,1,1,0,0))
+# set sigma
+sigma <- 0.01
+# storage for detected distances
+detected_distances <- c()
+for(i in 1:5){
+  # calculate next strip location
+  lt$x <- lt$x+0.15
+  # plot the line transect
+  lines(x=rep(mean(range(lt$x)),2), y=c(0,1), col="blue",lty=2)
+  # calculate the distances to objects from the line
+  distances <- abs(lt$x - x)
+  # randomly decide which were detected
+  detected <- exp(-distances^2/(2*sigma^2)) > runif(length(distances))
+  # plot those objects detected
+  points(x[detected], y[detected], pch=19, cex=0.6, col="red")
+  # collect the distances to detected objects
+  detected_distances <- c(detected_distances, distances[detected])
+}
+
+#par(opar2)
+par(mar=c(4, 4, 1, 2) + 0.1)
+hist(detected_distances, main="", xlab="Observed perpendicular distances", freq = FALSE)
+
+par(opar)
+
+```
+
+The probability of detecting an object may not only depend on how far it is from the obsever but also on other factors such as weather conditions, ground cover, cluster size etc.
+The \pkg{Distance} package also allows the incorporation of such covariates into the modelling process of the detection function.
+
+_(LM How does Distance use the detection function to estimate abundance is it any different to mrds does it infact use dht?)_
+
+<!--
+
+(LM: do we need to talk about assumptions here?)
+
+(LM: detail what Distance deals with?)
+
 Field methods are chosen to be suitable to species and habitat constraints [@buckland2015distance].
 
-```{r points-and-lines, fig.height=6, fig.width=7, echo=FALSE, message=FALSE, fig.cap="Left side plots show an example of a survey of an area containing a population of 500 objects; blue lines (top plot) and crosses (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals.
+```{r points-and-lines, fig.height=6, fig.width=7, echo=FALSE, message=FALSE, fig.cap="Left side plots show an example of a survey of an area containing a population of 500 objects; blue lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals.
 The right side of the figure shows histograms of observed distances (again, lines top and points bottom).\\label{fig:pointslines}"}
@@
 par(mar=c(4, 4, 1, 2) + 0.1)
-hist(detected_distances, main="", xlab="Observed distances")
+hist(detected_distances, main="", xlab="Observed perpendicular distances")
 
@@
 par(mar=c(4, 4, 1, 2) + 0.1)
-hist(detected_distances, main="", xlab="Observed distances")
+hist(detected_distances, main="", xlab="Observed radial distances")
 
@@
 
-(LT: stranded one-sentence paragraph -- needs integrating.) Figure \ref{fig:pointslines} shows sampling of a population of 500 individuals using line and point transects and their corresponding histograms.
+(LT: stranded one-sentence paragraph -- needs integrating.)
 
@@
 
-(LT: Only got this far: will do more later...)
+(LT: Only got this far: will do more later...) -->
 
@@
 
-We show examples of analyses in \pkg{Distance} using two data sets: one line transect and one point transect.
-(LT: I think it might be better to delete this next sentence -- after all, two datasets can't possibly represent all the possibilities there are in the real world -- CDS, MCDS, MRDS, with and without covariates, problems such as spiked data, animal movement, measurement error, etc etc.) These data sets have been chosen to be representative of data seen in practice.
+We introduce two example analyses performed in \pkg{Distance}: one line transect and one point transect. These data sets have been chosen as they represent typical data seen in practice.
 
@@
 
-The line transect data is based on a survey of Antarctic minke whales (*Balaenoptera bonaerensis*).
-The data are simulated from models fitted to data from the International Whaling Commission's International Decade of Cetacean Research Southern Ocean Whale and Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer surveys [@Branch:2001ua].
-They consist of 99 observations and include a stratification based on location (near or distant from ice edge) and effort data (transect lengths).
+The line transect data have been simulated from models fitted to Antarctic minke whale (*Balaenoptera bonaerensis*) data (_LM why is the Minke whale data simulated rather than ust using original data, just data ownership/permissions?_).
+These data were collected as part of the International Whaling Commission's International Decade of Cetacean Research Southern Ocean Whale and Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer surveys [@Branch:2001ua].
+This data set consists of 99 observations that are stratified based on location (near or distant from ice edge) and effort data (transect lengths)(LM did the original dataset have 99 observations or the simulated dataset or both?).
 The survey is shown in Figure \ref{fig:minke-strata}.
+_(LM wonder if we need more info on how the data were simulated...
+based only on detection function or on spatial model too?
+I notice in the figure text that the simulated data are based on ``1992/93 Area III'' should this info go here too?)_
 
@@
 
-The two example data sets used here are distributed with \pkg{Distance} so readers can reproduce our results.
-However, generally data will be collected in the field and will need to be formatted correctly for use with \pkg{Distance}.
-The package allows for a flexible format for data input ranging from very simple to complex:
+The two example data sets used here are distributed preformatted with \pkg{Distance} so readers can reproduce our analyses.
+Generally, data collected in the field will require some formatting for use with \pkg{Distance}.
+However, the package allows for a flexible format of data input ranging from very simple to complex dependent on the type of analysis you wish to perform:
 

### 2016-01-29 paper.Rmd
Added availability plot
to help explain the histogram of detection distances for point transects

@@
 The detection function can be used to infer how many objects were missed and thereby produce estimates of density and/or abundance.
-Further information on distance sampling, including the theory, various extensions to deal with more complex situations and recent methodological developments can be found at http://distancesampling.org/dbib.html
 
@@
 
-## Distance sampling in R?
-
-_(LM I wonder if here we should focus on comparing Distance to other R packages offering distance sampling analyses here?
-Some of the info below could be incorporated into the paragraphs above (I have done this to some extent..
-do we need more on distance sampling theory above?)...
-not done much with this section in case we want to do something different here.)_
+## Distance sampling
 
@@
 Further information on distance sampling, including the theory, various extensions to deal with more complex situations and recent methodological developments can be found at http://distancesampling.org/dbib.html -->
+
+Conventional distance sampling relies on a number of assumptions.
+Firstly, like plot sampling, distance sampling assumes that the samplers have been located at random within the study area.
+By ensuring that the samplers are located independently of the objects in the study region we know that the "availability" of objects with respect to line transect samplers should be constant with increasing perpendicular distance.
+In contrast,
 
@@
 
-```{r lines, fig.height=3, fig.width=7, echo=FALSE, message=FALSE, fig.cap="Left side plot shows an example of a survey of an area containing a population of 500 objects; blue lines indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals.
-The right side of the figure shows a histogram of observed distances.\\label{fig:lines}"}
-opar <- par(mfrow=c(1,2))
-# line and point transect examples
-# top row points/samplers in space
-# bottom row histograms
+```{r detectdists, fig.height=5.5, fig.width=8, echo=FALSE, message=FALSE, fig.cap="Top panels show an example detection function (left), the expected avaiability of objects (middle) and the resulting destribution of detections (right) for line transects.
+Bottom panels show an example detection function (left), the expected avaiability of objects (middle) and the resulting destribution of detections (right) for point transects.
+\\label{fig:detectiondist}"}
 
-# generate some animals to sample
-set.seed(131) # same results every time
-library("mgcv") # for inSide
+opar <- par(mfrow=c(2,3))
+#line and point transect examples
 
-# generate population locations
-N <- 500
-x <- runif(N)
-y <- runif(N)
+x <- seq(0,500, length = 100)
+y <- exp(-((x^2)/(2*(200^2))))
 
-## line transects
-opar2 <- par(mar=c(1, 1, 1, 1) + 0.1)
-plot(x, y, pch=19, asp=1, cex=0.6, main="",col="grey", axes=FALSE, xlab="", ylab="")
-polygon(x=c(0,0,1,1,0), y=c(0,1,1,0,0))
-# generate some lines
-# in this case we don't randomise the offset of the grid
-lt <- list(x=c(-0.0125,-0.0125,0.0125,0.0125,-0.0125), y=c(0,1,1,0,0))
-# set sigma
-sigma <- 0.01
-# storage for detected distances
-detected_distances <- c()
-for(i in 1:5){
-  # calculate next strip location
-  lt$x <- lt$x+0.15
-  # plot the line transect
-  lines(x=rep(mean(range(lt$x)),2), y=c(0,1), col="blue",lty=2)
-  # calculate the distances to objects from the line
-  distances <- abs(lt$x - x)
-  # randomly decide which were detected
-  detected <- exp(-distances^2/(2*sigma^2)) > runif(length(distances))
-  # plot those objects detected
-  points(x[detected], y[detected], pch=19, cex=0.6, col="red")
-  # collect the distances to detected objects
-  detected_distances <- c(detected_distances, distances[detected])
-}
+#Line transect example
+plot(x, y, xlab = "Perpendicular distance (m)", ylab = "Probability of Detection", type = "l", ylim = c(0,1.1), main = "Probability of Detection", xaxp = c(0,1,5))
+line.availability <- rep(1,100)
+plot(x, line.availability, xlab = "Perpendicular distance (m)", ylab = "Availability", type = "l", ylim = c(0,1.1), xlim = c(0,500), yaxt = "n", main = "Expected Availability")
+plot(x,y, xlab = "Perpendicular distance (m)", ylab= "", type = "l", ylim = c(0,1.1), yaxt = "n", main = "Observed Distribution")
 
-#par(opar2)
-par(mar=c(4, 4, 1, 2) + 0.1)
-hist(detected_distances, main="", xlab="Observed perpendicular distances", freq = FALSE)
-
-par(opar)
+#Point Transect example
+plot(x, y, xlab = "Radial distance (m)", ylab = "Probability of Detection", type = "l", ylim = c(0,1.1), main = "Probability of Detection", xaxp = c(0,1,5))
+point.availability <- 2*pi*x
+plot(x, point.availability, xlab = "Radial distance (m)", ylab = "Availability", type = "l", ylim = c(0,3300), xlim = c(0,500), yaxt = "n", main = "Expected Availability")
+point.observed <- y*point.availability
+plot(x, point.observed, xlab = "Radial distance (m)", ylab= "", type = "l", ylim = c(0,800), yaxt = "n", main = "Observed Distribution")
 
@@
 _(LM How does Distance use the detection function to estimate abundance is it any different to mrds does it infact use dht?)_
-
-<!--
 
@@
 
-(LT: Only got this far: will do more later...) -->
+(LT: Only got this far: will do more later...)
 

### 2016-01-29 paper.Rmd
First go at section 1.1
…. more to come

@@
 
-Conventional distance sampling relies on a number of assumptions.
-Firstly, like plot sampling, distance sampling assumes that the samplers have been located at random within the study area.
-By ensuring that the samplers are located independently of the objects in the study region we know that the "availability" of objects with respect to line transect samplers should be constant with increasing perpendicular distance.
-In contrast,
+Like plot sampling, distance sampling assumes that the samplers have been located at random within the study area.
+Given that the samplers are located independently of the objects, for line transect studies we know that the "availability" of objects should be constant with increasing perpendicular distance.
+However, for point transect surveys the available survey area increases linearly with radial distance implying that the number of objects available for detection should also increase linerary.
+Figure \ref{fig:detectiondist} shows how these availability functions when combined with a detection function give rise to the observed distribution of recorded distances.
 
-Conventional distance sampling analysis simply relies on estimating the probability of detection given how far an object is from the sampler.
-Note that in the case of populations that occur in groups (clusters) the whole group or cluster is treated as an individual object.
-Figure \ref{fig:lines} shows example data collection on a simulated population of 500 individuals using line transects and the corresponding histogram of distances recorded by the observer.
-As expected we see in the histogram of Fig.
-\ref{fig:lines} that as the perpendicular distance from the transect line increases, the number of detections made in each distance interval decreases.
-\pkg{Distance} provides a selection of functions to model this probability of detection detailed in Section...
-below.
+Given that we can assume these particular availability function there are a number of ways we can use either the estimated detection function or the composite function we fit to our data to estimate abundance.
+One simple idea begins with finding the area under the composite function we fit to our data out to a truncation distance w.
+If we then find the area under the availability function out to distance w then by dividing the former area by the latter gives us an estimate of the propoertion of objects seen within distance w of the transect.
+We can then scale up the number of objects we saw, first to give us an estimate of how many objects were within distance w of the transects and then again by the proportion of the study area we have sampled to give an estimate of abundance for the entire study area.
 
@@
 
-The probability of detecting an object may not only depend on how far it is from the obsever but also on other factors such as weather conditions, ground cover, cluster size etc.
-The \pkg{Distance} package also allows the incorporation of such covariates into the modelling process of the detection function.
-
-_(LM How does Distance use the detection function to estimate abundance is it any different to mrds does it infact use dht?)_
-
-(LM: do we need to talk about assumptions here?)
-
-(LM: detail what Distance deals with?)
-
-Field methods are chosen to be suitable to species and habitat constraints [@buckland2015distance].
+Figure \ref{fig:pointslines} shows simulated sampling of a population of 500 objects using line and point transects and their corresponding histograms of observed detection distances.
+Note that for the purposes of distance sampling an object may either refer to an individual in a population or a cluster of individuals.
+In the latter instance, the density or abundance of clusters is estimated separately to the mean cluster size which can later be used to estimate the density or abundance of individuals.
 
@@
 
-(LT: What is said about the decreasing number of observations is incorrect, really, and the "taking account of" just makes it confusing, I think.
+\pkg{Distance} provides a selection of candidate functions to describe the probability of detection and estimates the parameters using maximum likelihood estimation.
+The probability of detecting an object may not only depend on how far it is from the obsever but also on other factors such as weather conditions, ground cover, cluster size etc.
+The \pkg{Distance} package also allows the incorporation of such covariates into the detection function allowing the detection function parameters to vary based on these covariates.
+
+Field methods are chosen to be suitable to species and habitat constraints [@buckland2015distance].
+
+The rest of the paper has the following structure: we describe data formatting for \pkg{Distance}; candidate detection function models are described in terms of formulation and fitted examples in \proglang{R}.
+We then show how to perform model checking, goodness of fit and model selection.
+We go on to show how to estimate abundance, including stratified estimates of abundance.
+The final two sections of the article look at extensions (both in terms of methodology and software) and put the package in a broader context amongst other software packages used for estimating the abundance of biological populations.
+(LT: Do we mean software packages in general (e.g., Mark) or R packages?
+Do we actually do this -- i.e., talk about how Distance fits in among all this, or is this now an appendix, and does it cover all methods or just distance-based methods.)
+
+<!-- (LT: What is said about the decreasing number of observations is incorrect, really, and the "taking account of" just makes it confusing, I think.
 Needs a bit more editing - possibly re-order to explain what we see first.) For both points and lines, given the geometry of the sampler (see "Detection functions"), the histogram of distances should show a decreasing number of observations with increasing distance from the sampler (having taken into account sample geometry, see below).
@@
 
-(LT: stranded one-sentence paragraph -- needs integrating.)
-
-The rest of the paper has the following structure: we describe data organisation for use with \pkg{Distance}; models for the detection function are described in terms of formulation and examples of fitting in \proglang{R}.
-We then show how to perform model checking, goodness of fit and model selection.
-We go on to show how to estimate abundance, including stratified estimates of abundance.
-The final two sections of the article look at extensions (both in terms of methodology and software) and put the package in a broader context amongst other software packages used for estimating the abundance of biological populations.
-(LT: Do we mean software packages in general (e.g., Mark) or R packages?
-Do we actually do this -- i.e., talk about how Distance fits in among all this, or is this now an appendix, and does it cover all methods or just distance-based methods.)
+(LT: stranded one-sentence paragraph -- needs integrating.) -->
 

### 2016-01-30 paper.Rmd
Edited up to end of section 3.1

@@
 Given that we can assume these particular availability function there are a number of ways we can use either the estimated detection function or the composite function we fit to our data to estimate abundance.
-One simple idea begins with finding the area under the composite function we fit to our data out to a truncation distance w.
-If we then find the area under the availability function out to distance w then by dividing the former area by the latter gives us an estimate of the propoertion of objects seen within distance w of the transect.
-We can then scale up the number of objects we saw, first to give us an estimate of how many objects were within distance w of the transects and then again by the proportion of the study area we have sampled to give an estimate of abundance for the entire study area.
+One simple idea begins with finding the area under the composite function we fit to our data out to a truncation distance $w$.
+If we then find the area under the availability function out to distance $w$, then dividing the former area by the latter gives us an estimate of the propoertion of objects seen within distance $w$ of the transect.
+We can then scale up the number of objects we saw, first to give us an estimate of how many objects were within distance $w$ of the transects and then again by the proportion of the study area we have sampled to give an estimate of abundance for the entire study area.
 
@@
 Bottom panels show an example detection function (left), the expected avaiability of objects (middle) and the resulting destribution of detections (right) for point transects.
+The distributions in the right hand plots are obtained by multiplying the detection function by the availability function and are referred to in the text as the 'composite function'.
 \\label{fig:detectiondist}"}
@@
 
-Field methods are chosen to be suitable to species and habitat constraints [@buckland2015distance].
+Other important assumptions in conventional distance sampling are as follows: all objects exactly on the transect line or point are detected, objects are stationary, measurements are exact.
+Field methods must therefore be chosen carefully to satisfy these assumptions within the species and habitat constraints [@buckland2015distance].
 
@@
 
-(LT: stranded one-sentence paragraph -- needs integrating.) -->
+(LT: stranded one-sentence paragraph -- needs integrating.)
 
-(LT: Only got this far: will do more later...)
+(LT: Only got this far: will do more later...) -->
 
@@
 * In the simplest case, where the objective is to estimate a detection function, all that is needed is a vector of distances.
-* To include additional covariates into the detection function (see "Detection functions") we use a \code{data.frame}.
-The \code{data.frame} contains a column called \code{distance} (containing the observed distances) and additional named columns for covariates potentially useful in modelling detectability (for example \code{observer} or \code{seastate}).
-Some column names are reserved: \code{object} for an observation identifier (see "Extensions"), \code{size} for group or cluster size (see "Detection functions" and "Abundance and variance estimation"), \code{detected} for whether an observation was detected (see "Extensions") and the columns described in the next bullet.
-* To estimate abundance beyond the area covered by expended survey effort, additional information is required.
-This consists of information about transect and occasion the observation was made (the \code{Sample.Label}), a column named \code{Effort} which gives the effort associated with that sample (for lines their length and for points the number of times that point was visited), the stratum the sample was located in (the \code{Region.Label}, which may be from pre- or post-survey stratification, see "Estimating abundance and variance") and the stratum areas (\code{Area}).
-For transects which were surveyed but have no observations we enter \code{NA} in their \code{distance} column.
+* To include additional covariates into the detection function (see "Detection functions") a \code{data.frame} is required.
+The \code{data.frame} must contain a column named \code{distance} (containing the observed distances) and additional named columns for any covariates that may affect detectability (for example \code{observer} or \code{seastate}).
+The column name \code{size} is reserved for the cluster sizes for when the detection distances relate to observations of clusters rather than individuals.
+Additional reserved names include \code{object} and \code{detected}, these are not required for conventional distance sampling and should be avoided (see "Extensions" for an explanation of their use).
+* To estimate density or to estimate abundance beyond the sampled area, additional information is required.
+Additional columns should be included in the \code{data.frame} specifying: \code{Sample.Label}, the ID of the transect; \code{Effort}, transect effort (for lines their length and for points the number of times that point was visited); \code{Region.Label}, the strata containing the transect (which may be from pre- or post-survey stratification, see "Estimating abundance and variance"); \code{Area}, the area of the strata.
+For transects which were surveyed but have no observations must be included in the data set with \code{NA} for \code{distance} and any other covariates.
 We refer to this data format (where all information is contained in one table) as "flatfile" as it can be easily created in a single spreadsheet.
 
-As we will see in "Extensions", further information is also required when using more complex models.
+As we will see in "Extensions", further information is also required for fitting more complex models.
 
 It is also possible to use distances collected in intervals ("binned" or "grouped" data, as opposed to "exact" distances) to model the detection function.
-In this case a set of cutpoints for the bins must be provided rather than the \code{distance} column, these are specified as two columns \code{distbegin} and \code{distend}.
+In this case the column \code{distance} is replaced by two columns \code{distbegin} and \code{distend} referring to the distance interval start and end cutpoints.
 More information on binned data is included in @Buckland:2001vm sections 4.5 and 7.4.1.2.
@@
 For this reason, we recommended that these are converted to SI units before starting any analysis to ensure that resulting abundance and density estimates have sensible units.
+_(LM: does this mean that the units for distance must be the same for effort, and if these units are km then the area should be in square km?
+If so might be useful to give an e.g.)_
 
@@
 
-The detection function describes the relationship between observed distances and probability of detection.
-The detection function itself models the probability $\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)$ and is usually denoted $g(y; \boldsymbol{\theta})$ where $y$ is distance (from a line or point) and $\boldsymbol{\theta}$ is a vector of parameters to be estimated.
-Our goal is to estimate an *average probability of detection* ($p$, average in the sense of an average over distance from $0$ to $w$), so we must integrate out distance ($y$) from the detection function:
+The detection function models the probability $\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)$ and is usually denoted $g(y; \boldsymbol{\theta})$ where $y$ is distance (from a line or point) and $\boldsymbol{\theta}$ is a vector of parameters to be estimated.
+Our goal is to estimate an *average probability of detection* ($p$, average in the sense of an average over distance from $0$ to truncation distance $w$), so we must integrate out distance ($y$) from the detection function:
 $$
@@
 It is crucial that the detection function accurately models detectability at small distances; we are less worried by its behaviour further away from 0.
-To ensure that the model is not overly influenced by distances far from zero, we truncate the distances beyond a given distance $w$ (known as the *truncation distance*).
+To ensure that the model is not overly influenced by distances far from zero, we truncate the distances beyond a given distance $w$.
 Including these larger distances in our analysis does not demonstrably improve the precision of abundance estimates [@Buckland:2001vm, sections 4.3 and 5.3].
@@
 * *Non-increasing*: we do not think that observers should be more likely to see distant objects than those nearer the transect.
-If this occurs, it usually indicates an issue with field procedure (that the distribution of objects with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this.
+If this occurs, it usually indicates an issue with survey design or field procedures (for example that the distribution of objects with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this.
+_(LM: Hmm could be avilability but could also be detectability...
+either way it is bad news but don't want to give people the impression they should just chop these bumps off without thought do we?
+Ideally spot early in data collection and resolve)_
 * *Model robust*: models should be flexible enough to fit many different shapes.
@@
 $$
-Possible modelling options for key and adjustments are given in Table \ref{tab:keyadj} and illustrated in Figure \ref{fig:hnhr} and \ref{fig:keyadj}.
+Possible modelling options for key and adjustments are given in Table \ref{tab:keyadj} and illustrated in Figures \ref{fig:hnhr} and \ref{fig:keyadj}.
 We select the number of adjustment terms ($K$) by AIC (further details in "Model checking and model selection").
@@
 
-We do not always include adjustments (except in the case of the uniform key), in which case we refer to "key only" models (see the next section and "Model checking and model selection").
+It is not always necessary to include adjustments (except in the case of the uniform key) and in such cases we refer to these as "key only" models (see the next section and "Model checking and model selection").
 

### 2016-01-30 paper.Rmd
Oops now edited up to end 3.1

@@
 There are many factors that can affect the probability of detecting an object.
-These include things like the observer, size of group (if objects occur in groups), the vessel or platform used, the sea state, weather conditions and time of day.
-We assume that these variables affect detection only via the scale of the detection function (and do not affect the shape).
+These include things like observer skill, cluster/group size (if objects occur in groups), the vessel or platform used, sea state, other weather conditions and time of day.
+In \pkg{Distance} we assume that these variables affect detection only via the scale of the detection function (and do not affect the shape).
 
@@
 where $\mathbf{z_i}$ is the vector of $J$ covariates associated with observation $i$.
-For covariate models, we are calculating a value of "average" probability of detection (average in the sense of distance being integrated out) per observation.
-There will be as many unique values of $p(\mathbf{z_i})$ as there are unique covariate combinations in our data.
+For covariate models, we calculate a value of "average" probability of detection (average in the sense of distance being integrated out) per observation.
+There are as many unique values of $p(\mathbf{z_i})$ as there are unique covariate combinations in our data.
 
-K+A models that include covariates and one or more adjustments cannot be guaranteed to be monotonic non-increasing for all covariate combinations, as we do not have any model for the distribution of the covariates.
-Without a distribution for the covariates, it's not possible to know what the behaviour of the detection function will be, so we cannot setup meaningful constraints -- so we cannot constrain the detection function to be monotonic.
+K+A models that include covariates and one or more adjustments cannot be guaranteed to be monotonic non-increasing for all covariate combinations.
+Without a model for the distribution of the covariates, it is not possible to know what the behaviour of the detection function will be across the ranges of the covariates.
+As such we cannot set meaningful constraints on monotonicity.
 For this reason, we advise against using both adjustments and covariates in a detection function [see @Miller:2015hw for an example of when this can be problematic].

### 2016-01-31 paper.Rmd
Edited section 3.2

@@
 
-The workhorse of detection function fitting in \pkg{Distance} is the \code{ds} function.
+Fitting a detection function in \pkg{Distance} is done using the \code{ds} function.
 Here we show some of the possible formulations for the detection function we have seen above applied to the minke whale and amakihi data.
@@
 
-We can fit a model to the minke whale data, setting the truncation at 1.5km and using the default options in \code{ds} very simply:
+First we fit a model to the minke whale data, setting the truncation at 1.5km and using the default options in \code{ds} very simply:
 ```{r minke-hn}
@@
 ```
-Note that \code{ds} will automatically select adjustment terms by AIC and show the selection steps.
+Note that \code{ds} will automatically select adjustment terms using minimum AIC and show the selection steps.
 
@@
 
-We can specify the form of the detection function via the \code{key=} argument to \code{ds}. For example, a hazard rate model can be fitted as:
+A different form for the detection function can be specified via the \code{key=} argument to \code{ds}. For example, a hazard rate model can be fitted as:
 ```{r minke-hr-cos}
@@
 ```
-\code{ds} fits a model with cosine adjustments (the default) but finds the AIC improvement to be insufficient to select the adjustment.
+Here \code{ds} had also fitted a model with cosine adjustments (the default) but finds the AIC improvement to be insufficient to select the adjustment.
 
@@
 
-By default \code{ds} assumes the data given to it has been collected as line transects, but we can switch to point transects using \code{transect="point"}.
+By default \code{ds} assumes the data given to it has been collected as line transects, but we can switch to point transects using the argument \code{transect="point"}.
 We can include covariates in the scale parameter via the \code{formula=~...} argument to \code{ds}.
-A hazard-rate model for the amakihi that includes observer as a covariate can be specified by [truncating sightings at 82.5m, given in @Marques:2007ey]:
+A hazard-rate model for the amakihi that includes observer as a covariate and a truncation distance of 82.5m [@Marques:2007ey] can be specified using :
 ```{r amakihi-hr-obs}
@@
 ```
-Note that here \code{ds} warns us that we have not supplied enough information to estimate abundance, unlike with the minke whale data.
+Note that here, unlike with the minke whale data, \code{ds} warns us that we have only supplied enough information to estimate the detection function (not density or abundance).
 
-As with the minke whale model, we can plot the resulting model (Figure \ref{fig:minkeamakihi}, middle panel).
-Rather than detection functions, Figure \ref{fig:minkeamakihi} shows the probability density functions for the amakihi models.
-Probability density function plots give a better sense of fit of the model than the detection functions, as for point data we must rescale the histogram when plotting the detection function to take into account the geometry of the point sampler.
-The amakihi model included covariates, so the plot shows the detection function averaged over levels/values of the covariate.
-Points on the plot indicate probability of detection for each observation.
-For the `amakihi_hr_obs` model we see fairly clear levels of the observer covariate in the points.
-Looking at the right panel of Figure \ref{fig:minkeamakihi}, we can see this is less clear when adding minutes after sunrise as a covariate into the model:
+ _(LM: this next model was thrown in without much explanation, I have tried to add some intro to it... is this correct?)_
+
+While automatic AIC selection is performed on adjustment terms, model selection for covarariates must be performed manually.
+Here we introduce a second covariate model with both observer and minutes after sunrise explaining detectability.
+We will compare these models further in the following section.
+
 ```{r amakihi-hr-obs-mas, warning=FALSE}
@@
 ```
+
+As with the minke whale model, we can plot the resulting models (Figure \ref{fig:minkeamakihi}, middle and right panels).
+However, for point transect studies, probability density function plots give a better sense of model fit than the detection function plots.
+This is because when plotting the detection function for point transect data, the histogram must be rescaled to account for the geometry of the point sampler (i.e.
+the increasing availability with distance).
+The amakihi models included covariates, so the plots show the detection function averaged over levels/values of the covariate.
+Points on the plot indicate probability of detection for each observation.
+For the `amakihi_hr_obs` model we see fairly clear levels of the observer covariate in the points.
+Looking at the right panel of Figure \ref{fig:minkeamakihi}, we can see this is less clear when adding minutes after sunrise as a covariate into the model.
+
+_(LM: I found the plots a bit confusing to start is it worth putting titles on them to remind the reader that these are from both datasets?
+I couldn't figure out how to make the text bigger if we want to do this.
+I have also clarified the text in the para above as it referred to Figure 6 generally at one point that I think is what threw me off a bit and the fact that the other model hadn't been introduced till after it'a plot was introduced.)_
 
@@
 par(mfrow=c(1, 3))
-plot(minke_hn, showpoints=FALSE)
-plot(amakihi_hr_obs, pdf=TRUE, cex=0.7)
-plot(amakihi_hr_obs_mas, pdf=TRUE, cex=0.7)
+plot(minke_hn, showpoints=FALSE, main = "Minke~distance", cex = 2)
+plot(amakihi_hr_obs, pdf=TRUE, cex=0.7, main = "Amakihi~distance+observer")
+plot(amakihi_hr_obs_mas, pdf=TRUE, cex=0.7, main = "Amakihi~distance+observer+minutes")
 ```

### 2016-02-01 paper.Rmd
Edits to Section 4

@@
 As with models fitted using \code{lm} or \code{glm} in \proglang{R}, we can use \code{summary} to give useful information about our fitted model.
-For our hazard-rate model for the amakihi, with observer as a covariate:
+For our hazard-rate model of the amakihi data, with observer as a covariate:
 ```{r amakihi-summary}
@@
 
-To judge goodness of fit for detection functions when exact distances are used, we want to compare the cumulative distribution function (CDF) and empirical distribution function (EDF) for the detection function via a quantile-quantile plot (Q-Q plot).
-The CDF evaluates the probability of observing an object at a distance less than or equal to some value.
-The EDF gives the proportion of observations for which the CDF is less than or equal to that of a given distance.
-This can be interpreted as assessing whether the number of observations up to a given distance is in line with what the model says they should be (where the "given values" are the observed distances).
+To judge how well a detection functions fits the data when we have exact distances, we use a quantile-quantile plot (Q-Q plot).
+Here the Q-Q plot compares the the cumulative distribution function (CDF) of the fitted detection function to the distribution of the data (empirical distribution function or EDF).
+The Q-Q plots in \pkg{Distance} plot a point for every observation.
+The EDF is is the proportion of points that have been observed at a distance equal to or less than the distance of that observation.
+The CFD is calculated from the fitted detection function as the probability of observing an object at a distance less than or equal to that of the given observation.
+This can be interpreted as assessing whether the number of observations up to a given distance is in line with what the model says they should be.
 As usual for Q-Q plots, "good" models will have values close to the line $y=x$, poor models will show greater deviations from that line.
 
-We can inspect Q-Q plots visually, though this is prone to subjective judgments.
-Instead we can quantify the Q-Q plot's information using a Kolmogorov-Smirnov or Cramer-von Mises test [@Burnham:2004vd].
+Q-Q plots can be inspected visually, though this is prone to subjective judgments.
+Therefore, we also quantify the Q-Q plot's information using a Kolmogorov-Smirnov or Cramer-von Mises test [@Burnham:2004vd].
 Both test whether points from the EDF and CDF are from the same distribution.
@@
 As it takes into account more information and is therefore more powerful, the Cramer-von Mises is generally preferred.
-A significant result from either test indicates that the EDF and CDF do not come from the same distribution (and therefore the model is does not fit the data well).
+A significant result (usually $P < 0.05$) from either test gives evidence against the null hypothesis (that the data arose from the fitted model), suggesting that the model does not fit the data well.
 
@@
 ```
-So we can therefore conclude that the half-normal model does not pass our goodness of fit tests and should be discarded.
+We can conclude that the half-normal model does not pass our goodness of fit tests and should be discarded.
+Both hazard-rate models (output only shown for hazard-rate model with observer and minutes after sunrise) had non-significant goodness of fit test statistics and are both therefore deemed plausible models.
 The corresponding Q-Q plots are shown in Figure \ref{amakihi-qq}, comparing the half-normal model with the hazard-rate model with observer and minutes after sunrise included.
@@
 
-Once we have a set of models which fit well, we can use Akaike's Information Criterion (AIC) to select between models.
+Once we have a set of plausible models, we can use Akaike's Information Criterion (AIC) to select between models.
 \pkg{Distance} includes a function to create table of summary information for fitted models, making it easy to get an overview of a large number of models.
@@
 An example of this output is shown in Table \ref{tab:amakihi} and was generated by the following call to \code{summarize_ds_models}:
+
+_(LM: hmm we say that the hn model should be discarded based on fit but it is then included in the table...
+is this confusing?
+There is also no guidance here on model selection...
+although the AIC is smallest for amakihi.hr.obs.mas, some would argue that a difference of less than 2 does not really justify the inclusion of the additional covariate.
+Are we simply saying minimum AIC wins (if so this is not clear)?
+Should we point them at further discussion on the topic?)_
 ```{r summary-table-null, eval=FALSE}

### 2016-02-01 paper.Rmd
Section 5 edits

@@
 
-Though fitting the detection function is the primary modelling step in distance sampling, we are really interested in estimating detectability, and from that abundance.
-We also wish to calculate our uncertainty for each abundance estimate.
+Though fitting the detection function is the primary modelling step in distance sampling, we are really interested in estimating detectability, and from that density or abundance.
+It is also important to calculate our uncertainty for these estimates.
 This section addresses these issues mathematically before showing how to estimate abundance and its variance in \proglang{R}.
@@
 We wish to obtain the abundance in a study region, of which we have sampled a (representative) subset.
-To do this we first calculate the abundance in the area we have surveyed (the *covered area*) to obtain $\hat{N}_\text{C}$, we can then scale this up (based on the survey design) to the full study area by multiplying it by the ratio of covered area to study area.
+To do this we first calculate the abundance in the area we have surveyed (the *covered area*) to obtain $\hat{N}_\text{C}$, we can then scale this up to the full study area by multiplying it by the ratio of covered area to study area.
 We discuss other methods for spatially explicit abundance estimation in "Extensions".
@@
 
-As an example the minke whale data consists of two strata: \code{North} and \code{South} relating to a stratum further away and nearer the Antarctic ice edge, respectively.
+As an example, the minke whale data contain two strata: \code{North} and \code{South} relating to strata further away from and nearer to the Antarctic ice edge, respectively.
 Figure \ref{fig:minke-strata} shows the two strata, along with observations and transect lines.
@@
 
-Here we take an intuitive approach to uncertainty estimation, for a full derivations consult @Marques:2003vb. Uncertainty in $\hat{N}$ comes from two sources:
+We take an intuitive approach to uncertainty estimation, for a full derivation consult @Marques:2003vb. Uncertainty in $\hat{N}$ comes from two sources:
 
-1. *Model parameter uncertainty*, from the estimation of the detection function parameters $\boldsymbol{\theta}$.
-2. *Sampling uncertainty*, from the distribution of objects along the transect lines or between visiting occasions for points.
+1. *Detection function*: Uncertainty in parameter ($\boldsymbol{\theta}$) estimation.
+2. *Encounter rate*: Sampling variability arising from differences in the rate of observations between transects.
 
@@
 
-Sampling uncertainty can be characterised by the *encounter rate*: the number of objects per unit transect (rather than just $n$).
+*Encounter rate* is the number of objects per unit transect (rather than just $n$).
 When covariates are not included in the detection function we can define the encounter rate as $n/L$ for line transects (where $L$ is the total line length) or $n/T$ for point transects (where $T$ is the total number of visits summed over all points).
@@
 where $l_k$ are the lengths of the $K$ transects (such that $L = \sum_{k=1}^K l_k$) and $\hat{N_{C,k}}$ is the abundance in the covered area for transect $k$.
-Whereas for points we use estimator "P3" from @Fewster:2009ku but again replacing $n$ by $\hat{N}_\text{C}$ in the encounter rate definition, we obtain the following estimator:
+For point transects we use the estimator "P3" from @Fewster:2009ku but again replace $n$ by $\hat{N}_\text{C}$ in the encounter rate definition, to obtain the following estimator:
 $$
@@
 The summary can be more concisely expressed by extracting information from the summary object.
-This object is a \code{list} of \code{data.frames}, so we can again use the \code{kable} function from \pkg{knitr} to create summary tables of abundance estimates and measures of precision, such as Table \ref{minke-abund}.
+This object is a \code{list} of \code{data.frames}, so we can use the \code{kable} function from \pkg{knitr} to create summary tables of abundance estimates and measures of precision, such as Table \ref{minke-abund}.
 We prepare the \code{data.frame} as follows before using \code{kable}:
@@
 
+<!-- LM The location of this Table needs changing it appears in the middle of the output. -->
+
 # Extensions
@@
 
-The abundance estimates calculated here are based on the assumption that within a given stratum density is uniform.
+The abundance estimates calculated here are based on the assumption that within a given stratum density is uniform _(LM: previous sentence incorrect)_.
 We may extend this approach to many strata, making the area of each very small to account for small-scale variation in space.

### 2016-02-01 paper.Rmd
Edits to Section 6

@@
 The features of \pkg{Distance} are deliberately limited to provide a simplified interface for users.
-For more complex analyses of distance sampling data, there are further related packages for modelling in \proglang{R}.
+For more complex analyses of distance sampling data, we provide additional packages for modelling in \proglang{R}.
 
-We noted at the start of the article that \pkg{Distance} is a simpler wrapper around the package \pkg{mrds}.
-Additional features are available in \pkg{mrds} including the ability to model data where the assumption that detection is certain at zero distance from the line or point is violated, using mark-recapture type methods for double observer surveys [see @Burt:2014gu for an introduction].
+We noted at the start of the article that \pkg{Distance} is a simple-to-use wrapper around the package \pkg{mrds}.
+Additional features available in \pkg{mrds} include  models that relax the assumption that detection is certain at zero distance from the transect.
+This is done using mark-recapture type methods which require a different survey technique, known as double observer surveys [see @Burt:2014gu for an introduction].
 
-The abundance estimates calculated here are based on the assumption that within a given stratum density is uniform _(LM: previous sentence incorrect)_.
-We may extend this approach to many strata, making the area of each very small to account for small-scale variation in space.
-A more rigorous approach is to build a spatial model incorporating spatially-referenced environmental data (for example derived from GIS products).
-\pkg{Distance} interfaces with one such package to perform this type of analysis: \pkg{dsm} [@dsm-pkg].
+\pkg{Distance} can provide us with estimates of abundance or density for each strata as a whole but tells us nothing about the distribution of animals within strata.
+One option is to divide the study area into smaller and smaller strata to try to detect patterns in spatial distribution, however, a more rigorous approach is to build a spatial model.
+Such models incorporate spatially-referenced environmental data (for example derived from GIS products).
+\pkg{Distance} interfaces with one such package used to perform this type of analysis: \pkg{dsm} [@dsm-pkg].
 So-called "density surface modelling" uses the generalized additive model framework [e.g.
@@
 In addition, \pkg{mads} can also incorporate sightings for which species identification could not be achieved.
-This is done by estimating the abundance of these 'unidentified' sightings and pro-rating them as described in [@Gerrodette:2005].
+This is done by estimating the abundance of these 'unidentified' sightings and pro-rating them to the known species in the analysis as described in [@Gerrodette:2005].
 
-In addition to the analyses, an equally if not more important stage of a distance sampling survey is the survey design.
+In addition to the analysis, an equally important stage of a distance sampling survey is the survey design.
 We have therefore developed a package which allows users to test out different designs in their specific study region and tailor population attributes to reflect the species they are working with.
-In this way \pkg{DSsim} [@DSsim-pkg] allows users to more easily identify challenges unique to their study and select a survey design which is likely to yield the most accurate and precise results.
+In this way \pkg{DSsim} [@DSsim-pkg] allows users to more easily identify challenges unique to their study and select a survey design which is likely to yield the most accurate and precise estimates.
 
@@
 For that reason we have created the \pkg{readdst} [@readdst-pkg] package to interface with projects created by Distance for Windows.
-The package can take analyses created using the CDS, MCDS and MRDS engines in Distance for Windows and extract data and create equivalent models in \proglang{R}.
+The package can take analyses created using the CDS, MCDS and MRDS engines in Distance for Windows, extract the data and create equivalent models in \proglang{R}.
 \pkg{readdst} can also run these analyses and test the resulting statistics (for example, $\hat{N}$ or $\hat{P_a}$) calculated in \proglang{R} against those calculated by Distance for Windows.

### 2016-02-01 paper.Rmd
Edits to Section 7 and Acknowledgements

@@
 In combination with tools such as \pkg{knitr} and \pkg{rmarkdown} [@rmarkdown-pkg], the helper functions in \pkg{Distance} provide a useful set of tools to perform reproducible analyses of wildlife abundance for both managers and ecologists.
-We hope that this paper provides useful examples for those wishing to pursue this.
+In addition, users should remember the other basic functions in \proglang{R} such as \code{hist} that enable straight forward exploratory data analysis.
+We hope that this paper provides useful examples for those wishing to pursue distance sampling in \proglang{R}.
 More information on distance sampling can be found at \url{http://distancesampling.org} and a mailing list is maintained at \url{https://groups.google.com/forum/#!forum/distance-sampling}.
 
-We note that there are other packages available for performing distance sampling analyses in \proglang{R} but believe that \pkg{Distance} is the most flexible and feature-complete.
+We note that there are other packages available for performing distance sampling analyses in \proglang{R} but believe that \pkg{Distance} is the most flexible and feature-complete, and provides pathways to a range of more complicated analyses.
 Appendix A gives a feature comparison between \pkg{Distance} and other \proglang{R} packages for analysis of distance sampling data.
@@
 
-The authors would like to thank the many users of \pkg{Distance}, \pkg{mrds} and DISTANCE who have contributed bug reports and suggestions for improvements over the years.
+The authors would like to thank the many users of \pkg{Distance}, \pkg{mrds} and Distance for Windows who have contributed bug reports and suggestions for improvements over the years.
 We would particularly like to thank Steve Buckland, David Borchers, Tiago Marques, Jon Bishop and Lorenzo Milazzo for their contributions.

### 2016-02-01 paper.Rmd
A few more edits... now I'm finished
I have left a few comments in, they appear in italics in the pdf and
inside _(LM … )_ in the .Rmd so are pink. I have commented out Len’s
comments as I have tried to address them but they are still in the .Rmd
file in <!— —> and appear in blue. Actually colours may vary depending
on your personal setup.

@@
 Like the histogram function, the most basic detection function estimation only requires a numeric vector of distances.
-Here we demonstrate how to use \pkg{Distance} to fit detection functions, perform model checking and selection, and estimate abundance and/or density.
+Here we demonstrate how to use \pkg{Distance} to fit detection functions, perform model checking and selection, and estimate abundance.
 
@@
 
-Given that we can assume these particular availability function there are a number of ways we can use either the estimated detection function or the composite function we fit to our data to estimate abundance.
-One simple idea begins with finding the area under the composite function we fit to our data out to a truncation distance $w$.
-If we then find the area under the availability function out to distance $w$, then dividing the former area by the latter gives us an estimate of the propoertion of objects seen within distance $w$ of the transect.
+Given that we can assume these availability functions there are a number of ways we can use either the estimated detection function or the composite function we fit to our data to estimate abundance.
+One simple idea begins with finding the area under the composite function out to a truncation distance $w$.
+If we then find the area under the availability function out to distance $w$, then dividing the former area by the latter gives us an estimate of the proportion of objects seen within distance $w$ of the transect.
 We can then scale up the number of objects we saw, first to give us an estimate of how many objects were within distance $w$ of the transects and then again by the proportion of the study area we have sampled to give an estimate of abundance for the entire study area.
 
-```{r detectdists, fig.height=5.5, fig.width=8, echo=FALSE, message=FALSE, fig.cap="Top panels show an example detection function (left), the expected avaiability of objects (middle) and the resulting destribution of detections (right) for line transects.
-Bottom panels show an example detection function (left), the expected avaiability of objects (middle) and the resulting destribution of detections (right) for point transects.
-The distributions in the right hand plots are obtained by multiplying the detection function by the availability function and are referred to in the text as the 'composite function'.
+```{r detectdists, fig.height=5.5, fig.width=8, echo=FALSE, message=FALSE, fig.cap="Top panels show an example detection function (left), the expected availability of objects (middle) and the resulting distribution of detections (right) for line transects.
+Bottom panels show an example detection function (left), the expected availability of objects (middle) and the resulting distribution of detections (right) for point transects.
+The distributions in the right hand plots are obtained by multiplying the detection function by the availability function and are referred to in the text as the `composite function'.
 \\label{fig:detectiondist}"}
@@
 Note that for the purposes of distance sampling an object may either refer to an individual in a population or a cluster of individuals.
-In the latter instance, the density or abundance of clusters is estimated separately to the mean cluster size which can later be used to estimate the density or abundance of individuals.
 
@@
 
-\pkg{Distance} provides a selection of candidate functions to describe the probability of detection and estimates the parameters using maximum likelihood estimation.
+\pkg{Distance} provides a selection of candidate functions to describe the probability of detection and estimates the associated parameters using maximum likelihood estimation.
 The probability of detecting an object may not only depend on how far it is from the obsever but also on other factors such as weather conditions, ground cover, cluster size etc.
-The \pkg{Distance} package also allows the incorporation of such covariates into the detection function allowing the detection function parameters to vary based on these covariates.
+The \pkg{Distance} package also allows the incorporation of such covariates into the detection function allowing the detection function scale parameter to vary based on these covariates.
 
-Other important assumptions in conventional distance sampling are as follows: all objects exactly on the transect line or point are detected, objects are stationary, measurements are exact.
-Field methods must therefore be chosen carefully to satisfy these assumptions within the species and habitat constraints [@buckland2015distance].
+In addition to randomly placed samplers distance sampling relies on three other main assumptions.
+Firstly, all objects directly on the transect line or point (i.e.
+those at zero distance) are detected (see "Extensions" for methods to deal with the situation when this is not possible).
+Secondly, objects are stationary or detected prior to any movement.
+Thirdly, distance to the object must be measured accurately, or the observation allocated to the correct distance bin for grouped data.
+Depending on your survey species some of these assumptions may be more difficult to meet than others.
+Further information on field methods to help meet these assumptions can be found in [@buckland2015distance].
 
@@
 We go on to show how to estimate abundance, including stratified estimates of abundance.
-The final two sections of the article look at extensions (both in terms of methodology and software) and put the package in a broader context amongst other software packages used for estimating the abundance of biological populations.
-(LT: Do we mean software packages in general (e.g., Mark) or R packages?
-Do we actually do this -- i.e., talk about how Distance fits in among all this, or is this now an appendix, and does it cover all methods or just distance-based methods.)
+The final two sections of the article look at extensions (both in terms of methodology and software) and put the package in a broader context amongst other \proglang{R} packages used for estimating the abundance of biological populations.
+<!-- (LT: Do we mean software packages in general (e.g., Mark) or R packages?
+Do we actually do this -- i.e., talk about how Distance fits in among all this, or is this now an appendix, and does it cover all methods or just distance-based methods.) LM: changed this to R packages) -->
 

### 2016-02-09 paper.Rmd
first run through after Laura's edits

@@
 
-  The study of plants or animals often involves the estimation of population size and/or their spatial distribution.
-These characteristics are important considerations in wildlife management.
-We introduce the \proglang{R} package \pkg{Distance} that uses distance sampling data to estimate populations size (abundance) and is compatible with another \proglang{R} package \pkg{dsm} which can be used to model spatial distribution.
-We describe how users can obtain estimates of abundance and/or density using the package \pkg{Distance} as well documenting the links it provides with other \proglang{R} packages offering specialist solutions to specific distance sampling challenges.
-We demonstrate how this package provides a migration pathway from previous software, thereby allowing us to deliver cutting edge methods to the users more quickly.
+  Estimating the abundance and spatial distribution of animal and plant populations is essential for conservation and management.
+We introduce the \proglang{R} package \pkg{Distance} that uses distance sampling methods to estimate abundance.
+We describe how users can obtain estimates of abundance (and density) using the package as well documenting the links it provides with other more specialized \proglang{R} packages.
+We also demonstrate how \pkg{Distance} provides a migration pathway from previous software, thereby allowing us to deliver cutting edge methods to the users more quickly.
 
@@
 (LM: The paper on Distance in JAppEcol?
-There is quite a big intro on distance sampling there.))) -->
+There is quite a big intro on distance sampling there.)))
+
+DLM: if you look at the unmarked paper they do a fair bit of explanation of the methods.-->
 
@@
 
-Conventional distance sampling assumes the observer is either stood at a point or moving along a line and will only see everything that occurs exactly at that point or on the line.
-It is then expected that the further away an object is from the point or line (also known as the sampler or transect) the less likely it is that the observer will see it.
-By recording the distance to each of the detected objects we can fit a model to describe the probability of detection given distance from the sampler, which we refer to as the ‘detection function’.
+Conventional distance sampling assumes the observer is either stood at a point or moving along a line and will observe all objects (animals, plants, golf tees etc) that occur at the point or on the line.
+The further away an object is from the point or line (also known as the sampler or transect) the less likely it is that the observer will see it.
+We can use the distances to each of the detected objects from the line or point to build a model of the probability of detection given distance from the sampler -- the *detection function*.
 The detection function can be used to infer how many objects were missed and thereby produce estimates of density and/or abundance.
 
-The Windows program Distance (or "DISTANCE"; for clarity henceforth "Distance for Windows") can be used to fit detection functions to such distance sampling data.
+The Windows program Distance (or "DISTANCE"; for clarity henceforth "Distance for Windows") can be used to fit detection functions to distance sampling data.
 It was first released in November 1998 as Distance for Windows version 3.5.
 Since this time it has evolved to include various design and analysis features [@Thomas:2010cf].
-Distance for Windows versions 5 onwards have included \proglang{R} [@rcore]  packages as the analysis engines providing additional, more complex analysis options than those offered by the original FORTRAN engine.
+Distance for Windows versions 5 onwards have included \proglang{R} [@rcore]  packages as the analysis engines providing additional, more complex analysis options than those offered by the original (FORTRAN) code.
 
-As Distance for Windows becomes increasingly reliant on analyses performed in \proglang{R} and new methods are being developed at a rate where it is unfeasible to make all these available in the Windows interface, we are encouraging the use of our \proglang{R} packages directly.
-In addition, the core packages in \proglang{R} provide a variety of functions useful for data exploration with minimal data formatting.
-The plotting of a histogram of detection distances only requires these values to be stored in a numerical vector, an exercise that is recommended not only prior to analyses but also early on in the data collection process _(LM need ref here @buckland2015distance ?)_.
+As Distance for Windows becomes increasingly reliant on analyses performed in \proglang{R} and many new methods are being developed, we are encouraging the use of our \proglang{R} packages directly.
+\proglang{R} provides a huge variety of functionality for data exploration and reproducable research, much more than is possible in Distance for Windows.
 
-Until now those wishing to use our \proglang{R} packages for straight forward distance sampling analyses would have had to negotiate the complex package \pkg{mrds} [@mrds-pkg] designed for mark-recapture distance sampling.
-In addition, \pkg{mrds} requires a complex data structure before analyses can be completed.
+Until now those wishing to use our \proglang{R} packages for straight forward distance sampling analyses would have had to negotiate the package \pkg{mrds} [@mrds-pkg] designed for mark-recapture distance sampling [@Burt:2014gu], requiring a complex data structure to perform analyses.
 \pkg{Distance} is a wrapper package around \pkg{mrds} making it easier to get started with basic distance sampling analyses in \proglang{R}.
@@
 
-<!-- While plot sampling techniques are adequate for static populations occuring at high density they are inefficient for more sparsely distributed populations.
-Distance sampling provides a more efficient solution in such circumstances.
-In addition, for cryptic species it simply may not be possible to detect all individuals within a given plot and we must find a way to account for imperfect detectability when estimating abundance [@LahozMonfort:2013bg].
-This paper does not provide a detailed explanation of the theory behind distance sampling but instead concentrates on it's application using the package \pkg{Distance}.
-Further information on distance sampling, including the theory, various extensions to deal with more complex situations and recent methodological developments can be found at http://distancesampling.org/dbib.html -->
+The distribution of the observed distances is a product of detectability (sometimes referred to as "perception bias", @Marsh:1989ho) and the distribution of the animals (observed or unobserved) with respect to the line or point.
+We hope that our survey design will ensure that we can assume a distribution for the animals.
 
-Like plot sampling, distance sampling assumes that the samplers have been located at random within the study area.
-Given that the samplers are located independently of the objects, for line transect studies we know that the "availability" of objects should be constant with increasing perpendicular distance.
-However, for point transect surveys the available survey area increases linearly with radial distance implying that the number of objects available for detection should also increase linerary.
-Figure \ref{fig:detectiondist} shows how these availability functions when combined with a detection function give rise to the observed distribution of recorded distances.
-
-Given that we can assume these availability functions there are a number of ways we can use either the estimated detection function or the composite function we fit to our data to estimate abundance.
-One simple idea begins with finding the area under the composite function out to a truncation distance $w$.
-If we then find the area under the availability function out to distance $w$, then dividing the former area by the latter gives us an estimate of the proportion of objects seen within distance $w$ of the transect.
-We can then scale up the number of objects we saw, first to give us an estimate of how many objects were within distance $w$ of the transects and then again by the proportion of the study area we have sampled to give an estimate of abundance for the entire study area.
+Given that the lines or points are located independently of the objects, for line transect studies we assume that objects are uniformly distributed with respect to the line (i.e.
+the number of animals available for detection is constant).
+For point transect surveys the survey area increases linearly with radial distance, implying a triangular distribution with respect to the point.
+Figure \ref{fig:detectiondist} shows how these distributions, when combined with a detection function, give rise to the observed distribution of recorded distances.
 
@@
 
+Having estimated the detection function's parameters, one can then integrate out distance from the function (as the detection function describes the probability of detection _given_ distance) to get an "average" probability of detection, which can be used to correct the observed counts.
+Summing the corrected counts gives an estimate of abundance in the area covered by surveys, which can be multiplied up to the total study area.
+
 In addition to randomly placed samplers distance sampling relies on three other main assumptions.
@@
 Thirdly, distance to the object must be measured accurately, or the observation allocated to the correct distance bin for grouped data.
-Depending on your survey species some of these assumptions may be more difficult to meet than others.
+Depending on the survey species some of these assumptions may be more difficult to meet than others.
 Further information on field methods to help meet these assumptions can be found in [@buckland2015distance].
@@
 We go on to show how to estimate abundance, including stratified estimates of abundance.
-The final two sections of the article look at extensions (both in terms of methodology and software) and put the package in a broader context amongst other \proglang{R} packages used for estimating the abundance of biological populations.
-<!-- (LT: Do we mean software packages in general (e.g., Mark) or R packages?
-Do we actually do this -- i.e., talk about how Distance fits in among all this, or is this now an appendix, and does it cover all methods or just distance-based methods.) LM: changed this to R packages) -->
-
-<!-- (LT: What is said about the decreasing number of observations is incorrect, really, and the "taking account of" just makes it confusing, I think.
-Needs a bit more editing - possibly re-order to explain what we see first.) For both points and lines, given the geometry of the sampler (see "Detection functions"), the histogram of distances should show a decreasing number of observations with increasing distance from the sampler (having taken into account sample geometry, see below).
-For line transects we expect objects to be distributed uniformly with respect to distance from the line (assuming that lines and points are placed independently of animal distribution) with detectability causing the number of detections to decrease (as distance from the sampler increases).
-For point transects we recognise that as distance from the point increases, the area of the circle with a corresponding radius increases with distance squared, hence the number of objects available to be detected is a linearly increasing function of distance from the point.
-
-(LT: A detection function is not a line; it is also scaled differently.
-This needs editing.) Plotting a histogram of observed distances, we can crudely estimate the drop-off in detectability by eye if we trace a line that approximates the tops of the histogram bars: this is a detection function.
-\pkg{Distance} estimates the parameters for a fixed-form detection function using maximum likelihood estimation.
-We address possible models below.
-
-(LT: stranded one-sentence paragraph -- needs integrating.)
-
-(LT: Only got this far: will do more later...) -->
+The final two sections of the article look at extensions (both in terms of methodology and software) and put the package in a broader context amongst other \proglang{R} packages used for estimating the abundance of biological populations from distance sampling data.
 
@@
 
-The line transect data have been simulated from models fitted to Antarctic minke whale (*Balaenoptera bonaerensis*) data (_LM why is the Minke whale data simulated rather than ust using original data, just data ownership/permissions?_).
+The line transect data have been simulated from models fitted to Antarctic minke whale (*Balaenoptera bonaerensis*) data.
 These data were collected as part of the International Whaling Commission's International Decade of Cetacean Research Southern Ocean Whale and Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer surveys [@Branch:2001ua].
-This data set consists of 99 observations that are stratified based on location (near or distant from ice edge) and effort data (transect lengths)(LM did the original dataset have 99 observations or the simulated dataset or both?).
+This data set consists of 99 observations that are stratified based on location (near or distant from ice edge) and effort data (transect lengths).
 The survey is shown in Figure \ref{fig:minke-strata}.
-_(LM wonder if we need more info on how the data were simulated...
-based only on detection function or on spatial model too?
-I notice in the figure text that the simulated data are based on ``1992/93 Area III'' should this info go here too?)_
 
@@
 
-The two example data sets used here are distributed preformatted with \pkg{Distance} so readers can reproduce our analyses.
-Generally, data collected in the field will require some formatting for use with \pkg{Distance}.
-However, the package allows for a flexible format of data input ranging from very simple to complex dependent on the type of analysis you wish to perform:
+The two example data sets used here are distributed with \pkg{Distance} so readers can reproduce our analyses.
+Generally, data collected in the field will require some formatting before use with \pkg{Distance}, though there are a range of possible formats, dependent on the model specification and the output required:
 
-* In the simplest case, where the objective is to estimate a detection function, all that is needed is a vector of distances.
+* In the simplest case, where the objective is to estimate a detection function and exact distances are collected, all that is required is a numeric vector of distances.
 * To include additional covariates into the detection function (see "Detection functions") a \code{data.frame} is required.
 The \code{data.frame} must contain a column named \code{distance} (containing the observed distances) and additional named columns for any covariates that may affect detectability (for example \code{observer} or \code{seastate}).
-The column name \code{size} is reserved for the cluster sizes for when the detection distances relate to observations of clusters rather than individuals.
+The column name \code{size} is reserved for the cluster sizes (sometimes referred to as group sizes) for when the detection distances relate to observations of clusters rather than individuals.
 Additional reserved names include \code{object} and \code{detected}, these are not required for conventional distance sampling and should be avoided (see "Extensions" for an explanation of their use).
 * To estimate density or to estimate abundance beyond the sampled area, additional information is required.
-Additional columns should be included in the \code{data.frame} specifying: \code{Sample.Label}, the ID of the transect; \code{Effort}, transect effort (for lines their length and for points the number of times that point was visited); \code{Region.Label}, the strata containing the transect (which may be from pre- or post-survey stratification, see "Estimating abundance and variance"); \code{Area}, the area of the strata.
-For transects which were surveyed but have no observations must be included in the data set with \code{NA} for \code{distance} and any other covariates.
+Additional columns should be included in the \code{data.frame} specifying: \code{Sample.Label}, the ID of the transect; \code{Effort}, transect effort (for lines their length and for points the number of times that point was visited); \code{Region.Label}, the stratum containing the transect (which may be from pre- or post-survey stratification, see "Estimating abundance and variance"); \code{Area}, the area of the strata.
+Transects which were surveyed but have no observations must be included in the data set with \code{NA} for \code{distance} and any other covariates.
 We refer to this data format (where all information is contained in one table) as "flatfile" as it can be easily created in a single spreadsheet.
@@
 
-The columns \code{distance}, \code{Area} and (in the case of line transects) \code{Effort} have associated units (though these are not explicitly included in a \pkg{Distance} analysis.
-For this reason, we recommended that these are converted to SI units before starting any analysis to ensure that resulting abundance and density estimates have sensible units.
-_(LM: does this mean that the units for distance must be the same for effort, and if these units are km then the area should be in square km?
-If so might be useful to give an e.g.)_
+The columns \code{distance}, \code{Area} and (in the case of line transects) \code{Effort} have associated units (though these are not explicitly included in a \pkg{Distance} analysis).
+We recommended that data in these columns are converted to SI units before starting any analysis to ensure that resulting abundance and density estimates have sensible units.
+For example, if distances from a line transect survey are recorded in metres, the \code{Effort} columns should contain line lengths also in metres and the \code{Area} column gives the stratum areas in square metres.
+This would lead to density estimates of animals per square metre.
 
@@
 
-We will explore the consequences of including effort and stratum data during analysis below.
+We will explore the consequences of including effort and stratum data in the analysis below.
 
@@
 $$
-where $\pi(y)$ describes the distribution of objects with respect to the sampler; $\pi(x)=1/w$ for line transects and $\pi(r)=\frac{2r}{w^2}$ for point transects, taking into account the geometry of the sampler [@Buckland:2001vm, Chapter 3].
+where $\pi(y)$ describes the distribution of objects with respect to the sampler; $\pi(x)=1/w$ for line transects and $\pi(r)=\frac{2r}{w^2}$ for point transects, taking into account the geometry of the sampler [usually referred to as the _probability density function of distances_; @Buckland:2001vm, Chapter 3].
 When considering a particular transect type we let $x$ denote a perpendicular distance from a line and $r$ denote radial distance from a point (rather than using $y$).
@@
 * *Non-increasing*: we do not think that observers should be more likely to see distant objects than those nearer the transect.
-If this occurs, it usually indicates an issue with survey design or field procedures (for example that the distribution of objects with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this.
-_(LM: Hmm could be avilability but could also be detectability...
-either way it is bad news but don't want to give people the impression they should just chop these bumps off without thought do we?
-Ideally spot early in data collection and resolve)_
+If this occurs, it usually indicates an issue with survey design or field procedures (for example that the distribution of objects with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this [see @Miller:2015hw, @Marques:2010he and @Marques:2012fy for more on this topic].
 * *Model robust*: models should be flexible enough to fit many different shapes.
@@
 
-There are many factors that can affect the probability of detecting an object.
-These include things like observer skill, cluster/group size (if objects occur in groups), the vessel or platform used, sea state, other weather conditions and time of day.
+There are many factors that can affect the probability of detecting an object: observer skill, cluster/group size (if objects occur in groups), the vessel or platform used, sea state, other weather conditions, time of day and more.
 In \pkg{Distance} we assume that these variables affect detection only via the scale of the detection function (and do not affect the shape).
@@
 
-Fitting a detection function in \pkg{Distance} is done using the \code{ds} function.
-Here we show some of the possible formulations for the detection function we have seen above applied to the minke whale and amakihi data.
+A detection function can be fitted in \pkg{Distance} using the \code{ds} function.
+Here we apply some of the possible formulations for the detection function we have seen above to the minke whale and amakihi data.
 
@@
 ```
-Note that \code{ds} will automatically select adjustment terms using minimum AIC and show the selection steps.
+Note that when there are no covariates in the model, \code{ds} will add adjustment terms to the model until there is no improvements in AIC and print the selection steps.
 
@@
 
-By default \code{ds} assumes the data given to it has been collected as line transects, but we can switch to point transects using the argument \code{transect="point"}.
+\code{ds} assumes the data given to it has been collected as line transects, but we can switch to point transects using the argument \code{transect="point"}.
 We can include covariates in the scale parameter via the \code{formula=~...} argument to \code{ds}.
@@
 
- _(LM: this next model was thrown in without much explanation, I have tried to add some intro to it... is this correct?)_
-
 While automatic AIC selection is performed on adjustment terms, model selection for covarariates must be performed manually.
-Here we introduce a second covariate model with both observer and minutes after sunrise explaining detectability.
-We will compare these models further in the following section.
+Here we add a second covariate: minutes after sunrrise.
+We will compare these two models further in the following section.
 
@@
 However, for point transect studies, probability density function plots give a better sense of model fit than the detection function plots.
-This is because when plotting the detection function for point transect data, the histogram must be rescaled to account for the geometry of the point sampler (i.e.
-the increasing availability with distance).
+This is because when plotting the detection function for point transect data, the histogram must be rescaled to account for the geometry of the point sampler.
 The amakihi models included covariates, so the plots show the detection function averaged over levels/values of the covariate.
@@
 For the `amakihi_hr_obs` model we see fairly clear levels of the observer covariate in the points.
-Looking at the right panel of Figure \ref{fig:minkeamakihi}, we can see this is less clear when adding minutes after sunrise as a covariate into the model.
-
-_(LM: I found the plots a bit confusing to start is it worth putting titles on them to remind the reader that these are from both datasets?
-I couldn't figure out how to make the text bigger if we want to do this.
-I have also clarified the text in the para above as it referred to Figure 6 generally at one point that I think is what threw me off a bit and the fact that the other model hadn't been introduced till after it'a plot was introduced.)_
+Looking at the right panel of Figure \ref{fig:minkeamakihi}, this is less clear when adding minutes after sunrise to the model.
 
@@
 par(mfrow=c(1, 3))
-plot(minke_hn, showpoints=FALSE, main = "Minke~distance", cex = 2)
-plot(amakihi_hr_obs, pdf=TRUE, cex=0.7, main = "Amakihi~distance+observer")
-plot(amakihi_hr_obs_mas, pdf=TRUE, cex=0.7, main = "Amakihi~distance+observer+minutes")
+plot(minke_hn, showpoints=FALSE)
+title("Minke whales", cex.main=1.5)
+plot(amakihi_hr_obs, pdf=TRUE, cex=0.7)
+title("Amakihi\n(observer)", cex.main=1.5)
+plot(amakihi_hr_obs_mas, pdf=TRUE, cex=0.7)
+title("Amakihi\n(observer+minutes)", cex.main=1.5)
 ```
@@
 This summary information includes details of the data and model specification, as well as the values of the coefficients ($\beta_j$) and their uncertainties, an "average" value for the detectability (see "Estimating abundance and variance" for details on how this is calculated) and its uncertainty.
-The final line gives an estimate of abundance for the area covered by the survey (see the next section).
+The final line gives an estimate of abundance in the area covered by the survey (see the next section).
 
@@
 
-To judge how well a detection functions fits the data when we have exact distances, we use a quantile-quantile plot (Q-Q plot).
+To visually assess how well a detection functions fits the data when we have exact distances, we use a quantile-quantile plot (Q-Q plot).
 Here the Q-Q plot compares the the cumulative distribution function (CDF) of the fitted detection function to the distribution of the data (empirical distribution function or EDF).
@@
 The EDF is is the proportion of points that have been observed at a distance equal to or less than the distance of that observation.
-The CFD is calculated from the fitted detection function as the probability of observing an object at a distance less than or equal to that of the given observation.
+The CDF is calculated from the fitted detection function as the probability of observing an object at a distance less than or equal to that of the given observation.
 This can be interpreted as assessing whether the number of observations up to a given distance is in line with what the model says they should be.
@@
 
-Once we have a set of plausible models, we can use Akaike's Information Criterion (AIC) to select between models.
+Once we have a set of plausible models, we can use Akaike's Information Criterion (AIC) to select between models [see e.g.
+@burnham2003model].
 \pkg{Distance} includes a function to create table of summary information for fitted models, making it easy to get an overview of a large number of models.
 The \code{summarize_ds_models} function takes models as input and can be especially useful when paired with \pkg{knitr}'s \code{kable} function to create summary tables for publication [@knitr-pkg].
-An example of this output is shown in Table \ref{tab:amakihi} and was generated by the following call to \code{summarize_ds_models}:
+An example of this output (with all models included) is shown in Table \ref{tab:amakihi} and was generated by the following call to \code{summarize_ds_models}:
 
-_(LM: hmm we say that the hn model should be discarded based on fit but it is then included in the table...
-is this confusing?
-There is also no guidance here on model selection...
-although the AIC is smallest for amakihi.hr.obs.mas, some would argue that a difference of less than 2 does not really justify the inclusion of the additional covariate.
-Are we simply saying minimum AIC wins (if so this is not clear)?
-Should we point them at further discussion on the topic?)_
 ```{r summary-table-null, eval=FALSE}
@@
 ```
+
+In this case we may be skeptical about the top model as selected by AIC being truly better than the second best model, as there is only a very small difference in AICs.
+Generally, if the difference between AICs is less than 2, we may investigate multiple "best" models, potentially resorting to the simplest of these models.
 
@@
 We may wish to calculate abundance estimates for some sub-regions of the study region, we call these areas *strata*.
-For example, strata may be defined by habitat types or animal gender (or some combination) which may be of interest for biological or management reasons.
-To calculate estimates for a given stratification each observation must occur in a stratum which must be labelled with a \code{Region.Label} and have a corresponding \code{Area} (if we are using an animal characteristic like gender, we would have the areas be the same but if we were using say forested vs.
+For example, strata may be defined by habitat types or animal sex (or some combination) which may be of interest for biological or management reasons.
+To calculate estimates for a given stratification each observation must occur in a stratum which must be labelled with a \code{Region.Label} and have a corresponding \code{Area} (if we are using an animal characteristic like sex, we would have the areas be the same but if we were using say forested vs.
 wetland habitat the areas of those strata would be different).
@@
 
-We combine these two sources of variance by noting that squared coefficients of variation (approximately) add [@Goodman:1960dd] (sometimes referred to as "the delta method").
+We combine these two sources of variance by noting that squared coefficients of variation (approximately) add [@Goodman:1960dd] (often referred to as "the delta method").
 
@@
 
-<!-- LM The location of this Table needs changing it appears in the middle of the output. -->
-
 # Extensions
@@
 We noted at the start of the article that \pkg{Distance} is a simple-to-use wrapper around the package \pkg{mrds}.
-Additional features available in \pkg{mrds} include  models that relax the assumption that detection is certain at zero distance from the transect.
-This is done using mark-recapture type methods which require a different survey technique, known as double observer surveys [see @Burt:2014gu for an introduction].
+Additional features available in \pkg{mrds} include  models that relax the assumption that detection is certain at zero distance from the transect (by including data from additional observers).
+This is done using mark-recapture type methods which require additional survey methodology, known as double observer surveys [see @Burt:2014gu for an introduction].
 
@@
 
-Further distance sampling complexities such as uncertainty in measured covariates (e.g.
+Uncertainty in measured covariates (e.g.
 cluster size) and model uncertainty (when two models have similar fit but substantially different estimates) can be incorporated using the multi-analysis distance sampling package \pkg{mads} [@mads-pkg].
-In addition, \pkg{mads} can also incorporate sightings for which species identification could not be achieved.
-This is done by estimating the abundance of these 'unidentified' sightings and pro-rating them to the known species in the analysis as described in [@Gerrodette:2005].
+In addition, \pkg{mads} can also incorporate sightings with unknown species identification.
+This is done by estimating the abundance of these 'unidentified' sightings and pro-rating them to the known species [@Gerrodette:2005].
 
-In addition to the analysis, an equally important stage of a distance sampling survey is the survey design.
-We have therefore developed a package which allows users to test out different designs in their specific study region and tailor population attributes to reflect the species they are working with.
-In this way \pkg{DSsim} [@DSsim-pkg] allows users to more easily identify challenges unique to their study and select a survey design which is likely to yield the most accurate and precise estimates.
+Survey design is an important part of any survey.
+\pkg{DSsim} allows users to test out different designs in their study region and tailor population attributes to reflect the species they are working with.
+\pkg{DSsim} [@DSsim-pkg] allows users to more easily identify challenges unique to their study and select a survey design which is more likely to yield accurate and precise estimates.
 
@@
 In combination with tools such as \pkg{knitr} and \pkg{rmarkdown} [@rmarkdown-pkg], the helper functions in \pkg{Distance} provide a useful set of tools to perform reproducible analyses of wildlife abundance for both managers and ecologists.
-In addition, users should remember the other basic functions in \proglang{R} such as \code{hist} that enable straight forward exploratory data analysis.
+\proglang{R} and its extension packages provide many tools exploratory data analysis that can be useful for a distance sampling analysis.
 We hope that this paper provides useful examples for those wishing to pursue distance sampling in \proglang{R}.

### 2016-02-09 paper.Rmd
minor change to ds intro

@@
 
-The distribution of the observed distances is a product of detectability (sometimes referred to as "perception bias", @Marsh:1989ho) and the distribution of the animals (observed or unobserved) with respect to the line or point.
-We hope that our survey design will ensure that we can assume a distribution for the animals.
+The distribution of the observed distances is a product of detectability [sometimes referred to as "perception bias"; @Marsh:1989ho] and the distribution of the animals (observed or unobserved) with respect to the line or point.
+Our survey design allow us to assume a distribution for the animals with respect to the sampler.
 
@@
 
-```{r points-and-lines, fig.height=6, fig.width=7, echo=FALSE, message=FALSE, fig.cap="Left side plots show an example of a survey of an area containing a population of 500 objects; blue lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals.
+```{r points-and-lines, fig.height=6, fig.width=7, echo=FALSE, message=FALSE, fig.cap="Left side plots show an example of a survey of an area containing a population of 500 objects; blue dashed lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals.
 The right side of the figure shows histograms of observed distances (again, lines top and points bottom).\\label{fig:pointslines}"}

### 2016-03-18 paper.Rmd
Updated Figure 1
See #39

@@
 
-```{r detectdists, fig.height=5.5, fig.width=8, echo=FALSE, message=FALSE, fig.cap="Top panels show an example detection function (left), the expected availability of objects (middle) and the resulting distribution of detections (right) for line transects.
-Bottom panels show an example detection function (left), the expected availability of objects (middle) and the resulting distribution of detections (right) for point transects.
-The distributions in the right hand plots are obtained by multiplying the detection function by the availability function and are referred to in the text as the `composite function'.
+```{r detectdists, fig.height=5.5, fig.width=9, echo=FALSE, message=FALSE, fig.cap="Top panels show an example detection function (left), the pdf of object distances (middle) and the resulting pdf of detection distances (right) for line transects.
+Bottom panels show an example detection function (left), the pdf of object distances (middle) and the resulting pdf of detection distances (right) for point transects.
+The pdf of observed detection distances in the right hand plots are obtained by multiplying the detection function by the pdf of object distances.
 \\label{fig:detectiondist}"}
 
-opar <- par(mfrow=c(2,3))
+par(mfrow=c(2,3))
 #line and point transect examples
@@
 
+set.cex  = 1
+set.cex.lab = 1.2
+set.cex.main = 2
 #Line transect example
-plot(x, y, xlab = "Perpendicular distance (m)", ylab = "Probability of Detection", type = "l", ylim = c(0,1.1), main = "Probability of Detection", xaxp = c(0,1,5))
+plot(x, y, xlab = "Perpendicular distance", ylab = "Probability of detection", type = "l", ylim = c(0,1.1), xaxp = c(0,1,5), cex = set.cex, cex.lab = set.cex.lab)
 line.availability <- rep(1,100)
-plot(x, line.availability, xlab = "Perpendicular distance (m)", ylab = "Availability", type = "l", ylim = c(0,1.1), xlim = c(0,500), yaxt = "n", main = "Expected Availability")
-plot(x,y, xlab = "Perpendicular distance (m)", ylab= "", type = "l", ylim = c(0,1.1), yaxt = "n", main = "Observed Distribution")
+plot(x, line.availability, xlab = "Perpendicular distance", ylab = "pdf of object distances", type = "l", ylim = c(0,1.1), xlim = c(0,500), yaxt = "n", cex = set.cex, cex.lab = set.cex.lab, main = "Line Transect", cex.main = set.cex.main)
+plot(x,y, xlab = "Perpendicular distance", ylab= "pdf of observed distances", type = "l", ylim = c(0,1.1), yaxt = "n", cex = set.cex, cex.lab = set.cex.lab)
 
 #Point Transect example
-plot(x, y, xlab = "Radial distance (m)", ylab = "Probability of Detection", type = "l", ylim = c(0,1.1), main = "Probability of Detection", xaxp = c(0,1,5))
+plot(x, y, xlab = "Radial distance", ylab = "Probability of detection", type = "l", ylim = c(0,1.1), xaxp = c(0,1,5), cex = set.cex, cex.lab = set.cex.lab)
 point.availability <- 2*pi*x
-plot(x, point.availability, xlab = "Radial distance (m)", ylab = "Availability", type = "l", ylim = c(0,3300), xlim = c(0,500), yaxt = "n", main = "Expected Availability")
+plot(x, point.availability, xlab = "Radial distance", ylab = "pdf of object distances", type = "l", ylim = c(0,3300), xlim = c(0,500), yaxt = "n", cex = set.cex, cex.lab = set.cex.lab, main = "Point Transect", cex.main = set.cex.main)
 point.observed <- y*point.availability
-plot(x, point.observed, xlab = "Radial distance (m)", ylab= "", type = "l", ylim = c(0,800), yaxt = "n", main = "Observed Distribution")
+plot(x, point.observed, xlab = "Radial distance", ylab= "pdf of observed distances", type = "l", ylim = c(0,800), yaxt = "n", cex = set.cex, cex.lab = set.cex.lab)
 

### 2016-03-18 paper.Rmd
Fix x-axis ticks in Figure 1
#39

@@
 #Line transect example
-plot(x, y, xlab = "Perpendicular distance", ylab = "Probability of detection", type = "l", ylim = c(0,1.1), xaxp = c(0,1,5), cex = set.cex, cex.lab = set.cex.lab)
+plot(x, y, xlab = "Perpendicular distance", ylab = "Probability of detection", type = "l", ylim = c(0,1.1), cex = set.cex, cex.lab = set.cex.lab)
 line.availability <- rep(1,100)
@@
 #Point Transect example
-plot(x, y, xlab = "Radial distance", ylab = "Probability of detection", type = "l", ylim = c(0,1.1), xaxp = c(0,1,5), cex = set.cex, cex.lab = set.cex.lab)
+plot(x, y, xlab = "Radial distance", ylab = "Probability of detection", type = "l", ylim = c(0,1.1), cex = set.cex, cex.lab = set.cex.lab)
 point.availability <- 2*pi*x

### 2016-03-18 paper.Rmd
Updated Figure 1 caption
added note about arbitrary distance
Closes #39
first run at Len's comments (easy bits)

@@
   - name: David L Miller
-    affiliation: University of St Andrews
+    affiliation: Integrated Statistics and University of St Andrews
     address: >
-      Centre for Research into Ecological and Environmental Modelling,
-      The Observatory,
-      St Andrews, Fife KY16 9LZ, Scotland
+      16 Sumner Street, Woods Hole, MA 02543, USA
     email: dave@ninepointeightone.net
@@
     email: len.thomas@st-andrews.ac.uk
+    url: http://lenthomas.org
   - name: Laura Marshall
@@
   - name: Jeffrey L Laake
-    affiliation: NOAA
+    affiliation: National Oceanic and Atmospheric Administration
     address: >
@@
 We describe how users can obtain estimates of abundance (and density) using the package as well documenting the links it provides with other more specialized \proglang{R} packages.
-We also demonstrate how \pkg{Distance} provides a migration pathway from previous software, thereby allowing us to deliver cutting edge methods to the users more quickly.
+We also demonstrate how \pkg{Distance} provides a migration pathway from previous software, thereby allowing us to deliver cutting-edge methods to the users more quickly.
 
@@
   # at least one keyword must be supplied
-  formatted: [distance sampling, abundance estimation, line transects, point transects, detection functions, Horvitz-Thompson, "\\proglang{R}", Distance]
-  plain:     [distance sampling, abundance estimation, line transects, point transects, detection functions, Horvitz-Thompson, R, Distance]
+  formatted: [distance sampling, abundance estimation, line transect, point transect, detection function, Horvitz-Thompson, "\\proglang{R}", Distance]
+  plain:     [distance sampling, abundance estimation, line transect, point transecs, detection function, Horvitz-Thompson, R, Distance]
 preamble: >
@@
 
-<!--
-(LT: I wonder who the intended audience is for this paper?
-(LM: - Distance for Windows users and other people looking do distance sampling in R?) The way it is currently written, it ``kinda'' introduces distance sampling as well as the relevant commands.
-Is this standard for a paper in this journal -- to introduce the methods as well as the software?
-Can we assume a reasonable level of analytical sophistication on the part of the reader, and so start with a few more formulae?
-Can we assume they know the methods, more or less, or point them at somewher where this is already introduced.
-(Perhaps I should go back and have a look at the EcolApps paper to see what approach we took there...
-(LM: The paper on Distance in JAppEcol?
-There is quite a big intro on distance sampling there.)))
-
-DLM: if you look at the unmarked paper they do a fair bit of explanation of the methods.-->
-
 Distance sampling [@Buckland:2001vm; @Buckland:2004ts; @buckland2015distance] encompasses a suite of methods used to estimate the density and/or abundance of biological populations.
@@
 One of the key assumptions of plot sampling is that all objects within each of the plots are counted.
-Distance sampling relaxes this assumption in that observers are no longer required to see and count everything within selected plots.
+Distance sampling relaxes this assumption in that observers are no longer required to detect (either by eye, video/audio recording etc) and count everything within selected plots.
 While plot sampling techniques are adequate for static populations occuring at high density they are inefficient for more sparsely distributed populations.
@@
 
-Conventional distance sampling assumes the observer is either stood at a point or moving along a line and will observe all objects (animals, plants, golf tees etc) that occur at the point or on the line.
+Conventional distance sampling assumes the observer is located either at a point or moving along a line and will observe all objects that occur at the point or on the line.
 The further away an object is from the point or line (also known as the sampler or transect) the less likely it is that the observer will see it.
@@
 
-Given that the lines or points are located independently of the objects, for line transect studies we assume that objects are uniformly distributed with respect to the line (i.e.
-the number of animals available for detection is constant).
+Given that the lines or points are located independently of the objects, for line transect studies we assume that objects are uniformly distributed with respect to the line (i.e., the number of animals available for detection is the same at all distances).
 For point transect surveys the survey area increases linearly with radial distance, implying a triangular distribution with respect to the point.
@@
 The pdf of observed detection distances in the right hand plots are obtained by multiplying the detection function by the pdf of object distances.
+Note, distances shown on x-axis for all plots are arbitrary.
 \\label{fig:detectiondist}"}
@@
 Depending on the survey species some of these assumptions may be more difficult to meet than others.
-Further information on field methods to help meet these assumptions can be found in [@buckland2015distance].
+Further information on field methods to help meet these assumptions can be found in @Buckland:2001vm and @buckland2015distance.
 
@@
 
-We introduce two example analyses performed in \pkg{Distance}: one line transect and one point transect. These data sets have been chosen as they represent typical data seen in practice.
+We introduce two example analyses performed in \pkg{Distance}: one line transect and one point transect.
+These data sets have been chosen as they represent typical data seen in practice.
+The below example analyses are not intended to serve as guidelinies for practice, but to demonstrate features of the software.
+Extentive practical advice on approaches to analysis is given in @Thomas:2010cf.
 
@@
 It is crucial that the detection function accurately models detectability at small distances; we are less worried by its behaviour further away from 0.
-To ensure that the model is not overly influenced by distances far from zero, we truncate the distances beyond a given distance $w$.
-Including these larger distances in our analysis does not demonstrably improve the precision of abundance estimates [@Buckland:2001vm, sections 4.3 and 5.3].
+To ensure that the model is not overly influenced by distances far from zero, we truncate the distances beyond a given distance $w$ (this can be done during analysis or while collecting data in the field).
 
@@
 * *Non-increasing*: we do not think that observers should be more likely to see distant objects than those nearer the transect.
-If this occurs, it usually indicates an issue with survey design or field procedures (for example that the distribution of objects with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this [see @Miller:2015hw, @Marques:2010he and @Marques:2012fy for more on this topic].
+If this occurs, it usually indicates an issue with survey design or field procedures (for example that the distribution of objects with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this [@Miller:2015hw; @Marques:2010he; @Marques:2012fy].
 * *Model robust*: models should be flexible enough to fit many different shapes.
@@
 
-When adjustment terms are used it may be necessary to standardise the results to ensure that $g(0)=1$, so we can redefine the detection function as:
+When adjustment terms are used it is necessary to standardise the results to ensure that $g(0)=1$, so we can redefine the detection function as:
 $$
@@
 Including covariates has an important implication for our calculation of detectability.
-Because we do not know the true distribution of the covariates,we must calculate the probability of detection conditional on the observed values of the covariates:
+Because we do not know the true distribution of the covariates, we can either model the distribution of the covariates and integrate the covariates out of the joint density, or we can calculate the probability of detection conditional on the observed values of the covariates [@Marques:2003vb].
+We opt for the latter:
 $$
@@
 ```
-Here \code{ds} had also fitted a model with cosine adjustments (the default) but finds the AIC improvement to be insufficient to select the adjustment.
+Here \code{ds} also fits the hazard-rate model then hazard-rate with a cosine adjustment but the AIC improvement is insufficient to select the adjustment, so the hazard-rate key-only model is returned.
+
+Other adjustment series can be selected using the `adjustment=` argument and specific orders of adjustments can be set using `order=`.
+For example, to specify a uniform model with cosine adjustments of order 1 and 2 we can write:
+```{r minke-unif-cos}
+minke_unifcos <- ds(minke, truncation=1.5, key="unif", adjustment="cos", order=c(1,2))
+```
+Hermite polynomial adjustments use the code `"herm"` and simple polynomials `"poly"`, adjustment order should be in line with Table \ref{tab:keyadj}.
 
@@
 
-To visually assess how well a detection functions fits the data when we have exact distances, we use a quantile-quantile plot (Q-Q plot).
+We use a quantile-quantile plot (Q-Q plot) to visually assess how well a detection functions fits the data when we have exact distances.
 Here the Q-Q plot compares the the cumulative distribution function (CDF) of the fitted detection function to the distribution of the data (empirical distribution function or EDF).
@@
 Q-Q plots can be inspected visually, though this is prone to subjective judgments.
-Therefore, we also quantify the Q-Q plot's information using a Kolmogorov-Smirnov or Cramer-von Mises test [@Burnham:2004vd].
+Therefore, we also quantify the Q-Q plot's information using a Kolmogorov-Smirnov or Cramér-von Mises test [@Burnham:2004vd].
 Both test whether points from the EDF and CDF are from the same distribution.
-The Kolmogorov-Smirnov uses the test statistic of the largest difference between a point on the Q-Q plot and the line $y=x$, whereas the Cramer-von Mises test uses the sum of all the differences.
-As it takes into account more information and is therefore more powerful, the Cramer-von Mises is generally preferred.
+The Kolmogorov-Smirnov uses the test statistic of the largest difference between a point on the Q-Q plot and the line $y=x$, whereas the Cramér-von Mises test uses the sum of all the differences.
+As it takes into account more information and is therefore more powerful, the Cramér-von Mises is generally preferred.
 A significant result (usually $P < 0.05$) from either test gives evidence against the null hypothesis (that the data arose from the fitted model), suggesting that the model does not fit the data well.
@@
 Generally, if the difference between AICs is less than 2, we may investigate multiple "best" models, potentially resorting to the simplest of these models.
+In the authors' experience, it is often the case that models with similar AICs will have similar estimates probabilities of detection, so in practice there is little difference in selecting between these models.
 
@@
       caption="Summary for the detection function models fitted to the amakihi data.
-``C-vM'' stands for Cramer-von Mises, $P_a$ indicates average detectability (see ``Estimating abundance and variance''), se indicates standard error.
+``C-vM'' stands for Cramér-von Mises, $P_a$ indicates average detectability (see ``Estimating abundance and variance''), se indicates standard error.
 Models are sorted according to AIC.\\label{tab:amakihi}")
@@
 
-Though fitting the detection function is the primary modelling step in distance sampling, we are really interested in estimating detectability, and from that density or abundance.
+Though fitting the detection function is the primary modelling step in distance sampling, we are really interested in estimating density or abundance.
 It is also important to calculate our uncertainty for these estimates.
@@
 
-We wish to obtain the abundance in a study region, of which we have sampled a (representative) subset.
+We wish to obtain the abundance in a study region, of which we have sampled a random subset.
 To do this we first calculate the abundance in the area we have surveyed (the *covered area*) to obtain $\hat{N}_\text{C}$, we can then scale this up to the full study area by multiplying it by the ratio of covered area to study area.
@@
 We may wish to calculate abundance estimates for some sub-regions of the study region, we call these areas *strata*.
+Stratification can be used to increase the precision of estimates if we know *a priori* that density varies between different parts of the study area.
 For example, strata may be defined by habitat types or animal sex (or some combination) which may be of interest for biological or management reasons.
@@
 The \code{varn} manual page gives further advice and technical detail on encounter rate variance.
+For example for systematic survey designs, estimators S1, S2 and O1, O2 and O3 will typically provide lower variance estimates.
 
@@
 We would particularly like to thank Steve Buckland, David Borchers, Tiago Marques, Jon Bishop and Lorenzo Milazzo for their contributions.
+We also thank Ken Burnham and David Anderson for fundamental contributions to the early development of these methods.
 

### 2016-03-18 paper.Rmd
remove minke strata figure

@@
 The line transect data have been simulated from models fitted to Antarctic minke whale (*Balaenoptera bonaerensis*) data.
-These data were collected as part of the International Whaling Commission's International Decade of Cetacean Research Southern Ocean Whale and Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer surveys [@Branch:2001ua].
-This data set consists of 99 observations that are stratified based on location (near or distant from ice edge) and effort data (transect lengths).
-The survey is shown in Figure \ref{fig:minke-strata}.
-
-\begin{figure}
-\includegraphics{minke-strata}
-\caption{Strata used for the minke whale data adapted from Hedley and Buckland (2004).
-Points show the (simulated) locations of observations along transect lines.
-The stepped line shows the boundary between North and South strata.
-Further details on the survey are available in Branch and Butterworth (2001) (simulated data is based on ``1992/93 Area III'' therein).\label{fig:minke-strata}}
-\end{figure}
+These data were collected as part of the International Whaling Commission's International Decade of Cetacean Research Southern Ocean Whale and Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer surveys.
+These data consist of 99 observations that are stratified based on location (near or distant from ice edge) and effort data (transect lengths).
+Further details on the survey are available in @Branch:2001ua (simulated data is based on ``1992/93 Area III'' therein).
 
@@
 As an example, the minke whale data contain two strata: \code{North} and \code{South} relating to strata further away from and nearer to the Antarctic ice edge, respectively.
-Figure \ref{fig:minke-strata} shows the two strata, along with observations and transect lines.
 

### 2016-03-18 paper.Rmd
Distance history

@@
 The Windows program Distance (or "DISTANCE"; for clarity henceforth "Distance for Windows") can be used to fit detection functions to distance sampling data.
-It was first released in November 1998 as Distance for Windows version 3.5.
+It was first released (versions 1.0 - 3.0; principally programmed by Jeff Laake while working at the National Marine Mammal Laboratory) as a console-based application (this in turn was based on earlier software TRANSECT, @Burnham:1980wz and algorithms developed in @Buckland:1992fa), before the first graphical interface (Distance for Windows 3.5) was released in November 1998.
 Since this time it has evolved to include various design and analysis features [@Thomas:2010cf].
@@
 
-The two example data sets used here are distributed with \pkg{Distance} so readers can reproduce our analyses.
 Generally, data collected in the field will require some formatting before use with \pkg{Distance}, though there are a range of possible formats, dependent on the model specification and the output required:

### 2016-03-18 paper.Rmd
add Eric's extensions

@@
 
+The examples of applications for which distance sampling has been used is large.
+Objects being detected need not be animals; plants [@Buckland:2007kl], beer cans [@Otto:1990gs], elephant dung [@NchPlu01] and bricks on a lake bottom [@Ber90] have been subjects of distance sampling investigations.
+The method of detection need not be visual sightings.
+Detections of objects can be through auditory as well as visual means [@Marques2013b].
+Songs of birds [@Buckland2006] and whale vocalisations [@BorSub1] are just two examples.
+Blows made by whales are examples of processes that can be modelled using distance sampling.
+Songs and blows are indirect sampling methods producing estimates of cue density.
+Cue densities can be converted to animal densities with additional data needed to estimate the rate at which cues are produced and the rate at which they disappear.
+These are but a few examples of the applications to which distance sampling has been applied.
+
 The features of \pkg{Distance} are deliberately limited to provide a simplified interface for users.

### 2016-03-18 paper.Rmd
Len's corrections to Eric's extensions

@@
 
-The examples of applications for which distance sampling has been used is large.
+The examples of applications for which distance sampling has been used are many.
 Objects being detected need not be animals; plants [@Buckland:2007kl], beer cans [@Otto:1990gs], elephant dung [@NchPlu01] and bricks on a lake bottom [@Ber90] have been subjects of distance sampling investigations.
@@
 Cue densities can be converted to animal densities with additional data needed to estimate the rate at which cues are produced and the rate at which they disappear.
-These are but a few examples of the applications to which distance sampling has been applied.
+These are but a few examples of the applications to which distance sampling has been applied (a larger, but incomplete list of references is given at \url{http://distancesampling.org/dbib.html}).
 

### 2016-03-18 paper.Rmd
better lt/pt diagram?

@@
 # generate some animals to sample
-set.seed(131) # same results every time
+set.seed(15) # same results every time
 library("mgcv") # for inSide
@@
 # set sigma
-sigma <- 0.01
+sigma <- 0.035
 # storage for detected distances
 detected_distances <- c()
-for(i in 1:5){
+for(i in 1:6){
   # calculate next strip location
@@
   detected <- exp(-distances^2/(2*sigma^2)) > runif(length(distances))
+#  detected <- (1-exp(-(distances/sigma)^(-1.2))) > runif(length(distances))
   # plot those objects detected

### 2016-03-18 paper.Rmd
figure 1 changes

@@
 
-```{r detectdists, fig.height=5.5, fig.width=9, echo=FALSE, message=FALSE, fig.cap="Top panels show an example detection function (left), the pdf of object distances (middle) and the resulting pdf of detection distances (right) for line transects.
-Bottom panels show an example detection function (left), the pdf of object distances (middle) and the resulting pdf of detection distances (right) for point transects.
-The pdf of observed detection distances in the right hand plots are obtained by multiplying the detection function by the pdf of object distances.
-Note, distances shown on x-axis for all plots are arbitrary.
+```{r detectdists, fig.height=5.5, fig.width=9, echo=FALSE, message=FALSE, fig.cap="Panels show an example detection function (left), the pdf of object distances (middle) and the resulting pdf of detection distances (right) for line transects (top row) and point transects (bottom row).
+The pdf of observed detection distances in the right hand plots are obtained by multiplying the detection function by the pdf of object distances and rescaling.
+In this example, detection probability becomes effectively zero at 500 (distances shown on $x$-axis are arbitrary).
 \\label{fig:detectiondist}"}
@@
 line.availability <- rep(1,100)
-plot(x, line.availability, xlab = "Perpendicular distance", ylab = "pdf of object distances", type = "l", ylim = c(0,1.1), xlim = c(0,500), yaxt = "n", cex = set.cex, cex.lab = set.cex.lab, main = "Line Transect", cex.main = set.cex.main)
-plot(x,y, xlab = "Perpendicular distance", ylab= "pdf of observed distances", type = "l", ylim = c(0,1.1), yaxt = "n", cex = set.cex, cex.lab = set.cex.lab)
+plot(x, line.availability, xlab = "Perpendicular distance", ylab = "pdf of object distances", type = "l", ylim = c(0,1.1), xlim = c(0,500), cex = set.cex, cex.lab = set.cex.lab, main = "Line Transect", cex.main = set.cex.main)
+plot(x,y, xlab = "Perpendicular distance", ylab= "pdf of observed distances", type = "l", ylim = c(0,1.1), cex = set.cex, cex.lab = set.cex.lab)
 
@@
 point.availability <- 2*pi*x
-plot(x, point.availability, xlab = "Radial distance", ylab = "pdf of object distances", type = "l", ylim = c(0,3300), xlim = c(0,500), yaxt = "n", cex = set.cex, cex.lab = set.cex.lab, main = "Point Transect", cex.main = set.cex.main)
+plot(x, point.availability, xlab = "Radial distance", ylab = "pdf of object distances", type = "l", ylim = c(0,3300), xlim = c(0,500), cex = set.cex, cex.lab = set.cex.lab, main = "Point Transect", cex.main = set.cex.main)
 point.observed <- y*point.availability
-plot(x, point.observed, xlab = "Radial distance", ylab= "pdf of observed distances", type = "l", ylim = c(0,800), yaxt = "n", cex = set.cex, cex.lab = set.cex.lab)
+plot(x, point.observed, xlab = "Radial distance", ylab= "pdf of observed distances", type = "l", ylim = c(0,800), cex = set.cex, cex.lab = set.cex.lab)
 

### 2016-03-18 paper.Rmd
survey design stuff

@@
 The distribution of the observed distances is a product of detectability [sometimes referred to as "perception bias"; @Marsh:1989ho] and the distribution of the animals (observed or unobserved) with respect to the line or point.
-Our survey design allow us to assume a distribution for the animals with respect to the sampler.
+Our survey design allows us to assume a distribution for the animals with respect to the sampler.
 
-Given that the lines or points are located independently of the objects, for line transect studies we assume that objects are uniformly distributed with respect to the line (i.e., the number of animals available for detection is the same at all distances).
+For line transect studies we assume that objects are uniformly distributed with respect to the line (i.e., the number of animals available for detection is the same at all distances).
 For point transect surveys the survey area increases linearly with radial distance, implying a triangular distribution with respect to the point.
@@
 
-```{r points-and-lines, fig.height=6, fig.width=7, echo=FALSE, message=FALSE, fig.cap="Left side plots show an example of a survey of an area containing a population of 500 objects; blue dashed lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals.
+```{r points-and-lines, fig.height=6, fig.width=7, echo=FALSE, message=FALSE, fig.cap="Left side plots show an example of a survey of a (1x1) area containing a population of 500 objects; blue dashed lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals.
 The right side of the figure shows histograms of observed distances (again, lines top and points bottom).\\label{fig:pointslines}"}
@@
 par(opar)
+```
 
-```
+Good survey design is essential to produce reliable density and abundance estimates from distance sampling data.
+Survey design is beyond the scope of this article but we refer readers to @Buckland:2001vm [Chapter 7] and @buckland2015distance [Chapter 2] for introductory information; @Buckland:2004ts [Chapter 7] contains information on automated survey design using geographical information systems (GIS); @Thomas:2007wz gives an example of designing a distance sampling survey in a complex region.
 

### 2016-03-18 paper.Rmd
add GIS in Distance for Windows
Laura's adjustments to figure 1

@@
 
+hn.detect <- function(x, sigma) exp(-((x^2)/(2*(sigma^2))))
+hn.detect.scaled <- function(x, sigma) hn.detect(x,sigma)/integrate(hn.detect, 0, 500, sigma = sigma)$value
+
 x <- seq(0,500, length = 100)
-y <- exp(-((x^2)/(2*(200^2))))
+y <- hn.detect(x, sigma = 200)
+y.scaled <- hn.detect.scaled(x, sigma = 200)
 
@@
 plot(x, y, xlab = "Perpendicular distance", ylab = "Probability of detection", type = "l", ylim = c(0,1.1), cex = set.cex, cex.lab = set.cex.lab)
-line.availability <- rep(1,100)
-plot(x, line.availability, xlab = "Perpendicular distance", ylab = "pdf of object distances", type = "l", ylim = c(0,1.1), xlim = c(0,500), cex = set.cex, cex.lab = set.cex.lab, main = "Line Transect", cex.main = set.cex.main)
-plot(x,y, xlab = "Perpendicular distance", ylab= "pdf of observed distances", type = "l", ylim = c(0,1.1), cex = set.cex, cex.lab = set.cex.lab)
+line.availability <- rep(0.002,100)
+plot(x, line.availability, xlab = "Perpendicular distance", ylab = "pdf of object distances", type = "l", ylim = c(0,0.0022), xlim = c(0,500), cex = set.cex, cex.lab = set.cex.lab, main = "Line Transect", cex.main = set.cex.main)
+plot(x,y.scaled, xlab = "Perpendicular distance", ylab= "pdf of observed distances", type = "l", ylim = c(0,0.0044), cex = set.cex, cex.lab = set.cex.lab)
 
@@
 plot(x, y, xlab = "Radial distance", ylab = "Probability of detection", type = "l", ylim = c(0,1.1), cex = set.cex, cex.lab = set.cex.lab)
-point.availability <- 2*pi*x
-plot(x, point.availability, xlab = "Radial distance", ylab = "pdf of object distances", type = "l", ylim = c(0,3300), xlim = c(0,500), cex = set.cex, cex.lab = set.cex.lab, main = "Point Transect", cex.main = set.cex.main)
-point.observed <- y*point.availability
-plot(x, point.observed, xlab = "Radial distance", ylab= "pdf of observed distances", type = "l", ylim = c(0,800), cex = set.cex, cex.lab = set.cex.lab)
 
+point.area <- function(x) 2*pi*x
+
+point.availability <- 2*pi*x/integrate(point.area, 0, 500)$value
+
+plot(x, point.availability, xlab = "Radial distance", ylab = "pdf of object distances", type = "l", ylim = c(0,0.0044), xlim = c(0,500), cex = set.cex, cex.lab = set.cex.lab, main = "Point Transect", cex.main = set.cex.main)
+
+point.observed.shape <- function(x, sigma) 2*pi*x*hn.detect(x, sigma)
+
+pdf.point.observed <- function(x, sigma) point.observed.shape(x, sigma = sigma)/integrate(point.observed.shape, 0, 500, sigma = sigma)$value
+
+pdf.observed <- pdf.point.observed(x, sigma = 200)
+
+plot(x, pdf.observed, xlab = "Radial distance", ylab= "pdf of observed distances", type = "l", ylim = c(0,0.00345), cex = set.cex, cex.lab = set.cex.lab)
 ```
@@
 Survey design is beyond the scope of this article but we refer readers to @Buckland:2001vm [Chapter 7] and @buckland2015distance [Chapter 2] for introductory information; @Buckland:2004ts [Chapter 7] contains information on automated survey design using geographical information systems (GIS); @Thomas:2007wz gives an example of designing a distance sampling survey in a complex region.
+We also note that Distance for Windows includes a GIS survey design tool [@Thomas:2010cf].
 
@@
 \begin{table}
-\caption{Modelling options for key plus adjustment series models for the detection function. Adapted from Buckland et al. (2001), section 2.4.}
+\caption{Modelling options for key plus adjustment series models for the detection function.
+Adapted from Buckland et al.
+(2001), section 2.4.
+For each key function the default adjustments are cosine in \pkg{Distance}.}
 \begin{tabular}{llll}

### 2016-03-18 paper.Rmd
axes fix?

@@
 #Line transect example
-plot(x, y, xlab = "Perpendicular distance", ylab = "Probability of detection", type = "l", ylim = c(0,1.1), cex = set.cex, cex.lab = set.cex.lab)
-line.availability <- rep(0.002,100)
-plot(x, line.availability, xlab = "Perpendicular distance", ylab = "pdf of object distances", type = "l", ylim = c(0,0.0022), xlim = c(0,500), cex = set.cex, cex.lab = set.cex.lab, main = "Line Transect", cex.main = set.cex.main)
-plot(x,y.scaled, xlab = "Perpendicular distance", ylab= "pdf of observed distances", type = "l", ylim = c(0,0.0044), cex = set.cex, cex.lab = set.cex.lab)
+plot(x, y, xlab = "Perpendicular distance", ylab = "Probability of detection", type = "l", ylim = c(0,1.1), cex = set.cex, cex.lab = set.cex.lab, axes=FALSE)
+axis(2, at=c(0,1/2,1)); axis(1); box()
+
+w <- 1/500
+line.availability <- rep(w,100)
+plot(x, line.availability, xlab = "Perpendicular distance", ylab = "pdf of object distances", type = "l", ylim = c(0,w), xlim = c(0,500), cex = set.cex, cex.lab = set.cex.lab, main = "Line Transect", cex.main = set.cex.main, axes=FALSE)
+axis(2, at=c(0,w/2,w)); axis(1); box()
+
+plot(x,y.scaled, xlab = "Perpendicular distance", ylab= "pdf of observed distances", type = "l", ylim = c(0,0.0044), cex = set.cex, cex.lab = set.cex.lab, axes=FALSE)
+axis(2, at=round(c(0,median(y.scaled),max(y.scaled)),3)); axis(1); box()
 
 #Point Transect example
-plot(x, y, xlab = "Radial distance", ylab = "Probability of detection", type = "l", ylim = c(0,1.1), cex = set.cex, cex.lab = set.cex.lab)
+plot(x, y, xlab = "Radial distance", ylab = "Probability of detection", type = "l", ylim = c(0,1.1), cex = set.cex, cex.lab = set.cex.lab, axes=FALSE)
+axis(2, at=c(0,1/2,1)); axis(1); box()
 
@@
 
-plot(x, point.availability, xlab = "Radial distance", ylab = "pdf of object distances", type = "l", ylim = c(0,0.0044), xlim = c(0,500), cex = set.cex, cex.lab = set.cex.lab, main = "Point Transect", cex.main = set.cex.main)
+plot(x, point.availability, xlab = "Radial distance", ylab = "pdf of object distances", type = "l", ylim = c(0,2*w), xlim = c(0,500), cex = set.cex, cex.lab = set.cex.lab, main = "Point Transect", cex.main = set.cex.main, axes=FALSE)
+axis(2, at=c(0,w,2*w)); axis(1); box()
 
@@
 pdf.observed <- pdf.point.observed(x, sigma = 200)
-
-plot(x, pdf.observed, xlab = "Radial distance", ylab= "pdf of observed distances", type = "l", ylim = c(0,0.00345), cex = set.cex, cex.lab = set.cex.lab)
+plot(x, pdf.observed, xlab = "Radial distance", ylab= "pdf of observed distances", type = "l", ylim = c(0,0.00345), cex = set.cex, cex.lab = set.cex.lab, axes=FALSE)
+axis(2, at=round(c(0,max(pdf.observed)/2,max(pdf.observed)),4)); axis(1); box()
 ```

### 2016-03-18 paper.Rmd
Minor modification Figure 1
Ah I see you had done that any way @dill I just modified the ylims
slightly  think that looks better but undo if you don’t like #39

@@
 line.availability <- rep(w,100)
-plot(x, line.availability, xlab = "Perpendicular distance", ylab = "pdf of object distances", type = "l", ylim = c(0,w), xlim = c(0,500), cex = set.cex, cex.lab = set.cex.lab, main = "Line Transect", cex.main = set.cex.main, axes=FALSE)
+plot(x, line.availability, xlab = "Perpendicular distance", ylab = "pdf of object distances", type = "l", ylim = c(0,w*1.1), xlim = c(0,500), cex = set.cex, cex.lab = set.cex.lab, main = "Line Transect", cex.main = set.cex.main, axes=FALSE)
 axis(2, at=c(0,w/2,w)); axis(1); box()
@@
 
-plot(x, point.availability, xlab = "Radial distance", ylab = "pdf of object distances", type = "l", ylim = c(0,2*w), xlim = c(0,500), cex = set.cex, cex.lab = set.cex.lab, main = "Point Transect", cex.main = set.cex.main, axes=FALSE)
+plot(x, point.availability, xlab = "Radial distance", ylab = "pdf of object distances", type = "l", ylim = c(0,2*w*1.1), xlim = c(0,500), cex = set.cex, cex.lab = set.cex.lab, main = "Point Transect", cex.main = set.cex.main, axes=FALSE)
 axis(2, at=c(0,w,2*w)); axis(1); box()

### 2016-03-19 paper.Rmd
add survey studd, mess with abstract

--- 2016-03-18 paper.Rmd
@@
   Estimating the abundance and spatial distribution of animal and plant populations is essential for conservation and management.
-We introduce the \proglang{R} package \pkg{Distance} that uses distance sampling methods to estimate abundance.
+We introduce the \proglang{R} package \pkg{Distance} that implements distance sampling methods to estimate abundance.
 We describe how users can obtain estimates of abundance (and density) using the package as well documenting the links it provides with other more specialized \proglang{R} packages.
@@
 
-Covariates can be included in this formulation by considering the scale parameter from the half-normal or hazard-rate detection functions as a linear model (on the exponential scale) of the ($J$) covariates ($\mathbf{z}$; a vector of length $J$ for each observation):
+Covariates can be included in this formulation by considering the scale parameter from the half-normal or hazard-rate detection functions as a(n exponentiated) linear model of the ($J$) covariates ($\mathbf{z}$; a vector of length $J$ for each observation):
 $$
@@
 As such we cannot set meaningful constraints on monotonicity.
-For this reason, we advise against using both adjustments and covariates in a detection function [see @Miller:2015hw for an example of when this can be problematic].
+For this reason, we advise caution when using both adjustments and covariates in a detection function [see @Miller:2015hw for an example of when this can be problematic].
 
@@
 
-Survey design is an important part of any survey.
+As mentioned above, survey design is critical to ensuring that resulting distance sampling data can be reliably analysed.
 \pkg{DSsim} allows users to test out different designs in their study region and tailor population attributes to reflect the species they are working with.

### 2016-03-20 paper.Rmd
more annoying bibtext stuff
clarify distance scaling

@@
 \caption{Modelling options for key plus adjustment series models for the detection function.
-Adapted from Buckland et al.
-(2001), section 2.4.
-For each key function the default adjustments are cosine in \pkg{Distance}.}
+For each key function the default adjustments are cosine in \pkg{Distance}.
+Note that in the adjustments functions distance is divided by the width or the scale parameter to ensure the shape of adjustment independent of the units of $y$ (Marques et al., 2007); defaults are shown here, though either can be selected to rescale the distances.}
 \begin{tabular}{llll}
@@
 We would particularly like to thank Steve Buckland, David Borchers, Tiago Marques, Jon Bishop and Lorenzo Milazzo for their contributions.
+We thank Tiago Marques a second time for helpful comments on an early version of the paper.
 We also thank Ken Burnham and David Anderson for fundamental contributions to the early development of these methods.

### 2016-03-20 paper.Rmd
getting bored of messing with these plots

@@
 
-```{r adjust-mix-comp, fig.width=10, fig.height=3, fig.cap="Possible (though not necessarily plausible) shapes for the detection function when adjustments are included for half-normal and hazard-rate models.\\label{fig:keyadj}", echo=FALSE}
+```{r adjust-mix-comp, fig.width=10, fig.height=3, fig.cap="Possible shapes for the detection function when cosine adjustments are included for half-normal and hazard-rate models.\\label{fig:keyadj}", echo=FALSE}
 par(mfrow=c(1,4), mar=c(3.5, 3, 3, 1) + 0.1)
@@
 # 2 cosines
-g <- function(x) exp(-x^2/(2*0.01^2))*(1-0.06*cos((2*pi*x)/0.025)+0.25*cos((3*pi*x)/0.025))
+g <- function(x) exp(-x^2/(2*0.011^2))*(1-0.06*cos((2*pi*x)/0.025)+0.25*cos((3*pi*x)/0.025))
 f <- function(x) g(x)/g(0)
@@
 ## hazard-rate with cosine
-g <- function(x) (1 - exp(-(x/0.005)^-1.1))*(1-0.05*cos((2*pi*x)/0.025))
+g <- function(x) (1 - exp(-(x/0.011)^-3))*(1+0.05*cos((2*pi*x)/0.025))
 f <- function(x) g(x)/g(0)
@@
 # 2 cosines
-g <- function(x) (1 - exp(-(x/0.005)^-1.9))*
-                  (1+0.05*cos((2*pi*x)/0.025) + 0.25*cos((3*pi*x)/0.025))
+g <- function(x) (1 - exp(-(x/0.0199)^-1.4))*
+                 (1-0.08*cos((2*pi*x)/0.025)+0.08*cos((3*pi*x)/0.025))
 f <- function(x) g(x)/g(0)
-curve(f, from=0, to=0.025, xlab="", ylab="", main="Hazard-rate with \n2 cosine adjustments", xlim=c(0,0.025), ylim=c(0,1))
+curve(f, from=0, to=0.025, xlab="", ylab="", main="Hazard-rate with \n2 cosine adjustment", xlim=c(0,0.025), ylim=c(0,1))
 title(xlab="Distance", line=2)

### 2016-03-20 paper.Rmd
chi-sq tests
small addition to group size stuff

@@
 * To include additional covariates into the detection function (see "Detection functions") a \code{data.frame} is required.
+Each row of the \code{data.frame} contains the data on one observation.
 The \code{data.frame} must contain a column named \code{distance} (containing the observed distances) and additional named columns for any covariates that may affect detectability (for example \code{observer} or \code{seastate}).
-The column name \code{size} is reserved for the cluster sizes (sometimes referred to as group sizes) for when the detection distances relate to observations of clusters rather than individuals.
+The column name \code{size} is reserved for the cluster sizes (sometimes referred to as group sizes) for when the detection distances relate to observations of clusters rather than individuals (see @Buckland:2001vm, Section 1.4.3 for more on defining clusters and "Extensions" below for one approach to dealing with uncertain cluster size).
 Additional reserved names include \code{object} and \code{detected}, these are not required for conventional distance sampling and should be avoided (see "Extensions" for an explanation of their use).
@@
 
+For non-exact distance data, a $\chi^2$-test can be used to assess goodness of fit [see @Buckland:2001vm, Section 3.4.4].
+$\chi^2$-test results are produced by `gof_ds` when binned/grouped distances are provided.
+
 ## Model selection

### 2016-03-20 paper.Rmd
that's enough thinking about variance for one evening

@@
 These are: the detectability $\hat{p}(\mathbf{z}_i)$ (and hence the parameters of the detection function it is derived from) and $n$, the number of observations.
-We assume that the observed group size ($s_i$) is recorded without error.
 
@@
 
-For line transects, by default, \pkg{Distance} uses a variation of the estimator "R2" from @Fewster:2009ku replacing number of observations per sample with the estimated abundance per sample [@Innes:2002ka; @Marques:2003vb]:
+For line transects, by default, \pkg{Distance} uses a variation of the estimator "R2" from @Fewster:2009ku replacing number of observations per sample with the estimated abundance per sample, thus taking into account cluster size if this is recorded [@Innes:2002ka; @Marques:2003vb]:
 $$
@@
 
-We combine these two sources of variance by noting that squared coefficients of variation (approximately) add [@Goodman:1960dd] (often referred to as "the delta method").
+We combine these two variances by noting that squared coefficients of variation (approximately) add [@Goodman:1960dd] (often referred to as "the delta method").
 

### 2016-03-20 paper.Rmd
spell check

@@
 Distance sampling relaxes this assumption in that observers are no longer required to detect (either by eye, video/audio recording etc) and count everything within selected plots.
-While plot sampling techniques are adequate for static populations occuring at high density they are inefficient for more sparsely distributed populations.
+While plot sampling techniques are adequate for static populations occurring at high density they are inefficient for more sparsely distributed populations.
 Distance sampling provides a more efficient solution in such circumstances.
@@
 As Distance for Windows becomes increasingly reliant on analyses performed in \proglang{R} and many new methods are being developed, we are encouraging the use of our \proglang{R} packages directly.
-\proglang{R} provides a huge variety of functionality for data exploration and reproducable research, much more than is possible in Distance for Windows.
+\proglang{R} provides a huge variety of functionality for data exploration and reproducible research, much more than is possible in Distance for Windows.
 
@@
 \pkg{Distance} provides a selection of candidate functions to describe the probability of detection and estimates the associated parameters using maximum likelihood estimation.
-The probability of detecting an object may not only depend on how far it is from the obsever but also on other factors such as weather conditions, ground cover, cluster size etc.
+The probability of detecting an object may not only depend on how far it is from the observer but also on other factors such as weather conditions, ground cover, cluster size etc.
 The \pkg{Distance} package also allows the incorporation of such covariates into the detection function allowing the detection function scale parameter to vary based on these covariates.
@@
 These data sets have been chosen as they represent typical data seen in practice.
-The below example analyses are not intended to serve as guidelinies for practice, but to demonstrate features of the software.
-Extentive practical advice on approaches to analysis is given in @Thomas:2010cf.
+The below example analyses are not intended to serve as guidelines for practice, but to demonstrate features of the software.
+Extensive practical advice on approaches to analysis is given in @Thomas:2010cf.
 
@@
 These data consist of 99 observations that are stratified based on location (near or distant from ice edge) and effort data (transect lengths).
-Further details on the survey are available in @Branch:2001ua (simulated data is based on ``1992/93 Area III'' therein).
+Further details on the survey are available in @Branch:2001ua (simulated data is based on "1992/93 Area III" therein).
 
@@
 
-While automatic AIC selection is performed on adjustment terms, model selection for covarariates must be performed manually.
-Here we add a second covariate: minutes after sunrrise.
+While automatic AIC selection is performed on adjustment terms, model selection for covariates must be performed manually.
+Here we add a second covariate: minutes after sunrise.
 We will compare these two models further in the following section.
@@
 We use a quantile-quantile plot (Q-Q plot) to visually assess how well a detection functions fits the data when we have exact distances.
-Here the Q-Q plot compares the the cumulative distribution function (CDF) of the fitted detection function to the distribution of the data (empirical distribution function or EDF).
+Here the Q-Q plot compares the cumulative distribution function (CDF) of the fitted detection function to the distribution of the data (empirical distribution function or EDF).
 The Q-Q plots in \pkg{Distance} plot a point for every observation.
@@
 The \code{varn} manual page gives further advice and technical detail on encounter rate variance.
-For example for systematic survey designs, estimators S1, S2 and O1, O2 and O3 will typically provide lower variance estimates.
+For example for systematic survey designs, estimators S1, S2 and O1, O2 and O3 will typically provide smaller estimates of the variance.
 

### 2016-03-21 paper.Rmd
last changes post final read through

@@
     address: >
-      16 Sumner Street, Woods Hole, MA 02543, USA
+      16 Sumner Street, Woods Hole, MA 02543, USA and
+      Centre for Research into Ecological and Environmental Modelling,
+      The Observatory,
+      St Andrews, Fife KY16 9LZ, Scotland
     email: dave@ninepointeightone.net
@@
 \pkg{Distance} is a wrapper package around \pkg{mrds} making it easier to get started with basic distance sampling analyses in \proglang{R}.
-Like the histogram function, the most basic detection function estimation only requires a numeric vector of distances.
+The most basic detection function estimation only requires a numeric vector of distances.
 Here we demonstrate how to use \pkg{Distance} to fit detection functions, perform model checking and selection, and estimate abundance.
@@
 
-The distribution of the observed distances is a product of detectability [sometimes referred to as "perception bias"; @Marsh:1989ho] and the distribution of the animals (observed or unobserved) with respect to the line or point.
+The distribution of the observed distances is a product of detectability [sometimes referred to as "perception bias"; @Marsh:1989ho] and the distribution of the animals with respect to the line or point.
 Our survey design allows us to assume a distribution for the animals with respect to the sampler.
@@
 For line transect studies we assume that objects are uniformly distributed with respect to the line (i.e., the number of animals available for detection is the same at all distances).
-For point transect surveys the survey area increases linearly with radial distance, implying a triangular distribution with respect to the point.
+For point transect surveys, area increases linearly with radial distance, implying a triangular distribution with respect to the point.
 Figure \ref{fig:detectiondist} shows how these distributions, when combined with a detection function, give rise to the observed distribution of recorded distances.
@@
 Figure \ref{fig:pointslines} shows simulated sampling of a population of 500 objects using line and point transects and their corresponding histograms of observed detection distances.
-Note that for the purposes of distance sampling an object may either refer to an individual in a population or a cluster of individuals.
+Note that for the purposes of distance sampling an "object" may either refer to an individual in a population or a cluster of individuals.
 
@@
 Good survey design is essential to produce reliable density and abundance estimates from distance sampling data.
-Survey design is beyond the scope of this article but we refer readers to @Buckland:2001vm [Chapter 7] and @buckland2015distance [Chapter 2] for introductory information; @Buckland:2004ts [Chapter 7] contains information on automated survey design using geographical information systems (GIS); @Thomas:2007wz gives an example of designing a distance sampling survey in a complex region.
+Survey design is beyond the scope of this article but we refer readers to @Buckland:2001vm [, Chapter 7] and @buckland2015distance [, Chapter 2] for introductory information; @Strindberg:2004vr contains information on automated survey design using geographical information systems (GIS); @Thomas:2007wz gives an example of designing a distance sampling survey in a complex region.
 We also note that Distance for Windows includes a GIS survey design tool [@Thomas:2010cf].
@@
 These data sets have been chosen as they represent typical data seen in practice.
-The below example analyses are not intended to serve as guidelines for practice, but to demonstrate features of the software.
-Extensive practical advice on approaches to analysis is given in @Thomas:2010cf.
+The below example analyses are not intended to serve as guidelines, but to demonstrate features of the software.
+Practical advice on approaches to analysis is given in @Thomas:2010cf.
 
@@
 These data consist of 99 observations that are stratified based on location (near or distant from ice edge) and effort data (transect lengths).
-Further details on the survey are available in @Branch:2001ua (simulated data is based on "1992/93 Area III" therein).
+Further details on the survey are available in @Branch:2001ua (data is simulated based on the design used for "1992/93 Area III" therein).
 
@@
 The \code{data.frame} must contain a column named \code{distance} (containing the observed distances) and additional named columns for any covariates that may affect detectability (for example \code{observer} or \code{seastate}).
-The column name \code{size} is reserved for the cluster sizes (sometimes referred to as group sizes) for when the detection distances relate to observations of clusters rather than individuals (see @Buckland:2001vm, Section 1.4.3 for more on defining clusters and "Extensions" below for one approach to dealing with uncertain cluster size).
+The column name \code{size} is reserved for the cluster sizes (sometimes referred to as group sizes) in this case each row represents an observation of a cluster rather than individual (see @Buckland:2001vm, Section 1.4.3 for more on defining clusters and "Extensions" below for one approach to dealing with uncertain cluster size).
 Additional reserved names include \code{object} and \code{detected}, these are not required for conventional distance sampling and should be avoided (see "Extensions" for an explanation of their use).
@@
 * *Non-increasing*: we do not think that observers should be more likely to see distant objects than those nearer the transect.
-If this occurs, it usually indicates an issue with survey design or field procedures (for example that the distribution of objects with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this [@Miller:2015hw; @Marques:2010he; @Marques:2012fy].
+If this occurs, it usually indicates an issue with survey design or field procedures (for example that the distribution of objects with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this [@Marques:2010he; @Marques:2012fy; @Miller:2015hw].
 * *Model robust*: models should be flexible enough to fit many different shapes.
@@
 
-There are many factors that can affect the probability of detecting an object: observer skill, cluster/group size (if objects occur in groups), the vessel or platform used, sea state, other weather conditions, time of day and more.
+There are many factors that can affect the probability of detecting an object: observer skill, cluster size (if objects occur in clusters), the vessel or platform used, sea state, other weather conditions, time of day and more.
 In \pkg{Distance} we assume that these variables affect detection only via the scale of the detection function (and do not affect the shape).
@@
 Including covariates has an important implication for our calculation of detectability.
-Because we do not know the true distribution of the covariates, we can either model the distribution of the covariates and integrate the covariates out of the joint density, or we can calculate the probability of detection conditional on the observed values of the covariates [@Marques:2003vb].
+We do not know the true distribution of the covariates, we therefore must either: $(i)$ model the distribution of the covariates and integrate the covariates out of the joint density (thus making strong assumptions about their distribution), or $(ii)$ calculate the probability of detection conditional on the observed values of the covariates [@Marques:2003vb].
 We opt for the latter:
@@
 ```
-Note that when there are no covariates in the model, \code{ds} will add adjustment terms to the model until there is no improvements in AIC and print the selection steps.
+Note that when there are no covariates in the model, \code{ds} will add adjustment terms to the model until there is no improvement in AIC.
 
 Figure \ref{fig:minkeamakihi} (left panel) shows the result of calling \code{plot} on the resulting model object.
-We can also call \code{summary} on the model object to get summary information about the fitted model (we postpone this to the next section).
+We can also call \code{summary} on the model object to get summary information about the fitted model (we postpone this until the next section).
 
@@
 We will compare these two models further in the following section.
-
 ```{r amakihi-hr-obs-mas, warning=FALSE}
@@
 ```
-
 As with the minke whale model, we can plot the resulting models (Figure \ref{fig:minkeamakihi}, middle and right panels).
@@
 For the `amakihi_hr_obs` model we see fairly clear levels of the observer covariate in the points.
-Looking at the right panel of Figure \ref{fig:minkeamakihi}, this is less clear when adding minutes after sunrise to the model.
+Looking at the right panel of Figure \ref{fig:minkeamakihi}, this is less clear when adding minutes after sunrise (a continuous covariates) to the model.
 
@@
 Centre, hazard-rate with observer as a covariate; right, hazard-rate model with observer and minutes after sunrise as covariates.
-Points indicate probability of detection for a given observation (given that observations covariate values) and lines indicate the average detection function.\\label{fig:minkeamakihi}"}
+Points indicate probability of detection for a given observation (given that observation's covariate values) and lines indicate the average detection function.\\label{fig:minkeamakihi}"}
 par(mfrow=c(1, 3))
@@
 This summary information includes details of the data and model specification, as well as the values of the coefficients ($\beta_j$) and their uncertainties, an "average" value for the detectability (see "Estimating abundance and variance" for details on how this is calculated) and its uncertainty.
-The final line gives an estimate of abundance in the area covered by the survey (see the next section).
+The final line gives an estimate of abundance in the area covered by the survey (see the next section; though note this estimate does not take into account cluster size).
 
@@
 The Q-Q plots in \pkg{Distance} plot a point for every observation.
-The EDF is is the proportion of points that have been observed at a distance equal to or less than the distance of that observation.
+The EDF is the proportion of points that have been observed at a distance equal to or less than the distance of that observation.
 The CDF is calculated from the fitted detection function as the probability of observing an object at a distance less than or equal to that of the given observation.
@@
 As it takes into account more information and is therefore more powerful, the Cramér-von Mises is generally preferred.
-A significant result (usually $P < 0.05$) from either test gives evidence against the null hypothesis (that the data arose from the fitted model), suggesting that the model does not fit the data well.
+A significant result from either test gives evidence against the null hypothesis (that the data arose from the fitted model), suggesting that the model does not fit the data well.
 
@@
       caption="Summary for the detection function models fitted to the amakihi data.
-``C-vM'' stands for Cramér-von Mises, $P_a$ indicates average detectability (see ``Estimating abundance and variance''), se indicates standard error.
+``C-vM'' stands for Cramér-von Mises, $P_a$ is average detectability (see ``Estimating abundance and variance''), se is standard error.
 Models are sorted according to AIC.\\label{tab:amakihi}")
@@
 \end{equation}
-where $s_i$ are the sizes of the observed groups of objects, which is equal to 1 if objects only occur singly [@Borchers:2004wr].
+where $s_i$ are the sizes of the observed clusters of objects, which are all equal to 1 if objects only occur singly [@Borchers:2004wr].
 @thompson2012sampling is the canonical reference to this type of estimator.
-Intuitively, we can think of the estimates of detectability ($\hat{p}(\mathbf{z}_i)$) as "inflating" the group sizes ($s_i$), we then sum over the detections ($i$) to obtain the abundance estimate.
-For models that do not include covariates, $\hat{p}(\mathbf{z}_i)$ is equal for all $i$, so this is equivalent to summing the groups and inflating that sum by dividing through by the corresponding $\hat{p} (=\hat{p}(\mathbf{z}_i) \forall i)$.
+Intuitively, we can think of the estimates of detectability ($\hat{p}(\mathbf{z}_i)$) as "inflating" the cluster sizes ($s_i$), we then sum over the detections ($i$) to obtain the abundance estimate.
+For models that do not include covariates, $\hat{p}(\mathbf{z}_i)$ is equal for all $i$, so this is equivalent to summing the clusters and inflating that sum by dividing through by the corresponding $\hat{p} (=\hat{p}(\mathbf{z}_i) \forall i)$.
 
@@
 1. *Detection function*: Uncertainty in parameter ($\boldsymbol{\theta}$) estimation.
-2. *Encounter rate*: Sampling variability arising from differences in the rate of observations between transects.
+2. *Encounter rate*: Sampling variability arising from differences in the number of observations per transect.
 
@@
 
-1.
-\code{Summary statistics}: giving the areas, covered areas, effort, number of observations, number of transects, encounter rate, its standard error and coefficient of variation for each stratum, then a total for the whole study area.
-2.
-\code{Abundance}: giving estimates, standard errors, coefficients of variation, lower and upper confidence intervals and finally the degrees of freedom for each stratum's abundance estimate, then a total for the whole study area.
+1. \code{Summary statistics}: giving the areas, covered areas, effort, number of observations, number of transects, encounter rate, its standard error and coefficient of variation for each stratum. 
+2. \code{Abundance}: giving estimates, standard errors, coefficients of variation, lower and upper confidence intervals and finally the degrees of freedom for each stratum's abundance estimate.
 3. \code{Density}: lists the same statistics as \code{Abundance} but for a density estimate.
+
+In each table the bottom row gives a total for the whole study area.
 
@@
 We would particularly like to thank Steve Buckland, David Borchers, Tiago Marques, Jon Bishop and Lorenzo Milazzo for their contributions.
-We thank Tiago Marques a second time for helpful comments on an early version of the paper.
+We thank Tiago Marques a second time for helpful comments on an early version of the paper and Colin Beale who suggested the "flatfile" data format.
 We also thank Ken Burnham and David Anderson for fundamental contributions to the early development of these methods.

### 2016-03-30 paper.Rmd
Jeff comments

@@
     address: >
-      National Marine Mammal Laboratory
+      Marine Mammal Laboratory
       Alaska Fisheries Science Center
@@
 
-output: rticles::jss_article
+output:
+  rticles::jss_article:
+    citation_package: natbib
 bibliography: jstatsoft.bib
@@
 set.seed(15) # same results every time
-library("mgcv") # for inSide
+library("mgcv", quietly=TRUE) # for inSide
 
@@
 The columns \code{distance}, \code{Area} and (in the case of line transects) \code{Effort} have associated units (though these are not explicitly included in a \pkg{Distance} analysis).
-We recommended that data in these columns are converted to SI units before starting any analysis to ensure that resulting abundance and density estimates have sensible units.
+We recommend that data in these columns are converted to SI units before starting any analysis to ensure that resulting abundance and density estimates have sensible units.
 For example, if distances from a line transect survey are recorded in metres, the \code{Effort} columns should contain line lengths also in metres and the \code{Area} column gives the stratum areas in square metres.
@@
 In the authors' experience, it is often the case that models with similar AICs will have similar estimates probabilities of detection, so in practice there is little difference in selecting between these models.
+It is important to note that comparing AICs between models with different truncations is not appropriate, as models with different truncation use different data.
 
 ```{r summary-table, results="asis", echo=FALSE}
-library("knitr")
 kable(summarize_ds_models(amakihi_hn, amakihi_hr_obs, amakihi_hr_obs_mas),
@@
 
-We combine these two variances by noting that squared coefficients of variation (approximately) add [@Goodman:1960dd] (often referred to as "the delta method").
+We combine these two variances by noting that squared coefficients of variation (approximately) add [often referred to as "the delta method"; @seber2002estimation].
 
@@
 
-The examples of applications for which distance sampling has been used are many.
+Distance sampling has been applied in a wide variety of situations.
 Objects being detected need not be animals; plants [@Buckland:2007kl], beer cans [@Otto:1990gs], elephant dung [@NchPlu01] and bricks on a lake bottom [@Ber90] have been subjects of distance sampling investigations.
@@
 In combination with tools such as \pkg{knitr} and \pkg{rmarkdown} [@rmarkdown-pkg], the helper functions in \pkg{Distance} provide a useful set of tools to perform reproducible analyses of wildlife abundance for both managers and ecologists.
+We also direct readers' attention to \pkg{DsShiny} [@dsshiny-pkg], a package that builds on \pkg{Shiny} [@shiny-pkg] and \pkg{mrds} to allow users to build and fit models in a graphical interface, with output to RMarkdown.
+
 \proglang{R} and its extension packages provide many tools exploratory data analysis that can be useful for a distance sampling analysis.

### 2016-03-31 paper.Rmd
code formatting stuff

@@
   - name: Jeffrey L Laake
-    affiliation: National Oceanic and Atmospheric Administration
+    affiliation: Marine Mammal Laboratory
     address: >
-      Marine Mammal Laboratory
       Alaska Fisheries Science Center
+      National Oceanic and Atmospheric Administration
       7600 Sand Point Way N.E., Seattle, WA 98115, USA
@@
 ```{r minke-hn}
-minke_hn <- ds(minke, truncation=1.5)
+minke_hn <- ds(minke, truncation = 1.5)
 ```
@@
 ```{r minke-hr-cos}
-minke_hrcos <- ds(minke, truncation=1.5, key="hr")
+minke_hrcos <- ds(minke, truncation = 1.5, key = "hr")
 ```
@@
 ```{r minke-unif-cos}
-minke_unifcos <- ds(minke, truncation=1.5, key="unif", adjustment="cos", order=c(1,2))
+minke_unifcos <- ds(minke, truncation = 1.5, key = "unif",
+                    adjustment = "cos", order = c(1, 2))
 ```
@@
 ```{r amakihi-hr-obs}
-amakihi_hr_obs <- ds(amakihi, truncation=82.5, transect="point",
-                     key="hr", formula=~obs)
+amakihi_hr_obs <- ds(amakihi, truncation = 82.5, transect = "point",
+                     key = "hr", formula = ~obs)
 ```
@@
 ```{r amakihi-hr-obs-mas, warning=FALSE}
-amakihi_hr_obs_mas <- ds(amakihi, truncation=82.5, transect="point",
-                         key="hr", formula=~obs+mas)
+amakihi_hr_obs_mas <- ds(amakihi, truncation = 82.5, transect = "point",
+                         key = "hr", formula = ~obs + mas)
 ```
@@
 ```{r amakihi-gof, fig.keep="none"}
-amakihi_hn <- ds(amakihi, truncation=82.5, transect="point", key="hn", adjustment=NULL)
+amakihi_hn <- ds(amakihi, truncation = 82.5, transect = "point",
+                 key = "hn", adjustment = NULL)
 gof_ds(amakihi_hn)
@@
 
-```{r, amakihi-qq-comp, fig.width=6, fig.height=3, fig.cap="Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates; left) and hazard-rate model with observer and minutes after sunrise (right) for the amakihi data.\\label{amakihi-qq}", echo=FALSE, results="hide"}
+```{r amakihi-qq-comp, fig.width=6, fig.height=3, fig.cap="Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates; left) and hazard-rate model with observer and minutes after sunrise (right) for the amakihi data.\\label{amakihi-qq}", echo=FALSE, results="hide"}
 par(mfrow=c(1,2))

### 2016-03-31 paper.Rmd
Fortran not FORTRAN

@@
 Since this time it has evolved to include various design and analysis features [@Thomas:2010cf].
-Distance for Windows versions 5 onwards have included \proglang{R} [@rcore]  packages as the analysis engines providing additional, more complex analysis options than those offered by the original (FORTRAN) code.
+Distance for Windows versions 5 onwards have included \proglang{R} [@rcore]  packages as the analysis engines providing additional, more complex analysis options than those offered by the original (Fortran) code.
 

### 2016-03-31 paper.tex
remove temp files add to .gitignore
stop ignoring the paper.tex file now

@@
 \author{
-David L Miller\\University of St Andrews \And Eric Rexstad\\University of St Andrews \And Jeffrey L Laake\\National Marine Mammal Laboratory
+David L Miller\\Integrated Statistics and University of St Andrews \AND Eric Rexstad\\University of St Andrews \And Len Thomas\\University of St Andrews \AND Laura Marshall\\University of St Andrews \And Jeffrey L Laake\\Marine Mammal Laboratory
 }
 \title{Distance Sampling in \proglang{R}}
-\Keywords{distance sampling, abundance estimation, line transects, point transects, \proglang{R}}
+\Keywords{distance sampling, abundance estimation, line transect, point transect, detection function, Horvitz-Thompson, \proglang{R}, Distance}
 
 \Abstract{
-The abstract of the article.
+Estimating the abundance and spatial distribution of animal and plant
+populations is essential for conservation and management. We introduce
+the \proglang{R} package \pkg{Distance} that implements distance
+sampling methods to estimate abundance. We describe how users can obtain
+estimates of abundance (and density) using the package as well
+documenting the links it provides with other more specialized
+\proglang{R} packages. We also demonstrate how \pkg{Distance} provides a
+migration pathway from previous software, thereby allowing us to deliver
+cutting-edge methods to the users more quickly.
 }
 
-\Plainauthor{David L Miller, Eric Rexstad, Jeffrey L Laake}
+\Plainauthor{David L Miller, Eric Rexstad, Len Thomas, Laura Marshall, Jeffrey L Laake}
 \Plaintitle{Distance Sampling in R}
-\Plainkeywords{distance sampling, abundance estimation, line transects, point transects, R}
+\Plainkeywords{distance sampling, abundance estimation, line transect, point transecs, detection function, Horvitz-Thompson, R, Distance}
 
@@
     David L Miller\\
-  University of St Andrews\\
-  Centre for Research into Ecological and Environmental Modelling, The
-  Observatory, St Andrews, Fife KY16 9LZ, Scotland\\
+  Integrated Statistics and University of St Andrews\\
+  16 Sumner Street, Woods Hole, MA 02543, USA and Centre for Research into
+  Ecological and Environmental Modelling, The Observatory, St Andrews,
+  Fife KY16 9LZ, Scotland\\
   E-mail: \href{mailto:dave@ninepointeightone.net}{\nolinkurl{dave@ninepointeightone.net}}\\
@@
 
+      Len Thomas\\
+  University of St Andrews\\
+  Centre for Research into Ecological and Environmental Modelling, The
+  Observatory, St Andrews, Fife KY16 9LZ, Scotland\\
+  E-mail: \href{mailto:len.thomas@st-andrews.ac.uk}{\nolinkurl{len.thomas@st-andrews.ac.uk}}\\
+  URL: \url{http://lenthomas.org}\\~\\
+      Laura Marshall\\
+  University of St Andrews\\
+  Centre for Research into Ecological and Environmental Modelling, The
+  Observatory, St Andrews, Fife KY16 9LZ, Scotland\\
+  E-mail: \href{mailto:lhm@st-andrews.ac.uk}{\nolinkurl{lhm@st-andrews.ac.uk}}\\
+
       Jeffrey L Laake\\
-  National Marine Mammal Laboratory\\
-  Alaska Fisheries Science Center 7600 Sand Point Way N.E., Seattle, WA
-  98115, USA\\
+  Marine Mammal Laboratory\\
+  Alaska Fisheries Science Center National Oceanic and Atmospheric
+  Administration 7600 Sand Point Way N.E., Seattle, WA 98115, USA\\
   E-mail: \href{mailto:Jeff.Laake@noaa.gov}{\nolinkurl{Jeff.Laake@noaa.gov}}\\
@@
 \usepackage{booktabs}
+\usepackage[colorinlistoftodos, color=green!40, textwidth=60, textsize=small]{todonotes}
 
@@
 
-Distance sampling (Buckland et al. 2001; Buckland et al. 2004)
-encompasses a suite of field methods and statistical models used to
-estimate the abundance of biological populations. Distance sampling
-field procedure can be thought of as an extension of plot sampling,
-where we wish to take into account the decreasing probability of
-detecting objects at increasing distance from the sampler. We do this by
-building a model for detectability and use quantities calculated from
-the detection function to adjust the observed counts to obtain an
-estimate of abundance.
-
-For many years distance sampling analyses have been available via the
-Windows program Distance (or ``DISTANCE''); for clarity henceforth
-``Distance for Windows'' Thomas et al. (2010)). From version 5 of
-Distance for Windows, R packages have been included to perform
-particular analyses (CITE Distance user manual). This paper shows how to
-fit detection functions, perform model checking and selection and
-estimate abundance in \proglang{R} using the package \pkg{Distance}.
-\pkg{Distance} is a wrapper package around the more complex (and more
-powerful) \pkg{mrds} and offers a subset of the analyses possible with
-that package.
+Distance sampling
+\citep{Buckland:2001vm, Buckland:2004ts, buckland2015distance}
+encompasses a suite of methods used to estimate the density and/or
+abundance of biological populations. Distance sampling can be thought of
+as an extension of plot sampling. Plot sampling involves selecting a
+number of plots (small areas) at random within the study area and
+counting the objects of interest that are contained within each plot. By
+selecting the plots at random we can assume that the density of objects
+in the plots is representative of the study area as a whole. One of the
+key assumptions of plot sampling is that all objects within each of the
+plots are counted. Distance sampling relaxes this assumption in that
+observers are no longer required to detect (either by eye, video/audio
+recording etc) and count everything within selected plots. While plot
+sampling techniques are adequate for static populations occurring at
+high density they are inefficient for more sparsely distributed
+populations. Distance sampling provides a more efficient solution in
+such circumstances.
+
+Conventional distance sampling assumes the observer is located either at
+a point or moving along a line and will observe all objects that occur
+at the point or on the line. The further away an object is from the
+point or line (also known as the sampler or transect) the less likely it
+is that the observer will see it. We can use the distances to each of
+the detected objects from the line or point to build a model of the
+probability of detection given distance from the sampler -- the
+\emph{detection function}. The detection function can be used to infer
+how many objects were missed and thereby produce estimates of density
+and/or abundance.
+
+The Windows program Distance (or ``DISTANCE''; for clarity henceforth
+``Distance for Windows'') can be used to fit detection functions to
+distance sampling data. It was first released (versions 1.0 - 3.0;
+principally programmed by Jeff Laake while working at the National
+Marine Mammal Laboratory) as a console-based application (this in turn
+was based on earlier software TRANSECT, \citet{Burnham:1980wz} and
+algorithms developed in \citet{Buckland:1992fa}), before the first
+graphical interface (Distance for Windows 3.5) was released in November
+1998. Since this time it has evolved to include various design and
+analysis features \citep{Thomas:2010cf}. Distance for Windows versions 5
+onwards have included \proglang{R} \citep{rcore} packages as the
+analysis engines providing additional, more complex analysis options
+than those offered by the original (FORTRAN) code.
+
+As Distance for Windows becomes increasingly reliant on analyses
+performed in \proglang{R} and many new methods are being developed, we
+are encouraging the use of our \proglang{R} packages directly.
+\proglang{R} provides a huge variety of functionality for data
+exploration and reproducible research, much more than is possible in
+Distance for Windows.
+
+Until now those wishing to use our \proglang{R} packages for straight
+forward distance sampling analyses would have had to negotiate the
+package \pkg{mrds} \citep{mrds-pkg} designed for mark-recapture distance
+sampling \citep{Burt:2014gu}, requiring a complex data structure to
+perform analyses. \pkg{Distance} is a wrapper package around \pkg{mrds}
+making it easier to get started with basic distance sampling analyses in
+\proglang{R}. The most basic detection function estimation only requires
+a numeric vector of distances. Here we demonstrate how to use
+\pkg{Distance} to fit detection functions, perform model checking and
+selection, and estimate abundance.
 
@@
 
-Census-type surveys (quadrat or strip transects assuming perfect
-detectability within strips or quadrats) are inefficient (requiring
-considerable field effort) and we should expect that not all objects
-(animals, plants, dung, etc) can be observed. Accounting for imperfect
-detectability is an important consideration when obtaining accurate
-estimates of abundance (Lahoz-Monfort, Guillera-Arroita, and Wintle
-2013). Using the extra information gained by recording distance from the
-sampler to the observation, it is possible to model detectability.
-Because we expect detectability to decrease with increasing distance
-from the sampler, we model detectability as a function of distance (plus
-perhaps other covariates, see below). We use the probabilities of
-detection calculated from the detection function to estimate an
-abundance for the area covered by the survey, which we then scale-up to
-the study region of interest.
-
-Distance sampling surveys are conducted using two transect types: line
-and point transects. In line transect sampling observers walk (or fly,
-sail, etc) lines observing objects and recording the distances to the
-line; whereas in point transect sampling observers remain stationary at
-a location and record distances from that point. Field methods are
-chosen to be suitable to species and habitat constraints (Buckland et
-al. 2015).
+The distribution of the observed distances is a product of detectability
+\citep[sometimes referred to as ``perception bias'';][]{Marsh:1989ho}
+and the distribution of the animals with respect to the line or point.
+Our survey design allows us to assume a distribution for the animals
+with respect to the sampler.
+
+For line transect studies we assume that objects are uniformly
+distributed with respect to the line (i.e., the number of animals
+available for detection is the same at all distances). For point
+transect surveys, area increases linearly with radial distance, implying
+a triangular distribution with respect to the point. Figure
+\ref{fig:detectiondist} shows how these distributions, when combined
+with a detection function, give rise to the observed distribution of
+recorded distances.
 
@@
 
+{\centering \includegraphics{paper_files/figure-latex/detectdists-1}
+
+}
+
+\caption{Panels show an example detection function (left), the pdf of object distances (middle) and the resulting pdf of detection distances (right) for line transects (top row) and point transects (bottom row).
+The pdf of observed detection distances in the right hand plots are obtained by multiplying the detection function by the pdf of object distances and rescaling.
+In this example, detection probability becomes effectively zero at 500 (distances shown on $x$-axis are arbitrary).
+\label{fig:detectiondist}}\label{fig:detectdists}
+\end{figure}
+\end{CodeChunk}
+
+Figure \ref{fig:pointslines} shows simulated sampling of a population of
+500 objects using line and point transects and their corresponding
+histograms of observed detection distances. Note that for the purposes
+of distance sampling an ``object'' may either refer to an individual in
+a population or a cluster of individuals.
+
+\begin{CodeChunk}
+\begin{CodeOutput}
+Warning: package 'nlme' was built under R version 3.2.4
+\end{CodeOutput}
+\begin{figure}
+
 {\centering \includegraphics{paper_files/figure-latex/points-and-lines-1}
@@
 
-\caption{Left side plots show an example of a survey of an area containing a population of 500 objects, blue indicates sampler placement (top lines, bottom points) and red dots indicate detected individuals.
+\caption{Left side plots show an example of a survey of a (1x1) area containing a population of 500 objects; blue dashed lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals.
 The right side of the figure shows histograms of observed distances (again, lines top and points bottom).\label{fig:pointslines}}\label{fig:points-and-lines}
@@
 
-For both points and lines, given the geometry of the sampler (see
-``Detection functions''), the histogram of distances should show a
-decreasing number of observations with increasing distance from the
-sampler. For line transects we expect objects to be distributed
-uniformly with respect to distance from the line with detectability
-causing the number of detections to decrease. For point transects, we
-recognise that as distance from the point increases, the area of the
-circle encompassed increases with distance squared; hence the number of
-objects available to be detected is a linearly increasing function of
-distance from the point.
-
-Using this histogram we can crudely estimate the drop-off in
-detectability by eye by tracing a line that approximates the tops of the
-histogram bars -- this is the detection function. \pkg{Distance}
-estimates the parameters for a fixed-form detection function using
-maximum likelihood estimation. We address possible models below.
-
-Figure \ref{fig:pointslines} shows sampling of a population of 500
-individuals using line and point transects (left column) and their
-corresponding histograms (right column).
-
-\subsection{Data}\label{data}
-
-We demonstrate \pkg{Distance} using two data sets: one line transect and
-one point transect. These data sets have been chosen to be
-representative of data seen in practice.
+Good survey design is essential to produce reliable density and
+abundance estimates from distance sampling data. Survey design is beyond
+the scope of this article but we refer readers to \citet[Chapter
+7]{Buckland:2001vm} and \citet[Chapter 2]{buckland2015distance} for
+introductory information; \citet{Strindberg:2004vr} contains information
+on automated survey design using geographical information systems (GIS);
+\citet{Thomas:2007wz} gives an example of designing a distance sampling
+survey in a complex region. We also note that Distance for Windows
+includes a GIS survey design tool \citep{Thomas:2010cf}.
+
+\pkg{Distance} provides a selection of candidate functions to describe
+the probability of detection and estimates the associated parameters
+using maximum likelihood estimation. The probability of detecting an
+object may not only depend on how far it is from the observer but also
+on other factors such as weather conditions, ground cover, cluster size
+etc. The \pkg{Distance} package also allows the incorporation of such
+covariates into the detection function allowing the detection function
+scale parameter to vary based on these covariates.
+
+Having estimated the detection function's parameters, one can then
+integrate out distance from the function (as the detection function
+describes the probability of detection \emph{given} distance) to get an
+``average'' probability of detection, which can be used to correct the
+observed counts. Summing the corrected counts gives an estimate of
+abundance in the area covered by surveys, which can be multiplied up to
+the total study area.
+
+In addition to randomly placed samplers distance sampling relies on
+three other main assumptions. Firstly, all objects directly on the
+transect line or point (i.e.~those at zero distance) are detected (see
+``Extensions'' for methods to deal with the situation when this is not
+possible). Secondly, objects are stationary or detected prior to any
+movement. Thirdly, distance to the object must be measured accurately,
+or the observation allocated to the correct distance bin for grouped
+data. Depending on the survey species some of these assumptions may be
+more difficult to meet than others. Further information on field methods
+to help meet these assumptions can be found in \citet{Buckland:2001vm}
+and \citet{buckland2015distance}.
+
+The rest of the paper has the following structure: we describe data
+formatting for \pkg{Distance}; candidate detection function models are
+described in terms of formulation and fitted examples in \proglang{R}.
+We then show how to perform model checking, goodness of fit and model
+selection. We go on to show how to estimate abundance, including
+stratified estimates of abundance. The final two sections of the article
+look at extensions (both in terms of methodology and software) and put
+the package in a broader context amongst other \proglang{R} packages
+used for estimating the abundance of biological populations from
+distance sampling data.
+
+\section{Data}\label{data}
+
+We introduce two example analyses performed in \pkg{Distance}: one line
+transect and one point transect. These data sets have been chosen as
+they represent typical data seen in practice. The below example analyses
+are not intended to serve as guidelines, but to demonstrate features of
+the software. Practical advice on approaches to analysis is given in
+\citet{Thomas:2010cf}.
 
@@
 
-The line transect data is based on a survey of Antarctic minke whales
-(\emph{Balaenoptera bonaerensis}). The data are simulated from models
-fitted to data from the International Whaling Commission's International
-Decade of Cetacean Research Southern Ocean Whale and Ecosystem Research
-(IWC IDCR-SOWER) programme 1992-1993 austral summer surveys (Branch and
-Butterworth 2001). They consist of 99 observations and include
-geographical strata (near or distant from ice edge) and transect lengths
-on the effort expended.
+The line transect data have been simulated from models fitted to
+Antarctic minke whale (\emph{Balaenoptera bonaerensis}) data. These data
+were collected as part of the International Whaling Commission's
+International Decade of Cetacean Research Southern Ocean Whale and
+Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer
+surveys. These data consist of 99 observations that are stratified based
+on location (near or distant from ice edge) and effort data (transect
+lengths). Further details on the survey are available in
+\citet{Branch:2001ua} (data is simulated based on the design used for
+``1992/93 Area III'' therein).
 
@@
 sunrise (transformed to minutes (continuous) or hours (factor)
-covariates). Data are analysed comprehensively in Marques et al. (2007).
-
-\subsection{The rest of the paper}\label{the-rest-of-the-paper}
-
-The rest of the paper has this structure: we describe data organisation
-for use with \pkg{Distance}; models for the detection function are
-described in terms of formulation and examples of fitting in
-\proglang{R}. We then examine model checking, goodness of fit and model
-selection. Having illustrated how to obtain a good detection function,
-we show how to estimate abundance using that model, including
-stratification. The final two sections of the article look at extensions
-(both in terms of methodology and software) and put the package in a
-broader context amongst other software packages used for estimating the
-abundance of biological populations.
-
-\section{Data setup}\label{data-setup}
-
-The two example data sets used here are distributed with \pkg{Distance}
-so readers can reproduce our results. However, data will be collected in
-the field and formatted correctly for use with \pkg{Distance}. The
-package allows for a flexible format for data input ranging from very
-simple to complex:
+covariates). Data are analysed comprehensively in
+\citet{Marques:2007ey}.
+
+\subsection{Data setup}\label{data-setup}
+
+Generally, data collected in the field will require some formatting
+before use with \pkg{Distance}, though there are a range of possible
+formats, dependent on the model specification and the output required:
 
 \begin{itemize}
-\itemsep1pt\parskip0pt\parsep0pt
+\tightlist
 \item
   In the simplest case, where the objective is to estimate a detection
-  function, all that is needed is a vector of distances.
+  function and exact distances are collected, all that is required is a
+  numeric vector of distances.
 \item
   To include additional covariates into the detection function (see
-  ``Detection functions'') we use a \code{data.frame}. The data.frame
-  contains a column called \code{distance} and additional named columns
-  for covariates potentially useful in modelling detectability (for
-  example \code{observer} or \code{seastate}). Some column names are
-  reserved: \code{object} for an observation identifier (see
-  ``Extensions''), \code{size} for group or cluster size (see
-  ``Detection functions'' and ``Abundance and variance estimation''),
-  \code{detected} for whether an observation was detected (see
-  ``Extensions'') and the columns described in the next bullet.
+  ``Detection functions'') a \code{data.frame} is required. Each row of
+  the \code{data.frame} contains the data on one observation. The
+  \code{data.frame} must contain a column named \code{distance}
+  (containing the observed distances) and additional named columns for
+  any covariates that may affect detectability (for example
+  \code{observer} or \code{seastate}). The column name \code{size} is
+  reserved for the cluster sizes (sometimes referred to as group sizes)
+  in this case each row represents an observation of a cluster rather
+  than individual (see \citet{Buckland:2001vm}, Section 1.4.3 for more
+  on defining clusters and ``Extensions'' below for one approach to
+  dealing with uncertain cluster size). Additional reserved names
+  include \code{object} and \code{detected}, these are not required for
+  conventional distance sampling and should be avoided (see
+  ``Extensions'' for an explanation of their use).
 \item
-  To estimate abundance beyond area covered by expended survey effort,
-  additional information is required. This consists of information about
-  which transect and occasion the observation was made (the
-  \code{Sample.Label}), a column named \code{Effort} which gives the
-  effort associated with that sample (for lines their length and for
-  points the number of times that point was visited), the stratum the
-  sample was located in (this may have any name and may be from pre- or
-  post-survey stratification, see ``Estimating abundance and variance'')
-  and strata areas (which has the same name as the stratum column,
-  appended with \code{.Area}). We refer to this data format as
-  ``flatfile'' where all information is contained in one table.
+  To estimate density or to estimate abundance beyond the sampled area,
+  additional information is required. Additional columns should be
+  included in the \code{data.frame} specifying: \code{Sample.Label}, the
+  ID of the transect; \code{Effort}, transect effort (for lines their
+  length and for points the number of times that point was visited);
+  \code{Region.Label}, the stratum containing the transect (which may be
+  from pre- or post-survey stratification, see ``Estimating abundance
+  and variance''); \code{Area}, the area of the strata. Transects which
+  were surveyed but have no observations must be included in the data
+  set with \code{NA} for \code{distance} and any other covariates. We
+  refer to this data format (where all information is contained in one
+  table) as ``flatfile'' as it can be easily created in a single
+  spreadsheet.
 \end{itemize}
@@
 As we will see in ``Extensions'', further information is also required
-when using more complex models.
+for fitting more complex models.
+
+It is also possible to use distances collected in intervals (``binned''
+or ``grouped'' data, as opposed to ``exact'' distances) to model the
+detection function. In this case the column \code{distance} is replaced
+by two columns \code{distbegin} and \code{distend} referring to the
+distance interval start and end cutpoints. More information on binned
+data is included in \citet{Buckland:2001vm} sections 4.5 and 7.4.1.2.
+
+The columns \code{distance}, \code{Area} and (in the case of line
+transects) \code{Effort} have associated units (though these are not
+explicitly included in a \pkg{Distance} analysis). We recommend that
+data in these columns are converted to SI units before starting any
+analysis to ensure that resulting abundance and density estimates have
+sensible units. For example, if distances from a line transect survey
+are recorded in metres, the \code{Effort} columns should contain line
+lengths also in metres and the \code{Area} column gives the stratum
+areas in square metres. This would lead to density estimates of animals
+per square metre.
 
@@
 \begin{CodeInput}
-library(Distance)
+library("Distance")
 head(minke)
@@
 
-We will explore below the effects of including effort data during
-analysis.
+We will explore the consequences of including effort and stratum data in
+the analysis below.
 
@@
 
-The detection function describes the relationship between observed
-distances and probability of detection. The detection function itself
-models the probability
+The detection function models the probability
 \(\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)\)
 and is usually denoted \(g(y; \boldsymbol{\theta})\) where \(y\) is
-distance (either from a line or point) and \(\boldsymbol{\theta}\) is a
-vector of parameters to be estimated. Our goal is to estimate an
-\emph{average probability of detection} (\(p\), average in the sense of
-an average over distances), so we must integrate out distance (\(y\))
-from the detection function: \[
+distance (from a line or point) and \(\boldsymbol{\theta}\) is a vector
+of parameters to be estimated. Our goal is to estimate an \emph{average
+probability of detection} (\(p\), average in the sense of an average
+over distance from \(0\) to truncation distance \(w\)), so we must
+integrate out distance (\(y\)) from the detection function: \[
 p = \int_0^w \pi(y) g(y; \boldsymbol{\theta}) dy
 \] where \(\pi(y)\) describes the distribution of objects with respect
-to the sampler and \(\pi(x)=1/w\) for line transects and
+to the sampler; \(\pi(x)=1/w\) for line transects and
 \(\pi(r)=\frac{2r}{w^2}\) for point transects, taking into account the
-geometry of the sampler (Buckland et al. 2001, Chapter 3) (letting \(x\)
-denote a perpendicular distance from a line and \(r\) denote radial
-distance from a point).
+geometry of the sampler \citep[usually referred to as the
+\emph{probability density function of distances};][Chapter
+3]{Buckland:2001vm}. When considering a particular transect type we let
+\(x\) denote a perpendicular distance from a line and \(r\) denote
+radial distance from a point (rather than using \(y\)).
 
 It is crucial that the detection function accurately models
-detectability at small distances. We are less worried by its behaviour
-further away from 0. To ensure that the model is not overly influence by
-distances far from zero, we truncate the distances beyond a given
-distance \(w\) (known as the \emph{truncation distance}). Fitting to
-distances at great distances from zero does not demonstrably improve the
-precision of abundance estimates (Buckland et al. 2001, 103--8,
-151--53).
+detectability at small distances; we are less worried by its behaviour
+further away from 0. To ensure that the model is not overly influenced
+by distances far from zero, we truncate the distances beyond a given
+distance \(w\) (this can be done during analysis or while collecting
+data in the field).
 
 Models for the detection function are expected to have the following
-properties (Buckland et al. 2015, Chapter 5):
+properties \citep[Chapter 5]{buckland2015distance}:
 
 \begin{itemize}
-\itemsep1pt\parskip0pt\parsep0pt
+\tightlist
 \item
   \emph{Shoulder}: we expect observers to be able to see objects near
-  them, not just those directly adjacent to them. For this reason, we
-  expect the detection function to be horizontal near zero distance.
+  them, not just those directly in front of them. For this reason, we
+  expect the detection function to be flat near the line or point.
 \item
   \emph{Non-increasing}: we do not think that observers should be more
-  likely to see things distant from the transect than those near the
-  transect. If this occurs, it usually indicates an issue with field
-  procedure (that the distribution of objects with respect to the line,
-  \(\pi(y)\) is not what we expect), so we do not want the detection
-  function to model this.
+  likely to see distant objects than those nearer the transect. If this
+  occurs, it usually indicates an issue with survey design or field
+  procedures (for example that the distribution of objects with respect
+  to the line, \(\pi(y)\) is not what we expect), so we do not want the
+  detection function to model this
+  \citep{Marques:2010he, Marques:2012fy, Miller:2015hw}.
 \item
-  \emph{Model robust}: models should be flexible enough to make many
+  \emph{Model robust}: models should be flexible enough to fit many
   different shapes.
@@
   \emph{Estimator efficiency}: we would like models to have low
-  variances, but only given they satisfy the other properties above
-  (which, if they are satisfied, would give low bias).
+  variances, given they satisfy the other properties above (which, if
+  satisfied, would give low bias).
 \end{itemize}
@@
 There is a wide literature on possible formulations for the detection
-function (Buckland 1992; Eidous 2005; Becker and Quang 2009; Giammarino
-and Quatto 2014; Miller and Thomas 2015; Becker and Christ 2015).
-\code{Distance} includes the most popular of these models. Here we will
+function
+\citep{Buckland:1992fa, Eidous:2005bj, Becker:2009cj, Giammarino:2014eg, Miller:2015hw, Becker:2015fi}.
+\code{Distance} includes the most popular of these models. Here we
 detail the most popular detection function approach: ``key function plus
@@
 formulated by taking a ``key'' function and optionally adding
-``adjustments'' to it to improve the fit (Buckland 1992). Mathematically
-we formulate this as: \[
+``adjustments'' to it to improve the fit \citep{Buckland:1992fa}.
+Mathematically we formulate this as: \[
 g(y; \boldsymbol{\theta}) = k(y; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(y; \boldsymbol{\theta}_\text{adjust})\right),
@@
 \] Possible modelling options for key and adjustments are given in Table
-\ref{tab:keyadj} and illustrated in Figure \ref{fig:hnhr} and
+\ref{tab:keyadj} and illustrated in Figures \ref{fig:hnhr} and
 \ref{fig:keyadj}. We select the number of adjustment terms (\(K\)) by
@@
 \begin{table}
-\caption{Modelling options for key plus adjustment series models for the detection function.}
 \begin{tabular}{llll}
@@
 \label{tab:keyadj}
+\caption{Modelling options for key plus adjustment series models for the detection function.
+For each key function the default adjustments are cosine in \pkg{Distance}.
+Note that in the adjustments functions distance is divided by the width or the scale parameter to ensure the shape of adjustment independent of the units of $y$ (Marques et al., 2007); defaults are shown here, though either can be selected to rescale the distances.}
 \end{table}
 
-When adjustment terms are used it may be necessary to standardise the
+When adjustment terms are used it is necessary to standardise the
 results to ensure that \(g(0)=1\), so we can redefine the detection
@@
 A disadvantage of K+A models is that we must resort to constrained
-optimisation (via the \pkg{Rsolnp} package) to ensure that the resulting
-detection function is monotonic non-increasing over the whole range.
-
-We do not always include adjustments (except in the case of the uniform
-key), in which case we refer to ``key only'' models (see the next
-section and ``Model checking and model selection'').
+optimisation \citep[via the \pkg{Rsolnp} package;][]{rsolnp-pkg} to
+ensure that the resulting detection function is monotonic non-increasing
+over its range.
+
+It is not always necessary to include adjustments (except in the case of
+the uniform key) and in such cases we refer to these as ``key only''
+models (see the next section and ``Model checking and model
+selection'').
 
@@
 
-\caption{Possible shapes for the detection function when adjustments are included for half-normal and hazard-rate models.\label{fig:keyadj}}\label{fig:adjust-mix-comp}
+\caption{Possible shapes for the detection function when cosine adjustments are included for half-normal and hazard-rate models.\label{fig:keyadj}}\label{fig:adjust-mix-comp}
 \end{figure}
@@
 There are many factors that can affect the probability of detecting an
-object. These include things like the observer, the vessel or platform
-used, the sea state, weather conditions and time of day. We assume that
-these variables affect detection only via the scale of the detection
-function (and do not affect the shape).
+object: observer skill, cluster size (if objects occur in clusters), the
+vessel or platform used, sea state, other weather conditions, time of
+day and more. In \pkg{Distance} we assume that these variables affect
+detection only via the scale of the detection function (and do not
+affect the shape).
 
 Covariates can be included in this formulation by considering the scale
-parameter from the half-normal or hazard-rate detection functions as a
-linear model (on the exponential scale) of the (\(J\)) covariates
-(\(\mathbf{z}\); a vector of length \(J\) for each observation): \[
+parameter from the half-normal or hazard-rate detection functions as a(n
+exponentiated) linear model of the (\(J\)) covariates (\(\mathbf{z}\); a
+vector of length \(J\) for each observation): \[
 \sigma(\mathbf{z}) = \exp(\beta_0 + \sum_{j=1}^J \beta_j z_j).
-\] In the next section we will discuss model selection.
+\]
 
 Including covariates has an important implication for our calculation of
-detectability. Because we do not know what the true distribution of the
-covariates is we must calculate the probability of detection conditional
-on the observed values of the covariates: \[
+detectability. We do not know the true distribution of the covariates,
+we therefore must either: \((i)\) model the distribution of the
+covariates and integrate the covariates out of the joint density (thus
+making strong assumptions about their distribution), or \((ii)\)
+calculate the probability of detection conditional on the observed
+values of the covariates \citep{Marques:2003vb}. We opt for the latter:
+\[
 p(\mathbf{z_i}) = \int_0^w \pi(y) g(y, \mathbf{z_i}; \boldsymbol{\theta}) dy,
 \] where \(\mathbf{z_i}\) is the vector of \(J\) covariates associated
-with observation \(i\). For covariate models, we are calculating a value
-of ``average'' probability of detection (average in the sense of
-distance) per observation. There will be as many unique values of
-\(p(\mathbf{z_i})\) as there are unique covariate combinations in our
+with observation \(i\). For covariate models, we calculate a value of
+``average'' probability of detection (average in the sense of distance
+being integrated out) per observation. There are as many unique values
+of \(p(\mathbf{z_i})\) as there are unique covariate combinations in our
 data.
@@
 guaranteed to be monotonic non-increasing for all covariate
-combinations, as we do not have any model for the distribution of the
-covariates. For this reason, we advise against using both adjustments
-and covariates in a detection function (see Miller and Thomas 2015 for
-an example of when this can be problematic).
+combinations. Without a model for the distribution of the covariates, it
+is not possible to know what the behaviour of the detection function
+will be across the ranges of the covariates. As such we cannot set
+meaningful constraints on monotonicity. For this reason, we advise
+caution when using both adjustments and covariates in a detection
+function \citep[see][ for an example of when this can be
+problematic]{Miller:2015hw}.
 
@@
 
-The workhorse of detection function fitting in \pkg{Distance} is the
-\code{ds} function. Here we show off the formulations for the detection
-function we have seen above applied to the minke whale and amakihi data.
+A detection function can be fitted in \pkg{Distance} using the \code{ds}
+function. Here we apply some of the possible formulations for the
+detection function we have seen above to the minke whale and amakihi
+data.
 
@@
 
-We can fit a model to the minke whale data, setting the truncation at
+First we fit a model to the minke whale data, setting the truncation at
 1.5km and using the default options in \code{ds} very simply:
@@
 \begin{CodeInput}
-minke_hn <- ds(minke, truncation=1.5)
+minke_hn <- ds(minke, truncation = 1.5)
 \end{CodeInput}
@@
 Starting AIC adjustment term selection.
+\end{CodeOutput}
+\begin{CodeOutput}
 Fitting half-normal key function
-Key only models do not require monotonicity contraints. Not constraining model for monotonicity.
+\end{CodeOutput}
+\begin{CodeOutput}
+Key only model: not constraining for monotonicity.
+\end{CodeOutput}
+\begin{CodeOutput}
 AIC= 46.872
+\end{CodeOutput}
+\begin{CodeOutput}
 Fitting half-normal key function with cosine(2) adjustments
+\end{CodeOutput}
+\begin{CodeOutput}
 AIC= 48.872
-
-half-normal key function selected!
-\end{CodeOutput}
-\end{CodeChunk}
-
-Note that \code{ds} will automatically select adjustment terms by AIC
-and shows the selection steps.
-
-Figure \ref{fig:minkeamakihi} (top left plot) shows the result of
-calling \code{plot} on the resulting model object. We can also call
+\end{CodeOutput}
+\begin{CodeOutput}
+
+Half-normal key function selected.
+\end{CodeOutput}
+\end{CodeChunk}
+
+Note that when there are no covariates in the model, \code{ds} will add
+adjustment terms to the model until there is no improvement in AIC.
+
+Figure \ref{fig:minkeamakihi} (left panel) shows the result of calling
+\code{plot} on the resulting model object. We can also call
 \code{summary} on the model object to get summary information about the
-fitted model (we postpone this to the next section).
-
-We can specify the form of the detection function via the \code{key=}
-argument to \code{ds}. For example, a hazard rate model can be fitted
-as:
+fitted model (we postpone this until the next section).
+
+A different form for the detection function can be specified via the
+\code{key=} argument to \code{ds}. For example, a hazard rate model can
+be fitted as:
 
@@
 \begin{CodeInput}
-minke_hrcos <- ds(minke, truncation=1.5, key="hr")
+minke_hrcos <- ds(minke, truncation = 1.5, key = "hr")
 \end{CodeInput}
@@
 Starting AIC adjustment term selection.
+\end{CodeOutput}
+\begin{CodeOutput}
 Fitting hazard-rate key function
-Key only models do not require monotonicity contraints. Not constraining model for monotonicity.
+\end{CodeOutput}
+\begin{CodeOutput}
+Key only model: not constraining for monotonicity.
+\end{CodeOutput}
+\begin{CodeOutput}
 AIC= 48.637
+\end{CodeOutput}
+\begin{CodeOutput}
 Fitting hazard-rate key function with cosine(2) adjustments
+\end{CodeOutput}
+\begin{CodeOutput}
 AIC= 50.386
-
-hazard-rate key function selected!
-\end{CodeOutput}
-\end{CodeChunk}
-
-\code{ds} fits a model with cosine adjustments (the default) but finds
-the AIC improvement to be insufficient to select the adjustment.
+\end{CodeOutput}
+\begin{CodeOutput}
+
+Hazard-rate key function selected.
+\end{CodeOutput}
+\end{CodeChunk}
+
+Here \code{ds} also fits the hazard-rate model then hazard-rate with a
+cosine adjustment but the AIC improvement is insufficient to select the
+adjustment, so the hazard-rate key-only model is returned.
+
+Other adjustment series can be selected using the \texttt{adjustment=}
+argument and specific orders of adjustments can be set using
+\texttt{order=}. For example, to specify a uniform model with cosine
+adjustments of order 1 and 2 we can write:
+
+\begin{CodeChunk}
+\begin{CodeInput}
+minke_unifcos <- ds(minke, truncation = 1.5, key = "unif",
+                    adjustment = "cos", order = c(1, 2))
+\end{CodeInput}
+\begin{CodeOutput}
+Fitting uniform key function with cosine(1,2) adjustments
+\end{CodeOutput}
+\begin{CodeOutput}
+AIC= 48.268
+\end{CodeOutput}
+\end{CodeChunk}
+
+Hermite polynomial adjustments use the code \texttt{"herm"} and simple
+polynomials \texttt{"poly"}, adjustment order should be in line with
+Table \ref{tab:keyadj}.
 
@@
 
-By default \code{ds} assumes the data given to it is line transect, but
-we can switch to points using \code{transect="point"}. Including
-covariates in the scale is via the \code{formula=~...} argument to
-\code{ds}; a hazard-rate model for the amakihi that includes observer as
-a covariate can be specified by (truncating sightings at 82.5m, given in
-Marques et al. 2007):
+\code{ds} assumes the data given to it has been collected as line
+transects, but we can switch to point transects using the argument
+\code{transect="point"}. We can include covariates in the scale
+parameter via the \code{formula=~...} argument to \code{ds}. A
+hazard-rate model for the amakihi that includes observer as a covariate
+and a truncation distance of 82.5m \citep{Marques:2007ey} can be
+specified using :
 
@@
 \begin{CodeInput}
-amakihi_hr_obs <- ds(amakihi, truncation=82.5, transect="point",
-                     key="hr", formula=~obs)
+amakihi_hr_obs <- ds(amakihi, truncation = 82.5, transect = "point",
+                     key = "hr", formula = ~obs)
 \end{CodeInput}
 \begin{CodeOutput}
-Cannot perform AIC adjustment term selection when covariates are used.
+Model contains covariate term(s): no adjustment terms will be included.
+\end{CodeOutput}
+\begin{CodeOutput}
 Fitting hazard-rate key function
+\end{CodeOutput}
+\begin{CodeOutput}
 AIC= 10778.448
+\end{CodeOutput}
+\begin{CodeOutput}
 No survey area information supplied, only estimating detection function.
@@
 
-As with the minke whale model, we can plot the resulting detection
-function (Figure \ref{fig:minkeamakihi}, centre column). Because the
-amakihi model included covariates, the plot shows the detection function
-averaged over levels/values of the covariate. Points on the plot
-indicate probability of detection for each observation. For the
+Note that here, unlike with the minke whale data, \code{ds} warns us
+that we have only supplied enough information to estimate the detection
+function (not density or abundance).
+
+While automatic AIC selection is performed on adjustment terms, model
+selection for covariates must be performed manually. Here we add a
+second covariate: minutes after sunrise. We will compare these two
+models further in the following section.
+
+\begin{CodeChunk}
+\begin{CodeInput}
+amakihi_hr_obs_mas <- ds(amakihi, truncation = 82.5, transect = "point",
+                         key = "hr", formula = ~obs + mas)
+\end{CodeInput}
+\begin{CodeOutput}
+Model contains covariate term(s): no adjustment terms will be included.
+\end{CodeOutput}
+\begin{CodeOutput}
+Fitting hazard-rate key function
+\end{CodeOutput}
+\begin{CodeOutput}
+AIC= 10777.376
+\end{CodeOutput}
+\begin{CodeOutput}
+No survey area information supplied, only estimating detection function.
+\end{CodeOutput}
+\end{CodeChunk}
+
+As with the minke whale model, we can plot the resulting models (Figure
+\ref{fig:minkeamakihi}, middle and right panels). However, for point
+transect studies, probability density function plots give a better sense
+of model fit than the detection function plots. This is because when
+plotting the detection function for point transect data, the histogram
+must be rescaled to account for the geometry of the point sampler. The
+amakihi models included covariates, so the plots show the detection
+function averaged over levels/values of the covariate. Points on the
+plot indicate probability of detection for each observation. For the
 \texttt{amakihi\_hr\_obs} model we see fairly clear levels of the
-observer covariate in the points. Looking at the left panel of Figure
-\ref{fig:minkeamakihi}, we can see this is less clear when adding
-minutes after sunrise as a covariate into the model:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-amakihi_hr_obs_mas <- ds(amakihi, truncation=82.5, transect="point",
-                         key="hr", formula=~obs+mas)
-\end{CodeInput}
-\begin{CodeOutput}
-Cannot perform AIC adjustment term selection when covariates are used.
-Fitting hazard-rate key function
-AIC= 10777.376
-No survey area information supplied, only estimating detection function.
-\end{CodeOutput}
-\end{CodeChunk}
-
-The bottom row of Figure \ref{fig:minkeamakihi} shows the probability
-density functions for the amakihi models. Probability density function
-plots give a better sense of fit of the model than the detection
-functions, as for point data we must rescale the histogram when plotting
-the detection function to take into account the geometry of the point
-sampler.
-
-** ER suggests the g(x) plots for the amakihi be deleted **
+observer covariate in the points. Looking at the right panel of Figure
+\ref{fig:minkeamakihi}, this is less clear when adding minutes after
+sunrise (a continuous covariates) to the model.
 
@@
 
-\caption{Top row plots of fitted detection functions overlayed on the histograms of observed distances; left minke whale data, half-normal model; centre, amakihi data hazard-rate with observer as a covariate; right, amakihi data, hazard-rate model with observer and minutes after sunrise as covariates (for amakihi histograms are rescale to take into account sampler geometry).
-Bottom row shows plots of the probability density function for the amakihi models (as top row).
-For covariate models (centre and right columns) points indicate probability of detection for a given observation (given that observations covariate values) and lines indicate the average detection function.\label{fig:minkeamakihi}}\label{fig:minke-amakihi-hn-plot}
+\caption{Left: fitted detection function overlayed on the histogram of observed distances for the minke whale data using half-normal model.
+Centre and right: plots of the probability density function for the amakihi models.
+Centre, hazard-rate with observer as a covariate; right, hazard-rate model with observer and minutes after sunrise as covariates.
+Points indicate probability of detection for a given observation (given that observation's covariate values) and lines indicate the average detection function.\label{fig:minkeamakihi}}\label{fig:minke-amakihi-hn-plot}
 \end{figure}
@@
 can use \code{summary} to give useful information about our fitted
-model. For example for our hazard-rate model for the amakihi, with
-observer as a covariate:
+model. For our hazard-rate model of the amakihi data, with observer as a
+covariate:
 
@@
 calculated) and its uncertainty. The final line gives an estimate of
-abundance for the area covered by the survey (covered in the next
-section).
+abundance in the area covered by the survey (see the next section;
+though note this estimate does not take into account cluster size).
 
@@
 
-To judge goodness of fit for detection functions when exact distances
-are used, we want to compare the cumulative distribution function (CDF)
-and empirical distribution function (EDF) for the detection function via
-a quantile-quantile plot (Q-Q plot). In our case the CDF evaluates the
-probability of observing an object at a distance less than or equal to
-some given value. The EDF gives the proportion of observations for which
-the CDF is less than or equal to that of a given distance. This can be
-interpreted as assessing whether the number of observations up to a
-given distance in line with what the model says they should be (where
-`given values' are the observed distances). As usual for Q-Q plots,
-``good'' models will have values close to the line \(y=x\), poor models
-will show more deviations from that line.
-
-We can inspect Q-Q plots visually, though this is prone to subjective
-judgments. Instead we can quantify the Q-Q plot's information using a
-Kolmogorov-Smirnov or Cramer-von Mises test (Burnham et al. 2004). Both
-test whether points from the EDF and CDF are from the same distribution.
-The Kolmogorov-Smirnov uses the test statistic of the largest difference
-between a point on the Q-Q plot and the line \(y=x\), whereas the
-Cramer-von Mises test uses the sum of all the differences. As it takes
-into account more information and is therefore more powerful, the
-Cramer-von Mises is generally preferred. A significant result from
-either tests indicates that the EDF and CDF do not come from the same
-distribution (and therefore the model is does not fit the data well).
-
-We can generate a Q-Q plot and test results using the \code{gof_ddf}
-function. We can see the goodness of fit tests below for two models for
-the amakihi data (Figure \ref{amakihi-qq}), first fitting a half-normal
-model without covariates or adjustments (not setting
-\code{adjustment=NULL} will force \code{ds} to fit a model with no
-adjustments), then calculating goodness of fit for that model and our
-hazard-rate model with observer and minutes after sunrise included:
+We use a quantile-quantile plot (Q-Q plot) to visually assess how well a
+detection functions fits the data when we have exact distances. Here the
+Q-Q plot compares the cumulative distribution function (CDF) of the
+fitted detection function to the distribution of the data (empirical
+distribution function or EDF). The Q-Q plots in \pkg{Distance} plot a
+point for every observation. The EDF is the proportion of points that
+have been observed at a distance equal to or less than the distance of
+that observation. The CDF is calculated from the fitted detection
+function as the probability of observing an object at a distance less
+than or equal to that of the given observation. This can be interpreted
+as assessing whether the number of observations up to a given distance
+is in line with what the model says they should be. As usual for Q-Q
+plots, ``good'' models will have values close to the line \(y=x\), poor
+models will show greater deviations from that line.
+
+Q-Q plots can be inspected visually, though this is prone to subjective
+judgments. Therefore, we also quantify the Q-Q plot's information using
+a Kolmogorov-Smirnov or Cramér-von Mises test \citep{Burnham:2004vd}.
+Both test whether points from the EDF and CDF are from the same
+distribution. The Kolmogorov-Smirnov uses the test statistic of the
+largest difference between a point on the Q-Q plot and the line \(y=x\),
+whereas the Cramér-von Mises test uses the sum of all the differences.
+As it takes into account more information and is therefore more
+powerful, the Cramér-von Mises is generally preferred. A significant
+result from either test gives evidence against the null hypothesis (that
+the data arose from the fitted model), suggesting that the model does
+not fit the data well.
+
+We can generate a Q-Q plot and test results using the \code{gof_ds}
+function. Figure \ref{amakihi-qq} shows the goodness of fit tests for
+two models for the amakihi data. We first fit a half-normal model
+without covariates or adjustments (setting \code{adjustment=NULL} will
+force \code{ds} to fit a model with no adjustments):
 
@@
 \begin{CodeInput}
-amakihi_hr <- ds(amakihi, truncation=82.5, transect="point", key="hn", adjustment=NULL)
+amakihi_hn <- ds(amakihi, truncation = 82.5, transect = "point",
+                 key = "hn", adjustment = NULL)
 \end{CodeInput}
@@
 Fitting half-normal key function
-Key only models do not require monotonicity contraints. Not constraining model for monotonicity.
+\end{CodeOutput}
+\begin{CodeOutput}
+Key only model: not constraining for monotonicity.
+\end{CodeOutput}
+\begin{CodeOutput}
 AIC= 10833.841
+\end{CodeOutput}
+\begin{CodeOutput}
 No survey area information supplied, only estimating detection function.
@@
 \begin{CodeInput}
-gof_ds(amakihi_hr)
+gof_ds(amakihi_hn)
 \end{CodeInput}
@@
 
-The corresponding Q-Q plots are shown in Figure \ref{amakihi-qq}.
+We can conclude that the half-normal model does not pass our goodness of
+fit tests and should be discarded. Both hazard-rate models (output only
+shown for hazard-rate model with observer and minutes after sunrise) had
+non-significant goodness of fit test statistics and are both therefore
+deemed plausible models. The corresponding Q-Q plots are shown in Figure
+\ref{amakihi-qq}, comparing the half-normal model with the hazard-rate
+model with observer and minutes after sunrise included.
 
@@
 
-\caption[Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates) and hazard-rate model with observer and minutes after sunrise for the amakihi data]{Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates) and hazard-rate model with observer and minutes after sunrise for the amakihi data.\label{amakihi-qq}}\label{fig:amakihi-qq-comp}
+\caption[Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates]{Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates; left) and hazard-rate model with observer and minutes after sunrise (right) for the amakihi data.\label{amakihi-qq}}\label{fig:amakihi-qq-comp}
 \end{figure}
@@
 
+For non-exact distance data, a \(\chi^2\)-test can be used to assess
+goodness of fit \citep[see][Section 3.4.4]{Buckland:2001vm}.
+\(\chi^2\)-test results are produced by \texttt{gof\_ds} when
+binned/grouped distances are provided.
+
 \subsection{Model selection}\label{model-selection}
 
-Once we have a set of models which fit well, we can use Akaike's
-Information Criterion (AIC) to select between models. \pkg{Distance}
-includes a function to create table of summary information for fitted
-models, making it easy to get an overview of a large number of models at
-once. The \code{summarize_ds_models} function takes models as input and
-can be especially useful when paired with \pkg{knitr}'s \code{kable}
-function to create summary tables for publication. An example of this
-output is shown in Table \ref{tab:amakihi} and was generated by:
+Once we have a set of plausible models, we can use Akaike's Information
+Criterion (AIC) to select between models \citep[see
+e.g.][]{burnham2003model}. \pkg{Distance} includes a function to create
+table of summary information for fitted models, making it easy to get an
+overview of a large number of models. The \code{summarize_ds_models}
+function takes models as input and can be especially useful when paired
+with \pkg{knitr}'s \code{kable} function to create summary tables for
+publication \citep{knitr-pkg}. An example of this output (with all
+models included) is shown in Table \ref{tab:amakihi} and was generated
+by the following call to \code{summarize_ds_models}:
 
@@
 \begin{CodeInput}
-library(knitr)
-summarize_ds_models(amakihi_hr, amakihi_hr_obs, amakihi_hr_obs_mas)
+summarize_ds_models(amakihi_hn, amakihi_hr_obs, amakihi_hr_obs_mas)
 \end{CodeInput}
@@
 
+In this case we may be skeptical about the top model as selected by AIC
+being truly better than the second best model, as there is only a very
+small difference in AICs. Generally, if the difference between AICs is
+less than 2, we may investigate multiple ``best'' models, potentially
+resorting to the simplest of these models. In the authors' experience,
+it is often the case that models with similar AICs will have similar
+estimates probabilities of detection, so in practice there is little
+difference in selecting between these models. It is important to note
+that comparing AICs between models with different truncations is not
+appropriate, as models with different truncation use different data.
+
 \begin{table}
 
-\caption{Summary for the detection function models fitted to the amakihi data.
-``C-vM'' stands for Cramer-von Mises, $P_a$ indicates average detectability (see ``Estimating abundance and variance''), se indicates standard error.
-Models are sorted according to AIC.\label{tab:amakihi}}
-\begin{tabular}{llrrrrr}
+\centering
+\begin{tabular}[t]{lllrrrr}
 \toprule
-Key function & Formula & C-vM $p$-value & $\hat{P_a}$ & se($\hat{P_a}$) & AIC & $\Delta$AIC\\
+Model & Key function & Formula & C-vM $p$-value & $\hat{P_a}$ & se($\hat{P_a}$) & $\Delta$AIC\\
 \midrule
-Hazard-rate key function & ~obs + mas & 0.389 & 0.319 & 0.020 & 10777.38 & 0.000\\
-Hazard-rate key function & ~obs & 0.271 & 0.314 & 0.020 & 10778.45 & 1.073\\
-Half-normal key function & ~1 & 0.004 & 0.351 & 0.011 & 10833.84 & 56.465\\
+\texttt{amakihi\char`_hr\char`_obs\char`_mas} & Hazard-rate & ~obs + mas & 0.389 & 0.319 & 0.020 & 0.000\\
+\texttt{amakihi\char`_hr\char`_obs} & Hazard-rate & ~obs & 0.271 & 0.314 & 0.020 & 1.073\\
+\texttt{amakihi\char`_hn} & Half-normal & ~1 & 0.004 & 0.351 & 0.011 & 56.465\\
 \bottomrule
 \end{tabular}
+\caption{Summary for the detection function models fitted to the amakihi data.
+``C-vM'' stands for Cramér-von Mises, $P_a$ is average detectability (see ``Estimating abundance and variance''), se is standard error.
+Models are sorted according to AIC.\label{tab:amakihi}}
 \end{table}
@@
 Though fitting the detection function is the primary modelling step in
-distance sampling, we are really interested in estimating detectability,
-and from that abundance. We also wish to calculate our certainty in each
-abundance estimate. This section addresses these issues mathematically
-before showing how to estimate abundance and its variance in
-\proglang{R}.
+distance sampling, we are really interested in estimating density or
+abundance. It is also important to calculate our uncertainty for these
+estimates. This section addresses these issues mathematically before
+showing how to estimate abundance and its variance in \proglang{R}.
 
@@
 We wish to obtain the abundance in a study region, of which we have
-sampled a (representative) subset. To do this we first calculate the
-abundance in the area we have surveyed (the \emph{covered area}) to
-obtain \(\hat{N}_\text{C}\), we can then scale this up to the full study
-area by multiplying it by the ratio of covered area to study area.
+sampled a random subset. To do this we first calculate the abundance in
+the area we have surveyed (the \emph{covered area}) to obtain
+\(\hat{N}_\text{C}\), we can then scale this up to the full study area
+by multiplying it by the ratio of covered area to study area. We discuss
+other methods for spatially explicit abundance estimation in
+``Extensions''.
 
@@
 we use the estimates of detection probability (the
-\(\{p_i; i=1,\ldots,n\}\), above) in a Horvitz-Thompson like estimator:
+\(\{\hat{p}(\mathbf{z}_i); i=1,\ldots,n\}\), above) in a
+Horvitz-Thompson-like estimator:
 
 \begin{equation}
-\hat{N}_\text{C} = \sum_{i=1}^n\frac{s_i}{\hat{p}_i},
+\hat{N}_\text{C} = \sum_{i=1}^n\frac{s_i}{\hat{p}(\mathbf{z}_i)},
 \label{ht}
@@
 
-where \(s_i\) are the sizes of the observed groups of objects, which is
-equal to 1 if objects only occur singly (Borchers and Burnham 2004).
-Thompson (2002) is the canonical reference to this type of estimator;
-intuitively, we can think of the estimates of detectability
-(\(\hat{p}_i\)) as ``inflating'' the group sizes (\(s_i\)) to account
-for incomplete detection -- we then sum over all \(n\) detections to
-obtain the abundance estimate. For models that do not include
-covariates, \(\hat{p}_i\) is equal for all \(i\), so this is equivalent
-to summing the groups and inflating that sum by the corresponding
-\(\hat{p} (=\hat{p}_i \forall i)\).
+where \(s_i\) are the sizes of the observed clusters of objects, which
+are all equal to 1 if objects only occur singly \citep{Borchers:2004wr}.
+\citet{thompson2012sampling} is the canonical reference to this type of
+estimator. Intuitively, we can think of the estimates of detectability
+(\(\hat{p}(\mathbf{z}_i)\)) as ``inflating'' the cluster sizes
+(\(s_i\)), we then sum over the detections (\(i\)) to obtain the
+abundance estimate. For models that do not include covariates,
+\(\hat{p}(\mathbf{z}_i)\) is equal for all \(i\), so this is equivalent
+to summing the clusters and inflating that sum by dividing through by
+the corresponding \(\hat{p} (=\hat{p}(\mathbf{z}_i) \forall i)\).
 
@@
 \hat{N} = \frac{A}{a} \hat{N}_\text{C},
-\] where \(A\) is the area of the study region to extrapolate the
-abundance estimate to and \(a\) is the covered area. For line transects
+\] where \(A\) is the area of the study region to which to extrapolate
+the abundance estimate and \(a\) is the covered area. For line transects
 \(a=2wL\) (twice the truncation distance multiplied by the total length
-of transects surveyed) and for points \(a=\pi w^2 K\).
-
-\emph{have L and K been previously defined?}
+of transects surveyed, \(L\)) and for points \(a=\pi w^2 T\) (where
+\(\pi w^2\) is the area of a single surveyed circle and \(T\) is the sum
+of the number of visits to the sampled points).
 
@@
 consider what single detectability value would give the estimated
-\(\hat{N}\) and therefore calculate: \[
+\(\hat{N}_\text{C}\) and therefore calculate: \[
 \hat{P_a} = n/\hat{N}_\text{C}.
-\] This can be a useful summary statistic and is included in the
-\code{summary} output and the table produced by
-\code{summarize_ds_models}. \emph{anything else to say about this, seems
-to mess up folks, including me}
+\] This can be a useful summary statistic, giving us an idea of how
+detectable our \(n\) observed animals would have to be to estimate the
+same \(\hat{N}\) if there were no observed covariates. It can also be
+compared to similar estimates in mark-recapture studies. \(P_a\) is
+included in the \code{summary} output and the table produced by
+\code{summarize_ds_models}.
 
@@
 We may wish to calculate abundance estimates for some sub-regions of the
-study region, we call these areas \emph{strata}. For example, strata may
-be defined by habitat types or animal gender (or some combination) which
-may be of interest for biological or management reasons. To calculate
-estimates for a given stratification each observation must occur in a
-stratum which must be labelled with a \code{Region.Label} and have a
-corresponding \code{Area} (if we are using an animal characteristic like
-gender, we would have the areas be the same but if we were using say
-forested vs.~wetland habitat the areas of those strata would be
-different). Finally, we must also know the stratum in which transects
-lie.
-
-As an example the minke whale data consists of two strata: \code{North}
-and \code{South} relating to a stratum further away and nearer the
-Antarctic ice edge, respectively. Figure \ref{fig:minke-strata} shows
-the two strata, along with observations and transect lines. \emph{figure
-number messed up}
-
-\begin{figure}
-\includegraphics{minke-strata}
-\label{fig:minke-strata}
-\caption{Strata used for the minke whale data adapted from @Hedley:2004et.
-Points show the locations of observations along transect lines.
-The stepped line shows the boundary between North and South strata.
-Further details on the survey are available in @Branch:2001ua (simulated data is based on ``1992/93 Area III'' therein).}
-\end{figure}
+study region, we call these areas \emph{strata}. Stratification can be
+used to increase the precision of estimates if we know \emph{a priori}
+that density varies between different parts of the study area. For
+example, strata may be defined by habitat types or animal sex (or some
+combination) which may be of interest for biological or management
+reasons. To calculate estimates for a given stratification each
+observation must occur in a stratum which must be labelled with a
+\code{Region.Label} and have a corresponding \code{Area} (if we are
+using an animal characteristic like sex, we would have the areas be the
+same but if we were using say forested vs.~wetland habitat the areas of
+those strata would be different). Finally, we must also know the stratum
+in which each observation occurs.
+
+As an example, the minke whale data contain two strata: \code{North} and
+\code{South} relating to strata further away from and nearer to the
+Antarctic ice edge, respectively.
 
@@
 
-Here we take an intuitive approach to uncertainty estimation, for a full
-derivations consult F. F. C. Marques and Buckland (2003). Uncertainty in
-\(\hat{N}\) comes from two sources:
+We take an intuitive approach to uncertainty estimation, for a full
+derivation consult \citet{Marques:2003vb}. Uncertainty in \(\hat{N}\)
+comes from two sources:
 
@@
 \def\labelenumi{\arabic{enumi}.}
-\itemsep1pt\parskip0pt\parsep0pt
+\tightlist
 \item
-  \emph{Model uncertainty}, from the estimation of the detection
-  function parameters \(\boldsymbol{\theta}\)
+  \emph{Detection function}: Uncertainty in parameter
+  (\(\boldsymbol{\theta}\)) estimation.
 \item
-  \emph{Sampling uncertainty}, from the distribution of objects along
-  the transect lines or between visiting occasions for points.
+  \emph{Encounter rate}: Sampling variability arising from differences
+  in the number of observations per transect.
 \end{enumerate}
 
-We can see this by looking at the Horvitz-Thompson estimation in
+We can see this by looking at the Horvitz-Thompson-like estimation in
 (\ref{ht}) and consider the terms which are random. These are: the
-detectability \(\hat{p}_i\) (and hence the parameters of the detection
-function it is derived from) and \(n\), the number of observations. We
-assume that the observed group size (\(s_i\)) is recorded without error.
-
-Model uncertainty can be addressed using standard maximum likelihood
-theory. We can invert the Hessian matrix of the likelihood to obtain a
-variance-covariance matrix. We can then pre- and post-multiply this by
-the derivatives of \(\hat{N}_\text{C}\) with respect to the parameters
-of the detection function \[
+detectability \(\hat{p}(\mathbf{z}_i)\) (and hence the parameters of the
+detection function it is derived from) and \(n\), the number of
+observations.
+
+Model parameter uncertainty can be addressed using standard maximum
+likelihood theory. We can invert the Hessian matrix of the likelihood to
+obtain a variance-covariance matrix. We can then pre- and post-multiply
+this by the derivatives of \(\hat{N}_\text{C}\) with respect to the
+parameters of the detection function \[
 \widehat{\text{Var}}_\text{model}\left( \hat{N}_\text{C}\right) = \left(\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}\right)^\text{T} \left(\hat{\mathbf{H}}(\hat{\boldsymbol{\theta}})^{-1} \right)\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}
@@
 Hessian (outer product of first derivatives of the log likelihood) for
-numerical stability (Buckland et al. 2001, p 62). Note that although we
+numerical stability \citep[p 62]{Buckland:2001vm}. Note that although we
 calculate uncertainty in \(\hat{N}_\text{C}\), we can trivially scale-up
-to variance of \(\hat{N}\).
-
-Sampling uncertainty can be characterised (rather than as simply \(n\))
-by the \emph{encounter rate}: the number of objects per unit transect.
-When covariates are not included in the detection function we can define
-the encounter rate as \(n/L\) for line transects (where \(L\) is the
-total line length) or \(n/T\) for point transects (where \(T\) is the
-total number of visits to all points) \emph{earlier eqn uses notation
-\(K\) for number of points}. When covariates are included in the
-detection function, it is recommended that we substitute the \(n\) in
-the previous expressions with the estimated abundance \(\hat{N}\) as
-this will take into account the effects of the covariates.
+to variance of \(\hat{N}\) (by noting that
+\(\hat{N} = \frac{A}{a} \hat{N}_\text{C}\) and therefore
+\(\widehat{\text{Var}}_\text{model}\left(\hat{N}\right) = \left( \frac{A}{a} \right)^2 \widehat{\text{Var}}_\text{model} \left(\hat{N}_\text{C} \right)\)).
+
+\emph{Encounter rate} is the number of objects per unit transect (rather
+than just \(n\)). When covariates are not included in the detection
+function we can define the encounter rate as \(n/L\) for line transects
+(where \(L\) is the total line length) or \(n/T\) for point transects
+(where \(T\) is the total number of visits summed over all points). When
+covariates are included in the detection function, it is recommended
+that we substitute the \(n\) in the encounter rate with the estimated
+abundance \(\hat{N}_\text{C}\) as this will take into account the
+effects of the covariates \citep{Innes:2002ka}.
 
 For line transects, by default, \pkg{Distance} uses a variation of the
-estimator ``R2'' from Fewster et al. (2009) replacing number of
-observations per sample with the estimated abundance per sample (Innes
-et al. 2002; F. F. C. Marques and Buckland 2003): \[
-\widehat{\text{Var}}_{\text{encounter},R2} = \frac{K}{L^2(K-1)} \sum_{k=1}^{K} l_k^2 \left( \frac{N_k}{l_k} - \frac{N}{L}\right)^2,
+estimator ``R2'' from \citet{Fewster:2009ku} replacing number of
+observations per sample with the estimated abundance per sample, thus
+taking into account cluster size if this is recorded
+\citep{Innes:2002ka, Marques:2003vb}: \[
+\widehat{\text{Var}}_{\text{encounter},R2}\left( \hat{N}_\text{C}\right) = \frac{K}{L^2(K-1)} \sum_{k=1}^{K} l_k^2 \left( \frac{\hat{N_{\text{C}, k}}}{l_k} - \frac{\hat{N_\text{C}}}{L}\right)^2,
 \] where \(l_k\) are the lengths of the \(K\) transects (such that
-\(L = \sum_{k=1}^K l_k\)). Whereas for points we use estimator ``P3''
-from Fewster et al. (2009) but again replacing \(n\) by \(\hat{N}\) to
-obtain the following estimator: \[
-\widehat{\text{Var}}_{\text{encounter},P3} = \frac{1}{T(K-1)} \sum_{k=1}^{K} t_k \left( \frac{N_k}{t_k} - \frac{N}{T}\right)^2,
+\(L = \sum_{k=1}^K l_k\)) and \(\hat{N_{C,k}}\) is the abundance in the
+covered area for transect \(k\). For point transects we use the
+estimator ``P3'' from \citet{Fewster:2009ku} but again replace \(n\) by
+\(\hat{N}_\text{C}\) in the encounter rate definition, to obtain the
+following estimator: \[
+\widehat{\text{Var}}_{\text{encounter},P3}\left( \hat{N}_\text{C}\right) = \frac{1}{T(K-1)} \sum_{k=1}^{K} t_k \left( \frac{\hat{N_{\text{C}, k}}}{t_k} - \frac{\hat{N_\text{C}}}{T}\right)^2,
 \] where \(t_k\) is the number of visits to point \(k\) and
-\(T = \sum_{k=1}^K t_k\). \emph{this messes with the \(K\) notation even
-further}
-
-Various formulations for the encounter rate variance are discussed in
-detail in Fewster et al. (2009). \pkg{Distance} implements all of the
+\(T = \sum_{k=1}^K t_k\) (the total number of visits to all points is
+the sum of the visits to each point). Again, it is straightforward to
+calculate the encounter rate variance for \(\hat{N}\) from the encounter
+rate variance for \(\hat{N}_\text{C}\).
+
+Other formulations for the encounter rate variance are discussed in
+detail in \citet{Fewster:2009ku}. \pkg{Distance} implements all of the
 estimators of encounter rate variance given in that article. The
 \code{varn} manual page gives further advice and technical detail on
-encounter rate variance.
-
-We combine detection function variance with encounter rate variance by
-use of the \(\delta\)-method approximation (Goodman 1960).
+encounter rate variance. For example for systematic survey designs,
+estimators S1, S2 and O1, O2 and O3 will typically provide smaller
+estimates of the variance.
+
+We combine these two variances by noting that squared coefficients of
+variation (approximately) add \citep[often referred to as ``the delta
+method'';][]{seber2002estimation}.
 
@@
 
-This prints a rather large amount of information. The summary can be
-more concisely expressed by extracting information from the summary
-object. This object is a \code{list} of \code{data.frames}, so we can
-again use \pkg{knitr}s \code{kable} function to create summary tables of
-abundance estimates and measures of precision, such as Table
-\ref{minke-abund}.
+This prints a rather large amount of information: first the detection
+function summary, then three tables:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  \code{Summary statistics}: giving the areas, covered areas, effort,
+  number of observations, number of transects, encounter rate, its
+  standard error and coefficient of variation for each stratum.
+\item
+  \code{Abundance}: giving estimates, standard errors, coefficients of
+  variation, lower and upper confidence intervals and finally the
+  degrees of freedom for each stratum's abundance estimate.
+\item
+  \code{Density}: lists the same statistics as \code{Abundance} but for
+  a density estimate.
+\end{enumerate}
+
+In each table the bottom row gives a total for the whole study area.
+
+The summary can be more concisely expressed by extracting information
+from the summary object. This object is a \code{list} of
+\code{data.frames}, so we can use the \code{kable} function from
+\pkg{knitr} to create summary tables of abundance estimates and measures
+of precision, such as Table \ref{minke-abund}. We prepare the
+\code{data.frame} as follows before using \code{kable}:
 
@@
                            "$\\text{CV}(\\hat{N}$)")
-kable(minke_table, digits=3, format="latex", booktabs=TRUE,
-      row.names=FALSE, escape=FALSE,caption="Summary of abundance estimation for the half-normal model for the minke whale data.\\label{minke-abund}")
 \end{CodeInput}
+\end{CodeChunk}
+
+\begin{CodeChunk}
 \begin{table}
 
-\caption{Summary of abundance estimation for the half-normal model for the minke whale data.\label{minke-abund}}
-\begin{tabular}{lrrr}
+\centering
+\begin{tabular}[t]{lrrr}
 \toprule
@@
 \end{tabular}
+\caption{Summary of abundance estimation for the half-normal model for the minke whale data.\label{minke-abund}}
 \end{table}
@@
 \section{Extensions}\label{extensions}
+
+Distance sampling has been applied in a wide variety of situations.
+Objects being detected need not be animals; plants
+\citep{Buckland:2007kl}, beer cans \citep{Otto:1990gs}, elephant dung
+\citep{NchPlu01} and bricks on a lake bottom \citep{Ber90} have been
+subjects of distance sampling investigations. The method of detection
+need not be visual sightings. Detections of objects can be through
+auditory as well as visual means \citep{Marques2013b}. Songs of birds
+\citep{Buckland2006} and whale vocalisations \citep{BorSub1} are just
+two examples. Blows made by whales are examples of processes that can be
+modelled using distance sampling. Songs and blows are indirect sampling
+methods producing estimates of cue density. Cue densities can be
+converted to animal densities with additional data needed to estimate
+the rate at which cues are produced and the rate at which they
+disappear. These are but a few examples of the applications to which
+distance sampling has been applied (a larger, but incomplete list of
+references is given at \url{http://distancesampling.org/dbib.html}).
 
@@
 simplified interface for users. For more complex analyses of distance
-sampling data, there are further related packages for modelling in
+sampling data, we provide additional packages for modelling in
 \proglang{R}.
 
-We noted at the start of the article that \pkg{Distance} is a simpler
-wrapper around the package \pkg{mrds}. Additional features are available
-in \pkg{mrds} including the ability to model data where the assumption
-that detection is certain at zero distance from the line or point by
-using mark-recapture type methods when two observers are used (see Burt
-et al. 2014 for an introduction).
-
-The abundance estimates calculated here are based on the assumption that
-within a given stratum density is uniform. We may extend this approach
-to many strata, making the area of each very small to account for
-small-scale variation in space. A more rigourous approach is to build a
-spatial model incorporating spatially-referenced environmental data (for
-example derived from GIS products). \pkg{Distance} interfaces with once
-such package to perform this type of analysis: \pkg{dsm}. So-called
-``density surface modelling'' uses the generalized additive model
-framework (e.g. Wood 2006) to build models of abundance (adjusting
-counts for imperfect detectability) as a function of environmental
-covariates, as part of a two stage model (Hedley and Buckland 2004;
-Miller et al. 2013).
+We noted at the start of the article that \pkg{Distance} is a
+simple-to-use wrapper around the package \pkg{mrds}. Additional features
+available in \pkg{mrds} include models that relax the assumption that
+detection is certain at zero distance from the transect (by including
+data from additional observers). This is done using mark-recapture type
+methods which require additional survey methodology, known as double
+observer surveys \citep[see][ for an introduction]{Burt:2014gu}.
+
+\pkg{Distance} can provide us with estimates of abundance or density for
+each strata as a whole but tells us nothing about the distribution of
+animals within strata. One option is to divide the study area into
+smaller and smaller strata to try to detect patterns in spatial
+distribution, however, a more rigorous approach is to build a spatial
+model. Such models incorporate spatially-referenced environmental data
+(for example derived from GIS products). \pkg{Distance} interfaces with
+one such package used to perform this type of analysis: \pkg{dsm}
+\citep{dsm-pkg}. So-called ``density surface modelling'' uses the
+generalized additive model framework \citep[e.g.][]{Wood:2006vg} to
+build models of abundance (adjusting counts for imperfect detectability)
+as a function of environmental covariates, as part of a two stage model
+\citep{Hedley:2004et, Miller:2013fq}.
+
+Uncertainty in measured covariates (e.g.~cluster size) and model
+uncertainty (when two models have similar fit but substantially
+different estimates) can be incorporated using the multi-analysis
+distance sampling package \pkg{mads} \citep{mads-pkg}. In addition,
+\pkg{mads} can also incorporate sightings with unknown species
+identification. This is done by estimating the abundance of these
+`unidentified' sightings and pro-rating them to the known species
+\citep{Gerrodette:2005}.
+
+As mentioned above, survey design is critical to ensuring that resulting
+distance sampling data can be reliably analysed. \pkg{DSsim} allows
+users to test out different designs in their study region and tailor
+population attributes to reflect the species they are working with.
+\pkg{DSsim} \citep{DSsim-pkg} allows users to more easily identify
+challenges unique to their study and select a survey design which is
+more likely to yield accurate and precise estimates.
+
+Distance for Windows has many users (over 45,000 downloads since 2002)
+and they may be overwhelmed by the prospect of switching existing
+analyses to \proglang{R}. For that reason we have created the
+\pkg{readdst} \citep{readdst-pkg} package to interface with projects
+created by Distance for Windows. The package can take analyses created
+using the CDS, MCDS and MRDS engines in Distance for Windows, extract
+the data and create equivalent models in \proglang{R}. \pkg{readdst} can
+also run these analyses and test the resulting statistics (for example,
+\(\hat{N}\) or \(\hat{P_a}\)) calculated in \proglang{R} against those
+calculated by Distance for Windows. We hope that \pkg{readdst} will
+provide a useful transition to \proglang{R} for interested users.
+\pkg{readdst} is currently available on GitHub at
+\url{https://github.com/distancedevelopment/readdst}.
 
@@
 
-In combination with tools such as \pkg{knitr} and \pkg{rmarkdown}, the
-helper functions in \pkg{Distance} provide a useful set of tools to
-perform reproducible analyses of wildlife abundance for both managers
-and ecologists. We hope that this paper provides useful examples for
-those wishing to pursue this.
+In combination with tools such as \pkg{knitr} and \pkg{rmarkdown}
+\citep{rmarkdown-pkg}, the helper functions in \pkg{Distance} provide a
+useful set of tools to perform reproducible analyses of wildlife
+abundance for both managers and ecologists. We also direct readers'
+attention to \pkg{DsShiny} \citep{dsshiny-pkg}, a package that builds on
+\pkg{Shiny} \citep{shiny-pkg} and \pkg{mrds} to allow users to build and
+fit models in a graphical interface, with output to RMarkdown.
+
+\proglang{R} and its extension packages provide many tools exploratory
+data analysis that can be useful for a distance sampling analysis. We
+hope that this paper provides useful examples for those wishing to
+pursue distance sampling in \proglang{R}. More information on distance
+sampling can be found at \url{http://distancesampling.org} and a mailing
+list is maintained at
+\url{https://groups.google.com/forum/#!forum/distance-sampling}.
 
@@
 sampling analyses in \proglang{R} but believe that \pkg{Distance} is the
-most flexible and feature-complete. \textbf{Appendix ???} gives a
-feature comparison between \pkg{Distance} and other \proglang{R}
-packages for analysis of distance sampling data.
+most flexible and feature-complete, and provides pathways to a range of
+more complicated analyses. Appendix A gives a feature comparison between
+\pkg{Distance} and other \proglang{R} packages for analysis of distance
+sampling data.
 
@@
 The authors would like to thank the many users of \pkg{Distance},
-\pkg{mrds} and DISTANCE who have contributed bug reports and suggestions
-for improvements over the years.
-
-\section*{Bibliography}\label{bibliography}
-\addcontentsline{toc}{section}{Bibliography}
-
-Becker, Earl F, and Aaron M Christ. 2015. ``A Unimodal Model for Double
-Observer Distance Sampling Surveys.'' \emph{PLoS ONE} 10 (8):
-e0136403--18.
-
-Becker, Earl F, and P X Quang. 2009. ``A gamma-shaped detection function
-for line-transect surveys with mark-recapture and covariate data.''
-\emph{Journal of Agricultural, Biological, and Environmental Statistics}
-14 (2): 207--23.
-
-Borchers, David L, and Kenneth P Burnham. 2004. ``General formulation
-for distance sampling.'' In \emph{Advanced Distance Sampling}, 6--30.
-Oxford University Press, Oxford, UK.
-
-Branch, T A, and D S Butterworth. 2001. ``Southern Hemisphere minke
-whales: standardised abundance estimates from the 1978/79 to 1997/98
-IDCR-SOWER surveys.'' \emph{Journal of Cetacean Research and
-Management}.
-
-Buckland, Stephen T. 1992. ``Fitting Density Functions with
-Polynomials.'' \emph{Applied Statistics} 41 (1): 63.
-
-Buckland, Stephen T, David R Anderson, Kenneth P Burnham, David L
-Borchers, and Len Thomas. 2001. \emph{Introduction to Distance
-Sampling}. Estimating Abundance of Biological Populations. Oxford
-University Press, Oxford, UK.
-
-Buckland, Stephen T, David R Anderson, Kenneth P Burnham, Jeffrey L
-Laake, David L Borchers, and Len Thomas. 2004. \emph{Advanced Distance
-Sampling}. Estimating Abundance of Biological Populations. Oxford
-University Press, Oxford, UK.
-
-Buckland, Stephen T, E A Rexstad, Tiago A Marques, and Cornelia S
-Oedekoven. 2015. \emph{Distance Sampling: Methods and Applications}.
-Methods in Statistical Ecology. Springer International Publishing.
-
-Burnham, Kenneth P, Stephen T Buckland, Jeffrey L Laake, David L
-Borchers, Jon R B Bishop, and Len Thomas. 2004. ``Further topics in
-distance sampling.'' In \emph{Advanced Distance Sampling}, 385--89.
-Oxford University Press, Oxford, UK.
-
-Burt, Mary Louise, David L Borchers, Kurt J Jenkins, and Tiago A
-Marques. 2014. ``Using mark-recapture distance sampling methods on line
-transect surveys.'' \emph{Methods in Ecology and Evolution} 5 (11):
-1180--91.
-
-Eidous, Omar M. 2005. ``On Improving Kernel Estimators Using Line
-Transect Sampling.'' \emph{Communications in Statistics - Theory and
-Methods} 34 (4): 931--41.
-
-Fewster, Rachel M, Stephen T Buckland, Kenneth P Burnham, David L
-Borchers, Peter E Jupp, Jeffrey L Laake, and Len Thomas. 2009.
-``Estimating the Encounter Rate Variance in Distance Sampling.''
-\emph{Biometrics} 65 (1): 225--36.
-
-Giammarino, Mauro, and Piero Quatto. 2014. ``On estimating Hooded crow
-density from line transect data through exponential mixture models.''
-\emph{Environmental and Ecological Statistics} 21 (4): 689--96.
-
-Goodman, Leo A. 1960. ``On the Exact Variance of Products.''
-\emph{Journal of the American Statistical Association} 55 (292): 708.
-
-Hedley, Sharon L, and Stephen T Buckland. 2004. ``Spatial models for
-line transect sampling.'' \emph{Journal of Agricultural, Biological, and
-Environmental Statistics} 9 (2): 181--99.
-
-Innes, Stuart, M P Heide-J{ø}rgensen, Jeffrey L Laake, Kristin L Laidre,
-Holly J Cleator, Pierre Richard, and Robert EA Stewart. 2002. ``Surveys
-of belugas and narwhals in the Canadian High Arctic in 1996.''
-\emph{NAMMCO Scientific Publications} 4 (0): 169--90.
-
-Lahoz-Monfort, Jos{é} J, Gurutzeta Guillera-Arroita, and Brendan A
-Wintle. 2013. ``Imperfect detection impacts the performance of species
-distribution models.'' \emph{Global Ecology and Biogeography} 23 (4):
-504--15.
-
-Marques, Fernanda F C, and Stephen T Buckland. 2003. ``Incorporating
-covariates into standard line transect analyses.'' \emph{Biometrics} 59
-(4): 924--35.
-
-Marques, Tiago A, Len Thomas, Steven G Fancy, and Stephen T Buckland.
-2007. ``Improving estimates of bird density using multiple-covariate
-distance sampling.'' \emph{The Auk} 124 (4): 1229.
-
-Miller, David L, and Len Thomas. 2015. ``Mixture models for distance
-sampling detection functions.'' \emph{PLoS ONE}.
-
-Miller, David L, Mary Louise Burt, Eric A Rexstad, and Len Thomas. 2013.
-``Spatial models for distance sampling data: recent developments and
-future directions.'' \emph{Methods in Ecology and Evolution} 4 (11):
-1001--10.
-
-Thomas, Len, Stephen T Buckland, Eric A Rexstad, Jeffrey L Laake,
-Samantha Strindberg, Sharon L Hedley, Jon R B Bishop, Tiago A Marques,
-and Kenneth P Burnham. 2010. ``Distance software: design and analysis of
-distance sampling surveys for estimating population size.''
-\emph{Journal of Applied Ecology} 47 (1): 5--14.
-
-Thompson, S K. 2002. \emph{Sampling}. 2nd ed. Wiley.
-
-Wood, Simon N. 2006. \emph{Generalized Additive Models}. An Introduction
-with R. CRC Press.
+\pkg{mrds} and Distance for Windows who have contributed bug reports and
+suggestions for improvements over the years. We would particularly like
+to thank Steve Buckland, David Borchers, Tiago Marques, Jon Bishop and
+Lorenzo Milazzo for their contributions. We thank Tiago Marques a second
+time for helpful comments on an early version of the paper and Colin
+Beale who suggested the ``flatfile'' data format. We also thank Ken
+Burnham and David Anderson for fundamental contributions to the early
+development of these methods.
+
+\renewcommand\refname{Bibliography}
+\bibliography{jstatsoft.bib}
 

### 2016-03-31 paper.tex
fix shortnames for bibliography

@@
-\documentclass[article]{jss}
+\documentclass[article,shortnames]{jss}
 \usepackage[utf8]{inputenc}
@@
 analysis engines providing additional, more complex analysis options
-than those offered by the original (FORTRAN) code.
+than those offered by the original (Fortran) code.
 

### 2016-04-05 paper.Rmd
suppress nlme start message

@@
 
-```{r setup, include=FALSE}
+```{r setup, include=FALSE, cache=FALSE}
 library("knitr")
-opts_chunk$set(cache=TRUE)
+knitr::opts_chunk$set(cache=TRUE)
+knitr::opts_chunk$set(prompt = TRUE)
+options(prompt = 'R> ')
 ```
@@
 set.seed(15) # same results every time
-library("mgcv", quietly=TRUE) # for inSide
+suppressPackageStartupMessages(library("mgcv")) # for inSide
 

### 2016-04-05 paper.tex
manually add "R> " to code, urgh!

@@
 \begin{CodeChunk}
-\begin{CodeOutput}
-Warning: package 'nlme' was built under R version 3.2.4
-\end{CodeOutput}
 \begin{figure}
@@
 \begin{CodeInput}
-library("Distance")
-head(minke)
+R> library("Distance")
+R> head(minke)
 \end{CodeInput}
@@
 \begin{CodeInput}
-head(amakihi)
+R> head(amakihi)
 \end{CodeInput}
@@
 \begin{CodeInput}
-minke_hn <- ds(minke, truncation = 1.5)
+R> minke_hn <- ds(minke, truncation = 1.5)
 \end{CodeInput}
@@
 \begin{CodeInput}
-minke_hrcos <- ds(minke, truncation = 1.5, key = "hr")
+R> minke_hrcos <- ds(minke, truncation = 1.5, key = "hr")
 \end{CodeInput}
@@
 \begin{CodeInput}
-minke_unifcos <- ds(minke, truncation = 1.5, key = "unif",
+R> minke_unifcos <- ds(minke, truncation = 1.5, key = "unif",
                     adjustment = "cos", order = c(1, 2))
@@
 \begin{CodeInput}
-amakihi_hr_obs <- ds(amakihi, truncation = 82.5, transect = "point",
+R> amakihi_hr_obs <- ds(amakihi, truncation = 82.5, transect = "point",
                      key = "hr", formula = ~obs)
@@
 \begin{CodeInput}
-amakihi_hr_obs_mas <- ds(amakihi, truncation = 82.5, transect = "point",
+R> amakihi_hr_obs_mas <- ds(amakihi, truncation = 82.5, transect = "point",
                          key = "hr", formula = ~obs + mas)
@@
 \begin{CodeInput}
-summary(amakihi_hr_obs)
+R> summary(amakihi_hr_obs)
 \end{CodeInput}
@@
 \begin{CodeInput}
-amakihi_hn <- ds(amakihi, truncation = 82.5, transect = "point",
+R> amakihi_hn <- ds(amakihi, truncation = 82.5, transect = "point",
                  key = "hn", adjustment = NULL)
@@
 \begin{CodeInput}
-gof_ds(amakihi_hn)
+R> gof_ds(amakihi_hn)
 \end{CodeInput}
@@
 \begin{CodeInput}
-gof_ds(amakihi_hr_obs_mas)
+R> gof_ds(amakihi_hr_obs_mas)
 \end{CodeInput}
@@
 \begin{CodeInput}
-summarize_ds_models(amakihi_hn, amakihi_hr_obs, amakihi_hr_obs_mas)
+R> summarize_ds_models(amakihi_hn, amakihi_hr_obs, amakihi_hr_obs_mas)
 \end{CodeInput}
@@
 \begin{CodeInput}
-summary(minke_hn)
+R> summary(minke_hn)
 \end{CodeInput}
@@
 \begin{CodeInput}
-minke_table <- summary(minke_hn)$dht$individuals$N
-minke_table$lcl <- minke_table$ucl <- minke_table$df <- NULL
-colnames(minke_table) <- c("Stratum", "$\\hat{N}$", "$\\text{se}(\\hat{N}$)", 
+R> minke_table <- summary(minke_hn)$dht$individuals$N
+R> minke_table$lcl <- minke_table$ucl <- minke_table$df <- NULL
+R> colnames(minke_table) <- c("Stratum", "$\\hat{N}$", "$\\text{se}(\\hat{N}$)", 
                            "$\\text{CV}(\\hat{N}$)")

### 2016-04-05 paper.Rmd
use data from package, remove nlme warning

@@
 
-```{r points-and-lines, fig.height=6, fig.width=7, echo=FALSE, message=FALSE, fig.cap="Left side plots show an example of a survey of a (1x1) area containing a population of 500 objects; blue dashed lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals.
+```{r points-and-lines, fig.height=6, fig.width=7, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Left side plots show an example of a survey of a (1x1) area containing a population of 500 objects; blue dashed lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals.
 The right side of the figure shows histograms of observed distances (again, lines top and points bottom).\\label{fig:pointslines}"}
@@
 The minke whale data follows the "flatfile" format given in the last bullet point:
-```{r load-data, echo=FALSE}
-# quietly load the data
-load("data/minke.RData")
-load("data/amakihi.RData")
-```
 ```{r minke-data-head, message=FALSE}
 library("Distance")
+data(minke)
 head(minke)
@@
 ```{r amakihi-data-head}
+data(amakihi)
 head(amakihi)

### 2016-04-05 appendix-A.Rmd
appendix A

---
author:
  - name: David L Miller
    affiliation: University of St Andrews
  - name: Eric Rexstad
    affiliation: University of St Andrews
  - name: Len Thomas
    affiliation: University of St Andrews
  - name: Laura Marshall
    affiliation: University of St Andrews
  - name: Jeffrey L Laake
    affiliation: Marine Mammal Laboratory
title: "Appendix A: Feature comparison between \\pkg{Distance} and other packages"
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{longtable}
  \usepackage{booktabs}
output: html_document
bibliography: jstatsoft.bib

---

There are four packages available for analysis of distance sampling data in R that we are aware of. All are available on CRAN.

Package  | Description  | Version  | Homepage
-------: | -----------: | -------: | ---------:
`Distance` | package described in the main paper | 0.9.6 | https://github.com/distancedevelopment/Distance
`Rdistance` | uses a similar approach to `Distance` | 1.3.2 | https://github.com/tmcd82070/Rdistance
`unmarked` | uses a heirarchical modelling approach | 0.11.0 | https://sites.google.com/site/unmarkedinfo/
`mrds`     | on which `Distance` is based  | 2.1.15 | https://github.com/distancedevelopment/mrds

This table provides a feature comparison of the packages:

Feature    |  `Distance` | `Rdistance` | `unmarked` | `mrds`
-------: | --------:  | ---------: | --------: | ----:
Line transects | x | x | x | x
Point transets | x | | x | x
Interval (binned) distances | x | | x | x
Exact distances | x | x | | x
Continuous individual level covariates | x |  | | x
Factor individual level covariates | x | x | | x
Transect level covariates | x |  | x | x
Objects in clusters | x | x | x | x
Left truncation | x  | x | | x
Half-normal key | x | x | x | x
Hazard-rate key | x | x |  x | x
Uniform key | x | x | x | x
Gamma key |  | x |  | x
Negative exponential key |  |  | x  |
Adjustment terms | x | x | | x
AIC adjustment selection | x | x | 
Monotonicity constraints | x |  | | x
Availability bias model | | | x |
Perception bias model | | | | x
Abundance estimation | x | x | x | x
Density estimation | x | x | x | x
User-defined likelihood functions | | x | |

## Abundance estimation

`Distance` and `mrds` models can be used as part of a density surface model using `dsm`, which allows abundance to be modelled as a function of spatially varying covariates (such as location, sea depth, altitude etc). See @Miller:2013fq for more information.

`unmarked` also allows abundance to vary accoring to covariates, via the abundance part of the likelihood. See @Fiske:2011gv for more information on the package and @Royle:2004cd for more information on methodology.

Results `Rdistance` models can be use in combination with R modelling functions such as `lm`, `glm` etc to build abundance estimates which vary according to covariates. More information is available on the project's wiki.

# References

### 2016-04-05 appendix-A.Rmd
revise markup and affiliation

@@
   - name: David L Miller
-    affiliation: University of St Andrews
+    affiliation: Integrated Statistics and University of St Andrews
   - name: Eric Rexstad
@@
     affiliation: Marine Mammal Laboratory
-title: "Appendix A: Feature comparison between \\pkg{Distance} and other packages"
+title: "Appendix A: Feature comparison between `Distance` and other packages"
 preamble: >

### 2016-05-15 paper.Rmd
e.g. and i.e. -> e.g., and i.e.,

@@
 In addition to randomly placed samplers distance sampling relies on three other main assumptions.
-Firstly, all objects directly on the transect line or point (i.e.
-those at zero distance) are detected (see "Extensions" for methods to deal with the situation when this is not possible).
+Firstly, all objects directly on the transect line or point (i.e., those at zero distance) are detected (see "Extensions" for methods to deal with the situation when this is not possible).
 Secondly, objects are stationary or detected prior to any movement.
@@
 where $k$ is the key function and $\alpha_O$ is sum series of functions (given in Table \ref{tab:keyadj}), described as an *adjustment of order $O$*.
-Subscripts on the parameter vector indicate those parameters belonging to each part of the model (i.e.
-$\boldsymbol{\theta} = (\boldsymbol{\theta}_\text{key}, \boldsymbol{\theta}_\text{adjust})$).
+Subscripts on the parameter vector indicate those parameters belonging to each part of the model (i.e., $\boldsymbol{\theta} = (\boldsymbol{\theta}_\text{key}, \boldsymbol{\theta}_\text{adjust})$).
 
@@
 
-Once we have a set of plausible models, we can use Akaike's Information Criterion (AIC) to select between models [see e.g.
-@burnham2003model].
+Once we have a set of plausible models, we can use Akaike's Information Criterion (AIC) to select between models [see e.g., @burnham2003model].
 \pkg{Distance} includes a function to create table of summary information for fitted models, making it easy to get an overview of a large number of models.
@@
 \pkg{Distance} interfaces with one such package used to perform this type of analysis: \pkg{dsm} [@dsm-pkg].
-So-called "density surface modelling" uses the generalized additive model framework [e.g.
-@Wood:2006vg] to build models of abundance (adjusting counts for imperfect detectability) as a function of environmental covariates, as part of a two stage model [@Hedley:2004et; @Miller:2013fq].
+So-called "density surface modelling" uses the generalized additive model framework [e.g., @Wood:2006vg] to build models of abundance (adjusting counts for imperfect detectability) as a function of environmental covariates, as part of a two stage model [@Hedley:2004et; @Miller:2013fq].
 

### 2016-05-15 paper.Rmd
quotes when using data()

@@
 library("Distance")
-data(minke)
+data("minke")
 head(minke)
@@
 ```{r amakihi-data-head}
-data(amakihi)
+data("amakihi")
 head(amakihi)

### 2016-05-15 paper.Rmd
transposes

@@
 $$
-\widehat{\text{Var}}_\text{model}\left( \hat{N}_\text{C}\right) = \left(\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}\right)^\text{T} \left(\hat{\mathbf{H}}(\hat{\boldsymbol{\theta}})^{-1} \right)\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}
+\widehat{\text{Var}}_\text{model}\left( \hat{N}_\text{C}\right) = \left(\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}\right)^\top \left(\hat{\mathbf{H}}(\hat{\boldsymbol{\theta}})^{-1} \right)\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}
 $$

### 2016-05-29 paper.Rnw
compiling Rnw, though far from perfect

--- 2016-05-15 paper.Rmd
@@
----
-author:
-  - name: David L Miller
-    affiliation: Integrated Statistics and University of St Andrews
-    address: >
-      16 Sumner Street, Woods Hole, MA 02543, USA and
-      Centre for Research into Ecological and Environmental Modelling,
-      The Observatory,
-      St Andrews, Fife KY16 9LZ, Scotland
-    email: dave@ninepointeightone.net
-    url: http://converged.yt
-  - name: Eric Rexstad
-    affiliation: University of St Andrews
-    address: >
-      Centre for Research into Ecological and Environmental Modelling,
-      The Observatory,
-      St Andrews, Fife KY16 9LZ, Scotland
-    email: Eric.Rexstad@st-andrews.ac.uk
-  - name: Len Thomas
-    affiliation: University of St Andrews
-    address: >
-      Centre for Research into Ecological and Environmental Modelling,
-      The Observatory,
-      St Andrews, Fife KY16 9LZ, Scotland
-    email: len.thomas@st-andrews.ac.uk
-    url: http://lenthomas.org
-  - name: Laura Marshall
-    affiliation: University of St Andrews
-    address: >
-      Centre for Research into Ecological and Environmental Modelling,
-      The Observatory,
-      St Andrews, Fife KY16 9LZ, Scotland
-    email: lhm@st-andrews.ac.uk
-  - name: Jeffrey L Laake
-    affiliation: Marine Mammal Laboratory
-    address: >
-      Alaska Fisheries Science Center
-      National Oceanic and Atmospheric Administration
-      7600 Sand Point Way N.E., Seattle, WA 98115, USA
-    email: Jeff.Laake@noaa.gov
-title:
-  formatted: "Distance Sampling in \\proglang{R}"
-  # If you use tex in the formatted title, also supply version without
-  plain:     "Distance Sampling in R"
-  # For running headers, if needed
-  #short:     "\\pkg{foo}: A Capitalized Title"
-abstract: >
+\documentclass[article]{jss}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\usepackage{amsmath}
+\usepackage{amssymb}
+%% need no \usepackage{Sweave.sty}
+
+%\usepackage{longtable}
+\usepackage{booktabs}
+
+%% almost as usual
+\author{David L Miller\\Integrated Statistics and University of St Andrews \AND 
+        Eric Rexstad\\University of St Andrews \And 
+        Len Thomas\\University of St Andrews \AND 
+        Laura Marshall\\University of St Andrews \And 
+        Jeffrey L Laake\\Marine Mammal Laboratory}
+
+\title{Distance Sampling in \proglang{R}}
+
+%% for pretty printing and a nice hypersummary also set:
+\Plainauthor{David L Miller, Eric Rexstad, Len Thomas, Laura Marshall, Jeffrey L Laake}
+\Plaintitle{Distance Sampling in R} %% without formatting
+%\Shorttitle{\pkg{foo}: A Capitalized Title} %% a short title (if necessary)
+
+%% an abstract and keywords
+\Abstract{
 Estimating the abundance and spatial distribution of animal and plant populations is essential for conservation and management.
@@
 We also demonstrate how \pkg{Distance} provides a migration pathway from previous software, thereby allowing us to deliver cutting-edge methods to the users more quickly.
+}
+\Keywords{distance sampling, abundance estimation, line transect, point transect, detection function, Horvitz-Thompson, \proglang{R}, Distance}
+\Plainkeywords{distance sampling, abundance estimation, line transect, point transecs, detection function, Horvitz-Thompson, R, Distance} %% without formatting
+%% at least one keyword must be supplied
 
-keywords:
-  # at least one keyword must be supplied
-  formatted: [distance sampling, abundance estimation, line transect, point transect, detection function, Horvitz-Thompson, "\\proglang{R}", Distance]
-  plain:     [distance sampling, abundance estimation, line transect, point transecs, detection function, Horvitz-Thompson, R, Distance]
-preamble: >
-  \usepackage{amsmath}
-  \usepackage{amssymb}
-  \usepackage{longtable}
-  \usepackage{booktabs}
-  \usepackage[colorinlistoftodos, color=green!40, textwidth=60, textsize=small]{todonotes}
+%% publication information
+%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
+%% \Volume{50}
+%% \Issue{9}
+%% \Month{June}
+%% \Year{2012}
+%% \Submitdate{2012-06-04}
+%% \Acceptdate{2012-06-04}
 
-output:
-  rticles::jss_article:
-    citation_package: natbib
-bibliography: jstatsoft.bib
----
+%% The address of (at least) one author should be given
+%% in the following format:
+\Address{
+  David L Miller\\
+  Integrated Statistics and University of St Andrews\\
+  16 Sumner Street, Woods Hole, MA 02543, USA and\\
+  Centre for Research into Ecological and Environmental Modelling,\\
+  The Observatory,\\
+  St Andrews, Fife KY16 9LZ, Scotland\\
+  E-mail: \email{dave@ninepointeightone.net}\\
+  URL: \url{http://converged.yt} %\And
+%  Eric Rexstad\\
+%  University of St Andrews\\
+%  Centre for Research into Ecological and Environmental Modelling,\\
+%  The Observatory,\\
+%  St Andrews, Fife KY16 9LZ, Scotland\\
+%  E-mail: \email{Eric.Rexstad@st-andrews.ac.uk} \And
+%  Len Thomas\\
+%  University of St Andrews\\
+%  Centre for Research into Ecological and Environmental Modelling,\\
+%  The Observatory,\\
+%  St Andrews, Fife KY16 9LZ, Scotland\\
+%  E-mail: \email{len.thomas@st-andrews.ac.uk}\\
+%  URL: \url{http://lenthomas.org} \And
+%  Laura Marshall\\
+%  University of St Andrews\\
+%  Centre for Research into Ecological and Environmental Modelling,\\
+%  The Observatory,\\
+%  St Andrews, Fife KY16 9LZ, Scotland\\
+%  E-mail: \email{lhm@st-andrews.ac.uk}\And
+%  Jeffrey L Laake\\
+%  Marine Mammal Laboratory\\
+%  Alaska Fisheries Science Center\\
+%  National Oceanic and Atmospheric Administration\\
+%  7600 Sand Point Way N.E., Seattle, WA 98115, USA\\
+%  E-mail: \email{Jeff.Laake@noaa.gov}
+}
+%% It is also possible to add a telephone and fax number
+%% before the e-mail in the following format:
+%% Telephone: +43/512/507-7103
+%% Fax: +43/512/507-2851
 
-# Introduction
+%% for those who use Sweave please include the following line (with % symbols):
+%% need no \usepackage{Sweave.sty}
 
-```{r setup, include=FALSE, cache=FALSE}
-library("knitr")
-knitr::opts_chunk$set(cache=TRUE)
-knitr::opts_chunk$set(prompt = TRUE)
+%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{document}
+%\SweaveOpts{concordance=TRUE}
+
+\section{Introduction}
+<<label=setup, echo=FALSE>>=
 options(prompt = 'R> ')
-```
+@
 
-Distance sampling [@Buckland:2001vm; @Buckland:2004ts; @buckland2015distance] encompasses a suite of methods used to estimate the density and/or abundance of biological populations.
+Distance sampling \citep{Buckland:2001vm, Buckland:2004ts, buckland2015distance} encompasses a suite of methods used to estimate the density and/or abundance of biological populations.
 Distance sampling can be thought of as an extension of plot sampling.
@@
 The further away an object is from the point or line (also known as the sampler or transect) the less likely it is that the observer will see it.
-We can use the distances to each of the detected objects from the line or point to build a model of the probability of detection given distance from the sampler -- the *detection function*.
+We can use the distances to each of the detected objects from the line or point to build a model of the probability of detection given distance from the sampler --- the \textit{detection function}.
 The detection function can be used to infer how many objects were missed and thereby produce estimates of density and/or abundance.
@@
 
-## Distance sampling
+\subsection{Distance sampling}
 
@@
 
-```{r detectdists, fig.height=5.5, fig.width=9, echo=FALSE, message=FALSE, fig.cap="Panels show an example detection function (left), the pdf of object distances (middle) and the resulting pdf of detection distances (right) for line transects (top row) and point transects (bottom row).
-The pdf of observed detection distances in the right hand plots are obtained by multiplying the detection function by the pdf of object distances and rescaling.
-In this example, detection probability becomes effectively zero at 500 (distances shown on $x$-axis are arbitrary).
-\\label{fig:detectiondist}"}
+\begin{figure}
+\begin{center}
+<<label=detectdists, fig=TRUE, echo=FALSE, message=FALSE>>=
 
@@
 axis(2, at=round(c(0,max(pdf.observed)/2,max(pdf.observed)),4)); axis(1); box()
-```
+@
+\caption{Panels show an example detection function (left), the pdf of object distances (middle) and the resulting pdf of detection distances (right) for line transects (top row) and point transects (bottom row).
+The pdf of observed detection distances in the right hand plots are obtained by multiplying the detection function by the pdf of object distances and rescaling.
+In this example, detection probability becomes effectively zero at 500 (distances shown on $x$-axis are arbitrary).}
+\label{fig:detectiondist}
+\end{center}
+\end{figure}
 
@@
 
-```{r points-and-lines, fig.height=6, fig.width=7, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Left side plots show an example of a survey of a (1x1) area containing a population of 500 objects; blue dashed lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals.
-The right side of the figure shows histograms of observed distances (again, lines top and points bottom).\\label{fig:pointslines}"}
+\begin{figure}
+\begin{center}
+<<label=points-and-lines, fig=TRUE, echo=FALSE, message=FALSE, warning=FALSE>>=
 opar <- par(mfrow=c(2,2))
@@
 par(opar)
-```
+@
+\caption{"Left side plots show an example of a survey of a (1x1) area containing a population of 500 objects; blue dashed lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals.
+The right side of the figure shows histograms of observed distances (again, lines top and points bottom).}
+\label{fig:pointslines}
+\end{center}
+\end{figure}
 
@@
 
-Having estimated the detection function's parameters, one can then integrate out distance from the function (as the detection function describes the probability of detection _given_ distance) to get an "average" probability of detection, which can be used to correct the observed counts.
+Having estimated the detection function's parameters, one can then integrate out distance from the function (as the detection function describes the probability of detection \textit{given} distance) to get an ``average'' probability of detection, which can be used to correct the observed counts.
 Summing the corrected counts gives an estimate of abundance in the area covered by surveys, which can be multiplied up to the total study area.
@@
 
-# Data
+\section{Data}
 
@@
 
-### Minke whales
+\subsection{Minke whales}
 
-The line transect data have been simulated from models fitted to Antarctic minke whale (*Balaenoptera bonaerensis*) data.
+The line transect data have been simulated from models fitted to Antarctic minke whale (\textit{Balaenoptera bonaerensis}) data.
 These data were collected as part of the International Whaling Commission's International Decade of Cetacean Research Southern Ocean Whale and Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer surveys.
@@
 
-### Amakihi
+\subsection{Amakihi}
 
-The point transect data set consists of 1485 observations of Amakihi (*Hemignathus virens*; a Hawaiian songbird), collected at 41 points between 1992 and 1995.
+The point transect data set consists of 1485 observations of Amakihi (\textit{Hemignathus virens}; a Hawaiian songbird), collected at 41 points between 1992 and 1995.
 The data include distances and two covariates collected during the survey: observer (a three level factor), time after sunrise (transformed to minutes (continuous) or hours (factor) covariates).
@@
 
-## Data setup
+\subsection{Data setup}
 
@@
 
-* In the simplest case, where the objective is to estimate a detection function and exact distances are collected, all that is required is a numeric vector of distances.
-* To include additional covariates into the detection function (see "Detection functions") a \code{data.frame} is required.
+\begin{itemize}
+\item In the simplest case, where the objective is to estimate a detection function and exact distances are collected, all that is required is a numeric vector of distances.
+\item To include additional covariates into the detection function (see "Detection functions") a \code{data.frame} is required.
 Each row of the \code{data.frame} contains the data on one observation.
@@
 Additional reserved names include \code{object} and \code{detected}, these are not required for conventional distance sampling and should be avoided (see "Extensions" for an explanation of their use).
-* To estimate density or to estimate abundance beyond the sampled area, additional information is required.
+\item To estimate density or to estimate abundance beyond the sampled area, additional information is required.
 Additional columns should be included in the \code{data.frame} specifying: \code{Sample.Label}, the ID of the transect; \code{Effort}, transect effort (for lines their length and for points the number of times that point was visited); \code{Region.Label}, the stratum containing the transect (which may be from pre- or post-survey stratification, see "Estimating abundance and variance"); \code{Area}, the area of the strata.
@@
 We refer to this data format (where all information is contained in one table) as "flatfile" as it can be easily created in a single spreadsheet.
+\end{itemize}
 
@@
 The minke whale data follows the "flatfile" format given in the last bullet point:
-```{r minke-data-head, message=FALSE}
+<<label=minke-data-head, message=FALSE>>=
 library("Distance")
@@
 head(minke)
-```
+@
 Whereas the amakihi data lacks effort and stratum data:
-```{r amakihi-data-head}
+<<label=amakihi-data-head>>=
 data("amakihi")
 head(amakihi)
-```
+@
 
@@
 
-# Detection functions
+\section{Detection functions}
 
 The detection function models the probability $\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)$ and is usually denoted $g(y; \boldsymbol{\theta})$ where $y$ is distance (from a line or point) and $\boldsymbol{\theta}$ is a vector of parameters to be estimated.
-Our goal is to estimate an *average probability of detection* ($p$, average in the sense of an average over distance from $0$ to truncation distance $w$), so we must integrate out distance ($y$) from the detection function:
+Our goal is to estimate an \textit{average probability of detection} ($p$, average in the sense of an average over distance from $0$ to truncation distance $w$), so we must integrate out distance ($y$) from the detection function:
 $$
@@
 $$
-where $\pi(y)$ describes the distribution of objects with respect to the sampler; $\pi(x)=1/w$ for line transects and $\pi(r)=\frac{2r}{w^2}$ for point transects, taking into account the geometry of the sampler [usually referred to as the _probability density function of distances_; @Buckland:2001vm, Chapter 3].
+where $\pi(y)$ describes the distribution of objects with respect to the sampler; $\pi(x)=1/w$ for line transects and $\pi(r)=\frac{2r}{w^2}$ for point transects, taking into account the geometry of the sampler [usually referred to as the \textit{probability density function of distances}; @Buckland:2001vm, Chapter 3].
 When considering a particular transect type we let $x$ denote a perpendicular distance from a line and $r$ denote radial distance from a point (rather than using $y$).
@@
 
-* *Shoulder*: we expect observers to be able to see objects near them, not just those directly in front of them. For this reason, we expect the detection function to be flat near the line or point.
-* *Non-increasing*: we do not think that observers should be more likely to see distant objects than those nearer the transect.
+\begin{itemize}
+\item \textit{Shoulder}: we expect observers to be able to see objects near them, not just those directly in front of them.
+For this reason, we expect the detection function to be flat near the line or point.
+\item \textit{Non-increasing}: we do not think that observers should be more likely to see distant objects than those nearer the transect.
 If this occurs, it usually indicates an issue with survey design or field procedures (for example that the distribution of objects with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this [@Marques:2010he; @Marques:2012fy; @Miller:2015hw].
-* *Model robust*: models should be flexible enough to fit many different shapes.
-* *Pooling robust*: many factors can affect the probability of detection and it is not possible to measure all of these.
+\item \textit{Model robust}: models should be flexible enough to fit many different shapes.
+\item \textit{Pooling robust}: many factors can affect the probability of detection and it is not possible to measure all of these.
 We would like models to produce unbiased results without inclusion of these factors.
-* *Estimator efficiency*: we would like models to have low variances, given they satisfy the other properties above (which, if satisfied, would give low bias).
+\item \textit{Estimator efficiency}: we would like models to have low variances, given they satisfy the other properties above (which, if satisfied, would give low bias).
+\end{itemize}
 
@@
 
-## Formulations
+\subsection{Formulations}
 
@@
 
-### Key function plus adjustments (K+A)
+\subsubsection{Key function plus adjustments (K+A)}
 
@@
 $$
-where $k$ is the key function and $\alpha_O$ is sum series of functions (given in Table \ref{tab:keyadj}), described as an *adjustment of order $O$*.
+where $k$ is the key function and $\alpha_O$ is sum series of functions (given in Table \ref{tab:keyadj}), described as an \textit{adjustment of order $O$}.
 Subscripts on the parameter vector indicate those parameters belonging to each part of the model (i.e., $\boldsymbol{\theta} = (\boldsymbol{\theta}_\text{key}, \boldsymbol{\theta}_\text{adjust})$).
@@
 
-```{r hn-hr-par-comp, fig.width=10, fig.height=5, fig.cap="Half-normal (top row) and hazard-rate (bottom row) detection functions without adjustments, varying scale ($\\sigma$) and (for hazard-rate) shape ($b$) parameters (values are given above the plots).
-On the top row from left to right, the study species becomes more detectable (higher probability of detection at larger distances).
-The bottom row shows the hazard-rate model's more pronounced shoulder.\\label{fig:hnhr}", echo=FALSE}
+\begin{figure}
+\begin{center}
+<<label=hn-hr-par-comp, fig=TRUE, echo=FALSE>>=
 par(mfrow=c(2,4), mar=c(3.5, 3, 2, 1) + 0.1)
@@
 }
-```
+@
+\caption{Half-normal (top row) and hazard-rate (bottom row) detection functions without adjustments, varying scale ($\sigma$) and (for hazard-rate) shape ($b$) parameters (values are given above the plots).
+On the top row from left to right, the study species becomes more detectable (higher probability of detection at larger distances).
+The bottom row shows the hazard-rate model's more pronounced shoulder.}
+\label{fig:hnhr}
+\end{center}
+\end{figure}
 
@@
 
-It is not always necessary to include adjustments (except in the case of the uniform key) and in such cases we refer to these as "key only" models (see the next section and "Model checking and model selection").
+It is not always necessary to include adjustments (except in the case of the uniform key) and in such cases we refer to these as ``key only'' models (see the next section and "Model checking and model selection").
 
-```{r adjust-mix-comp, fig.width=10, fig.height=3, fig.cap="Possible shapes for the detection function when cosine adjustments are included for half-normal and hazard-rate models.\\label{fig:keyadj}", echo=FALSE}
+\begin{figure}
+\begin{center}
+<<label=adjust-mix-comp, fig=TRUE, echo=FALSE>>=
 par(mfrow=c(1,4), mar=c(3.5, 3, 3, 1) + 0.1)
@@
 title(ylab="Detection probability", line=2)
-```
+@
+\caption{Possible shapes for the detection function when cosine adjustments are included for half-normal and hazard-rate models.}
+\label{fig:keyadj}
+\end{center}
+\end{figure}
 
-### Covariates
+\subsubsection{Covariates}
 
@@
 where $\mathbf{z_i}$ is the vector of $J$ covariates associated with observation $i$.
-For covariate models, we calculate a value of "average" probability of detection (average in the sense of distance being integrated out) per observation.
+For covariate models, we calculate a value of ``average'' probability of detection (average in the sense of distance being integrated out) per observation.
 There are as many unique values of $p(\mathbf{z_i})$ as there are unique covariate combinations in our data.
@@
 
-## Fitting detection functions in R
+\subsection{Fitting detection functions in R}
 
@@
 
-### Minke whale
+\subsubsection{Minke whale}
 
 First we fit a model to the minke whale data, setting the truncation at 1.5km and using the default options in \code{ds} very simply:
-```{r minke-hn}
+<<label=minke-hn>>=
 minke_hn <- ds(minke, truncation = 1.5)
-```
+@
 Note that when there are no covariates in the model, \code{ds} will add adjustment terms to the model until there is no improvement in AIC.
@@
 A different form for the detection function can be specified via the \code{key=} argument to \code{ds}. For example, a hazard rate model can be fitted as:
-```{r minke-hr-cos}
+<<label=minke-hr-cos>>=
 minke_hrcos <- ds(minke, truncation = 1.5, key = "hr")
-```
+@
 Here \code{ds} also fits the hazard-rate model then hazard-rate with a cosine adjustment but the AIC improvement is insufficient to select the adjustment, so the hazard-rate key-only model is returned.
@@
 For example, to specify a uniform model with cosine adjustments of order 1 and 2 we can write:
-```{r minke-unif-cos}
+<<label=minke-unif-cos>>=
 minke_unifcos <- ds(minke, truncation = 1.5, key = "unif",
                     adjustment = "cos", order = c(1, 2))
-```
+@
 Hermite polynomial adjustments use the code `"herm"` and simple polynomials `"poly"`, adjustment order should be in line with Table \ref{tab:keyadj}.
 
-### Amakihi
+\subsubsection{Amakihi}
 
@@
 A hazard-rate model for the amakihi that includes observer as a covariate and a truncation distance of 82.5m [@Marques:2007ey] can be specified using :
-```{r amakihi-hr-obs}
+<<label=amakihi-hr-obs>>=
 amakihi_hr_obs <- ds(amakihi, truncation = 82.5, transect = "point",
                      key = "hr", formula = ~obs)
-```
+@
 Note that here, unlike with the minke whale data, \code{ds} warns us that we have only supplied enough information to estimate the detection function (not density or abundance).
@@
 We will compare these two models further in the following section.
-```{r amakihi-hr-obs-mas, warning=FALSE}
+<<label=amakihi-hr-obs-mas, warning=FALSE>>=
 amakihi_hr_obs_mas <- ds(amakihi, truncation = 82.5, transect = "point",
                          key = "hr", formula = ~obs + mas)
-```
+@
 As with the minke whale model, we can plot the resulting models (Figure \ref{fig:minkeamakihi}, middle and right panels).
@@
 Points on the plot indicate probability of detection for each observation.
-For the `amakihi_hr_obs` model we see fairly clear levels of the observer covariate in the points.
+For the \code{amakihi_hr_obs} model we see fairly clear levels of the observer covariate in the points.
 Looking at the right panel of Figure \ref{fig:minkeamakihi}, this is less clear when adding minutes after sunrise (a continuous covariates) to the model.
 
-```{r minke-amakihi-hn-plot, echo=FALSE, warning=FALSE, fig.width=9, fig.height=4, fig.cap="Left: fitted detection function overlayed on the histogram of observed distances for the minke whale data using half-normal model.
-Centre and right: plots of the probability density function for the amakihi models.
-Centre, hazard-rate with observer as a covariate; right, hazard-rate model with observer and minutes after sunrise as covariates.
-Points indicate probability of detection for a given observation (given that observation's covariate values) and lines indicate the average detection function.\\label{fig:minkeamakihi}"}
+\begin{figure}
+\begin{center}
+<<label=minke-amakihi-hn-plot, echo=FALSE, warning=FALSE, fig=TRUE>>=
 par(mfrow=c(1, 3))
@@
 title("Amakihi\n(observer+minutes)", cex.main=1.5)
-```
+@
+\caption{Left: fitted detection function overlayed on the histogram of observed distances for the minke whale data using half-normal model.
+Centre and right: plots of the probability density function for the amakihi models.
+Centre, hazard-rate with observer as a covariate; right, hazard-rate model with observer and minutes after sunrise as covariates.
+Points indicate probability of detection for a given observation (given that observation's covariate values) and lines indicate the average detection function.}
+\label{fig:minkeamakihi}
+\end{center}
+\end{figure}
 
-# Model checking and model selection
+\section{Model checking and model selection}
 
@@
 For our hazard-rate model of the amakihi data, with observer as a covariate:
-```{r amakihi-summary}
+<<label=amakihi-summary>>=
 summary(amakihi_hr_obs)
-```
+@
 This summary information includes details of the data and model specification, as well as the values of the coefficients ($\beta_j$) and their uncertainties, an "average" value for the detectability (see "Estimating abundance and variance" for details on how this is calculated) and its uncertainty.
@@
 
-## Goodness of fit
+\subsection{Goodness of fit}
 
@@
 We first fit a half-normal model without covariates or adjustments (setting \code{adjustment=NULL} will force \code{ds} to fit a model with no adjustments):
-```{r amakihi-gof, fig.keep="none"}
+<<label=amakihi-gof, grdevice=NULL>>=
 amakihi_hn <- ds(amakihi, truncation = 82.5, transect = "point",
@@
 gof_ds(amakihi_hr_obs_mas)
-```
+@
 We can conclude that the half-normal model does not pass our goodness of fit tests and should be discarded.
@@
 
-```{r amakihi-qq-comp, fig.width=6, fig.height=3, fig.cap="Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates; left) and hazard-rate model with observer and minutes after sunrise (right) for the amakihi data.\\label{amakihi-qq}", echo=FALSE, results="hide"}
+\begin{figure}
+\begin{center}
+<<label=amakihiqq, fig=TRUE, echo=FALSE, message=FALSE>>=
 par(mfrow=c(1,2))
@@
 gof_ds(amakihi_hr_obs_mas, asp=1, cex=0.3)
-```
+@
+\caption{Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates; left) and hazard-rate model with observer and minutes after sunrise (right) for the amakihi data.}
+\label{amakihi-qq}
+\end{center}
+\end{figure}
 
 For non-exact distance data, a $\chi^2$-test can be used to assess goodness of fit [see @Buckland:2001vm, Section 3.4.4].
-$\chi^2$-test results are produced by `gof_ds` when binned/grouped distances are provided.
+$\chi^2$-test results are produced by \code{gof_ds} when binned/grouped distances are provided.
 
-## Model selection
+\subsection{Model selection}
 
@@
 
-```{r summary-table-null, eval=FALSE}
+\begin{Code}
 summarize_ds_models(amakihi_hn, amakihi_hr_obs, amakihi_hr_obs_mas)
-```
+\end{Code}
 
 In this case we may be skeptical about the top model as selected by AIC being truly better than the second best model, as there is only a very small difference in AICs.
-Generally, if the difference between AICs is less than 2, we may investigate multiple "best" models, potentially resorting to the simplest of these models.
+Generally, if the difference between AICs is less than 2, we may investigate multiple ``best'' models, potentially resorting to the simplest of these models.
 In the authors' experience, it is often the case that models with similar AICs will have similar estimates probabilities of detection, so in practice there is little difference in selecting between these models.
@@
 
-```{r summary-table, results="asis", echo=FALSE}
+<<label=summary-table, results="tex", echo=FALSE>>=
 kable(summarize_ds_models(amakihi_hn, amakihi_hr_obs, amakihi_hr_obs_mas),
@@
 Models are sorted according to AIC.\\label{tab:amakihi}")
-```
+@
 
-# Estimating abundance and variance
+\section{Estimating abundance and variance}
 
@@
 
-## Abundance
+\subsection{Abundance}
 
 We wish to obtain the abundance in a study region, of which we have sampled a random subset.
-To do this we first calculate the abundance in the area we have surveyed (the *covered area*) to obtain $\hat{N}_\text{C}$, we can then scale this up to the full study area by multiplying it by the ratio of covered area to study area.
+To do this we first calculate the abundance in the area we have surveyed (the \textit{covered area}) to obtain $\hat{N}_\text{C}$, we can then scale this up to the full study area by multiplying it by the ratio of covered area to study area.
 We discuss other methods for spatially explicit abundance estimation in "Extensions".
@@
 
-### Stratification
+\subsubsection{Stratification}
 
-We may wish to calculate abundance estimates for some sub-regions of the study region, we call these areas *strata*.
-Stratification can be used to increase the precision of estimates if we know *a priori* that density varies between different parts of the study area.
+We may wish to calculate abundance estimates for some sub-regions of the study region, we call these areas \textit{strata}.
+Stratification can be used to increase the precision of estimates if we know \textit{a priori} that density varies between different parts of the study area.
 For example, strata may be defined by habitat types or animal sex (or some combination) which may be of interest for biological or management reasons.
@@
 
-## Variance
+\subsection{Variance}
 
@@
 
-1. *Detection function*: Uncertainty in parameter ($\boldsymbol{\theta}$) estimation.
-2. *Encounter rate*: Sampling variability arising from differences in the number of observations per transect.
+\begin{enumerate}
+\item \textit{Detection function}: Uncertainty in parameter ($\boldsymbol{\theta}$) estimation.
+\item \textit{Encounter rate}: Sampling variability arising from differences in the number of observations per transect.
+\end{enumerate}
 
@@
 
-*Encounter rate* is the number of objects per unit transect (rather than just $n$).
+\textit{Encounter rate} is the number of objects per unit transect (rather than just $n$).
 When covariates are not included in the detection function we can define the encounter rate as $n/L$ for line transects (where $L$ is the total line length) or $n/T$ for point transects (where $T$ is the total number of visits summed over all points).
@@
 
-For line transects, by default, \pkg{Distance} uses a variation of the estimator "R2" from @Fewster:2009ku replacing number of observations per sample with the estimated abundance per sample, thus taking into account cluster size if this is recorded [@Innes:2002ka; @Marques:2003vb]:
+For line transects, by default, \pkg{Distance} uses a variation of the estimator ``R2'' from @Fewster:2009ku replacing number of observations per sample with the estimated abundance per sample, thus taking into account cluster size if this is recorded [@Innes:2002ka; @Marques:2003vb]:
 $$
@@
 where $l_k$ are the lengths of the $K$ transects (such that $L = \sum_{k=1}^K l_k$) and $\hat{N_{C,k}}$ is the abundance in the covered area for transect $k$.
-For point transects we use the estimator "P3" from @Fewster:2009ku but again replace $n$ by $\hat{N}_\text{C}$ in the encounter rate definition, to obtain the following estimator:
+For point transects we use the estimator ``P3'' from @Fewster:2009ku but again replace $n$ by $\hat{N}_\text{C}$ in the encounter rate definition, to obtain the following estimator:
 $$
@@
 
-## Estimating abundance and variance in R
+\subsection[Estimating abundance and variance in R]{Estimating abundance and variance in \proglang{R}}
 
@@
 Having already fitted a model to the minke whale data, we can see the results of the abundance estimation by viewing the model summary:
-```{r minke-N-summary}
+<<label=minke-N-summary>>=
 summary(minke_hn)
-```
+@
 This prints a rather large amount of information: first the detection function summary, then three tables:
@@
 
-```{r minke-abundance-table-code, eval=FALSE}
+\begin{Code}
 minke_table <- summary(minke_hn)$dht$individuals$N
@@
                            "$\\text{CV}(\\hat{N}$)")
-```
+\end{Code}
 
-```{r minke-abundance-table, echo=FALSE}
+<<label=minke-abundance-table, echo=FALSE>>=
 minke_table <- summary(minke_hn)$dht$individuals$N
 minke_table$lcl <- minke_table$ucl <- minke_table$df <- NULL
-colnames(minke_table) <- c("Stratum", "$\\hat{N}$", "$\\text{se}(\\hat{N}$)", 
-                           "$\\text{CV}(\\hat{N}$)")
-kable(minke_table, digits=3, format="latex", booktabs=TRUE,
-      row.names=FALSE, escape=FALSE,caption="Summary of abundance estimation for the half-normal model for the minke whale data.\\label{minke-abund}")
-```
+colnames(minke_table) <- c("Stratum", "$\\hat{N}$", "$\\text{se}(\\hat{N}$)", "$\\text{CV}(\\hat{N}$)")
+kable(minke_table, digits=3, format="latex", booktabs=TRUE, row.names=FALSE, escape=FALSE,caption="Summary of abundance estimation for the half-normal model for the minke whale data.\\label{minke-abund}")
+@
 
-# Extensions
+\section{Extensions}
 
@@
 
-# Conclusion
+\section{Conclusion}
 
@@
 
-# Acknowledgements
+\section{Acknowledgements}
 
@@
 
-# Bibliography
+\bibliography{jstatsoft}
+
+\end{document}

### 2016-06-08 paper.Rnw
got through some of the requirements

@@
   E-mail: \email{dave@ninepointeightone.net}\\
-  URL: \url{http://converged.yt} %\And
-%  Eric Rexstad\\
-%  University of St Andrews\\
-%  Centre for Research into Ecological and Environmental Modelling,\\
-%  The Observatory,\\
-%  St Andrews, Fife KY16 9LZ, Scotland\\
-%  E-mail: \email{Eric.Rexstad@st-andrews.ac.uk} \And
-%  Len Thomas\\
-%  University of St Andrews\\
-%  Centre for Research into Ecological and Environmental Modelling,\\
-%  The Observatory,\\
-%  St Andrews, Fife KY16 9LZ, Scotland\\
-%  E-mail: \email{len.thomas@st-andrews.ac.uk}\\
-%  URL: \url{http://lenthomas.org} \And
-%  Laura Marshall\\
-%  University of St Andrews\\
-%  Centre for Research into Ecological and Environmental Modelling,\\
-%  The Observatory,\\
-%  St Andrews, Fife KY16 9LZ, Scotland\\
-%  E-mail: \email{lhm@st-andrews.ac.uk}\And
-%  Jeffrey L Laake\\
-%  Marine Mammal Laboratory\\
-%  Alaska Fisheries Science Center\\
-%  National Oceanic and Atmospheric Administration\\
-%  7600 Sand Point Way N.E., Seattle, WA 98115, USA\\
-%  E-mail: \email{Jeff.Laake@noaa.gov}
+  URL: \url{http://converged.yt} \\ \\
+  Eric Rexstad\\
+  University of St Andrews\\
+  Centre for Research into Ecological and Environmental Modelling,\\
+  The Observatory,\\
+  St Andrews, Fife KY16 9LZ, Scotland\\
+  E-mail: \email{Eric.Rexstad@st-andrews.ac.uk} \\ \\
+  Len Thomas\\
+  University of St Andrews\\
+  Centre for Research into Ecological and Environmental Modelling,\\
+  The Observatory,\\
+  St Andrews, Fife KY16 9LZ, Scotland\\
+  E-mail: \email{len.thomas@st-andrews.ac.uk}\\
+  URL: \url{http://lenthomas.org} \\ \\
+  Laura Marshall\\
+  University of St Andrews\\
+  Centre for Research into Ecological and Environmental Modelling,\\
+  The Observatory,\\
+  St Andrews, Fife KY16 9LZ, Scotland\\
+  E-mail: \email{lhm@st-andrews.ac.uk}\\ \\
+  Jeffrey L Laake\\
+  Marine Mammal Laboratory\\
+  Alaska Fisheries Science Center\\
+  National Oceanic and Atmospheric Administration\\
+  7600 Sand Point Way N.E., Seattle, WA 98115, USA\\
+  E-mail: \email{Jeff.Laake@noaa.gov}
 }
@@
 \begin{document}
-%\SweaveOpts{concordance=TRUE}
 
@@
 
-The Windows program Distance (or "DISTANCE"; for clarity henceforth "Distance for Windows") can be used to fit detection functions to distance sampling data.
-It was first released (versions 1.0 - 3.0; principally programmed by Jeff Laake while working at the National Marine Mammal Laboratory) as a console-based application (this in turn was based on earlier software TRANSECT, @Burnham:1980wz and algorithms developed in @Buckland:1992fa), before the first graphical interface (Distance for Windows 3.5) was released in November 1998.
-Since this time it has evolved to include various design and analysis features [@Thomas:2010cf].
-Distance for Windows versions 5 onwards have included \proglang{R} [@rcore]  packages as the analysis engines providing additional, more complex analysis options than those offered by the original (Fortran) code.
+The Windows program Distance (or ``DISTANCE''; for clarity henceforth ``Distance for Windows'') can be used to fit detection functions to distance sampling data.
+It was first released (versions 1.0 - 3.0; principally programmed by Jeff Laake while working at the National Marine Mammal Laboratory) as a console-based application (this in turn was based on earlier software TRANSECT, \citealp{Burnham:1980wz} and algorithms developed in \citealp{Buckland:1992fa}), before the first graphical interface (Distance for Windows 3.5) was released in November 1998.
+Since this time it has evolved to include various design and analysis features \citep{Thomas:2010cf}.
+Distance for Windows versions 5 onwards have included \proglang{R} \citep{rcore} packages as the analysis engines providing additional, more complex analysis options than those offered by the original (\proglang{Fortran}) code.
 
@@
 
-Until now those wishing to use our \proglang{R} packages for straight forward distance sampling analyses would have had to negotiate the package \pkg{mrds} [@mrds-pkg] designed for mark-recapture distance sampling [@Burt:2014gu], requiring a complex data structure to perform analyses.
+Until now those wishing to use our \proglang{R} packages for straight forward distance sampling analyses would have had to negotiate the package \pkg{mrds} \citep{mrds-pkg} designed for mark-recapture distance sampling \citep{Burt:2014gu}, requiring a complex data structure to perform analyses.
 \pkg{Distance} is a wrapper package around \pkg{mrds} making it easier to get started with basic distance sampling analyses in \proglang{R}.
@@
 
-The distribution of the observed distances is a product of detectability [sometimes referred to as "perception bias"; @Marsh:1989ho] and the distribution of the animals with respect to the line or point.
+The distribution of the observed distances is a product of detectability \citep[sometimes referred to as ``perception bias'';][]{Marsh:1989ho} and the distribution of the animals with respect to the line or point.
 Our survey design allows us to assume a distribution for the animals with respect to the sampler.
@@
 \begin{center}
-<<label=detectdists, fig=TRUE, echo=FALSE, message=FALSE>>=
+<<label=detectdists, fig=TRUE, echo=FALSE, message=FALSE, height=5.5, width=9>>=
 
@@
 Figure \ref{fig:pointslines} shows simulated sampling of a population of 500 objects using line and point transects and their corresponding histograms of observed detection distances.
-Note that for the purposes of distance sampling an "object" may either refer to an individual in a population or a cluster of individuals.
+Note that for the purposes of distance sampling an ``object'' may either refer to an individual in a population or a cluster of individuals.
 
@@
 \begin{center}
-<<label=points-and-lines, fig=TRUE, echo=FALSE, message=FALSE, warning=FALSE>>=
+<<label=points-and-lines, fig=TRUE, echo=FALSE, message=FALSE, warning=FALSE, width=7, height=6>>=
 opar <- par(mfrow=c(2,2))
@@
 @
-\caption{"Left side plots show an example of a survey of a (1x1) area containing a population of 500 objects; blue dashed lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals.
+\caption{Left side plots show an example of a survey of a (1x1) area containing a population of 500 objects; blue dashed lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals.
 The right side of the figure shows histograms of observed distances (again, lines top and points bottom).}
@@
 Good survey design is essential to produce reliable density and abundance estimates from distance sampling data.
-Survey design is beyond the scope of this article but we refer readers to @Buckland:2001vm [, Chapter 7] and @buckland2015distance [, Chapter 2] for introductory information; @Strindberg:2004vr contains information on automated survey design using geographical information systems (GIS); @Thomas:2007wz gives an example of designing a distance sampling survey in a complex region.
-We also note that Distance for Windows includes a GIS survey design tool [@Thomas:2010cf].
+Survey design is beyond the scope of this article but we refer readers to \cite[][Chapter 7]{Buckland:2001vm} and \cite[][Chapter 2]{buckland2015distance} for introductory information; \cite{Strindberg:2004vr} contains information on automated survey design using geographical information systems (GIS); \cite{Thomas:2007wz} gives an example of designing a distance sampling survey in a complex region.
+We also note that Distance for Windows includes a GIS survey design tool \citep{Thomas:2010cf}.
 
@@
 In addition to randomly placed samplers distance sampling relies on three other main assumptions.
-Firstly, all objects directly on the transect line or point (i.e., those at zero distance) are detected (see "Extensions" for methods to deal with the situation when this is not possible).
+Firstly, all objects directly on the transect line or point (i.e., those at zero distance) are detected (see ``Extensions'' for methods to deal with the situation when this is not possible).
 Secondly, objects are stationary or detected prior to any movement.
@@
 Depending on the survey species some of these assumptions may be more difficult to meet than others.
-Further information on field methods to help meet these assumptions can be found in @Buckland:2001vm and @buckland2015distance.
+Further information on field methods to help meet these assumptions can be found in \cite{Buckland:2001vm} and \cite{buckland2015distance}.
 
@@
 The below example analyses are not intended to serve as guidelines, but to demonstrate features of the software.
-Practical advice on approaches to analysis is given in @Thomas:2010cf.
+Practical advice on approaches to analysis is given in \cite{Thomas:2010cf}.
 
@@
 These data consist of 99 observations that are stratified based on location (near or distant from ice edge) and effort data (transect lengths).
-Further details on the survey are available in @Branch:2001ua (data is simulated based on the design used for "1992/93 Area III" therein).
+Further details on the survey are available in \cite{Branch:2001ua} (data is simulated based on the design used for ``1992/93 Area III'' therein).
 
@@
 The data include distances and two covariates collected during the survey: observer (a three level factor), time after sunrise (transformed to minutes (continuous) or hours (factor) covariates).
-Data are analysed comprehensively in @Marques:2007ey.
+Data are analysed comprehensively in \cite{Marques:2007ey}.
 
@@
 \item In the simplest case, where the objective is to estimate a detection function and exact distances are collected, all that is required is a numeric vector of distances.
-\item To include additional covariates into the detection function (see "Detection functions") a \code{data.frame} is required.
+\item To include additional covariates into the detection function (see ``Detection functions'') a \code{data.frame} is required.
 Each row of the \code{data.frame} contains the data on one observation.
 The \code{data.frame} must contain a column named \code{distance} (containing the observed distances) and additional named columns for any covariates that may affect detectability (for example \code{observer} or \code{seastate}).
-The column name \code{size} is reserved for the cluster sizes (sometimes referred to as group sizes) in this case each row represents an observation of a cluster rather than individual (see @Buckland:2001vm, Section 1.4.3 for more on defining clusters and "Extensions" below for one approach to dealing with uncertain cluster size).
-Additional reserved names include \code{object} and \code{detected}, these are not required for conventional distance sampling and should be avoided (see "Extensions" for an explanation of their use).
+The column name \code{size} is reserved for the cluster sizes (sometimes referred to as group sizes) in this case each row represents an observation of a cluster rather than individual \citep[see][Section 1.4.3 for more on defining clusters and ``Extensions'' below for one approach to dealing with uncertain cluster size]{Buckland:2001vm}.
+Additional reserved names include \code{object} and \code{detected}, these are not required for conventional distance sampling and should be avoided (see ``Extensions'' for an explanation of their use).
 \item To estimate density or to estimate abundance beyond the sampled area, additional information is required.
-Additional columns should be included in the \code{data.frame} specifying: \code{Sample.Label}, the ID of the transect; \code{Effort}, transect effort (for lines their length and for points the number of times that point was visited); \code{Region.Label}, the stratum containing the transect (which may be from pre- or post-survey stratification, see "Estimating abundance and variance"); \code{Area}, the area of the strata.
+Additional columns should be included in the \code{data.frame} specifying: \code{Sample.Label}, the ID of the transect; \code{Effort}, transect effort (for lines their length and for points the number of times that point was visited); \code{Region.Label}, the stratum containing the transect (which may be from pre- or post-survey stratification, see ``Estimating abundance and variance''); \code{Area}, the area of the strata.
 Transects which were surveyed but have no observations must be included in the data set with \code{NA} for \code{distance} and any other covariates.
-We refer to this data format (where all information is contained in one table) as "flatfile" as it can be easily created in a single spreadsheet.
+We refer to this data format (where all information is contained in one table) as ``flatfile'' as it can be easily created in a single spreadsheet.
 \end{itemize}
 
-As we will see in "Extensions", further information is also required for fitting more complex models.
+As we will see in ``Extensions'', further information is also required for fitting more complex models.
 
-It is also possible to use distances collected in intervals ("binned" or "grouped" data, as opposed to "exact" distances) to model the detection function.
+It is also possible to use distances collected in intervals (``binned'' or ``grouped'' data, as opposed to ``exact'' distances) to model the detection function.
 In this case the column \code{distance} is replaced by two columns \code{distbegin} and \code{distend} referring to the distance interval start and end cutpoints.
-More information on binned data is included in @Buckland:2001vm sections 4.5 and 7.4.1.2.
+More information on binned data is included in \cite{Buckland:2001vm} sections 4.5 and 7.4.1.2.
 
@@
 
-The minke whale data follows the "flatfile" format given in the last bullet point:
+The minke whale data follows the ``flatfile'' format given in the last bullet point:
 <<label=minke-data-head, message=FALSE>>=
@@
 $$
-where $\pi(y)$ describes the distribution of objects with respect to the sampler; $\pi(x)=1/w$ for line transects and $\pi(r)=\frac{2r}{w^2}$ for point transects, taking into account the geometry of the sampler [usually referred to as the \textit{probability density function of distances}; @Buckland:2001vm, Chapter 3].
+where $\pi(y)$ describes the distribution of objects with respect to the sampler; $\pi(x)=1/w$ for line transects and $\pi(r)=\frac{2r}{w^2}$ for point transects, taking into account the geometry of the sampler \citep[usually referred to as the \textit{probability density function of distances}][Chapter 3]{Buckland:2001vm}, .
 When considering a particular transect type we let $x$ denote a perpendicular distance from a line and $r$ denote radial distance from a point (rather than using $y$).
@@
 
-Models for the detection function are expected to have the following properties [@buckland2015distance, Chapter 5]:
+Models for the detection function are expected to have the following properties \citep[][Chapter 5]{buckland2015distance}:
 
@@
 \item \textit{Non-increasing}: we do not think that observers should be more likely to see distant objects than those nearer the transect.
-If this occurs, it usually indicates an issue with survey design or field procedures (for example that the distribution of objects with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this [@Marques:2010he; @Marques:2012fy; @Miller:2015hw].
+If this occurs, it usually indicates an issue with survey design or field procedures (for example that the distribution of objects with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this \citep{Marques:2010he, Marques:2012fy, Miller:2015hw}.
 \item \textit{Model robust}: models should be flexible enough to fit many different shapes.
@@
 
-There is a wide literature on possible formulations for the detection function [@Buckland:1992fa; @Eidous:2005bj; @Becker:2009cj; @Giammarino:2014eg; @Miller:2015hw; @Becker:2015fi].
+There is a wide literature on possible formulations for the detection function \citep{Buckland:1992fa, Eidous:2005bj, Becker:2009cj, Giammarino:2014eg, Miller:2015hw, Becker:2015fi}.
 \code{Distance} includes the most popular of these models.
-Here we detail the most popular detection function approach: "key function plus adjustments" (K+A).
+Here we detail the most popular detection function approach: ``key function plus adjustments'' (K+A).
 
@@
 
-Key function plus adjustment terms (or adjustment series) models are formulated by taking a "key" function and optionally adding "adjustments" to it to improve the fit [@Buckland:1992fa].
+Key function plus adjustment terms (or adjustment series) models are formulated by taking a ``key'' function and optionally adding ``adjustments'' to it to improve the fit \citep{Buckland:1992fa}.
 Mathematically we formulate this as:
@@
 Possible modelling options for key and adjustments are given in Table \ref{tab:keyadj} and illustrated in Figures \ref{fig:hnhr} and \ref{fig:keyadj}.
-We select the number of adjustment terms ($K$) by AIC (further details in "Model checking and model selection").
+We select the number of adjustment terms ($K$) by AIC (further details in ``Model checking and model selection'').
 
@@
 \begin{center}
-<<label=hn-hr-par-comp, fig=TRUE, echo=FALSE>>=
+<<label=hn-hr-par-comp, fig=TRUE, echo=FALSE, height=5, width=10>>=
 par(mfrow=c(2,4), mar=c(3.5, 3, 2, 1) + 0.1)
@@
 For each key function the default adjustments are cosine in \pkg{Distance}.
-Note that in the adjustments functions distance is divided by the width or the scale parameter to ensure the shape of adjustment independent of the units of $y$ (Marques et al., 2007); defaults are shown here, though either can be selected to rescale the distances.}
+Note that in the adjustments functions distance is divided by the width or the scale parameter to ensure the shape of adjustment independent of the units of $y$ \citep{Marques:2007ey}; defaults are shown here, though either can be selected to rescale the distances.}
 \begin{tabular}{llll}
@@
 
-A disadvantage of K+A models is that we must resort to constrained optimisation [via the \pkg{Rsolnp} package; @rsolnp-pkg] to ensure that the resulting detection function is monotonic non-increasing over its range.
+A disadvantage of K+A models is that we must resort to constrained optimisation \citep[via the \pkg{Rsolnp} package;][]{rsolnp-pkg} to ensure that the resulting detection function is monotonic non-increasing over its range.
 
-It is not always necessary to include adjustments (except in the case of the uniform key) and in such cases we refer to these as ``key only'' models (see the next section and "Model checking and model selection").
+It is not always necessary to include adjustments (except in the case of the uniform key) and in such cases we refer to these as ``key only'' models (see the next section and ``Model checking and model selection'').
 
@@
 \begin{center}
-<<label=adjust-mix-comp, fig=TRUE, echo=FALSE>>=
+<<label=adjust-mix-comp, fig=TRUE, echo=FALSE, height=3, width=10>>=
 par(mfrow=c(1,4), mar=c(3.5, 3, 3, 1) + 0.1)
@@
 Including covariates has an important implication for our calculation of detectability.
-We do not know the true distribution of the covariates, we therefore must either: $(i)$ model the distribution of the covariates and integrate the covariates out of the joint density (thus making strong assumptions about their distribution), or $(ii)$ calculate the probability of detection conditional on the observed values of the covariates [@Marques:2003vb].
+We do not know the true distribution of the covariates, we therefore must either: $(i)$ model the distribution of the covariates and integrate the covariates out of the joint density (thus making strong assumptions about their distribution), or $(ii)$ calculate the probability of detection conditional on the observed values of the covariates \citep{Marques:2003vb}.
 We opt for the latter:
@@
 As such we cannot set meaningful constraints on monotonicity.
-For this reason, we advise caution when using both adjustments and covariates in a detection function [see @Miller:2015hw for an example of when this can be problematic].
+For this reason, we advise caution when using both adjustments and covariates in a detection function \citep[see][for an example of when this can be problematic]{Miller:2015hw}.
 
@@
 
-Other adjustment series can be selected using the `adjustment=` argument and specific orders of adjustments can be set using `order=`.
+Other adjustment series can be selected using the \code{adjustment=} argument and specific orders of adjustments can be set using \code{order=}.
 For example, to specify a uniform model with cosine adjustments of order 1 and 2 we can write:
@@
 @
-Hermite polynomial adjustments use the code `"herm"` and simple polynomials `"poly"`, adjustment order should be in line with Table \ref{tab:keyadj}.
+Hermite polynomial adjustments use the code \code{"herm"} and simple polynomials \code{"poly"}, adjustment order should be in line with Table \ref{tab:keyadj}.
 
@@
 We can include covariates in the scale parameter via the \code{formula=~...} argument to \code{ds}.
-A hazard-rate model for the amakihi that includes observer as a covariate and a truncation distance of 82.5m [@Marques:2007ey] can be specified using :
+A hazard-rate model for the amakihi that includes observer as a covariate and a truncation distance of 82.5m \citep{Marques:2007ey} can be specified using :
 <<label=amakihi-hr-obs>>=
@@
 \begin{center}
-<<label=minke-amakihi-hn-plot, echo=FALSE, warning=FALSE, fig=TRUE>>=
+<<label=minke-amakihi-hn-plot, echo=FALSE, warning=FALSE, fig=TRUE, height=4, width=9>>=
 par(mfrow=c(1, 3))
@@
 @
-This summary information includes details of the data and model specification, as well as the values of the coefficients ($\beta_j$) and their uncertainties, an "average" value for the detectability (see "Estimating abundance and variance" for details on how this is calculated) and its uncertainty.
+This summary information includes details of the data and model specification, as well as the values of the coefficients ($\beta_j$) and their uncertainties, an ``average'' value for the detectability (see ``Estimating abundance and variance'' for details on how this is calculated) and its uncertainty.
 The final line gives an estimate of abundance in the area covered by the survey (see the next section; though note this estimate does not take into account cluster size).
@@
 This can be interpreted as assessing whether the number of observations up to a given distance is in line with what the model says they should be.
-As usual for Q-Q plots, "good" models will have values close to the line $y=x$, poor models will show greater deviations from that line.
+As usual for Q-Q plots, ``good'' models will have values close to the line $y=x$, poor models will show greater deviations from that line.
 
 Q-Q plots can be inspected visually, though this is prone to subjective judgments.
-Therefore, we also quantify the Q-Q plot's information using a Kolmogorov-Smirnov or Cramér-von Mises test [@Burnham:2004vd].
+Therefore, we also quantify the Q-Q plot's information using a Kolmogorov-Smirnov or Cram\'{e}r-von Mises test \citep{Burnham:2004vd}.
 Both test whether points from the EDF and CDF are from the same distribution.
-The Kolmogorov-Smirnov uses the test statistic of the largest difference between a point on the Q-Q plot and the line $y=x$, whereas the Cramér-von Mises test uses the sum of all the differences.
-As it takes into account more information and is therefore more powerful, the Cramér-von Mises is generally preferred.
+The Kolmogorov-Smirnov uses the test statistic of the largest difference between a point on the Q-Q plot and the line $y=x$, whereas the Cram\'{e}r-von Mises test uses the sum of all the differences.
+As it takes into account more information and is therefore more powerful, the Cram\'{e}r-von Mises is generally preferred.
 A significant result from either test gives evidence against the null hypothesis (that the data arose from the fitted model), suggesting that the model does not fit the data well.
@@
 \begin{center}
-<<label=amakihiqq, fig=TRUE, echo=FALSE, message=FALSE>>=
+<<label=amakihiqq, fig=TRUE, echo=FALSE, message=FALSE, height=3, width=6>>=
 par(mfrow=c(1,2))
@@
 
-For non-exact distance data, a $\chi^2$-test can be used to assess goodness of fit [see @Buckland:2001vm, Section 3.4.4].
+For non-exact distance data, a $\chi^2$-test can be used to assess goodness of fit \citep[see][Section 3.4.4]{Buckland:2001vm}.
 $\chi^2$-test results are produced by \code{gof_ds} when binned/grouped distances are provided.
@@
 
-Once we have a set of plausible models, we can use Akaike's Information Criterion (AIC) to select between models [see e.g., @burnham2003model].
+Once we have a set of plausible models, we can use Akaike's Information Criterion (AIC) to select between models \citep[see e.g.][]{burnham2003model}.
 \pkg{Distance} includes a function to create table of summary information for fitted models, making it easy to get an overview of a large number of models.
-The \code{summarize_ds_models} function takes models as input and can be especially useful when paired with \pkg{knitr}'s \code{kable} function to create summary tables for publication [@knitr-pkg].
+The \code{summarize_ds_models} function takes models as input and can be especially useful when paired with \pkg{knitr}'s \code{kable} function to create summary tables for publication \citep{knitr-pkg}.
 An example of this output (with all models included) is shown in Table \ref{tab:amakihi} and was generated by the following call to \code{summarize_ds_models}:
@@
 
-<<label=summary-table, results="tex", echo=FALSE>>=
+<<label=summary-table, results=tex, echo=FALSE>>=
+library("knitr")
 kable(summarize_ds_models(amakihi_hn, amakihi_hr_obs, amakihi_hr_obs_mas),
@@
       caption="Summary for the detection function models fitted to the amakihi data.
-``C-vM'' stands for Cramér-von Mises, $P_a$ is average detectability (see ``Estimating abundance and variance''), se is standard error.
+``C-vM'' stands for Cram\\'{e}r-von Mises, $P_a$ is average detectability (see ``Estimating abundance and variance''), se is standard error.
 Models are sorted according to AIC.\\label{tab:amakihi}")
@@
 To do this we first calculate the abundance in the area we have surveyed (the \textit{covered area}) to obtain $\hat{N}_\text{C}$, we can then scale this up to the full study area by multiplying it by the ratio of covered area to study area.
-We discuss other methods for spatially explicit abundance estimation in "Extensions".
+We discuss other methods for spatially explicit abundance estimation in ``Extensions''.
 
@@
 \end{equation}
-where $s_i$ are the sizes of the observed clusters of objects, which are all equal to 1 if objects only occur singly [@Borchers:2004wr].
-@thompson2012sampling is the canonical reference to this type of estimator.
-Intuitively, we can think of the estimates of detectability ($\hat{p}(\mathbf{z}_i)$) as "inflating" the cluster sizes ($s_i$), we then sum over the detections ($i$) to obtain the abundance estimate.
+where $s_i$ are the sizes of the observed clusters of objects, which are all equal to 1 if objects only occur singly \citep{Borchers:2004wr}.
+\cite{thompson2012sampling} is the canonical reference to this type of estimator.
+Intuitively, we can think of the estimates of detectability ($\hat{p}(\mathbf{z}_i)$) as ``inflating'' the cluster sizes ($s_i$), we then sum over the detections ($i$) to obtain the abundance estimate.
 For models that do not include covariates, $\hat{p}(\mathbf{z}_i)$ is equal for all $i$, so this is equivalent to summing the clusters and inflating that sum by dividing through by the corresponding $\hat{p} (=\hat{p}(\mathbf{z}_i) \forall i)$.
@@
 
-We can use the Horvitz-Thompson-like estimator to calculate the "average" detectability for models which include covariates.
+We can use the Horvitz-Thompson-like estimator to calculate the ``average'' detectability for models which include covariates.
 We can consider what single detectability value would give the estimated $\hat{N}_\text{C}$ and therefore calculate:
@@
 
-We take an intuitive approach to uncertainty estimation, for a full derivation consult @Marques:2003vb. Uncertainty in $\hat{N}$ comes from two sources:
+We take an intuitive approach to uncertainty estimation, for a full derivation consult \cite{Marques:2003vb}. Uncertainty in $\hat{N}$ comes from two sources:
 
@@
 $$
-where the partial derivatives of $\hat{N}_\text{C}$ are evaluated at the MLE ($\hat{\theta}$) and $\mathbf{H}$ is the first partial Hessian (outer product of first derivatives of the log likelihood) for numerical stability [@Buckland:2001vm, p 62].
+where the partial derivatives of $\hat{N}_\text{C}$ are evaluated at the MLE ($\hat{\theta}$) and $\mathbf{H}$ is the first partial Hessian (outer product of first derivatives of the log likelihood) for numerical stability \citep[][p 62]{Buckland:2001vm}.
 Note that although we calculate uncertainty in $\hat{N}_\text{C}$, we can trivially scale-up to variance of $\hat{N}$ (by noting that $\hat{N} = \frac{A}{a} \hat{N}_\text{C}$ and therefore $\widehat{\text{Var}}_\text{model}\left(\hat{N}\right) = \left( \frac{A}{a} \right)^2 \widehat{\text{Var}}_\text{model} \left(\hat{N}_\text{C} \right)$).
@@
 When covariates are not included in the detection function we can define the encounter rate as $n/L$ for line transects (where $L$ is the total line length) or $n/T$ for point transects (where $T$ is the total number of visits summed over all points).
-When covariates are included in the detection function, it is recommended that we substitute the $n$ in the encounter rate with the estimated abundance $\hat{N}_\text{C}$ as this will take into account the effects of the covariates [@Innes:2002ka].
+When covariates are included in the detection function, it is recommended that we substitute the $n$ in the encounter rate with the estimated abundance $\hat{N}_\text{C}$ as this will take into account the effects of the covariates \citep{Innes:2002ka}.
 
-For line transects, by default, \pkg{Distance} uses a variation of the estimator ``R2'' from @Fewster:2009ku replacing number of observations per sample with the estimated abundance per sample, thus taking into account cluster size if this is recorded [@Innes:2002ka; @Marques:2003vb]:
+For line transects, by default, \pkg{Distance} uses a variation of the estimator ``R2'' from \cite{Fewster:2009ku} replacing number of observations per sample with the estimated abundance per sample, thus taking into account cluster size if this is recorded \citep{Innes:2002ka, Marques:2003vb}:
 $$
@@
 where $l_k$ are the lengths of the $K$ transects (such that $L = \sum_{k=1}^K l_k$) and $\hat{N_{C,k}}$ is the abundance in the covered area for transect $k$.
-For point transects we use the estimator ``P3'' from @Fewster:2009ku but again replace $n$ by $\hat{N}_\text{C}$ in the encounter rate definition, to obtain the following estimator:
+For point transects we use the estimator ``P3'' from \cite{Fewster:2009ku} but again replace $n$ by $\hat{N}_\text{C}$ in the encounter rate definition, to obtain the following estimator:
 $$
@@
 
-Other formulations for the encounter rate variance are discussed in detail in @Fewster:2009ku.
+Other formulations for the encounter rate variance are discussed in detail in \cite{Fewster:2009ku}.
 \pkg{Distance} implements all of the estimators of encounter rate variance given in that article.
@@
 
-We combine these two variances by noting that squared coefficients of variation (approximately) add [often referred to as "the delta method"; @seber2002estimation].
+We combine these two variances by noting that squared coefficients of variation (approximately) add \citep[often referred to as ``the delta method'';][]{seber2002estimation}.
 
@@
 Returning to the minke whale data, we have the necessary information to calculate $A$ and $a$ above, so we can estimate abundance and its variance.
-When we supply data to \code{ds} in the "flatfile" format given above, \code{ds} will automatically calculate abundance estimates based on the survey information in the data.
+When we supply data to \code{ds} in the ``flatfile'' format given above, \code{ds} will automatically calculate abundance estimates based on the survey information in the data.
 
@@
 
-<<label=minke-abundance-table, echo=FALSE>>=
+<<label=minke-abundance-table, results=tex, echo=FALSE>>=
 minke_table <- summary(minke_hn)$dht$individuals$N
@@
 colnames(minke_table) <- c("Stratum", "$\\hat{N}$", "$\\text{se}(\\hat{N}$)", "$\\text{CV}(\\hat{N}$)")
-kable(minke_table, digits=3, format="latex", booktabs=TRUE, row.names=FALSE, escape=FALSE,caption="Summary of abundance estimation for the half-normal model for the minke whale data.\\label{minke-abund}")
+kable(minke_table, digits=3, format="latex", booktabs=TRUE, row.names=FALSE, escape=FALSE, caption="Summary of abundance estimation for the half-normal model for the minke whale data.\\label{minke-abund}")
 @
@@
 Distance sampling has been applied in a wide variety of situations.
-Objects being detected need not be animals; plants [@Buckland:2007kl], beer cans [@Otto:1990gs], elephant dung [@NchPlu01] and bricks on a lake bottom [@Ber90] have been subjects of distance sampling investigations.
+Objects being detected need not be animals; plants \citep{Buckland:2007kl}, beer cans \citep{Otto:1990gs}, elephant dung \citep{NchPlu01} and bricks on a lake bottom \citep{Ber90} have been subjects of distance sampling investigations.
 The method of detection need not be visual sightings.
-Detections of objects can be through auditory as well as visual means [@Marques2013b].
-Songs of birds [@Buckland2006] and whale vocalisations [@BorSub1] are just two examples.
+Detections of objects can be through auditory as well as visual means \citep{Marques2013b}.
+Songs of birds \citep{Buckland2006} and whale vocalisations \citep{BorSub1} are just two examples.
 Blows made by whales are examples of processes that can be modelled using distance sampling.
@@
 Additional features available in \pkg{mrds} include  models that relax the assumption that detection is certain at zero distance from the transect (by including data from additional observers).
-This is done using mark-recapture type methods which require additional survey methodology, known as double observer surveys [see @Burt:2014gu for an introduction].
+This is done using mark-recapture type methods which require additional survey methodology, known as double observer surveys \citep[see][for an introduction]{Burt:2014gu}.
 
@@
 Such models incorporate spatially-referenced environmental data (for example derived from GIS products).
-\pkg{Distance} interfaces with one such package used to perform this type of analysis: \pkg{dsm} [@dsm-pkg].
-So-called "density surface modelling" uses the generalized additive model framework [e.g., @Wood:2006vg] to build models of abundance (adjusting counts for imperfect detectability) as a function of environmental covariates, as part of a two stage model [@Hedley:2004et; @Miller:2013fq].
+\pkg{Distance} interfaces with one such package used to perform this type of analysis: \pkg{dsm} \citep{dsm-pkg}.
+So-called ``density surface modelling'' uses the generalized additive model framework \citep[e.g.][]{Wood:2006vg} to build models of abundance (adjusting counts for imperfect detectability) as a function of environmental covariates, as part of a two stage model \citep{Hedley:2004et, Miller:2013fq}.
 
 Uncertainty in measured covariates (e.g.
-cluster size) and model uncertainty (when two models have similar fit but substantially different estimates) can be incorporated using the multi-analysis distance sampling package \pkg{mads} [@mads-pkg].
+cluster size) and model uncertainty (when two models have similar fit but substantially different estimates) can be incorporated using the multi-analysis distance sampling package \pkg{mads} \citep{mads-pkg}.
 In addition, \pkg{mads} can also incorporate sightings with unknown species identification.
-This is done by estimating the abundance of these 'unidentified' sightings and pro-rating them to the known species [@Gerrodette:2005].
+This is done by estimating the abundance of these 'unidentified' sightings and pro-rating them to the known species \citep{Gerrodette:2005}.
 
@@
 \pkg{DSsim} allows users to test out different designs in their study region and tailor population attributes to reflect the species they are working with.
-\pkg{DSsim} [@DSsim-pkg] allows users to more easily identify challenges unique to their study and select a survey design which is more likely to yield accurate and precise estimates.
+\pkg{DSsim} \citep{DSsim-pkg} allows users to more easily identify challenges unique to their study and select a survey design which is more likely to yield accurate and precise estimates.
 
 Distance for Windows has many users (over 45,000 downloads since 2002) and they may be overwhelmed by the prospect of switching existing analyses to \proglang{R}.
-For that reason we have created the \pkg{readdst} [@readdst-pkg] package to interface with projects created by Distance for Windows.
+For that reason we have created the \pkg{readdst} \citep{readdst-pkg} package to interface with projects created by Distance for Windows.
 The package can take analyses created using the CDS, MCDS and MRDS engines in Distance for Windows, extract the data and create equivalent models in \proglang{R}.
@@
 
-In combination with tools such as \pkg{knitr} and \pkg{rmarkdown} [@rmarkdown-pkg], the helper functions in \pkg{Distance} provide a useful set of tools to perform reproducible analyses of wildlife abundance for both managers and ecologists.
-We also direct readers' attention to \pkg{DsShiny} [@dsshiny-pkg], a package that builds on \pkg{Shiny} [@shiny-pkg] and \pkg{mrds} to allow users to build and fit models in a graphical interface, with output to RMarkdown.
+In combination with tools such as \pkg{knitr} and \pkg{rmarkdown} \citep{rmarkdown-pkg}, the helper functions in \pkg{Distance} provide a useful set of tools to perform reproducible analyses of wildlife abundance for both managers and ecologists.
+We also direct readers' attention to \pkg{DsShiny} \citep{dsshiny-pkg}, a package that builds on \pkg{Shiny} \citep{shiny-pkg} and \pkg{mrds} to allow users to build and fit models in a graphical interface, with output to RMarkdown.
 
@@
 We would particularly like to thank Steve Buckland, David Borchers, Tiago Marques, Jon Bishop and Lorenzo Milazzo for their contributions.
-We thank Tiago Marques a second time for helpful comments on an early version of the paper and Colin Beale who suggested the "flatfile" data format.
+We thank Tiago Marques a second time for helpful comments on an early version of the paper and Colin Beale who suggested the ``flatfile'' data format.
 We also thank Ken Burnham and David Anderson for fundamental contributions to the early development of these methods.

### 2016-06-08 paper.tex
fix-up
got through some of the requirements

--- 2016-03-31 paper.tex
@@
 \documentclass[article]{jss}
-\usepackage[utf8]{inputenc}
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-
-\author{
-David L Miller\\Integrated Statistics and University of St Andrews \AND Eric Rexstad\\University of St Andrews \And Len Thomas\\University of St Andrews \AND Laura Marshall\\University of St Andrews \And Jeffrey L Laake\\Marine Mammal Laboratory
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\usepackage{amsmath}
+\usepackage{amssymb}
+%% need no \usepackage{Sweave.sty}
+
+%\usepackage{longtable}
+\usepackage{booktabs}
+
+%% almost as usual
+\author{David L Miller\\Integrated Statistics and University of St Andrews \AND 
+        Eric Rexstad\\University of St Andrews \And 
+        Len Thomas\\University of St Andrews \AND 
+        Laura Marshall\\University of St Andrews \And 
+        Jeffrey L Laake\\Marine Mammal Laboratory}
+
+\title{Distance Sampling in \proglang{R}}
+
+%% for pretty printing and a nice hypersummary also set:
+\Plainauthor{David L Miller, Eric Rexstad, Len Thomas, Laura Marshall, Jeffrey L Laake}
+\Plaintitle{Distance Sampling in R} %% without formatting
+%\Shorttitle{\pkg{foo}: A Capitalized Title} %% a short title (if necessary)
+
+%% an abstract and keywords
+\Abstract{
+Estimating the abundance and spatial distribution of animal and plant populations is essential for conservation and management.
+We introduce the \proglang{R} package \pkg{Distance} that implements distance sampling methods to estimate abundance.
+We describe how users can obtain estimates of abundance (and density) using the package as well documenting the links it provides with other more specialized \proglang{R} packages.
+We also demonstrate how \pkg{Distance} provides a migration pathway from previous software, thereby allowing us to deliver cutting-edge methods to the users more quickly.
 }
-\title{Distance Sampling in \proglang{R}}
 \Keywords{distance sampling, abundance estimation, line transect, point transect, detection function, Horvitz-Thompson, \proglang{R}, Distance}
-
-\Abstract{
-Estimating the abundance and spatial distribution of animal and plant
-populations is essential for conservation and management. We introduce
-the \proglang{R} package \pkg{Distance} that implements distance
-sampling methods to estimate abundance. We describe how users can obtain
-estimates of abundance (and density) using the package as well
-documenting the links it provides with other more specialized
-\proglang{R} packages. We also demonstrate how \pkg{Distance} provides a
-migration pathway from previous software, thereby allowing us to deliver
-cutting-edge methods to the users more quickly.
-}
-
-\Plainauthor{David L Miller, Eric Rexstad, Len Thomas, Laura Marshall, Jeffrey L Laake}
-\Plaintitle{Distance Sampling in R}
-\Plainkeywords{distance sampling, abundance estimation, line transect, point transecs, detection function, Horvitz-Thompson, R, Distance}
+\Plainkeywords{distance sampling, abundance estimation, line transect, point transecs, detection function, Horvitz-Thompson, R, Distance} %% without formatting
+%% at least one keyword must be supplied
 
 %% publication information
+%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
 %% \Volume{50}
@@
 %% \Year{2012}
-\Submitdate{}
+%% \Submitdate{2012-06-04}
 %% \Acceptdate{2012-06-04}
 
+%% The address of (at least) one author should be given
+%% in the following format:
 \Address{
@@
   Integrated Statistics and University of St Andrews\\
-  16 Sumner Street, Woods Hole, MA 02543, USA and Centre for Research into
-  Ecological and Environmental Modelling, The Observatory, St Andrews,
-  Fife KY16 9LZ, Scotland\\
-  E-mail: \href{mailto:dave@ninepointeightone.net}{\nolinkurl{dave@ninepointeightone.net}}\\
-  URL: \url{http://converged.yt}\\~\\
+  16 Sumner Street, Woods Hole, MA 02543, USA and\\
+  Centre for Research into Ecological and Environmental Modelling,\\
+  The Observatory,\\
+  St Andrews, Fife KY16 9LZ, Scotland\\
+  E-mail: \email{dave@ninepointeightone.net}\\
+  URL: \url{http://converged.yt} \\ \\
   Eric Rexstad\\
   University of St Andrews\\
-  Centre for Research into Ecological and Environmental Modelling, The
-  Observatory, St Andrews, Fife KY16 9LZ, Scotland\\
-  E-mail: \href{mailto:Eric.Rexstad@st-andrews.ac.uk}{\nolinkurl{Eric.Rexstad@st-andrews.ac.uk}}\\
-
+  Centre for Research into Ecological and Environmental Modelling,\\
+  The Observatory,\\
+  St Andrews, Fife KY16 9LZ, Scotland\\
+  E-mail: \email{Eric.Rexstad@st-andrews.ac.uk} \\ \\
   Len Thomas\\
   University of St Andrews\\
-  Centre for Research into Ecological and Environmental Modelling, The
-  Observatory, St Andrews, Fife KY16 9LZ, Scotland\\
-  E-mail: \href{mailto:len.thomas@st-andrews.ac.uk}{\nolinkurl{len.thomas@st-andrews.ac.uk}}\\
-  URL: \url{http://lenthomas.org}\\~\\
+  Centre for Research into Ecological and Environmental Modelling,\\
+  The Observatory,\\
+  St Andrews, Fife KY16 9LZ, Scotland\\
+  E-mail: \email{len.thomas@st-andrews.ac.uk}\\
+  URL: \url{http://lenthomas.org} \\ \\
   Laura Marshall\\
   University of St Andrews\\
-  Centre for Research into Ecological and Environmental Modelling, The
-  Observatory, St Andrews, Fife KY16 9LZ, Scotland\\
-  E-mail: \href{mailto:lhm@st-andrews.ac.uk}{\nolinkurl{lhm@st-andrews.ac.uk}}\\
-
+  Centre for Research into Ecological and Environmental Modelling,\\
+  The Observatory,\\
+  St Andrews, Fife KY16 9LZ, Scotland\\
+  E-mail: \email{lhm@st-andrews.ac.uk}\\ \\
   Jeffrey L Laake\\
   Marine Mammal Laboratory\\
-  Alaska Fisheries Science Center National Oceanic and Atmospheric
-  Administration 7600 Sand Point Way N.E., Seattle, WA 98115, USA\\
-  E-mail: \href{mailto:Jeff.Laake@noaa.gov}{\nolinkurl{Jeff.Laake@noaa.gov}}\\
-
+  Alaska Fisheries Science Center\\
+  National Oceanic and Atmospheric Administration\\
+  7600 Sand Point Way N.E., Seattle, WA 98115, USA\\
+  E-mail: \email{Jeff.Laake@noaa.gov}
 }
-
-\usepackage{amsmath} \usepackage{amssymb} \usepackage{longtable}
-\usepackage{booktabs}
-\usepackage[colorinlistoftodos, color=green!40, textwidth=60, textsize=small]{todonotes}
+%% It is also possible to add a telephone and fax number
+%% before the e-mail in the following format:
+%% Telephone: +43/512/507-7103
+%% Fax: +43/512/507-2851
+
+%% for those who use Sweave please include the following line (with % symbols):
+%% need no \usepackage{Sweave.sty}
+
+%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\section{Introduction}\label{introduction}
-
-Distance sampling
-\citep{Buckland:2001vm, Buckland:2004ts, buckland2015distance}
-encompasses a suite of methods used to estimate the density and/or
-abundance of biological populations. Distance sampling can be thought of
-as an extension of plot sampling. Plot sampling involves selecting a
-number of plots (small areas) at random within the study area and
-counting the objects of interest that are contained within each plot. By
-selecting the plots at random we can assume that the density of objects
-in the plots is representative of the study area as a whole. One of the
-key assumptions of plot sampling is that all objects within each of the
-plots are counted. Distance sampling relaxes this assumption in that
-observers are no longer required to detect (either by eye, video/audio
-recording etc) and count everything within selected plots. While plot
-sampling techniques are adequate for static populations occurring at
-high density they are inefficient for more sparsely distributed
-populations. Distance sampling provides a more efficient solution in
-such circumstances.
-
-Conventional distance sampling assumes the observer is located either at
-a point or moving along a line and will observe all objects that occur
-at the point or on the line. The further away an object is from the
-point or line (also known as the sampler or transect) the less likely it
-is that the observer will see it. We can use the distances to each of
-the detected objects from the line or point to build a model of the
-probability of detection given distance from the sampler -- the
-\emph{detection function}. The detection function can be used to infer
-how many objects were missed and thereby produce estimates of density
-and/or abundance.
-
-The Windows program Distance (or ``DISTANCE''; for clarity henceforth
-``Distance for Windows'') can be used to fit detection functions to
-distance sampling data. It was first released (versions 1.0 - 3.0;
-principally programmed by Jeff Laake while working at the National
-Marine Mammal Laboratory) as a console-based application (this in turn
-was based on earlier software TRANSECT, \citet{Burnham:1980wz} and
-algorithms developed in \citet{Buckland:1992fa}), before the first
-graphical interface (Distance for Windows 3.5) was released in November
-1998. Since this time it has evolved to include various design and
-analysis features \citep{Thomas:2010cf}. Distance for Windows versions 5
-onwards have included \proglang{R} \citep{rcore} packages as the
-analysis engines providing additional, more complex analysis options
-than those offered by the original (FORTRAN) code.
-
-As Distance for Windows becomes increasingly reliant on analyses
-performed in \proglang{R} and many new methods are being developed, we
-are encouraging the use of our \proglang{R} packages directly.
-\proglang{R} provides a huge variety of functionality for data
-exploration and reproducible research, much more than is possible in
-Distance for Windows.
-
-Until now those wishing to use our \proglang{R} packages for straight
-forward distance sampling analyses would have had to negotiate the
-package \pkg{mrds} \citep{mrds-pkg} designed for mark-recapture distance
-sampling \citep{Burt:2014gu}, requiring a complex data structure to
-perform analyses. \pkg{Distance} is a wrapper package around \pkg{mrds}
-making it easier to get started with basic distance sampling analyses in
-\proglang{R}. The most basic detection function estimation only requires
-a numeric vector of distances. Here we demonstrate how to use
-\pkg{Distance} to fit detection functions, perform model checking and
-selection, and estimate abundance.
-
-\subsection{Distance sampling}\label{distance-sampling}
-
-The distribution of the observed distances is a product of detectability
-\citep[sometimes referred to as ``perception bias'';][]{Marsh:1989ho}
-and the distribution of the animals with respect to the line or point.
-Our survey design allows us to assume a distribution for the animals
-with respect to the sampler.
-
-For line transect studies we assume that objects are uniformly
-distributed with respect to the line (i.e., the number of animals
-available for detection is the same at all distances). For point
-transect surveys, area increases linearly with radial distance, implying
-a triangular distribution with respect to the point. Figure
-\ref{fig:detectiondist} shows how these distributions, when combined
-with a detection function, give rise to the observed distribution of
-recorded distances.
-
-\begin{CodeChunk}
+\section{Introduction}
+
+Distance sampling \citep{Buckland:2001vm, Buckland:2004ts, buckland2015distance} encompasses a suite of methods used to estimate the density and/or abundance of biological populations.
+Distance sampling can be thought of as an extension of plot sampling.
+Plot sampling involves selecting a number of plots (small areas) at random within the study area and counting the objects of interest that are contained within each plot.
+By selecting the plots at random we can assume that the density of objects in the plots is representative of the study area as a whole.
+One of the key assumptions of plot sampling is that all objects within each of the plots are counted.
+Distance sampling relaxes this assumption in that observers are no longer required to detect (either by eye, video/audio recording etc) and count everything within selected plots.
+While plot sampling techniques are adequate for static populations occurring at high density they are inefficient for more sparsely distributed populations.
+Distance sampling provides a more efficient solution in such circumstances.
+
+Conventional distance sampling assumes the observer is located either at a point or moving along a line and will observe all objects that occur at the point or on the line.
+The further away an object is from the point or line (also known as the sampler or transect) the less likely it is that the observer will see it.
+We can use the distances to each of the detected objects from the line or point to build a model of the probability of detection given distance from the sampler --- the \textit{detection function}.
+The detection function can be used to infer how many objects were missed and thereby produce estimates of density and/or abundance.
+
+The Windows program Distance (or ``DISTANCE''; for clarity henceforth ``Distance for Windows'') can be used to fit detection functions to distance sampling data.
+It was first released (versions 1.0 - 3.0; principally programmed by Jeff Laake while working at the National Marine Mammal Laboratory) as a console-based application (this in turn was based on earlier software TRANSECT, \citealp{Burnham:1980wz} and algorithms developed in \citealp{Buckland:1992fa}), before the first graphical interface (Distance for Windows 3.5) was released in November 1998.
+Since this time it has evolved to include various design and analysis features \citep{Thomas:2010cf}.
+Distance for Windows versions 5 onwards have included \proglang{R} \citep{rcore} packages as the analysis engines providing additional, more complex analysis options than those offered by the original (\proglang{Fortran}) code.
+
+As Distance for Windows becomes increasingly reliant on analyses performed in \proglang{R} and many new methods are being developed, we are encouraging the use of our \proglang{R} packages directly.
+\proglang{R} provides a huge variety of functionality for data exploration and reproducible research, much more than is possible in Distance for Windows.
+
+Until now those wishing to use our \proglang{R} packages for straight forward distance sampling analyses would have had to negotiate the package \pkg{mrds} \citep{mrds-pkg} designed for mark-recapture distance sampling \citep{Burt:2014gu}, requiring a complex data structure to perform analyses.
+\pkg{Distance} is a wrapper package around \pkg{mrds} making it easier to get started with basic distance sampling analyses in \proglang{R}.
+The most basic detection function estimation only requires a numeric vector of distances.
+Here we demonstrate how to use \pkg{Distance} to fit detection functions, perform model checking and selection, and estimate abundance.
+
+\subsection{Distance sampling}
+
+The distribution of the observed distances is a product of detectability \citep[sometimes referred to as ``perception bias'';][]{Marsh:1989ho} and the distribution of the animals with respect to the line or point.
+Our survey design allows us to assume a distribution for the animals with respect to the sampler.
+
+For line transect studies we assume that objects are uniformly distributed with respect to the line (i.e., the number of animals available for detection is the same at all distances).
+For point transect surveys, area increases linearly with radial distance, implying a triangular distribution with respect to the point.
+Figure \ref{fig:detectiondist} shows how these distributions, when combined with a detection function, give rise to the observed distribution of recorded distances.
+
 \begin{figure}
-
-{\centering \includegraphics{paper_files/figure-latex/detectdists-1}
-
-}
-
+\begin{center}
+\includegraphics{paper-detectdists}
 \caption{Panels show an example detection function (left), the pdf of object distances (middle) and the resulting pdf of detection distances (right) for line transects (top row) and point transects (bottom row).
 The pdf of observed detection distances in the right hand plots are obtained by multiplying the detection function by the pdf of object distances and rescaling.
-In this example, detection probability becomes effectively zero at 500 (distances shown on $x$-axis are arbitrary).
-\label{fig:detectiondist}}\label{fig:detectdists}
+In this example, detection probability becomes effectively zero at 500 (distances shown on $x$-axis are arbitrary).}
+\label{fig:detectiondist}
+\end{center}
 \end{figure}
-\end{CodeChunk}
-
-Figure \ref{fig:pointslines} shows simulated sampling of a population of
-500 objects using line and point transects and their corresponding
-histograms of observed detection distances. Note that for the purposes
-of distance sampling an ``object'' may either refer to an individual in
-a population or a cluster of individuals.
-
-\begin{CodeChunk}
-\begin{CodeOutput}
-Warning: package 'nlme' was built under R version 3.2.4
-\end{CodeOutput}
+
+Figure \ref{fig:pointslines} shows simulated sampling of a population of 500 objects using line and point transects and their corresponding histograms of observed detection distances.
+Note that for the purposes of distance sampling an ``object'' may either refer to an individual in a population or a cluster of individuals.
+
 \begin{figure}
-
-{\centering \includegraphics{paper_files/figure-latex/points-and-lines-1}
-
-}
-
+\begin{center}
+\includegraphics{paper-points-and-lines}
 \caption{Left side plots show an example of a survey of a (1x1) area containing a population of 500 objects; blue dashed lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals.
-The right side of the figure shows histograms of observed distances (again, lines top and points bottom).\label{fig:pointslines}}\label{fig:points-and-lines}
+The right side of the figure shows histograms of observed distances (again, lines top and points bottom).}
+\label{fig:pointslines}
+\end{center}
 \end{figure}
-\end{CodeChunk}
-
-Good survey design is essential to produce reliable density and
-abundance estimates from distance sampling data. Survey design is beyond
-the scope of this article but we refer readers to \citet[Chapter
-7]{Buckland:2001vm} and \citet[Chapter 2]{buckland2015distance} for
-introductory information; \citet{Strindberg:2004vr} contains information
-on automated survey design using geographical information systems (GIS);
-\citet{Thomas:2007wz} gives an example of designing a distance sampling
-survey in a complex region. We also note that Distance for Windows
-includes a GIS survey design tool \citep{Thomas:2010cf}.
-
-\pkg{Distance} provides a selection of candidate functions to describe
-the probability of detection and estimates the associated parameters
-using maximum likelihood estimation. The probability of detecting an
-object may not only depend on how far it is from the observer but also
-on other factors such as weather conditions, ground cover, cluster size
-etc. The \pkg{Distance} package also allows the incorporation of such
-covariates into the detection function allowing the detection function
-scale parameter to vary based on these covariates.
-
-Having estimated the detection function's parameters, one can then
-integrate out distance from the function (as the detection function
-describes the probability of detection \emph{given} distance) to get an
-``average'' probability of detection, which can be used to correct the
-observed counts. Summing the corrected counts gives an estimate of
-abundance in the area covered by surveys, which can be multiplied up to
-the total study area.
-
-In addition to randomly placed samplers distance sampling relies on
-three other main assumptions. Firstly, all objects directly on the
-transect line or point (i.e.~those at zero distance) are detected (see
-``Extensions'' for methods to deal with the situation when this is not
-possible). Secondly, objects are stationary or detected prior to any
-movement. Thirdly, distance to the object must be measured accurately,
-or the observation allocated to the correct distance bin for grouped
-data. Depending on the survey species some of these assumptions may be
-more difficult to meet than others. Further information on field methods
-to help meet these assumptions can be found in \citet{Buckland:2001vm}
-and \citet{buckland2015distance}.
-
-The rest of the paper has the following structure: we describe data
-formatting for \pkg{Distance}; candidate detection function models are
-described in terms of formulation and fitted examples in \proglang{R}.
-We then show how to perform model checking, goodness of fit and model
-selection. We go on to show how to estimate abundance, including
-stratified estimates of abundance. The final two sections of the article
-look at extensions (both in terms of methodology and software) and put
-the package in a broader context amongst other \proglang{R} packages
-used for estimating the abundance of biological populations from
-distance sampling data.
-
-\section{Data}\label{data}
-
-We introduce two example analyses performed in \pkg{Distance}: one line
-transect and one point transect. These data sets have been chosen as
-they represent typical data seen in practice. The below example analyses
-are not intended to serve as guidelines, but to demonstrate features of
-the software. Practical advice on approaches to analysis is given in
-\citet{Thomas:2010cf}.
-
-\subsubsection{Minke whales}\label{minke-whales}
-
-The line transect data have been simulated from models fitted to
-Antarctic minke whale (\emph{Balaenoptera bonaerensis}) data. These data
-were collected as part of the International Whaling Commission's
-International Decade of Cetacean Research Southern Ocean Whale and
-Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer
-surveys. These data consist of 99 observations that are stratified based
-on location (near or distant from ice edge) and effort data (transect
-lengths). Further details on the survey are available in
-\citet{Branch:2001ua} (data is simulated based on the design used for
-``1992/93 Area III'' therein).
-
-\subsubsection{Amakihi}\label{amakihi}
-
-The point transect data set consists of 1485 observations of Amakihi
-(\emph{Hemignathus virens}; a Hawaiian songbird), collected at 41 points
-between 1992 and 1995. The data include distances and two covariates
-collected during the survey: observer (a three level factor), time after
-sunrise (transformed to minutes (continuous) or hours (factor)
-covariates). Data are analysed comprehensively in
-\citet{Marques:2007ey}.
-
-\subsection{Data setup}\label{data-setup}
-
-Generally, data collected in the field will require some formatting
-before use with \pkg{Distance}, though there are a range of possible
-formats, dependent on the model specification and the output required:
+
+Good survey design is essential to produce reliable density and abundance estimates from distance sampling data.
+Survey design is beyond the scope of this article but we refer readers to \cite[][Chapter 7]{Buckland:2001vm} and \cite[][Chapter 2]{buckland2015distance} for introductory information; \cite{Strindberg:2004vr} contains information on automated survey design using geographical information systems (GIS); \cite{Thomas:2007wz} gives an example of designing a distance sampling survey in a complex region.
+We also note that Distance for Windows includes a GIS survey design tool \citep{Thomas:2010cf}.
+
+\pkg{Distance} provides a selection of candidate functions to describe the probability of detection and estimates the associated parameters using maximum likelihood estimation.
+The probability of detecting an object may not only depend on how far it is from the observer but also on other factors such as weather conditions, ground cover, cluster size etc.
+The \pkg{Distance} package also allows the incorporation of such covariates into the detection function allowing the detection function scale parameter to vary based on these covariates.
+
+Having estimated the detection function's parameters, one can then integrate out distance from the function (as the detection function describes the probability of detection \textit{given} distance) to get an ``average'' probability of detection, which can be used to correct the observed counts.
+Summing the corrected counts gives an estimate of abundance in the area covered by surveys, which can be multiplied up to the total study area.
+
+In addition to randomly placed samplers distance sampling relies on three other main assumptions.
+Firstly, all objects directly on the transect line or point (i.e., those at zero distance) are detected (see ``Extensions'' for methods to deal with the situation when this is not possible).
+Secondly, objects are stationary or detected prior to any movement.
+Thirdly, distance to the object must be measured accurately, or the observation allocated to the correct distance bin for grouped data.
+Depending on the survey species some of these assumptions may be more difficult to meet than others.
+Further information on field methods to help meet these assumptions can be found in \cite{Buckland:2001vm} and \cite{buckland2015distance}.
+
+The rest of the paper has the following structure: we describe data formatting for \pkg{Distance}; candidate detection function models are described in terms of formulation and fitted examples in \proglang{R}.
+We then show how to perform model checking, goodness of fit and model selection.
+We go on to show how to estimate abundance, including stratified estimates of abundance.
+The final two sections of the article look at extensions (both in terms of methodology and software) and put the package in a broader context amongst other \proglang{R} packages used for estimating the abundance of biological populations from distance sampling data.
+
+\section{Data}
+
+We introduce two example analyses performed in \pkg{Distance}: one line transect and one point transect.
+These data sets have been chosen as they represent typical data seen in practice.
+The below example analyses are not intended to serve as guidelines, but to demonstrate features of the software.
+Practical advice on approaches to analysis is given in \cite{Thomas:2010cf}.
+
+\subsection{Minke whales}
+
+The line transect data have been simulated from models fitted to Antarctic minke whale (\textit{Balaenoptera bonaerensis}) data.
+These data were collected as part of the International Whaling Commission's International Decade of Cetacean Research Southern Ocean Whale and Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer surveys.
+These data consist of 99 observations that are stratified based on location (near or distant from ice edge) and effort data (transect lengths).
+Further details on the survey are available in \cite{Branch:2001ua} (data is simulated based on the design used for ``1992/93 Area III'' therein).
+
+\subsection{Amakihi}
+
+The point transect data set consists of 1485 observations of Amakihi (\textit{Hemignathus virens}; a Hawaiian songbird), collected at 41 points between 1992 and 1995.
+The data include distances and two covariates collected during the survey: observer (a three level factor), time after sunrise (transformed to minutes (continuous) or hours (factor) covariates).
+Data are analysed comprehensively in \cite{Marques:2007ey}.
+
+\subsection{Data setup}
+
+Generally, data collected in the field will require some formatting before use with \pkg{Distance}, though there are a range of possible formats, dependent on the model specification and the output required:
 
 \begin{itemize}
-\tightlist
-\item
-  In the simplest case, where the objective is to estimate a detection
-  function and exact distances are collected, all that is required is a
-  numeric vector of distances.
-\item
-  To include additional covariates into the detection function (see
-  ``Detection functions'') a \code{data.frame} is required. Each row of
-  the \code{data.frame} contains the data on one observation. The
-  \code{data.frame} must contain a column named \code{distance}
-  (containing the observed distances) and additional named columns for
-  any covariates that may affect detectability (for example
-  \code{observer} or \code{seastate}). The column name \code{size} is
-  reserved for the cluster sizes (sometimes referred to as group sizes)
-  in this case each row represents an observation of a cluster rather
-  than individual (see \citet{Buckland:2001vm}, Section 1.4.3 for more
-  on defining clusters and ``Extensions'' below for one approach to
-  dealing with uncertain cluster size). Additional reserved names
-  include \code{object} and \code{detected}, these are not required for
-  conventional distance sampling and should be avoided (see
-  ``Extensions'' for an explanation of their use).
-\item
-  To estimate density or to estimate abundance beyond the sampled area,
-  additional information is required. Additional columns should be
-  included in the \code{data.frame} specifying: \code{Sample.Label}, the
-  ID of the transect; \code{Effort}, transect effort (for lines their
-  length and for points the number of times that point was visited);
-  \code{Region.Label}, the stratum containing the transect (which may be
-  from pre- or post-survey stratification, see ``Estimating abundance
-  and variance''); \code{Area}, the area of the strata. Transects which
-  were surveyed but have no observations must be included in the data
-  set with \code{NA} for \code{distance} and any other covariates. We
-  refer to this data format (where all information is contained in one
-  table) as ``flatfile'' as it can be easily created in a single
-  spreadsheet.
+\item In the simplest case, where the objective is to estimate a detection function and exact distances are collected, all that is required is a numeric vector of distances.
+\item To include additional covariates into the detection function (see ``Detection functions'') a \code{data.frame} is required.
+Each row of the \code{data.frame} contains the data on one observation.
+The \code{data.frame} must contain a column named \code{distance} (containing the observed distances) and additional named columns for any covariates that may affect detectability (for example \code{observer} or \code{seastate}).
+The column name \code{size} is reserved for the cluster sizes (sometimes referred to as group sizes) in this case each row represents an observation of a cluster rather than individual \citep[see][Section 1.4.3 for more on defining clusters and ``Extensions'' below for one approach to dealing with uncertain cluster size]{Buckland:2001vm}.
+Additional reserved names include \code{object} and \code{detected}, these are not required for conventional distance sampling and should be avoided (see ``Extensions'' for an explanation of their use).
+\item To estimate density or to estimate abundance beyond the sampled area, additional information is required.
+Additional columns should be included in the \code{data.frame} specifying: \code{Sample.Label}, the ID of the transect; \code{Effort}, transect effort (for lines their length and for points the number of times that point was visited); \code{Region.Label}, the stratum containing the transect (which may be from pre- or post-survey stratification, see ``Estimating abundance and variance''); \code{Area}, the area of the strata.
+Transects which were surveyed but have no observations must be included in the data set with \code{NA} for \code{distance} and any other covariates.
+We refer to this data format (where all information is contained in one table) as ``flatfile'' as it can be easily created in a single spreadsheet.
 \end{itemize}
 
-As we will see in ``Extensions'', further information is also required
-for fitting more complex models.
-
-It is also possible to use distances collected in intervals (``binned''
-or ``grouped'' data, as opposed to ``exact'' distances) to model the
-detection function. In this case the column \code{distance} is replaced
-by two columns \code{distbegin} and \code{distend} referring to the
-distance interval start and end cutpoints. More information on binned
-data is included in \citet{Buckland:2001vm} sections 4.5 and 7.4.1.2.
-
-The columns \code{distance}, \code{Area} and (in the case of line
-transects) \code{Effort} have associated units (though these are not
-explicitly included in a \pkg{Distance} analysis). We recommend that
-data in these columns are converted to SI units before starting any
-analysis to ensure that resulting abundance and density estimates have
-sensible units. For example, if distances from a line transect survey
-are recorded in metres, the \code{Effort} columns should contain line
-lengths also in metres and the \code{Area} column gives the stratum
-areas in square metres. This would lead to density estimates of animals
-per square metre.
-
-The minke whale data follows the ``flatfile'' format given in the last
-bullet point:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-library("Distance")
-head(minke)
-\end{CodeInput}
-\begin{CodeOutput}
+As we will see in ``Extensions'', further information is also required for fitting more complex models.
+
+It is also possible to use distances collected in intervals (``binned'' or ``grouped'' data, as opposed to ``exact'' distances) to model the detection function.
+In this case the column \code{distance} is replaced by two columns \code{distbegin} and \code{distend} referring to the distance interval start and end cutpoints.
+More information on binned data is included in \cite{Buckland:2001vm} sections 4.5 and 7.4.1.2.
+
+The columns \code{distance}, \code{Area} and (in the case of line transects) \code{Effort} have associated units (though these are not explicitly included in a \pkg{Distance} analysis).
+We recommend that data in these columns are converted to SI units before starting any analysis to ensure that resulting abundance and density estimates have sensible units.
+For example, if distances from a line transect survey are recorded in metres, the \code{Effort} columns should contain line lengths also in metres and the \code{Area} column gives the stratum areas in square metres.
+This would lead to density estimates of animals per square metre.
+
+The minke whale data follows the ``flatfile'' format given in the last bullet point:
+\begin{Schunk}
+\begin{Sinput}
+R> library("Distance")
+R> data("minke")
+R> head(minke)
+\end{Sinput}
+\begin{Soutput}
   Region.Label  Area Sample.Label Effort distance
@@
 6        South 84734            1  86.75     0.95
-\end{CodeOutput}
-\end{CodeChunk}
-
+\end{Soutput}
+\end{Schunk}
 Whereas the amakihi data lacks effort and stratum data:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-head(amakihi)
-\end{CodeInput}
-\begin{CodeOutput}
+\begin{Schunk}
+\begin{Sinput}
+R> data("amakihi")
+R> head(amakihi)
+\end{Sinput}
+\begin{Soutput}
    survey object distance obs mas has detected
@@
 6 July 92      6      120 TJS  50   1        1
-\end{CodeOutput}
-\end{CodeChunk}
-
-We will explore the consequences of including effort and stratum data in
-the analysis below.
-
-\section{Detection functions}\label{detection-functions}
-
-The detection function models the probability
-\(\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)\)
-and is usually denoted \(g(y; \boldsymbol{\theta})\) where \(y\) is
-distance (from a line or point) and \(\boldsymbol{\theta}\) is a vector
-of parameters to be estimated. Our goal is to estimate an \emph{average
-probability of detection} (\(p\), average in the sense of an average
-over distance from \(0\) to truncation distance \(w\)), so we must
-integrate out distance (\(y\)) from the detection function: \[
+\end{Soutput}
+\end{Schunk}
+
+We will explore the consequences of including effort and stratum data in the analysis below.
+
+\section{Detection functions}
+
+The detection function models the probability $\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)$ and is usually denoted $g(y; \boldsymbol{\theta})$ where $y$ is distance (from a line or point) and $\boldsymbol{\theta}$ is a vector of parameters to be estimated.
+Our goal is to estimate an \textit{average probability of detection} ($p$, average in the sense of an average over distance from $0$ to truncation distance $w$), so we must integrate out distance ($y$) from the detection function:
+$$
 p = \int_0^w \pi(y) g(y; \boldsymbol{\theta}) dy
-\] where \(\pi(y)\) describes the distribution of objects with respect
-to the sampler; \(\pi(x)=1/w\) for line transects and
-\(\pi(r)=\frac{2r}{w^2}\) for point transects, taking into account the
-geometry of the sampler \citep[usually referred to as the
-\emph{probability density function of distances};][Chapter
-3]{Buckland:2001vm}. When considering a particular transect type we let
-\(x\) denote a perpendicular distance from a line and \(r\) denote
-radial distance from a point (rather than using \(y\)).
-
-It is crucial that the detection function accurately models
-detectability at small distances; we are less worried by its behaviour
-further away from 0. To ensure that the model is not overly influenced
-by distances far from zero, we truncate the distances beyond a given
-distance \(w\) (this can be done during analysis or while collecting
-data in the field).
-
-Models for the detection function are expected to have the following
-properties \citep[Chapter 5]{buckland2015distance}:
+$$
+where $\pi(y)$ describes the distribution of objects with respect to the sampler; $\pi(x)=1/w$ for line transects and $\pi(r)=\frac{2r}{w^2}$ for point transects, taking into account the geometry of the sampler \citep[usually referred to as the \textit{probability density function of distances}][Chapter 3]{Buckland:2001vm}, .
+When considering a particular transect type we let $x$ denote a perpendicular distance from a line and $r$ denote radial distance from a point (rather than using $y$).
+
+It is crucial that the detection function accurately models detectability at small distances; we are less worried by its behaviour further away from 0.
+To ensure that the model is not overly influenced by distances far from zero, we truncate the distances beyond a given distance $w$ (this can be done during analysis or while collecting data in the field).
+
+Models for the detection function are expected to have the following properties \citep[][Chapter 5]{buckland2015distance}:
 
 \begin{itemize}
-\tightlist
-\item
-  \emph{Shoulder}: we expect observers to be able to see objects near
-  them, not just those directly in front of them. For this reason, we
-  expect the detection function to be flat near the line or point.
-\item
-  \emph{Non-increasing}: we do not think that observers should be more
-  likely to see distant objects than those nearer the transect. If this
-  occurs, it usually indicates an issue with survey design or field
-  procedures (for example that the distribution of objects with respect
-  to the line, \(\pi(y)\) is not what we expect), so we do not want the
-  detection function to model this
-  \citep{Marques:2010he, Marques:2012fy, Miller:2015hw}.
-\item
-  \emph{Model robust}: models should be flexible enough to fit many
-  different shapes.
-\item
-  \emph{Pooling robust}: many factors can affect the probability of
-  detection and it is not possible to measure all of these. We would
-  like models to produce unbiased results without inclusion of these
-  factors.
-\item
-  \emph{Estimator efficiency}: we would like models to have low
-  variances, given they satisfy the other properties above (which, if
-  satisfied, would give low bias).
+\item \textit{Shoulder}: we expect observers to be able to see objects near them, not just those directly in front of them.
+For this reason, we expect the detection function to be flat near the line or point.
+\item \textit{Non-increasing}: we do not think that observers should be more likely to see distant objects than those nearer the transect.
+If this occurs, it usually indicates an issue with survey design or field procedures (for example that the distribution of objects with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this \citep{Marques:2010he, Marques:2012fy, Miller:2015hw}.
+\item \textit{Model robust}: models should be flexible enough to fit many different shapes.
+\item \textit{Pooling robust}: many factors can affect the probability of detection and it is not possible to measure all of these.
+We would like models to produce unbiased results without inclusion of these factors.
+\item \textit{Estimator efficiency}: we would like models to have low variances, given they satisfy the other properties above (which, if satisfied, would give low bias).
 \end{itemize}
 
-Given these criteria, we can formulate models for \(g\).
-
-\subsection{Formulations}\label{formulations}
-
-There is a wide literature on possible formulations for the detection
-function
-\citep{Buckland:1992fa, Eidous:2005bj, Becker:2009cj, Giammarino:2014eg, Miller:2015hw, Becker:2015fi}.
-\code{Distance} includes the most popular of these models. Here we
-detail the most popular detection function approach: ``key function plus
-adjustments'' (K+A).
-
-\subsubsection{Key function plus adjustments
-(K+A)}\label{key-function-plus-adjustments-ka}
-
-Key function plus adjustment terms (or adjustment series) models are
-formulated by taking a ``key'' function and optionally adding
-``adjustments'' to it to improve the fit \citep{Buckland:1992fa}.
-Mathematically we formulate this as: \[
+Given these criteria, we can formulate models for $g$.
+
+\subsection{Formulations}
+
+There is a wide literature on possible formulations for the detection function \citep{Buckland:1992fa, Eidous:2005bj, Becker:2009cj, Giammarino:2014eg, Miller:2015hw, Becker:2015fi}.
+\code{Distance} includes the most popular of these models.
+Here we detail the most popular detection function approach: ``key function plus adjustments'' (K+A).
+
+\subsubsection{Key function plus adjustments (K+A)}
+
+Key function plus adjustment terms (or adjustment series) models are formulated by taking a ``key'' function and optionally adding ``adjustments'' to it to improve the fit \citep{Buckland:1992fa}.
+Mathematically we formulate this as:
+$$
 g(y; \boldsymbol{\theta}) = k(y; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(y; \boldsymbol{\theta}_\text{adjust})\right),
-\] where \(k\) is the key function and \(\alpha_O\) is sum series of
-functions (given in Table \ref{tab:keyadj}), described as an
-\emph{adjustment of order \(O\)}. Subscripts on the parameter vector
-indicate those parameters belonging to each part of the model (i.e.
-\(\boldsymbol{\theta} = (\boldsymbol{\theta}_\text{key}, \boldsymbol{\theta}_\text{adjust})\)).
-
-Available models for the key are as follows: \[
+$$
+where $k$ is the key function and $\alpha_O$ is sum series of functions (given in Table \ref{tab:keyadj}), described as an \textit{adjustment of order $O$}.
+Subscripts on the parameter vector indicate those parameters belonging to each part of the model (i.e., $\boldsymbol{\theta} = (\boldsymbol{\theta}_\text{key}, \boldsymbol{\theta}_\text{adjust})$).
+
+Available models for the key are as follows:
+$$
 k(y) = \left\{
@@
 \end{array} \right.
-\] Possible modelling options for key and adjustments are given in Table
-\ref{tab:keyadj} and illustrated in Figures \ref{fig:hnhr} and
-\ref{fig:keyadj}. We select the number of adjustment terms (\(K\)) by
-AIC (further details in ``Model checking and model selection'').
-
-\begin{CodeChunk}
+$$
+Possible modelling options for key and adjustments are given in Table \ref{tab:keyadj} and illustrated in Figures \ref{fig:hnhr} and \ref{fig:keyadj}.
+We select the number of adjustment terms ($K$) by AIC (further details in ``Model checking and model selection'').
+
 \begin{figure}
-
-{\centering \includegraphics{paper_files/figure-latex/hn-hr-par-comp-1}
-
-}
-
+\begin{center}
+\includegraphics{paper-hn-hr-par-comp}
 \caption{Half-normal (top row) and hazard-rate (bottom row) detection functions without adjustments, varying scale ($\sigma$) and (for hazard-rate) shape ($b$) parameters (values are given above the plots).
 On the top row from left to right, the study species becomes more detectable (higher probability of detection at larger distances).
-The bottom row shows the hazard-rate model's more pronounced shoulder.\label{fig:hnhr}}\label{fig:hn-hr-par-comp}
+The bottom row shows the hazard-rate model's more pronounced shoulder.}
+\label{fig:hnhr}
+\end{center}
 \end{figure}
-\end{CodeChunk}
 
 \begin{table}
+\caption{Modelling options for key plus adjustment series models for the detection function.
+For each key function the default adjustments are cosine in \pkg{Distance}.
+Note that in the adjustments functions distance is divided by the width or the scale parameter to ensure the shape of adjustment independent of the units of $y$ \citep{Marques:2007ey}; defaults are shown here, though either can be selected to rescale the distances.}
 \begin{tabular}{llll}
@@
 \label{tab:keyadj}
-\caption{Modelling options for key plus adjustment series models for the detection function.
-For each key function the default adjustments are cosine in \pkg{Distance}.
-Note that in the adjustments functions distance is divided by the width or the scale parameter to ensure the shape of adjustment independent of the units of $y$ (Marques et al., 2007); defaults are shown here, though either can be selected to rescale the distances.}
 \end{table}
 
-When adjustment terms are used it is necessary to standardise the
-results to ensure that \(g(0)=1\), so we can redefine the detection
-function as: \[
+When adjustment terms are used it is necessary to standardise the results to ensure that $g(0)=1$, so we can redefine the detection function as:
+$$
 g(y; \boldsymbol{\theta}) = \frac{k(y; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(y; \boldsymbol{\theta}_\text{adjust})\right)}{k(0; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(0; \boldsymbol{\theta}_\text{adjust})\right)}.
-\]
-
-A disadvantage of K+A models is that we must resort to constrained
-optimisation \citep[via the \pkg{Rsolnp} package;][]{rsolnp-pkg} to
-ensure that the resulting detection function is monotonic non-increasing
-over its range.
-
-It is not always necessary to include adjustments (except in the case of
-the uniform key) and in such cases we refer to these as ``key only''
-models (see the next section and ``Model checking and model
-selection'').
-
-\begin{CodeChunk}
+$$
+
+A disadvantage of K+A models is that we must resort to constrained optimisation \citep[via the \pkg{Rsolnp} package;][]{rsolnp-pkg} to ensure that the resulting detection function is monotonic non-increasing over its range.
+
+It is not always necessary to include adjustments (except in the case of the uniform key) and in such cases we refer to these as ``key only'' models (see the next section and ``Model checking and model selection'').
+
 \begin{figure}
-
-{\centering \includegraphics{paper_files/figure-latex/adjust-mix-comp-1}
-
-}
-
-\caption{Possible shapes for the detection function when cosine adjustments are included for half-normal and hazard-rate models.\label{fig:keyadj}}\label{fig:adjust-mix-comp}
+\begin{center}
+\includegraphics{paper-adjust-mix-comp}
+\caption{Possible shapes for the detection function when cosine adjustments are included for half-normal and hazard-rate models.}
+\label{fig:keyadj}
+\end{center}
 \end{figure}
-\end{CodeChunk}
-
-\subsubsection{Covariates}\label{covariates}
-
-There are many factors that can affect the probability of detecting an
-object: observer skill, cluster size (if objects occur in clusters), the
-vessel or platform used, sea state, other weather conditions, time of
-day and more. In \pkg{Distance} we assume that these variables affect
-detection only via the scale of the detection function (and do not
-affect the shape).
-
-Covariates can be included in this formulation by considering the scale
-parameter from the half-normal or hazard-rate detection functions as a(n
-exponentiated) linear model of the (\(J\)) covariates (\(\mathbf{z}\); a
-vector of length \(J\) for each observation): \[
+
+\subsubsection{Covariates}
+
+There are many factors that can affect the probability of detecting an object: observer skill, cluster size (if objects occur in clusters), the vessel or platform used, sea state, other weather conditions, time of day and more.
+In \pkg{Distance} we assume that these variables affect detection only via the scale of the detection function (and do not affect the shape).
+
+Covariates can be included in this formulation by considering the scale parameter from the half-normal or hazard-rate detection functions as a(n exponentiated) linear model of the ($J$) covariates ($\mathbf{z}$; a vector of length $J$ for each observation):
+$$
 \sigma(\mathbf{z}) = \exp(\beta_0 + \sum_{j=1}^J \beta_j z_j).
-\]
-
-Including covariates has an important implication for our calculation of
-detectability. We do not know the true distribution of the covariates,
-we therefore must either: \((i)\) model the distribution of the
-covariates and integrate the covariates out of the joint density (thus
-making strong assumptions about their distribution), or \((ii)\)
-calculate the probability of detection conditional on the observed
-values of the covariates \citep{Marques:2003vb}. We opt for the latter:
-\[
+$$
+
+Including covariates has an important implication for our calculation of detectability.
+We do not know the true distribution of the covariates, we therefore must either: $(i)$ model the distribution of the covariates and integrate the covariates out of the joint density (thus making strong assumptions about their distribution), or $(ii)$ calculate the probability of detection conditional on the observed values of the covariates \citep{Marques:2003vb}.
+We opt for the latter:
+$$
 p(\mathbf{z_i}) = \int_0^w \pi(y) g(y, \mathbf{z_i}; \boldsymbol{\theta}) dy,
-\] where \(\mathbf{z_i}\) is the vector of \(J\) covariates associated
-with observation \(i\). For covariate models, we calculate a value of
-``average'' probability of detection (average in the sense of distance
-being integrated out) per observation. There are as many unique values
-of \(p(\mathbf{z_i})\) as there are unique covariate combinations in our
-data.
-
-K+A models that include covariates and one or more adjustments cannot be
-guaranteed to be monotonic non-increasing for all covariate
-combinations. Without a model for the distribution of the covariates, it
-is not possible to know what the behaviour of the detection function
-will be across the ranges of the covariates. As such we cannot set
-meaningful constraints on monotonicity. For this reason, we advise
-caution when using both adjustments and covariates in a detection
-function \citep[see][ for an example of when this can be
-problematic]{Miller:2015hw}.
-
-\subsection{Fitting detection functions in
-R}\label{fitting-detection-functions-in-r}
-
-A detection function can be fitted in \pkg{Distance} using the \code{ds}
-function. Here we apply some of the possible formulations for the
-detection function we have seen above to the minke whale and amakihi
-data.
-
-\subsubsection{Minke whale}\label{minke-whale}
-
-First we fit a model to the minke whale data, setting the truncation at
-1.5km and using the default options in \code{ds} very simply:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-minke_hn <- ds(minke, truncation = 1.5)
-\end{CodeInput}
-\begin{CodeOutput}
-Starting AIC adjustment term selection.
-\end{CodeOutput}
-\begin{CodeOutput}
-Fitting half-normal key function
-\end{CodeOutput}
-\begin{CodeOutput}
-Key only model: not constraining for monotonicity.
-\end{CodeOutput}
-\begin{CodeOutput}
-AIC= 46.872
-\end{CodeOutput}
-\begin{CodeOutput}
-Fitting half-normal key function with cosine(2) adjustments
-\end{CodeOutput}
-\begin{CodeOutput}
-AIC= 48.872
-\end{CodeOutput}
-\begin{CodeOutput}
-
-Half-normal key function selected.
-\end{CodeOutput}
-\end{CodeChunk}
-
-Note that when there are no covariates in the model, \code{ds} will add
-adjustment terms to the model until there is no improvement in AIC.
-
-Figure \ref{fig:minkeamakihi} (left panel) shows the result of calling
-\code{plot} on the resulting model object. We can also call
-\code{summary} on the model object to get summary information about the
-fitted model (we postpone this until the next section).
-
-A different form for the detection function can be specified via the
-\code{key=} argument to \code{ds}. For example, a hazard rate model can
-be fitted as:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-minke_hrcos <- ds(minke, truncation = 1.5, key = "hr")
-\end{CodeInput}
-\begin{CodeOutput}
-Starting AIC adjustment term selection.
-\end{CodeOutput}
-\begin{CodeOutput}
-Fitting hazard-rate key function
-\end{CodeOutput}
-\begin{CodeOutput}
-Key only model: not constraining for monotonicity.
-\end{CodeOutput}
-\begin{CodeOutput}
-AIC= 48.637
-\end{CodeOutput}
-\begin{CodeOutput}
-Fitting hazard-rate key function with cosine(2) adjustments
-\end{CodeOutput}
-\begin{CodeOutput}
-AIC= 50.386
-\end{CodeOutput}
-\begin{CodeOutput}
-
-Hazard-rate key function selected.
-\end{CodeOutput}
-\end{CodeChunk}
-
-Here \code{ds} also fits the hazard-rate model then hazard-rate with a
-cosine adjustment but the AIC improvement is insufficient to select the
-adjustment, so the hazard-rate key-only model is returned.
-
-Other adjustment series can be selected using the \texttt{adjustment=}
-argument and specific orders of adjustments can be set using
-\texttt{order=}. For example, to specify a uniform model with cosine
-adjustments of order 1 and 2 we can write:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-minke_unifcos <- ds(minke, truncation = 1.5, key = "unif",
-                    adjustment = "cos", order = c(1, 2))
-\end{CodeInput}
-\begin{CodeOutput}
-Fitting uniform key function with cosine(1,2) adjustments
-\end{CodeOutput}
-\begin{CodeOutput}
-AIC= 48.268
-\end{CodeOutput}
-\end{CodeChunk}
-
-Hermite polynomial adjustments use the code \texttt{"herm"} and simple
-polynomials \texttt{"poly"}, adjustment order should be in line with
-Table \ref{tab:keyadj}.
-
-\subsubsection{Amakihi}\label{amakihi-1}
-
-\code{ds} assumes the data given to it has been collected as line
-transects, but we can switch to point transects using the argument
-\code{transect="point"}. We can include covariates in the scale
-parameter via the \code{formula=~...} argument to \code{ds}. A
-hazard-rate model for the amakihi that includes observer as a covariate
-and a truncation distance of 82.5m \citep{Marques:2007ey} can be
-specified using :
-
-\begin{CodeChunk}
-\begin{CodeInput}
-amakihi_hr_obs <- ds(amakihi, truncation = 82.5, transect = "point",
-                     key = "hr", formula = ~obs)
-\end{CodeInput}
-\begin{CodeOutput}
-Model contains covariate term(s): no adjustment terms will be included.
-\end{CodeOutput}
-\begin{CodeOutput}
-Fitting hazard-rate key function
-\end{CodeOutput}
-\begin{CodeOutput}
-AIC= 10778.448
-\end{CodeOutput}
-\begin{CodeOutput}
-No survey area information supplied, only estimating detection function.
-\end{CodeOutput}
-\end{CodeChunk}
-
-Note that here, unlike with the minke whale data, \code{ds} warns us
-that we have only supplied enough information to estimate the detection
-function (not density or abundance).
-
-While automatic AIC selection is performed on adjustment terms, model
-selection for covariates must be performed manually. Here we add a
-second covariate: minutes after sunrise. We will compare these two
-models further in the following section.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-amakihi_hr_obs_mas <- ds(amakihi, truncation = 82.5, transect = "point",
-                         key = "hr", formula = ~obs + mas)
-\end{CodeInput}
-\begin{CodeOutput}
-Model contains covariate term(s): no adjustment terms will be included.
-\end{CodeOutput}
-\begin{CodeOutput}
-Fitting hazard-rate key function
-\end{CodeOutput}
-\begin{CodeOutput}
-AIC= 10777.376
-\end{CodeOutput}
-\begin{CodeOutput}
-No survey area information supplied, only estimating detection function.
-\end{CodeOutput}
-\end{CodeChunk}
-
-As with the minke whale model, we can plot the resulting models (Figure
-\ref{fig:minkeamakihi}, middle and right panels). However, for point
-transect studies, probability density function plots give a better sense
-of model fit than the detection function plots. This is because when
-plotting the detection function for point transect data, the histogram
-must be rescaled to account for the geometry of the point sampler. The
-amakihi models included covariates, so the plots show the detection
-function averaged over levels/values of the covariate. Points on the
-plot indicate probability of detection for each observation. For the
-\texttt{amakihi\_hr\_obs} model we see fairly clear levels of the
-observer covariate in the points. Looking at the right panel of Figure
-\ref{fig:minkeamakihi}, this is less clear when adding minutes after
-sunrise (a continuous covariates) to the model.
-
-\begin{CodeChunk}
+$$
+where $\mathbf{z_i}$ is the vector of $J$ covariates associated with observation $i$.
+For covariate models, we calculate a value of ``average'' probability of detection (average in the sense of distance being integrated out) per observation.
+There are as many unique values of $p(\mathbf{z_i})$ as there are unique covariate combinations in our data.
+
+K+A models that include covariates and one or more adjustments cannot be guaranteed to be monotonic non-increasing for all covariate combinations.
+Without a model for the distribution of the covariates, it is not possible to know what the behaviour of the detection function will be across the ranges of the covariates.
+As such we cannot set meaningful constraints on monotonicity.
+For this reason, we advise caution when using both adjustments and covariates in a detection function \citep[see][for an example of when this can be problematic]{Miller:2015hw}.
+
+\subsection{Fitting detection functions in R}
+
+A detection function can be fitted in \pkg{Distance} using the \code{ds} function.
+Here we apply some of the possible formulations for the detection function we have seen above to the minke whale and amakihi data.
+
+\subsubsection{Minke whale}
+
+First we fit a model to the minke whale data, setting the truncation at 1.5km and using the default options in \code{ds} very simply:
+\begin{Schunk}
+\begin{Sinput}
+R> minke_hn <- ds(minke, truncation = 1.5)
+\end{Sinput}
+\end{Schunk}
+Note that when there are no covariates in the model, \code{ds} will add adjustment terms to the model until there is no improvement in AIC.
+
+Figure \ref{fig:minkeamakihi} (left panel) shows the result of calling \code{plot} on the resulting model object.
+We can also call \code{summary} on the model object to get summary information about the fitted model (we postpone this until the next section).
+
+A different form for the detection function can be specified via the \code{key=} argument to \code{ds}. For example, a hazard rate model can be fitted as:
+\begin{Schunk}
+\begin{Sinput}
+R> minke_hrcos <- ds(minke, truncation = 1.5, key = "hr")
+\end{Sinput}
+\end{Schunk}
+Here \code{ds} also fits the hazard-rate model then hazard-rate with a cosine adjustment but the AIC improvement is insufficient to select the adjustment, so the hazard-rate key-only model is returned.
+
+Other adjustment series can be selected using the \code{adjustment=} argument and specific orders of adjustments can be set using \code{order=}.
+For example, to specify a uniform model with cosine adjustments of order 1 and 2 we can write:
+\begin{Schunk}
+\begin{Sinput}
+R> minke_unifcos <- ds(minke, truncation = 1.5, key = "unif",
++                     adjustment = "cos", order = c(1, 2))
+\end{Sinput}
+\end{Schunk}
+Hermite polynomial adjustments use the code \code{"herm"} and simple polynomials \code{"poly"}, adjustment order should be in line with Table \ref{tab:keyadj}.
+
+\subsubsection{Amakihi}
+
+\code{ds} assumes the data given to it has been collected as line transects, but we can switch to point transects using the argument \code{transect="point"}.
+We can include covariates in the scale parameter via the \code{formula=~...} argument to \code{ds}.
+A hazard-rate model for the amakihi that includes observer as a covariate and a truncation distance of 82.5m \citep{Marques:2007ey} can be specified using :
+\begin{Schunk}
+\begin{Sinput}
+R> amakihi_hr_obs <- ds(amakihi, truncation = 82.5, transect = "point",
++                      key = "hr", formula = ~obs)
+\end{Sinput}
+\end{Schunk}
+Note that here, unlike with the minke whale data, \code{ds} warns us that we have only supplied enough information to estimate the detection function (not density or abundance).
+
+While automatic AIC selection is performed on adjustment terms, model selection for covariates must be performed manually.
+Here we add a second covariate: minutes after sunrise.
+We will compare these two models further in the following section.
+\begin{Schunk}
+\begin{Sinput}
+R> amakihi_hr_obs_mas <- ds(amakihi, truncation = 82.5, transect = "point",
++                          key = "hr", formula = ~obs + mas)
+\end{Sinput}
+\end{Schunk}
+As with the minke whale model, we can plot the resulting models (Figure \ref{fig:minkeamakihi}, middle and right panels).
+However, for point transect studies, probability density function plots give a better sense of model fit than the detection function plots.
+This is because when plotting the detection function for point transect data, the histogram must be rescaled to account for the geometry of the point sampler.
+The amakihi models included covariates, so the plots show the detection function averaged over levels/values of the covariate.
+Points on the plot indicate probability of detection for each observation.
+For the \code{amakihi_hr_obs} model we see fairly clear levels of the observer covariate in the points.
+Looking at the right panel of Figure \ref{fig:minkeamakihi}, this is less clear when adding minutes after sunrise (a continuous covariates) to the model.
+
 \begin{figure}
-
-{\centering \includegraphics{paper_files/figure-latex/minke-amakihi-hn-plot-1}
-
-}
-
+\begin{center}
+\includegraphics{paper-minke-amakihi-hn-plot}
 \caption{Left: fitted detection function overlayed on the histogram of observed distances for the minke whale data using half-normal model.
@@
 Centre, hazard-rate with observer as a covariate; right, hazard-rate model with observer and minutes after sunrise as covariates.
-Points indicate probability of detection for a given observation (given that observation's covariate values) and lines indicate the average detection function.\label{fig:minkeamakihi}}\label{fig:minke-amakihi-hn-plot}
+Points indicate probability of detection for a given observation (given that observation's covariate values) and lines indicate the average detection function.}
+\label{fig:minkeamakihi}
+\end{center}
 \end{figure}
-\end{CodeChunk}
-
-\section{Model checking and model
-selection}\label{model-checking-and-model-selection}
-
-As with models fitted using \code{lm} or \code{glm} in \proglang{R}, we
-can use \code{summary} to give useful information about our fitted
-model. For our hazard-rate model of the amakihi data, with observer as a
-covariate:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-summary(amakihi_hr_obs)
-\end{CodeInput}
-\begin{CodeOutput}
-
+
+\section{Model checking and model selection}
+
+As with models fitted using \code{lm} or \code{glm} in \proglang{R}, we can use \code{summary} to give useful information about our fitted model.
+For our hazard-rate model of the amakihi data, with observer as a covariate:
+\begin{Schunk}
+\begin{Sinput}
+R> summary(amakihi_hr_obs)
+\end{Sinput}
+\begin{Soutput}
 Summary for distance analysis 
@@
 N in covered region 3955.1685709 274.2284029 0.06933419
-\end{CodeOutput}
-\end{CodeChunk}
-
-This summary information includes details of the data and model
-specification, as well as the values of the coefficients (\(\beta_j\))
-and their uncertainties, an ``average'' value for the detectability (see
-``Estimating abundance and variance'' for details on how this is
-calculated) and its uncertainty. The final line gives an estimate of
-abundance in the area covered by the survey (see the next section;
-though note this estimate does not take into account cluster size).
-
-\subsection{Goodness of fit}\label{goodness-of-fit}
-
-We use a quantile-quantile plot (Q-Q plot) to visually assess how well a
-detection functions fits the data when we have exact distances. Here the
-Q-Q plot compares the cumulative distribution function (CDF) of the
-fitted detection function to the distribution of the data (empirical
-distribution function or EDF). The Q-Q plots in \pkg{Distance} plot a
-point for every observation. The EDF is the proportion of points that
-have been observed at a distance equal to or less than the distance of
-that observation. The CDF is calculated from the fitted detection
-function as the probability of observing an object at a distance less
-than or equal to that of the given observation. This can be interpreted
-as assessing whether the number of observations up to a given distance
-is in line with what the model says they should be. As usual for Q-Q
-plots, ``good'' models will have values close to the line \(y=x\), poor
-models will show greater deviations from that line.
-
-Q-Q plots can be inspected visually, though this is prone to subjective
-judgments. Therefore, we also quantify the Q-Q plot's information using
-a Kolmogorov-Smirnov or Cramér-von Mises test \citep{Burnham:2004vd}.
-Both test whether points from the EDF and CDF are from the same
-distribution. The Kolmogorov-Smirnov uses the test statistic of the
-largest difference between a point on the Q-Q plot and the line \(y=x\),
-whereas the Cramér-von Mises test uses the sum of all the differences.
-As it takes into account more information and is therefore more
-powerful, the Cramér-von Mises is generally preferred. A significant
-result from either test gives evidence against the null hypothesis (that
-the data arose from the fitted model), suggesting that the model does
-not fit the data well.
-
-We can generate a Q-Q plot and test results using the \code{gof_ds}
-function. Figure \ref{amakihi-qq} shows the goodness of fit tests for
-two models for the amakihi data. We first fit a half-normal model
-without covariates or adjustments (setting \code{adjustment=NULL} will
-force \code{ds} to fit a model with no adjustments):
-
-\begin{CodeChunk}
-\begin{CodeInput}
-amakihi_hn <- ds(amakihi, truncation = 82.5, transect = "point",
-                 key = "hn", adjustment = NULL)
-\end{CodeInput}
-\begin{CodeOutput}
-Fitting half-normal key function
-\end{CodeOutput}
-\begin{CodeOutput}
-Key only model: not constraining for monotonicity.
-\end{CodeOutput}
-\begin{CodeOutput}
-AIC= 10833.841
-\end{CodeOutput}
-\begin{CodeOutput}
-No survey area information supplied, only estimating detection function.
-\end{CodeOutput}
-\begin{CodeInput}
-gof_ds(amakihi_hn)
-\end{CodeInput}
-\begin{CodeOutput}
-
+\end{Soutput}
+\end{Schunk}
+This summary information includes details of the data and model specification, as well as the values of the coefficients ($\beta_j$) and their uncertainties, an ``average'' value for the detectability (see ``Estimating abundance and variance'' for details on how this is calculated) and its uncertainty.
+The final line gives an estimate of abundance in the area covered by the survey (see the next section; though note this estimate does not take into account cluster size).
+
+\subsection{Goodness of fit}
+
+We use a quantile-quantile plot (Q-Q plot) to visually assess how well a detection functions fits the data when we have exact distances.
+Here the Q-Q plot compares the cumulative distribution function (CDF) of the fitted detection function to the distribution of the data (empirical distribution function or EDF).
+The Q-Q plots in \pkg{Distance} plot a point for every observation.
+The EDF is the proportion of points that have been observed at a distance equal to or less than the distance of that observation.
+The CDF is calculated from the fitted detection function as the probability of observing an object at a distance less than or equal to that of the given observation.
+This can be interpreted as assessing whether the number of observations up to a given distance is in line with what the model says they should be.
+As usual for Q-Q plots, ``good'' models will have values close to the line $y=x$, poor models will show greater deviations from that line.
+
+Q-Q plots can be inspected visually, though this is prone to subjective judgments.
+Therefore, we also quantify the Q-Q plot's information using a Kolmogorov-Smirnov or Cram\'{e}r-von Mises test \citep{Burnham:2004vd}.
+Both test whether points from the EDF and CDF are from the same distribution.
+The Kolmogorov-Smirnov uses the test statistic of the largest difference between a point on the Q-Q plot and the line $y=x$, whereas the Cram\'{e}r-von Mises test uses the sum of all the differences.
+As it takes into account more information and is therefore more powerful, the Cram\'{e}r-von Mises is generally preferred.
+A significant result from either test gives evidence against the null hypothesis (that the data arose from the fitted model), suggesting that the model does not fit the data well.
+
+We can generate a Q-Q plot and test results using the \code{gof_ds} function.
+Figure \ref{amakihi-qq} shows the goodness of fit tests for two models for the amakihi data.
+We first fit a half-normal model without covariates or adjustments (setting \code{adjustment=NULL} will force \code{ds} to fit a model with no adjustments):
+\begin{Schunk}
+\begin{Sinput}
+R> amakihi_hn <- ds(amakihi, truncation = 82.5, transect = "point",
++                  key = "hn", adjustment = NULL)
+R> gof_ds(amakihi_hn)
+\end{Sinput}
+\begin{Soutput}
 Goodness of fit results for ddf object
@@
 Test statistic =  0.93083  P =  0.003578 
-\end{CodeOutput}
-\begin{CodeInput}
-gof_ds(amakihi_hr_obs_mas)
-\end{CodeInput}
-\begin{CodeOutput}
-
+\end{Soutput}
+\begin{Sinput}
+R> gof_ds(amakihi_hr_obs_mas)
+\end{Sinput}
+\begin{Soutput}
 Goodness of fit results for ddf object
@@
 Test statistic =  0.15016  P =  0.38908 
-\end{CodeOutput}
-\end{CodeChunk}
-
-We can conclude that the half-normal model does not pass our goodness of
-fit tests and should be discarded. Both hazard-rate models (output only
-shown for hazard-rate model with observer and minutes after sunrise) had
-non-significant goodness of fit test statistics and are both therefore
-deemed plausible models. The corresponding Q-Q plots are shown in Figure
-\ref{amakihi-qq}, comparing the half-normal model with the hazard-rate
-model with observer and minutes after sunrise included.
-
-\begin{CodeChunk}
+\end{Soutput}
+\end{Schunk}
+We can conclude that the half-normal model does not pass our goodness of fit tests and should be discarded.
+Both hazard-rate models (output only shown for hazard-rate model with observer and minutes after sunrise) had non-significant goodness of fit test statistics and are both therefore deemed plausible models.
+The corresponding Q-Q plots are shown in Figure \ref{amakihi-qq}, comparing the half-normal model with the hazard-rate model with observer and minutes after sunrise included.
+
 \begin{figure}
-
-{\centering \includegraphics{paper_files/figure-latex/amakihi-qq-comp-1}
-
-}
-
-\caption[Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates]{Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates; left) and hazard-rate model with observer and minutes after sunrise (right) for the amakihi data.\label{amakihi-qq}}\label{fig:amakihi-qq-comp}
+\begin{center}
+\begin{Schunk}
+\begin{Soutput}
+Goodness of fit results for ddf object
+
+Distance sampling Kolmogorov-Smirnov test
+Test statistic =  0.059345  P =  0.00031527
+
+Distance sampling Cramer-von Mises test (unweighted)
+Test statistic =  0.93083  P =  0.003578 
+\end{Soutput}
+\begin{Soutput}
+Goodness of fit results for ddf object
+
+Distance sampling Kolmogorov-Smirnov test
+Test statistic =  0.036251  P =  0.076237
+
+Distance sampling Cramer-von Mises test (unweighted)
+Test statistic =  0.15016  P =  0.38908 
+\end{Soutput}
+\end{Schunk}
+\includegraphics{paper-amakihiqq}
+\caption{Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates; left) and hazard-rate model with observer and minutes after sunrise (right) for the amakihi data.}
+\label{amakihi-qq}
+\end{center}
 \end{figure}
-\end{CodeChunk}
-
-For non-exact distance data, a \(\chi^2\)-test can be used to assess
-goodness of fit \citep[see][Section 3.4.4]{Buckland:2001vm}.
-\(\chi^2\)-test results are produced by \texttt{gof\_ds} when
-binned/grouped distances are provided.
-
-\subsection{Model selection}\label{model-selection}
-
-Once we have a set of plausible models, we can use Akaike's Information
-Criterion (AIC) to select between models \citep[see
-e.g.][]{burnham2003model}. \pkg{Distance} includes a function to create
-table of summary information for fitted models, making it easy to get an
-overview of a large number of models. The \code{summarize_ds_models}
-function takes models as input and can be especially useful when paired
-with \pkg{knitr}'s \code{kable} function to create summary tables for
-publication \citep{knitr-pkg}. An example of this output (with all
-models included) is shown in Table \ref{tab:amakihi} and was generated
-by the following call to \code{summarize_ds_models}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
+
+For non-exact distance data, a $\chi^2$-test can be used to assess goodness of fit \citep[see][Section 3.4.4]{Buckland:2001vm}.
+$\chi^2$-test results are produced by \code{gof_ds} when binned/grouped distances are provided.
+
+\subsection{Model selection}
+
+Once we have a set of plausible models, we can use Akaike's Information Criterion (AIC) to select between models \citep[see e.g.][]{burnham2003model}.
+\pkg{Distance} includes a function to create table of summary information for fitted models, making it easy to get an overview of a large number of models.
+The \code{summarize_ds_models} function takes models as input and can be especially useful when paired with \pkg{knitr}'s \code{kable} function to create summary tables for publication \citep{knitr-pkg}.
+An example of this output (with all models included) is shown in Table \ref{tab:amakihi} and was generated by the following call to \code{summarize_ds_models}:
+
+\begin{Code}
 summarize_ds_models(amakihi_hn, amakihi_hr_obs, amakihi_hr_obs_mas)
-\end{CodeInput}
-\end{CodeChunk}
-
-In this case we may be skeptical about the top model as selected by AIC
-being truly better than the second best model, as there is only a very
-small difference in AICs. Generally, if the difference between AICs is
-less than 2, we may investigate multiple ``best'' models, potentially
-resorting to the simplest of these models. In the authors' experience,
-it is often the case that models with similar AICs will have similar
-estimates probabilities of detection, so in practice there is little
-difference in selecting between these models. It is important to note
-that comparing AICs between models with different truncations is not
-appropriate, as models with different truncation use different data.
+\end{Code}
+
+In this case we may be skeptical about the top model as selected by AIC being truly better than the second best model, as there is only a very small difference in AICs.
+Generally, if the difference between AICs is less than 2, we may investigate multiple ``best'' models, potentially resorting to the simplest of these models.
+In the authors' experience, it is often the case that models with similar AICs will have similar estimates probabilities of detection, so in practice there is little difference in selecting between these models.
+It is important to note that comparing AICs between models with different truncations is not appropriate, as models with different truncation use different data.
 
@@
 
+\caption{Summary for the detection function models fitted to the amakihi data.
+``C-vM'' stands for Cram\'{e}r-von Mises, $P_a$ is average detectability (see ``Estimating abundance and variance''), se is standard error.
+Models are sorted according to AIC.\label{tab:amakihi}}
 \centering
@@
 \end{tabular}
-\caption{Summary for the detection function models fitted to the amakihi data.
-``C-vM'' stands for Cramér-von Mises, $P_a$ is average detectability (see ``Estimating abundance and variance''), se is standard error.
-Models are sorted according to AIC.\label{tab:amakihi}}
 \end{table}
-
-\section{Estimating abundance and
-variance}\label{estimating-abundance-and-variance}
-
-Though fitting the detection function is the primary modelling step in
-distance sampling, we are really interested in estimating density or
-abundance. It is also important to calculate our uncertainty for these
-estimates. This section addresses these issues mathematically before
-showing how to estimate abundance and its variance in \proglang{R}.
-
-\subsection{Abundance}\label{abundance}
-
-We wish to obtain the abundance in a study region, of which we have
-sampled a random subset. To do this we first calculate the abundance in
-the area we have surveyed (the \emph{covered area}) to obtain
-\(\hat{N}_\text{C}\), we can then scale this up to the full study area
-by multiplying it by the ratio of covered area to study area. We discuss
-other methods for spatially explicit abundance estimation in
-``Extensions''.
-
-First, to estimate abundance in the covered area (\(\hat{N}_\text{C}\)),
-we use the estimates of detection probability (the
-\(\{\hat{p}(\mathbf{z}_i); i=1,\ldots,n\}\), above) in a
-Horvitz-Thompson-like estimator:
-
+\section{Estimating abundance and variance}
+
+Though fitting the detection function is the primary modelling step in distance sampling, we are really interested in estimating density or abundance.
+It is also important to calculate our uncertainty for these estimates.
+This section addresses these issues mathematically before showing how to estimate abundance and its variance in \proglang{R}.
+
+\subsection{Abundance}
+
+We wish to obtain the abundance in a study region, of which we have sampled a random subset.
+To do this we first calculate the abundance in the area we have surveyed (the \textit{covered area}) to obtain $\hat{N}_\text{C}$, we can then scale this up to the full study area by multiplying it by the ratio of covered area to study area.
+We discuss other methods for spatially explicit abundance estimation in ``Extensions''.
+
+First, to estimate abundance in the covered area ($\hat{N}_\text{C}$), we use the estimates of detection probability (the $\{\hat{p}(\mathbf{z}_i); i=1,\ldots,n\}$, above) in a Horvitz-Thompson-like estimator:
 \begin{equation}
@@
 \end{equation}
-
-where \(s_i\) are the sizes of the observed clusters of objects, which
-are all equal to 1 if objects only occur singly \citep{Borchers:2004wr}.
-\citet{thompson2012sampling} is the canonical reference to this type of
-estimator. Intuitively, we can think of the estimates of detectability
-(\(\hat{p}(\mathbf{z}_i)\)) as ``inflating'' the cluster sizes
-(\(s_i\)), we then sum over the detections (\(i\)) to obtain the
-abundance estimate. For models that do not include covariates,
-\(\hat{p}(\mathbf{z}_i)\) is equal for all \(i\), so this is equivalent
-to summing the clusters and inflating that sum by dividing through by
-the corresponding \(\hat{p} (=\hat{p}(\mathbf{z}_i) \forall i)\).
-
-Having obtained the abundance in the covered area, we can then scale-up
-to the study area: \[
+where $s_i$ are the sizes of the observed clusters of objects, which are all equal to 1 if objects only occur singly \citep{Borchers:2004wr}.
+\cite{thompson2012sampling} is the canonical reference to this type of estimator.
+Intuitively, we can think of the estimates of detectability ($\hat{p}(\mathbf{z}_i)$) as ``inflating'' the cluster sizes ($s_i$), we then sum over the detections ($i$) to obtain the abundance estimate.
+For models that do not include covariates, $\hat{p}(\mathbf{z}_i)$ is equal for all $i$, so this is equivalent to summing the clusters and inflating that sum by dividing through by the corresponding $\hat{p} (=\hat{p}(\mathbf{z}_i) \forall i)$.
+
+Having obtained the abundance in the covered area, we can then scale-up to the study area:
+$$
 \hat{N} = \frac{A}{a} \hat{N}_\text{C},
-\] where \(A\) is the area of the study region to which to extrapolate
-the abundance estimate and \(a\) is the covered area. For line transects
-\(a=2wL\) (twice the truncation distance multiplied by the total length
-of transects surveyed, \(L\)) and for points \(a=\pi w^2 T\) (where
-\(\pi w^2\) is the area of a single surveyed circle and \(T\) is the sum
-of the number of visits to the sampled points).
-
-We can use the Horvitz-Thompson-like estimator to calculate the
-``average'' detectability for models which include covariates. We can
-consider what single detectability value would give the estimated
-\(\hat{N}_\text{C}\) and therefore calculate: \[
+$$
+where $A$ is the area of the study region to which to extrapolate the abundance estimate and $a$ is the covered area.
+For line transects $a=2wL$ (twice the truncation distance multiplied by the total length of transects surveyed, $L$) and for points $a=\pi w^2 T$ (where $\pi w^2$ is the area of a single surveyed circle and $T$ is the sum of the number of visits to the sampled points).
+
+We can use the Horvitz-Thompson-like estimator to calculate the ``average'' detectability for models which include covariates.
+We can consider what single detectability value would give the estimated $\hat{N}_\text{C}$ and therefore calculate:
+$$
 \hat{P_a} = n/\hat{N}_\text{C}.
-\] This can be a useful summary statistic, giving us an idea of how
-detectable our \(n\) observed animals would have to be to estimate the
-same \(\hat{N}\) if there were no observed covariates. It can also be
-compared to similar estimates in mark-recapture studies. \(P_a\) is
-included in the \code{summary} output and the table produced by
-\code{summarize_ds_models}.
-
-\subsubsection{Stratification}\label{stratification}
-
-We may wish to calculate abundance estimates for some sub-regions of the
-study region, we call these areas \emph{strata}. Stratification can be
-used to increase the precision of estimates if we know \emph{a priori}
-that density varies between different parts of the study area. For
-example, strata may be defined by habitat types or animal sex (or some
-combination) which may be of interest for biological or management
-reasons. To calculate estimates for a given stratification each
-observation must occur in a stratum which must be labelled with a
-\code{Region.Label} and have a corresponding \code{Area} (if we are
-using an animal characteristic like sex, we would have the areas be the
-same but if we were using say forested vs.~wetland habitat the areas of
-those strata would be different). Finally, we must also know the stratum
-in which each observation occurs.
-
-As an example, the minke whale data contain two strata: \code{North} and
-\code{South} relating to strata further away from and nearer to the
-Antarctic ice edge, respectively.
-
-\subsection{Variance}\label{variance}
-
-We take an intuitive approach to uncertainty estimation, for a full
-derivation consult \citet{Marques:2003vb}. Uncertainty in \(\hat{N}\)
-comes from two sources:
+$$
+This can be a useful summary statistic, giving us an idea of how detectable our $n$ observed animals would have to be to estimate the same $\hat{N}$ if there were no observed covariates.
+It can also be compared to similar estimates in mark-recapture studies.
+$P_a$ is included in the \code{summary} output and the table produced by \code{summarize_ds_models}.
+
+\subsubsection{Stratification}
+
+We may wish to calculate abundance estimates for some sub-regions of the study region, we call these areas \textit{strata}.
+Stratification can be used to increase the precision of estimates if we know \textit{a priori} that density varies between different parts of the study area.
+For example, strata may be defined by habitat types or animal sex (or some combination) which may be of interest for biological or management reasons.
+To calculate estimates for a given stratification each observation must occur in a stratum which must be labelled with a \code{Region.Label} and have a corresponding \code{Area} (if we are using an animal characteristic like sex, we would have the areas be the same but if we were using say forested vs.
+wetland habitat the areas of those strata would be different).
+Finally, we must also know the stratum in which each observation occurs.
+
+As an example, the minke whale data contain two strata: \code{North} and \code{South} relating to strata further away from and nearer to the Antarctic ice edge, respectively.
+
+\subsection{Variance}
+
+We take an intuitive approach to uncertainty estimation, for a full derivation consult \cite{Marques:2003vb}. Uncertainty in $\hat{N}$ comes from two sources:
 
 \begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  \emph{Detection function}: Uncertainty in parameter
-  (\(\boldsymbol{\theta}\)) estimation.
-\item
-  \emph{Encounter rate}: Sampling variability arising from differences
-  in the number of observations per transect.
+\item \textit{Detection function}: Uncertainty in parameter ($\boldsymbol{\theta}$) estimation.
+\item \textit{Encounter rate}: Sampling variability arising from differences in the number of observations per transect.
 \end{enumerate}
 
-We can see this by looking at the Horvitz-Thompson-like estimation in
-(\ref{ht}) and consider the terms which are random. These are: the
-detectability \(\hat{p}(\mathbf{z}_i)\) (and hence the parameters of the
-detection function it is derived from) and \(n\), the number of
-observations.
-
-Model parameter uncertainty can be addressed using standard maximum
-likelihood theory. We can invert the Hessian matrix of the likelihood to
-obtain a variance-covariance matrix. We can then pre- and post-multiply
-this by the derivatives of \(\hat{N}_\text{C}\) with respect to the
-parameters of the detection function \[
-\widehat{\text{Var}}_\text{model}\left( \hat{N}_\text{C}\right) = \left(\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}\right)^\text{T} \left(\hat{\mathbf{H}}(\hat{\boldsymbol{\theta}})^{-1} \right)\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}
-\] where the partial derivatives of \(\hat{N}_\text{C}\) are evaluated
-at the MLE (\(\hat{\theta}\)) and \(\mathbf{H}\) is the first partial
-Hessian (outer product of first derivatives of the log likelihood) for
-numerical stability \citep[p 62]{Buckland:2001vm}. Note that although we
-calculate uncertainty in \(\hat{N}_\text{C}\), we can trivially scale-up
-to variance of \(\hat{N}\) (by noting that
-\(\hat{N} = \frac{A}{a} \hat{N}_\text{C}\) and therefore
-\(\widehat{\text{Var}}_\text{model}\left(\hat{N}\right) = \left( \frac{A}{a} \right)^2 \widehat{\text{Var}}_\text{model} \left(\hat{N}_\text{C} \right)\)).
-
-\emph{Encounter rate} is the number of objects per unit transect (rather
-than just \(n\)). When covariates are not included in the detection
-function we can define the encounter rate as \(n/L\) for line transects
-(where \(L\) is the total line length) or \(n/T\) for point transects
-(where \(T\) is the total number of visits summed over all points). When
-covariates are included in the detection function, it is recommended
-that we substitute the \(n\) in the encounter rate with the estimated
-abundance \(\hat{N}_\text{C}\) as this will take into account the
-effects of the covariates \citep{Innes:2002ka}.
-
-For line transects, by default, \pkg{Distance} uses a variation of the
-estimator ``R2'' from \citet{Fewster:2009ku} replacing number of
-observations per sample with the estimated abundance per sample, thus
-taking into account cluster size if this is recorded
-\citep{Innes:2002ka, Marques:2003vb}: \[
+We can see this by looking at the Horvitz-Thompson-like estimation in (\ref{ht}) and consider the terms which are random.
+These are: the detectability $\hat{p}(\mathbf{z}_i)$ (and hence the parameters of the detection function it is derived from) and $n$, the number of observations.
+
+Model parameter uncertainty can be addressed using standard maximum likelihood theory.
+We can invert the Hessian matrix of the likelihood to obtain a variance-covariance matrix.
+We can then pre- and post-multiply this by the derivatives of $\hat{N}_\text{C}$ with respect to the parameters of the detection function
+$$
+\widehat{\text{Var}}_\text{model}\left( \hat{N}_\text{C}\right) = \left(\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}\right)^\top \left(\hat{\mathbf{H}}(\hat{\boldsymbol{\theta}})^{-1} \right)\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}
+$$
+where the partial derivatives of $\hat{N}_\text{C}$ are evaluated at the MLE ($\hat{\theta}$) and $\mathbf{H}$ is the first partial Hessian (outer product of first derivatives of the log likelihood) for numerical stability \citep[][p 62]{Buckland:2001vm}.
+Note that although we calculate uncertainty in $\hat{N}_\text{C}$, we can trivially scale-up to variance of $\hat{N}$ (by noting that $\hat{N} = \frac{A}{a} \hat{N}_\text{C}$ and therefore $\widehat{\text{Var}}_\text{model}\left(\hat{N}\right) = \left( \frac{A}{a} \right)^2 \widehat{\text{Var}}_\text{model} \left(\hat{N}_\text{C} \right)$).
+
+\textit{Encounter rate} is the number of objects per unit transect (rather than just $n$).
+When covariates are not included in the detection function we can define the encounter rate as $n/L$ for line transects (where $L$ is the total line length) or $n/T$ for point transects (where $T$ is the total number of visits summed over all points).
+When covariates are included in the detection function, it is recommended that we substitute the $n$ in the encounter rate with the estimated abundance $\hat{N}_\text{C}$ as this will take into account the effects of the covariates \citep{Innes:2002ka}.
+
+For line transects, by default, \pkg{Distance} uses a variation of the estimator ``R2'' from \cite{Fewster:2009ku} replacing number of observations per sample with the estimated abundance per sample, thus taking into account cluster size if this is recorded \citep{Innes:2002ka, Marques:2003vb}:
+$$
 \widehat{\text{Var}}_{\text{encounter},R2}\left( \hat{N}_\text{C}\right) = \frac{K}{L^2(K-1)} \sum_{k=1}^{K} l_k^2 \left( \frac{\hat{N_{\text{C}, k}}}{l_k} - \frac{\hat{N_\text{C}}}{L}\right)^2,
-\] where \(l_k\) are the lengths of the \(K\) transects (such that
-\(L = \sum_{k=1}^K l_k\)) and \(\hat{N_{C,k}}\) is the abundance in the
-covered area for transect \(k\). For point transects we use the
-estimator ``P3'' from \citet{Fewster:2009ku} but again replace \(n\) by
-\(\hat{N}_\text{C}\) in the encounter rate definition, to obtain the
-following estimator: \[
+$$
+where $l_k$ are the lengths of the $K$ transects (such that $L = \sum_{k=1}^K l_k$) and $\hat{N_{C,k}}$ is the abundance in the covered area for transect $k$.
+For point transects we use the estimator ``P3'' from \cite{Fewster:2009ku} but again replace $n$ by $\hat{N}_\text{C}$ in the encounter rate definition, to obtain the following estimator:
+$$
 \widehat{\text{Var}}_{\text{encounter},P3}\left( \hat{N}_\text{C}\right) = \frac{1}{T(K-1)} \sum_{k=1}^{K} t_k \left( \frac{\hat{N_{\text{C}, k}}}{t_k} - \frac{\hat{N_\text{C}}}{T}\right)^2,
-\] where \(t_k\) is the number of visits to point \(k\) and
-\(T = \sum_{k=1}^K t_k\) (the total number of visits to all points is
-the sum of the visits to each point). Again, it is straightforward to
-calculate the encounter rate variance for \(\hat{N}\) from the encounter
-rate variance for \(\hat{N}_\text{C}\).
-
-Other formulations for the encounter rate variance are discussed in
-detail in \citet{Fewster:2009ku}. \pkg{Distance} implements all of the
-estimators of encounter rate variance given in that article. The
-\code{varn} manual page gives further advice and technical detail on
-encounter rate variance. For example for systematic survey designs,
-estimators S1, S2 and O1, O2 and O3 will typically provide smaller
-estimates of the variance.
-
-We combine these two variances by noting that squared coefficients of
-variation (approximately) add \citep[often referred to as ``the delta
-method'';][]{seber2002estimation}.
-
-\subsection{Estimating abundance and variance in
-R}\label{estimating-abundance-and-variance-in-r}
-
-Returning to the minke whale data, we have the necessary information to
-calculate \(A\) and \(a\) above, so we can estimate abundance and its
-variance. When we supply data to \code{ds} in the ``flatfile'' format
-given above, \code{ds} will automatically calculate abundance estimates
-based on the survey information in the data.
-
-Having already fitted a model to the minke whale data, we can see the
-results of the abundance estimation by viewing the model summary:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-summary(minke_hn)
-\end{CodeInput}
-\begin{CodeOutput}
-
+$$
+where $t_k$ is the number of visits to point $k$ and $T = \sum_{k=1}^K t_k$ (the total number of visits to all points is the sum of the visits to each point).
+Again, it is straightforward to calculate the encounter rate variance for $\hat{N}$ from the encounter rate variance for $\hat{N}_\text{C}$.
+
+Other formulations for the encounter rate variance are discussed in detail in \cite{Fewster:2009ku}.
+\pkg{Distance} implements all of the estimators of encounter rate variance given in that article.
+The \code{varn} manual page gives further advice and technical detail on encounter rate variance.
+For example for systematic survey designs, estimators S1, S2 and O1, O2 and O3 will typically provide smaller estimates of the variance.
+
+We combine these two variances by noting that squared coefficients of variation (approximately) add \citep[often referred to as ``the delta method'';][]{seber2002estimation}.
+
+\subsection[Estimating abundance and variance in R]{Estimating abundance and variance in \proglang{R}}
+
+Returning to the minke whale data, we have the necessary information to calculate $A$ and $a$ above, so we can estimate abundance and its variance.
+When we supply data to \code{ds} in the ``flatfile'' format given above, \code{ds} will automatically calculate abundance estimates based on the survey information in the data.
+
+Having already fitted a model to the minke whale data, we can see the results of the abundance estimation by viewing the model summary:
+\begin{Schunk}
+\begin{Sinput}
+R> summary(minke_hn)
+\end{Sinput}
+\begin{Soutput}
 Summary for distance analysis 
@@
 3 Total 0.02403400 0.007179465 0.2987212 0.012838347 0.04499280 14.00459
-\end{CodeOutput}
-\end{CodeChunk}
-
-This prints a rather large amount of information: first the detection
-function summary, then three tables:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  \code{Summary statistics}: giving the areas, covered areas, effort,
-  number of observations, number of transects, encounter rate, its
-  standard error and coefficient of variation for each stratum.
-\item
-  \code{Abundance}: giving estimates, standard errors, coefficients of
-  variation, lower and upper confidence intervals and finally the
-  degrees of freedom for each stratum's abundance estimate.
-\item
-  \code{Density}: lists the same statistics as \code{Abundance} but for
-  a density estimate.
-\end{enumerate}
+\end{Soutput}
+\end{Schunk}
+This prints a rather large amount of information: first the detection function summary, then three tables:
+
+1. \code{Summary statistics}: giving the areas, covered areas, effort, number of observations, number of transects, encounter rate, its standard error and coefficient of variation for each stratum. 
+2. \code{Abundance}: giving estimates, standard errors, coefficients of variation, lower and upper confidence intervals and finally the degrees of freedom for each stratum's abundance estimate.
+3. \code{Density}: lists the same statistics as \code{Abundance} but for a density estimate.
 
@@
 
-The summary can be more concisely expressed by extracting information
-from the summary object. This object is a \code{list} of
-\code{data.frames}, so we can use the \code{kable} function from
-\pkg{knitr} to create summary tables of abundance estimates and measures
-of precision, such as Table \ref{minke-abund}. We prepare the
-\code{data.frame} as follows before using \code{kable}:
-
-\begin{CodeChunk}
-\begin{CodeInput}
+The summary can be more concisely expressed by extracting information from the summary object.
+This object is a \code{list} of \code{data.frames}, so we can use the \code{kable} function from \pkg{knitr} to create summary tables of abundance estimates and measures of precision, such as Table \ref{minke-abund}.
+We prepare the \code{data.frame} as follows before using \code{kable}:
+
+\begin{Code}
 minke_table <- summary(minke_hn)$dht$individuals$N
@@
                            "$\\text{CV}(\\hat{N}$)")
-\end{CodeInput}
-\end{CodeChunk}
-
-\begin{CodeChunk}
+\end{Code}
+
 \begin{table}
 
+\caption{Summary of abundance estimation for the half-normal model for the minke whale data.\label{minke-abund}}
 \centering
@@
 \end{tabular}
-\caption{Summary of abundance estimation for the half-normal model for the minke whale data.\label{minke-abund}}
 \end{table}
-
-\end{CodeChunk}
-
-\section{Extensions}\label{extensions}
+\section{Extensions}
 
 Distance sampling has been applied in a wide variety of situations.
-Objects being detected need not be animals; plants
-\citep{Buckland:2007kl}, beer cans \citep{Otto:1990gs}, elephant dung
-\citep{NchPlu01} and bricks on a lake bottom \citep{Ber90} have been
-subjects of distance sampling investigations. The method of detection
-need not be visual sightings. Detections of objects can be through
-auditory as well as visual means \citep{Marques2013b}. Songs of birds
-\citep{Buckland2006} and whale vocalisations \citep{BorSub1} are just
-two examples. Blows made by whales are examples of processes that can be
-modelled using distance sampling. Songs and blows are indirect sampling
-methods producing estimates of cue density. Cue densities can be
-converted to animal densities with additional data needed to estimate
-the rate at which cues are produced and the rate at which they
-disappear. These are but a few examples of the applications to which
-distance sampling has been applied (a larger, but incomplete list of
-references is given at \url{http://distancesampling.org/dbib.html}).
-
-The features of \pkg{Distance} are deliberately limited to provide a
-simplified interface for users. For more complex analyses of distance
-sampling data, we provide additional packages for modelling in
-\proglang{R}.
-
-We noted at the start of the article that \pkg{Distance} is a
-simple-to-use wrapper around the package \pkg{mrds}. Additional features
-available in \pkg{mrds} include models that relax the assumption that
-detection is certain at zero distance from the transect (by including
-data from additional observers). This is done using mark-recapture type
-methods which require additional survey methodology, known as double
-observer surveys \citep[see][ for an introduction]{Burt:2014gu}.
-
-\pkg{Distance} can provide us with estimates of abundance or density for
-each strata as a whole but tells us nothing about the distribution of
-animals within strata. One option is to divide the study area into
-smaller and smaller strata to try to detect patterns in spatial
-distribution, however, a more rigorous approach is to build a spatial
-model. Such models incorporate spatially-referenced environmental data
-(for example derived from GIS products). \pkg{Distance} interfaces with
-one such package used to perform this type of analysis: \pkg{dsm}
-\citep{dsm-pkg}. So-called ``density surface modelling'' uses the
-generalized additive model framework \citep[e.g.][]{Wood:2006vg} to
-build models of abundance (adjusting counts for imperfect detectability)
-as a function of environmental covariates, as part of a two stage model
-\citep{Hedley:2004et, Miller:2013fq}.
-
-Uncertainty in measured covariates (e.g.~cluster size) and model
-uncertainty (when two models have similar fit but substantially
-different estimates) can be incorporated using the multi-analysis
-distance sampling package \pkg{mads} \citep{mads-pkg}. In addition,
-\pkg{mads} can also incorporate sightings with unknown species
-identification. This is done by estimating the abundance of these
-`unidentified' sightings and pro-rating them to the known species
-\citep{Gerrodette:2005}.
-
-As mentioned above, survey design is critical to ensuring that resulting
-distance sampling data can be reliably analysed. \pkg{DSsim} allows
-users to test out different designs in their study region and tailor
-population attributes to reflect the species they are working with.
-\pkg{DSsim} \citep{DSsim-pkg} allows users to more easily identify
-challenges unique to their study and select a survey design which is
-more likely to yield accurate and precise estimates.
-
-Distance for Windows has many users (over 45,000 downloads since 2002)
-and they may be overwhelmed by the prospect of switching existing
-analyses to \proglang{R}. For that reason we have created the
-\pkg{readdst} \citep{readdst-pkg} package to interface with projects
-created by Distance for Windows. The package can take analyses created
-using the CDS, MCDS and MRDS engines in Distance for Windows, extract
-the data and create equivalent models in \proglang{R}. \pkg{readdst} can
-also run these analyses and test the resulting statistics (for example,
-\(\hat{N}\) or \(\hat{P_a}\)) calculated in \proglang{R} against those
-calculated by Distance for Windows. We hope that \pkg{readdst} will
-provide a useful transition to \proglang{R} for interested users.
-\pkg{readdst} is currently available on GitHub at
-\url{https://github.com/distancedevelopment/readdst}.
-
-\section{Conclusion}\label{conclusion}
-
-We have given an introduction as to how to perform a distance sampling
-analysis in \proglang{R}. We have covered the possible models for
-detectability, model checking and selection and finally abundance and
-variance estimation.
-
-In combination with tools such as \pkg{knitr} and \pkg{rmarkdown}
-\citep{rmarkdown-pkg}, the helper functions in \pkg{Distance} provide a
-useful set of tools to perform reproducible analyses of wildlife
-abundance for both managers and ecologists. We also direct readers'
-attention to \pkg{DsShiny} \citep{dsshiny-pkg}, a package that builds on
-\pkg{Shiny} \citep{shiny-pkg} and \pkg{mrds} to allow users to build and
-fit models in a graphical interface, with output to RMarkdown.
-
-\proglang{R} and its extension packages provide many tools exploratory
-data analysis that can be useful for a distance sampling analysis. We
-hope that this paper provides useful examples for those wishing to
-pursue distance sampling in \proglang{R}. More information on distance
-sampling can be found at \url{http://distancesampling.org} and a mailing
-list is maintained at
-\url{https://groups.google.com/forum/#!forum/distance-sampling}.
-
-We note that there are other packages available for performing distance
-sampling analyses in \proglang{R} but believe that \pkg{Distance} is the
-most flexible and feature-complete, and provides pathways to a range of
-more complicated analyses. Appendix A gives a feature comparison between
-\pkg{Distance} and other \proglang{R} packages for analysis of distance
-sampling data.
-
-\section{Acknowledgements}\label{acknowledgements}
-
-The authors would like to thank the many users of \pkg{Distance},
-\pkg{mrds} and Distance for Windows who have contributed bug reports and
-suggestions for improvements over the years. We would particularly like
-to thank Steve Buckland, David Borchers, Tiago Marques, Jon Bishop and
-Lorenzo Milazzo for their contributions. We thank Tiago Marques a second
-time for helpful comments on an early version of the paper and Colin
-Beale who suggested the ``flatfile'' data format. We also thank Ken
-Burnham and David Anderson for fundamental contributions to the early
-development of these methods.
-
-\renewcommand\refname{Bibliography}
-\bibliography{jstatsoft.bib}
+Objects being detected need not be animals; plants \citep{Buckland:2007kl}, beer cans \citep{Otto:1990gs}, elephant dung \citep{NchPlu01} and bricks on a lake bottom \citep{Ber90} have been subjects of distance sampling investigations.
+The method of detection need not be visual sightings.
+Detections of objects can be through auditory as well as visual means \citep{Marques2013b}.
+Songs of birds \citep{Buckland2006} and whale vocalisations \citep{BorSub1} are just two examples.
+Blows made by whales are examples of processes that can be modelled using distance sampling.
+Songs and blows are indirect sampling methods producing estimates of cue density.
+Cue densities can be converted to animal densities with additional data needed to estimate the rate at which cues are produced and the rate at which they disappear.
+These are but a few examples of the applications to which distance sampling has been applied (a larger, but incomplete list of references is given at \url{http://distancesampling.org/dbib.html}).
+
+The features of \pkg{Distance} are deliberately limited to provide a simplified interface for users.
+For more complex analyses of distance sampling data, we provide additional packages for modelling in \proglang{R}.
+
+We noted at the start of the article that \pkg{Distance} is a simple-to-use wrapper around the package \pkg{mrds}.
+Additional features available in \pkg{mrds} include  models that relax the assumption that detection is certain at zero distance from the transect (by including data from additional observers).
+This is done using mark-recapture type methods which require additional survey methodology, known as double observer surveys \citep[see][for an introduction]{Burt:2014gu}.
+
+\pkg{Distance} can provide us with estimates of abundance or density for each strata as a whole but tells us nothing about the distribution of animals within strata.
+One option is to divide the study area into smaller and smaller strata to try to detect patterns in spatial distribution, however, a more rigorous approach is to build a spatial model.
+Such models incorporate spatially-referenced environmental data (for example derived from GIS products).
+\pkg{Distance} interfaces with one such package used to perform this type of analysis: \pkg{dsm} \citep{dsm-pkg}.
+So-called ``density surface modelling'' uses the generalized additive model framework \citep[e.g.][]{Wood:2006vg} to build models of abundance (adjusting counts for imperfect detectability) as a function of environmental covariates, as part of a two stage model \citep{Hedley:2004et, Miller:2013fq}.
+
+Uncertainty in measured covariates (e.g.
+cluster size) and model uncertainty (when two models have similar fit but substantially different estimates) can be incorporated using the multi-analysis distance sampling package \pkg{mads} \citep{mads-pkg}.
+In addition, \pkg{mads} can also incorporate sightings with unknown species identification.
+This is done by estimating the abundance of these 'unidentified' sightings and pro-rating them to the known species \citep{Gerrodette:2005}.
+
+As mentioned above, survey design is critical to ensuring that resulting distance sampling data can be reliably analysed.
+\pkg{DSsim} allows users to test out different designs in their study region and tailor population attributes to reflect the species they are working with.
+\pkg{DSsim} \citep{DSsim-pkg} allows users to more easily identify challenges unique to their study and select a survey design which is more likely to yield accurate and precise estimates.
+
+Distance for Windows has many users (over 45,000 downloads since 2002) and they may be overwhelmed by the prospect of switching existing analyses to \proglang{R}.
+For that reason we have created the \pkg{readdst} \citep{readdst-pkg} package to interface with projects created by Distance for Windows.
+The package can take analyses created using the CDS, MCDS and MRDS engines in Distance for Windows, extract the data and create equivalent models in \proglang{R}.
+\pkg{readdst} can also run these analyses and test the resulting statistics (for example, $\hat{N}$ or $\hat{P_a}$) calculated in \proglang{R} against those calculated by Distance for Windows.
+We hope that \pkg{readdst} will provide a useful transition to \proglang{R} for interested users.
+\pkg{readdst} is currently available on GitHub at \url{https://github.com/distancedevelopment/readdst}.
+
+\section{Conclusion}
+
+We have given an introduction as to how to perform a distance sampling analysis in \proglang{R}.
+We have covered the possible models for detectability, model checking and selection and finally abundance and variance estimation.
+
+In combination with tools such as \pkg{knitr} and \pkg{rmarkdown} \citep{rmarkdown-pkg}, the helper functions in \pkg{Distance} provide a useful set of tools to perform reproducible analyses of wildlife abundance for both managers and ecologists.
+We also direct readers' attention to \pkg{DsShiny} \citep{dsshiny-pkg}, a package that builds on \pkg{Shiny} \citep{shiny-pkg} and \pkg{mrds} to allow users to build and fit models in a graphical interface, with output to RMarkdown.
+
+\proglang{R} and its extension packages provide many tools exploratory data analysis that can be useful for a distance sampling analysis.
+We hope that this paper provides useful examples for those wishing to pursue distance sampling in \proglang{R}.
+More information on distance sampling can be found at \url{http://distancesampling.org} and a mailing list is maintained at \url{https://groups.google.com/forum/#!forum/distance-sampling}.
+
+We note that there are other packages available for performing distance sampling analyses in \proglang{R} but believe that \pkg{Distance} is the most flexible and feature-complete, and provides pathways to a range of more complicated analyses.
+Appendix A gives a feature comparison between \pkg{Distance} and other \proglang{R} packages for analysis of distance sampling data.
+
+\section{Acknowledgements}
+
+The authors would like to thank the many users of \pkg{Distance}, \pkg{mrds} and Distance for Windows who have contributed bug reports and suggestions for improvements over the years.
+We would particularly like to thank Steve Buckland, David Borchers, Tiago Marques, Jon Bishop and Lorenzo Milazzo for their contributions.
+We thank Tiago Marques a second time for helpful comments on an early version of the paper and Colin Beale who suggested the ``flatfile'' data format.
+We also thank Ken Burnham and David Anderson for fundamental contributions to the early development of these methods.
+
+\bibliography{jstatsoft}
 

### 2016-06-08 paper.Rnw
merge appendix A

@@
 
+\appendix
+
+\section*{Appendix A: Feature comparison}
+
+There are four packages available for analysis of distance sampling data in \proglang{R} that we are aware of.
+All are available on CRAN.
+On top of \pkg{Distance} and \pkg{mrds} \citep{mrds-pkg} described above, they are \pkg{Rdistance} \citep{Rdistance-pkg} and \pkg{unmarked} \citep{Fiske:2011gv}.
+Table \ref{app-comp-tab} provides a feature comparison of these packages.
+
+\begin{table}
+\begin{tabular}{ccccc}
+Feature    &  \pkg{Distance} & \pkg{Rdistance} & \pkg{unmarked} & \pkg{mrds}\\
+\hline
+Line transects & x & x & x & x\\
+Point transets & x & & x & x\\
+Interval (binned) distances & x & & x & x\\
+Exact distances & x & x & & x\\
+Continuous individual level covariates & x &  & & x\\
+Factor individual level covariates & x & x & & x\\
+Transect level covariates & x &  & x & x\\
+Objects in clusters & x & x & x & x\\
+Left truncation & x  & x & & x\\
+Half-normal key & x & x & x & x\\
+Hazard-rate key & x & x &  x & x\\
+Uniform key & x & x & x & x\\
+Gamma key &  & x &  & x\\
+Negative exponential key &  &  & x  &\\
+Adjustment terms & x & x & & x\\
+AIC adjustment selection & x & x & \\
+Monotonicity constraints & x &  & & x\\
+Availability bias model & & & x &\\
+Perception bias model & & & & x\\
+Abundance estimation & x & x & x & x\\
+Density estimation & x & x & x & x\\
+User-defined likelihood functions & & x & & \\
+\end{tabular}
+\caption{Feature comparison of the available packages to perform distance sampling analyses.}
+\label{app-comp-tab}
+\end{table}
+
+\subsection*{Abundance estimation}
+
+\pkg{Distance} and \pkg{mrds} models can be used as part of a density surface model using \pkg{dsm}, which allows abundance to be modelled as a function of spatially varying covariates (such as location, sea depth, altitude etc).
+See \cite{Miller:2013fq} for more information.
+
+\pkg{unmarked} also allows abundance to vary according to covariates, via the abundance part of the likelihood.
+See \cite{Fiske:2011gv} for more information on the package and \cite{Royle:2004cd} for more information on methodology.
+
+Results \pkg{Rdistance} models can be use in combination with \proglang{R} modelling functions such as \code{lm}, \code{glm} etc to build abundance estimates which vary according to covariates.
+More information is available on the project's wiki.
+
 \bibliography{jstatsoft}

### 2016-06-08 paper.Rnw
getting there...

@@
 A different form for the detection function can be specified via the \code{key=} argument to \code{ds}. For example, a hazard rate model can be fitted as:
-<<label=minke-hr-cos>>=
+<<label=minke-hr-cos, message=TRUE>>=
 minke_hrcos <- ds(minke, truncation = 1.5, key = "hr")

### 2016-06-08 paper.Rnw
now in knitr format

@@
 \Plaintitle{Distance Sampling in R} %% without formatting
-%\Shorttitle{\pkg{foo}: A Capitalized Title} %% a short title (if necessary)
 
@@
 }
-%% It is also possible to add a telephone and fax number
-%% before the e-mail in the following format:
-%% Telephone: +43/512/507-7103
-%% Fax: +43/512/507-2851
-
 %% for those who use Sweave please include the following line (with % symbols):
@@
 \section{Introduction}
-<<label=setup, echo=FALSE>>=
+<<label=setup, echo=FALSE, include=FALSE>>=
 options(prompt = 'R> ')
+knitr::opts_knit$set(out.format="sweave", cache=TRUE,prompt = TRUE)
+render_sweave()
 @
@@
 \begin{center}
-<<label=detectdists, fig=TRUE, echo=FALSE, message=FALSE, height=5.5, width=9>>=
+<<label=detectdists, echo=FALSE, message=FALSE, fig.height=5.5, fig.width=9>>=
 
@@
 \begin{center}
-<<label=points-and-lines, fig=TRUE, echo=FALSE, message=FALSE, warning=FALSE, width=7, height=6>>=
+<<label=points-and-lines, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=6>>=
 opar <- par(mfrow=c(2,2))
@@
 \begin{center}
-<<label=hn-hr-par-comp, fig=TRUE, echo=FALSE, height=5, width=10>>=
+<<label=hn-hr-par-comp, echo=FALSE, fig.height=5, fig.width=10>>=
 par(mfrow=c(2,4), mar=c(3.5, 3, 2, 1) + 0.1)
@@
 \begin{center}
-<<label=adjust-mix-comp, fig=TRUE, echo=FALSE, height=3, width=10>>=
+<<label=adjust-mix-comp, echo=FALSE, fig.height=3, fig.width=10>>=
 par(mfrow=c(1,4), mar=c(3.5, 3, 3, 1) + 0.1)
@@
 \begin{center}
-<<label=minke-amakihi-hn-plot, echo=FALSE, warning=FALSE, fig=TRUE, height=4, width=9>>=
+<<label=minke-amakihi-hn-plot, echo=FALSE, warning=FALSE, fig.height=4, fig.width=9>>=
 par(mfrow=c(1, 3))
@@
 \begin{center}
-<<label=amakihiqq, fig=TRUE, echo=FALSE, message=FALSE, height=3, width=6>>=
+<<label=amakihiqq, echo=FALSE, message=FALSE, fig.height=3, fig.width=6>>=
 par(mfrow=c(1,2))
@@
 
-<<label=summary-table, results=tex, echo=FALSE>>=
+<<label=summary-table, results='asis', echo=FALSE>>=
 library("knitr")
@@
 
-<<label=minke-abundance-table, results=tex, echo=FALSE>>=
+<<label=minke-abundance-table, results='asis', echo=FALSE>>=
 minke_table <- summary(minke_hn)$dht$individuals$N

### 2016-06-08 paper.Rnw
okay really nearly there...

@@
 <<label=setup, echo=FALSE, include=FALSE>>=
+library("knitr")
 options(prompt = 'R> ')
-knitr::opts_knit$set(out.format="sweave", cache=TRUE,prompt = TRUE)
+knitr::opts_knit$set(out.format="sweave", prompt=TRUE)
+knitr::opts_chunk$set(cache=TRUE)
 render_sweave()
@@
 We first fit a half-normal model without covariates or adjustments (setting \code{adjustment=NULL} will force \code{ds} to fit a model with no adjustments):
-<<label=amakihi-gof, grdevice=NULL>>=
+<<label=amakihi-gof, fig.keep="none">>=
 amakihi_hn <- ds(amakihi, truncation = 82.5, transect = "point",
@@
 \begin{center}
-<<label=amakihiqq, echo=FALSE, message=FALSE, fig.height=3, fig.width=6>>=
+<<label=amakihiqq, echo=FALSE, message=FALSE, fig.height=3, fig.width=6, results="hide">>=
 par(mfrow=c(1,2))
@@
 <<label=summary-table, results='asis', echo=FALSE>>=
-library("knitr")
 kable(summarize_ds_models(amakihi_hn, amakihi_hr_obs, amakihi_hr_obs_mas),

### 2016-06-08 paper.tex
okay really nearly there...

@@
-\documentclass[article]{jss}
+\documentclass[article]{jss}\usepackage[]{graphicx}\usepackage[]{color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\usepackage{Sweave}
 
@@
 %% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 \begin{center}
-\includegraphics{paper-detectdists}
+\begin{Schunk}
+
+\includegraphics[width=\maxwidth]{figure/detectdists-1} \end{Schunk}
 \caption{Panels show an example detection function (left), the pdf of object distances (middle) and the resulting pdf of detection distances (right) for line transects (top row) and point transects (bottom row).
@@
 \begin{center}
-\includegraphics{paper-points-and-lines}
+\begin{Schunk}
+
+\includegraphics[width=\maxwidth]{figure/points-and-lines-1} \end{Schunk}
 \caption{Left side plots show an example of a survey of a (1x1) area containing a population of 500 objects; blue dashed lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals.
@@
 \begin{center}
-\includegraphics{paper-hn-hr-par-comp}
+\begin{Schunk}
+
+\includegraphics[width=\maxwidth]{figure/hn-hr-par-comp-1} \end{Schunk}
 \caption{Half-normal (top row) and hazard-rate (bottom row) detection functions without adjustments, varying scale ($\sigma$) and (for hazard-rate) shape ($b$) parameters (values are given above the plots).
@@
 \begin{center}
-\includegraphics{paper-adjust-mix-comp}
+\begin{Schunk}
+
+\includegraphics[width=\maxwidth]{figure/adjust-mix-comp-1} \end{Schunk}
 \caption{Possible shapes for the detection function when cosine adjustments are included for half-normal and hazard-rate models.}
@@
 \end{Sinput}
+\begin{Soutput}
+Starting AIC adjustment term selection.
+\end{Soutput}
+\begin{Soutput}
+Fitting half-normal key function
+\end{Soutput}
+\begin{Soutput}
+Key only model: not constraining for monotonicity.
+\end{Soutput}
+\begin{Soutput}
+AIC= 46.872
+\end{Soutput}
+\begin{Soutput}
+Fitting half-normal key function with cosine(2) adjustments
+\end{Soutput}
+\begin{Soutput}
+AIC= 48.872
+\end{Soutput}
+\begin{Soutput}
+
+Half-normal key function selected.
+\end{Soutput}
 \end{Schunk}
@@
 \end{Sinput}
+\begin{Soutput}
+Starting AIC adjustment term selection.
+\end{Soutput}
+\begin{Soutput}
+Fitting hazard-rate key function
+\end{Soutput}
+\begin{Soutput}
+Key only model: not constraining for monotonicity.
+\end{Soutput}
+\begin{Soutput}
+AIC= 48.637
+\end{Soutput}
+\begin{Soutput}
+Fitting hazard-rate key function with cosine(2) adjustments
+\end{Soutput}
+\begin{Soutput}
+AIC= 50.386
+\end{Soutput}
+\begin{Soutput}
+
+Hazard-rate key function selected.
+\end{Soutput}
 \end{Schunk}
@@
 \end{Sinput}
+\begin{Soutput}
+Fitting uniform key function with cosine(1,2) adjustments
+\end{Soutput}
+\begin{Soutput}
+AIC= 48.268
+\end{Soutput}
 \end{Schunk}
@@
 \end{Sinput}
+\begin{Soutput}
+Model contains covariate term(s): no adjustment terms will be included.
+\end{Soutput}
+\begin{Soutput}
+Fitting hazard-rate key function
+\end{Soutput}
+\begin{Soutput}
+AIC= 10778.448
+\end{Soutput}
+\begin{Soutput}
+No survey area information supplied, only estimating detection function.
+\end{Soutput}
 \end{Schunk}
@@
 \end{Sinput}
+\begin{Soutput}
+Model contains covariate term(s): no adjustment terms will be included.
+\end{Soutput}
+\begin{Soutput}
+Fitting hazard-rate key function
+\end{Soutput}
+\begin{Soutput}
+AIC= 10777.376
+\end{Soutput}
+\begin{Soutput}
+No survey area information supplied, only estimating detection function.
+\end{Soutput}
 \end{Schunk}
@@
 \begin{center}
-\includegraphics{paper-minke-amakihi-hn-plot}
+\begin{Schunk}
+
+\includegraphics[width=\maxwidth]{figure/minke-amakihi-hn-plot-1} \end{Schunk}
 \caption{Left: fitted detection function overlayed on the histogram of observed distances for the minke whale data using half-normal model.
@@
 \begin{Soutput}
+
 Summary for distance analysis 
@@
 +                  key = "hn", adjustment = NULL)
+\end{Sinput}
+\begin{Soutput}
+Fitting half-normal key function
+\end{Soutput}
+\begin{Soutput}
+Key only model: not constraining for monotonicity.
+\end{Soutput}
+\begin{Soutput}
+AIC= 10833.841
+\end{Soutput}
+\begin{Soutput}
+No survey area information supplied, only estimating detection function.
+\end{Soutput}
+\begin{Sinput}
 R> gof_ds(amakihi_hn)
@@
 \begin{Soutput}
+
 Goodness of fit results for ddf object
@@
 \begin{Schunk}
-\begin{Soutput}
-Goodness of fit results for ddf object
 
-Distance sampling Kolmogorov-Smirnov test
-Test statistic =  0.059345  P =  0.00031527
-
-Distance sampling Cramer-von Mises test (unweighted)
-Test statistic =  0.93083  P =  0.003578 
-\end{Soutput}
-\begin{Soutput}
-Goodness of fit results for ddf object
-
-Distance sampling Kolmogorov-Smirnov test
-Test statistic =  0.036251  P =  0.076237
-
-Distance sampling Cramer-von Mises test (unweighted)
-Test statistic =  0.15016  P =  0.38908 
-\end{Soutput}
-\end{Schunk}
-\includegraphics{paper-amakihiqq}
+\includegraphics[width=\maxwidth]{figure/amakihiqq-1} \end{Schunk}
 \caption{Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates; left) and hazard-rate model with observer and minutes after sunrise (right) for the amakihi data.}
@@
 \end{table}
+
 \section{Estimating abundance and variance}
@@
 \end{table}
+
 \section{Extensions}

### 2016-06-08 paper.Rnw
table captions below

@@
 \usepackage{booktabs}
+
+% make captions be at the bottom
+\usepackage{float}
+\floatstyle{plain}
+\restylefloat{table}
 
@@
 \begin{center}
-<<label=amakihiqq, echo=FALSE, message=FALSE, fig.height=3, fig.width=6, results="hide">>=
+<<label=amakihiqq, echo=FALSE, message=FALSE, fig.height=4, fig.width=7, results="hide">>=
 par(mfrow=c(1,2))
@@
 
-\begin{table}
+\begin{table}[h]
 \begin{tabular}{ccccc}

### 2016-06-29 paper.Rnw
missing word in abstract
"as well documenting" -> "as well as documenting"

@@
 We introduce the \proglang{R} package \pkg{Distance} that implements distance sampling methods to estimate abundance.
-We describe how users can obtain estimates of abundance (and density) using the package as well documenting the links it provides with other more specialized \proglang{R} packages.
+We describe how users can obtain estimates of abundance (and density) using the package as well as documenting the links it provides with other more specialized \proglang{R} packages.
 We also demonstrate how \pkg{Distance} provides a migration pathway from previous software, thereby allowing us to deliver cutting-edge methods to the users more quickly.

### 2016-06-29 paper.Rnw
sum series -> some series

@@
 $$
-where $k$ is the key function and $\alpha_O$ is sum series of functions (given in Table \ref{tab:keyadj}), described as an \textit{adjustment of order $O$}.
+where $k$ is the key function and $\alpha_O$ is some series of functions (given in Table \ref{tab:keyadj}), described as an \textit{adjustment of order $O$}.
 Subscripts on the parameter vector indicate those parameters belonging to each part of the model (i.e., $\boldsymbol{\theta} = (\boldsymbol{\theta}_\text{key}, \boldsymbol{\theta}_\text{adjust})$).

### 2016-06-29 paper.Rnw
missing hat on P_a just before stratification subsubsect

@@
 $$
-where $\pi(y)$ describes the distribution of objects with respect to the sampler; $\pi(x)=1/w$ for line transects and $\pi(r)=\frac{2r}{w^2}$ for point transects, taking into account the geometry of the sampler \citep[usually referred to as the \textit{probability density function of distances}][Chapter 3]{Buckland:2001vm}, .
+where $\pi(y)$ describes the distribution of objects with respect to the sampler; $\pi(x)=1/w$ for line transects and $\pi(r)=\frac{2r}{w^2}$ for point transects, taking into account the geometry of the sampler \citep[usually referred to as the \textit{probability density function of distances}][Chapter 3]{Buckland:2001vm}.
 When considering a particular transect type we let $x$ denote a perpendicular distance from a line and $r$ denote radial distance from a point (rather than using $y$).
@@
 It can also be compared to similar estimates in mark-recapture studies.
-$P_a$ is included in the \code{summary} output and the table produced by \code{summarize_ds_models}.
+$\hat{P_a}$ is included in the \code{summary} output and the table produced by \code{summarize_ds_models}.
 

### 2017-09-04 paper.Rnw
reviewer comment alterations

--- 2016-06-29 paper.Rnw
@@
 %% almost as usual
-\author{David L Miller\\Integrated Statistics and University of St Andrews \AND 
+\author{David L Miller\\University of St Andrews \AND 
         Eric Rexstad\\University of St Andrews \And 
@@
 We introduce the \proglang{R} package \pkg{Distance} that implements distance sampling methods to estimate abundance.
-We describe how users can obtain estimates of abundance (and density) using the package as well as documenting the links it provides with other more specialized \proglang{R} packages.
+We describe how users can obtain estimates of abundance (and density) using the package as well documenting the links it provides with other more specialized \proglang{R} packages.
 We also demonstrate how \pkg{Distance} provides a migration pathway from previous software, thereby allowing us to deliver cutting-edge methods to the users more quickly.
@@
   David L Miller\\
-  Integrated Statistics and University of St Andrews\\
-  16 Sumner Street, Woods Hole, MA 02543, USA and\\
+  University of St Andrews\\
   Centre for Research into Ecological and Environmental Modelling,\\
@@
 \begin{center}
-<<label=detectdists, echo=FALSE, message=FALSE, fig.height=5.5, fig.width=9>>=
-
+<<label=detectdists, echo=FALSE, message=FALSE, fig.height=5.5, fig.width=8>>=
 par(mfrow=c(2,3))
@@
 @
-\caption{Left side plots show an example of a survey of a (1x1) area containing a population of 500 objects; blue dashed lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals.
-The right side of the figure shows histograms of observed distances (again, lines top and points bottom).}
+\caption{Examples of line (top row) and point (bottom row) transect sampling.
+Left side plots show an example of a survey of the unit square containing a population of 500 objects; blue dashed lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals.
+The right side of the figure shows histograms of observed distances.}
 \label{fig:pointslines}
@@
 
-Having estimated the detection function's parameters, one can then integrate out distance from the function (as the detection function describes the probability of detection \textit{given} distance) to get an ``average'' probability of detection, which can be used to correct the observed counts.
+Having estimated the detection function's parameters, one can then integrate out distance from the function (as the detection function describes the probability of detection \textit{given} distance) to get an ``average'' probability of detection (in the sense of averaging over distances), which can be used to correct the observed counts.
 Summing the corrected counts gives an estimate of abundance in the area covered by surveys, which can be multiplied up to the total study area.
@@
 
-The detection function models the probability $\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)$ and is usually denoted $g(y; \boldsymbol{\theta})$ where $y$ is distance (from a line or point) and $\boldsymbol{\theta}$ is a vector of parameters to be estimated.
-Our goal is to estimate an \textit{average probability of detection} ($p$, average in the sense of an average over distance from $0$ to truncation distance $w$), so we must integrate out distance ($y$) from the detection function:
+The detection function models the probability $\Prob(\text{object detected } \vert \text{ object at distance } y)$ and is usually denoted $g(y; \boldsymbol{\theta})$ where $y$ is distance (from a line or point) and $\boldsymbol{\theta}$ is a vector of parameters to be estimated.
+The parameter(s) $\boldsymbol{\theta}$ are used in a detection function (see Section 3.1), serving as the basis to produce an estimate of the probability that an object within the truncation distance of the point or line transect is detected during the survey.
+Our goal is to estimate an \textit{average probability of detection} ($p$, average in the sense of an average over distance from $0$ to truncation distance $w$), so we must integrate out distance from the detection function.
+Letting $x$ denote a perpendicular distance from a line and $r$ denote radial distance from a point:
 $$
-p = \int_0^w \pi(y) g(y; \boldsymbol{\theta}) dy
+p = \left\{
+\begin{array}{l l}
+\int_0^w \frac{1}{w}    g(x; \boldsymbol{\theta}) dx & \quad \text{for lines,} \\
+\int_0^w \frac{2r}{w^2} g(r; \boldsymbol{\theta}) dr & \quad \text{for points,}
+\end{array} \right.
 $$
-where $\pi(y)$ describes the distribution of objects with respect to the sampler; $\pi(x)=1/w$ for line transects and $\pi(r)=\frac{2r}{w^2}$ for point transects, taking into account the geometry of the sampler \citep[usually referred to as the \textit{probability density function of distances}][Chapter 3]{Buckland:2001vm}, .
-When considering a particular transect type we let $x$ denote a perpendicular distance from a line and $r$ denote radial distance from a point (rather than using $y$).
+where the fractions pre-multiplying the detection function describes the distribution of objects with respect to the sampler, taking into account the geometry of the sampler \citep[usually referred to as the \textit{probability density function of (object) distances} and denoted $\pi(y)$;][Chapter 3]{Buckland:2001vm}.
+Figure \ref{fig:detectdists} shows the relationship between the detection function, the probability density function of object distances and the probability density function of observed distances.
 
@@
 It is not always necessary to include adjustments (except in the case of the uniform key) and in such cases we refer to these as ``key only'' models (see the next section and ``Model checking and model selection'').
+Adjustment terms increase the flexibility of the detection functioni but this added flexibility comes at the expense of additional parameters to be estimated.
+The analyst must consider whether the process giving rise to the distribution of detection distances should be modelled with shapes depicted in Figure \ref{fig:keyadj}.
 
@@
 f <- function(x) g(x)/g(0)
-curve(f, from=0, to=0.025, xlab="", ylab="", main="Hazard-rate with \n2 cosine adjustment", xlim=c(0,0.025), ylim=c(0,1))
+curve(f, from=0, to=0.025, xlab="", ylab="", main="Hazard-rate with \n2 cosine adjustments", xlim=c(0,0.025), ylim=c(0,1))
 title(xlab="Distance", line=2)
@@
 There are many factors that can affect the probability of detecting an object: observer skill, cluster size (if objects occur in clusters), the vessel or platform used, sea state, other weather conditions, time of day and more.
-In \pkg{Distance} we assume that these variables affect detection only via the scale of the detection function (and do not affect the shape).
+In \pkg{Distance} we assume that these covariates affect detection only via the scale of the detection function (and do not affect the shape).
 
@@
 $$
-p(\mathbf{z_i}) = \int_0^w \pi(y) g(y, \mathbf{z_i}; \boldsymbol{\theta}) dy,
+p(\mathbf{z}_i) = \left\{
+\begin{array}{l l}
+\int_0^w \frac{1}{w} g(x, \mathbf{z}_i; \boldsymbol{\theta}) dx & \quad \text{for lines,} \\
+\int_0^w \frac{2r}{w^2} g(r, \mathbf{z}_i; \boldsymbol{\theta}) dr & \quad \text{for points,}
+\end{array} \right.
 $$
-where $\mathbf{z_i}$ is the vector of $J$ covariates associated with observation $i$.
+where $\mathbf{z}_i$ is the vector of $J$ covariates associated with observation $i$.
 For covariate models, we calculate a value of ``average'' probability of detection (average in the sense of distance being integrated out) per observation.
-There are as many unique values of $p(\mathbf{z_i})$ as there are unique covariate combinations in our data.
+There are as many unique values of $p(\mathbf{z}_i)$ as there are unique covariate combinations in our data.
 
@@
 par(mfrow=c(1, 3))
-plot(minke_hn, showpoints=FALSE)
+plot(minke_hn, showpoints=FALSE, xlab="Distance (km)")
 title("Minke whales", cex.main=1.5)
-plot(amakihi_hr_obs, pdf=TRUE, cex=0.7)
+plot(amakihi_hr_obs, pdf=TRUE, cex=0.7, xlab="Distance (m)")
 title("Amakihi\n(observer)", cex.main=1.5)
-plot(amakihi_hr_obs_mas, pdf=TRUE, cex=0.7)
+plot(amakihi_hr_obs_mas, pdf=TRUE, cex=0.7, xlab="Distance (m)")
 title("Amakihi\n(observer+minutes)", cex.main=1.5)
@@
 Centre, hazard-rate with observer as a covariate; right, hazard-rate model with observer and minutes after sunrise as covariates.
-Points indicate probability of detection for a given observation (given that observation's covariate values) and lines indicate the average detection function.}
+Points indicate probability of detection for a given observation (given that observation's covariate values) and lines indicate the average detection function (averaged over covariates, observer or observer+minutes after sunrise).}
 \label{fig:minkeamakihi}
@@
 Stratification can be used to increase the precision of estimates if we know \textit{a priori} that density varies between different parts of the study area.
-For example, strata may be defined by habitat types or animal sex (or some combination) which may be of interest for biological or management reasons.
-To calculate estimates for a given stratification each observation must occur in a stratum which must be labelled with a \code{Region.Label} and have a corresponding \code{Area} (if we are using an animal characteristic like sex, we would have the areas be the same but if we were using say forested vs.
-wetland habitat the areas of those strata would be different).
+For example, strata may be defined by habitat types which may be of interest for biological or management reasons.
+To calculate estimates for a given stratification each observation must occur in a stratum which must be labelled with a \code{Region.Label} and have a corresponding \code{Area}.
 Finally, we must also know the stratum in which each observation occurs.
@@
 $$
-\widehat{\text{Var}}_\text{model}\left( \hat{N}_\text{C}\right) = \left(\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}\right)^\top \left(\hat{\mathbf{H}}(\hat{\boldsymbol{\theta}})^{-1} \right)\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}
+\widehat{\text{Var}}_\text{model}\left( \hat{N}_\text{C}\right) = \left(\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}\right)^\top \left(\hat{\mathbf{H}}(\hat{\boldsymbol{\theta}})^{-1} \right)\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}},
 $$
 where the partial derivatives of $\hat{N}_\text{C}$ are evaluated at the MLE ($\hat{\theta}$) and $\mathbf{H}$ is the first partial Hessian (outer product of first derivatives of the log likelihood) for numerical stability \citep[][p 62]{Buckland:2001vm}.
-Note that although we calculate uncertainty in $\hat{N}_\text{C}$, we can trivially scale-up to variance of $\hat{N}$ (by noting that $\hat{N} = \frac{A}{a} \hat{N}_\text{C}$ and therefore $\widehat{\text{Var}}_\text{model}\left(\hat{N}\right) = \left( \frac{A}{a} \right)^2 \widehat{\text{Var}}_\text{model} \left(\hat{N}_\text{C} \right)$).
+Note that although we calculate uncertainty in $\hat{N}_\text{C}$, we can scale-up to variance of $\hat{N}$ (by noting that $\hat{N} = \frac{A}{a} \hat{N}_\text{C}$ and therefore $\widehat{\text{Var}}_\text{model}\left(\hat{N}\right) = \left( \frac{A}{a} \right)^2 \widehat{\text{Var}}_\text{model} \left(\hat{N}_\text{C} \right)$).
 
@@
 
-Results \pkg{Rdistance} models can be use in combination with \proglang{R} modelling functions such as \code{lm}, \code{glm} etc to build abundance estimates which vary according to covariates.
+Resulting \pkg{Rdistance} models can be use in combination with \proglang{R} modelling functions such as \code{lm}, \code{glm} etc to build abundance estimates which vary according to covariates.
 More information is available on the project's wiki.

### 2017-09-04 paper.tex
table captions below
reviewer comment alterations

@@
 
+% make captions be at the bottom
+\usepackage{float}
+\floatstyle{plain}
+\restylefloat{table}
+
 %% almost as usual
-\author{David L Miller\\Integrated Statistics and University of St Andrews \AND 
+\author{David L Miller\\University of St Andrews \AND 
         Eric Rexstad\\University of St Andrews \And 
@@
 \includegraphics[width=\maxwidth]{figure/points-and-lines-1} \end{Schunk}
-\caption{Left side plots show an example of a survey of a (1x1) area containing a population of 500 objects; blue dashed lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals.
-The right side of the figure shows histograms of observed distances (again, lines top and points bottom).}
+\caption{Examples of line (top row) and point (bottom row) transect sampling.
+Left side plots show an example of a survey of the unit square containing a population of 500 objects; blue dashed lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals.
+The right side of the figure shows histograms of observed distances.}
 \label{fig:pointslines}
@@
 
-The detection function models the probability $\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)$ and is usually denoted $g(y; \boldsymbol{\theta})$ where $y$ is distance (from a line or point) and $\boldsymbol{\theta}$ is a vector of parameters to be estimated.
-Our goal is to estimate an \textit{average probability of detection} ($p$, average in the sense of an average over distance from $0$ to truncation distance $w$), so we must integrate out distance ($y$) from the detection function:
-$$
-p = \int_0^w \pi(y) g(y; \boldsymbol{\theta}) dy
-$$
-where $\pi(y)$ describes the distribution of objects with respect to the sampler; $\pi(x)=1/w$ for line transects and $\pi(r)=\frac{2r}{w^2}$ for point transects, taking into account the geometry of the sampler \citep[usually referred to as the \textit{probability density function of distances}][Chapter 3]{Buckland:2001vm}, .
-When considering a particular transect type we let $x$ denote a perpendicular distance from a line and $r$ denote radial distance from a point (rather than using $y$).
+The detection function models the probability $\Prob(\text{object detected } \vert \text{ object at distance } y)$ and is usually denoted $g(y; \boldsymbol{\theta})$ where $y$ is distance (from a line or point) and $\boldsymbol{\theta}$ is a vector of parameters to be estimated.
+The parameter(s) $\boldsymbol{\theta}$ are used in a detection function (see Section 3.1), serving as the basis to produce an estimate of the probability that an object within the truncation distance of the point or line transect is detected during the survey.
+Our goal is to estimate an \textit{average probability of detection} ($p$, average in the sense of an average over distance from $0$ to truncation distance $w$), so we must integrate out distance from the detection function.
+Letting $x$ denote a perpendicular distance from a line and $r$ denote radial distance from a point:
+$$
+p = \left\{
+\begin{array}{l l}
+\int_0^w \frac{1}{w}    g(x; \boldsymbol{\theta}) dx & \quad \text{for lines,} \\
+\int_0^w \frac{2r}{w^2} g(r; \boldsymbol{\theta}) dr & \quad \text{for points,}
+\end{array} \right.
+$$
+where the fractions pre-multiplying the detection function describes the distribution of objects with respect to the sampler, taking into account the geometry of the sampler \citep[usually referred to as the \textit{probability density function of (object) distances} and denoted $\pi(y)$;][Chapter 3]{Buckland:2001vm}.
+Figure \ref{fig:detectdists} shows the relationship between the detection function, the probability density function of object distances and the probability density function of observed distances.
 
@@
 $$
-p(\mathbf{z_i}) = \int_0^w \pi(y) g(y, \mathbf{z_i}; \boldsymbol{\theta}) dy,
-$$
-where $\mathbf{z_i}$ is the vector of $J$ covariates associated with observation $i$.
+p(\mathbf{z}_i) = \left\{
+\begin{array}{l l}
+\int_0^w \frac{1}{w} g(x, \mathbf{z}_i; \boldsymbol{\theta}) dx & \quad \text{for lines,} \\
+\int_0^w \frac{2r}{w^2} g(r, \mathbf{z}_i; \boldsymbol{\theta}) dr & \quad \text{for points,}
+\end{array} \right.
+$$
+where $\mathbf{z}_i$ is the vector of $J$ covariates associated with observation $i$.
 For covariate models, we calculate a value of ``average'' probability of detection (average in the sense of distance being integrated out) per observation.
-There are as many unique values of $p(\mathbf{z_i})$ as there are unique covariate combinations in our data.
+There are as many unique values of $p(\mathbf{z}_i)$ as there are unique covariate combinations in our data.
 

### 2017-11-02 paper.Rnw
omg, FINALLY

@@
 gof_ds(amakihi_hn)
-gof_ds(amakihi_hr_obs_mas)
+gof_ds(amakihi_hr_obs)
 @
@@
 par(mfrow=c(1,2))
-gof_ds(amakihi_hn, asp=1, cex=0.3)  # shrinking dot size to obliterate 45-deg line less
-gof_ds(amakihi_hr_obs_mas, asp=1, cex=0.3)
+gof_ds(amakihi_hn, cex=0.3)  # shrinking dot size to obliterate 45-deg line less
+gof_ds(amakihi_hr_obs, cex=0.3)
 @
-\caption{Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates; left) and hazard-rate model with observer and minutes after sunrise (right) for the amakihi data.}
+\caption{Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates; left) and hazard-rate model with observer as a covariate (right) for the amakihi data.}
 \label{amakihi-qq}

### 2017-11-02 paper.tex
omg, FINALLY

@@
 Distance sampling Kolmogorov-Smirnov test
-Test statistic =  0.059345  P =  0.00031527
+Test statistic = 0.0593449 p-value = 0
 
 Distance sampling Cramer-von Mises test (unweighted)
-Test statistic =  0.93083  P =  0.003578 
-\end{Soutput}
-\begin{Sinput}
-R> gof_ds(amakihi_hr_obs_mas)
+Test statistic = 0.930827 p-value = 0.00357797
+\end{Soutput}
+\begin{Sinput}
+R> gof_ds(amakihi_hr_obs)
 \end{Sinput}
@@
 Distance sampling Kolmogorov-Smirnov test
-Test statistic =  0.036251  P =  0.076237
+Test statistic = 0.0420268 p-value = 0.48
 
 Distance sampling Cramer-von Mises test (unweighted)
-Test statistic =  0.15016  P =  0.38908 
+Test statistic = 0.198335 p-value = 0.270719
 \end{Soutput}
@@
 \includegraphics[width=\maxwidth]{figure/amakihiqq-1} \end{Schunk}
-\caption{Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates; left) and hazard-rate model with observer and minutes after sunrise (right) for the amakihi data.}
+\caption{Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates; left) and hazard-rate model with observer as a covariate (right) for the amakihi data.}
 \label{amakihi-qq}
@@
 
-\caption{Summary for the detection function models fitted to the amakihi data.
+\caption{\label{tab:summary-table}Summary for the detection function models fitted to the amakihi data.
 ``C-vM'' stands for Cram\'{e}r-von Mises, $P_a$ is average detectability (see ``Estimating abundance and variance''), se is standard error.
@@
 \toprule
-Model & Key function & Formula & C-vM $p$-value & $\hat{P_a}$ & se($\hat{P_a}$) & $\Delta$AIC\\
+Model & Key function & Formula & C-vM p-value & $\hat{P_a}$ & se($\hat{P_a}$) & $\Delta$AIC\\
 \midrule
@@
 
-\caption{Summary of abundance estimation for the half-normal model for the minke whale data.\label{minke-abund}}
+\caption{\label{tab:minke-abundance-table}Summary of abundance estimation for the half-normal model for the minke whale data.\label{minke-abund}}
 \centering

### 2017-11-02 paper-code.R

@@ -1,300 +1,0 @@
-### R code for "Distance Sampling in R"
-### David L Miller, Eric A Rexstad, Len Thomas,
-### Laura Marshall, Jeffrey L Laake
-
-### code chunk number 1: setup prompt for paper
-options(prompt = 'R> ')
-
-
-### code chunk number 2: detectdists
-### this builds figure 1
-par(mfrow=c(2,3))
-
-# detection functions
-hn.detect <- function(x, sigma) exp(-((x^2)/(2*(sigma^2))))
-hn.detect.scaled <- function(x, sigma) hn.detect(x,sigma)/integrate(hn.detect, 0, 500, sigma = sigma)$value
-
-# limits
-x <- seq(0,500, length = 100)
-# evaluate and rescale the detection function
-y <- hn.detect(x, sigma = 200)
-y.scaled <- hn.detect.scaled(x, sigma = 200)
-
-# set plotting options
-set.cex = 1
-set.cex.lab = 1.2
-set.cex.main = 2
-
-# now plot some detection functions and pdfs
-# Line transect example
-plot(x, y, xlab = "Perpendicular distance", ylab = "Probability of detection", type = "l", ylim = c(0,1.1), cex = set.cex, cex.lab = set.cex.lab, axes=FALSE)
-axis(2, at=c(0,1/2,1)); axis(1); box()
-
-w <- 1/500
-line.availability <- rep(w,100)
-plot(x, line.availability, xlab = "Perpendicular distance", ylab = "pdf of object distances", type = "l", ylim = c(0,w*1.1), xlim = c(0,500), cex = set.cex, cex.lab = set.cex.lab, main = "Line Transect", cex.main = set.cex.main, axes=FALSE)
-axis(2, at=c(0,w/2,w)); axis(1); box()
-
-plot(x,y.scaled, xlab = "Perpendicular distance", ylab= "pdf of observed distances", type = "l", ylim = c(0,0.0044), cex = set.cex, cex.lab = set.cex.lab, axes=FALSE)
-axis(2, at=round(c(0,median(y.scaled),max(y.scaled)),3)); axis(1); box()
-
-# Point Transect example
-plot(x, y, xlab = "Radial distance", ylab = "Probability of detection", type = "l", ylim = c(0,1.1), cex = set.cex, cex.lab = set.cex.lab, axes=FALSE)
-axis(2, at=c(0,1/2,1)); axis(1); box()
-
-point.area <- function(x) 2*pi*x
-
[…]

### 2017-11-05 paper.Rnw
EAR comments

@@
 where the fractions pre-multiplying the detection function describes the distribution of objects with respect to the sampler, taking into account the geometry of the sampler \citep[usually referred to as the \textit{probability density function of (object) distances} and denoted $\pi(y)$;][Chapter 3]{Buckland:2001vm}.
-Figure \ref{fig:detectdists} shows the relationship between the detection function, the probability density function of object distances and the probability density function of observed distances.
+Figure \ref{fig:detectiondist} shows the relationship between the detection function, the probability density function of object distances and the probability density function of observed distances.
 
@@
 It is not always necessary to include adjustments (except in the case of the uniform key) and in such cases we refer to these as ``key only'' models (see the next section and ``Model checking and model selection'').
-Adjustment terms increase the flexibility of the detection functioni but this added flexibility comes at the expense of additional parameters to be estimated.
+Adjustment terms increase the flexibility of the detection function but this added flexibility comes at the expense of additional parameters to be estimated.
 The analyst must consider whether the process giving rise to the distribution of detection distances should be modelled with shapes depicted in Figure \ref{fig:keyadj}.
@@
 
-1. \code{Summary statistics}: giving the areas, covered areas, effort, number of observations, number of transects, encounter rate, its standard error and coefficient of variation for each stratum. 
-2. \code{Abundance}: giving estimates, standard errors, coefficients of variation, lower and upper confidence intervals and finally the degrees of freedom for each stratum's abundance estimate.
-3. \code{Density}: lists the same statistics as \code{Abundance} but for a density estimate.
+\begin{enumerate}
+\item \code{Summary statistics}: giving the areas, covered areas, effort, number of observations, number of transects, encounter rate, its standard error and coefficient of variation for each stratum.
+\item \code{Abundance}: giving estimates, standard errors, coefficients of variation, lower and upper confidence intervals and finally the degrees of freedom for each stratum's abundance estimate.
+\item \code{Density}: lists the same statistics as \code{Abundance} but for a density estimate.
+\end{enumerate}
 

### 2017-11-05 make.sh

@@ -2,5 +2,6 @@
 
 Rscript -e "library(knitr);knit('paper.Rnw')"
 pdflatex paper.tex
+bibtex paper
 pdflatex paper.tex
 open paper.pdf

### 2017-11-14 paper.Rnw
remove K-S tests from paper

@@
 Q-Q plots can be inspected visually, though this is prone to subjective judgments.
-Therefore, we also quantify the Q-Q plot's information using a Kolmogorov-Smirnov or Cram\'{e}r-von Mises test \citep{Burnham:2004vd}.
-Both test whether points from the EDF and CDF are from the same distribution.
-The Kolmogorov-Smirnov uses the test statistic of the largest difference between a point on the Q-Q plot and the line $y=x$, whereas the Cram\'{e}r-von Mises test uses the sum of all the differences.
-As it takes into account more information and is therefore more powerful, the Cram\'{e}r-von Mises is generally preferred.
+Therefore, we also quantify the Q-Q plot's information using a Cram\'{e}r-von Mises test \citep{Burnham:2004vd} to test whether points from the EDF and CDF are from the same distribution.
+The Cram\'{e}r-von Mises test uses the sum of all the distances between a point on the Q-Q plot and the line $y=x$ to form a test statistic.
+As it takes into account more information and is therefore more powerful, the Cram\'{e}r-von Mises is generally more powerful than the Kolmogorov-Smirnov test, which uses the largest difference between a point on the Q-Q plot and the line $y=x$ (which is also available in \pkg{Distance}, though is not produced by default as it requires computationally demanding bootstraps).
 A significant result from either test gives evidence against the null hypothesis (that the data arose from the fitted model), suggesting that the model does not fit the data well.
@@
 
+The authors thank the two anonymous reviewers for their constructive comments, which have greatly improved the paper and package.
 The authors would like to thank the many users of \pkg{Distance}, \pkg{mrds} and Distance for Windows who have contributed bug reports and suggestions for improvements over the years.

### 2017-11-14 paper.tex
remove K-S tests from paper

@@
 
-Distance sampling Kolmogorov-Smirnov test
-Test statistic = 0.0593449 p-value = 0
-
 Distance sampling Cramer-von Mises test (unweighted)
@@
 Goodness of fit results for ddf object
-
-Distance sampling Kolmogorov-Smirnov test
-Test statistic = 0.0420268 p-value = 0.48
 

### 2017-11-27 paper.Rnw
len comments round 2?

@@
 One of the key assumptions of plot sampling is that all objects within each of the plots are counted.
-Distance sampling relaxes this assumption in that observers are no longer required to detect (either by eye, video/audio recording etc) and count everything within selected plots.
+Distance sampling relaxes this assumption in that observers are no longer required to detect (i.e., either by eye, video/audio recording, etc.) and count everything within selected plots.
 While plot sampling techniques are adequate for static populations occurring at high density they are inefficient for more sparsely distributed populations.
@@
 The detection function can be used to infer how many objects were missed and thereby produce estimates of density and/or abundance.
+Exact distances can be recorded or distances can be collected in bins if exact distances are hard to estimate (sometimes referred to as ``grouped'' or ``interval'' data).
+To ensure that the model is not overly influenced by distances far from zero and that observer time is not spent looking for far away objects, we discard or do not record observations beyond a given \textit{truncation distance} (during analysis or while collecting data in the field).
 
@@
 
-Until now those wishing to use our \proglang{R} packages for straight forward distance sampling analyses would have had to negotiate the package \pkg{mrds} \citep{mrds-pkg} designed for mark-recapture distance sampling \citep{Burt:2014gu}, requiring a complex data structure to perform analyses.
+Until now those wishing to use our \proglang{R} packages for straightforward distance sampling analyses would have had to negotiate the package \pkg{mrds} \citep{mrds-pkg} designed for mark-recapture distance sampling \citep{Burt:2014gu}, requiring a complex data structure to perform analyses.
 \pkg{Distance} is a wrapper package around \pkg{mrds} making it easier to get started with basic distance sampling analyses in \proglang{R}.
@@
 @
-\caption{Panels show an example detection function (left), the pdf of object distances (middle) and the resulting pdf of detection distances (right) for line transects (top row) and point transects (bottom row).
+\caption{Panels show an example detection function (left), the probability density function of object distances (middle) and the resulting pdf of detection distances (right) for line transects (top row) and point transects (bottom row).
 The pdf of observed detection distances in the right hand plots are obtained by multiplying the detection function by the pdf of object distances and rescaling.
@@
 Figure \ref{fig:pointslines} shows simulated sampling of a population of 500 objects using line and point transects and their corresponding histograms of observed detection distances.
-Note that for the purposes of distance sampling an ``object'' may either refer to an individual in a population or a cluster of individuals.
+Note that for the purposes of distance sampling an ``object'' may either refer to an individual in a population or a cluster (or group) of individuals.
 
@@
 \pkg{Distance} provides a selection of candidate functions to describe the probability of detection and estimates the associated parameters using maximum likelihood estimation.
-The probability of detecting an object may not only depend on how far it is from the observer but also on other factors such as weather conditions, ground cover, cluster size etc.
+The probability of detecting an object may not only depend on how far it is from the sampler but also on other factors such as weather conditions, ground cover, cluster size etc \citep{Marques:2007ey}.
 The \pkg{Distance} package also allows the incorporation of such covariates into the detection function allowing the detection function scale parameter to vary based on these covariates.
 
-Having estimated the detection function's parameters, one can then integrate out distance from the function (as the detection function describes the probability of detection \textit{given} distance) to get an ``average'' probability of detection (in the sense of averaging over distances), which can be used to correct the observed counts.
+Having estimated the detection function's parameters, one can then integrate out distance from the function (as the detection function describes the probability of detection \textit{given} distance) to get an ``average'' probability of detection (in the sense of averaging over distances, conditional on any observed covariates), which can be used to correct the observed counts.
 Summing the corrected counts gives an estimate of abundance in the area covered by surveys, which can be multiplied up to the total study area.
@@
 These data were collected as part of the International Whaling Commission's International Decade of Cetacean Research Southern Ocean Whale and Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer surveys.
-These data consist of 99 observations that are stratified based on location (near or distant from ice edge) and effort data (transect lengths).
-Further details on the survey are available in \cite{Branch:2001ua} (data is simulated based on the design used for ``1992/93 Area III'' therein).
+Data consist of 99 observations on 25 transects, which were stratified based on location (near or distant from ice edge) and effort data (transect lengths).
+Further details on the survey are available in \cite{Branch:2001ua} (data are simulated based on the design used for ``1992/93 Area III'' therein).
 
@@
 
-It is also possible to use distances collected in intervals (``binned'' or ``grouped'' data, as opposed to ``exact'' distances) to model the detection function.
-In this case the column \code{distance} is replaced by two columns \code{distbegin} and \code{distend} referring to the distance interval start and end cutpoints.
+If distances were collected in bins the column \code{distance} is replaced by two columns \code{distbegin} and \code{distend} referring to the distance interval start and end cutpoints.
 More information on binned data is included in \cite{Buckland:2001vm} sections 4.5 and 7.4.1.2.
@@
 The detection function models the probability $\Prob(\text{object detected } \vert \text{ object at distance } y)$ and is usually denoted $g(y; \boldsymbol{\theta})$ where $y$ is distance (from a line or point) and $\boldsymbol{\theta}$ is a vector of parameters to be estimated.
-The parameter(s) $\boldsymbol{\theta}$ are used in a detection function (see Section 3.1), serving as the basis to produce an estimate of the probability that an object within the truncation distance of the point or line transect is detected during the survey.
 Our goal is to estimate an \textit{average probability of detection} ($p$, average in the sense of an average over distance from $0$ to truncation distance $w$), so we must integrate out distance from the detection function.
@@
 $$
-where the fractions pre-multiplying the detection function describes the distribution of objects with respect to the sampler, taking into account the geometry of the sampler \citep[usually referred to as the \textit{probability density function of (object) distances} and denoted $\pi(y)$;][Chapter 3]{Buckland:2001vm}.
+where the fractions pre-multiplying the detection function describe the distribution of objects with respect to the sampler, taking into account the geometry of the sampler \citep[usually referred to as the \textit{probability density function of (object) distances} and denoted $\pi(y)$;][Chapter 3]{Buckland:2001vm}.
 Figure \ref{fig:detectiondist} shows the relationship between the detection function, the probability density function of object distances and the probability density function of observed distances.
-
-It is crucial that the detection function accurately models detectability at small distances; we are less worried by its behaviour further away from 0.
-To ensure that the model is not overly influenced by distances far from zero, we truncate the distances beyond a given distance $w$ (this can be done during analysis or while collecting data in the field).
 
@@
 \item \textit{Shoulder}: we expect observers to be able to see objects near them, not just those directly in front of them.
-For this reason, we expect the detection function to be flat near the line or point.
+We therefore expect the detection function to be flat near the line or point.
 \item \textit{Non-increasing}: we do not think that observers should be more likely to see distant objects than those nearer the transect.
@@
 
+The shoulder condition also implies that it is crucial that the detection function accurately models detectability at small distances and that we are less worried by its behaviour further away from 0.
 Given these criteria, we can formulate models for $g$.
@@
 
-Key function plus adjustment terms (or adjustment series) models are formulated by taking a ``key'' function and optionally adding ``adjustments'' to it to improve the fit \citep{Buckland:1992fa}.
+Key function plus adjustment terms (or adjustment series) models are formulated by taking a ``key'' function and optionally including ``adjustments'' to it to improve the fit \citep{Buckland:1992fa}.
 Mathematically we formulate this as:
 $$
-g(y; \boldsymbol{\theta}) = k(y; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(y; \boldsymbol{\theta}_\text{adjust})\right),
+g(y; \boldsymbol{\theta}) \propto k(y; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(y; \boldsymbol{\theta}_\text{adjust})\right),
 $$
@@
 
-When adjustment terms are used it is necessary to standardise the results to ensure that $g(0)=1$, so we can redefine the detection function as:
+When adjustment terms are used it is necessary to standardise the results to ensure that $g(0)=1$:
 $$
@@
 As such we cannot set meaningful constraints on monotonicity.
-For this reason, we advise caution when using both adjustments and covariates in a detection function \citep[see][for an example of when this can be problematic]{Miller:2015hw}.
+For this reason, we advise caution when using both adjustments and covariates in a detection function \citep[see][for an example of when this can be problematic and an alternative detection function formulation to solve this issue]{Miller:2015hw}.
 
@@
 We use a quantile-quantile plot (Q-Q plot) to visually assess how well a detection functions fits the data when we have exact distances.
-Here the Q-Q plot compares the cumulative distribution function (CDF) of the fitted detection function to the distribution of the data (empirical distribution function or EDF).
+The Q-Q plot compares the cumulative distribution function (CDF) of the fitted detection function to the distribution of the data (empirical distribution function or EDF).
 The Q-Q plots in \pkg{Distance} plot a point for every observation.
@@
 
-We may wish to calculate abundance estimates for some sub-regions of the study region, we call these areas \textit{strata}.
+We may wish to calculate abundance estimates for some sub-regions of the study region, we call these areas \textit{strata} and can be defined at the design stage or post hoc.
 Stratification can be used to increase the precision of estimates if we know \textit{a priori} that density varies between different parts of the study area.
@@
 Cue densities can be converted to animal densities with additional data needed to estimate the rate at which cues are produced and the rate at which they disappear.
-These are but a few examples of the applications to which distance sampling has been applied (a larger, but incomplete list of references is given at \url{http://distancesampling.org/dbib.html}).
+These are but a few examples of the applications to which distance sampling has been applied (an incomplete list of references is given at \url{http://distancesampling.org/dbib.html}).
 
@@
 Additional features available in \pkg{mrds} include  models that relax the assumption that detection is certain at zero distance from the transect (by including data from additional observers).
-This is done using mark-recapture type methods which require additional survey methodology, known as double observer surveys \citep[see][for an introduction]{Burt:2014gu}.
+This is done using mark-recapture type methods which require additional survey methodology, known as double observer surveys or mark-recapture distance sampling \citep[see][for an introduction]{Burt:2014gu}.
 
@@
 In addition, \pkg{mads} can also incorporate sightings with unknown species identification.
-This is done by estimating the abundance of these 'unidentified' sightings and pro-rating them to the known species \citep{Gerrodette:2005}.
+This is done by estimating the abundance of these unidentified sightings and pro-rating them to the known species \citep{Gerrodette:2005}.
 
@@
 
-Distance for Windows has many users (over 45,000 downloads since 2002) and they may be overwhelmed by the prospect of switching existing analyses to \proglang{R}.
+Distance for Windows has many users (over 50,000 downloads since 2002) and they may be overwhelmed by the prospect of switching existing analyses to \proglang{R}.
 For that reason we have created the \pkg{readdst} \citep{readdst-pkg} package to interface with projects created by Distance for Windows.

### 2017-11-27 paper.tex
len comments round 2?

@@
 \includegraphics[width=\maxwidth]{figure/detectdists-1} \end{Schunk}
-\caption{Panels show an example detection function (left), the pdf of object distances (middle) and the resulting pdf of detection distances (right) for line transects (top row) and point transects (bottom row).
+\caption{Panels show an example detection function (left), the probability density function of object distances (middle) and the resulting pdf of detection distances (right) for line transects (top row) and point transects (bottom row).
 The pdf of observed detection distances in the right hand plots are obtained by multiplying the detection function by the pdf of object distances and rescaling.

### 2017-11-27 paper.Rnw
dave edits

@@
 Conventional distance sampling assumes the observer is located either at a point or moving along a line and will observe all objects that occur at the point or on the line.
-The further away an object is from the point or line (also known as the sampler or transect) the less likely it is that the observer will see it.
+The further away an object is from the point or line (more generally, the sampler or transect) the less likely it is that the observer will see it.
 We can use the distances to each of the detected objects from the line or point to build a model of the probability of detection given distance from the sampler --- the \textit{detection function}.
@@
 
-The rest of the paper has the following structure: we describe data formatting for \pkg{Distance}; candidate detection function models are described in terms of formulation and fitted examples in \proglang{R}.
-We then show how to perform model checking, goodness of fit and model selection.
+The rest of the paper has the following structure: we describe data formatting for \pkg{Distance}; candidate detection function models are described and examples fitted in \proglang{R}.
+We then show how to perform model checking, goodness of fit testing and model selection.
 We go on to show how to estimate abundance, including stratified estimates of abundance.
@@
 \item \textit{Non-increasing}: we do not think that observers should be more likely to see distant objects than those nearer the transect.
-If this occurs, it usually indicates an issue with survey design or field procedures (for example that the distribution of objects with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this \citep{Marques:2010he, Marques:2012fy, Miller:2015hw}.
+If this occurs, it usually indicates an issue with survey design or field procedures (for example that the distribution of objects with respect to the line, $\pi(y)$, is not what we expect), so we do not want the detection function to model this \citep{Marques:2010he, Marques:2012fy, Miller:2015hw}.
 \item \textit{Model robust}: models should be flexible enough to fit many different shapes.
@@
 @
-We can conclude that the half-normal model does not pass our goodness of fit tests and should be discarded.
+We conclude that the half-normal model should be discarded.
 Both hazard-rate models (output only shown for hazard-rate model with observer and minutes after sunrise) had non-significant goodness of fit test statistics and are both therefore deemed plausible models.

### 2017-11-27 paper.tex
dave edits

@@
 \end{Schunk}
-We can conclude that the half-normal model does not pass our goodness of fit tests and should be discarded.
+We conclude that the half-normal model should be discarded.
 Both hazard-rate models (output only shown for hazard-rate model with observer and minutes after sunrise) had non-significant goodness of fit test statistics and are both therefore deemed plausible models.

### 2017-11-28 paper.Rnw
blerg.

@@
 $$
-where the partial derivatives of $\hat{N}_\text{C}$ are evaluated at the MLE ($\hat{\theta}$) and $\mathbf{H}$ is the first partial Hessian (outer product of first derivatives of the log likelihood) for numerical stability \citep[][p 62]{Buckland:2001vm}.
-Note that although we calculate uncertainty in $\hat{N}_\text{C}$, we can scale-up to variance of $\hat{N}$ (by noting that $\hat{N} = \frac{A}{a} \hat{N}_\text{C}$ and therefore $\widehat{\text{Var}}_\text{model}\left(\hat{N}\right) = \left( \frac{A}{a} \right)^2 \widehat{\text{Var}}_\text{model} \left(\hat{N}_\text{C} \right)$).
+where the partial derivatives of $\hat{N}_\text{C}$ are evaluated at the MLE ($\hat{\boldsymbol{\theta}}$) and $\mathbf{H}$ is the first partial Hessian (outer product of first derivatives of the log likelihood) for numerical stability \citep[][p 62]{Buckland:2001vm}.
+Note that although we calculate uncertainty in $\hat{N}_\text{C}$, we can scale-up to variance of $\hat{N}$ (by noting that $\hat{N} = \frac{A}{a} \hat{N}_\text{C}$ and therefore $\widehat{\text{Var}}_\text{model}\left(\hat{N}\right) = \left( \frac{A}{a} \right)^2 \widehat{\text{Var}}_\text{model} \left(\hat{N}_\text{C} \right)$), at least in the case where covariates are independent between strata.
 

### 2017-11-28 paper.Rnw
yikes. resubmitted!

@@
 where the partial derivatives of $\hat{N}_\text{C}$ are evaluated at the MLE ($\hat{\boldsymbol{\theta}}$) and $\mathbf{H}$ is the first partial Hessian (outer product of first derivatives of the log likelihood) for numerical stability \citep[][p 62]{Buckland:2001vm}.
-Note that although we calculate uncertainty in $\hat{N}_\text{C}$, we can scale-up to variance of $\hat{N}$ (by noting that $\hat{N} = \frac{A}{a} \hat{N}_\text{C}$ and therefore $\widehat{\text{Var}}_\text{model}\left(\hat{N}\right) = \left( \frac{A}{a} \right)^2 \widehat{\text{Var}}_\text{model} \left(\hat{N}_\text{C} \right)$), at least in the case where covariates are independent between strata.
+Note that although we calculate uncertainty in $\hat{N}_\text{C}$, we can scale-up to variance of $\hat{N}$ (by noting that $\hat{N} = \frac{A}{a} \hat{N}_\text{C}$ and therefore $\widehat{\text{Var}}_\text{model}\left(\hat{N}\right) = \left( \frac{A}{a} \right)^2 \widehat{\text{Var}}_\text{model} \left(\hat{N}_\text{C} \right)$), at least in the case where covariates are independent between strata  \cite[see e.g.,][]{JPE:JPE12065}.
 

### 2018-01-10 jss2630.Rnw
journal copy editing

--- 2017-11-28 paper.Rnw
@@
 knitr::opts_knit$set(out.format="sweave", prompt=TRUE)
-knitr::opts_chunk$set(cache=TRUE)
+knitr::opts_chunk$set(cache=TRUE, fig.path="Figures/")
 render_sweave()
@@
 
-Uncertainty in measured covariates (e.g.
-cluster size) and model uncertainty (when two models have similar fit but substantially different estimates) can be incorporated using the multi-analysis distance sampling package \pkg{mads} \citep{mads-pkg}.
+Uncertainty in measured covariates (e.g., cluster size) and model uncertainty (when two models have similar fit but substantially different estimates) can be incorporated using the multi-analysis distance sampling package \pkg{mads} \citep{mads-pkg}.
 In addition, \pkg{mads} can also incorporate sightings with unknown species identification.
@@
 
-\bibliography{jstatsoft}
+\bibliography{jss2630}
 

### 2018-01-10 jss2630.tex
journal copy editing

--- 2017-11-28 paper.tex
@@
 
-\includegraphics[width=\maxwidth]{figure/detectdists-1} \end{Schunk}
+\includegraphics[width=\maxwidth]{Figures/detectdists-1} \end{Schunk}
 \caption{Panels show an example detection function (left), the probability density function of object distances (middle) and the resulting pdf of detection distances (right) for line transects (top row) and point transects (bottom row).
@@
 
-\includegraphics[width=\maxwidth]{figure/points-and-lines-1} \end{Schunk}
+\includegraphics[width=\maxwidth]{Figures/points-and-lines-1} \end{Schunk}
 \caption{Examples of line (top row) and point (bottom row) transect sampling.
@@
 
-\includegraphics[width=\maxwidth]{figure/hn-hr-par-comp-1} \end{Schunk}
+\includegraphics[width=\maxwidth]{Figures/hn-hr-par-comp-1} \end{Schunk}
 \caption{Half-normal (top row) and hazard-rate (bottom row) detection functions without adjustments, varying scale ($\sigma$) and (for hazard-rate) shape ($b$) parameters (values are given above the plots).
@@
 
-\includegraphics[width=\maxwidth]{figure/adjust-mix-comp-1} \end{Schunk}
+\includegraphics[width=\maxwidth]{Figures/adjust-mix-comp-1} \end{Schunk}
 \caption{Possible shapes for the detection function when cosine adjustments are included for half-normal and hazard-rate models.}
@@
 
-\includegraphics[width=\maxwidth]{figure/minke-amakihi-hn-plot-1} \end{Schunk}
+\includegraphics[width=\maxwidth]{Figures/minke-amakihi-hn-plot-1} \end{Schunk}
 \caption{Left: fitted detection function overlayed on the histogram of observed distances for the minke whale data using half-normal model.
@@
               estimate         se
-(Intercept) 3.06441705 0.10878121
-obsTJS      0.53017364 0.09956539
-obsTKP      0.08885471 0.18071851
+(Intercept) 3.06441773 0.10878048
+obsTJS      0.53017905 0.09956471
+obsTKP      0.08885528 0.18071707
 
@@
              estimate         se
-(Intercept) 0.8690009 0.06261764
+(Intercept) 0.8690066 0.06261754
 
                        Estimate           SE         CV
-Average p              0.3142723   0.0204413 0.06504326
-N in covered region 3955.1685709 274.2284029 0.06933419
+Average p              0.314273   0.02044127 0.06504304
+N in covered region 3955.160439 274.22700349 0.06933398
 \end{Soutput}
@@
 Distance sampling Cramer-von Mises test (unweighted)
-Test statistic = 0.930827 p-value = 0.00357797
+Test statistic = 0.930828 p-value = 0.00357795
 \end{Soutput}
@@
 Distance sampling Cramer-von Mises test (unweighted)
-Test statistic = 0.198335 p-value = 0.270719
+Test statistic = 0.198333 p-value = 0.270724
 \end{Soutput}
@@
 
-\includegraphics[width=\maxwidth]{figure/amakihiqq-1} \end{Schunk}
+\includegraphics[width=\maxwidth]{Figures/amakihiqq-1} \end{Schunk}
 \caption{Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates; left) and hazard-rate model with observer as a covariate (right) for the amakihi data.}

### 2018-01-10 Code/jss2630.R

@@ -1,0 +1,253 @@
+## ----label=setup, echo=FALSE, include=FALSE------------------------------
+library("knitr")
+options(prompt = 'R> ')
+knitr::opts_knit$set(out.format="sweave", prompt=TRUE)
+knitr::opts_chunk$set(cache=TRUE, fig.path="Figures/")
+render_sweave()
+
+## ----label=detectdists, echo=FALSE, message=FALSE, fig.height=5.5, fig.width=8----
+par(mfrow=c(2,3))
+#line and point transect examples
+
+hn.detect <- function(x, sigma) exp(-((x^2)/(2*(sigma^2))))
+hn.detect.scaled <- function(x, sigma) hn.detect(x,sigma)/integrate(hn.detect, 0, 500, sigma = sigma)$value
+
+x <- seq(0,500, length = 100)
+y <- hn.detect(x, sigma = 200)
+y.scaled <- hn.detect.scaled(x, sigma = 200)
+
+set.cex = 1
+set.cex.lab = 1.2
+set.cex.main = 2
+#Line transect example
+plot(x, y, xlab = "Perpendicular distance", ylab = "Probability of detection", type = "l", ylim = c(0,1.1), cex = set.cex, cex.lab = set.cex.lab, axes=FALSE)
+axis(2, at=c(0,1/2,1)); axis(1); box()
+
+w <- 1/500
+line.availability <- rep(w,100)
+plot(x, line.availability, xlab = "Perpendicular distance", ylab = "pdf of object distances", type = "l", ylim = c(0,w*1.1), xlim = c(0,500), cex = set.cex, cex.lab = set.cex.lab, main = "Line Transect", cex.main = set.cex.main, axes=FALSE)
+axis(2, at=c(0,w/2,w)); axis(1); box()
+
+plot(x,y.scaled, xlab = "Perpendicular distance", ylab= "pdf of observed distances", type = "l", ylim = c(0,0.0044), cex = set.cex, cex.lab = set.cex.lab, axes=FALSE)
+axis(2, at=round(c(0,median(y.scaled),max(y.scaled)),3)); axis(1); box()
+
+#Point Transect example
+plot(x, y, xlab = "Radial distance", ylab = "Probability of detection", type = "l", ylim = c(0,1.1), cex = set.cex, cex.lab = set.cex.lab, axes=FALSE)
+axis(2, at=c(0,1/2,1)); axis(1); box()
+
+point.area <- function(x) 2*pi*x
+
[…]

### 2018-01-10 Distance/R/AIC.dsmodel.R

@@ -1,0 +1,42 @@
+#' Akaike's An Information Criterion for detection functions
+#'
+#' Extract the AIC from a fitted detection function.
+#'
+#' @param object a fitted detection function object
+#' @param k penalty per parameter to be used; the default \code{k = 2} is the "classical" AIC
+#' @param \dots optionally more fitted model objects.
+#' @author David L Miller
+#' @export
+#' @importFrom stats logLik
+#' @examples
+#' \dontrun{
+#' library(Distance)
+#' data(minke)
+#' model <- ds(minke, truncation=4)
+#' model_hr <- ds(minke, truncation=4, key="hr")
+#' # extract the AIC for 2 models
+#' AIC(model, model_hr)
+#' }
+AIC.dsmodel <- function(object, ..., k=2){
+
+  # get the models
+  models <- list(object, ...)
+  models$k <- NULL
+
+  # build the table
+  aics <- matrix(NA, nrow=length(models), ncol=2)
+  for(i in seq_along(models)){
+    ll <- logLik(models[[i]])
+    aics[i, 1] <- attr(ll, "df")
+    aics[i, 2] <- -2*ll + k*attr(ll, "df")
+  }
+  # make it a data.frame
+  aics <- as.data.frame(aics)
+  names(aics) <- c("df", "AIC")
+  # add row names
+  call <- match.call(expand.dots=TRUE)
+  call$k <- NULL
+  rownames(aics) <- as.character(call)[-1]
+
+  return(aics)
+}

### 2018-01-10 Distance/R/Distance-package.R

@@ -1,0 +1,129 @@
+#' Distance sampling
+#'
+#' \code{Distance} is a simple way to fit detection functions and estimate abundance using distance sampling methodology.
+#'
+#' Underlying \code{Distance} is the package \code{mrds}, for more advanced analyses (such as those involving double observer surveys) one may find it necessary to use \code{mrds}.
+#'
+#' Further information on distance sampling methods and example code is available at \url{http://distancesampling.org/R/}.
+#'
+#' For help with distance sampling and this package, there is a Google Group \url{https://groups.google.com/forum/#!forum/distance-sampling}.
+#'
+#' @name Distance-package
+#' @import mrds
+#' @aliases Distance-package Distance
+#' @docType package
+#' @author David L. Miller <dave@@ninepointeightone.net>
+#' @references Laake, J.L. and D.L. Borchers. 2004. Methods for incomplete
+#'   detection at distance zero. In: Advanced Distance Sampling, eds. S.T.
+#'   Buckland, D.R.Anderson, K.P. Burnham, J.L. Laake, D.L. Borchers, and L.
+#'   Thomas. Oxford University Press.
+#'
+#' Marques, F.F.C. and S.T. Buckland. 2004. Covariate models for the detection
+#'   function. In: Advanced Distance Sampling, eds. S.T. Buckland,
+#'   D.R.Anderson, K.P. Burnham, J.L. Laake, D.L. Borchers, and L. Thomas.
+#'   Oxford University Press.
+#'
+#' @keywords Statistical Models
+#'
+NULL
+
+
+#' Simulated minke whale data
+#'
+#' Data simulated from models fitted to 1992/1993 Southern Hemisphere minke whale data collected by the International Whaling Commission. See Branch and Butterworth (2001) for survey details (survey design is shown in figure 1(e)). Data simulated by David Borchers.
+#'
+#' Data are included here as both R data and as an Excel spreadsheet to illustrate the "flat file" input method. See \code{\link{flatfile}} for how to load this data and an example analysis.
+#'
+#' @references Branch, T.A. and D.S. Butterworth (2001) Southern Hemisphere minke whales: standardised abundance estimates from the 1978/79 to 1997/98 IDCR-SOWER surveys. Journal of Cetacean Research and Management 3(2): 143-174
+#'
[…]

### 2018-01-10 Distance/R/checkdata.R

@@ -1,0 +1,136 @@
+#' Check that the data supplied to \code{ds} is correct
+#'
+#' This is an internal function that checks the \code{data.frame}s supplied
+#' to \code{ds} are "correct".
+#'
+#' @param data as in \code{ds}
+#' @param sample.table as in \code{ds}
+#' @param region.table as in \code{ds}
+#' @param obs.table as in \code{ds}
+#' @param formula formula for the covariates
+#'
+#' @return Throws an error if something goes wrong, otherwise returns a
+#'  \code{data.frame}.
+#'
+#' @author David L. Miller
+checkdata<-function(data, region.table=NULL, sample.table=NULL, obs.table=NULL,
+                    formula=~1){
+
+  #Check if the user has passed in a numeric vector
+  if(class(data) != "data.frame" & class(data) != "list"){
+    if(is.numeric(data)){
+      data <- data.frame(distance = data)
+    }else{
+      stop("data is not a data.frame nor are the values numeric")
+    }
+  }
+
+  ## make sure that the data are in the right format first
+  if(is.null(data$distance)){
+    stop("Your data must (at least) have a column called 'distance'!")
+  }
+
+  # make sure that we have a data.frame()
+  data <- as.data.frame(data)
+
+  # first see if the data has detected/observer/object columns, if not add them
+  if(!any(names(data)=="observer")){
+    data<-cbind(data,observer=rep(1,nrow(data)))
+  }
+  if(!any(names(data)=="detected")){
+    data<-cbind(data,detected=rep(1,nrow(data)))
+  }
+  if(!any(names(data)=="object")){
+    data<-cbind(data,object=1:nrow(data))
+  }else{
+    # check that the object IDs are unique
+    # first need to remove the rows with NA distances used for padding
+    # below
+    data_no_NA <- data[!is.na(data$distance),]
+    if(length(data_no_NA$object)!=length(unique(data_no_NA$object))){
+      stop("Not all object IDs are unique, check data.")
+    }
+  }
+
+  ## check that the covariates that are specified exist in the data
+  formula <- as.formula(formula)
+  if(formula!=~1){
+    vars_in_data <- all.vars(formula) %in% names(data)
+    if(!all(vars_in_data)){
+      stop(paste("Variable(s):", paste(all.vars(formula)[!vars_in_data],collapse=", "),
+                 "are in the model formula but not in the data."))
+    }
+  }
+
+  if(is.null(region.table) & is.null(sample.table) & is.null(obs.table)){
[…]

### 2018-01-10 Distance/R/create.bins.R

@@ -1,0 +1,72 @@
+#' Create bins from a set of binned distances and a set of cutpoints.
+#'
+#' This is an internal routine and shouldn't be necessary in normal analyses.
+#'
+#' @param data \code{data.frame} with at least the column \code{distance}.
+#' @param cutpoints vector of cutpoints for the bins
+#'
+#' @return data \code{data} with two extra columns \code{distbegin} and
+#'        \code{distend}.
+#'
+#' @author David L. Miller
+#' @export
+#' @examples
+#' \dontrun{
+#' library(Distance)
+#' data(minke)
+#'
+#' # put the minke data into bins 0-1, 1-2, 2-3 km
+#' minke_cuts <- create.bins(minke[!is.na(minke$distance),], c(0,1,2,3))
+#' }
+create.bins <- function(data, cutpoints){
+
+  # don't do anything if there are NAs in the distance column
+  if(any(is.na(data$distance))){
+    stop("Some distances are NA, can't create bins.")
+  }
+
+  # lazy typist
+  cp <- cutpoints
+
+  # remove distances outside bins
+  in.cp.ind <- data$distance>=cp[1] & data$distance<=cp[length(cp)]
+  if(!all(in.cp.ind)){
+    warning("Some distances were outside bins and have been removed.")
+  }
+  data <- data[in.cp.ind, , drop=FALSE]
+
+  # pull out the distances (removing the NAs for now)
+  na.ind <- is.na(data$distance)
+  d <- data$distance[!na.ind]
+
+  distbegin<-rep(NA,length(d))
+  distend<-rep(NA,length(d))
+
+  for(i in 1:(length(cp)-1)){
+    # which elements of d lie between cutpoints i and i+1
+    ind <- which(d>=cp[i] & d<cp[i+1])
+
+    distbegin[ind] <- cp[i]
+    distend[ind]   <- cp[i+1]
+  }
+  # last cutpoint, include those observations AT the truncation point
+  ind <- which(d>=cp[i] & d<=cp[i+1])
+
+  distbegin[ind] <- cp[i]
+  distend[ind]   <- cp[i+1]
+
+
+  # handle NA distances, that we need to preserve
+  distbegin.na <- rep(NA,length(data$distance))
+  distend.na <- rep(NA,length(data$distance))
+  distbegin.na[!na.ind] <- distbegin
+  distend.na[!na.ind] <- distend
+
+  # put all that together and make a data.frame
+  data <- cbind(data,
[…]

### 2018-01-10 Distance/R/ds.R

@@ -1,0 +1,645 @@
+#' Fit detection functions and calculate abundance from line or point transect data
+#'
+#' This function fits detection functions to line or point transect data and then (provided that survey information is supplied) calculates abundance and density estimates. The examples below illustrate some basic types of analysis using \code{ds()}.
+#'
+#' @param data a \code{data.frame} containing at least a column called
+#'        \code{distance} or a numeric vector containing the distances. 
+#'        NOTE! If there is a column called \code{size} in
+#'        the data then it will be interpreted as group/cluster size, see the
+#'        section "Clusters/groups", below. One can supply data as a "flat file"
+#'        and not supply \code{region.table}, \code{sample.table} and
+#'        \code{obs.table}, see "Data format", below and \code{\link{flatfile}}.
+#' @param truncation either truncation distance (numeric, e.g. 5) or percentage (as a string, e.g. "15\%"). Can be supplied as a \code{list} with elements \code{left} and \code{right} if left truncation is required (e.g. \code{list(left=1,right=20)} or \code{list(left="1\%",right="15\%")} or even \code{list(left="1",right="15\%")}).
+#' By default for exact distances the maximum observed distance is used as the right truncation. When the data is binned, the right truncation is the largest bin end point. Default left truncation is set to zero.
+#' @param transect indicates transect type "line" (default) or "point".
+#' @param formula formula for the scale parameter. For a CDS analysis leave this as its default \code{~1}.
+#' @param key key function to use; "hn" gives half-normal (default), "hr" gives hazard-rate and "unif" gives uniform. Note that if uniform key is used, covariates cannot be included in the model.
+#' @param adjustment adjustment terms to use; \code{"cos"} gives cosine (default), \code{"herm"} gives Hermite polynomial and \code{"poly"} gives simple polynomial. \code{"cos"} is recommended. A value of \code{NULL} indicates that no adjustments are to be fitted.
[…]

### 2018-01-10 Distance/R/ds.gof.R

@@ -1,0 +1,28 @@
+#' Goodness of fit tests for distance sampling models
+#'
+#' Chi-square, Kolmogorov-Smirnov (if \code{ks=TRUE}) and Cramer-von Mises goodness of fit tests for detection function models.
+#'
+#' @export
+#' @param model fitted model object
+#' @param breaks Cutpoints to use for binning data
+#' @param nc Number of distance classes
+#' @param qq Flag to indicate whether quantile-quantile plot is desired
+#' @param ks perform the Kolmogorov-Smirnov test (this involves many bootstraps so can take a while)
+#' @param \dots Graphics parameters to pass into qqplot function
+#' @return List of test results and a plot.
+#' @author David L Miller
+#' @seealso qqplot.ddf ddf.gof
+#' @keywords utility
+#' @examples
+#' \dontrun{
+#' # fit and test a simple model for the golf tee data
+#' library(Distance)
+#' data(book.tee.data)
+#' tee.data<-book.tee.data$book.tee.dataframe[book.tee.data$book.tee.dataframe$observer==1,]
+#' ds.model <- ds(tee.data,4)
+#' ds.gof(ds.model)
+#'}
+ds.gof <- function(model, breaks=NULL, nc=NULL, qq=TRUE, ks=FALSE, ...){
+  return(suppressMesages(ddf.gof(model$ddf, breaks=breaks, nc=nc, qq=qq, ks=ks,
+                                 ...)))
+}

### 2018-01-10 Distance/R/gof_ds.R

@@ -1,0 +1,40 @@
+#' Goodness of fit testing and quantile-quantile plots
+#'
+#' Formal goodness of fit testing for detection function models using Kolmogorov-Smirnov and Cramer-von Mises tests. Both tests are based on looking at the quantile-quantile plot produced by \code{\link{qqplot.ddf}} and deviations from the line x=y.
+#'
+#' The Kolmogorov-Smirnov test asks the question “what’s the largest vertical distance between a point and the y = x line?” It uses this distance as a statistic to test the null hypothesis that the samples (EDF and CDF in our case) are from the same distribution (and hence our model fits well). If the deviation between the y = x line and the points is too large we reject the null hypothesis and say the model doesn’t have a good fit.
+#'
+#' Rather than looking at the single biggest difference between the y = x line and the points in the Q-Q plot, we might prefer to think about all the differences between line and points, since there may be many smaller differences that we want to take into account rather than looking for one large deviation. Its null hypothesis is the same, but the statistic it uses is the sum of the deviations from each of the point to the line.
+#
+#' @section Details:
+#'
+#' Note that a bootstrap procedure is required for the Kolmogorov-Smirnov test to ensure that the p-values from the procedure are correct as the we are comparing the cumulative distribution function (CDF) and empirical distribution function (EDF) and we have estimated the parameters of the detection function. The \code{nboot} parameter controls the number of bootstraps to use. Set to \code{0} to avoid computing bootstraps (much faster but with no Kolmogorov-Smirnov results, of course).
+#'
+#' @param model a fitted detection function.
+#' @param plot if \code{TRUE} the Q-Q plot is plotted
+#' @param chisq if \code{TRUE} then chi-squared statistic is calculated even for models that use exact distances. Ignored for models that use binned distances
+#' @param nboot number of replicates to use to calculate p-values for the Kolmogorov-Smirnov goodness of fit test statistics
+#' @param ks perform the Kolmogorov-Smirnov test (this involves many bootstraps so can take a while)
+#' @param ... other arguments to be passed to \code{\link{ddf.gof}}
+#' @export
+#' @examples
+#' \dontrun{
+#' # fit and test a simple model for the golf tee data
+#' library(Distance)
[…]

### 2018-01-10 Distance/R/logLik.dsmodel.R

@@ -1,0 +1,29 @@
+#' log-likelihood value for a fitted detection function
+#'
+#' Extract the log-likelihood from a fitted detection function.
+#'
+#' @param object a fitted detection function model object
+#' @param \dots included for S3 completeness, but ignored
+#' @return a numeric value giving the log-likelihood with two attributes: \code{"df"} the "degrees of freedom" for the model (number of parameters) and \code{"nobs"} the number of observations used to fit the model
+#' @export
+#' @author David L Miller
+#' @examples
+#' \dontrun{
+#' library(Distance)
+#' data(minke)
+#' model <- ds(minke, truncation=4)
+#' # extract the log likelihood
+#' logLik(model)
+#' }
+logLik.dsmodel <- function(object, ...){
+
+  # see ?logLik for information on why
+
+  ret <- object$ddf$lnl
+
+  attr(ret, "df") <- length(object$ddf$par)
+  attr(ret, "nobs") <- nrow(object$ddf$data)
+
+  class(ret) <- "logLik"
+  return(ret)
+}

### 2018-01-10 Distance/R/model.description.R

@@ -1,0 +1,27 @@
+model.description<-function(model){
+# builds the description of the model to be printed
+# only used internally.
+
+  key <- switch(model$ds$aux$ddfobj$type,
+                unif = "Uniform",
+                hn   = "Half-normal",
+                hr   = "Hazard-rate")
+
+  mod.str <- paste(key, "key function")
+
+  if(!is.null(model$ds$aux$ddfobj$adjustment)){
+    adj.series <- switch(model$ds$aux$ddfobj$adjustment$series,
+                       cos  = "cosine",
+                       herm = "Hermite polynomial",
+                       poly = "simple polynomial")
+    mod.str <- paste(mod.str, "with", adj.series, "adjustment term")
+
+    adj.order <- model$ds$aux$ddfobj$adjustment$order
+    if(length(adj.order)>1){
+      mod.str <- paste(mod.str, "s", sep="")
+    }
+    mod.str <- paste(mod.str, "of order", paste(adj.order, collapse=","))
+  }
+
+  return(mod.str)
+}

### 2018-01-10 Distance/R/plot.dsmodel.R

@@ -1,0 +1,19 @@
+#' Plot a fitted detection function
+#'
+#' This is just a simple wrapper around \code{\link{plot.ds}}. See the manual
+#' page for that function for more information.
+#'
+#' @param x an object of class \code{dsmodel}.
+#' @param pl.den shading density for histogram (default \code{0}, no shading)
+#' @param ... extra arguments to be passed to \code{\link{plot.ds}}.
+#' @return \code{NULL}, just produces a plot.
+#' @aliases plot.dsmodel
+#' @export
+#' @author David L. Miller
+#' @importFrom graphics plot
+plot.dsmodel <- function(x, pl.den=0, ...){
+
+  plot(x$ddf, pl.den=pl.den, ...)
+
+  invisible()
+}

### 2018-01-10 Distance/R/print.summary.dsmodel.R

@@ -1,0 +1,110 @@
+#' Print summary of distance detection function model object
+#' 
+#' Provides a brief summary of a distance sampling analysis. Including: detection 
+#' function parameters, model selection criterion, and optionally abundance in the
+#' covered (sampled) region and its standard error.
+#' 
+#' @aliases print.summary.dsmodel
+#' @param x a summary of distance sampling analysis
+#' @param \dots unspecified and unused arguments for S3 consistency
+#' @return Nothing, just prints the summary.
+#' @author David L. Miller and Jeff Laake
+#' @seealso \code{\link{summary.ds}}
+#' @keywords utility
+#' @export
+print.summary.dsmodel <- function (x,...){
+
+  # split up the object
+  dht.obj <- x$dht
+  model <- x$ddf
+  x <- x$ds
+
+  # routine from dht to print the tables...
+  # stolen from print.dht, removed vcmatrices and cor arguments
+  print.tables <- function(x,bysample){
+     cat("\nSummary statistics:\n")
+     print(x$summary)
+     if("N" %in% names(x))
+     {
+        cat("\nAbundance:\n")
+        print(x$N)
+     }
+     cat("\nDensity:\n")
+     print(x$D)
+     if(bysample)
+     {
+        cat("\nEstimates by sample:\n")
+        print(x$bysample)
+     }
+  }
+
+  cat("\nSummary for distance analysis \n")
+  cat("Number of observations : ", x$n,"\n")
+  cat("Distance range         : ", x$left, " - ",x$width,"\n")
+
+  
+  cat("\nModel :",model.description(model),"\n")
+  # Remind the user that monotonicity constraints were enforced
+  if(x$mono & x$mono.strict){
+    cat("\nStrict monotonicity constraints were enforced.\n")
+  }else if(x$mono){
+    cat("\nMonotonicity constraints were enforced.\n")
+  }
+  cat("AIC   :", x$aic, "\n")
+
+  # parameter summaries
+  cat("\nDetection function parameters\n")
+  cat("Scale coefficient(s): ", "\n")
+  print(x$coeff$key.scale)
+  if(x$key %in% c("gamma","hr")) {
+    cat("\nShape coefficient(s): ", "\n")
+    print(x$coeff$key.shape)
+  }
+  if(!is.null(x$coeff$adj.parm)) {
[…]

### 2018-01-10 Distance/R/summarize_ds_models.R

@@ -1,0 +1,153 @@
+#' Make a table of summary statistics for detection function models
+#'
+#' Provide a summary table of useful information about fitted detection functions. This can be useful when paired with \code{knitr}s \code{kable} function. By default models are sorted by AIC and will therefore not allow models with different truncations and distance binning.
+#'
+#' Note that the column names are in LaTeX format, so if you plan to manipulate the resulting \code{data.frame} in R, you may wish to rename the columns for ease of access.
+#'
+#' @param ... models to be summarised
+#' @param sort column to sort by (default \code{"AIC"})
+#' @param output should the output be given in \code{"latex"} compatible format or as \code{"plain"} text?
+#' @param delta_only only output AIC differences (default \code{TRUE})
+#' @author David L Miller
+#' @export
+#' @examples
+#' \dontrun{
+#' # fit some models to the golf tee data
+#' library(Distance)
+#' data(book.tee.data)
+#' tee.data<-book.tee.data$book.tee.dataframe[book.tee.data$book.tee.dataframe$observer==1,]
+#' model_hn <- ds(tee.data,4)
+#' model_hr <- ds(tee.data,4, key="hr")
+#' summarize_ds_models(model_hr, model_hn, output="plain")
+#'}
+summarize_ds_models <- function(..., sort="AIC", output="latex", delta_only=TRUE){
+
+  # get the models
+  models <- list(...)
+
+  # get the model names
+  model_names <- setdiff(as.character(match.call(expand.dots=TRUE)),
+                         as.character(match.call(expand.dots=FALSE)))
+
+
+  ## checking
+  # can't compare models with different truncations
+  r_truncs <- unlist(lapply(models, function(x) x$ddf$meta.data$width))
+  l_truncs <- unlist(lapply(models, function(x) x$ddf$meta.data$left))
+  if(!all(abs(c(r_truncs-mean(r_truncs),
+                l_truncs-mean(l_truncs))) < 1e-8)){
+    stop("All truncations must be the same for AIC comparison.")
+  }
+  # check all binned
+  binned <- unlist(lapply(models, function(x) x$ddf$meta.data$binned))
+  if((any(binned) & !all(binned)) | (any(!binned) & !all(!binned))){
+    stop("Can't compare binned and unbinned distances")
+  }
+  # check all binning is the same
+  if(all(binned)){
[…]

### 2018-01-10 Distance/R/summary.dsmodel.R

@@ -1,0 +1,32 @@
+#' Summary of distance sampling analysis 
+#' 
+#' Provides a brief summary of a distance sampling analysis. This includes 
+# parameters, model selection criterion, and optionally abundance in the
+# covered (sampled) region and its standard error.
+# 
+# The argument \code{N} is used to suppress computation of
+# abundance and average detection probability in calls to summarize the
+# \code{ds} and either \code{io.fi} or \code{trial.fi} for summaries of
+# \code{io} and \code{trial} objects respectively which are composed of a
+# \code{ds} model object and a mark-recapture model object. The corresponding
+# print function is called to print the summary results.
+# 
+#' @aliases summary.dsmodel
+#' @param object a distance analysis
+#' @param \dots unspecified and unused arguments for S3 consistency
+#' @return list of extracted and summarized objects
+#' @note This function just calls \code{\link{summary.ds}} and \code{\link{dht}}, collates and prints the results in a nice way.
+#' @author David L. Miller
+#' @keywords utility
+#' @export
+summary.dsmodel <- function(object,...){
+
+  #  se if TRUE, computes standard errors
+  #  N if TRUE, computes abundance in covered (sampled) region
+  ans <- list(ds=summary(object$ddf, se=TRUE, N=TRUE),
+              dht=object$dht,
+              ddf=object$ddf)
+
+  class(ans) <- "summary.dsmodel"
+  return(ans)
+}

### 2018-01-10 Distance/tests/testthat/test_dht.R

@@ -1,0 +1,30 @@
+context("Testing dht")
+
+## first need to take the golf tee data
+data(book.tee.data)
+region <- book.tee.data$book.tee.region
+samples <- book.tee.data$book.tee.samples
+obs <- book.tee.data$book.tee.obs
+obs <- obs[order(obs$object),]
+egdata <- book.tee.data$book.tee.dataframe
+# take only observer 1 data
+egdata <- egdata[egdata$observer==1,]
+
+test_that("dht without obs.table",{
+
+  # first with obs table
+  ds.dht.model<-suppressMessages(ds(egdata,4,region.table=region, monotonicity="strict", sample.table=samples,obs.table=obs))
+
+  # check if you just forget obs.table an error is thrown
+  # No longer throws error just displays to the user that it is only estimating detection function
+  #expect_error(suppressMessages(ds(egdata,4,region.table=region, monotonicity="strict", sample.table=samples)))
+
+  # test that the result is the same if we do the correct merge
+  dat <- merge(egdata, obs, by="object")
+  ds.dht.model2 <- suppressMessages(ds(dat, 4, region.table=region,
+                                       monotonicity="strict",
+                                       sample.table=samples))
+  expect_equivalent(ds.dht.model$dht, ds.dht.model2$dht)
+
+})
+

### 2018-01-10 Distance/tests/testthat/test_pt.R

@@ -1,0 +1,44 @@
+# point transect testing
+
+par.tol <- 1e-5
+p.tol <- 1e-4
+lnl.tol <- 1e-4
+
+context("Point transects")
+
+test_that("ptexample.distance from mrds gives same results",{
+
+  set.seed(123)
+  data(ptdata.distance)
+  # mrds code for the same example
+  #xx <- ddf(dsmodel=~cds(key="hn", formula=~1), data=ptdata.distance,
+  #          method="ds",
+  #          meta.data=list(point=TRUE,width=max(ptdata.distance$distance)))#,
+  xx <- ds(ptdata.distance, key="hn", transect="point", adjustment=NULL)
+
+  expect_that(xx$ddf$par,equals(2.283007,tol=par.tol))
+  expect_that(xx$ddf$lnl,equals(-458.5701,tol=lnl.tol))
+  expect_that(summary(xx$ddf)$average.p,equals(0.1644303,tol=p.tol))
+
+})
+
+test_that("ptexample.single from mrds gives same results -- binned",{
+
+  data(ptdata.single)
+
+  set.seed(345)
+
+  # mrds code for the same example
+  #ptdata.single$distbegin <- (as.numeric(cut(ptdata.single$distance,10*(0:10)))-1)*10
+  #ptdata.single$distend <- (as.numeric(cut(ptdata.single$distance,10*(0:10))))*10
+  #model <- ddf(data=ptdata.single, dsmodel=~cds(key="hn"),
+  #                meta.data=list(point=TRUE,binned=TRUE,breaks=10*(0:10)))
+
+  xx <-suppressMessages(ds(ptdata.single,key="hn",transect="point",
+                           cutpoints=10*(0:10),order=0))
+
+  expect_that(xx$ddf$par,equals(3.397266,tol=par.tol))
+  expect_that(xx$ddf$lnl,equals(-687.7673,tol=lnl.tol))
+  expect_that(summary(xx$ddf)$average.p,equals(0.1779359,tol=p.tol))
+
+})

### 2018-01-10 Distance/tests/testthat/test_simpletable.R

@@ -1,0 +1,66 @@
+context("Testing simplified data tables (tee data)")
+
+## first need to take the golf tee data
+data(book.tee.data)
+region <- book.tee.data$book.tee.region
+samples <- book.tee.data$book.tee.samples
+obs <- book.tee.data$book.tee.obs
+obs <- obs[order(obs$object),]
+egdata <- book.tee.data$book.tee.dataframe
+# take only observer 1 data
+egdata <- egdata[egdata$observer==1,]
+
+# merge sample onto region
+flatdat <- merge(region,samples,by="Region.Label")
+# merge obs table onto that
+flatdat <- merge(flatdat,obs,by=c("Sample.Label","Region.Label"),all.x=TRUE)
+# finally merge the distances onto that
+flatdat <- merge(flatdat,egdata,by="object",all.x=TRUE)
+
+
+
+
+test_that("Recover structure for tee data",{
+
+  ## check we can recover the 4 data frames
+  sepdat <- Distance:::checkdata(flatdat)
+
+  # this should be a factor really, as is elsewhere
+  obs$Region.Label <- as.factor(obs$Region.Label)
+
+  ## now egdata should actually have the Region.Label and Sample.Label columns
+  egdata <- merge(egdata, obs)
+  egdata <- egdata[,c(1,9,8,2:7)]
+
+  expect_equivalent(sepdat$region.table,region)
+  expect_equivalent(sepdat$obs.table,obs)
+  expect_equivalent(sepdat$sample.table,samples)
+  expect_equivalent(sepdat$data,egdata)
+
+
+})
+
+test_that("Errors work",{
+
+  flatdat[nrow(flatdat)+1,] <- flatdat[1,]
+  flatdat[nrow(flatdat),]$Effort <- 1322
+
+  expect_error(Distance:::checkdata(flatdat))
+
+  # check we get an error if the columns are right but their case is wrong
+  flatdat$effort <- flatdat$Effort
+  flatdat$Effort <- NULL
+  expect_error(Distance:::checkdata(flatdat))
+
+})
+#test_that("Extra regions",{
+#
+#  # add an extra region
+#  region.e <- data.frame(Region.Label=1:3,Area=c(1040,640,500))
+#  samples
+#
+#  # estimate using all tables
+#  dsmod.e <- ds(egdata,4,region.table=region.e, sample.table=samples,
+#                obs.table=obs)
+#
+#}

### 2018-01-10 Distance/tests/testthat/test_summarize.R

@@ -1,0 +1,56 @@
+# test ds()
+library(Distance)
+
+par.tol <- 1e-5
+N.tol <- 1e-3
+lnl.tol <- 1e-4
+
+context("summarize_ds_models()")
+
+# data setup
+data(book.tee.data)
+egdata <- book.tee.data$book.tee.dataframe
+egdata <- egdata[!duplicated(egdata$object),]
+
+test_that("Error on different truncation distance",{
+  set.seed(100)
+  # some models
+  t4 <- ds(egdata,4)
+  t42 <- t4
+  t4hr <- ds(egdata,4, key="hr")
+  t3 <- ds(egdata,3)
+  t14 <- ds(egdata,list(left=1, right=4))
+
+  # when things are okay!
+  out <- "           Model                                       Key function Formula\n1   \\\\texttt{t4}                                        Half-normal      ~1\n2  \\\\texttt{t42}                                        Half-normal      ~1\n3 \\\\texttt{t4hr} Hazard-rate with cosine adjustment term of order 2      ~1\n  C-vM p-value $\\\\hat{P_a}$ se($\\\\hat{P_a}$) $\\\\Delta$AIC\n1    0.7789695    0.5842744       0.04637627     0.000000\n2    0.7789695    0.5842744       0.04637627     0.000000\n3    0.8170780    0.5556728       0.07197512     2.501712"
+  expect_output(print(summarize_ds_models(t4, t42, t4hr)), out, fixed=TRUE)
+
+  # right truncation different
+  expect_error(summarize_ds_models(t3,t4))
+  # left
+  expect_error(summarize_ds_models(t4,t14))
+  # both
+  expect_error(summarize_ds_models(t3,t4, t41))
+})
+
+
+test_that("Binning",{
+
+  # largest cutpoint
+  cp1 <- suppressMessages(ds(egdata,key="hn",order=0,
+                  cutpoints=c(0,1,2,3,3.84)))
+  cp11 <- cp1
+  cp2 <- suppressMessages(ds(egdata,key="hn",order=0,
+                  cutpoints=c(0,2,3,3.84)))
+
[…]

### 2018-01-10 make.sh

@@ -1,7 +1,7 @@
 #!/bin/sh
 
-Rscript -e "library(knitr);knit('paper.Rnw')"
-pdflatex paper.tex
-bibtex paper
-pdflatex paper.tex
-open paper.pdf
+Rscript -e "library(knitr);knit('jss2630.Rnw')"
+pdflatex jss2630.tex
+bibtex jss2630
+pdflatex jss2630.tex
+open jss2630.pdf

### 2018-01-10 mrds/R/DeltaMethod.R

@@ -1,0 +1,61 @@
+#' Numeric Delta Method approximation for the variance-covariance matrix
+#'
+#' Computes delta method variance-covariance matrix of results of any generic function \code{fct} that computes a vector of estimates as a function of a set of estimated parameters \code{par}.
+#'
+#' The delta method (aka propogation of errors is based on Taylor series approximation - see Seber's book on Estimation of Animal Abundance). It uses the first derivative of \code{fct} with respect to \code{par} which is computed in this function numerically using the central-difference formula. It also uses the variance-covariance matrix of the estimated parameters which is derived in estimating the parameters and is an input argument.
+#'
+#' The first argument of \code{fct} should be \code{par} which is a vector of parameter estimates. It should return a single value (or vector) of estimate(s).  The remaining arguments of \code{fct} if any can be passed to \code{fct} by including them at the end of the call to \code{DeltaMethod} as \code{name=value} pairs.
+#'
+#' @param par vector of parameter values at which estimates should be constructed
+#' @param fct function that constructs estimates from parameters \code{par}
+#' @param vcov variance-covariance matrix of the parameters
+#' @param delta proportional change in parameters used to numerically estimate first derivative with central-difference formula
+#' @param \dots any additional arguments needed by \code{fct}
+#' @export
+#' @return a list with values \item{variance}{estimated variance-covariance matrix of estimates derived by \code{fct}} \item{partial}{ matrix (or vector) of partial derivatives of \code{fct} with respect to the parameters \code{par}}
+#' @note This is a generic function that can be used in any setting beyond the \code{mrds} package. However this is an internal function for \code{mrds} and the user does not need to call it explicitly.
+#' @author Jeff Laake
+#' @keywords utility
+DeltaMethod <- function(par, fct, vcov, delta, ...){
+
+  # Construct theta call to fct
+  theta <- function(par) fct(par,...)
+
+  # Numerically compute the first derivative of the estimator with
+  # respect to each parameter using a central difference formula
+  savepar <- par
+  value1 <- theta(par)
+  partial <- matrix(0,nrow=length(par),ncol=length(value1))
+  for(i in seq_along(par)){
+    # Store the original parameters into par and then adjust the
+    # ith parameter by adding the proportion based on delta
+    if(savepar[i]!=0){
+      deltap <- delta*savepar[i]
+    }else{
+      deltap <- delta
+    }
+
+    par <- savepar
[…]

### 2018-01-10 mrds/R/NCovered.io.R

@@ -1,0 +1,35 @@
+NCovered.io <- function(par,model,data,group=TRUE,...){
+  # see docs in NCovered
+  #
+  # computes abundance in covered region for io model object
+  #
+  # par   - parameter values (used when computing derivatives wrt 
+  #         parameter uncertainty)
+  # model - ddf model object
+  # group - if TRUE computes group abundance and if FALSE individual abundance
+  #
+  # result - abundance estimate
+  #
+  # Functions Used: predict (predict.io), compute.Nht 
+
+  # set pars if we are differentiating
+  # then extract fitted values
+  if(!is.null(par)){
+    model$mr$mr$coefficients <- par[seq_along(model$mr$mr$coefficients)]
+    model$ds$par <- par[(length(model$mr$mr$coefficients)+1):length(par)]
+    model$ds$ds$aux$ddfobj <- assign.par(model$ds$ds$aux$ddfobj,model$ds$par)
+    fitted <- predict(model, compute=TRUE)$fitted
+  }else{
+    fitted <- model$fitted
+  }
+
+  if(!group){
+    size <- model$data$size[model$data$observer==1&model$data$object %in%
+                            as.numeric(names(model$fitted))]
+    Nhat <- sum(compute.Nht(fitted,FALSE,size))
+  }else{
+    Nhat <- sum(compute.Nht(fitted,TRUE,size=NULL))
+  }
+
+  return(Nhat)
+}

### 2018-01-10 mrds/R/NCovered.rem.R

@@ -1,0 +1,28 @@
+# Compute abundance in covered region for trial model object
+#
+# par    - parameter values (used when computing derivatives wrt parameter uncertainty)
+# model  - ddf model object
+# group  - if TRUE computes group abundance and if FALSE individual abundance
+#
+# result - abundance estimate
+NCovered.rem <- function(par, model, group=TRUE, ...){
+  if(!is.null(par)){
+    model$mr$mr$coefficients <- par[seq_along(model$mr$mr$coefficients)]
+    model$ds$par <- par[(length(model$mr$mr$coefficients)+1):length(par)]
+    model$ds$ds$aux$ddfobj <- assign.par(model$ds$ds$aux$ddfobj, model$ds$par)
+    fitted <- predict(model, compute=TRUE, integrate=TRUE)$fitted
+  }else{
+    fitted <- model$fitted
+  }
+
+  if(!group){
+    size <- model$data$size[model$data$observer==1 &
+                            model$data$object %in%
+                              as.numeric(names(model$fitted))]
+    Nhat <- sum(compute.Nht(fitted,FALSE,size))
+  }else{
+    Nhat <- sum(compute.Nht(fitted,TRUE,size=NULL))
+  }
+
+  return(Nhat)
+}

### 2018-01-10 mrds/R/NCovered.trial.R

@@ -1,0 +1,28 @@
+NCovered.trial <- function(par, model, group=TRUE, ...){
+# Compute abundance in covered region for trial model object
+#
+# par   - parameter values (used when computing derivatives wrt par uncertainty)
+# model - ddf model object
+# group - if TRUE computes group abundance and if FALSE individual abundance
+#
+# result - abundance estimate
+  if(!is.null(par)){
+    model$mr$mr$coefficients <- par[1:length(model$mr$mr$coefficients)]
+    model$ds$par <- par[(length(model$mr$mr$coefficients)+1):length(par)]
+    model$ds$ds$aux$ddfobj <- assign.par(model$ds$ds$aux$ddfobj, model$ds$par)
+    fitted <- predict(model, compute=TRUE, integrate=TRUE)$fitted
+  }else{
+    fitted <- model$fitted
+  }
+
+  if(!group){
+    size <- model$data$size[model$data$observer==1 &
+                            model$data$object %in%
+                             as.numeric(names(model$fitted))]
+    Nhat <- sum(compute.Nht(fitted, FALSE, size))
+  }else{
+    Nhat <- sum(compute.Nht(fitted, TRUE, size=NULL))
+  }
+
+  return(Nhat)
+}

### 2018-01-10 mrds/R/adjfct.cos.R

@@ -1,0 +1,29 @@
+# Cosine adjustment terms
+#
+# distance  - perpendicular distance vector
+# scaling     - scale parameter
+# adj.order - vector of orders of Cosine terms to fit
+# adj.parm  - vector of parameters (a_j)
+adjfct.cos <- function(distance,scaling=1,adj.order,adj.parm=NULL,
+                       adj.exp=FALSE){
+
+  # Check the adjustment parameters
+  if(is.null(adj.parm)){
+    adj.parm <- as.vector(rep(1,length(adj.order)))
+  }
+
+  adj.order <- as.vector(adj.order)
+
+  cossum <- 0
+
+  for(i in seq_along(adj.order)){
+    cossum <- cossum + (adj.parm[i]*cos((adj.order[i]*pi*distance)/scaling))
+  }
+
+  # if adj.exp return exp(cossum) to keep f(x)>0  
+  if(adj.exp){
+    return(exp(cossum))
+  }else{
+    return(cossum)
+  }
+}

### 2018-01-10 mrds/R/adjfct.herm.R

@@ -1,0 +1,32 @@
+# adjfct.herm
+#
+# Hermite polynomial adjustment terms
+#
+# distance  - perpendicular distance vector
+# scaling     - scale parameter
+# adj.order - vector of orders of polynomials to fit
+# adj.parm  - vector of parameters (a_j)
+adjfct.herm <- function(distance, scaling=1, adj.order, adj.parm=NULL,
+                        adj.exp=FALSE){
+
+  # Check the adjustment parameters
+  if(is.null(adj.parm)){
+    adj.parm <- as.vector(rep(1,length(adj.order)))
+  }
+
+  adj.order <- as.vector(adj.order)
+
+  hermsum <- 0
+
+  for(i in seq_along(adj.order)){
+    hermsum <- hermsum + (adj.parm[i] * hermite.poly((distance/scaling),
+                                                     adj.order[i]))
+  }
+
+  # if adj.exp return exp(hermsum) to keep f(x)>0
+  if(adj.exp){
+    return(exp(hermsum))
+  }else{
+    return(hermsum)
+  }
+}

### 2018-01-10 mrds/R/adjfct.poly.R

@@ -1,0 +1,32 @@
+# adjfct.poly
+#
+# Polynomial adjustment terms
+#
+# distance  - perpendicular distance vector
+# scaling     - scale parameter
+# adj.order - vector of orders of polynomials to fit
+# adj.parm  - vector of parameters (a_j)
+adjfct.poly <- function(distance,scaling=1,adj.order,adj.parm=NULL,
+                        adj.exp=FALSE){
+
+  # Check the adjustment parameters
+  if(is.null(adj.parm)){
+    adj.parm <- as.vector(rep(1,length(adj.order)))
+  }
+
+  adj.order <- as.vector(adj.order)
+
+  # Should have checked the order beforehand
+  polysum <- 0
+
+  for(i in seq_along(adj.order)){
+    polysum <- polysum + (adj.parm[i]*(distance/scaling)^adj.order[i])
+  }
+
+  # if adj.exp return exp(polysum) to keep f(x)>0 
+  if(adj.exp){
+    return(exp(polysum))
+  }else{
+    return(polysum)
+  }
+}

### 2018-01-10 mrds/R/assign.par.R

@@ -1,0 +1,36 @@
+#' Extraction and assignment of parameters to vector
+#'
+#' Assigns parameters of a particular type (scale, shape, adjustments or g0 (p(0))) from the vector of parameters in \code{ddfobj}. All of the parameters are kept in a single vector for optimization even though they have very different uses. \code{assign.par} parses them from the vector based on a known structure and assigns them into \code{ddfobj}.
+#'
+#' \code{getpar} extracts the requested types to be extracted from \code{ddfobj}.
+#'
+#' @aliases assign.par
+#' @param ddfobj distance sampling object (see \code{\link{create.ddfobj}})
+#' @param fpar parameter vector
+#' @return if \code{index==FALSE}, vector of parameters that were requested or \code{index==TRUE}, vector of 3 indices for scale, shape, adjustment
+#' @note Internal functions not intended to be called by user.
+#' @seealso getpar
+#' @author Jeff Laake
+#' @keywords utility internal
+#' @export
+assign.par <- function(ddfobj, fpar){
+  index <- 1
+  # Shape parameters
+  if(!is.null(ddfobj$shape)){
+    ddfobj$shape$parameters <- fpar[1:ncol(ddfobj$shape$dm)]
+    index <- index+ncol(ddfobj$shape$dm)
+  }
+
+  # Scale parameters
+  if(!is.null(ddfobj$scale)){
+    ddfobj$scale$parameters <- fpar[index:(index+ncol(ddfobj$scale$dm)-1)]
+    index <- index+ncol(ddfobj$scale$dm)
+  }
+
+  # Adjustment parameters
+  if(!is.null(ddfobj$adjustment)){
+    ddfobj$adjustment$parameters <- fpar[index:length(fpar)]
+  }
+
+  return(ddfobj)
+}

### 2018-01-10 mrds/R/calc.se.Np.R

@@ -1,0 +1,49 @@
+#' Find se of average p and N
+#'
+#' @param model a \code{ddf} model object
+#' @param avgp average p function
+#' @param n sample size
+#' @param average.p the average probability of detection for the model
+#'
+#' @author David L. Miller
+calc.se.Np <- function(model, avgp, n, average.p){
+
+  # if we have any NAs in the hessian then everything is
+  # messed up, so we should just return NAs
+  if(any(is.na(model$hessian))){
+    se.obj <- list()
+    se.obj$Nhat.se <- NA
+    se.obj$average.p.se <- NA
+    model$hessian <- as.matrix(model$hessian)
+    se.obj$Nhatvar.list <-list(variance = matrix(NA, nrow(model$hessian),
+                                                 ncol(model$hessian)),
+                               partial  = matrix(NA, length(model$par),
+                                                 length(model$par)))
+    se.obj$vcov <- matrix(NA, nrow(model$hessian), ncol(model$hessian))
+  }else{
+    # calculate the variance-covariate matrix
+    vcov <- solvecov(model$hessian)$inv
+
+    # calculate Nhat uncertainty
+    Nhatvar.list <- DeltaMethod(model$par, NCovered, vcov, 0.001,
+                                model=model, group=TRUE)
+    Nhatvar <- Nhatvar.list$variance + sum((1-model$fitted)/model$fitted^2)
+    cvN <- sqrt(Nhatvar)/model$Nhat
+
+    # calculate the average p uncertainty
+    var.pbar.list <- prob.se(model, avgp, vcov)
+    covar <- t(Nhatvar.list$partial) %*% vcov %*% var.pbar.list$partial+
+                 var.pbar.list$covar
+    var.pbar <- average.p^2*(cvN^2 + var.pbar.list$var/n^2-
+                                  2*covar/(n*model$Nhat))
+
+    # what should we return?
+    se.obj <- list()
+    se.obj$Nhat.se <- sqrt(Nhatvar)
+    se.obj$average.p.se <- sqrt(var.pbar)
+    se.obj$Nhatvar.list <- Nhatvar.list
+    se.obj$vcov <- vcov
+  }
+
+  return(se.obj)
+}

### 2018-01-10 mrds/R/check.bounds.R

@@ -1,0 +1,55 @@
+#' Check parameters bounds during optimsations
+#'
+#' Simple internal function to check that the optimisation didn't hit bounds. Based on code that used to live in \code{detfct.fit.opt}.
+#'
+#' @param lt optimisation object
+#' @param lowerbounds current lower bounds
+#' @param upperbounds current upper bounds
+#' @param ddfobj ddf object
+#' @param showit debug level
+#' @param setlower were lower bounds set by the user
+#' @param setupper were upper bounds set by the user
+#' @return \code{TRUE} if bounded (ie parameters close to bound), else \code{FALSE}
+#'
+#' @author Dave Miller; Jeff Laake
+check.bounds <- function(lt, lowerbounds, upperbounds, ddfobj, showit,
+                         setlower, setupper){
+
+  tol <- 1e-6
+
+  # function to check upper/lower bounds
+  chk.bnds <- function(par, bounds, bound.label, set, tol){
+
+    if(set) return(FALSE)
+
+    if(bound.label=="lower" && (any(par<bounds) | any(abs((bounds-par))<tol))){
+      bounded <- TRUE
+    }else if(bound.label=="upper" && (any(par>bounds) | any(abs(((bounds-par)<tol))))){
+      bounded <- TRUE
+    }else{
+      bounded <-FALSE
+    }
+
+    # Issue message if any of the parameters are at their bounds
+    if(bounded & showit>=1){
+      message(paste("One or more parameters was at a", bound.label, "bound\n",
+                    "Parameters:", paste(par, collapse=", "), "\n",
+                    bound.label, "bounds:", paste(bounds, collapse=", ")))
+    }
+    return(bounded)
+  }
+
+  ## check lower bounds
+  # handle hazard rate power par
+  if(ddfobj$type=="hr"){
+    bounded <- chk.bnds(lt$par[2:length(lt$par)],
+                        lowerbounds[2:length(lt$par)], "lower", setlower, tol)
+  }else{
+    bounded <- chk.bnds(lt$par, lowerbounds, "lower", setlower, tol)
+  }
+
+  ## check upper bounds
+  bounded <- bounded | chk.bnds(lt$par, upperbounds, "upper", setupper, tol)
+
+  return(bounded)
+}

### 2018-01-10 mrds/R/check.mono.R

@@ -1,0 +1,211 @@
+#' Check that a detection function is monotone
+#'
+#' Check that a fitted detection function is monotone non-increasing.
+#'
+#' Evaluates a series of points over the range of the detection function (left to right truncation) then determines:
+#'
+#' 1. If the detection function is always less than or equal to its value at the left truncation poin (\code{g(x)<=g(left)}, or usually \code{g(x)<=g(0)}).
+#' 2. (Optionally) The detection function is always monotone decreasing (\code{g(x[i])<=g(x[i-1])}). This check is only performed when \code{strict=TRUE} (the default).
+#' 3. The detection function is never less than 0 (\code{g(x)>=0}).
+#' 4. The detection function is never greater than 1 (\code{g(x)<=1}).
+#'
+#' For models with covarates in the scale parameter of the detection function is evaluated at all observed covariate combinations.
+#'
+#' Currently covariates in the shape parameter are not supported.
+#'
+#' @param df a fitted detection function object
+#' @param strict if \code{TRUE} (default) the detection function must be "strictly" monotone, that is that (\code{g(x[i])<=g(x[i-1])}) over the whole range (left to right truncation points).
+#' @param n.pts number of equally-spaced points between left and right truncation at which to evaluate the detection function (default 100)
+#' @param tolerance numerical tolerance for monotonicity checks (default 1e-6)
+#' @param plot plot a diagnostic highlighting the non-monotonic areas (default FALSE)
+#' @param max.plots when \code{plot=TRUE}, what is the maximum number of plots of non-monotone covariate combinations that should be plotted? Plotted combinations are a random sample of the non-monotonic subset of evaluations. No effect for non-covariate models.
+#'
+#' @return \code{TRUE} if the detection function is monotone, \code{FALSE} if it's not. \code{warning}s are issued to warn the user that the function is non-monotonic.
+#'
+#' @keywords utility
+#' @author David L. Miller
+#' @importFrom stats as.formula model.matrix
+#' @importFrom graphics polygon rug
+#' @importFrom grDevices rgb
+#' @export
+check.mono <- function(df, strict=TRUE, n.pts=100, tolerance=1e-6, plot=FALSE,
+                       max.plots=6){
+
[…]

### 2018-01-10 mrds/R/coef.trial.R

@@ -1,0 +1,30 @@
+#' @export
+coef.trial <- function(object,...){
+  par <- coef(object$mr)
+  row.names(par) <- paste("mr      :",row.names(coef(object$mr)))
+
+  if(!is.null(coef(object$ds)$exponent)){
+     rn <- row.names(par)
+     par <- rbind(par,coef(object$ds)$exponent)
+     row.names(par) <- c(rn, paste("ds expon:",
+                                   row.names(coef(object$ds)$exponent)))
+  }
+
+  rn <- row.names(par)
+  ds.scale <- coef(object$ds)$scale
+
+  if(!is.null(ds.scale)){
+    par <- rbind(par,coef(object$ds)$scale)
+    row.names(par) <- c(rn,paste("ds scale:",
+            row.names(coef(object$ds)$scale)))
+  }
+
+  if(!is.null(coef(object$ds)$adjustment)){
+    rn <- row.names(par)
+    par <- rbind(par,coef(object$ds)$adjustment)
+    row.names(par) <- c(rn, paste("ds adjust:",
+                               row.names(coef(object$ds)$adjustment)))
+  }
+
+  return(par)
+}

### 2018-01-10 mrds/R/covered.region.dht.R

@@ -1,0 +1,49 @@
+#' Covered region estimate of abundance from Horvitz-Thompson-like estimator
+#'
+#' Computes H-T abundance within covered region by sample.
+#'
+#'
+#' @param obs observations table
+#' @param samples samples table
+#' @param group if TRUE compute abundance of group otherwise abundance of
+#'   individuals
+#' @return Nhat.by.sample - dataframe of abundance by sample
+#' @note Internal function called by \code{\link{dht}} and related functions
+#' @author Jeff Laake
+#' @keywords utility
+covered.region.dht <- function(obs, samples, group){
+
+  # Compute abundance in covered region depending on value of
+  # group = TRUE (do group abundance); F(do individual abundance)
+  # if there are observations of this species
+  if(nrow(obs) > 0){
+    Nhats <- compute.Nht(obs$pdot, group, obs$size)
+
+    # Sum abundances by sample within region
+    Nhats <- by(Nhats, obs$Label, sum)
+
+    # Sum observations by sample within region
+    if(group){
+      sum.obs <- by(obs$object, obs$Label, length)
+    }else{
+      sum.obs <- by(obs$size, obs$Label, sum)
+    }
+
+    # Merge with samples
+    num.obs <- data.frame(Label = names(sum.obs), n = as.vector(sum.obs))
+    Nhats <- data.frame(Label = names(Nhats), Nhat = as.vector(Nhats))
+    Nhat.by.sample <- merge(samples, Nhats,by.x = "Label", by.y = "Label",
+                            all.x = TRUE)
+    Nhat.by.sample <- merge(Nhat.by.sample, num.obs, by.x = "Label",
+                            by.y = "Label", all.x = TRUE)
+    Nhat.by.sample$Nhat[is.na(Nhat.by.sample$Nhat)] <- 0
+    Nhat.by.sample$n[is.na(Nhat.by.sample$n)] <- 0
+
+  # Create Nhat.by.sample in the case where there were no sightings
+  }else{
+    Nhat.by.sample <- cbind(samples,
+                            Nhat = rep(0,nrow(samples)),
+                            n = rep(0,nrow(samples)))
+  }
+  return(Nhat.by.sample)
+}

### 2018-01-10 mrds/R/create.ddfobj.R

@@ -1,0 +1,163 @@
+#' Create detection function object
+#'
+#' Creates and populates a specific list structure to define a detection
+#' function object and its data. The \code{ddfobj} is used throughout the
+#' package as a calling argument to various functions.
+#'
+#' @param model model list with key function and possibly adjustment functions,
+#'   scale formula, and shape formula
+#' @param xmat model data frame
+#' @param meta.data list of options describing data like width, etc
+#' @param initial named list of initial values for parameters of the detection function (should have at least \code{"scale"}, maybe also have \code{"shape"} and \code{"adjustments"})
+#' @return Distance sampling function object list with elements that all can be
+#'   null except type: \item{type}{type of detection function
+#'   hn,hr,gamma,unif,logistic} \item{xmat}{model data frame}
+#'   \item{intercept.only}{TRUE if scale = ~1 and any shape formula =~1}
+#'   \item{scale}{sublist with elements (can be NULL i.e., unif key):formula,
+#'   parameters, design matrix (dm)} \item{shape}{sublist with elements (power
+#'   of hazard rate or gamma) (can be NULL i.e., unif or hn key):formula,
+#'   parameters, design matrix (dm)} \item{adjustment}{sublist with elements
+#'   (is NULL if no adjustments used):series,order,scale,parameters}
+#'   \item{g0}{sublist with elements (not used at present):formula,parameters,
+#'   design matrix(dm), link}
+#' @note Internal function not meant to be called by user
+#' @author Jeff Laake
+#' @seealso \code{\link{detfct}}, \code{\link{ddf}}
+#' @export
+#' @keywords internal
+create.ddfobj <- function(model, xmat, meta.data, initial){
+
+  # Create empty object and get values from cds or mcds function
+  ddfobj <- vector("list")
+  point <- meta.data$point
+  modpaste <- paste(model)
+  modelvalues <- eval(parse(text=modpaste[2:length(modpaste)]))
+
+  # Specify key function type
+  ddfobj$type <- modelvalues$key
+
+  if(ddfobj$type == "logistic"){
+    stop("Logistic detection function has been temporarily disabled")
+  }
+
+  if(!ddfobj$type %in% c("gamma", "hn", "hr", "unif", "th1", "th2")){
[…]

### 2018-01-10 mrds/R/create.model.frame.R

@@ -1,0 +1,97 @@
+#' Create a model frame for ddf fitting
+#'
+#' Creates a model.frame for distance detection function fitting. It includes
+#' some pre-specified and computed variables with those included in the model
+#' specified by user (formula)
+#'
+#' The following fields are always included: detected, observer, binned, and
+#' optionally distance (unless null), timesdetected (if present in data). If
+#' the distance data were binned, include distbegin and distend point fields.
+#' If the integration width varies also include int.begin and int.end and
+#' include an offset field for an iterative glm, if used.  Beyond these fields
+#' only fields used in the model formula are included.
+#'
+#' @param xmat dataframe for ddf
+#' @param scale.formula user specified formula for scale of distance detection
+#'   function
+#' @param meta.data user-specified meta.data (see \code{\link{ddf}}
+#' @param shape.formula user specified formula for shape parameter of distance
+#'   detection function
+#' @return model frame for analysis
+#' @note Internal function and not called by user
+#' @author Jeff Laake
+#' @keywords utility
+create.model.frame <- function(xmat,scale.formula,meta.data,shape.formula=NULL){
+  # Create data frame with variables used in the formula and object #
+  # (Object id); this code fix: jll 18-Nov-04; a different approach was used
+  # to allow for use of as.factor(x) etc in formula
+  if(!is.null(shape.formula)){
+    varlist <- unique(c(all.vars(scale.formula),all.vars(shape.formula)))
+  }else{
+    varlist <- all.vars(scale.formula)
+  }
+
+  if(any(! varlist %in% names(xmat))){
+     stop("The following variables used in the formula are not in the data: ",
+          paste(varlist[!varlist%in%names(xmat)], collapse=", "))
+  }
+
+  data <- cbind(object = xmat$object,
+                xmat[,varlist,drop=FALSE])
+  colnames(data) <- c("object",varlist)
+  data <- as.data.frame(data)
+
+  # fix: ljt 21-Sep-05; row.names of xmat need to be copied over to data,
+  #     otherwise some later stuff in e.g., ddf.ds doesn't work
+  row.names(data) <- row.names(xmat)
+
+  # Include other fields such as detected, observer, binned, and optionally
+  # distance (unless null)
+  if(is.null(data$distance)){
+    data <- cbind(data,
+                  distance=xmat$distance,
+                  binned=xmat$binned)
+  }else{
[…]

### 2018-01-10 mrds/R/ddf.R

@@ -1,0 +1,297 @@
+#' Distance Detection Function Fitting
+#'
+#' Generic function for fitting detection functions for distance sampling with
+#' single and double observer configurations. Independent observer, trial and
+#' dependent observer (removal) configurations are included. This is a generic
+#' function which does little other than to validate the calling arguments and
+#' methods and then calls the appropriate \code{method} specific function to do
+#' the analysis.
+#'
+#' The fitting code has certain expectations about \code{data}.  It should be a
+#' dataframe with at least the following fields named and defined as follows:
+#' \tabular{ll}{ \code{object} \tab object number \cr
+#'               \code{observer} \tab observer number (1 or 2) for double observer; only 1 if single observer \cr
+#'               \code{detected} \tab 1 if detected by the observer and 0 if missed; always 1 for single observer \cr
+#'               \code{distance} \tab perpendicular distance\cr }
+#' If the data are for clustered objects, the dataframe should also contain a
+#' field named \code{size} that gives the observed number in the cluster. If
+#' the data are for a double observer survey, then there are two records for
+#' each observation and each should have the same object number. The code
+#' assumes the observations are listed in the same order for each observer such
+#' that if the data are subsetted by \code{observer} there will be the same
+#' number of records in each and each subset will be in the same \code{object}
+#' order. In addition to these predefined and pre-named fields, the dataframe
+#' can have any number and type of fields that are used as covariates in the
+#' \code{dsmodel} and \code{mrmodel}. At present, discrepancies between
+#' observations in \code{distance}, \code{size} and any user-specified
+#' covariates cannot be assimilated into the uncertainty of the estimate. The
+#' code presumes the values for those fields are the same for both records
+#' (observer=1 and observer=2) and it uses the value from observer 1. Thus it
+#' makes sense to make the values the same for both records in each pair even
+#' when both detect the object or when observer 1 doesn't detect the object the
+#' data would have to be taken from observer 2 and would not be consistent.
+#'
+#' Five different fitting methods are currently available and these in turn
+#' define whether \code{dsmodel} and \code{mrmodel} need to be defined.
+#'
+#' \tabular{llll}{
+#' Method          \tab Single/Double \tab \code{dsmodel} \tab \code{mrmodel}\cr
[…]

### 2018-01-10 mrds/R/ddf.ds.R

@@ -1,0 +1,279 @@
+#' CDS/MCDS Distance Detection Function Fitting
+#'
+#' Fits a conventional distance sampling (CDS) (likelihood eq 6.6 in Laake and
+#' Borchers 2004) or multi-covariate distance sampling (MCDS)(likelihood eq
+#' 6.14 in Laake and Borchers 2004) model for the detection function of
+#' observed distance data.  It only uses key functions and does not incorporate
+#' adjustment functions as in CDS/MCDS analysis engines in DISTANCE (Marques
+#' and Buckland 2004). Distance can be grouped (binned), ungrouped (unbinned)
+#' or mixture of the two.  This function is not called directly by the user and
+#' is called from \code{ddf},\code{ddf.io}, or \code{ddf.trial}.
+#'
+#' For a complete description of each of the calling arguments, see
+#' \code{\link{ddf}}.  The argument \code{model} in this function is the same
+#' as \code{dsmodel} in \code{ddf}.  The argument \code{dataname} is the name
+#' of the dataframe specified by the argument \code{data} in \code{ddf}. The
+#' arguments \code{control},\code{meta.data},and \code{method} are defined the
+#' same as in \code{ddf}.
+#'
+#' @export
+#' @method ddf ds
+#' @param model model list with key function and scale formula if any
+#' @param data analysis dataframe
+#' @param meta.data list containing settings controlling data structure
+#' @param control list containing settings controlling model fitting
+#' @param call original function call if this function not called directly from
+#'   \code{ddf} (e.g., called via \code{ddf.io}
+#' @param method analysis method; only needed if this function called from
+#'   \code{ddf.io} or \code{ddf.trial}
+#' @return result: a ds model object
+#' @note If mixture of binned and unbinned distance, width must be set to be >=
+#'   largest interval endpoint; this could be changed with a more complicated
+#'   analysis; likewise, if all binned and bins overlap, the above must also
+#'   hold; if bins don't overlap, width must be one of the interval endpoints;
+#'   same holds for left truncation Although the mixture analysis works in
+#'   principle it has not been tested via simulation.
+#' @author Jeff Laake
+#' @seealso \code{\link{flnl}}, \code{\link{summary.ds}}, \code{\link{coef.ds}},
+#'   \code{\link{plot.ds}},\code{\link{gof.ds}}
[…]

### 2018-01-10 mrds/R/ddf.gof.R

@@ -1,0 +1,59 @@
+#' Goodness of fit tests for distance sampling models
+#'
+#' Generic function that computes chi-square goodness of fit test for detection function models with binned data and Cramer-von Mises and Kolmogorov-Smirnov (if \code{ks=TRUE})tests for exact distance data. By default a Q-Q plot is generated for exact data (and can be suppressed using the \code{qq=FALSE} argument).
+#'
+#' Formal goodness of fit testing for detection function models using Kolmogorov-Smirnov and Cramer-von Mises tests. Both tests are based on looking at the quantile-quantile plot produced by \code{\link{qqplot.ddf}} and deviations from the line x=y.
+#'
+#' The Kolmogorov-Smirnov test asks the question "what's the largest vertical distance between a point and the y=x line?" It uses this distance as a statistic to test the null hypothesis that the samples (EDF and CDF in our case) are from the same distribution (and hence our model fits well). If the deviation between the y=x line and the points is too large we reject the null hypothesis and say the model doesn't have a good fit.
+#'
+#' Rather than looking at the single biggest difference between the y=x line and the points in the Q-Q plot, we might prefer to think about all the differences between line and points, since there may be many smaller differences that we want to take into account rather than looking for one large deviation. Its null hypothesis is the same, but the statistic it uses is the sum of the deviations from each of the point to the line.
+#
+#' @section Details:
+#'
+#' Note that a bootstrap procedure is required for the Kolmogorov-Smirnov test to ensure that the p-values from the procedure are correct as the we are comparing the cumulative distribution function (CDF) and empirical distribution function (EDF) and we have estimated the parameters of the detection function. The \code{nboot} parameter controls the number of bootstraps to use. Set to \code{0} to avoid computing bootstraps (much faster but with no Kolmogorov-Smirnov results, of course).
+#'
+#' @aliases ddf.gof gof.io gof.io.fi gof.trial gof.trial.fi gof.rem gof.rem.fi
+#' @export
+#' @param model model object
+#' @param breaks Cutpoints to use for binning data
+#' @param nc Number of distance classes
+#' @param qq Flag to indicate whether quantile-quantile plot is desired
[…]

### 2018-01-10 mrds/R/ddf.io.R

@@ -1,0 +1,123 @@
+#' Mark-Recapture Distance Sampling (MRDS) IO - PI
+#'
+#' Mark-Recapture Distance Sampling (MRDS) Analysis of Independent Observer
+#' Configuration and Point Independence
+#'
+#' MRDS analysis based on point independence involves two separate and
+#' independent analyses of the mark-recapture data and the distance sampling
+#' data.  For the independent observer configuration, the mark-recapture data
+#' are analysed with a call to \code{\link{ddf.io.fi}} (see likelihood eq 6.8
+#' and 6.16 in Laake and Borchers 2004) to fit conditional distance sampling
+#' detection functions to estimate p(0), detection probability at distance zero
+#' for the independent observer team based on independence at zero (eq 6.22 in
+#' Laake and Borchers 2004). Independently, the distance data, the union of the
+#' observations from the independent observers, are used to fit a conventional
+#' distance sampling (CDS) (likelihood eq 6.6) or multi-covariate distance
+#' sampling (MCDS) (likelihood eq 6.14) model for the detection function, g(y),
+#' such that g(0)=1. The detection function for the observer team is then
+#' created as p(y)=p(0)*g(y) (eq 6.28 of Laake and Borchers 2004) from which
+#' predictions are made. \code{ddf.io} is not called directly by the user and
+#' is called from \code{\link{ddf}} with \code{method="io"}.
+#'
+#' For a complete description of each of the calling arguments, see
+#' \code{\link{ddf}}.  The argument \code{dataname} is the name of the
+#' dataframe specified by the argument \code{data} in \code{ddf}. The arguments
+#' \code{dsmodel}, \code{mrmodel}, \code{control} and \code{meta.data} are
+#' defined the same as in \code{ddf}.
+#'
+#' @export
+#' @method ddf io
+#' @param dsmodel distance sampling model specification; model list with key
+#'   function and scale formula if any
+#' @param mrmodel mark-recapture model specfication; model list with formula
+#'   and link
+#' @param data analysis dataframe
+#' @param meta.data list containing settings controlling data structure
+#' @param control list containing settings controlling model fitting
+#' @param call original function call used to call \code{ddf}
+#' @return result: an io model object which is composed of io.fi and ds model
+#'   objects
[…]

### 2018-01-10 mrds/R/ddf.io.fi.R

@@ -1,0 +1,326 @@
+#' Mark-Recapture Distance Sampling (MRDS) IO - FI
+#'
+#' Mark-Recapture Analysis of Independent Observer Configuration with Full
+#' Independence
+#'
+#' The mark-recapture data derived from an independent observer distance
+#' sampling survey can be used to derive conditional detection functions
+#' (p_j(y)) for both observers (j=1,2).  They are conditional detection
+#' functions because detection probability for observer j is based on seeing or
+#' not seeing observations made by observer 3-j. Thus, p_1(y) is estimated by
+#' p_1|2(y).
+#'
+#' If detections by the observers are independent (full
+#' independence) then p_1(y)=p_1|2(y),p_2(y)=p_2|1(y) and for the union, full
+#' independence means that p(y)=p_1(y) + p_2(y) - p_1(y)*p_2(y) for each
+#' distance y.  In fitting the detection functions the likelihood given by eq
+#' 6.8 and 6.16 in Laake and Borchers (2004) is used. That analysis does not
+#' require the usual distance sampling assumption that perpendicular distances
+#' are uniformly distributed based on line placement that is random relative to
+#' animal distribution.  However, that assumption is used in computing
+#' predicted detection probability which is averaged based on a uniform
+#' distribution (see eq 6.11 of Laake and Borchers 2004).
+#'
+#' For a complete description of each of the calling arguments, see
+#' \code{\link{ddf}}.  The argument \code{model} in this function is the same
+#' as \code{mrmodel} in \code{ddf}.  The argument \code{dataname} is the name
+#' of the dataframe specified by the argument \code{data} in \code{ddf}. The
+#' arguments \code{control},\code{meta.data},and \code{method} are defined the
+#' same as in \code{ddf}.
+#'
+#' @export
+#' @method ddf io.fi
+#' @param model mark-recapture model specification
+#' @param data analysis dataframe
+#' @param meta.data list containing settings controlling data structure
+#' @param control list containing settings controlling model fitting
+#' @param call original function call used to call \code{ddf}
[…]

### 2018-01-10 mrds/R/ddf.trial.R

@@ -1,0 +1,126 @@
+#' Mark-Recapture Distance Sampling (MRDS) Trial Configuration - PI
+#'
+#' Mark-Recapture Distance Sampling (MRDS) Analysis of Trial Observer
+#' Configuration and Point Independence
+#'
+#' MRDS analysis based on point independence involves two separate and
+#' independent analyses of the mark-recapture data and the distance sampling
+#' data.  For the trial configuration, the mark-recapture data are analysed
+#' with a call to \code{\link{ddf.trial.fi}} (see likelihood eq 6.12 and 6.17
+#' in Laake and Borchers 2004) to fit a conditional distance sampling detection
+#' function for observer 1 based on trials (observations) from observer 2 to
+#' estimate p_1(0), detection probability at distance zero for observer 1.
+#' Independently, the distance data from observer 1 are used to fit a
+#' conventional distance sampling (CDS) (likelihood eq 6.6) or multi-covariate
+#' distance sampling (MCDS) (likelihood eq 6.14) model for the detection
+#' function, g(y), such that g(0)=1.  The detection function for observer 1 is
+#' then created as p_1(y)=p_1(0)*g(y) (eq 6.28 of Laake and Borchers 2004) from
+#' which predictions are made. \code{ddf.trial} is not called directly by the
+#' user and is called from \code{\link{ddf}} with \code{method="trial"}.
+#'
+#' For a complete description of each of the calling arguments, see
+#' \code{\link{ddf}}.  The argument \code{dataname} is the name of the
+#' dataframe specified by the argument \code{data} in \code{ddf}. The arguments
+#' \code{dsmodel}, \code{mrmodel}, \code{control} and \code{meta.data} are
+#' defined the same as in \code{ddf}.
+#'
+#' @export
+#' @method ddf trial
+#' @param dsmodel distance sampling model specification; model list with key
+#'   function and scale formula if any
+#' @param mrmodel mark-recapture model specfication; model list with formula
+#'   and link
+#' @param data analysis \code{data.frame}
+#' @param meta.data list containing settings controlling data structure
+#' @param control list containing settings controlling model fitting
+#' @param call original function call used to call \code{ddf}
+#' @return result: a trial model object which is composed of \code{trial.fi} and \code{ds} model objects
+#' @author Jeff Laake
[…]

### 2018-01-10 mrds/R/det.tables.R

@@ -1,0 +1,126 @@
+#' Observation detection tables
+#'
+#' Creates a series of tables for dual observer data that shows the number
+#' missed and detected for each observer within defined distance classes.
+#'
+#' @aliases det.tables
+#' @param model fitted model from \code{ddf}
+#' @param nc number of equal-width bins for histogram
+#' @param breaks user define breakpoints
+#' @return list object of class "det.tables" \item{Observer1}{table for observer 1}
+#' \item{Observer2}{table for observer 2} \item{Duplicates}{histogram counts for duplicates}
+#' \item{Pooled}{histogram counts for all observations by either observer}
+#' \item{Obs1_2}{table for observer 1 within subset seen by observer 2}
+#' \item{Obs2_1}{table for observer 2 within subset seen by observer 1}
+#' @author Jeff Laake
+#' @export
+#' @examples
+#' \donttest{
+#' data(book.tee.data)
+#' region<<-book.tee.data$book.tee.region
+#' egdata<<-book.tee.data$book.tee.dataframe
+#' samples<<-book.tee.data$book.tee.samples
+#' obs<<-book.tee.data$book.tee.obs
+#' xx=ddf(mrmodel=~glm(formula=~distance*observer),
+#'  dsmodel = ~mcds(key = "hn", formula = ~sex), data = egdata, method = "io",
+#' meta.data = list(width = 4))
+#' tabs=det.tables(xx,breaks=c(0,.5,1,2,3,4))
+#' par(mfrow=c(2,2))
+#' plot(tabs,new=FALSE,which=c(1,2,5,6))
+#' }
+det.tables <- function(model, nc=NULL, breaks=NULL){
+
+  # do nothing if we have single observer data
+  if(model$method=="ds"){
+    stop("This function does not work with single observer data.\n")
+  }
+
+  xmat <- process.data(model$data,model$meta.data)$xmat
+  xmat$Detected <- factor(xmat$detected,labels=c("Missed","Detected"))
+  left <- model$meta.data$left
+  width <- model$meta.data$width
+
+  # extract the method
+  this.method <- substr(model$method,1,2)
+
+  # Set up default number of classes unless specified
+  if(is.null(nc)){
+    if(this.method == "io"){
+      nc<-round(sqrt(min(
+           length(xmat$distance[xmat$observer==1&xmat$detected==1]),
[…]

### 2018-01-10 mrds/R/detfct.R

@@ -1,0 +1,261 @@
+#'  Detection functions
+#'
+#' Various functions used to specify key and adjustment functions for
+#' detection functions.
+#'
+#' Multi-covariate detection functions (MCDS) are represented by a function
+#' \eqn{g(x,w,\theta)} where x is distance, z is a set of covariates and
+#' \eqn{\theta} is the parameter vector.  The functions are defined such that
+#' \eqn{g(0,w,\theta)=1} and the covariates modify the scale \eqn{(x/\sigma)}
+#' where a log link is used to relate \eqn{\sigma} to the covariates,
+#' \eqn{\sigma=exp(\theta*w)}. A CDS function is obtained with a constant
+#' \eqn{\sigma} which is equivalent to an intercept  design matrix, z.
+#'
+#' \code{detfct} will call either a gamma, half-normal, hazard-rate or uniform
+#' function only returning the probability of detection at that distance. In
+#' addition to the simple model above, we may specify adjustment terms to fit
+#' the data better. These adjustments are either Cosine, Hermite and simple
+#' polynomials.
+#' These are specified as arguments to \code{detfct}, as detailed below.
+#'
+#' \code{detfct} function which calls the others and assembles the final result using either key(x)[1+series(x)] or (key(x)[1+series(x)])/(key(0)[1+series(0)]) (depending on the value of \code{standardize}).
+#'
+#' \code{keyfct.*} functions calculate key function values and \code{adjfct.*} calculate adjustment term values.
+#'
+#' \code{scalevalue} for either detection function it computes the scale with
+#' the log link using the parameters and the covariate design matrix
+#'
+#' \code{fx}, \code{fr} non-normalized probability density for line transects and point
+#' counts respectively
+#'
+#' @aliases detfct adjfct.cos adjfct.herm hermite.poly adjfct.poly keyfct.hn
+#'  keyfct.hz keyfct.gamma scalevalue fx fr distpdf
+#' @usage detfct(distance, ddfobj, select=NULL, index=NULL, width=NULL,
+#'               standardize = TRUE, stdint=FALSE, left=0)
+#'
+#' adjfct.cos(distance, scaling = 1, adj.order, adj.parm = NULL, adj.exp=FALSE)
+#'
[…]

### 2018-01-10 mrds/R/detfct.fit.R

@@ -1,0 +1,197 @@
+#' Fit detection function using key-adjustment functions
+#'
+#' Fit detection function to observed distances using the key-adjustment
+#' function approach.  If adjustment functions are included it will alternate
+#' between fitting parameters of key and adjustment functions and then all
+#' parameters much like the approach in the CDS and MCDS Distance FORTRAN code.
+#' To do so it calls \code{detfct.fit.opt} which uses the R optim function
+#' which does not allow non-linear constraints so inclusion of adjustments does
+#' allow the detection function to be non-monotone.
+#'
+#' @aliases detfct.fit
+#' @param ddfobj detection function object
+#' @param optim.options control options for optim
+#' @param bounds bounds for the parameters
+#' @param misc.options miscellaneous options
+#' @return fitted detection function model object with the following list
+#'   structure \item{par}{final parameter vector} \item{value}{final negative
+#'   log likelihood value} \item{counts}{number of function evaluations}
+#'   \item{convergence}{see codes in optim} \item{message}{string about
+#'   convergence} \item{hessian}{hessian evaluated at final parameter values}
+#'   \item{aux}{ a list with 20 elements \itemize{ \item maxit: maximum number
+#'   of iterations allowed for optimization \item lower: lower bound values for
+#'   parameters \item upper: upper bound values for parameters \item setlower:
+#'   TRUE if they are user set bounds \item setupper: TRUE if they are user set
+#'   bounds \item point: TRUE if point counts and FALSE if line transect \item
+#'   int.range: integration range values \item showit: integer value that
+#'   determines information printed during iteration \item integral.numeric
+#'   if TRUE compute logistic integrals numerically \item
+#'   breaks: breaks in distance for defined fixed bins for analysis \item
+#'   maxiter: maximum iterations used \item refit: if TRUE, detection function
+#'   will be fitted more than once if parameters are at a boundary or when
+#'   convergence is not achieved \item nrefits: number of refittings \item
+#'   \item mono: if TRUE montonicity will be enforced \item
+#'   mono.strict: if TRUE, then strict monotonicity is enforced; otherwise weak
+#'   \item width: radius of point count or half-width of strip \item
[…]

### 2018-01-10 mrds/R/detfct.fit.opt.R

@@ -1,0 +1,407 @@
+#' Fit detection function using key-adjustment functions
+#'
+#' Fit detection function to observed distances using the key-adjustment
+#' function approach. If adjustment functions are included it will alternate
+#' between fitting parameters of key and adjustment functions and then all
+#' parameters much like the approach in the CDS and MCDS Distance FORTRAN code.
+#' This function is called by the driver function \code{detfct.fit}, then
+#' calls \code{\link{optimx}} function.
+#'
+#' @import optimx Rsolnp
+#' @aliases detfct.fit.opt
+#' @param ddfobj detection function object
+#' @param optim.options control options for optim
+#' @param bounds bounds for the parameters
+#' @param misc.options miscellaneous options
+#' @param fitting character string with values "all","key","adjust" to
+#'   determine which parameters are allowed to vary in the fitting
+#' @return fitted detection function model object with the following list
+#'   structure \item{par}{final parameter vector} \item{value}{final negative
+#'   log likelihood value} \item{counts}{number of function evaluations}
+#'   \item{convergence}{see codes in optim} \item{message}{string about
+#'   convergence} \item{hessian}{hessian evaluated at final parameter values}
+#'   \item{aux}{ a list with 20 elements \itemize{ \item maxit: maximum number
+#'   of iterations allowed for optimization \item lower: lower bound values for
+#'   parameters \item upper: upper bound values for parameters \item setlower:
+#'   TRUE if they are user set bounds \item setupper: TRUE if they are user set
+#'   bounds \item point: TRUE if point counts and FALSE if line transect \item
+#'   int.range: integration range values \item showit: integer value that
+#'   determines information printed during iteration \item integral.numeric
+#'   if TRUE compute logistic integrals numerically \item
+#'   breaks: breaks in distance for defined fixed bins for analysis \item
+#'   maxiter: maximum iterations used \item refit: if TRUE, detection function
+#'   will be fitted more than once if parameters are at a boundary or when
+#'   convergence is not achieved \item nrefits: number of refittings \item
+#'   \item mono: if TRUE, montonicity will be enforced \item
+#'   mono.strict: if TRUE, then strict monotonicity is enforced; otherwise weak
+#'   \item width: radius of point count or half-width of strip \item
[…]

### 2018-01-10 mrds/R/dht.R

@@ -1,0 +1,470 @@
+#' Density and abundance estimates and variances
+#'
+#' Compute density and abundance estimates and variances based on Horvitz-Thompson-like estimator.
+#'
+#' Density and abundance within the sampled region is computed based on a Horvitz-Thompson-like estimator for groups and individuals (if a clustered population) and this is extrapolated to the entire survey region based on any defined regional stratification. The variance is based on replicate samples within any regional stratification. For clustered populations, \eqn{E(s)} and its standard error are also output.
+#'
+#' Abundance is estimated with a Horvitz-Thompson-like estimator (Huggins 1989, 1991; Borchers et al 1998; Borchers and Burnham 2004). The abundance in the sampled region is simply \eqn{1/p_1 + 1/p_2 + ... + 1/p_n} where \eqn{p_i} is the estimated detection probability for the \eqn{i}th detection of \eqn{n} total observations. It is not strictly a Horvitz-Thompson estimator because the \eqn{p_i} are estimated and not known. For animals observed in tight clusters, that estimator gives the abundance of groups (\code{group=TRUE} in \code{options}) and the abundance of individuals is estimated as \eqn{s_1/p_1 + s_2/p_2 + ... + s_n/p_n}, where \eqn{s_i} is the size (e.g., number of animals in the group) of each observation (\code{group=FALSE} in \code{options}).
+#'
+#' Extrapolation and estimation of abundance to the entire survey region is based on either a random sampling design or a stratified random sampling design. Replicate samples (lines) are specified within regional strata \code{region.table}, if any. If there is no stratification, \code{region.table} should contain only a single record with the \code{Area} for the entire survey region. The \code{sample.table} is linked to the \code{region.table} with the \code{Region.Label}. The \code{obs.table} is linked to the \code{sample.table} with the \code{Sample.Label} and \code{Region.Label}. Abundance can be restricted to a subset (e.g., for a particular species) of the population by limiting the list the observations in \code{obs.table} to those in the desired subset. Alternatively, if \code{Sample.Label} and \code{Region.Label} are in the \code{data.frame} used to fit the model, then a \code{subset} argument can be given in place of the \code{obs.table}. To use the \code{subset} argument but include all of the observations, use \code{subset=1==1} to avoid creating an \code{obs.table}.
+#'
[…]

### 2018-01-10 mrds/R/dht.deriv.R

@@ -1,0 +1,69 @@
+#' Computes abundance estimates at specified parameter values using
+#' Horvitz-Thompson-like estimator
+#'
+#' Computes abundance at specified values of parameters for numerical
+#' computation of first derivative with respect to parameters in detection
+#' function.  An internal function called by DeltaMethod which is invoked by
+#' dht.se
+#'
+#' @param par detection function parameter values
+#' @param model ddf model object
+#' @param obs observations table
+#' @param samples samples table
+#' @param options list of options as specified in \code{\link{dht}}
+#' @return vector of abundance estimates at values of parameters specified in
+#'   par
+#' @note Internal function; not intended to be called by user
+#' @author Jeff Laake
+#' @seealso \code{\link{dht}}, \code{\link{dht.se}}, \code{\link{DeltaMethod}}
+#' @keywords utility
+dht.deriv <- function(par,model,obs,samples,options=list()){
+  # Functions Used:  predict, covered.region.dht, survey.region.dht
+
+  #  Depending on model method store new parameter values
+  #  uses coefficients instead of coef; needed for trial.fi and io.fi 
+  model$par <- par
+  if(model$method!="ds"){
+    if(model$method%in%c("io","trial","rem")){
+      model$mr$mr$coefficients <- model$par[1:length(model$mr$mr$coefficients)]
+      model$ds$par <-model$par[(length(model$mr$mr$coefficients)+1):length(par)]
+    }else{
+      model$mr$coefficients <- model$par
+    }
+  }
+
+  # Get new predicted detection probabilities and put with observations
+  # 1/1/05 jll; change made to fi methods. need integrate=FALSE
+  # 1/24/06 jll; must have been on glue when I made the 1/1/05 change; clearly
+  # not what I wanted.  You would only not integrate if you didn't want to
+  # assume 1/w; otherwise, always want to integrate to get average detection
+  # probability
+  pdot <- predict(model,integrate=TRUE,compute=TRUE)
+
+  if(!is.null(pdot$fitted)){
+    pdot <- pdot$fitted
+  }
+
+  # probably should stop here if pdot$fitted is null? anyway...
+  # code below would not work if there is a pdot column there already, so remove
+  obs$pdot <- NULL
+
+  obs <- merge(obs,data.frame(object=as.numeric(names(model$fitted)),pdot=pdot))
+
+  # Compute covered region abundances by sample depending on value of group
[…]

### 2018-01-10 mrds/R/dht.se.R

@@ -1,0 +1,296 @@
+#' Variance and confidence intervals for density and abundance estimates
+#'
+#' Computes standard error, cv, and log-normal confidence intervals for abundance and density within each region (if any) and for the total of all the regions. It also produces the correlation matrix for regional and total estimates.
+#'
+#' The variance has two components:
+#' \itemize{
+#'   \item variation due to uncertainty from estimation of the detection function parameters;
+#'   \item variation in abundance due to random sample selection;
+#' }
+#' The first component (model parameter uncertainty) is computed using a delta method estimate of variance (Huggins 1989, 1991, Borchers et al. 1998) in which the first derivatives of the abundance estimator with respect to the parameters in the detection function are computed numerically (see \code{\link{DeltaMethod}}).
+#'
+#' The second component (encounter rate variance) can be computed in one of several ways depending on the form taken for the encounter rate and the estimator used. To begin with there three possible values for \code{varflag} to calculate encounter rate:
+#' \itemize{
+#'  \item \code{0} uses a binomial variance for the number of observations (equation 13 of Borchers et al. 1998). This estimator is only useful if the sampled region is the survey region and the objects are not clustered; this situation will not occur very often;
+#'  \item \code{1} uses the encounter rate \eqn{n/L} (objects observed per unit transect) from Buckland et al. (2001) pg 78-79 (equation 3.78) for line transects (see also Fewster et al, 2009 estimator R2). This variance estimator is not appropriate if \code{size} or a derivative of \code{size} is used in the detection function;
+#'  \item \code{2} is the default and uses the encounter rate estimator \eqn{\hat{N}/L} (estimated abundance per unit transect) suggested by Innes et al (2002) and Marques & Buckland (2004).
+#' }
+#'
+#' In general if any covariates are used in the models, the default \code{varflag=2} is preferable as the estimated abundance will take into account variability due to covariate effects. If the population is clustered the mean group size and standard error is also reported.
+#'
[…]

### 2018-01-10 mrds/R/ds.function.r

@@ -1,0 +1,114 @@
+#' Distance Sampling Functions
+#'
+#' Computes values of conditional and unconditional detection functions and
+#' probability density functions for for line/point data for single observer or
+#' dual observer in any of the 3 configurations (io,trial,rem).
+#'
+#' Placeholder -- Not functional ----
+#'
+#' @param model model object
+#' @param newdata dataframe at which to compute values; if NULL uses fitting data
+#' @param obs 1 or 2 for observer 1 or 2, 3 for duplicates, "." for combined and "All" to return all of the values
+#' @param conditional if FALSE, computes p(x) based on distance detection function and if TRUE based on mr detection function
+#' @param pdf if FALSE, returns p(x) and if TRUE, returns p(x)*pi(x)/integral p(x)*pi(x)
+#' @param finebr fine break values over which line is averaged
+#' @return List containing \item{xgrid }{grid of distance values} \item{values }{average detection fct values at the xgrid values}
+#' @author Jeff Laake
+ds.function <- function(model,newdata=NULL,obs="All",conditional=FALSE,
+                        pdf=TRUE,finebr){
+
+  xgrid <- NULL
+  linevalues <- NULL
+
+  # Depending on the type of model, setup data to be used for prediction
+  # selecting the specified observer and only those detected (detected=1).
+  if(model$method=="io"){
+    newdat <- model$mr$mr$data
+  }else if(model$method=="trial" | model$method=="trial.fi"){
+    newdat <- process.data(model$data,model$meta.data)$xmat
+    newdat <- newdat[newdat$observer==obs & newdat$detected==1,]
+  }else{
+    newdat <- model$mr$data
+  }
+
+  newdat$offsetvalue <- rep(0, dim(newdat)[1])
+
+  # Also, depending on the type of model, get det fct predicted values
+  # Note g0 = p(0) and prob.det are integrated values (mu)
+  if(model$method=="io" | model$method=="trial"| model$method=="rem"){
+    prob.det <- predict(model)$fitted
+    newdat$distance <- 0
+    ddfobj <- model$ds$ds$aux$ddfobj
+
+    if(ddfobj$type=="gamma"){
+      if(model$method=="io"){
+        newdat$distance <- rep(apex.gamma(ddfobj),2)
+      }else{
+        newdat$distance <- as.vector(apex.gamma(ddfobj))
[…]

### 2018-01-10 mrds/R/flnl.R

@@ -1,0 +1,76 @@
+#' Log-likelihood computation for distance sampling data
+#'
+#' For a specific set of parameter values, it computes and returns the negative
+#' log-likelihood for the distance sampling likelihood for distances that are
+#' unbinned, binned and a mixture of both.  The function \code{flnl} is the
+#' function minimized using \code{\link{optim}} from within
+#' \code{\link{ddf.ds}}.
+#'
+#' Most of the computation is in \code{flpt.lnl} in which the negative
+#' log-likelihood is computed for each observation. \code{flnl} is a wrapper
+#' that optionally outputs intermediate results and sums the individual
+#' log-likelihood values.
+#'
+#' \code{flnl} is the main routine that manipulates the parameters using
+#' \code{\link{getpar}} to handle fitting of key, adjustment or all of the
+#' parameters.  It then calls \code{flpt.lnl} to do the actual computation of
+#' the likelihood.  The probability density function for point counts is
+#' \code{fr} and for line transects is \code{fx}.
+#' \code{fx}=g(x)/mu (where g(x) is the detection function); whereas,
+#' f(r)=r*g(r)/mu where mu in both cases is the normalizing constant.  Both
+#' functions are in source code file for \code{link{detfct}} and are called from
+#' \code{distpdf} and the integral calculations are made with
+#' \code{\link{integratepdf}}.
+#'
+#' @aliases flnl flpt.lnl
+#' @param fpar parameter values for detection function at which negative log-likelihood should be evaluated
+#' @param ddfobj distance sampling object
+#' @param misc.options a \code{list} with the following elements: \code{width} transect width; \code{int.range} the integration range for observations; \code{showit} 0 to 3 controls level debug output; \code{integral.numeric} if \code{TRUE} integral is computed numerically rather than analytically; \code{point} is this a point transect?
+#' @param fitting character \code{"key"} if only fitting key function parameters, \code{"adjust"} if fitting adjustment parameters or \code{"all"} to fit both
+#' @return negative log-likelihood value at the parameter values specified in \code{fpar}
[…]

### 2018-01-10 mrds/R/flpt.lnl.r

@@ -1,0 +1,181 @@
+flpt.lnl <- function(fpar, ddfobj, misc.options){
+
+  # flpt.lnl - computes negative log-likelihood values of line or point
+  # transect binned/unbinned distances
+  #
+  # Arguments: see flnl for description of arguments
+  #
+  # value: vector of negative log-likelihood values for observations at current
+  #        set of parameters (fpar)
+  #
+  # Functions Used: assign.par, tablecgf, predict(predict.smooth.spline),
+  # detfct, integratedetfct, integratedetfct.logistic
+
+  # Assign parameter values into ddfobj
+  ddfobj <- assign.par(ddfobj, fpar)
+
+  # Setup integration ranges
+  int.range <- misc.options$int.range
+  if(is.vector(int.range)){
+    int.range <- matrix(int.range, nrow=1)
+    samelimits <- TRUE
+  }else{
+    #int.range <- int.range[2:nrow(int.range), , drop=FALSE]
+    samelimits <- FALSE
+  }
+  left <- int.range[, 1]
+  right <- int.range[, 2]
+
+  x <- ddfobj$xmat
+  z <- ddfobj$scale$dm
+  if(is.null(z)){
+    z <- matrix(1, nrow=nrow(x), ncol=1)
+  }
+
+  width <- misc.options$width
+
+  # Compute log-likelihood for any binned data
+  lnl <- rep(0, dim(x)[1])
+  if(any(x$binned)){
+    # Get bins and create unique set of bins/covariates and indices 
+    # (int.index) in that set
+    bins <- as.matrix(x[x$binned, c("distbegin", "distend")])
+    allbins <- apply(cbind(bins, z[x$binned, , drop=FALSE]), 1,
+                     paste, collapse="")
+
+    uniquevals <- !duplicated(allbins)
+    uniquebins <- bins[uniquevals, , drop=FALSE]
+
+    int.index <- match(allbins,
+                       apply(cbind(bins,
+                              z[x$binned, , drop=FALSE])[uniquevals, , drop=FALSE],
+                              1, paste, collapse=""))
+
+    which.obs <- x$binned
+    which.obs[!uniquevals] <- FALSE
+    int.bin <- integratepdf(ddfobj, select=which.obs, width=width,
+                            int.range=uniquebins,
+                            standardize=FALSE, point=misc.options$point,
+                            left=misc.options$left)
+
+    if(any(int.bin<0)){
+      int.bin <- integratepdf(ddfobj, select=which.obs, width=width,
+                              int.range=uniquebins,
[…]

### 2018-01-10 mrds/R/flt.var.R

@@ -1,0 +1,48 @@
+#' Hessian computation for fitted distance detection function model parameters
+#'
+#' Computes hessian to be used for variance-covariance matrix.  The hessian is
+#' the outer product of the vector of first partials (see pg 62 of Buckland et
+#' al 2002).
+#'
+#' @param ddfobj distance sampling object
+#' @param misc.options width-transect width (W); int.range-integration range
+#'   for observations; showit-0 to 3 controls level of iteration printing;
+#'   integral.numeric-if TRUE integral is computed numerically rather
+#'   than analytically
+#' @return variance-covariance matrix of parameters in the detection function
+#' @note This is an internal function used by \code{\link{ddf.ds}} to fit
+#'   distance sampling detection functions.  It is not intended for the user to
+#'   invoke this function but it is documented here for completeness.
+#' @author Jeff Laake
+#' @seealso \code{\link{flnl}},\code{\link{flpt.lnl}},\code{\link{ddf.ds}}
+#' @references Buckland et al. 2002
+#' @keywords utility
+flt.var <- function(ddfobj, misc.options){
+
+  fpar1 <- getpar(ddfobj)
+  fpar <- fpar1
+  parmat <- NULL
+
+  #   Compute first partial (numerically) of log(f(y)) for each observation
+  #   for each parameter and store in parmat (n by length(fpar))
+  for(i in seq_along(fpar)){
+    deltap <- .0001*fpar1[i]
+    if(deltap==0) deltap <- 0.0001
+    fpar[i] <- fpar1[i]- deltap
+    x1 <- -flpt.lnl(fpar, ddfobj,misc.options)
+    fpar[i] <- fpar1[i]+deltap
+    x2 <- -flpt.lnl(fpar, ddfobj,misc.options)
+    parmat <- cbind(parmat,(x2-x1)/(2*deltap))
+  }
+
+  # Compute varmat using first partial approach (pg 62 of Buckland et al 2002)
+  varmat <- matrix(0,ncol=length(fpar1),nrow=length(fpar1))
+
+  for(i in seq_along(fpar1)){
+    for(j in seq_along(fpar1)){
+      varmat[i,j] <- sum(parmat[,i]*parmat[,j])
+    }
+  }
+
+  return(varmat)
+}

### 2018-01-10 mrds/R/getpar.R

@@ -1,0 +1,78 @@
+#' Extraction and assignment of parameters to vector
+#'
+#' Extracts parameters of a particular type (scale,
+#' shape, adjustments or g0 (p(0))) from the vector of parameters in
+#' \code{ddfobj}. All of the parameters are kept in a single vector for
+#' optimization even though they have very different uses.  \code{assign.par}
+#' parses them from the vector based on a known structure and assigns them into
+#' \code{ddfobj}.  \code{getpar} extracts the requested types to be extracted
+#' from \code{ddfobj}.
+#'
+#' @aliases getpar
+#' @param ddfobj distance sampling object (see \code{\link{create.ddfobj}})
+#' @param fitting character string which is either "all","key","adjust" which
+#'   determines which parameters are retrieved
+#' @param index logical that determines whether parameters are returned (FALSE)
+#'   or starting indices in parameter vector for scale, shape, adjustment
+#'   parameters
+#' @return index==FALSE, vector of parameters that were requested or
+#'   index==TRUE, vector of 3 indices for shape, scale, adjustment
+#' @note Internal functions not intended to be called by user.
+#' @author Jeff Laake
+#' @seealso assign.par
+#' @keywords utility
+getpar <- function(ddfobj,fitting="all",index=FALSE){
+
+  pars <- NULL
+
+  if(!index){
+    if(fitting=="key" | fitting=="all"){
+      if(!is.null(ddfobj$shape)) pars <- c(pars,ddfobj$shape$parameters)
+      if(!is.null(ddfobj$scale)) pars <- c(pars,ddfobj$scale$parameters)
+    }else{
+      if(!is.null(ddfobj$shape)){
+        pars <- c(pars,rep(NA,length(ddfobj$shape$parameters)))
+      }
+      if(!is.null(ddfobj$scale)){
+        pars <- c(pars,rep(NA,length(ddfobj$scale$parameters)))
+      }
+    }
+
+    if(fitting=="adjust" | fitting=="all"){
+      if(!is.null(ddfobj$adjustment)){
+        pars <- c(pars,ddfobj$adjustment$parameters)
+      }
+    }else{
+      if(!is.null(ddfobj$adjustment)){
+        pars <- c(pars,rep(NA,length(ddfobj$adjustment$parameters)))
+      }
+    }
+
+    return(pars)
+
+  }else{
+
+    # WHAT IS GOING ON HERE?!
[…]

### 2018-01-10 mrds/R/gof.ds.R

@@ -1,0 +1,58 @@
+#' Compute chi-square goodness-of-fit test for ds models
+#'
+#' @param model \code{ddf} model object
+#' @param breaks distance cut points
+#' @param nc number of distance classes
+#'
+#' @return list with chi-square value, df and p-value
+#' @seealso ddf.gof
+#' @author Jeff Laake
+#' @importFrom stats pchisq
+gof.ds <- function(model, breaks=NULL, nc=NULL){
+  # Functions used: predict.ds
+
+  width <- model$meta.data$width
+  left <- model$meta.data$left
+  xmat <- model$ds$aux$ddfobj$xmat
+  n <- dim(xmat)[1]
+
+  # If breaks are given use those; otherwise create them
+  if(is.null(breaks)){
+    breaks <- model$ds$aux$breaks
+  }
+  if(is.null(nc)){
+    nc <- round( sqrt(length(xmat$distance)))
+  }
+  if(is.null(breaks)){
+    breaks <- left + ((width-left)/nc)*(0:nc)
+  }else{
+    nc <- length(breaks)-1
+  }
+
+  # Get predicted values for ds component
+  expected.1 <- rep(0,nc)
+  for(j in 1:nc){
+    expected.1[j] <- sum(predict(model, compute=TRUE, integrate=TRUE,
+                                 int.range=matrix(c(breaks[j], breaks[j+1]),
+                                 nrow=1))$fitted/model$fitted,
+                         na.rm=TRUE)
+  }
+
+  # Compute observed values of distance bins
+  observed.count.1 <- table(cut(xmat$distance,breaks,include.lowest=TRUE))
+  chisq.1 <- sum((observed.count.1-expected.1)^2/expected.1,na.rm=TRUE)
+  df.1 <- nc-1-length(model$par)
+  if(df.1 <= 0){
+    df.1 <- NA
+    p.1 <- NA
+  }else{
+    p.1 <- 1-pchisq(chisq.1,df.1)
+  }
+
+  # build and return list
+  return(list(chi1=list(observed=observed.count.1,
+                        expected=expected.1,
+                        chisq=chisq.1,
+                        p=p.1,
+                        df=df.1)))
+}

### 2018-01-10 mrds/R/gof.io.R

@@ -1,0 +1,102 @@
+# gof.io - Computes chi-square gof test for io models
+#
+# Arguments:
+#   model  - ddf model object
+#   breaks - distance cut points
+#   nc     - number of distance classes
+# Value:
+#   result - list with chi-square value, df and p-value
+gof.io <- function(model, breaks=NULL, nc=NULL){
+
+  width <- model$meta.data$width
+  left <- model$meta.data$left
+  xmat <- model$mr$mr$data
+  n <- dim(xmat)[1]/2
+
+  # Set up omega index; 1 - detected by primary only,
+  #                     2 - detected by secondary only,
+  #                     3 - detected by both
+  xmat <- xmat[xmat$observer==1,]
+  xmat$omega <- rep(1,dim(xmat)[1])
+  xmat$omega[xmat$timesdetected==2] <- 3
+  xmat$omega[xmat$timesdetected==1&xmat$detected==0] <- 2
+
+  # If number of classes for histogram intervals was not set
+  # compute a reasonable default
+  if(is.null(nc)){
+    nc<-round(sqrt(min(length(xmat$distance[xmat$observer==1 &
+                                            xmat$detected==1]),
+                       length(xmat$distance[xmat$observer==1 &
+                                            xmat$timesdetected==2]) )), 0)
+  }
+
+  # Set up default break points - need to allow user-defined values
+  if(is.null(breaks)){
+    breaks <- left + ((width-left)/nc)*(0:nc)
+  }else{
+    nc <- length(breaks)-1
+  }
+
+  # Get predicted values for mr component
+  predict.list <- predict(model$mr)
+  p1 <- predict.list$p1
+  p2 <- predict.list$p2
+  p.omega <- data.frame(object   = rep(1:n,3),
+                        omega    = c(rep(1,n),rep(2,n),rep(3,n)),
+                        distance = rep(xmat$distance,3),
+                        prob     = rep(0,3*n))
+  p.omega$prob[p.omega$omega==1] <- p1*(1-p2)/(p1+p2-p1*p2)
+  p.omega$prob[p.omega$omega==2] <- p2*(1-p1)/(p1+p2-p1*p2)
+  p.omega$prob[p.omega$omega==3] <- p1*p2/(p1+p2-p1*p2)
+  expected.2 <- by(p.omega$prob,
+                   list(as.factor(p.omega$omega),
+                        cut(p.omega$distance,breaks,include.lowest=TRUE)),
+                        sum,na.rm=TRUE)
+
[…]

### 2018-01-10 mrds/R/gof.pvalues.R

@@ -1,0 +1,68 @@
+# Goodness of fit testing for detection functions
+#
+# See ddf.gof
+#
+# @param model a fitted model object
+# @param nboot number of bootstraps to do
+# @param ks logical whether to calculate Kolmogorov-Smirnov test statistic
+# @param progress logical whether to display progress
+#
+# @return list with p-values for the two tests (\code{ks.p}, \code{cramer.p}) an dtest statistics (\code{Dn} for K-S and \code{W} for C-vM).
+#' @importFrom utils setTxtProgressBar txtProgressBar
+# @author David L Miller
+#
+gof.pvalues <- function(model, nboot=500, ks=FALSE, progress=FALSE){
+  
+  # storage
+  ks.boot <- rep(0, nboot)
+  cramer.boot <- rep(0, nboot)
+  
+  if(progress) pb <- txtProgressBar(0, nboot)
+  
+  boot_success <- 0
+  
+  # now calculate test statistics for the model
+  model_edf_cdf <- get_edf_cdf(model)
+  model_Dn <- max(c(abs(model_edf_cdf$lower.edf - model_edf_cdf$cdfvalues),
+                    abs(model_edf_cdf$upper.edf - model_edf_cdf$cdfvalues)))
+  model_W <- 1/(12*model_edf_cdf$n) + sum((model_edf_cdf$cdfvalues - ((1:model_edf_cdf$n)-.5)/model_edf_cdf$n)^2)
+  
+  # can skip the bootstraps if we want
+  if(ks && nboot>0){
+    for (i in 1:nboot) {
+      # simulate data from the model
+      # fit a model
+      refit <- sample_ddf(model)
+      
+      if(!inherits(refit, "try-error")){
+        ## calculate test statistics and store them
+        edf_cdf <- get_edf_cdf(refit)
+        # Kolmogorov-Smirnov test statistic
+        ks.boot[i] <- max(c(abs(model_edf_cdf$lower.edf - edf_cdf$cdfvalues),
+                            abs(model_edf_cdf$upper.edf - edf_cdf$cdfvalues)))
+        boot_success <- boot_success + 1
+      }
+      if(progress) setTxtProgressBar(pb, i)
+    }
+    
+    # do the boostrap test
+    ks.p <- mean(model_Dn <= ks.boot, na.rm=TRUE)
+  }else{
+    ks.p <- NA
+    model_Dn <- NA
+  }
+  
+  # do regular Cramer-von Mises thingo
[…]

### 2018-01-10 mrds/R/gof.rem.fi.R

@@ -1,0 +1,87 @@
+# Compute chi-square gof test for rem.fi models
+#
+# model ddf model object
+# breaks distance cut points
+# nc number of distance classes
+#
+# return list with chi-square value, df and p-value
+# documented in ?ddf.gof
+gof.rem.fi <- function(model,breaks=NULL,nc=NULL){
+  width <- model$meta.data$width 
+  left <- model$meta.data$left
+  xmat <- model$mr$data
+  n <- nrow(xmat)
+
+  # Set up omega index
+  #  1 - detected by primary only
+  #  2 - detected by secondary only
+  #  3 - detected by both
+  xmat$omega <- rep(2,dim(xmat)[1])
+  xmat$omega[xmat$detected==0] <- 1
+
+  # If number of classes for histogram intervals was not set
+  #  compute a reasonable default
+  if(is.null(nc)){
+    nc<-round( sqrt(min(length(xmat$distance[xmat$detected==1]),
+                length(xmat$distance[xmat$detected==0]) )),0)
+  }
+
+  # Set up default break points - need to allow user-defined values
+  if(is.null(breaks)){
+    breaks <- left + ((width-left)/nc)*(0:nc)
+  }else{
+    nc <- length(breaks)-1
+  }
+
+  # Get predicted values for mr component
+  predict.list <- predict(model)
+  p1 <- predict.list$p1
+  p2 <- predict.list$p2
+  p.omega <- data.frame(object=rep(1:n,2),
+                        omega=c(rep(1,n),rep(2,n)),
+                        distance=rep(xmat$distance,2),
+                        prob=rep(0,2*n))
+  p.omega$prob[p.omega$omega==2] <- p1*1/(p1+p2-p1*p2)
+  p.omega$prob[p.omega$omega==1] <- p2*(1-p1)/(p1+p2-p1*p2)
+  expected.2 <- by(p.omega$prob,list(as.factor(p.omega$omega),
+                                     cut(p.omega$distance,breaks,
+                                         include.lowest=TRUE)),
+                   sum,na.rm=TRUE)
+
+  # Get predicted values for ds component
+  expected.1 <- rep(0,nc)
+  for(j in 1:nc){
+     expected.1[j] <- sum(predict(model, integrate=TRUE, compute=TRUE,
+                                  int.range=breaks[j+1])$fitted/model$fitted)
+  }
+
+  n <- expected.1[nc]
+  expected.1[2:nc] <- expected.1[2:nc]- expected.1[1:(nc-1)]
[…]

### 2018-01-10 mrds/R/gof.trial.R

@@ -1,0 +1,99 @@
+# gof.trial - Computes chi-square gof test for trial models
+#
+# Arguments:
+#   model  - ddf model object
+#   breaks - distance cut points
+#   nc     - number of distance classes
+# Value:
+#   result - lists with observed,expected, chi-square value, df and p-value
+gof.trial <- function(model, breaks=NULL, nc=NULL){
+  width <- model$meta.data$width
+  left <- model$meta.data$left
+  xmat <- model$mr$mr$data
+  data <- eval(model$data)
+  data <- data[data$observer==1&data$object %in%
+            as.numeric(names(model$fitted)),]
+  n <- dim(xmat)[1]
+
+  # Set up omega index; 1 - detected by secondary only, 2 - detected by both
+  xmat$omega <- rep(1,dim(xmat)[1])
+  xmat$omega[xmat$timesdetected==2] <- 2
+
+  # If number of classes for histogram intervals was not set compute
+  # a reasonable default
+  if(is.null(nc)){
+    nc <- round(sqrt(min(length(xmat$distance[xmat$observer==1 &
+                                              xmat$detected==1]),
+                         length(xmat$distance[xmat$observer==1 &
+                                              xmat$timesdetected==2]))), 0)
+  }
+
+  # Set up default break points unless some are specified
+  if(is.null(breaks)){
+    breaks <- left + ((width-left)/nc)*(0:nc)
+  }else{
+    nc <- length(breaks)-1
+  }
+
+  # Get predicted values for mr component
+  xmat$detected <- 1
+  p1 <- predict(model$mr, xmat, compute=TRUE, integrate=FALSE)$fitted
+  p.omega <- data.frame(object   = rep(1:n, 2),
+                        omega    = c(rep(1, n), rep(2, n)),
+                        distance = rep(xmat$distance, 2),
+                        prob     = rep(0, 2*n))
+  p.omega$prob[p.omega$omega==1] <- (1-p1)
+  p.omega$prob[p.omega$omega==2] <- p1
+  expected.2 <- by(p.omega$prob,
+                   list(as.factor(p.omega$omega),
+                        cut(p.omega$distance, breaks, include.lowest=TRUE)),
+                   sum, na.rm=TRUE)
+
+  # Get predicted values for ds component
+  expected.1 <- rep(0,nc)
+  for(j in 1:nc){
+    expected.1[j] <- sum(predict(model, compute=TRUE,
[…]

### 2018-01-10 mrds/R/gof.trial.fi.R

@@ -1,0 +1,85 @@
+# Compute chi-square gof test for trial.fi models
+#
+# model ddf model object
+# breaks distance cut points
+# nc number of distance classes
+#
+# return lists with observed,expected, chi-square value, df and p-value
+# documented in ?ddf.gof
+gof.trial.fi <- function(model,breaks=NULL,nc=NULL){
+  width <- model$meta.data$width 
+  left <- model$meta.data$left
+  xmat <- model$mr$data
+  n <- dim(xmat)[1]
+
+  # Set up omega index; 1 - detected by secondary only, 2 - detected by both
+  xmat$omega <- rep(1,dim(xmat)[1])
+  xmat$omega[xmat$timesdetected==2] <- 2
+
+  # If number of classes for histogram intervals was not set
+  # compute a reasonable default
+  if(is.null(nc)){
+    nc<-round(sqrt(min(length(xmat$distance[xmat$observer==1&xmat$detected==1]),
+                 length(xmat$distance[xmat$observer==1&xmat$timesdetected==2]))),0)
+  }
+
+  # Set up default break points unless specified
+  if(is.null(breaks)){
+    breaks <- left + ((width-left)/nc)*(0:nc)
+  }else{
+    nc <- length(breaks)-1
+  }
+
+  # Get predicted values for mr component
+  xmat$detected <- 1
+  p1 <- predict(model,newdata=xmat,compute=TRUE,integrate=FALSE)$fitted
+  p.omega <- data.frame(object=rep(1:n,2),
+                        omega=c(rep(1,n),rep(2,n)),
+                        distance=rep(xmat$distance,2),
+                        prob=rep(0,2*n))
+  p.omega$prob[p.omega$omega==1] <- (1-p1)
+  p.omega$prob[p.omega$omega==2] <- p1
+  expected.2 <- by(p.omega$prob,list(as.factor(p.omega$omega),
+                                     cut(p.omega$distance,breaks,
+                                         include.lowest=TRUE)),
+                   sum,na.rm=TRUE)
+
+  # Get predicted values for ds component
+  expected.1 <- rep(0,nc)
+  for(j in 1:nc){
+    expected.1[j] <- sum(predict(model,integrate=TRUE,compute=TRUE,
+                                 int.range=breaks[j+1])$fitted/model$fitted)
+  }
+
+  n <- expected.1[nc]
+  expected.1[2:nc] <- expected.1[2:nc]- expected.1[1:(nc-1)]
[…]

### 2018-01-10 mrds/R/gstdint.R

@@ -1,0 +1,152 @@
+#' Integral of pdf of distances
+#'
+#' Computes the integral of \code{distpdf} with scale=1 (\code{stdint=TRUE}) or specified scale (\code{stdint=FALSE}).
+#'
+#' @param x lower, upper value for integration
+#' @param ddfobj distance detection function specification
+#' @param index specific data row index
+#' @param select logical vector for selection of data values
+#' @param width truncation width
+#' @param left left truncation width
+#' @param standardize if \code{TRUE}, divide through by the function evaluated at 0
+#' @param point logical to determine if point (\code{TRUE}) or line transect(\code{FALSE})
+#' @param stdint if \code{TRUE}, scale=1 otherwise specified scale used
+#' @param doeachint if \code{TRUE} perform integration using \code{\link{integrate}}
+#' @return vector of integral values of detection function
+#' @note This is an internal function that is not intended to be invoked directly.
+#' @author Jeff Laake and David L Miller
+#' @keywords utility
+#' @importFrom stats pnorm smooth.spline
+gstdint <- function(x, ddfobj, index=NULL,select=NULL, width,
+                    standardize=TRUE, point=FALSE, stdint=TRUE,
+                    doeachint=FALSE, left=left){
+
+  if(!is.matrix(x)){
+    x <- matrix(x, ncol=2)
+  }
+  ## NB this calculates the integral of:
+  ##       g(x)/w             for line transects
+  ##       2*r*g(r)/width^2   for point transects
+
+  # Set of observations for computation of detection function can
+  # be specified with logical (select) and numeric (index) values.
+  # Either or both can be specified although the latter is unlikely.
+  if(is.null(select)){
+    # use all
+    if(is.null(index)){
+      scale.dm <- ddfobj$scale$dm
+    }else{
+      # use only those with specific indices
+      scale.dm <- ddfobj$scale$dm[index, , drop=FALSE]
+    }
+  }else{
+    # Use those with select=TRUE
+    if(is.null(index)){
+      scale.dm <- ddfobj$scale$dm[select, , drop=FALSE]
+    }else{
+      # use the numeric index within those with select=TRUE
+      scale.dm <- ddfobj$scale$dm[select, , drop=FALSE][index, , drop=FALSE]
+    }
+  }
+
+  # when we have half-normal, key only use the exact analytic expression
[…]

### 2018-01-10 mrds/R/histline.R

@@ -1,0 +1,45 @@
+#' Plot histogram line
+#'
+#' Takes bar heights (height) and cutpoints (breaks), and constructs a
+#' line-only histogram from them using the function plot() (if lineonly==FALSE)
+#' or lines() (if lineonly==TRUE).
+#'
+#' @param height heights of histogram bars
+#' @param breaks cutpoints for x
+#' @param lineonly if TRUE, drawn with plot; otherwise with lines to allow
+#'   addition of current plot
+#' @param outline if TRUE, only outline of histogram is plotted
+#' @param ylim limits for y axis
+#' @param xlab label for x axis
+#' @param ylab label for y axis
+#' @param det.plot if TRUE, plot is of detection so yaxis limited to unit
+#'   interval
+#' @param add should this plot add to a previous window
+#' @param \dots Additional unspecified arguments for plot
+#' @return None
+#' @author Jeff Laake and David L Miller
+#' @importFrom graphics plot box
+histline <- function(height, breaks, lineonly=FALSE, outline=FALSE,
+                     ylim=range(height), xlab="x", ylab="y", det.plot=FALSE,
+                     add=FALSE,...){
+
+  # make hist object
+  hh <- list()
+  hh$breaks <- breaks
+  hh$counts <- height
+  hh$density <- height#/sum(height)
+  hh$mids <- breaks[-length(breaks)] + diff(breaks)/2
+  hh$xname <- "hh"
+  hh$equidist <- FALSE
+  class(hh) <- "histogram"
+
+  # make the plot
+  if(det.plot){
+    plot(hh, ylim=ylim, xlab=xlab, ylab=ylab, yaxp=c(0, 1, 5), main="",
+         add=add, freq=FALSE, ...)
+  }else{
+    plot(hh, ylim=ylim, xlab=xlab, ylab=ylab, main="", add=add, freq=FALSE, ...)
+  }
+  # put a box around
+  box()
+}

### 2018-01-10 mrds/R/integratelogistic.R

@@ -1,0 +1,8 @@
+integratelogistic <- function (x, models, beta, lower=0, width, point){
+  # computes integral (numerically) over x from 0 to width of a
+  # logistic detection function
+
+  integrate(logisticbyx, lower=lower, upper=width,
+            subdivisions=10, rel.tol=0.01, abs.tol=0.01,
+            x=x, models=models, beta=beta, point=point)$value
+}

### 2018-01-10 mrds/R/integratelogisticdup.R

@@ -1,0 +1,29 @@
+integratelogisticdup <- function(x1, x2, models, beta, lower=0, width, point){
+  # numerical integral of product of logistic detection functions
+
+  # computation speed-up when there is distance in the formula
+  # but only if there are no interactions with distance
+  if(sum(grepl("distance",names(beta)))==1){
+
+    # set parameter to be zero for distance
+    beta_distance <- beta[grepl("distance",names(beta))]
+    beta[grepl("distance",names(beta))] <- 0
+
+    # calculate the rest of the linear predictor
+    x1 <- setcov(x1,models$g0model)%*%beta
+    x2 <- setcov(x2,models$g0model)%*%beta
+
+    # do some integration
+    integrate(logisticdupbyx_fast, lower=lower, upper=width,
+#              subdivisions=10, rel.tol=0.01, abs.tol=0.01,
+              rel.tol=0.01,
+              x1=x1, x2=x2, models=models, beta=beta, point=point,
+              beta_distance=beta_distance)$value
+  }else{
+  # Otherwise just go ahead and do the numerical integration
+    integrate(logisticdupbyx, lower=lower, upper=width,
+#              subdivisions=10, rel.tol=0.01, abs.tol=0.01,
+              rel.tol=0.01,
+              x1=x1, x2=x2, models=models, beta=beta, point=point)$value
+  }
+}

### 2018-01-10 mrds/R/integratepdf.r

@@ -1,0 +1,110 @@
+#' Numerically integrate pdf of observed distances over specified ranges
+#'
+#' Computes integral of pdf of observed distances over x for each observation.
+#' The method of computation depends on argument switches set and the type of
+#' detection function.
+#'
+#' @param ddfobj distance detection function specification
+#' @param select logical vector for selection of data values
+#' @param width truncation width
+#' @param left left truncation width
+#' @param int.range integration range matrix; vector is converted to matrix
+#' @param standardize logical used to decide whether to divide through by the function evaluated at 0
+#' @param point logical to determine if point count (\code{TRUE}) or line transect (\code{FALSE})
+#' @param doeachint calculate each integral numerically
+#' @return vector of integral values - one for each observation
+#' @author Jeff Laake & Dave Miller
+#' @keywords utility
+# @importFrom mgcv uniquecombs
+#' @importFrom stats integrate
+integratepdf <- function(ddfobj, select, width, int.range, standardize=TRUE,
+                         point=FALSE, left=0, doeachint=FALSE){
+  # Make sure there is consistency between integration ranges and data
+  # It is ok to have a single observation with multiple ranges or a single range
+  # with multiple observations but otherwise the numbers must agree if both >1
+
+  if(!is.matrix(int.range)){
+    if(is.vector(int.range) && length(int.range)==2){
+      int.range <- matrix(int.range, ncol=2, nrow=1)
+    }else{
+      stop("int.range is not a matrix and cannot be given the required matrix structure")
+    }
+  }
+
+  ## select tells us which data we compute the integrals for,
+  ## if it's null then we compute for all data in ddfobj$xmat
+  if(is.null(select)){
+    nobs <- nrow(ddfobj$xmat)
+    index <- 1:nobs
+  }else{
+    nobs <- sum(select)
+    index <- which(select)
+  }
+
+  # number of integration ranges must be 1 or match number of observations
+  # OR only be 1 observation and many ranges
+  if(nrow(int.range)>1 && nobs>1 && nrow(int.range)!=nobs){
+    stop("\n Number of integration ranges (int.range) does not match number of observations\n")
+  }
+
+  ## Now compute the integrals
+
+  # if there is only 1 integral to compute (no covariates/1 set of covariates
+  # & only one set of integration ranges), that's easy
[…]

### 2018-01-10 mrds/R/io.glm.R

@@ -1,0 +1,137 @@
+#' Iterative offset GLM/GAM for fitting detection function
+#'
+#' Provides an iterative algorithm for finding the MLEs of detection (capture)
+#' probabilities for a two-occasion (double observer) mark-recapture experiment
+#' using standard algorithms GLM/GAM and an offset to compensate for
+#' conditioning on the set of observations.  While the likelihood can be
+#' formulated and solved numerically, the use of GLM/GAM provides all of the
+#' available tools for fitting, predictions, plotting etc without any further
+#' development.
+#'
+#' Note that currently the code in this function for GAMs has been commented
+#' out until the remainder of the mrds package will work with GAMs.  This is an
+#' internal function that is used as by \code{ddf.io.fi} to fit mark-recapture
+#' models with 2 occasions.  The argument \code{mrmodel} is used for
+#' \code{fitformula}.
+#'
+#' @import mgcv
+#' @param datavec dataframe
+#' @param fitformula logit link formula
+#' @param eps convergence criterion
+#' @param iterlimit maximum number of iterations allowed
+#' @param GAM uses GAM instead of GLM for fitting
+#' @param gamplot set to TRUE to get a gam plot object if \code{GAM=TRUE}
+#' @return list of class("ioglm","glm","lm") or class("ioglm","gam")
+#'   \item{glmobj}{GLM or GAM object} \item{offsetvalue}{offsetvalues from
+#'   iterative fit} \item{plotobj}{gam plot object (if GAM & gamplot==TRUE,
+#'   else NULL)}
+#' @author Jeff Laake, David Borchers, Charles Paxton
+#' @references Buckland, S.T., J.M. breiwick, K.L. Cattanach, and J.L. Laake.
+#'   1993. Estimated population size of the California gray whale.  Marine
+#'   Mammal Science, 9:235-249.
+#'
+#' Burnham, K.P., S.T. Buckland, J.L. Laake, D.L. Borchers, T.A. Marques,
+#'   J.R.B. Bishop, and L. Thomas. 2004.  Further topics in distance sampling.
+#'   pp: 360-363. In: Advanced Distance Sampling, eds. S.T. Buckland,
+#'   D.R.Anderson, K.P. Burnham, J.L. Laake, D.L. Borchers, and L. Thomas.
+#'   Oxford University Press.
+#' @keywords Statistical Models
+#' @importFrom stats glm plogis binomial
[…]

### 2018-01-10 mrds/R/logisticdupbyx_fast.R

@@ -1,0 +1,29 @@
+#' Logistic for duplicates as a function of covariates (fast)
+#'
+#' As \code{\link{logisticdupbyx}}, but faster when distance is a covariate (but no interactions with distance occur.
+#'
+#' @inheritParams logisticdupbyx
+#' @param beta_distance parameter for distance
+#' @param x1 linear predictor for 1, without distance
+#' @param x2 linear predictor for 2, without distance
+#' @author David L Miller
+logisticdupbyx_fast <- function(distance, x1, x2, models, beta, point,
+                                beta_distance){
+
+  # function to calculate p/(1+p)
+  ologit <- function(p) p/(1+p)
+
+  # vectorise for R 3.4 silliness
+  x1 <- as.vector(x1)
+  x2 <- as.vector(x2)
+
+  # first part of the function
+  gx1 <- ologit(exp(x1 + distance*beta_distance))
+
+  # calculate second and return
+  if(!point){
+    return(gx1 * ologit(exp(x2 + distance*beta_distance)))
+  }else{
+    return(gx1 * ologit(exp(x2 + distance*beta_distance))*2*distance)
+  }
+}

### 2018-01-10 mrds/R/model.description.R

@@ -1,0 +1,28 @@
+model.description <- function(model){
+# builds the description of the model to be printed
+# only used internally.
+
+  key <- switch(model$key,
+                hn   = "Half-normal",
+                hr   = "Hazard-rate",
+                unif = "Uniform",
+                th1  = "Threshold 1",
+                th2  = "Threshold 2")
+
+  mod.str <- paste(key, "key function")
+  if(!is.null(model$adjustment)){
+    adj.series <- switch(model$adjustment$series,
+                         cos="cosine",
+                         herm="Hermite polynomial",
+                         poly="simple polynomial")
+    mod.str <- paste(mod.str,"with",adj.series,"adjustment term")
+
+    adj.order <- model$adjustment$order
+    if(length(adj.order)>1){
+      mod.str <- paste(mod.str,"s",sep="")
+    }
+    mod.str <- paste(mod.str,"of order",paste(adj.order,collapse=","))
+  }
+
+  return(mod.str)
+}

### 2018-01-10 mrds/R/nlminb_wrapper.R

@@ -1,0 +1,123 @@
+#' Wrapper around \code{nlminb}
+#'
+#' This is a wrapper around nlminb to use scaling, as this is not available (nor will it be) in \code{\link{optimx}}.
+#'
+#' @import optimx
+#' @inheritParams optimx::nlminb
+#'
+#' @param par starting parameters
+#' @param ll log likelihood function
+#' @param ugr gradent function
+#' @param lower lower bounds on parameters
+#' @param upper upper bounds on parameters
+#' @param mcontrol control options
+#' @param hess hessian function
+#' @param ddfobj detection function specification object
+#' @param data the data
+#' @param \dots anything else to pass to \code{ll}
+#' @return \code{optimx} object
+#'
+#' @importFrom stats nlminb
+#' @author David L Miller, modified from \code{optimx.run} by JC Nash, R Varadhan, G Grothendieck.
+nlminb_wrapper <- function(par, ll, ugr=NULL, lower=NULL, upper=NULL,
+                           mcontrol, hess=NULL, ddfobj, data, ...){
+
+  # need to do this from optimx.R
+  optcfg <- optimx.setup(par, ll, ugr, hess, lower, upper, "nlminb",
+                         itnmax=NULL, hessian=FALSE, mcontrol, ...)
+  mcontrol <- optcfg$ctrl
+
+  ## most comments here are from optimx.run, optimx version 2014.5.4
+  ## downloaded from r-forge
+
+  # from top of optimx.run:
+  # And make sure that controls NOT in method are deleted (nulled)
+  mcontrol$follow.on <- NULL
+  mcontrol$usenumDeriv <- NULL # JN130207
+  mcontrol$save.failures <- NULL
+  mcontrol$kkt <- NULL
+  mcontrol$starttests <- NULL
+  mcontrol$all.methods <- NULL
+  mcontrol$dowarn <- NULL
+  mcontrol$kkttol <- NULL
+  mcontrol$kkt2tol <- NULL
+  mcontrol$maximize <- NULL # Even if method DOES have it
+  mcontrol$badval <- NULL
+  mcontrol$scaletol <- NULL
+
+
+  # from optimx.run nlminb section:
+  # different name for iteration limit in this routine
+  mcontrol$iter.max <- mcontrol$maxit
+  mcontrol$maxit <- NULL
+  mcontrol$abs.tol <- 0 # To fix issues when minimum is less than 0. 20100711
[…]

### 2018-01-10 mrds/R/optimx.setup.R

@@ -1,0 +1,204 @@
+#' @importFrom numDeriv grad
+optimx.setup <- function(par, fn, gr=NULL, hess=NULL, lower=-Inf, upper=Inf, 
+            method=c("Nelder-Mead","BFGS"), itnmax=NULL, hessian=FALSE,
+            control=list(),
+             ...) {
+### To return in optcfg: fname, npar ??, method, ufn, ugr, ctrl, have.bounds
+
+# Get real name of function to be minimized
+  fname<-deparse(substitute(fn))
+  if (!is.null(control$trace) && control$trace>0) {
+	cat("fn is ",fname,"\n")
+  }
+  optcfg<-list()
+  optcfg$fname<-fname
+
+# Only one ref to parameters -- to get npar here
+  npar <- length(par) # !! NOT CHECKED in case par not well-defined
+  optcfg$npar <- npar
+
+# Set control defaults
+    ctrl <- list(
+	follow.on=FALSE, 
+	save.failures=TRUE,
+	trace=0,
+	kkt=TRUE,
+	all.methods=FALSE,
+	starttests=TRUE,
+	maximize=FALSE,
+	dowarn=TRUE, 
+        usenumDeriv=FALSE,
+	kkttol=0.001,
+	kkt2tol=1.0E-6,
+	badval=(0.5)*.Machine$double.xmax,
+	scaletol=3
+    ) 
+    
+# Note that we do NOT want to check on the names, because we may introduce 
+#    new names in the control lists of added methods
+#    if (!all(namc %in% names(ctrl))) 
+#        stop("unknown names in control: ", namc[!(namc %in% names(ctrl))])
+# However, we do want to substitute the appropriate information. 
+# removed copy of hessian to control$kkt
+    ncontrol <- names(control)
+    nctrl <- names(ctrl)
+    for (onename in ncontrol) {
+       if (onename %in% nctrl) {
+           ctrl[onename]<-control[onename]
+       } else {
+           ctrl[onename]<-control[onename]
+       }
+    }
+    if (is.null(control$kkt)) { # turn off kkt for large matrices
+      ctrl$kkt<-TRUE # default it to compute KKT tests
+      if (is.null(gr)) { # no analytic gradient
+         if (npar > 50) {
+           ctrl$kkt <- FALSE # too much work when large number of parameters
+           if (ctrl$trace>0) cat("gr NULL, npar > 50, kkt set FALSE\n")
+         }
+      } else {
+         if (npar > 500) {
[…]

### 2018-01-10 mrds/R/pdot.dsr.integrate.logistic.R

@@ -1,0 +1,143 @@
+#' Compute probability that a object was detected by at least one observer
+#'
+#' Computes probability that a object was detected by at least one observer
+#' (\code{pdot} or p_.) for a logistic detection function that contains
+#' distance.
+#'
+#' @param right either an integration range for binned data (vector of 2) or the rightmost value for integration (from 0 to right)
+#' @param width transect width
+#' @param beta parameters of logistic detection function
+#' @param x data matrix
+#' @param integral.numeric set to TRUE unless data are binned (done in this fct) or the model is such that distance is not linear (eg distance^2), If integral.numeric is FALSE it will compute the integral analytically. It should only be FALSE if is.linear.logistic function is TRUE.
+#' @param BT FALSE except for the trial configuration; BT stands for Buckland-Turnock who initially proposed a trial configuration for dual observers
+#' @param models list of models including \code{g0model}
+#' @param GAM Not used at present. The idea was to be able to use a GAM for g(0) portion of detection function; shoudl always be F
+#' @param rem only TRUE for the removal configuration but not used and could be removed if pulled from the function calls. Originally thought the pdot integral would differ but it is the same as the io formula. The only thing that differes with removal is that p(2|1)=1. Observer 2 sees everything seen by observer 1,
+#' @param point \code{TRUE} for point transects
+#'
+#' @author Jeff Laake
+pdot.dsr.integrate.logistic <- function(right, width, beta, x,
+                                        integral.numeric, BT, models, GAM=FALSE,
+                                        rem=FALSE, point=FALSE){
+  # Functions called:
+  # integratelogistic - computes integral (numerically) over x from 0 to width
+  #                     of a logistic detection function
+  # integratelogisticdup - computes integral (numerically) over x from 0 to
+  #                        width of a duplicate logistic detection function
+  # integratelogistic.analytic - computes integral (analytically) over x from 0
+  #                              to width of a logistic detection function
+  # logisticbyx - computes the detection probability with a logistic det fct
+  #               for a single set of z covariates but multiple values of x
+  # logisticdupbyx - computes the duplicate detection probability with a
+  #                  logistic det fct for a single set of z covariates but
+  #                  multiple values of x
[…]

### 2018-01-10 mrds/R/plot.det.tables.R

@@ -1,0 +1,90 @@
+#' Observation detection tables
+#'
+#' Plot the tables created by \code{\link{det.tables}}. Produces a series of
+#' tables for dual observer data that shows the number missed and detected for
+#' each observer within defined distance classes.
+#'
+#' @aliases plot.det.tables
+#' @export
+#' @param x object returned by \code{\link{det.tables}}
+#' @param which items in x to plot (vector with values in 1:6)
+#' @param angle shading angle for hatching
+#' @param density shading density for hatching
+#' @param col1 plotting colour for observer 1 detections
+#' @param col2 plotting colour for observer 2 detections within observer 1
+#'  subset detections
+#' @param new if TRUE new plotting window for each plot
+#' @param \dots other graphical parameters, passed to plotting functions
+#' @return Just plots.
+#' @author Jeff Laake
+#' @importFrom grDevices dev.new
+#' @importFrom graphics legend
+#' @examples
+#' \donttest{
+#' data(book.tee.data)
+#' region <- book.tee.data$book.tee.region
+#' egdata <- book.tee.data$book.tee.dataframe
+#' samples <- book.tee.data$book.tee.samples
+#' obs <- book.tee.data$book.tee.obs
+#' xx <- ddf(mrmodel=~glm(formula=~distance*observer),
+#'           dsmodel = ~mcds(key = "hn", formula = ~sex),
+#'           data = egdata, method = "io", meta.data = list(width = 4))
+#' tabs <- det.tables(xx,breaks=c(0,.5,1,2,3,4))
+#' par(mfrow=c(2,3))
+#' plot(tabs,which=1:6,new=FALSE)
+#' }
+plot.det.tables <- function(x,which=1:6,angle=-45,density=20,col1="black",
+                            col2="blue", new=TRUE,...){
+
+  # plotting function that actually does the work
+  plot_seen <- function(x,col1,col2,leg.title,...){
+    if(new& .Platform$GUI=="Rgui")dev.new()
+    missed <- x[,"Missed"]
+    detected <- x[,"Detected"]
+    ymax <- max(missed+detected)
+    histline(detected,breaks=breaks,lineonly=FALSE,ylim=c(0,ymax),
+             xlab="Distance",ylab="Frequency",angle=angle,
+             density=density,col=col2,...)
+    histline(missed+detected,breaks,lineonly=TRUE,col=col1,add=TRUE,
[…]

### 2018-01-10 mrds/R/plot.ds.R

@@ -1,0 +1,378 @@
+#' Plot fit of detection functions and histograms of data from distance sampling model
+#'
+#' Plots the fitted detection function(s) with a histogram of the observed distances to compare visually the fitted model and data.
+#'
+#' The structure of the histogram can be controlled by the user-defined arguments \code{nc} or \code{breaks}. The observation specific detection probabilities along with the line representing the fitted average detection probability.
+#'
+#' It is not intended for the user to call \code{plot.ds} but its arguments are documented here. Instead the generic \code{plot} command should be used and it will call the appropriate function based on the class of the \code{ddf} object.
+#'
+#' @aliases plot.ds
+#' @export
+#' @param x fitted model from \code{ddf}.
+#' @param which index to specify which plots should be produced:
+#'  \tabular{ll}{1 \tab histogram of observed distances\cr
+#'               2 \tab histogram of observed distanes with fitted line and points (default)\cr}
+#' @param breaks user defined breakpoints
+#' @param nc number of equal width bins for histogram
+#' @param jitter.v scaling option for plotting points.  Jitter is applied to points by multiplying the fitted value by a random draw from a normal distribution with mean 1 and sd \code{jitter.v[j]}.  Where \code{j=1,2} corresponds to observer \code{j} and \code{j=3} corresponds to pooled/duplicate detections.
+#' @param showpoints logical variable; if \code{TRUE} plots predicted value for each observation.
+#' @param subset subset of data to plot.
+#' @param pl.col colours plotting colours for obs 1, obs 2 detections.
+#' @param bw.col grayscale plotting colours for obs 1, obs 2 detections.
+#' @param black.white logical variable; if \code{TRUE} plots are grayscale.
+#' @param pl.den shading density for plots of obs 1, obs 2 detections.
+#' @param pl.ang shading angle for plots of obs 1, obs 2 detections.
+#' @param main user-specfied plot title.
+#' @param ylim user-specified y axis limits.
[…]

### 2018-01-10 mrds/R/plot_cond.R

@@ -1,0 +1,92 @@
+#' Plot conditional detection function from distance sampling model
+#'
+#' Plot proportion of observations detected within distance intervals (for
+#' conditional detection functions) to compare visually the fitted model and
+#' data. Internal function called by \code{plot} methods.
+#'
+#' @aliases plot_cond
+#' @param obs obsever code
+#' @param xmat processed data
+#' @param gxvalues detection function values for each observation
+#' @param model   fitted model from \code{ddf}
+#' @param nc number of equal-width bins for histogram
+#' @param breaks user define breakpoints
+#' @param finebr fine break values over which line is averaged
+#' @param showpoints logical variable; if \code{TRUE} plots predicted value
+#'   for each observation
+#' @param showlines logical variable; if \code{TRUE} plots average predicted
+#'   value line
+#' @param maintitle main title line for each plot
+#' @param ylim range of y axis (default \code{c(0,1)})
+#' @param angle shading angle for hatching
+#' @param density shading density for hatching
+#' @param col plotting colour
+#' @param jitter scaling option for plotting points.  Jitter is applied to
+#'   points by multiplying the fitted value by a random draw from a normal
+#'   distribution with mean 1 and sd jitter.
+#' @param xlab label for x-axis
+#' @param ylab label for y-axis
+#' @param subtitle if TRUE, shows plot type as sub-title
+#' @param \dots other graphical parameters, passed to the plotting functions
+#'   (\code{plot}, \code{hist}, \code{lines}, \code{points}, etc)
+#' @return NULL
+#' @author Jeff Laake, Jon Bishop, David Borchers
+#' @keywords plot
+plot_cond <- function(obs,xmat,gxvalues,model,nc,breaks,finebr,showpoints,
+                      showlines,maintitle,ylim,angle=-45,density=20,col="black",
+                      jitter=NULL,xlab="Distance",ylab="Detection probability",
+                      subtitle=TRUE,...){
+
+  # build and plot the histogram
+  selection <- xmat$detected[xmat$observer!=obs]==1
+  selmat <- (xmat[xmat$observer==obs,])[selection,]
+  shist <- hist(xmat$distance[xmat$observer!= obs & xmat$detected==1],
+                breaks=breaks, plot = FALSE)
[…]

### 2018-01-10 mrds/R/plot_uncond.R

@@ -1,0 +1,108 @@
+#' Plot unconditional detection function from distance sampling model
+#'
+#' Plots unconditional detection function for observer=obs observations
+#' overlays histrogram, average detection function and values for individual
+#' observations data. Internal function called by \code{plot} methods.
+#'
+#' @aliases plot_uncond
+#' @param model fitted model from \code{ddf}
+#' @param obs value of observer for plot
+#' @param xmat processed data
+#' @param gxvalues detection function values for each observation
+#' @param nc number of equal-width bins for histogram
+#' @param finebr fine break values over which line is averaged
+#' @param breaks user define breakpoints
+#' @param showpoints logical variable; if TRUE plots predicted value for each
+#'   observation
+#' @param showlines logical variable; if TRUE plots average predicted value line
+#' @param maintitle main title line for each plot
+#' @param ylim range of y axis; defaults to (0,1)
+#' @param return.lines if TRUE, returns values for line
+#' @param angle shading angle for hatching
+#' @param density shading density for hatching
+#' @param col plotting colour
+#' @param jitter scaling option for plotting points.  Jitter is applied to
+#'   points by multiplying the fitted value by a random draw from a normal
+#'   distribution with mean 1 and sd jitter.
+#' @param xlab label for x-axis
+#' @param ylab label for y-axis
+#' @param subtitle if TRUE, shows plot type as sub-title
+#' @param \dots other graphical parameters, passed to the plotting functions
+#'   (\code{plot}, \code{hist}, \code{lines}, \code{points}, etc)
+#' @return if \code{return.lines==TRUE} returns dataframe \code{average.line}
+#'  otherwise just plots
+#' @author Jeff Laake, Jon Bishop, David Borchers
+#' @keywords plot
+plot_uncond <- function(model,obs,xmat,gxvalues,nc,finebr,breaks,showpoints,
+                        showlines,maintitle,ylim,return.lines=FALSE,
+                        angle=-45,density=20,col="black",jitter=NULL,
+                        xlab="Distance",ylab="Detection probability",
+                        subtitle=TRUE,...){
+
+  #  Functions Used: average.line
+
[…]

### 2018-01-10 mrds/R/predict.ds.R

@@ -1,0 +1,204 @@
+#' Predictions from \code{mrds} models
+#'
+#' Predict detection probabilities (or effective strip widths/effective areas of detection) from a fitted distance sampling model using either the original data (i.e. "fitted" values) or using new data.
+#'
+#' The first 4 arguments are the same in each predict function.  The latter 2 are specific to certain functions. For line transects, the effective strip half-width (\code{esw=TRUE}) is the integral of the fitted detection function over either 0 to W or the specified \code{int.range}.  The predicted detection probability is the average probability which is simply the integral divided by the distance range. For point transect models, \code{esw=TRUE} calculates the effective area of detection (commonly referred to as "nu", this is the integral of \code{2/width^2 * rg(r)}.
+#'
+#' Fitted detection probabilities are stored in the \code{model} object and these are returned unless \code{compute=TRUE} or \code{newdata} is specified. \code{compute=TRUE} is used to estimate numerical derivatives for use in delta method approximations to the variance.
+#'
+#' For \code{method="io.fi"} or \code{method="trial.fi"} if \code{integrate=FALSE}, \code{predict} returns the value of the conditional detection probability and if \code{integrate=TRUE}, it returns the average conditional detection probability by integrating over x (distance) with respect to a uniform distribution.
+#'
+#' Note that the ordering of the returned results when no new data is supplied (the "fitted" values) will not necessarily be the same as the data supplied to \code{\link{ddf}}, the data (and hence results from \code{predict}) will be sorted by object ID (\code{object}) then observer ID (\code{observer}).
+#'
+#' @aliases predict predict.ds predict.ddf predict.io predict.io.fi predict.trial predict.trial.fi predict.rem predict.rem.fi
+#' @param object \code{ddf} model object.
+#' @param newdata new \code{data.frame} for prediction, this must include a column called "\code{distance}".
+#' @param compute if \code{TRUE} compute values and don't use the fitted values stored in the model object.
+#' @param int.range integration range for variable range analysis; either vector or 2 column matrix.
[…]

### 2018-01-10 mrds/R/predict.io.fi.R

@@ -1,0 +1,81 @@
+#' @export
+# see predict.ds for documentation
+predict.io.fi <- function(object ,newdata=NULL, compute=FALSE, int.range=NULL,
+                          integrate=FALSE, ...){
+  # Functions Used: pdot.dsr.integrate.logistic, is.linear.logistic,
+  #                 predict.glm (could also use predict.gam eventually)
+  model <- object
+  width <- model$meta.data$width
+  point <- model$meta.data$point
+
+  # if no new data supplied, use the data from the model
+  if(is.null(newdata)){
+    newdata <- model$mr$data
+  }else{
+    if(!("observer" %in% names(newdata))){
+      stop("newdata does not contain a column named \"observer\"")
+    }
+  }
+
+  newdata$offsetvalue <- 0
+
+  GAM <- FALSE
+  if("gam" %in% class(model$mr)){
+    GAM <- TRUE
+  }
+
+  # integrate=FALSE -- predict p(y)
+  if(!integrate){
+    # predict here is predict.glm
+    # since model here is model$mr, so model$mr is model$mr$mr
+    fitted <- predict(model$mr,newdata,type="response")
+
+    p1 <- fitted[newdata$observer==1]
+    p2 <- fitted[newdata$observer==2]
+    fitted <- p1+p2-p1*p2
+
+    names(fitted) <- newdata$object[newdata$observer==1]
+
+    return(list(fitted = fitted,
+                p1     = p1,
+                p2     = p2))
+
+  }else{
+    # integrate=TRUE -- get the average probability of detection
+    left <- model$meta.data$left
+    formula <- paste("~",as.character(model$mr$formula)[3],collapse="")
+
+    if("gam" %in% class(model$mr)){
+      integral.numeric <- TRUE
+    }else{
+      integral.numeric <- is.linear.logistic(newdata,formula,
+                                             length(coef(model$mr)),width)
+    }
+
+    models <- list(g0model        = formula,
+                   scalemodel     = NULL,
+                   fullscalemodel = NULL)
+
+    # now int.range is a vector with lower and upper bounds
+    if(is.null(int.range)){
+      pdot.list <- pdot.dsr.integrate.logistic(width, width, model$mr$coef,
+                     newdata, integral.numeric, FALSE, models,GAM, point=point)
+    }else{
+      pdot.list <- pdot.dsr.integrate.logistic(int.range,width, model$mr$coef,
+                      newdata, integral.numeric, FALSE, models,GAM, point=point)
[…]

### 2018-01-10 mrds/R/print.ddf.gof.R

@@ -1,0 +1,106 @@
+#' Prints results of goodness of fit tests for detection functions
+#'
+#' Provides formatted output for results of goodness of fit tests: chi-square,
+#' Kolmogorv-Smirnov and Cramer-von Mises test as approrpriate.
+#'
+#' @method print ddf.gof
+#' @param x result of call to \code{\link{ddf.gof}}
+#' @param \dots unused unspecified arguments for generic print
+#' @export
+#' @return None
+#' @author Jeff Laake
+#' @seealso \code{\link{ddf.gof}}
+#' @keywords utility
+print.ddf.gof <- function(x,...){
+
+  chitable <- function(observed, expected){
+    x <- rbind(observed, expected, (observed-expected)^2/expected)
+    x <- cbind(x, apply(x, 1, sum))
+    colnames(x)[dim(x)[2]] <- "Total"
+    rownames(x)=c("Observed", "Expected", "Chisquare")
+    return(x)
+  }
+
+  gof <- x
+  cat("\nGoodness of fit results for ddf object\n")
+
+  if(!is.null(gof$chisquare)){
+    cat("\nChi-square tests\n")
+
+    if(!is.null(gof$chisquare$chi2)){
+      cat("\nDistance sampling component:\n")
+    }
+
+    print(chitable(gof$chisquare$chi1$observed,gof$chisquare$chi1$expected))
+
+    if(!is.na(gof$chisquare$chi1$p)){
+      cat(paste("\nP = ",format(gof$chisquare$chi1$p,digits=5),
+                " with ",gof$chisquare$chi1$df," degrees of freedom\n",sep=""))
+    }else{
+      cat("\nNo degrees of freedom for test\n")
+    }
+
+    if(!is.null(gof$chisquare$chi2)){
+      if(dim(gof$chisquare$chi2$observed)[1]==3){
+        Trial <- FALSE
+      }else{
+        Trial <- TRUE
+      }
+
+      cat("\nMark-recapture component:\n")
+
+      if(Trial){
+        cat("Capture History 01\n")
+      }else{
+        cat("Capture History 10\n")
+      }
+
+      print(chitable(gof$chisquare$chi2$observed[1, ],
+                     gof$chisquare$chi2$expected[1, ]))
+
+      if(Trial){
+        cat("Capture History 11\n")
+      }else{
+        cat("Capture History 01\n")
+      }
+
+      print(chitable(gof$chisquare$chi2$observed[2, ],
[…]

### 2018-01-10 mrds/R/print.summary.ds.R

@@ -1,0 +1,75 @@
+#' Print summary of distance detection function model object
+#'
+#' Provides a brief summary of data and fitted detection probability model
+#' parameters, model selection criterion, and optionally abundance in the
+#' covered (sampled) region and its standard error. What is printed depends
+#' on the corresponding call to summary.
+#'
+#' @aliases print.summary.ds
+#' @export
+#' @param x a summary of \code{ddf} model object
+#' @param \dots unspecified and unused arguments for S3 consistency
+#' @return NULL
+#' @author Jeff Laake
+#' @seealso \code{\link{summary.ds}}
+#' @keywords utility
+print.summary.ds <- function (x,...){
+  cat("\nSummary for ds object \n")
+  cat("Number of observations : ", x$n,"\n")
+  if(x$int.range){
+    cat("Distance range         : ", x$left, " - ",x$width,"(int.range set)\n")
+  }else{
+    cat("Distance range         : ", x$left, " - ",x$width,"\n")
+  }
+  cat("AIC                    : ", x$aic, "\n")
+  cat("\nDetection function:\n",model.description(x),"\n")
+  cat("\nDetection function parameters", "\n")
+  cat("Scale coefficient(s): ", "\n")
+  print(x$coeff$key.scale)
+
+  if(x$key %in% c("gamma","hr","th1","th2")) {
+    cat("\nShape coefficient(s): ", "\n")
+    print(x$coeff$key.shape)
+  }
+
+  if (!is.null(x$coeff$adj.parm)) {
+     cat("\nAdjustment term coefficient(s): ", "\n")
+     print(x$coeff$adj.parm)
+  }
+
+  cat("\n")
+
+  if(x$mono & x$mono.strict){
+    cat("\nStrict monotonicity constraints were enforced.\n")
+  }else if(x$mono){
+    cat("\nMonotonicity constraints were enforced.\n")
+  }
+
+  if(!is.null(x$Nhat)){
+    parameters <- data.frame(Estimate=c(x$average.p,x$Nhat))
+    row.names(parameters) <- c("Average p", "N in covered region")
+    if(!is.null(x$average.p.se)){
+      parameters$SE <- c(x$average.p.se,x$Nhat.se)
+      parameters$CV <- parameters$SE/parameters$Estimate
+    }
+  }else{
+    parameters <- data.frame(Estimate=c(x$average.p))
+    row.names(parameters) <- c("Average p")
[…]

### 2018-01-10 mrds/R/print.summary.rem.fi.R

@@ -1,0 +1,45 @@
+#' Print summary of distance detection function model object
+#'
+#' Provides a brief summary of data and fitted detection probability model
+#' parameters, model selection criterion, and optionally abundance in the
+#' covered (sampled) region and its standard error. What is printed depends
+#' on the corresponding call to summary.
+#'
+#' @export
+#' @param x a summary of \code{ddf} model object
+#' @param \dots unspecified and unused arguments for S3 consistency
+#' @return NULL
+#' @author Jeff Laake
+#' @seealso \code{\link{summary.rem.fi}}
+#' @keywords utility
+print.summary.rem.fi <- function(x,...){
+
+  cat("\nSummary for rem.fi object \n")
+  cat("Number of observations               : ", x$n,"\n")
+  cat("Number seen by primary               : ", x$n1,"\n")
+  cat("Number additional seen by secondary  : ", x$n2-x$n3,"\n")
+  cat("AIC                                  : ", x$aic, "\n")
+
+  cat("\n\nConditional detection function parameters:\n")
+  print(x$cond.det.coef)
+
+  cat("\n")
+  if(!is.null(x$Nhat)){
+    parameters <- data.frame(Estimate=c(x$average.p, x$average.p0.1, x$Nhat))
+    row.names(parameters) <- c("Average p", "Average primary p(0)",
+                               "N in covered region")
+    if(!is.null(x$average.p.se)){
+      parameters$SE <- c(x$average.p.se, x$average.p0.1.se, x$Nhat.se)
+      parameters$CV <- parameters$SE/parameters$Estimate
+    }
+  }else{
+    parameters <- data.frame(Estimate=c(x$average.p0.1))
+    row.names(parameters) <- c("Average primary p(0)")
+    if(!is.null(x$average.p0.1.se)){
+        parameters$SE <- c(x$average.p0.1.se)
+        parameters$CV <- parameters$SE/parameters$Estimate
+    }
+  }
+  print(parameters)
+  invisible(NULL)
+}

### 2018-01-10 mrds/R/prob.deriv.R

@@ -1,0 +1,79 @@
+#' Derivatives for variance of average p and average p(0) variance
+#'
+#' Used in call to DeltaMethod from prob.se to get first derivatives
+#'
+#' Need to add equations here as I do not think they exist in any of the texts.
+#' These should probably be checked with simulation.
+#'
+#' @param par detection function parameter values
+#' @param model ddf model object
+#' @param parfct function of detection probabilities; currently only
+#'   average (over covariates) detection probability p integrated over distance
+#'   or average (over covariates) detection probability at distance 0; p(0)
+#' @param observer 1,2,3 for primary, secondary, or duplicates for average
+#'   p(0); passed to fct
+#' @param fittedmodel full fitted ddf model when \code{trial.fi} or
+#'   \code{io.fi} is called from \code{trial} or \code{io} respectively
+#' @return Vector of values from fct at specified parameter values
+#' @seealso prob.se
+#' @author Jeff Laake
+prob.deriv <- function(par,model,parfct,observer=NULL,fittedmodel=NULL){
+  # Depending on model method store new parameter values
+  # change 1/1/05 jll: used coefficients instead of coef;
+  #                    needed for trial.fi and io.fi
+
+  set.par <- function(model,par){
+    model$par <- par
+    if(model$method!="ds"){
+      if(model$method=="io" | model$method=="trial" | model$method=="rem"){
+        model$mr$mr$coefficients<-model$par[seq_along(model$mr$mr$coefficients)]
+        model$ds$par <- model$par[(length(model$mr$mr$coefficients)+1):
+                                   length(par)]
+        model$ds$ds$aux$ddfobj <-assign.par(model$ds$ds$aux$ddfobj,model$ds$par)
+      }else{
+        model$mr$coefficients <- model$par
+      }
+    }
+   return(model)
+  }
+
+  model <- set.par(model,par)
+  if(!is.null(observer)){
+    newdat <- model$mr$data
+    newdat$distance <- rep(0,length(newdat$distance))
+    newdat$offsetvalue <- 0
+    pred.at0 <- predict(model,newdat)
+  }else{
+    if(!is.null(fittedmodel)){
+      if(class(fittedmodel)[1]!="rem"){
+        newdat <- model$data[model$data$observer==1&model$data$detected==1,]
+      }else{
+        newdat <- model$data
+      }
+
[…]

### 2018-01-10 mrds/R/process.data.R

@@ -1,0 +1,202 @@
+#' Process data for fitting distance sampling detection function
+#'
+#' Sets up dataframe and does some basic error checking. Adds needed fields to
+#' dataframe and to \code{meta.data}.
+#'
+#' The function does a number of error checking tasks, creating fields and
+#' adding to \code{meta.data} including:
+#'
+#' 1) If \code{check=TRUE}, check to make sure the record structure is okay for
+#' mrds data. The number of primary records (observer=1) must equal the number
+#' of secondary records (observer=2). Also, a field in the dataframe is created
+#' \code{timesseen} which counts the number of times an object was detected
+#' 0,1,2; if \code{timesseen=0} then the record is tossed from the analysis.
+#' Also if there are differences in the data (distance, size, covariates) for
+#' observer 1 and 2 a warning is issued that the analysis may fail.  The code
+#' assumes these values are the same for both observers.
+#'
+#' 2) Based on the presence of fields \code{distbegin} and \code{distend}, a
+#' determination is made of whether the data analysis should be based on binned
+#' distances and a field \code{binned} is created, which is \code{TRUE} if the
+#' distance for the observation is binned.  By assigning for each observation
+#' this allows an analysis of a mixture of binned and unbinned distances.
+#'
+#' 4) Data are restricted such that distances are not greater than \code{width}
+#' and not less than \code{left} if those values are specified in
+#' \code{meta.data}.  If they are not specified then \code{left} defaults to 0
+#' and \code{width} defaults to the largest distance measurement.
+#'
+#' 5) Determine if an integration range (\code{int.begin} and \code{int.end}
+#' has been specified for the observations.  If it has, add the structure to
+#' \code{meta.data}.  The integration range is typically used for aerial
+#' surveys in which the altitude varies such that the strip width (left to
+#' width) changes with a change in altitude.
+#'
+#' 6) Fields defined as factors are cleaned up such that any unused levels are
+#' eliminated.
+#'
+#' 7) If the restrictions placed on the data, eliminated all of the data, the
+#' function stops with an error message
+#'
+#' @param data dataframe object
+#' @param meta.data meta.data options; see \code{\link{ddf}} for a description
+#' @param check if \code{TRUE} check data for errors in the mrds structure; for
[…]

### 2018-01-10 mrds/R/rem.glm.R

@@ -1,0 +1,139 @@
+#' Iterative offset model fitting of mark-recapture with removal model
+#' 
+#' Detection function fitting from mark-recapture data with a removal
+#' configuration in which a secondary observer knows what the primary observer
+#' detects and detects objects missed by the primary observer.  The iterative
+#' offset glm/gam uses an offset to compensate for the conditioning on the set
+#' of objects seen by either observer (eg 00 those missed by both observers are
+#' not included in the analysis.  This function is similar to
+#' \code{\link{io.glm}}.
+#' 
+#' The only difference between this function and \code{\link{io.glm}} is the
+#' offset and the data construction because there is only one detection
+#' function being estimated for the primary observer. The two functions could
+#' be merged.
+#' 
+#' @param datavec dataframe containing records seen by either observer 1 or 2
+#' @param fitformula logit link formula
+#' @param eps convergence criterion
+#' @param iterlimit maximum number of iterations allowed
+#' @param GAM uses GAM instead of GLM for fitting
+#' @param gamplot set to TRUE to get a gam plot object if \code{GAM=TRUE}
+#' @param datavec2 dataframe containing all records for observer 1 and observer
+#'   2 as in io.glm form; this is used in case there is an observer(not
+#'   platform effect)
+#' @return list of class("remglm","glm","lm") or class("remglm","gam")
+#'   \item{glmobj}{GLM or GAM object} \item{offsetvalue}{offsetvalues from
+#'   iterative fit} \item{plotobj}{gam plot object (if GAM & gamplot==TRUE,
+#'   else NULL)}
+#' @note currently the code in this function for GAMs has been commented out
+#'   until the remainder of the mrds package will work with GAMs.
+#' @author Jeff Laake
+#' @references Buckland, S.T., J.M. breiwick, K.L. Cattanach, and J.L. Laake.
+#'   1993. Estimated population size of the California gray whale.  Marine
+#'   Mammal Science, 9:235-249.
+#' 
+#' Burnham, K.P., S.T. Buckland, J.L. Laake, D.L. Borchers, T.A. Marques,
+#'   J.R.B. Bishop, and L. Thomas. 2004.  Further topics in distance sampling.
+#'   pp: 360-363. In: Advanced Distance Sampling, eds. S.T. Buckland,
[…]

### 2018-01-10 mrds/R/rescale_pars.R

@@ -1,0 +1,43 @@
+#' Calculate the parameter rescaling for parameters associated with covariates
+#'
+#' This will calculate the rescaling needed when covariates to be included in the scale of the detection function are "too big". Based on code from \code{\link{optimx}}.
+#'
+#' Derivative-free methods like nlminb are sensitive to the parameters being poorly scaled. This can also cause problems for quasi-Newton methods too (at least, bad scaling won't _help_ the optimisation). So here we rescale the parameters if necessary (unless we already got scaling from control)
+#'
+#' @author David L Miller
+#' @param initialvalues starting values for the optimisation
+#' @param ddfobj detection function object
+#' @importFrom stats sd terms
+rescale_pars <- function(initialvalues, ddfobj){
+
+  par_scaling <- rep(1, length(initialvalues))
+
+  # from optimx:::optimx.setup, scaletol = 3 (so setting 3 here for consistency)
+  # here we use a local copy in scalecheck.R
+  if(scalecheck(initialvalues, NA, NA, dowarn=FALSE)$lpratio > 3){
+    # do the rescaling to the scale parameters only
+    # this is (still) a bit hackish
+    # divide by the standard deviation of the distances
+    #  -- this seems appropriate due to the form of the detection function
+    #  may not be appropriate in all settings?
+
+    # match parameter vector indices to indices of model matrix
+    ind <- getpar(ddfobj, index=TRUE)
+    ind <- (ind[1]+1):ind[2]
+
+    # find the scalings
+    par_scaling[ind] <- apply(ddfobj$scale$dm, 2, sd)/sd(ddfobj$xmat$distance)
+
+    # ensure that the intercept has scaling 1 & any zero is set back to 1
+    par_scaling[abs(par_scaling) < sqrt(.Machine$double.eps)] <- 1
+
+    # set the factor scalings to be 1
+    non_factors <- !(colnames(ddfobj$scale$dm) %in%
+                     rownames(attr(terms(as.formula(ddfobj$scale$formula)),
+                                   "factors")))
+    par_scaling[ind][non_factors] <- 1
+  }
+
+  # return the parameter rescaling vector
+  return(par_scaling)
+}

### 2018-01-10 mrds/R/sample_ddf.R

@@ -1,0 +1,198 @@
+#' Generate data from a fitted detection function and refit the model
+#'
+#' @param ds.object a fitted detection function object
+#'
+#' @note This function changes the random number generator seed. To avoid any potential side-effects, use something like: \code{seed <- get(".Random.seed",envir=.GlobalEnv)} before running code and \code{assign(".Random.seed",seed,envir=.GlobalEnv)} after.
+#' @author David L. Miller
+#' @importFrom stats runif rnorm fitted
+sample_ddf <- function(ds.object){
+
+  # this would make life a lot easier...
+  if(!inherits(ds.object, "ds")){
+    stop("Can only sample from ds models")
+  }
+
+  # how many samples do we need?
+  n.ds.samples <- nrow(ds.object$data)
+
+  # how many samples do we have so far?
+  n.samps <- 0
+  dists <- c()
+
+  # make sure that a model gets fitted
+  dud.df <- TRUE
+
+  covars <- FALSE
+  direct_sim <- FALSE
+  # can we directly simulate?
+  if(ds.object$ds$aux$ddfobj$type=="hn" &
+     is.null(ds.object$ds$aux$ddfobj$adjustment$parameters) &
+     ds.object$ds$aux$ddfobj$scale$formula=="~1" &
+     !ds.object$meta.data$point){
+    direct_sim <- TRUE
+  }
+  # do we have a covariate model?
+  if(ds.object$ds$aux$ddfobj$scale$formula != "~1"){
+    covars <- TRUE
+    formula <- as.formula(ds.object$ds$aux$ddfobj$scale$formula)
+    model_xmat <- ds.object$ds$aux$ddfobj$xmat
+  }
+
+  # create an object to hold the parameters
+  pars <- list()
+  pars$scale <- ds.object$ds$aux$ddfobj$scale$parameters
+  if(!is.null(ds.object$ds$aux$ddfobj$shape$parameters)){
+    pars$shape <- ds.object$ds$aux$ddfobj$shape$parameters
+  }
+  if(!is.null(ds.object$ds$aux$ddfobj$adjustment$parameters)){
+    pars$adjustment <- ds.object$ds$aux$ddfobj$adjustment$parameters
+  }
+
+
+  # okay now do the simulation
+  while(dud.df){
+    if(direct_sim){
+      # if we just have a half-normal key function then we can
+      # directly simulate...
+      while(length(dists) < n.ds.samples){
[…]

### 2018-01-10 mrds/R/setcov.R

@@ -1,0 +1,36 @@
+#' Creates design matrix for covariates in detection function
+#'
+#' This function creates a design matrix for the g(0) or scale covariates using
+#' the input model formula. It returns a list which contains 2 elements: 1)
+#' dim: the dimension (number of columns) of the design matrix, and 2) cov: the
+#' constructed design matrix. This function is relatively simple because it
+#' uses the built-in function \code{\link{model.matrix}} which does the
+#' majority of the work.  This function handles 2 exceptions "~.", the null
+#' model with 0 columns and "~1" the intercept only model - a column of 1s.  If
+#' a model other than the 2 exceptions is provided, it calls
+#' \code{\link{model.matrix}} to construct the columns. If any of the colums of
+#' the design matrix are all 0's the column is removed.  This occurs when there
+#' is no data for a particular factor.
+#'
+#'
+#' @param dmat data matrix
+#' @param model model formula
+#' @return a design matrix for the specified data and model
+#' @author Jeff Laake
+#' @keywords utility
+setcov <- function(dmat, model){
+  # Null Model
+  if(model=="~."){
+    n <- 0
+    x <- NULL
+  # Intercept Model
+  }else if(model == "~1"){
+    n <- 1
+    x <- matrix(rep(1, nrow(dmat)), ncol=1)
+    colnames(x) <- "(Intercept)"
+  # Covariate Model
+  }else{
+    x <- model.matrix(as.formula(model), data = dmat)
+  }
+  return(x)
+}

### 2018-01-10 mrds/R/sethazard.R

@@ -1,0 +1,70 @@
+sethazard <- function(ddfobj, dmat, width, left=0){
+
+  if("distend" %in% names(ddfobj$xmat)){
+    binned <- TRUE
+  }else{
+    binned <- FALSE
+  }
+
+  # temporary function to evaluate the likelihood for hazard rate
+  evallike <- function(fpar, ddfobj, binned, left){
+    ddfobj$intercept.only <- TRUE
+    ddfobj$shape$parameters[1] <- fpar[1]
+    ddfobj$scale$parameters[1] <- fpar[2]
+    # set adjustment parameters to zero for now...
+    if(!is.null(ddfobj$adjustment)){
+      ddfobj$adjustment$parameters <- rep(0, length(ddfobj$adjustment$order))
+    }
+
+    fpar <- getpar(ddfobj)
+    ddfobj$shape$parameters[1] <- NA
+    ddfobj$scale$parameters[1] <- NA
+    if(!is.null(ddfobj$adjustment)){
+      ddfobj$adjustment$parameters <- rep(NA, length(ddfobj$adjustment$order))
+    }
+
+
+    flnl(fpar, ddfobj, FALSE,
+         misc.options=list(width=width, int.range=c(0, width), showit=0,
+                           integral.numeric=FALSE, standardize=FALSE,
+                           binned=binned, left=left,
+                           fitting="none", point=FALSE))
+  }
+
+  # Using code from CDS in Distance.
+  #  First find the scale parameter as you (CDS) would for half-normal,
+  #  by finding:
+  ss <- sqrt(sum(dmat$distance^2)/length(dmat$distance))
+
+  # Then find the shape parameter by using the following algorithm:
+  for(j in 1:10){
+    scale <- log(ss/2+j*0.2*ss)
+    for(i in seq(1, 30, 2)){
+      shape <- log(i)
+      # evaluate likelihood at this value
+      newlnl <- evallike(c(shape, scale), ddfobj, binned, left)
+      if(is.nan(newlnl)){
+        break
+      }
+      if(i==1 & j==1){
+        lnl <- newlnl
+        ival <- 2
+        jval <- 1
+      }else{
+        if(newlnl < lnl){
+          lnl <- newlnl
+          ival <- i
+          jval <- j
+        }
+      }
+    }
+  }
+  if(is.nan(newlnl)){
+    shape <- log(2)
+  }else{
+    scale <- log(ss/2+jval*0.2*ss)
[…]

### 2018-01-10 mrds/R/sim.mix.R

@@ -1,0 +1,47 @@
+#' Simulation of distance sampling data via mixture models
+
+#' Allows one to simulate line transect distance sampling data using 
+#' a mixture of half-normal detection functions.
+#'
+#' @param n number of samples to generate
+#' @param sigma vector of scale parameters
+#' @param mix.prop vector of mixture proportions (same length as sigma)
+#' @param width truncation
+#' @param means vector of means (used to generate wacky, non-monotonic data)
+#' @return distances a vector of distances
+#'
+#' @author David Lawrence Miller
+#'
+#' @note At the moment this is TOTALLY UNSUPPORTED!
+#'       Please don't use it for anything important!
+
+sim.mix<-function(n, sigma, mix.prop, width, means=0){
+
+   # TODO:
+   #  * covariates
+   #  * From SNW: "All following could be made more efficient by not
+   #               generating so many more than needed..."
+
+
+   # code from Simon Wood for direct simulation of mixtures
+
+   ## detection function is of form \sum_i p_i exp(-d^2/(2 sigma_i^2))
+   ## truncated at width, say. This is equivalent to 
+   ## \sum_i \alpha_i /\sqrt{2 \pi sigma_i^2} exp(-d^2/(2 \sigma_i^2))
+   ## i.e. to a mixture of half normals with mixture weights proportional to
+   ## \alpha_i = p_i \sqrt{2 \pi sigma_i^2}
+   ## the truncated version of the latter is easy to simulate from.
+
+
+   ## mixture method
+   alpha <- mix.prop*(sqrt(2*pi)*sigma) ## transform to mixture weights
+   alpha <- alpha/(sum(alpha)) ## normalize
+   ## sample mixture index
+   ii <- sample(1:length(mix.prop), size=2*n, replace=TRUE, prob=alpha)
+   d <- abs(rnorm(ii, means, sigma[ii])) ## sample from components
+   d <- d[d<=width]   ## reject beyond truncation
+   d <- d[1:n] ## ditch the excess
+
+   return(d)
+}
+

### 2018-01-10 mrds/R/solvecov.R

@@ -1,0 +1,29 @@
+# solvecov code was taken from package fpc: Christian
+# Hennig chrish@@stats.ucl.ac.uk http://www.homepages.ucl.ac.uk/~ucakche/
+solvecov <- function (m, cmax = 1e+10){
+  options(show.error.messages = FALSE)
+  covinv <- try(solve(m))
+  if(class(covinv) != "try-error"){
+    coll <- FALSE
+  }else{
+    p <- nrow(m)
+    cove <- eigen(m, symmetric = TRUE)
+    coll <- TRUE
+    if(min(cove$values) < 1/cmax) {
+      covewi <- diag(p)
+      for (i in 1:p){
+        if (cove$values[i] < 1/cmax){
+          covewi[i, i] <- cmax
+        }else{
+         covewi[i, i] <- 1/cove$values[i]
+        }
+      }
+    }else{
+      covewi <- diag(1/cove$values, nrow = length(cove$values))
+    }
+    covinv <- cove$vectors %*% covewi %*% t(cove$vectors)
+  }
+  options(show.error.messages = TRUE)
+  out <- list(inv = covinv, coll = coll)
+  return(out)
+}

### 2018-01-10 mrds/R/summary.ds.R

@@ -1,0 +1,119 @@
+#' Summary of distance detection function model object
+#'
+#' Provides a brief summary of data and fitted detection probability model
+#' parameters, model selection criterion, and optionally abundance in the
+#' covered (sampled) region and its standard error.
+#'
+#' The argument \code{N} is used to suppress computation of
+#' abundance and average detection probability in calls to summarize the
+#' \code{ds} and either \code{io.fi} or \code{trial.fi} for summaries of
+#' \code{io} and \code{trial} objects respectively which are composed of a
+#' \code{ds} model object and a mark-recapture model object. The corresponding
+#' print function is called to print the summary results.
+#'
+#' @export
+#' @param object a \code{ddf} model object
+#' @param se if TRUE, computes standard errors
+#' @param N if TRUE, computes abundance in covered (sampled) region
+#' @param \dots unspecified and unused arguments for S3 consistency
+#' @return list of extracted and summarized objects
+#' @note This function is called by the generic function \code{summary} for any
+#'   \code{ddf} model object.  Each function can be called directly by the
+#'   user, but it is typically safest to use the generic function
+#'   \code{summary} which calls the appropriate function based on the type of
+#'   \code{ddf} model.
+#' @author Jeff Laake
+#' @keywords utility
+summary.ds <- function(object, se=TRUE, N=TRUE, ...){
+  # Uses: predict.ds (via predict), DeltaMethod, coef.ds (via coef)
+
+  # at present f(0) code is commented out; need to understand this further
+  #F0=function(model,pdot,...)
+  #{return(1/(pdot*model$meta.data$width))}
+
+  model <- object
+  avgp <- function(model, pdot, ...){return(pdot)}
+
+  ddfobj <- model$ds$aux$ddfobj
+
+  ans <- list()
+
+  # was monotonicity enforced?
+  ans$mono<-model$ds$aux$mono
+  # strict monotonicity?
+  ans$mono.strict<-model$ds$aux$mono.strict
+
+  # Number of observations
+  ans$n <- length(ddfobj$xmat$distance)
+
+  ## Average detection prob for mcds
+  #if(is.null(model$fitted)){
+  #  pdot <- predict(model,esw=FALSE)$fitted
+  #}else{
[…]

### 2018-01-10 mrds/R/varn.R

@@ -1,0 +1,181 @@
+#' Compute empirical variance of encounter rate
+#'
+#' Computes one of a series of possible variance estimates for the observed encounter rate for a set of sample measurements (e.g., line lengths) and number of observations per sample.
+#'
+#' The choice of type follows the notation of Fewster et al. (2009) in that there are 8 choices of encounter rate variance that can be computed for lines and one for points:
+#'
+#' \describe{
+#' \item{\code{R2}}{random line placement with unequal line lengths (design-assisted estimator)}
+#' \item{\code{R3}}{random line placement, model-assisted estimator, based on true contagion process}
+#' \item{\code{R4}}{random line placement, model-assisted estimator, based on apparent contagion process}
+#' \item{\code{S1}}{systematic line placement, post-stratification with no strata overlap}
+#' \item{\code{S2}}{systematic line placement, post-stratification with no strata overlap, variances weighted by line length per stratum}
+#' \item{\code{O1}}{systematic line placement, post-stratification with overlapping strata (akin to S1)}
+#' \item{\code{O2}}{systematic line placement, post-stratification with overlapping strata (weighted by line length per stratum, akin to S2)}
+#' \item{\code{O3}}{systematic line placement, post-stratification with overlapping strata, model-assisted estimator with trend in encounter rate with line length}
+#' \item{\code{P3}}{random point placement, potentially unequal number of visitsper point, model-based estimator}
+#' }
+#'
+#' Default value is \code{"R2"}, shown in Fewster et al. (2009) to have good performance for completely random designs for lines. For systematic parallel line transect designs, Fewster et al. recommend \code{"O2"}. For point transects the default (and currently only implemented option) is \code{"P3"}.
+#'
[…]

### 2018-01-10 mrds/tests/testthat.R

@@ -1,0 +1,6 @@
+library(testthat)
+library(mrds)
+
+test_check("mrds")
+
+

### 2018-01-10 mrds/tests/testthat/test_easy.R

@@ -1,0 +1,304 @@
+library(mrds)
+
+lnl.tol<-1e-3#4
+par.tol<-1e-6
+
+context("easy tests: golf tees")
+# taken from Laura Marshall's RUnit tests
+
+test_that("golf tee data gives the same results as Distance",{
+
+  data(book.tee.data)
+  egdata<-book.tee.data$book.tee.dataframe
+
+
+  #CDS
+  result.cds <- ddf(dsmodel=~cds(key="hn"), mrmodel=~glm(~distance),
+                    data=egdata, method="io", meta.data=list(width=4))
+  expect_that(result.cds$Nhat, equals(232.0015, tolerance=1e-6))
+
+  #GLM
+  result.glm <- ddf(mrmodel=~glm(~distance), data=egdata, method="io.fi",
+                  meta.data=list(width=4))
+  expect_that(result.glm$Nhat, equals(186.0947, tolerance=1e-6))
+
+  #MCDS
+  #checkException(ddf(dsmodel = ~mcds(key = "hn", formula = ~1), data = newdata, method = "ds", meta.data = list(width = 4)))
+
+  result.mcds <- ddf(dsmodel=~mcds(key="hn", formula=~1),
+                   data=egdata[egdata$observer==1, ], method="ds",
+                   meta.data=list(width=4))
+  expect_that(result.mcds$Nhat, equals(212.229, tolerance=1e-3))
+
+  #rr <- ddf(dsmodel = ~mcds(key = "hn", formula = ~sex),
+  #          data=subset(egdata, detected==1 & observer==1), method="ds",
+  #          meta.data=list(width=4))
+
+#  # check that uniform key works
+#  result.unif<-ddf(dsmodel = ~cds(key = "unif",adj.series="cos",adj.order=2), 
+#                  data = egdata, method = "ds", meta.data = list(width = 4))
+#  expect_that(result.unif$par, equals(0.7005934,tolerance=1e-6))
+  
+# there should be an error if we don't supply adjustments with uniform key
+# This was removed because it now works
+#  expect_error(ddf(dsmodel = ~cds(key = "unif"), 
+#                  data = egdata, method = "ds", meta.data = list(width = 4)))
+
+})
+
+context("errors thrown for invalid models")
[…]

### 2018-01-10 mrds/tests/testthat/test_gamma.R

@@ -1,0 +1,50 @@
+library(mrds)
+library(testthat)
+
+par.tol<-1e-3
+
+context("gamma detection function")
+
+test_that("standardisation is done correctly",{
+
+  # generate some gamma data where the mode is basically 0
+  set.seed(123)
+  gdata<-data.frame(object=1:1000,distance=rgamma(1000,scale=1,shape=1))
+
+  # this will complain about first partial hessian being singular
+  # suppress that
+  mm <- suppressWarnings(ddf(dsmodel=~cds(key="gamma"), data=gdata, method="ds",
+                             meta.data=list(width=max(gdata$distance))))
+
+  # check fitted parameters
+  parcomp <- mm$par
+  names(parcomp) <- NULL
+  expect_equal(c(-19.097429433, -0.001956123),parcomp,tol=par.tol)
+
+
+  # now with adjustments
+  # again this will complain about first partial hessian being singular
+  # suppress that
+  mm <- suppressWarnings(ddf(dsmodel=~cds(key="gamma",adj.series="cos",
+                                          adj.order=2),
+                             data=gdata, method="ds",
+                             meta.data=list(width=max(gdata$distance))))
+
+  # check fitted parameters
+  parcomp <- mm$par
+  names(parcomp) <- NULL
+  expect_equal(c(-21.6765623,0.1103963, 0.2100172),parcomp,tol=par.tol)
+
+  # generate some gamma data where the mode is away from 0
+  set.seed(123)
+  gdata<-data.frame(object=1:1000,distance=rgamma(1000,scale=2,shape=3))
+
+  mm <- ddf(dsmodel=~cds(key="gamma"), data=gdata, method="ds",
+            meta.data=list(width=max(gdata$distance)))
+
+  # check fitted parameters
+  parcomp <- mm$par
+  names(parcomp) <- NULL
+  expect_equal(c(0.7517943,1.9497688),parcomp,tol=par.tol)
+
+})

### 2018-01-10 mrds/tests/testthat/test_nosightings.R

@@ -1,0 +1,43 @@
+library(mrds)
+library(testthat)
+
+context("No sightings")
+
+test_that("a dht object with only 0's for estimates is returned", {
+
+  #datasetup
+  ex.filename<-system.file("testData/NoSightings/clusters.robj", package="mrds")
+  load(ex.filename)
+  ex.filename<-system.file("testData/NoSightings/no_clusters.robj", package="mrds")
+  load(ex.filename)
+  ex.filename<-system.file("testData/NoSightings/obs_subset.robj", package="mrds")
+  load(ex.filename)
+  ex.filename<-system.file("testData/NoSightings/region_table.robj", package="mrds")
+  load(ex.filename)
+  ex.filename<-system.file("testData/NoSightings/sample_table.robj", package="mrds")
+  load(ex.filename)
+  
+  #run analyses
+  ddf.clusters <- ddf(method='ds', dsmodel = ~mcds(key = 'hn', formula = ~scaledtotsize), data = clusters, meta.data = list(width = 5.5))
+  ddf.no.clusters <- ddf(method='ds', dsmodel = ~mcds(key = 'hn', formula = ~scaledtotsize), data = no.clusters, meta.data = list(width = 5.5))
+  cluster.result <- dht(ddf.clusters, region.table, sample.table, obs.table.subset)
+  no.cluster.result <- dht(ddf.no.clusters, region.table, sample.table, obs.table.subset)
+
+  #run tests
+  #checks there are NO results for clusters
+  expect_that(is.null(no.cluster.result$clusters), equals(TRUE))
+  #checks there ARE results for clusters
+  expect_that(is.null(cluster.result$clusters), equals(FALSE))
+  #checks all abundance estimates are 0 for individuals in clustered data
+  expect_that(length(which(cluster.result$individuals$N$Estimate != 0)), equals(0))
+  #checks all df are 0 for clusters for clustered data; removed I don't think this is useful nor desirable
+  #expect_that(length(which(cluster.result$clusters$N$df != 0)), equals(0))
+  #checks all encounter rate estimates are 0 for clusters for clustered data
+  expect_that(length(which(cluster.result$clusters$summary$ER != 0)), equals(0))
+  #checks all density estimates are 0 for clusters for non-clustered data
+  expect_that(length(which(no.cluster.result$clusters$D$se != 0)), equals(0))
+  #checks the region areas match for the cluster and individual summaries for clustered data
[…]

### 2018-01-10 mrds/tests/testthat/test_plotting.R

@@ -1,0 +1,132 @@
+# testing for plotting functions
+
+library(testthat)
+library(mrds)
+
+line.tol <- 1e-4
+
+context("Plotting tests: average.line")
+
+test_that("average.line for plot.io is correct (unconditional df)",{
+
+  # load some data as usual
+  data(book.tee.data)
+  egdata <- book.tee.data$book.tee.dataframe
+
+  # fit the model
+  result <- ddf(dsmodel = ~cds(key = "hn"), mrmodel = ~glm(~distance),
+                data = egdata, method = "io", meta.data = list(width = 4))
+
+  # make the grid (as in plot.io)
+  divisions <- 25 # default in plot.io
+  finebr <- (4/divisions)*(0:divisions)
+
+  ## observer 1
+  # make the line
+  aline <- mrds:::average.line(finebr,obs=1,result)
+  # "true" values
+  aline.truth <- c(0.942984310350384,0.931758960736299,0.916033134386296,
+                   0.895960681152277,0.871766127423631,0.843743459341331,
+                   0.812252811904826,0.777714776947772,0.740602126839018,
+                   0.701428901443524,0.660737024685639,0.619080889235704,
+                   0.577010639634571,0.535055145122935,0.493705825337338,
+                   0.453402523523292,0.414522484905807,0.377373199698558,
+                   0.342189456128584,0.309134492201619,0.278304718215714,
+                   0.249737174975323,0.223418735462668,0.199296054654547,
+                   0.177285396274316)
+  expect_that(aline$value,equals(aline.truth,tol=line.tol))
+
+  # obs == 3 (duplicates)
+  aline <- mrds:::average.line(finebr,obs=3,result)
+  aline.truth <- c(0.996674615155133,0.992654180180232,0.984661898293878,
+                   0.972793968270969,0.957192272380195,0.938041553280893,
+                   0.915565773567083,0.890023758238941,0.861704238583432,
+                   0.830920430012759,0.79800428592961,0.763300574439644,
+                   0.727160924687131,0.689937984912215,0.651979825357745,
+                   0.613624706385435,0.575196316218169,0.536999564325613,
+                   0.49931699639121,0.462405875840549,0.426495955861535,
+                   0.391787945447792,0.35845265391595,0.326630781154814,
+                   0.29643330601488)
+  expect_that(aline$value,equals(aline.truth,tol=line.tol))
+
+  # obs == 4 (pooled)
+  aline <- mrds:::average.line(finebr,obs=4,result)
[…]

### 2018-01-10 mrds/tests/testthat/test_pt.R

@@ -1,0 +1,142 @@
+# point transect testing
+
+library(mrds)
+library(testthat)
+
+par.tol <- 1e-5
+lnl.tol <- 1e-4
+
+context("Point transects: ptexample")
+
+test_that("Point transect example from distance gives same results",{
+
+  data(ptdata.distance)
+  xx <- ddf(dsmodel=~cds(key="hn", formula=~1), data=ptdata.distance,
+            method="ds",
+            meta.data=list(point=TRUE,width=max(ptdata.distance$distance)))#,
+            #control=list(upperbounds=Inf,lowerbounds=-Inf,showit=3))
+            #control=list(optimx.maxit=10000,showit=3))
+
+  expect_that(xx$par,equals(2.283007,tol=par.tol))
+  expect_that(xx$lnl,equals(-458.5701,tol=lnl.tol))
+  expect_that(summary(xx)$average.p,equals(0.1644288,tol=par.tol))
+
+
+})
+
+
+context("Point transects: simulated data")
+
+test_that("Nothing has changed?", {
+
+  # thanks to Tiago Marques for this example
+  set.seed(123)
+  n <- 1000
+  w <- 8000
+  sigma <- 3000
+  xs <- runif(n,-w,w)
+  ys <- runif(n,-w,w)
+  rs <- sqrt(xs^2+ys^2)
+  gx <- function(x,sigma){exp(-(x/sigma)^2)}
+  px <- gx(rs,sigma)
+  pdet <- runif(n,0,1)
+  det <- px>pdet
+  data4D <- data.frame(distance=rs[det==1])
+  data4D$object <- 1:nrow(data4D)
+  data4D$detected <- rep(1, nrow(data4D))
+
+
+  # half-normal
+  xx <- ddf(dsmodel=~cds(key="hn", formula=~1), data=data4D,
+            method="ds",
+            meta.data=list(point=TRUE, width=8000))
+
+  expect_that(xx$par, equals(7.6438325, tol=par.tol))
+  expect_that(xx$lnl, equals(-1066.120945, tol=lnl.tol))
+  expect_that(summary(xx)$average.p, equals(0.1361185, tol=par.tol))
+
+
+  # hazard-rate
+  xx <- ddf(dsmodel=~cds(key="hr", formula=~1), data=data4D,
+            method="ds",
+            meta.data=list(point=TRUE, width=8000))
+
[…]

### 2018-01-10 mrds/tests/testthat/test_summary.R

@@ -1,0 +1,85 @@
+# testing for summary.* functions
+
+library(testthat)
+library(mrds)
+
+tol <- 1e-6
+
+context("Summary tests")
+
+test_that("summary results are correct",{
+
+  # run the non-expect_that() code then the commented section below
+  #  to "update" results
+  #pt.result.summ<-summary(pt.result)
+  #tee.result.summ<-summary(tee.result)
+  #tee.result.trial.summ<-summary(tee.result.trial)
+  #tee.result.trial.fi.summ<-summary(tee.result.trial.fi)
+  #tee.result.rem.summ<-summary(tee.result.rem)
+  #save(pt.result.summ,tee.result.summ,tee.result.trial.summ,
+  #     tee.result.trial.fi.summ,tee.result.rem.summ,
+  #     file="mrds/inst/testData/summaryresults.rda")
+
+  ### load the results (summary objects)
+  res.filename<-system.file("testData/summaryresults.rda", package="mrds")
+  load(res.filename)
+
+
+  ### io method -- line transects
+  # load some data as usual
+  data(book.tee.data)
+  egdata <- book.tee.data$book.tee.dataframe
+
+  # fit the model
+  tee.result <- ddf(dsmodel = ~cds(key = "hn"), mrmodel = ~glm(~distance),
+                data = egdata, method = "io", meta.data = list(width = 4))
+
+  expect_that(summary(tee.result)$mr.summary,equals(tee.result.summ$mr.summary))
+  expect_that(summary(tee.result)$ds.summary,equals(tee.result.summ$ds.summary))
+  expect_that(summary(tee.result),equals(tee.result.summ))
+
+  ### trial.fi method -- line transect
+  tee.result.trial.fi <- ddf(mrmodel=~glm(formula=~distance),
+                             dsmodel = ~mcds(key = "hn", formula = ~sex),
+                             data=egdata, method="trial.fi",
+                             meta.data=list(width=4))
+  expect_that(summary(tee.result.trial.fi),equals(tee.result.trial.fi.summ))
+
+  #### trial method -- line transect
+  tee.result.trial <- ddf(mrmodel=~glm(formula=~distance),
+                          dsmodel = ~mcds(key = "hn", formula = ~sex),
+                          data=egdata, method="trial",
+                          meta.data=list(width=4))
+  expect_that(summary(tee.result.trial),equals(tee.result.trial.summ))
+
+
+  ### rem method -- line transect
[…]

### 2018-01-10 mrds/tests/testthat/test_trial.R

@@ -1,0 +1,112 @@
+# bugs in trial and trial.io go here...
+
+library(testthat)
+library(mrds)
+
+context("trial and trial.fi")
+
+# bug from Louise Burt 31/10/11
+test_that("detected and missed are the same in summary and det.tables", {
+
+   # bug appeared because there were zeros in the data but when using cut()
+   # you must specify include.lowest=TRUE for them to be included.
+
+   # grab some data
+   data(book.tee.data)
+   egdata<<-book.tee.data$book.tee.dataframe
+
+
+   # no zeros, just check it works anyway
+   xx<-ddf(mrmodel=~glm(formula=~distance),
+           dsmodel = ~mcds(key = "hn", formula = ~sex),
+           data = egdata, method = "trial.fi", meta.data = list(width = 4))
+
+   tabs<-det.tables(xx,breaks=c(0,.5,1,2,3,4))
+   summ<-summary(xx)
+
+   # the sums of the columns in tabs should match the summary numbers
+   expect_that(sum(tabs$Observer1[,2]), equals(summ$n1))
+   expect_that(sum(tabs$Observer2[,2]), equals(summ$n2))
+   expect_that(sum(tabs$Duplicates), equals(summ$n3))
+
+
+   # there are no zeros, so fix that and re-run the model
+   egdata$distance[egdata$distance==0.02]<-0
+   xx<-ddf(mrmodel=~glm(formula=~distance),
+           dsmodel = ~mcds(key = "hn", formula = ~sex),
+           data = egdata, method = "trial.fi", meta.data = list(width = 4))
+   tabs<-det.tables(xx,breaks=c(0,.5,1,2,3,4))
+   summ<-summary(xx)
+   # test the bug
+   expect_that(sum(tabs$Observer1[,2]), equals(summ$n1))
+   expect_that(sum(tabs$Observer2[,2]), equals(summ$n2))
+   expect_that(sum(tabs$Duplicates), equals(summ$n3))
+
+
+})
+
+# test results from golf tee data
+test_that("results from golf tee data (trial.fi mode) work (distance covar)", {
+
+  data(book.tee.data)
+  # extract the list elements from the dataset into easy-to-use objects
+  detections <- book.tee.data$book.tee.dataframe
+  # make sure sex and exposure are factor variables
+  detections$sex <- as.factor(detections$sex)
+  detections$exposure <- as.factor(detections$exposure)
+  region <- book.tee.data$book.tee.region
[…]
